### 2019-06-16 project/overleaf/paper.tex
Paper v1
Overleaf project and Octave scripts

% mnras_template.tex 
%
% LaTeX template for creating an MNRAS paper
%
% v3.0 released 14 May 2015
% (version numbers match those of mnras.cls)
%
% Copyright (C) Royal Astronomical Society 2015
% Authors:
% Keith T. Smith (Royal Astronomical Society)

% Change log
%
% v3.0 May 2015
%    Renamed to match the new package name
%    Version number matches mnras.cls
%    A few minor tweaks to wording
% v1.0 September 2013
%    Beta testing only - never publicly released
%    First version: a simple (ish) template for creating an MNRAS paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[fleqn,usenatbib]{mnras}

% MNRAS is set in Times font. If you don't have this installed (most LaTeX
% installations will be fine) or prefer the old Computer Modern fonts, comment
% out the following line
\usepackage{newtxtext,newtxmath}
% Depending on your LaTeX fonts installation, you might get better results with one of these:
%\usepackage{mathptmx}
%\usepackage{txfonts}

% Use vector fonts, so it zooms properly in on-screen viewing software
% Don't change these lines unless you know what you are doing
\usepackage[T1]{fontenc}
\usepackage{ae,aecompl}

%%%%% AUTHORS - PLACE YOUR OWN PACKAGES HERE %%%%%

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{nccmath}
\usepackage{isotope}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{multicol}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\raggedbottom
\setlength{\parskip}{1em}
%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\setlength{\parskip}{1em}
%%%%%%%%%%%%%%%%%%% TITLE PAGE %%%%%%%%%%%%%%%%%%%

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Short title, max. 45 characters]{Effects of rotation and magnetic braking on Li abundances for solar type stars}

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[R. Caballero Navarro et al.]{
R. Caballero Navarro,$^{1}$\thanks{E-mail: rcaballeron@correo.ugr.es}
A. Garc\'ia Hern\'andez,$^{1,2}$\thanks{E-mail: agh@ugr.es}
A. Ayala$^{2},$\thanks{E-mail: aayala@ugr.es}
J.~C. Su\'arez$^{1,2}$\thanks{E-mail: jcsuarez@ugr.es}
\\
% List of institutions
% Affiliations should be in the format ‘Department, Institution, Street
% Address, City and Postal Code, Country’.
$^{1}$Dept. Theoretical Physics and Cosmology, University of Granada (UGR), 18071, Granada, Spain\\
$^{2}$Instituto de Astrof\'isica de Andaluc\'ia (CSIC), Glorieta de la Astronom\'ia S/N, 18008, Granada, Spain\\
}

% These dates will be filled out by the publisher
\date{Accepted XXX. Received YYY; in original form ZZZ}

% Enter the current year, for the copyright statements etc.
\pubyear{2019}

% Don't change these lines
\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

% Abstract of the paper
\begin{abstract}
\textit{Goals.} The study the influence of rotation and magnetic braking on lithium (Li) depletion during pre-main sequence (PMS) and main sequence (MS) for solar-type stars.
\newline\textit{Methods.} The effects of rotational mixing and of the hydrostatic effects of rotation on Li abundances are studied by computing rotating PMS \& MS models. The impact of the magnetic braking effect is then analyzed by comparing the Li abundances of those models experiencing different magnetic field intensities ranging between 3.5 and 5.5 G.
\newline\textit{Results.} \textbf{TBD}
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
rotation -- magnetic fields -- abundances
\end{keywords}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%% BODY OF PAPER %%%%%%%%%%%%%%%%%%

\section{Introduction}
Despite decades of theoretical efforts, a coherent explanation can not be found from the models for the discrepancies found in the comparisons between the abundance of Li for stars belonging to clusters of different ages and that are also in the same evolutionary state either in the PMS or in the MS. Additionally, these theoretical models are not able to explain the abundances detected in the late stages of the MS \citep{Tschape2001}.\par

The denominated standard models, models that include convection only as a mixing process and do not consider other transport options like diffusion or angular momentum loss (AML), have been mainly involved in the elaboration of these predictions \citep{Sestito2005}. Li burns at a temperature $T_{Li} \approx 2.5 x 10^6\; K$, and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value. Solar-type stars are characterized by having a convective zone that covers much of the stellar radius during PMS, causing their lower limit to exceed the temperature $T_{Li}$ necessary to initiate the Li destruction \citep{Iben1965}. This depletion stops during the approach to the zero-age MS (ZAMS), when the convection zone retreats and again becomes cool. In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances, and remain nearly fixed until the terminal-age MS (TAMS). During this same period the convection mechanisms induce a mixing process that homogenizes the chemical composition of the convective zone, from its lower limit to the surface of the star. However different abundances of Li have been observed for different star populations (\citet{Somers2014} and references therein).\par

Since the twenties of the last century it has been accepted that rotation is capable of inducing the mixing of the elements present inside the stars. The studies of Eddington (1925) and Vogt (1925) have already revealed the fact that the southern circulation must necessarily arise in those stars in rotation as a consequence of the von Zeipel paradox: a star cannot be simultaneously in thermal and hydrostatic equilibrium if its rotation speed depends exclusively on its radius. The solution to this depends on the existence of meridian circulations. At the same time, this type of instability implies an impact on the convection mechanisms not considered by the standard models. In addition to this type of instability, there is today a catalogue of others that appear as a consequence of assuming models in rotation \citep{Maeder2003a}, especially the differential rotation as a function of the radius of the star or the deviation of the spherical symmetry. Although initially these effects could be candidates to explain the observations, the fact is that in the Sun the deviation of the measured spherical symmetry is of the order of $10^{-5}$ and the time scale for the southern circulation is of the order of $10^{12}$ years, a time period well above the nuclear combustion time scale \citep{Pinsonneault1997}. In view of these findings, it seems evident that there are still questions to be resolved on the subject of the processes that participate, directly or indirectly, in the mixing mechanisms that are not incorporated in the standard stellar models.\par

Rotation is the primary cause of deviation from spherical symmetry in the Sun. The Sun is a slow rotator, and slow rotation can be treated as a linear perturbation that makes it easy to determine how rotation changes mode frequencies. Slow rotation in this case is one for which the maximum centrifugal force is much smaller than the force of gravity and frequency u is much larger than the rotational frequency U. This works well for the Sun, for which the ratio of centrifugal force to gravity at the equator at the solar surface is on the order of $10^{-5}$

The impact of rotation on PMS Li depletion of solar-type stars has already been discussed (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}). These previous studies focused mainly on the hydrostatic effects and did not consider the influence of a coupling between the star magnetic field and a possible spin-down effect. As previously mentioned, it's thought that AML has a direct influence in the mixing processes and transport of the momentum could be produced by a number of mechanisms that could be important: mass loss, magnetic fields and waves (g-modes). Concerning the two latter, waves and magnetic fields have the property that they generally transmit angular momentum much more effectively than they induce mixing  \citep{Denissenkov2007}. As a result, the efficiency of these processes is most easily tested by their impact on angular momentum evolution. Of particular interest for this work is the impact caused by the magnetic field of the star in the loss of angular momentum, more specifically the effect produced on it by magnetic braking as a result of the solidary rotation that produces on the particles that constitute the solar wind and the torque derived from this coupling. These magnetic fields would originate in the limits between the radiative and convective zones of the star, in the so-called tachoclines, due to the rotational difference between both zones \citep{Guerrero2015} (On the role of tachoclines in solar and stellar dynamos). Models of rotationally driven dynamos in stellar radiative zones have suggested that magnetohydrodynamic transport of angular momentum and chemical composition can dominate over the otherwise purely hydrodynamic processes. A proper consideration of the interaction between rotation and magnetic fields is therefore essential. We have adapted our one-dimensional stellar rotation code, MESA, to model magnetic braking and include its effect on the equations for the evolution of the angular momentum distribution and stellar structure. This produces a more complete, though still reasonably simple, model for angular momentum evolution. In this study we are now focused in investigating not only the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the magnetic braking on the Li destruction as a result of its influence on the rotational history of solar-type stars.\par

In the remainder of this paper, we'll describe a physically sound yet computationally simple model extension to Modules for Experiments in Stellar Astrophysics star evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the angular momentum lost as a result of the torque applied by a magnetically-coupled stellar wind. In its simplest form, implementation of the model requires specification of both the surface magnetic field strength $B$ and the initial  rotation rate $\Omega$ at star surface. Our simulations to trace the rotational history and A(Li) of a $1 M_{\sun}$ star, for a variety of initial values for $B$ and $\Omega$.\par

We define some conventions and assumptions adopted in the paper. We use $M_i$ throughout to refer to the initial stellar mass of a model. Both Z and [Z/H] are used to refer to metallicities, where Z is the metal mass fraction. For the models presented in this paper, [Z/H] = [Fe/H] since we adopt solar-scaled abundances. The [Z/H] notation assumes the \citet{Asplund2009} protosolar birth cloud metallicity, not the current photospheric metallicity, as the reference value. Lastly, in accordance with the XXIXth IAU resolutions B3 \citep{Mamajek2015} we adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10} cm$ and $M_{\odot} = 1.988x10^{33} g$. We also adopt the protosolar abundances recommended by \citet{Asplund2009} as the reference scale for all metallicities, unless noted otherwise. In other words, [Z/H] is computed with respect to $Z = Z_{\odot, protosolar} = 0.0142$, not $Z = Z_{\odot, photosphere} = 0.0134$. For a detailed description of the physics adopted in this paper, see \citet{Choi2016}. This work has been used as a starting point for our work regarding the parameterization of the MESA project calibrated to reproduce the measured values on the solar surface, in particular the section 4.1 of the aforementioned paper.\par
The paper is organized as follows. Section 2 gives an overview of (TBD)

\section{Methods, Observations, Simulations etc.}
\begin{table}
	\centering
	\caption{Summary of adopted physics in MESA (based on \citet{Choi2016})}
	\label{tab:phy_mesa}
	\begin{tabular}{ll} 
		\hline
		Parameter & Adopted prescriptions and values\\
		\hline
		Solar Abundance & $X_{\odot}=0.7154, Y_{\odot}=0.2703, Z_{\odot}=0.0142$\\
		Equation of State & OPAL+SCVH+MacDonald+HELM+PC\\
		Opacity & OPAL Type I for log T $\geq$ 4 \\ & Ferguson for logT $\leq$ 4\\
		Reaction Rates & JINA REACLIB\\
		Boundary Conditions & ATLAS12; $\tau$=100 tables + photosphere\\
		Difussion & Track \isotope[1]{H}, \isotope[2]{He}, \isotope[7]{Li}, \isotope[7]{Be}\\
		Rotation & Differential rotation at PMS \& MS\\
		Convection & MLT + Ledoux, $\alpha_{MLT}$ = 1.82\\
		Overshoot & time-dependent, diffusive, \\ & $f_{ov,core}=0.0160$,\\ 
		& $f_{ov,sh}=0.0174$\\
		Semiconvection & $\alpha_{sc}=0.1$\\
		Thermohaline & $\alpha_{th}=666$\\
		Rotational Mixing & Include SH, ES, GSF, SSI \& DSI\\
		Magnetic Effects & Magnetic braking based on idealized \\ & monopole field\\
		Magnetic Field & B(G) variable between [3.5 - 5.5]\\
		Mass Loss & activated, $\Dot{M}_{max} = 10^{-3} \: M_{\sun} \: yr^{-1}$\\
		Angular Moment Loss & activated, $\Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A}$\\
		\hline
	\end{tabular}
\end{table}

In this section we review the relevant physic aspects simulated in MESA and the most relevant parameters adopted in the models simulations. In its simplest version, by activating rotation in the model, the evolutionary computational code MESA takes into account the following effects:
%\vspace{-1pc}
\begin{enumerate}
    \item Angular momentum transport from the radiative interior to the convective envelope in response to the rotational deceleration of the stellar surface layers\label{itm:1}.
    \item Angular momentum redistribution associated with changes in internal structure during the process of contraction to the main sequence\label{itm:2}.
    \item Angular momentum loss as a result of the torque applied to the convection zone by a magnetically coupled wind\label{itm:3}.
\end{enumerate}

The enumerate effects \ref{itm:1} and \ref{itm:2} are standard features offered by MESA. On the contrary, the effect \ref{itm:3} isn't available during the elaboration of this paper and represents our extension to MESA.

\subsection{Modelling rotation}

Mixing driven by stellar rotation is important for a variety of problems in stellar structure and evolution. Hydrodynamic mechanisms, such as meridional circulation and shear instabilities, induce mixing that is consistent with observational data (see \citet{Pinsonneault1997} \citet{Maeder2000}). However, the evolution of a rotating star depends on its internal angular momentum distribution, and existing theoretical models have difficulty matching empirical constraints on the timescale for transport in stellar radiative interiors \citep{Denissenkov2007}. \par

The stellar evolution code used for these computations is the MESA code, which includes a detailed treatment of stellar structure which deviates from spherical symmetry in the presence of rotation. While the structure is inherently 3D, it suffices to solve the stellar structure equations in one dimension if the angular velocity, $\Omega$, is constant over isobars (the so-called shellular approximation; see, e.g., \citet{Meynet1997} and to calculate the modification to the stellar equations due to centrifugal acceleration \citep{Endal1976}.

\subsection{Effects of rotation}
Rotation may affect the equations of stellar structure in four ways \citep{Endal1976}:
%\renewcommand{\theenumi}{\arabic{enumi}.}
\begin{enumerate}
    \item Centrifugal forces reduce the effective gravity at any point not on the axis of rotation.
    \item Since the centrifugal force is not, in general, parallel to the force of gravity, equipotential surfaces are no longer spheres.
    \item Because the radiative flux varies with the local effective gravity (the von Zeipel effect), the radiative flux is not constant on an equipotential surface.
    \item Rotation may inhibit certain modes of convective motions and, thus, directly affect the criterion for convective stability. Rotation may induce some mixing processes.
\end{enumerate}

Rotational mixing is a particularly interesting mechanism for explaining the Li variability during the stellar evolution. Stars are born with a range of initial rotation conditions and the rotation slows over time \citep{Skumanich1972}, so the rate of mixing differs from star to star in young systems, leading to the emergence of Li dispersion. Both the convective envelope and the radiative interior spin down over the course of 0.5-5 Gyr \citep{Somers2014a}. Throughout stellar evolution, the strength of mixing depends both on the absolute rotation rate of the star and on the degree of differential rotation within the star which both lead to non-uniform angular momentum (AM) histories. This makes Li depletion efficient at young ages and inefficient at old ages, given the progressive spin down of stars, naturally explaining the different measured Li depletion rates \citep{Sestito2005}. On the other hand, stellar rotation rates show a large dispersion at ZAMS \citep{Stauffer1984}, providing in this way a consistent argument for explaining the variance of Li abundancies among stars which have a similar mass and were formed within the same nebulas. Although this approach is not a novelty at all, it is no less true that in those studies based on it the effect caused by magnetic braking over the rotation on the PMS and MS is often neglected. \par

Additionally, and according to the prevalent theory of rotational mixing, meridional circulation and internal shears leads to a homogenization of the material depleted of Li found within the convective zone, causing the inner layers where there is a temperature higher than $T_{Li}$, and therefore where the Li is destroyed, been mixed with those cooler above them. These mechanism causes a positive feedback loop in which the cooler material sinks, is heated again and the remaining Li is destroyed, totally or partially, before proceeding to rise once more into cooler layers. \par

With regards to MESA, the transport of angular momentum and chemicals due to rotationally induced instabilities is implemented in a diffusion approximation \citet{Endal1978}. MESA calculates diffusion coefficients for five different rotationally induced mixing processes: dynamical shear instability, Solberg-H{\o}iland instability, secular shear instability, Eddington-Sweet circulation, and the Goldreich-Schubert-Fricke instability \citep{Paxton2013}.

\subsection{Modelling magnetic braking}
The origin of these magnetic fields is still unknown. The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields (see \citet{Dudorov2014} for details about the Theory of fossil magnetic field), or fields produced through a dynamo mechanism \citep{Cantiello2009}. It has long been a requirement for T Tauri stars to host kG magnetic fields \citep{Hussain2014} in order to explain several key observational characteristics in young, accreting PMS systems \citep{Johns-Krull2007}. One of this features is the long rotating periods  which is much longer than predicted from angular momentum conservation.  A key question is to understand how these young stars can accrete large amounts of marterial from their circumstellar disk with high specific angular momentum and at the same time they're able to keep slow rotation periods, between 7-10 days \citep{Hussain2014}. One plausible explanation for this behaviour considers the loss of angular momentum produced by the interaction between the magnetic field and the particles that form its circumstellar disk \citep{Zanni2012}. Once accretion stops or becomes less efficient then stars are free to spin up as they contract towards the MS.\par

An important difference between massive and low-mass stars is the presence of an outer convection zone. In the present paper. High-mass stars have no outer convective envelopes. While they may have convective cores, radiative energy transport dominates in the outer zones of these stars. Because dissipation timescales are long, strong magnetic fields may survive there, but fields are not generated, and fields that may be generated in the core find no easy way to the surface.\par

Low-mass stars, on the other hand, have outer convective envelopes in which magnetic fields decay within only a few decades or centuries \citep{Chabrier2006}, and where motion of ionized particles apparently manage to generate strong magnetic fields as for example in the Sun. The efficiency of magnetic field generation through a dynamo process depends on several conditions, but the details of these are not well known (e.g., \citet{Charbonneau2010}). \par

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/hussain_gaitee_fig1.png}
    \caption{H-R diagram showing Behrend \& Maeder (2001) pre-main sequence evolutionary tracks for stellar masses up to $4.0 \; M_{\sun}$. The dashed blue line marks the position of the birthline. All stars with $1.0 \; M_{\sun}$ will undergo a stage along their pre-main sequence evolution in which they have a fully convective interior near the birthline, to developing a radiative core when it's approaching the ZAMS (black dot-dashed line) \citet{Hussain2014}}
    \label{fig:hr_hussain}
\end{figure}

In the present work we do not take into account the existence of possible magnetic fields during the T-Tauri phase and we adopt a simple theoretical approach to establish their existence during the ZAMS approximation phase. The fixed criterion is based on the simultaneous existence of an extensive convective layer and a radiative core. From this moment on, the magnetic braking routine will be activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participate in the loss of the star's angular momentum. Additionally, based on our simple approach, we assume that the magnetic field does not vary its intensity throughout the evolution of the star until it reaches TAMS.\par

Our main objective for this paper is to contrast, in a first approximation, the degree of validity of an approach that includes magnetic braking as a complementary mechanism of angular moment loss that can contribute to explain the evolution of Li in solar type stars. More advanced models of the arrangement of the magnetic field as well as the evolution of its intensity throughout the life of the star are left for later work.\par

\subsubsection{Internal Magnetic Fields}
The Sun is governed by a strong magnetic field which is generated with a magnetic field strength of $B\approx10^5\; G$ in the tachocline, the thin shear layer between the radiative and the convective zone \citep{Aschwanden2014}. The differential rotation on the solar surface is thought to wind up the surface magnetic field, which then fragments under the magnetic stress, circulates meridionally to the poles, and reorients from the toroidally stressed state (with field lines oriented in the eastewest direction) at solar maximum into a poloidal dipole field (connecting the North with the South Pole) in the solar minimum. This process is called the solar dynamo, which swaps the magnetic polarity of the Sun every $\approx$ 11 years (the solar cycle), or returns to the same magnetic configuration every $\approx$ 22 years (the Hale cycle, discovered in 1925 by the US astronomer George Ellery Hale).\par

In MESA, magnetic fields generated by differential rotation in radiative regions have been implemented following the work of Spruit \citep{Paxton2013}. It has been suggested that differential rotation in the radiative layers of a star can amplify a magnetic field. Such a dynamo process was proposed by \citet{Spruit2002}. It's also worth to mention that a theoretical debate on this topic is still ongoing \citep{Denissenkov2007}. MESA accounts for transport by magnetic fields of angular momentum and chemicals due to the Spruit-Tayler dynamo.

\subsubsection{Surface Magnetic Fields}
The Sun has a very large and very complex magnetic field. The magnetic field at an average place on the Sun is $B\approx1\; G$, about twice as strong as the average field on the surface of Earth  $B\approx0.5\; G$. Buoyant magnetic flux tubes rise through the convection zone (due to the convective instability obeying the Schwarzschild criterion) and emerge at the solar surface in active regions, where they form sunspots with magnetic field strengths of $B\approx10^3\; G$ and coronal loops with field strengths of $B\approx10^2\; G$ at the photospheric, and $B\approx10 G$ in larger coronal heights \citep{Aschwanden2014}.\par

As appointed in the previous section, rotating stars that have a significant outer convective zone can produce surface magnetic fields through a dynamo mechanism (see, e.g. \citet{Brandenburg2004}, \citet{Charbonneau2010} or \citet{Brun2017} for a review on astrophysical dynamos). Whatever the origin of surface magnetic fields, these are expected to couple to the wind mass-loss and, if this is strong enough, to produce magnetic braking \citet{UdDoula2002}, \citet{Ud-Doula2007}, \citet{Ud-Doula2008} \citet{Meynet2010}.\par

In the presence of mass-loss, angular momentum is removed by the material in the stellar wind, resulting in a spin-down effect. This is quite important for example in massive stars, that are known to be rapidly rotating and have strong stellar winds. It has been also observed that strongly magnetic intermediate-mass stars typically have rotation rates much slower than other stars in their parent population \citep{Mathys2006}. In those stars, the presence of such magnetic field will interplay with the mass loss. If the Alfv\'{e}n radius $R_{A}$, defined as the point where the ratio between the magnetic field energy density $B^{2}/8\upi$ and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$ of the wind is equal to 1, is larger than the stellar radius, then the flow will have to follow the magnetic field. As a consequence the material will leave the stellar surface with a higher specific angular momentum, as the co-rotation radius has increased. This co-rotation radius will roughly correspond to the $R_{A}$. This is the effect known as magnetic braking.\par

This energy ratio (eq. \ref{eq:wind_conf}) denominated wind confinement magnetic parameter $\eta_*$, defines a characteristic parameter for the relative effectiveness of the magnetic fields in confining and/or channeling the wind outflow \citep{UdDoula2002}. The two most important parameters regarding a stellar wind that can be derived from the observations are the mass loss rate $\Dot{M}$, which is the amount of mass lost by the star per unit time, and terminal velocity $\nu_{esc}$, which is the velocity of the stellar wind at a large distance from the star. \par

\begin{ceqn}
\begin{equation}
    \eta_* = \frac{B^{2}/8\upi}{\rho\nu^2/2} \label{eq:wind_conf}
\end{equation}
\end{ceqn}

For a star with a stationary spherically symmetric wind, the mass loss rate is related to the density and the velocity at any point in the wind via the equation of mass continuity

\begin{ceqn}
\begin{equation}
    \Dot{M} = 4\upi r^2\rho\nu \label{eq:mass_loss}
\end{equation}
\end{ceqn}

Using (\ref{eq:wind_conf}) and (\ref{eq:mass_loss}) $\eta_*$ can be approximated by: 
\begin{ceqn}
\begin{equation}
    \eta_* = \frac{B^{2}r^{2}}{\Dot{M}\nu} \label{eq:wind_conf2}
\end{equation}
\end{ceqn}

The values of $\Dot{M}$ and $\nu_{esc}$ are important because:
 \begin{enumerate}
    \item $\Dot{M}$ describes how much material is lost by the star per unit of time. This is important for the evolution of the stars, because stars with high mass loss rates will evolve differently from those with low mass loss rates.
    \item The gas that escapes from the star into space carries kinetic energy into the interstellar medium. The amount of kinetic energy that a stellar wind deposits into the interstellar medium per unit of time is $0.5\Dot{M}\nu_{esc}$.
 \end{enumerate}

In general, a magnetically channeled outflow will have a complex flow geometry, but for convenience, the second equality in equation simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate $\Dot{M}=4\upi r^{2}\rho\nu$. $\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law $ \nu(r) = \nu_\infty (1-R_*/r)$, where $\nu_\infty$ is the terminal wind velocity defined as the velocity that the wind (the out-flowing matter) reaches at large distance from the central star, where it is not accelerated anymore by the wind driving force but its deceleration due to interaction with the interstellar medium (ISM) is negligible \citep{Niedzielski2002}.\par

The line-driven winds of massive OB stars have terminal velocities (see e.g. Lamers \& Cassinelli 2000) that scale with the photospheric escape velocity 
\begin{ceqn}
\begin{align}
\nu_{esc} &= \sqrt{\frac{2GM_*}{R_*}} \label{eq:vesc} \\
\nu_\infty &\simeq 1.92 \;\nu_{esc}\\
\nu_\infty &\simeq 1.92 \; x \; 618 \; \Bigg(\sqrt{\frac{R_{\sun}}{R_*}\frac{M_*}{M_{\sun}}} \;\Bigg) \label{eq:vinf}
\end{align}
\end{ceqn}
where (\ref{eq:vesc}) is the Newtonian escape velocity from the stellar surface and $G$ is the gravitational constant and (\ref{eq:vinf}) the terminal velocity expressed in terms of escape velocity, radius and mass of the Sun.\par

MESA does not include the physics of magnetic braking, as its only considers the evolution of stars without surface magnetic fields. On the other hand, the formalism of the magnetic braking law used here follows theoretical developments first made for the Sun, starting with \citet{Weber1967} who used an idealized monopole field to model the AML in the solar wind $\Dot{J}$. They found that:
\begin{ceqn}
\begin{equation}
 \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A} \label{eq:j_dot}
\end{equation}
\end{ceqn}
where $\Dot{M}$ is the mass loss rate and $R_A$ the Alfv\'{e}n radius. \par

In our simulations, we turn on Reimers mass loss at the beginning of the evolution, but a negligible amount of mass loss occurs thorough the MS (approx $10^{-13}M_{\odot} \; yr^{-1}$ for a solar metallicity $1 M_{\odot}$ star). With regards to the magnetic braking routine implementation, we've based on \citet{Ud-Doula2007}. A set of different simulation scenarios with magnetic field strength ranging from between 3 and 10 G have been simulated. We'd like to highlight (again) that the magnetic field remains constant during the PMS \& MS phases. We computed the evolution of $1M\sun$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.\par

\subsection{Evolving the angular velocity}
As stated in \citet{Paxton2015} initialization of rotation in MESA begins from a non-rotating model. The angular velocity $\omega$ is added to the set of model variables and initialized to a constant value throughout the model (i.e., solid body rotation). The initial value of $\omega$ can be specified as a surface rotational velocity (in $km\,s^{-1}$) or as a fraction of the surface critical rotation rate (we opted for this option). During the subsequent evolution, $\omega$ is changed at each time step by remeshing, mass adjustment, radius adjustment (as part of the structure evolution), optional extra angular momentum removal in the outer layers, and the transport of angular momentum optionally with user-defined source terms for external torques. Our magnetic braking routine influences the angular velocity evolution of the star making use of the MESA 'hook' which allow to incorporate that aforementioned external torque factor.\par

The angular velocity ($\omega$) is defined at cell boundaries. Thus $\omega_k$ is at the outer boundary of cell $k$, which is the same location as the radius ($r_k$), the specific angular moment of inertia ($I_k$), and the specific angular momentum ($J_k$). During the mass adjustment operation, when mass is added or removed from the model, cells in the outer layers are moved to new mass locations. As part of this process, the angular velocity values are updated to conserve angular momentum using the same scheme as for remeshing: sum the angular momentum in the original model and set $\omega_k$ in the new model to conserve it \citet{Paxton2015}.

\subsection{Evolving the angular momentum}
MESA performs the transport of angular momentum after any mass change and after solving for the new structure. Calculation of the new stellar structure changes the radii. Given the new radius ($r_k$), it calculates the new specific angular moment of inertia for that particular radius ($I_k$). Then using the unchanged angular momentum ($J_k$), the angular velocity ($\omega_k$) is set to $J_k/I_k$ to conserve specific angular momentum. Next, MESA applies an optional, user-specified amount of specific AML (see eq. \ref{eq:k_jdot}) in the outer surface layers ($\Dot{J}_{k_e}$) \citet{Paxton2015}.\par

\begin{ceqn}
\begin{align}
\Dot{J}_{k_e} &= \Dot{J}_*\;\frac{m^{}_{k} r^2_{k}}{m^{}_* r_*^2} \label{eq:k_jdot}
\end{align}
\end{ceqn}
where $m_{k}$ is the mass associated with the cell $k$.\par

\subsection{Effects of magnetic braking}
According to \citet{Meynet2010} the effects of magnetic braking depend a lot on how angular momentum, hence the chemical species, is transported inside the star and there are two limiting cases:
\begin{enumerate}
    \item Differential rotation: the angular momentum transport is driven by the meridional currents and the shear instabilities.
    \item Solid body rotation: when the transport of the angular momentum is very efficient, then solid body rotation is maintained during the whole MS phase. The chemical species are transported by the meridional currents.
\end{enumerate}

Magnetic braking is the process where mass being lost from the star via stellar winds is still bound to the star due to magnetic field of the star. Despite the amount of mass being relatively small for a solar like star (approx $10^{-13}M_{\odot} \; yr^{-1}$), the radius where the material is locked to the star's surface rotation is large. This results in an appreciable amount of angular momentum being lost when this mass leaves the system.\par

The default prescription for magnetic braking which is commonly referred to as the Skumanich magnetic braking law, is calibrated to work with a specific subset of stellar systems. Skumanich initially calibrated the rate of AML to single G type main sequence stars \citep{Skumanich}.\par

Convection, differential rotation and many not fully understood processes in stars can give rise to magnetic fields \citet{Langer2012}. In general magnetic fields are produced by moving charged particles, or currents. Convection produces local magnetic fields, if the surface layers of star is convective, this will lead to magnetic fields at the surface \citet{Langer2012}. If there is a stellar wind of charged particles leaving the star, and a strong magnetic field, the star will slow itself down. The charged wind will interact with the magnetic field as it is driven away from the star. The magnetic field and the star are bound in rotation, they have the same angular velocity. The angular velocity of the wind in respect to the rotation axis of the star will stay the same as on the envelope. It will then move in respect to the magnetic field and induce a torque on the star, due to its inertia. If the wind is strong and the magnetic fields are strong, it slows down the star significantly. In the Sun, magnetic braking results from solar wind material following the magnetic field lines that extend well beyond the stellar surface. This coupling exerts a torque on the surface layers of the Sun, and this slows down its rotation.\par

\section{Results}
\subsection{Angular velocity and Li evolution}
Li is a very fragile element that is burned via proton capture at temperatures as low as $2.5 x 10^6 K$. Non-convective mixing can take place in radiative regions, driven by angular AML, and causes Li depletion. Fast rotating ZAMS stars have suffered little AML and so would have the highest Li abundances. Slow rotators may have suffered low AML if they started with a less angular momentum, or a lot if they were magnetically coupled to a circumstellar disk for an extended period of time, and therefore may have a range of Li abundances. \par

Theoretical models of stars don't inform about the initial amount of Li that a star has, they only describe how it is exhausted. Therefore, in order to make an accurate estimate of the initial abundance of Li, it is a prerequisite to be able to compare observations and models beforehand. Our Sun represents a unique exception, since it allows us to know the current abundance of this element in its photosphere, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex$ \citep{Jeffries2004}, where $A(\isotope[7]{Li})$ is defined according to eq. (\ref{eq:A_Li})\par

\begin{ceqn}
\begin{align}
    A(\isotope[7]{Li}) &= log(N_{\isotope[7]{Li}} / N_{\isotope[1]{H}}) + 12
    \label{eq:A_Li}
\end{align}
\end{ceqn}

On the other hand, the initial abundance of $A(\isotope[7]{Li}) = 3.34 \, dex$ is obtained from meteorite measurements \citet{Randich2006}. According to the theory for newly born stars, we have that the initial abundance of Li can be estimated fairly accurately from photospherical measurements on T-Tauri type stars, or from the hottest F stars forming part of slightly older clusters. For the latter, the current theory of stellar evolution suggests that Li must not yet be exhausted. The results obtained from the measurements on both types of stars allow us to fix the initial abundance of Li in the interval $3.0 \, dex < A(\isotope[7]{Li}) < 3.4 \, dex$ \citet{Randich2006}.\par

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/li_var_vel_0_0g.eps}
    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models. The solid black line represents the fiducial model according to \citet{Choi2016}. The rest of lines are models which include PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively. The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
    \label{fig:li_var_vel_0g}
\end{figure}

As previously mentioned, standard stellar evolution models so far have not been able to successfully reproduce the solar surface Li abundance, indicating the need to include extra physical mechanisms. One way to account for missing physics in the models is to incorporate the effects of rotation, mixing processes and the magnetic braking. In Figure \ref{fig:li_var_vel_0g} we show the evolution of surface Li abundance relative to H as a function of time for several $1M_{\odot}$ models which take into account the effects of rotation and AML caused by stellar winds and. The purple star and square are surface Li abundance for the present-day Sun \citet{Asplund2009} and Pleiades surface Li abundance \citep{Sestito2005}.\par

In the figure (\ref{fig:li_var_vel_0g}) it can be clearly appreciated how the surface Li abundance decreases over time in all the simulated models due to the inclusion of diffusion. The solid black line represents the fiducial model that adopts the solar-calibrated envelope overshoot parameters as documented in \citet{Choi2016}. In general, all models burn too much Li early on and do not match with the Pleiades surface Li abundance. Later, the fidicual model (black line) does not deplete lithium efficiently on the MS and fail (again) to match the current solar surface Li abundance. On the other hand, the other models which include rotation during the PMS with values of $v/v_{crit}$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but it rotational velocity is much higher (see fig. \ref{fig:rot_vel_0g}) than the $2 kms^{-1}$ of the Sun \citep{Gill2012} \par

Figure (\ref{fig:rot_vel_0g}) shows the photosphere rotation profile at different ages for the different rotating models. At the beginning of the PMS phase, the star is fully convective and rotates as a solid body. As soon as the radiative core develops, differential rotation takes place in the radiative zone mainly as a result of the star contracting. This effects can be clearly recognized in the figure (\ref{fig:rot_vel_cz_0g}) which shows, in addition to the angular velocity of the star surface, the velocity corresponding to the upper and lower limits of the convective zone. Differential rotation in radiative zone is directly influenced by the initial angular velocity and leads to a steep rotation profile at the base of the convective zone; the higher the initial velocity, the bigger the velocity gradient between the bottom and top limits of the convective zone. As a consequence, the turbulence strength located at the bottom of the convective envelope also increases so that the resulting mixing can transport Li to deeper and hotter regions, where it is finally destroyed in a more efficient way (see figure (\ref{fig:li_var_vel_0g})).On the other hand, the quantitative impact of rotation on the PMS Li abundance of solar-type stars seems not to be very sensitive to the initial angular velocity until the radiative core develops.\par

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
    \caption {The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models in the approach to the ZAMS (dashed lines). The solid black line represents the fiducial model according to \citet{Choi2016}. The rest of lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.}
    \label{fig:li_var_vel_0g_z1}
\end{figure}

The evolutionary tracks in the HR diagram are shown for the models in figure (\ref{fig:hr_var_vel_0g}). A well known structural effect of rotation is the decrease of the effective temperature and to a less extent of stellar luminosity. These results are in line with those of previous studies (see e.g. \citet{Eggenberger2012}, \citet{Piau2001}, \citet{Pinsonneault1989}). Figure (\ref{fig:li_var_vel_0g_z1}) shows a zoomed-in view of evolution tracks from the ZAMS till the TAMS in which the aforementioned effects caused by rotation are easily recognized. As depicted in figure (\ref{fig:li_var_vel_0g_z1}) at the end of the PMS, the rotating models exhibits a lower surface Li abundance than the non-rotating one. We thus see that the inclusion of rotation results in a global increase of the Li depletion during the PMS. We also note that the Li depletion occurs earlier for the rotating model than for the non-rotating.\par

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/rot_vel_0g.eps}
    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models. The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}. The dashed lines make reference to the ZAMS.}
    \label{fig:rot_vel_0g}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/rot_vel_cz_0g.eps}
    \caption{The evolution of angular velocity at surface and limits of the uppermost convective zone, as a function of time for several 1 $M_{\sun}$ models. The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}. The dashed lines make reference to the ZAMS.}
    \label{fig:rot_vel_cz_0g}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/hr_var_vel_0g.eps}
    \caption{An example solar 1$M_{\sun}$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities. The rotation is activated in the models in the PMS.}
    \label{fig:hr_var_vel_0g}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/hr_var_vel_0g_z1.eps}
    \caption{Similar to figure (\ref{fig:hr_var_vel_0g}) but now showing in detail the effects of rotation on the evolutionary tracks.}
    \label{fig:hr_var_vel_0g_z1}
\end{figure}

\subsection{Magnetic braking and Li evolution}
In the rest of this document we are going to focus, unless otherwise indicated, on models that include rotation from the PMS since they are the ones that can offer a greater approximation to the observations. In turn, these models will serve as a basis for others in which our magnetic braking routine is used by simulating magnetic fields of different intensity, varying from 3.5G to 5.5G in 0.5G increments.\par

If we concentrate our attention on figure (\ref{fig:li_var_vel_4_0g}) and compare it with figure (\ref{fig:li_var_vel_0g}) we will see how the profiles of Li abundance are already altered during PMS and MS. In the first phase we can describe the effect as modest, somewhat expected and in line with the fact that the loss of angular momentum caused by magnetic braking (see equation (\ref{eq:j_dot})) depends directly on the loss of mass. If we take into account that for solar type stars the models predict a modest total loss of mass, this is even much lower during PMS. On the contrary, during the MS it is observed in a much more evident way how the loss of angular momentum is much more significant, causing the star to rotate more slowly and as a result a smaller amount of Li to be destroyed.\par

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/li_var_vel_4_0g.eps}
    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models. The models include a magnetic field with an intensity of 4G and PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively. The dashed lines make reference to the ZAMS.}
    \label{fig:li_var_vel_4_0g}
\end{figure}

The effect of the magnetic braking routine can be seen even more clearly in figure (\ref{fig:rot_vel_4g}). Here we represent the rotation profiles on the surface of the stars for different rotating models. As it can be observed, and in a similar way to the evolution profiles of the Li on the surface commented in the previous paragraph, the effect of the routine is much more accentuated once the ZAMS is reached. If the evolution of the curves presented here with those of figure (\ref{fig:rot_vel_0g}) we see how the star, instead of continuing to increase its angular velocity, begins to slow down after having reached its maximum in the passage through the ZAMS.\par

We also find it interesting to highlight the behaviour of angular velocity profiles when approaching TAMS. As can be seen in figure (\ref{fig:rot_vel_4g}) and in more detail in (\ref{fig:rot_vel_4g_z1}), we find that the evolution of angular velocity suffers from a series of ups and downs that do not appear previously. Everything seems to indicate that this behavior is due to a combined effect of the magnetic braking routine and the time-step (probably to big) used by MESA during the last part of the simulation. We based this affirmation on the fact that this effect is not observed in the series of simulations in which the routine is not included (figs. (\ref{fig:rot_vel_0g}) and (\ref{fig:rot_vel_cz_0g})).\par

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/rot_vel_4g.eps}
    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models. The models include a magnetic field with an intensity of 4G and PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. Additionally it's simulated the magnetic braking effect of magnetic field with an intensity of 4G. The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}. The dashed lines make reference to the ZAMS.}
    \label{fig:rot_vel_4g}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/rot_vel_cz_4g.eps}
    \caption{The evolution of angular velocity at surface and limits of the uppermost convective zone, as a function of time for several 1 $M_{\sun}$ models. The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. The dashed lines make reference to the ZAMS.}
    \label{fig:rot_vel_cz_4g}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/rot_vel_4g_z1.eps}
    \caption{Similar to figure (\ref{fig:rot_vel_4g}) but now showing in detail the surface rotational velocity as the star approaches TAMS.}
    \label{fig:rot_vel_4g_z1}
\end{figure}

It's also remarkable to highlight the lower angular velocity observed in figure (\ref{fig:rot_vel_cz_4g} at the star surface in comparison with the upper limit of the convective layer. This fact is most probably a consequence of the expansion of the superficial layers after the mass loss produced by stellar winds.\par

As mentioned above, the magnetic braking routine distributes the total amount of AML calculated according to equation (\ref{eq:k_jdot}) among the different layers that compose it. In figure (\ref{fig:cz_vc_028_var_b}) we can observe the evolution of the most external convective zone normalized with respect to the radius of the star. In accordance with the established models of stellar evolution, in a solar-type star the convective zone covers practically all of it for a large part of the PMS. As it approaches ZAMS it is decreasing, as a consequence of the appearance of a radiative core, maintaining an approximately constant radius until the final stage of the MS, where it increases significantly as a response to the generalized increase of the star's radius. Regarding the effect of magnetic braking on the size of the convective zone, we observe that as the intensity of the magnetic field increases, the size of the convective zone decreases. According to \citet{Jeffries2004} this direct consequence is expected because the magnetic fields in the convective zone raise the adiabatic temperature gradient and accelerate the formation of a radiative core. This radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the BCZ drops below the one required for burning Li. This effect is most evident during MS (fig. (\ref{fig:cz_vc_028_var_b_z1})). The decrease in the size of the convective zone is in line with the fact that less Li is destroyed by causing less star material to reach areas with temperatures above $T_{Li}$. \par

Similar to the evolution of the angular velocity towards the end of the MS discussed in the previous paragraph, we observe that the size of the convective zone also shows a series of ups and downs. Again we think that this effect is a direct consequence of the temporal resolution used by MESA in the stellar evolution equations. The specific origin of these fluctuations is a subject that we will deal with in more depth and detail the origin of these fluctuations in subsequent works to the current one.\par

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/cz_vc_028_var_b.eps}
    \caption{The evolution of convective zone size, as a function of time for several 1 $M_{\sun}$ models. The models include a magnetic field with an intensity of 4G and PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. Additionally it's simulated the magnetic braking effect of magnetic field with an intensity of 4G.The dashed lines make reference to the ZAMS.}
    \label{fig:cz_vc_028_var_b}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/cz_vc_028_var_b_z1.eps}
    \caption{Similar to figure (\ref{fig:cz_vc_028_var_b}) but now showing in detail the convection zone size from the near ZAMS till the TAMS. As the intensity of the magnetic field increases, the size of the convective zone decreases.}
    \label{fig:cz_vc_028_var_b_z1}
\end{figure}

Before ending this section we would also like to comment on the other factor that directly affects AML; the loss of mass, $\Dot{M}$. In figure (\ref{fig:mdot_vc_028_var_b}) we can see the evolution of $\Dot{M}$ during PMS and MS. In the first stages of PMS is where the greatest loss of mass is concentrated and this diminishes as it approaches the ZAMS. If we add to the scenario the fact that the star also decreases its radius during the PMS we have as a result that the star increases $\Omega$ obeying the principle of conservation of angular momentum. When reaching the ZAMS, the radius of the star remains more or less stable for much of the MS (except for its final stage) but continues to lose mass so it continues to increase its angular velocity, although in a less aggressive way if we compare it with the PMS. However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a convective zone (see fig. (\ref{fig:mb_act_var_vel_vc_028})), the magnetic braking routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS. The more intense the magnetic field, the greater the braking effect.\par

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/mdot_vc_028_var_b.eps}
    \caption{The evolution of mass loss $\Dot{M}$ , as a function of time for several 1 $M_{\sun}$ models. The models include a variable magnetic field intensity between 0G and 5.5G and PMS rotation with $v/v_{crit}=0.028$. The dashed lines make reference to the ZAMS.}
    \label{fig:mdot_vc_028_var_b}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/mdot_vc_028_var_b_z1.eps}
    \caption{The evolution of mass loss $\Dot{M}$ as the star is approaching TAMS, as a function of time for several 1 $M_{\sun}$ models. The models include a variable magnetic field intensity between 0G and 5.5G and PMS rotation with $v/v_{crit}=0.028$. The stronger the magnetic field, the lower both the angular velocity and $\Dot{M}$. The dashed lines make reference to the ZAMS.}
    \label{fig:mdot_vc_028_var_b_z1}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/mb_act_var_vel_vc_028.eps}
    \caption{The evolution of convective zone size, as a function of time for several 1 $M_{\sun}$ models. The models include a magnetic field with an intensity of 4G and PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. Additionally it's simulated the magnetic braking effect of magnetic field with an intensity of 4G. The solid lines signal the magnetic braking routine activation (on) and deactivation (off). The horizontal dashed lines inform about the star's core nature: radiative (rad) or convective (conv). The vertical dashed lines make reference to the ZAMS.}
    \label{fig:mb_act_var_vel_vc_028}
\end{figure}

\begin{figure*}
    \centering
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_var_vel_0_0g.eps}
    \label{fig:subim1}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_var_vel_3_5g.eps}
    \label{fig:subim2}
    \end{subfigure}

    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_var_vel_4_0g.eps}
    \label{fig:subim3}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_var_vel_4_5g.eps}
    \label{fig:subim4}
    \end{subfigure}

    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_var_vel_5_0g.eps}
    \label{fig:subim5}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_var_vel_5_5g.eps}
    \label{fig:subim6}
    \end{subfigure}
\caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models. The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. The purple star is the surface Li abundance for the present-day Sun \citep{Asplund2009}. The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
\label{fig:image2}
\end{figure*}
\par

\begin{figure*}
    \centering
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_vc_0084_var_b.eps}
    \label{fig:subim21}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_vc_014_var_b.eps}
    \label{fig:subim22}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_vc_0196_var_b.eps}
    \label{fig:subim23}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_vc_028_var_b.eps}
    \label{fig:subim24}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/li_vc_0336_var_b.eps}
    \label{fig:subim25}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/blank.eps}
    \label{fig:subim26}
    \end{subfigure}
\caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models. The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. The purple star is the surface Li abundance for the present-day Sun \citep{Asplund2009}. The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
\label{fig:image22}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/mb_act_3_5g.eps}
    \label{fig:subim31}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/mb_act_4_0g.eps}
    \label{fig:subim32}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/mb_act_4_5g.eps}
    \label{fig:subim33}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/mb_act_5_0g.eps}
    \label{fig:subim34}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/mb_act_5_5g.eps}
    \label{fig:subim35}
    \end{subfigure}
    \begin{subfigure}[h]{0.47\textwidth}
    \includegraphics[width=\textwidth]{figures/blank.eps}
    \label{fig:subim36}
    \end{subfigure}
\caption{The evolution of magnetic braking activation, as a function of time and the existence of a radiative core for several 1 $M_{\sun}$ models. The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
\label{fig:image32}
\end{figure*}

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/li_var_vel_3_5g.eps}
    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models. The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively. The purple star is the surface Li abundance for the present-day Sun \citep{Asplund2009}. The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
    \label{fig:li_var_vel_3_5g}
\end{figure}

In Figs. 2 and 3, we notice several interesting new facts for models with a magnetic field:
- the models have almost an internal solid body rotation, with a difference of only a few percent between the core and the bulk of the envelope, with a short transition at their inter- face. The differential rotation parameter q is very small, but still different from zero. The relative constancy of $\Omega$ is the result of the high value of the transport coefficient ν by the magnetic field (see below Fig. 8). We notice that these results are in agreement with those of Heger et al. (2004), who also find essentially solid body rotation until the end of the MS phase. Then in later phases, differential rotation appears together with a slowing down of the core;
- there is a general slowing down of rotation during evolution due to two effects: the mass loss which removes angular momentum and the expansion of the external layers. The strong internal coupling of rotation transports this slowing down to the core;
- models M 1 and M 5, although not strictly identical (see Fig. 4), exhibit similar $\Omega$ distributions;
- we notice a small decrease of $\Omega$ near the surface. This is likely resulting from the expansion of the superficial layers after the removal of some mass by stellar winds at the surface.

red lines are models with less and more efficient overshoot, resulting in reduced and enhanced lithium depletion, respectively. The three dot-dashed lines show additional variations in input
physics: the red lines correspond to a MLT = 1.7 , while the orange and green lines are models that include PMS rotation with v v crit = 0.01 and 0.10, respectively. As expected, a lower a MLT produces a puffier, cooler star, resulting in less lithium depletion. The inclusion of rotation during the PMS produces very different, potentially more promising behavior, though the current absence of magnetic braking (see Sec-
tions 3.6.5 and 3.5) in MESA results in an unrealistically high rotation speed ( v v crit between 0.1 and 1) on the MS. The models presented here fail to simultaneously match both the young and present-day solar surface lithium abundances.

\section{Conclusions}
The inclusion of rotation during the PMS and magnetic braking during MS produces very different, potentially more promising behavio, though magnetic braking during PMS seems to be also required for explaining Li abundances in young clusters.

The Teff’s from standard evolutionary tracks represent upper limits since these models do not include the magnetic field effect on the thermal structure.
Inclusion of the magnetic field leads to cooler models and lower Li depletions in the PMS.

¿Para la cabecera?
This generally neglected physical input ( thermal modifications that are due to the magnetic field) affects the location of the PMS: tracks that do not allow for the effect of the magnetic field on the convective Teff gradients provide an upper limit to the.
This new feature in MESA is not equivalent to an artificial parametric change of the convection efficiency.
We also point out that more complete models are needed, including a self-consistency model between the dynamo
magnetic field and the evolution of stellar rotation.

Mass loss by stellar winds drastically reduces vrot during the evolution (Packet
et al 1980; Langer \& Heger 1998; Figs. 3 and 4). Even if isotropic, the stellar winds carry away quite a lot of angular momentum, and this is even more important in the case of equatorial mass loss. The new surface layers then have a lower vrot as a result of expansion and redistribution.

Indicar que se hace un planteamiento relativamente sencillo para la existencia de un campo magnético: existencia de núcleo radiativo y envelope convectivo. Para un tratamiento más avanzado se puede recurrir al documento (pag 10) "Stellar Evolution with rotation and magnetic fields" (Maeder \& Meynet)

Página 6 de "On the revolution ofrotational velocity..." hay un planteamiento teórico para hacer dependiente la intensidad del campo magnético. Utilizar esto para futuros planteamientos. En la página 16 deriva la fórmula para el cálculo de la pérdida de momento angular. Utilizarla para enlazar con la versión simplificada ofrecida por Cantiello.

The last numbered section should briefly summarise what has been done, and describe
the final conclusions which the authors draw from their work.

A T Tauri star is a type of pre–main sequence star that is being heated through gravitational contraction and has not yet begun to burn hydrogen at its core. They are variable stars that are magnetically active. The magnetic field of these stars is thought to interact with its strong stellar wind, transferring angular momentum to the surrounding protoplanetary disk. This allows the star to brake its rotation rate as it collapses. PODRÍAMOS INDICAR QUE YA QUE NO ACTIVAMOS EL FRENADO MAGNÉTICO DURANTE ESTA FASE SERÍA FACTIBLE QUE LA CONCENTRACIÓN DE LI FUERA SUPERIOR A LA QUE TENEMOS EN EL MODELO, ACORDE CON PLEIADES, Y QUE EL FRENADO MAGNÉTICO DURANTE LA FASE T-TAURI PERMITIESE QUE SE LLEGASE A ESE VALOR Y LUEGO FUERA REDUCIENDO LA VELOCIDAD PARA QUE NO TERMINARA DESTRUYENDO MÁS DE LA CUENTA.

\section*{Acknowledgements}

The Acknowledgements section is not numbered. Here you can thank helpful
colleagues, acknowledge funding agencies, telescopes and facilities used etc.
Try to keep it short. xxxx

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%% REFERENCES %%%%%%%%%%%%%%%%%%

% The best way to enter references is to use BibTeX:

\bibliographystyle{mnras}
%\bibliography{magbrlitium} % if your bibtex file is called example.bib
\bibliography{mblithium}

% Alternatively you could enter them by hand, like this:
% This method is tedious and prone to error if you have lots of references
%\begin{thebibliography}{99}
%\bibitem[\protect\citeauthoryear{Author}{2012}]{Author2012}
%Author A.~N., 2013, Journal of Improbable Astronomy, 1, 1
%\bibitem[\protect\citeauthoryear{Others}{2013}]{Others2013}
%Others S., 2012, Journal of Interesting Stuff, 17, 198
%\end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%

\appendix

\section{Some extra material}

If you want to present additional material which would interrupt the flow of the main paper,
it can be placed in an Appendix which appears after the list of references.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Don't change these lines
\bsp	% typesetting comment
\label{lastpage}
\end{document}

% End of mnras_template.tex

### 2019-06-16 project/octave/plotting_routines.m

@@ -1,17 +1,22 @@
 format longEng;
 clear global;
 
-global star_age_col       = 3;
 global header_lines       = 6;
+
+%cols position in output files
+global star_age_col       = 3;
+global log_abs_m_dot_col  = 7;
 global mass_conv_core_col = 8;
 global log_LH_col         = 17;
 global log_Teff_col       = 33;
 global log_L_col          = 34;
+global log_R_col          = 35;
 global surf_avg_v_rot_col = 42;
 global center_h1_col      = 54;
 global surface_h1_col     = 55;
 global surface_li_col     = 59;
 global mb_activated_col   = 72;
+global j_dot_col          = 77;
 global sz_top_radius_col  = 79;
 global sz_bot_radius_col  = 80;
 global sz_top_zone_col    = 83;
@@ -22,7 +27,7 @@ global sz_bot_vrot_col    = 86;
 global data_parent_folder = '/home/rcaballeron/MESA/workspace/sun-jupiter-system/Docs/runs/run_paper';
 global filename = '1M_photosphere_history.data';
 global gauss_fields = ['0g'; '3.5g'; '4g'; '4.5g'; '5g'; '5.5g'];
-global rotational_vels = ['0kms'; '3kms'; '5kms'; '7kms'; '10kms'; '12kms'];
+global rotational_vels = ['0crit';'0084crit'; '014crit'; '0196crit'; '028crit'; '0336crit';];
 global colors = ['k'; 'r'; 'g'; 'b'; 'y'; 'm'; 'c'];
 
 %Sun constants
@@ -204,7 +209,7 @@ function result = calculate_A_Li7(A)
 end 
 
 
-function plot_A_Li7(A, B, color, width, axis_limits)
+function plot_A_Li7(A, B, color, width, ytick, axis_limits)
   %Plot values
   plot(A(:,1), B(:,1), color, 'linewidth', width);
 
@@ -215,7 +220,7 @@ function plot_A_Li7(A, B, color, width, axis_limits)
   axis(axis_limits);
   
   %Axis ticks
-  set(gca,'YTick',1.0:0.5:4.5);
+  set(gca,'YTick',1.0:ytick:4.5);
   xticks = get (gca, "xtick"); 
   xlabels = arrayfun (@(x) sprintf ("%.2e", x), xticks, "uniformoutput", false); 
   set (gca, "xticklabel", xlabels) ;
@@ -246,7 +251,7 @@ end
 
 
 
-function plot_vel_rot(A, color, width, axis_limits)
[…]

### 2019-06-22 project/overleaf/paper.tex
Paper v2

@@
 % Keep the title short and informative.
-\title[Short title, max. 45 characters]{Effects of rotation and magnetic braking on Li abundances for solar type stars}
+\title[Rotation, magnetic braking \& Li abundances]{Effects of rotation and magnetic braking on Li abundances for solar-type stars}
 
@@
 \begin{abstract}
-\textit{Goals.} The study the influence of rotation and magnetic braking on lithium (Li) depletion during pre-main sequence (PMS) and main sequence (MS) for solar-type stars.
+\textit{Goals.} The study the influence of rotation and magnetic braking (MB) on lithium (Li) depletion during pre-main sequence (PMS) and main sequence (MS) for solar-type stars.
 \newline\textit{Methods.} The effects of rotational mixing and of the hydrostatic effects of rotation on Li abundances are studied by computing rotating PMS \& MS models.
-The impact of the magnetic braking effect is then analyzed by comparing the Li abundances of those models experiencing different magnetic field intensities ranging between 3.5 and 5.5 G.
-\newline\textit{Results.} \textbf{TBD}
+The impact of the MB effect is then analyzed by comparing the Li abundances of those models experiencing different magnetic field intensities ranging between 3.5 and 5.5 G.
+\newline\textit{Results.} The surface Li abundance at the end of the PMS and throughout the MS is decreased when rotational effects are included.
+This results in an increment in the Li depletion rate for rotating models compared to non-rotating ones.
+This effect is attenuated when the braking effect produced by a magnetic field is added.
+This physical phenomena, generally neglected in star evolutionary models, impacts also the effective temperature ($T_{eff}$) of the star and therefore its location in the HR diagram.
+The impact of MB in Li depletion is sensitive to the field intensity: the higher it is, the lower the Li destruction.
+A relationship between the magnetic fields and the convective zone (CZ) size is observed: stronger magnetic fields produce shallower CZ's.
+Another important result of this study is that MB effect must be taken into consideration during PMS must if we aim to reproduce Li abundances in young clusters.
 \end{abstract}
@@
 \section{Introduction}
-Despite decades of theoretical efforts, a coherent explanation can not be found from the models for the discrepancies found in the comparisons between the abundance of Li for stars belonging to clusters of different ages and that are also in the same evolutionary state either in the PMS or in the MS.
+Despite decades of theoretical efforts, a coherent explanation cannot be found from the models for the discrepancies found in the comparisons between the abundance of Li for stars belonging to clusters of different ages and that are also in the same evolutionary state either in the PMS or in the MS.
 Additionally, these theoretical models are not able to explain the abundances detected in the late stages of the MS \citep{Tschape2001}.\par
@@
 Li burns at a temperature $T_{Li} \approx 2.5 x 10^6\; K$, and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value.
-Solar-type stars are characterized by having a convective zone that covers much of the stellar radius during PMS, causing their lower limit to exceed the temperature $T_{Li}$ necessary to initiate the Li destruction \citep{Iben1965}.
-This depletion stops during the approach to the zero-age MS (ZAMS), when the convection zone retreats and again becomes cool.
+solar-type stars are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $T_{Li}$ \citep{Iben1965}.
+This depletion stops during the approach to the zero-age MS (ZAMS), when the convection zone retreats and the temperature at BCZ is cooler that $T_{Li}$.
 In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances, and remain nearly fixed until the terminal-age MS (TAMS).
-During this same period the convection mechanisms induce a mixing process that homogenizes the chemical composition of the convective zone, from its lower limit to the surface of the star.
+During this same period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the convective envelope, from its lower limit to the star surface.
 However different abundances of Li have been observed for different star populations (\citet{Somers2014} and references therein).\par
 
-Since the twenties of the last century it has been accepted that rotation is capable of inducing the mixing of the elements present inside the stars.
-The studies of Eddington (1925) and Vogt (1925) have already revealed the fact that the southern circulation must necessarily arise in those stars in rotation as a consequence of the von Zeipel paradox: a star cannot be simultaneously in thermal and hydrostatic equilibrium if its rotation speed depends exclusively on its radius.
+From approximately the 1920s of the last century it has been accepted that rotation is capable of inducing the mixing of the elements present inside the stars.
+The studies of Eddington (1925) and Vogt (1925) had revealed the fact that the southern circulation must necessarily arise in those stars in rotation as a consequence of the von Zeipel paradox: a star cannot be simultaneously in thermal and hydrostatic equilibrium if its rotation speed depends exclusively on its radius.
 The solution to this depends on the existence of meridian circulations.
 At the same time, this type of instability implies an impact on the convection mechanisms not considered by the standard models.
-In addition to this type of instability, there is today a catalogue of others that appear as a consequence of assuming models in rotation \citep{Maeder2003a}, especially the differential rotation as a function of the radius of the star or the deviation of the spherical symmetry.
-Although initially these effects could be candidates to explain the observations, the fact is that in the Sun the deviation of the measured spherical symmetry is of the order of $10^{-5}$ and the time scale for the southern circulation is of the order of $10^{12}$ years, a time period well above the nuclear combustion time scale \citep{Pinsonneault1997}.
-In view of these findings, it seems evident that there are still questions to be resolved on the subject of the processes that participate, directly or indirectly, in the mixing mechanisms that are not incorporated in the standard stellar models.\par
-
-Rotation is the primary cause of deviation from spherical symmetry in the Sun.
-The Sun is a slow rotator, and slow rotation can be treated as a linear perturbation that makes it easy to determine how rotation changes mode frequencies.
-Slow rotation in this case is one for which the maximum centrifugal force is much smaller than the force of gravity and frequency u is much larger than the rotational frequency U.
-This works well for the Sun, for which the ratio of centrifugal force to gravity at the equator at the solar surface is on the order of $10^{-5}$
+Today it is known that in combination with this type of instability exists a catalogue of others that appear as a consequence of assuming models in rotation \citep{Maeder2003a}, especially the differential rotation as a function of the radius of the star or the deviation of the spherical symmetry.
+Although initially these effects could be candidates to explain the observations, the fact is that the Sun is a slow rotator, meaning slow rotation that one for which the maximum centrifugal force is much smaller than the force of gravity, its deviation of the measured spherical symmetry is of the order of $10^{-5}$ and the time scale for the southern circulation is of the order of $10^{12}$ years, a time period well above the nuclear combustion time scale \citep{Pinsonneault1997}.
+In view of these findings, it seems evident that there are still open questions to be resolved on the subject of the processes that participate, directly or indirectly, in the mixing mechanisms that are not incorporated in the standard stellar models.\par
 
 The impact of rotation on PMS Li depletion of solar-type stars has already been discussed (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}).
-These previous studies focused mainly on the hydrostatic effects and did not consider the influence of a coupling between the star magnetic field and a possible spin-down effect.
+These previous studies focused mainly on the hydrostatic effects and didn't consider the influence of a coupling between the star magnetic field and its possible spin-down effect.
 As previously mentioned, it's thought that AML has a direct influence in the mixing processes and transport of the momentum could be produced by a number of mechanisms that could be important: mass loss, magnetic fields and waves (g-modes).
-Concerning the two latter, waves and magnetic fields have the property that they generally transmit angular momentum much more effectively than they induce mixing  \citep{Denissenkov2007}.
-As a result, the efficiency of these processes is most easily tested by their impact on angular momentum evolution.
-Of particular interest for this work is the impact caused by the magnetic field of the star in the loss of angular momentum, more specifically the effect produced on it by magnetic braking as a result of the solidary rotation that produces on the particles that constitute the solar wind and the torque derived from this coupling.
-These magnetic fields would originate in the limits between the radiative and convective zones of the star, in the so-called tachoclines, due to the rotational difference between both zones \citep{Guerrero2015} (On the role of tachoclines in solar and stellar dynamos).
-Models of rotationally driven dynamos in stellar radiative zones have suggested that magnetohydrodynamic transport of angular momentum and chemical composition can dominate over the otherwise purely hydrodynamic processes.
+Concerning the two latter, waves and magnetic fields have the property that they generally transmit angular momentum (AM) much more effectively than they induce mixing \citep{Denissenkov2007}.
+As result, the efficiency of these processes is most easily tested by their impact on AM evolution.
+Of particular interest for this work is the impact caused by the magnetic field of the star in the AML, more specifically the effect produced on it by MB as a consequence of the solidary rotation that produces on the particles that constitute the solar wind and the torque derived from this coupling.
+These magnetic fields would originate in the limits between the radiative and CZ's of the star, in the so-called tachoclines, due to the rotational difference between both zones \citep{Guerrero2016}.
 A proper consideration of the interaction between rotation and magnetic fields is therefore essential.
-We have adapted our one-dimensional stellar rotation code, MESA, to model magnetic braking and include its effect on the equations for the evolution of the angular momentum distribution and stellar structure.
-This produces a more complete, though still reasonably simple, model for angular momentum evolution.
-In this study we are now focused in investigating not only the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the magnetic braking on the Li destruction as a result of its influence on the rotational history of solar-type stars.\par
+We have adapted our one-dimensional stellar rotation code to model MB and include its effect on the equations for the evolution of the AM distribution and stellar structure.
+This produces a more complete, though still reasonably simple, model for AM evolution.
+In this study we are now focused in investigating not only the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.\par
 
-In the remainder of this paper, we'll describe a physically sound yet computationally simple model extension to Modules for Experiments in Stellar Astrophysics star evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the angular momentum lost as a result of the torque applied by a magnetically-coupled stellar wind.
+In the remainder of this paper, we describe a physically sound yet computationally simple model extension to Modules for Experiments in Stellar Astrophysics star evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar wind.
 In its simplest form, implementation of the model requires specification of both the surface magnetic field strength $B$ and the initial  rotation rate $\Omega$ at star surface.
-Our simulations to trace the rotational history and A(Li) of a $1 M_{\sun}$ star, for a variety of initial values for $B$ and $\Omega$.\par
+Our simulations trace the rotational history and A(Li) of a $1 M_{\sun}$ star for a variety of initial values for $B$ and $\Omega$.\par
 
 We define some conventions and assumptions adopted in the paper.
-We use $M_i$ throughout to refer to the initial stellar mass of a model.
-Both Z and [Z/H] are used to refer to metallicities, where Z is the metal mass fraction.
-For the models presented in this paper, [Z/H] = [Fe/H] since we adopt solar-scaled abundances.
-The [Z/H] notation assumes the \citet{Asplund2009} protosolar birth cloud metallicity, not the current photospheric metallicity, as the reference value.
+For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity, not the current photospheric metallicity, as the reference value.
+In other words, [Z/H] is computed with respect to $Z = Z_{\odot, protosolar} = 0.0142$, not $Z = Z_{\odot, photosphere} = 0.0134$.
 Lastly, in accordance with the XXIXth IAU resolutions B3 \citep{Mamajek2015} we adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10} cm$ and $M_{\odot} = 1.988x10^{33} g$.
-We also adopt the protosolar abundances recommended by \citet{Asplund2009} as the reference scale for all metallicities, unless noted otherwise.
-In other words, [Z/H] is computed with respect to $Z = Z_{\odot, protosolar} = 0.0142$, not $Z = Z_{\odot, photosphere} = 0.0134$.
-For a detailed description of the physics adopted in this paper, see \citet{Choi2016}.
+For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}.
 This work has been used as a starting point for our work regarding the parameterization of the MESA project calibrated to reproduce the measured values on the solar surface, in particular the section 4.1 of the aforementioned paper.\par
-The paper is organized as follows. Section 2 gives an overview of (TBD)
 
-\section{Methods, Observations, Simulations etc.}
+The paper is organized as follows.
+Section 2 gives an overview of rotation and magnetic fields physical processes and how do they impact on the star evolution.
+We also summarized the most relevant parameters configured in MESA for the model simulations.
+In Section 3, the model outputs are presented and discussed in detail.
+Finally, Section 4 finishes this paper with a discussion of open points and future works.\par
+
+\section{Method}
 \begin{table}
@@
 
-In this section we review the relevant physic aspects simulated in MESA and the most relevant parameters adopted in the models simulations.
+In this section we review the core physic aspects simulated in MESA and the most relevant parameters adopted in the models simulations.
 In its simplest version, by activating rotation in the model, the evolutionary computational code MESA takes into account the following effects:
@@
 \begin{enumerate}
-    \item Angular momentum transport from the radiative interior to the convective envelope in response to the rotational deceleration of the stellar surface layers\label{itm:1}.
-    \item Angular momentum redistribution associated with changes in internal structure during the process of contraction to the main sequence\label{itm:2}.
-    \item Angular momentum loss as a result of the torque applied to the convection zone by a magnetically coupled wind\label{itm:3}.
+    \item AM transport from the radiative interior to the convective envelope in response to the rotational deceleration of the stellar surface layers\label{itm:1}.
+    \item AM redistribution associated with changes in internal structure during the process of contraction to the MS\label{itm:2}.
+    \item AML as a result of the torque applied to the convection zone by a magnetically coupled wind\label{itm:3}.
 \end{enumerate}
 
-The enumerate effects \ref{itm:1} and \ref{itm:2} are standard features offered by MESA.
-On the contrary, the effect \ref{itm:3} isn't available during the elaboration of this paper and represents our extension to MESA.
+The enumerated effects \ref{itm:1} and \ref{itm:2} are standard features offered by MESA.
+On the contrary, the effect \ref{itm:3} isn't available as an out-of-the-box feature during the elaboration of this paper and represents our extension to MESA.
 
 \subsection{Modelling rotation}
-
-Mixing driven by stellar rotation is important for a variety of problems in stellar structure and evolution.
-Hydrodynamic mechanisms, such as meridional circulation and shear instabilities, induce mixing that is consistent with observational data (see \citet{Pinsonneault1997} \citet{Maeder2000}).
-However, the evolution of a rotating star depends on its internal angular momentum distribution, and existing theoretical models have difficulty matching empirical constraints on the timescale for transport in stellar radiative interiors \citep{Denissenkov2007}.
+The rotation to which a star is subjected has profound implications for its inner structure and, therefore, for the way in which it evolves throughout its life.
+It is also responsible for inducing through hydrodynamic instabilities the mixing of the chemical elements that make it up.
+On the other hand, in order to be able to correctly model the effects induced in the mixing by rotation and for the results obtained from the theoretical models that implement this physical effect to be consistent with the observational data (see \citet{Pinsonneault1997}, \citet{Maeder2000}), it is necessary to analyze and understand more precisely the mechanisms that govern the distribution of angular momentum, since in this area the theoretical models also offer deficiencies to explain the observations \citep{Denissenkov2007}.
 \par
 
-The stellar evolution code used for these computations is the MESA code, which includes a detailed treatment of stellar structure which deviates from spherical symmetry in the presence of rotation.
-While the structure is inherently 3D, it suffices to solve the stellar structure equations in one dimension if the angular velocity, $\Omega$, is constant over isobars (the so-called shellular approximation; see, e.g., \citet{Meynet1997} and to calculate the modification to the stellar equations due to centrifugal acceleration \citep{Endal1976}.
+Angular velocity, or rather the way in which it evolves, also conditions the mechanism by which theoretical models solve the equations of stellar structure.
+It is evident that stars have a three-dimensional structure and for this reason the equations that model their evolution must take this fact into account.
+However, it has been proven sufficient to solve the equations in only one dimension, extrapolating the results to the other two, if it is fulfilled that the angular velocity is constant in regions subjected to the same pressure (isobaric).
+This is what is known as shellular approximation \citet{Meynet1997} and is based on calculating the impact of the centrifugal acceleration caused by the rotation in the stellar evolution equations \citep{Endal1976}.
+MESA uses this simplification to accelerate computational calculations.\par
 
 \subsection{Effects of rotation}
-Rotation may affect the equations of stellar structure in four ways \citep{Endal1976}:
+Rotation may affect the equations of stellar structure in four ways according to \citet{Endal1976}:
 %\renewcommand{\theenumi}{\arabic{enumi}.}
 \begin{enumerate}
-    \item Centrifugal forces reduce the effective gravity at any point not on the axis of rotation.
-    \item Since the centrifugal force is not, in general, parallel to the force of gravity, equipotential surfaces are no longer spheres.
-    \item Because the radiative flux varies with the local effective gravity (the von Zeipel effect), the radiative flux is not constant on an equipotential surface.
-    \item Rotation may inhibit certain modes of convective motions and, thus, directly affect the criterion for convective stability. Rotation may induce some mixing processes.
+    \item The centrifugal force causes the local effective gravity ($g_{eff}$) associated with a point of the star not located on the axis of rotation to be reduced.
+    \item In general, the centrifugal force will not be parallel to the gravitational force and as a consequence of this the equipotential layers will cease to be spherical.
+    \item As the von Zeipel theorem states, that the radiative flux in a uniformly rotating star is proportional to the $g_{eff}$, therefore the radiative flux is not constant on an equipotential surface.
+    \item Rotation can suppress certain mechanisms that trigger convection, indirectly affecting mixing processes.
 \end{enumerate}
 
-Rotational mixing is a particularly interesting mechanism for explaining the Li variability during the stellar evolution.
-Stars are born with a range of initial rotation conditions and the rotation slows over time \citep{Skumanich1972}, so the rate of mixing differs from star to star in young systems, leading to the emergence of Li dispersion.
-Both the convective envelope and the radiative interior spin down over the course of 0.5-5 Gyr \citep{Somers2014a}.
-Throughout stellar evolution, the strength of mixing depends both on the absolute rotation rate of the star and on the degree of differential rotation within the star which both lead to non-uniform angular momentum (AM) histories.
-This makes Li depletion efficient at young ages and inefficient at old ages, given the progressive spin down of stars, naturally explaining the different measured Li depletion rates \citep{Sestito2005}.
-On the other hand, stellar rotation rates show a large dispersion at ZAMS \citep{Stauffer1984}, providing in this way a consistent argument for explaining the variance of Li abundancies among stars which have a similar mass and were formed within the same nebulas.
-Although this approach is not a novelty at all, it is no less true that in those studies based on it the effect caused by magnetic braking over the rotation on the PMS and MS is often neglected.
+Observational data support the fact that stars are born with different angular velocities even if they come from the same interstellar cloud and that throughout evolution they undergo a deceleration process \citep{Skumanich1972} in the time period between 0.5 and 5 Gyr \citep{Somers2014a}.
+This range of different velocities implies that the mixing mechanisms induced by rotation vary their efficiency and effectiveness and, therefore, have a direct influence on the abundances of the different chemical elements observed on the surface of stars.
+This fact, together with the differential degree of rotation to which the limits between the radiative and convective zones of the star are subjected, is of particular interest for our study since Li, as mentioned in the introduction, is destroyed at a low temperature, so variations in the intensity of the mixing processes can cause convective zones to extend to interior zones with temperatures higher than $T_{Li}$ in which Li is destroyed.
+The deceleration process to which a star is subjected throughout its evolution and the direct link that the rotation has with the mixing processes also offer a natural and consistent explanation for the fact that the destruction of Li is more accentuated or efficient in young stars than in old stars \citep{Sestito2005}.
+In a similar way, stellar rotation rates show a large dispersion at ZAMS \citep{Stauffer1984}, providing in this way a consistent argument for explaining the variance of Li abundances among stars which have a similar mass and were formed within the same cloud.
+Although this approach is not a novelty at all, it is no less true that in those studies based on it the effect caused by MB over the rotation on the PMS and MS is often overlooked.
 \par
 
-Additionally, and according to the prevalent theory of rotational mixing, meridional circulation and internal shears leads to a homogenization of the material depleted of Li found within the convective zone, causing the inner layers where there is a temperature higher than $T_{Li}$, and therefore where the Li is destroyed, been mixed with those cooler above them.
+Additionally, and according to the prevalent theory of rotational mixing, meridional circulation and internal shears leads to a homogenization of the material depleted of Li found within the CZ, causing the inner layers where there is a temperature higher than $T_{Li}$, and therefore where the Li is destroyed, been mixed with those cooler above them.
 These mechanism causes a positive feedback loop in which the cooler material sinks, is heated again and the remaining Li is destroyed, totally or partially, before proceeding to rise once more into cooler layers.
@@
 
-With regards to MESA, the transport of angular momentum and chemicals due to rotationally induced instabilities is implemented in a diffusion approximation \citet{Endal1978}.
+With regards to MESA, the transport of AM and chemical elements due to rotationally induced instabilities is implemented in a diffusion approximation \citet{Endal1978}.
 MESA calculates diffusion coefficients for five different rotationally induced mixing processes: dynamical shear instability, Solberg-H{\o}iland instability, secular shear instability, Eddington-Sweet circulation, and the Goldreich-Schubert-Fricke instability \citep{Paxton2013}.
@@
 \subsection{Modelling magnetic braking}
-The origin of these magnetic fields is still unknown.
+The origin of magnetic fields is still unknown.
 The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields (see \citet{Dudorov2014} for details about the Theory of fossil magnetic field), or fields produced through a dynamo mechanism \citep{Cantiello2009}.
-It has long been a requirement for T Tauri stars to host kG magnetic fields \citep{Hussain2014} in order to explain several key observational characteristics in young, accreting PMS systems \citep{Johns-Krull2007}.
-One of this features is the long rotating periods  which is much longer than predicted from angular momentum conservation.
-A key question is to understand how these young stars can accrete large amounts of marterial from their circumstellar disk with high specific angular momentum and at the same time they're able to keep slow rotation periods, between 7-10 days \citep{Hussain2014}.
+Besides that, it has long been a requirement for T Tauri stars to host kG magnetic fields \citep{Hussain2014} in order to explain several key observational characteristics in young, accreting PMS systems \citep{Johns-Krull2007}.
+One of this features is the long rotating periods  which is much longer than predicted from AM conservation.
+A key question is to understand how these young stars can accrete large amounts of material from their circumstellar disk with high specific angular momentum and at the same time they're able to keep slow rotation periods, between 7-10 days \citep{Hussain2014}.
 One plausible explanation for this behaviour considers the loss of angular momentum produced by the interaction between the magnetic field and the particles that form its circumstellar disk \citep{Zanni2012}.
@@
 
-An important difference between massive and low-mass stars is the presence of an outer convection zone.
-In the present paper.
-High-mass stars have no outer convective envelopes.
-While they may have convective cores, radiative energy transport dominates in the outer zones of these stars.
-Because dissipation timescales are long, strong magnetic fields may survive there, but fields are not generated, and fields that may be generated in the core find no easy way to the surface.\par
-
-Low-mass stars, on the other hand, have outer convective envelopes in which magnetic fields decay within only a few decades or centuries \citep{Chabrier2006}, and where motion of ionized particles apparently manage to generate strong magnetic fields as for example in the Sun.
-The efficiency of magnetic field generation through a dynamo process depends on several conditions, but the details of these are not well known (e.g., \citet{Charbonneau2010}).
-\par
+The mass of stars plays a determining role in the existence or not of an external convective envelope.
+In turn, the presence of this convective region conditions the evolution of magnetic fields.
+As depicted in figure (\ref{fig:hr_hussain}), massive stars do not have external convective zones although they can expose a convective core.
+In these cases, the dominant type of energy transport in the outer zones of the star is radiative.
+On the other hand, the less massive stars, such as those of the solar-type, have convective covers and radiative core.
+In both types of stars strong magnetic fields can be generated in the limits of both zones (the so-called tachocline) but with notable differences.
+In the case of massive stars, these magnetic fields can survive for a long time but not reach the surface of the star.
+On the other hand in less massive stars these magnetic fields, although they are able to reach the surface of the star, decay quickly in time, in a few hundred years or even only in a few decades \citep{Chabrier2006}, so it is necessary a mechanism for generating magnetic field, e.g.
+a dynamo type, continued in time to explain its existence.
+The way in which this dynamo process is able to generate magnetic fields depends on different parameters and to this day is still not fully accepted and is subject of controversy (e.g.
+\citep{Charbonneau2010}).\par
 
@@
 In the present work we do not take into account the existence of possible magnetic fields during the T-Tauri phase and we adopt a simple theoretical approach to establish their existence during the ZAMS approximation phase.
-The fixed criterion is based on the simultaneous existence of an extensive convective layer and a radiative core.
-From this moment on, the magnetic braking routine will be activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participate in the loss of the star's angular momentum.
+The adopted criterion is based on the simultaneous existence of an extensive convective layer and a radiative core.
+From this moment on, the MB routine will be activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participate in the star AML.
 Additionally, based on our simple approach, we assume that the magnetic field does not vary its intensity throughout the evolution of the star until it reaches TAMS.\par
 
-Our main objective for this paper is to contrast, in a first approximation, the degree of validity of an approach that includes magnetic braking as a complementary mechanism of angular moment loss that can contribute to explain the evolution of Li in solar type stars.
+Our main objective for this paper is to contrast, in a first approximation, the degree of validity of an approach that includes MB as a complementary mechanism of AML that can contribute to explain the evolution of Li in solar-type stars.
 More advanced models of the arrangement of the magnetic field as well as the evolution of its intensity throughout the life of the star are left for later work.\par
@@
 \subsubsection{Internal Magnetic Fields}
-The Sun is governed by a strong magnetic field which is generated with a magnetic field strength of $B\approx10^5\; G$ in the tachocline, the thin shear layer between the radiative and the convective zone \citep{Aschwanden2014}.
-The differential rotation on the solar surface is thought to wind up the surface magnetic field, which then fragments under the magnetic stress, circulates meridionally to the poles, and reorients from the toroidally stressed state (with field lines oriented in the eastewest direction) at solar maximum into a poloidal dipole field (connecting the North with the South Pole) in the solar minimum.
-This process is called the solar dynamo, which swaps the magnetic polarity of the Sun every $\approx$ 11 years (the solar cycle), or returns to the same magnetic configuration every $\approx$ 22 years (the Hale cycle, discovered in 1925 by the US astronomer George Ellery Hale).\par
+Our Sun is an active star as far as the magnetic field is concerned.
+According to the theoretical models this magnetic field is generated by a dynamo type effect in the tachocline, the intermediate zone between the radiative core and the CZ \citep{Aschwanden2014}, and has an intensity of approximately $B\approx10^5\; G$.
+Additionally, these models predict that because of the differential rotation existing between the core and CZ, the lines of the magnetic field that at the moment of their generation have a poloidal orientation will be affected by the differential rotation and forced to adopt a toroidal typology; the $\omega$-effect.
+On the other hand, we have that the opposite case can occur, that is to say, starting from a toroidal magnetic field to obtain a poloidal configuration.
+This would be possible, on the one hand, thanks to the convective movements of material that are produced in the CZ of the star and, on the other, to assume that this material behaves as a perfect conductor.
+This second condition allows us to assume that the magnetic field lines are "frozen" in the conductive fluid and therefore must move in solidarity with it.
+Under these circumstances, the magnetic field lines will be pushed to higher areas of the CZ as the material moves vertically and, due to the combined effect of stellar rotation plus Coriolis forces, the magnetic lines will be twisted and give rise to a poloidal magnetic field.
+This is what is known as the $\alpha$-effect.\par
 
-In MESA, magnetic fields generated by differential rotation in radiative regions have been implemented following the work of Spruit \citep{Paxton2013}.
-It has been suggested that differential rotation in the radiative layers of a star can amplify a magnetic field.
-Such a dynamo process was proposed by \citet{Spruit2002}.
-It's also worth to mention that a theoretical debate on this topic is still ongoing \citep{Denissenkov2007}.
-MESA accounts for transport by magnetic fields of angular momentum and chemicals due to the Spruit-Tayler dynamo.
+A number of theoretical models have been proposed based on $\alpha$ and $\omega$ effects as an explanation to the generation of magnetic fields, among which is the so-called $\alpha\omega$ or also known as dynamo effect which was proposed by \citet{Spruit2002}.
+By means of this model it is possible to explain the change in polarity of the magnetic field that occurs in the Sun every 11 years, completing a full cycle every 22 years.
+It's also worth to mention that a theoretical debate on this topic is still ongoing \citep{Denissenkov2007} and with regards to MESA, an implementation according to the work of Spruit-Tayler has been implemented \citep{Paxton2013}.\par
 
-\subsubsection{Surface Magnetic Fields}
-The Sun has a very large and very complex magnetic field.
-The magnetic field at an average place on the Sun is $B\approx1\; G$, about twice as strong as the average field on the surface of Earth  $B\approx0.5\; G$.
-Buoyant magnetic flux tubes rise through the convection zone (due to the convective instability obeying the Schwarzschild criterion) and emerge at the solar surface in active regions, where they form sunspots with magnetic field strengths of $B\approx10^3\; G$ and coronal loops with field strengths of $B\approx10^2\; G$ at the photospheric, and $B\approx10 G$ in larger coronal heights \citep{Aschwanden2014}.\par
+\subsubsection{Surface Magnetic Fields} \label{surf_mf}
+The intensity of the magnetic field measured on the Sun's surface is about $B\approx1\; G$, twice the value of the Earth's one.
+Additionally its topology is varied and complex.
+As mentioned in the previous section, the dynamo model establishes that the magnetic field produced inside the star reaches the surface with the help of convective processes taken place in the CZ and gives rise to magnetically active regions on they reach the solar surface.
+Among them, we can find sunspots dominated by magnetic fields with an approximate an intensity of $B\approx10^3\; G$ and coronal loops with field strengths of $B\approx10^2\; G$ at the photosphere, and $B\approx10 G$ in larger coronal heights \citep{Aschwanden2014}.
+Additionally, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism (see, e.g.
+\citet{Brandenburg2004}, \citet{Charbonneau2010} or \citet{Brun2017}).
+Whatever the origin of surface magnetic fields, these are expected to couple to the wind mass-loss and, if this is strong enough, to produce MB \citet{UdDoula2002}, \citet{Ud-Doula2007}, \citet{Ud-Doula2008} \citet{Meynet2010}.\par
 
-As appointed in the previous section, rotating stars that have a significant outer convective zone can produce surface magnetic fields through a dynamo mechanism (see, e.g.
-\citet{Brandenburg2004}, \citet{Charbonneau2010} or \citet{Brun2017} for a review on astrophysical dynamos).
-Whatever the origin of surface magnetic fields, these are expected to couple to the wind mass-loss and, if this is strong enough, to produce magnetic braking \citet{UdDoula2002}, \citet{Ud-Doula2007}, \citet{Ud-Doula2008} \citet{Meynet2010}.\par
-
-In the presence of mass-loss, angular momentum is removed by the material in the stellar wind, resulting in a spin-down effect.
+In the presence of mass-loss, AM is removed by the material in the stellar wind, resulting in a spin-down effect.
 This is quite important for example in massive stars, that are known to be rapidly rotating and have strong stellar winds.
 It has been also observed that strongly magnetic intermediate-mass stars typically have rotation rates much slower than other stars in their parent population \citep{Mathys2006}.
-In those stars, the presence of such magnetic field will interplay with the mass loss.
-If the Alfv\'{e}n radius $R_{A}$, defined as the point where the ratio between the magnetic field energy density $B^{2}/8\upi$ and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$ of the wind is equal to 1, is larger than the stellar radius, then the flow will have to follow the magnetic field.
-As a consequence the material will leave the stellar surface with a higher specific angular momentum, as the co-rotation radius has increased.
-This co-rotation radius will roughly correspond to the $R_{A}$.
-This is the effect known as magnetic braking.\par
+In those stars, the presence of such magnetic field will interact with the mass loss and the Alfv\'{e}n radius $R_{A}$ play an important role in this interaction.
+The $R_{A}$ is defined as the point in which the magnetic field energy density $B^{2}/8\upi$ and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$ of the stellar wind are balanced, that is, where the coefficient between both is equal to 1.
+In case that $R_{A}$ is greater than the star radius, then the wind flow will have to follow the magnetic field.
+As a consequence, the material will leave the stellar surface with a higher specific AM, as the co-rotation radius has increased and it will roughly correspond to the $R_{A}$.
+This is the effect known as MB.\par
 
 This energy ratio (eq.
-\ref{eq:wind_conf}) denominated wind confinement magnetic parameter $\eta_*$, defines a characteristic parameter for the relative effectiveness of the magnetic fields in confining and/or channeling the wind outflow \citep{UdDoula2002}.
-The two most important parameters regarding a stellar wind that can be derived from the observations are the mass loss rate $\Dot{M}$, which is the amount of mass lost by the star per unit time, and terminal velocity $\nu_{esc}$, which is the velocity of the stellar wind at a large distance from the star.
-\par
+\ref{eq:wind_conf}) denominated wind confinement magnetic parameter $\eta_*$, defines a characteristic parameter for the relative effectiveness of the magnetic fields in circumscribing and/or channeling the wind outflow \citep{UdDoula2002}.
+In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval $\Dot{M}$, and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity $\nu_{esc}$.\par
 
@@
 
-For a star with a stationary spherically symmetric wind, the mass loss rate is related to the density and the velocity at any point in the wind via the equation of mass continuity
+Each of these factors has a decisive influence on the future of the star.
+On the one hand, stars with similar initial masses but different mass loss ratios will end up evolving very differently.
+On the other hand, the ionized particles carried by the solar wind not only contribute to the loss of mass but also to the loss of kinetic energy that is deposited in the interstellar medium.
+Given a star with a spherically symmetric wind, the mass loss rate and the kinetic energy are characterized by the following expressions:
 
 \begin{ceqn}
-\begin{equation}
-    \Dot{M} = 4\upi r^2\rho\nu \label{eq:mass_loss}
-\end{equation}
+\begin{align}
+    \Dot{M} &= 4\upi r^2\rho\nu \label{eq:mass_loss}\\
+    K_e &= 0.5\Dot{M}\nu_{esc}
+\end{align}
 \end{ceqn}
@@
 
-The values of $\Dot{M}$ and $\nu_{esc}$ are important because:
- \begin{enumerate}
-    \item $\Dot{M}$ describes how much material is lost by the star per unit of time.
-This is important for the evolution of the stars, because stars with high mass loss rates will evolve differently from those with low mass loss rates.
-    \item The gas that escapes from the star into space carries kinetic energy into the interstellar medium.
-The amount of kinetic energy that a stellar wind deposits into the interstellar medium per unit of time is $0.5\Dot{M}\nu_{esc}$.
- \end{enumerate}
-
-In general, a magnetically channeled outflow will have a complex flow geometry, but for convenience, the second equality in equation simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate $\Dot{M}=4\upi r^{2}\rho\nu$.
+In general, a magnetically channeled outflow will have a complex stream geometry, but for convenience, the second equality in equation simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate $\Dot{M}=4\upi r^{2}\rho\nu$.
 $\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law $ \nu(r) = \nu_\infty (1-R_*/r)$, where $\nu_\infty$ is the terminal wind velocity defined as the velocity that the wind (the out-flowing matter) reaches at large distance from the central star, where it is not accelerated anymore by the wind driving force but its deceleration due to interaction with the interstellar medium (ISM) is negligible \citep{Niedzielski2002}.\par
@@
 
-MESA does not include the physics of magnetic braking, as its only considers the evolution of stars without surface magnetic fields.
-On the other hand, the formalism of the magnetic braking law used here follows theoretical developments first made for the Sun, starting with \citet{Weber1967} who used an idealized monopole field to model the AML in the solar wind $\Dot{J}$.
-They found that:
+Given that the version of MESA used in this work does include neither the treatment of magnetic fields and, therefore, nor for magnetic braking effects, it is necessary to extend the simulator to include this phenomenon.
+In the present work we have followed the theoretical work on magnetic braking in the Sun carried out by \citet{Weber1967}, making use of the expression (\ref{eq:j_dot}) for the calculation of the AML.
 \begin{ceqn}
@@
 
-In our simulations, we turn on Reimers mass loss at the beginning of the evolution, but a negligible amount of mass loss occurs thorough the MS (approx $10^{-13}M_{\odot} \; yr^{-1}$ for a solar metallicity $1 M_{\odot}$ star).
-With regards to the magnetic braking routine implementation, we've based on \citet{Ud-Doula2007}.
-A set of different simulation scenarios with magnetic field strength ranging from between 3 and 10 G have been simulated.
+The above expression can be rewritten so that it depends on the $\eta_*$ instead of the $R_A$ as follows:
+\begin{ceqn}
+\begin{equation}
+ \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{*}\eta_* \label{eq:j_dot_mesa}
+\end{equation}
+\end{ceqn}
+which is a more convenient expression to be implemented in MESA because is based on values directly exposed during the simulations.
+
+As reported in table (\ref{tab:phy_mesa}), for the set of simulations performed by MESA and in particular for the mass loss estimation, the empirical formula developed by Reimers for stars in the asymptotic giant branch (AGB) is used.
+For a solar-type star the loss of mass during MS is relatively small, about  $10^{-13}M_{\odot} \; yr^{-1}$.
+With regards to the MB routine implementation, we've based on \citet{Ud-Doula2007}.
+A set of different simulation scenarios with magnetic field strength ranging from between 3 and 10G have been exercised.
 We'd like to highlight (again) that the magnetic field remains constant during the PMS \& MS phases.
-We computed the evolution of $1M\sun$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.\par
+We compute the evolution of $1M\sun$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.\par
 
 \subsection{Evolving the angular velocity}
-As stated in \citet{Paxton2015} initialization of rotation in MESA begins from a non-rotating model.
-The angular velocity $\omega$ is added to the set of model variables and initialized to a constant value throughout the model (i.e., solid body rotation).
-The initial value of $\omega$ can be specified as a surface rotational velocity (in $km\,s^{-1}$) or as a fraction of the surface critical rotation rate (we opted for this option).
-During the subsequent evolution, $\omega$ is changed at each time step by remeshing, mass adjustment, radius adjustment (as part of the structure evolution), optional extra angular momentum removal in the outer layers, and the transport of angular momentum optionally with user-defined source terms for external torques.
-Our magnetic braking routine influences the angular velocity evolution of the star making use of the MESA 'hook' which allow to incorporate that aforementioned external torque factor.\par
+As stated in the work of \citet{Paxton2015} makes its models evolve by default without taking into account the rotation but offers options to initialize models with a certain angular speed $\Omega$.
+The initial value for $\Omega$ can be indicated as a function of the rotation speed of the star surface expressed in $km\,s^{-1}$ or using the critical rotation speed as a reference (option used in our case).
+Once the model has been initialized with a specific $\Omega$, it is recalculated as the star evolves as a function of the mass, radius and AML that the simulator determines in each time step.
+\par
 
-The angular velocity ($\omega$) is defined at cell boundaries.
-Thus $\omega_k$ is at the outer boundary of cell $k$, which is the same location as the radius ($r_k$), the specific angular moment of inertia ($I_k$), and the specific angular momentum ($J_k$).
-During the mass adjustment operation, when mass is added or removed from the model, cells in the outer layers are moved to new mass locations.
-As part of this process, the angular velocity values are updated to conserve angular momentum using the same scheme as for remeshing: sum the angular momentum in the original model and set $\omega_k$ in the new model to conserve it \citet{Paxton2015}.
+MESA assigns an $\Omega$ value to each of the cells that make up the star model.
+The number of cells does not remain fixed throughout the simulation but varies according to certain evolutionary parameters (see \citet{Paxton2015} for more details).
+The $\Omega$ for each cell $k$ ($\Omega_k$) is adjusted so that the resulting angular momentum is retained after adjusting the new mass of the cell ($m_k$) according to the loss or gain suffered by the solar wind and calculate the distance from the cell to the center of the star ($r_k$).\par
 
 \subsection{Evolving the angular momentum}
-MESA performs the transport of angular momentum after any mass change and after solving for the new structure.
-Calculation of the new stellar structure changes the radii.
-Given the new radius ($r_k$), it calculates the new specific angular moment of inertia for that particular radius ($I_k$).
-Then using the unchanged angular momentum ($J_k$), the angular velocity ($\omega_k$) is set to $J_k/I_k$ to conserve specific angular momentum.
-Next, MESA applies an optional, user-specified amount of specific AML (see eq.
-\ref{eq:k_jdot}) in the outer surface layers ($\Dot{J}_{k_e}$) \citet{Paxton2015}.\par
+Similar to the situation with $\Omega$, MESA also assigns to each cell an AM value ($J_k$) after resolving the new interior structure of the star.
+The evolution of the AM for each of the cells is performed after knowing the new radius of the star and the distance at which each cell is from the star center.
+Known $m_k$ and $r_k$ proceed to calculate the AM of specific inertia ($I_k$), and from $I_k$ and using $J_k$ of the previous computational step, derives the new value of $\Omega_k$ by means of $J_k/I_k$ and thus manages to conserve the AM.
+It is only after these calculations that he proceeds to calculate the new value of $J_k$ (see \citet{Paxton2015} for more details).
+It's in this point when our MB routine plugs in and allows us to influence the internally calculated value by providing an additional contribution calculated by the user routine.
+Our implementation of the MB routine calculates the additional AML ($\Dot{J}_{k}$), derived from the external torque exerted by the magnetic field, to be taken into account and distributes it through the different layers that make up the CZ, according to the expression (\ref{eq:j_dot}.\par
 
@@
 \begin{align}
-\Dot{J}_{k_e} &= \Dot{J}_*\;\frac{m^{}_{k} r^2_{k}}{m^{}_* r_*^2} \label{eq:k_jdot}
+\Dot{J}_{k} &= \Dot{J}_*\;\frac{m^{}_{k} r^2_{k}}{m^{}_* r_*^2} \label{eq:k_jdot}
 \end{align}
 \end{ceqn}
-where $m_{k}$ is the mass associated with the cell $k$.\par
 
 \subsection{Effects of magnetic braking}
-According to \citet{Meynet2010} the effects of magnetic braking depend a lot on how angular momentum, hence the chemical species, is transported inside the star and there are two limiting cases:
+In its most widely accepted approach, the MB is linked to Skumanich's work in which he develops an empirical law of evolution of the MB according to the age of the star.
+To calibrate this law, Skumanich relied on data obtained from type G stars found in the MS \citep{Skumanich}.
+The influence of the MB on the evolution of the star is closely related to the transport of the AM and, by extension, influenced by how the chemical elements are transported within it.
+According to this approach \citep{Meynet2010}, two main mechanisms are distinguished, attending to how the transport of the AM is produced, which condition the mixing of these elements:
+
 \begin{enumerate}
-    \item Differential rotation: the angular momentum transport is driven by the meridional currents and the shear instabilities.
-    \item Solid body rotation: when the transport of the angular momentum is very efficient, then solid body rotation is maintained during the whole MS phase.
-The chemical species are transported by the meridional currents.
+    \item Differential rotation: the AM transport is pushed by southern currents and shear instabilities.
+    \item Solid body rotation: when the transport of AM is very efficient, the the solid body rotation is the prevailing mechanism maintained throughout the entire MS phase and the chemical elements are transported by the southern currents.
 \end{enumerate}
 
-Magnetic braking is the process where mass being lost from the star via stellar winds is still bound to the star due to magnetic field of the star.
-Despite the amount of mass being relatively small for a solar like star (approx $10^{-13}M_{\odot} \; yr^{-1}$), the radius where the material is locked to the star's surface rotation is large.
-This results in an appreciable amount of angular momentum being lost when this mass leaves the system.\par
-
-The default prescription for magnetic braking which is commonly referred to as the Skumanich magnetic braking law, is calibrated to work with a specific subset of stellar systems.
-Skumanich initially calibrated the rate of AML to single G type main sequence stars \citep{Skumanich}.\par
-
-Convection, differential rotation and many not fully understood processes in stars can give rise to magnetic fields \citet{Langer2012}.
-In general magnetic fields are produced by moving charged particles, or currents.
-Convection produces local magnetic fields, if the surface layers of star is convective, this will lead to magnetic fields at the surface \citet{Langer2012}.
-If there is a stellar wind of charged particles leaving the star, and a strong magnetic field, the star will slow itself down.
-The charged wind will interact with the magnetic field as it is driven away from the star.
-The magnetic field and the star are bound in rotation, they have the same angular velocity.
-The angular velocity of the wind in respect to the rotation axis of the star will stay the same as on the envelope.
-It will then move in respect to the magnetic field and induce a torque on the star, due to its inertia.
-If the wind is strong and the magnetic fields are strong, it slows down the star significantly.
-In the Sun, magnetic braking results from solar wind material following the magnetic field lines that extend well beyond the stellar surface.
-This coupling exerts a torque on the surface layers of the Sun, and this slows down its rotation.\par
+Returning to section \ref{surf_mf} on the estimated mass loss ratio for a solar-type star, approximately $10^{-13}M_{\odot} \; yr^{-1}$, and using this value to estimate $\Dot{J}$, the result obtained can be considered relatively modest as to decisively influence the evolution of the star.
+On the other hand, if we consider that the convective movements that take place in the stellar interior together with the presence of differential rotation can give rise to the generation of magnetic fields that end up reaching the surface of the star \citep{Langer2012}, the AML can be increased in several orders of magnitude as a consequence of the magnetic field that forces the ionized particles that compose the stellar wind to turn with the same angular speed and whose influence on the stellar wind extends a distance several times the radius of the star (see \citet{UdDoula2002}, \citet{Ud-Doula2007} and \citet{Ud-Doula2008} for more details).
+This distance is directly influenced by the radius of alven (see eq.
+(\ref{eq:j_dot})) or alternatively by the confinement parameter (see eq.
+(\ref{eq:j_dot_mesa})).
+As a result of this approach, if the value of $\eta_*$ is large, the obtained AML will be also large and the star undergoes a considerable deceleration.
+In our proposal, the value of $\eta_*$ can be directly influenced by establishing in the different simulations the value of the magnetic field ($B$) on the surface or the rotation speed of the star ($\Omega$).
 
@@
 \subsection{Angular velocity and Li evolution}
-Li is a very fragile element that is burned via proton capture at temperatures as low as $2.5 x 10^6 K$.
-Non-convective mixing can take place in radiative regions, driven by angular AML, and causes Li depletion.
+The destruction of Li occurs when a Li atom collides with a proton producing two He atoms, something that occurs during proton-proton type II reactions (P-P II) and at a temperature lower than that required for H fusion, as low as $2.5\, x\, 10^6\, K$.
+Those stars without a convective zone can still destroy Li by non-convective mixing processes which can take place in radiative regions and are driven by AML.
 Fast rotating ZAMS stars have suffered little AML and so would have the highest Li abundances.
-Slow rotators may have suffered low AML if they started with a less angular momentum, or a lot if they were magnetically coupled to a circumstellar disk for an extended period of time, and therefore may have a range of Li abundances.
+Slow rotators may have suffered low AML if they started with a less AM, or a lot if they were magnetically coupled to a circumstellar disk for an extended period of time, and therefore may have a range of Li abundances.
 \par
@@
 
-As previously mentioned, standard stellar evolution models so far have not been able to successfully reproduce the solar surface Li abundance, indicating the need to include extra physical mechanisms.
-One way to account for missing physics in the models is to incorporate the effects of rotation, mixing processes and the magnetic braking.
-In Figure \ref{fig:li_var_vel_0g} we show the evolution of surface Li abundance relative to H as a function of time for several $1M_{\odot}$ models which take into account the effects of rotation and AML caused by stellar winds and.
+Today, standard models are not able to replicate in a satisfactory way the observed values of Li abundance on star surfaces with the predictions that these models yield.
+This lack of agreement makes us think that certain physical mechanisms that influence the destruction of Li are either being modelled correctly or simply not being taken into account, e.g.
+the magnetic braking.
+Figure (\ref{fig:li_var_vel_0g}) shows the evolution of surface Li abundance relative to H as a function of time and for several $1M_{\odot}$ models which take into account the effects of rotation and AML caused by stellar winds but not MB.
 The purple star and square are surface Li abundance for the present-day Sun \citet{Asplund2009} and Pleiades surface Li abundance \citep{Sestito2005}.\par
 
-In the figure (\ref{fig:li_var_vel_0g}) it can be clearly appreciated how the surface Li abundance decreases over time in all the simulated models due to the inclusion of diffusion.
-The solid black line represents the fiducial model that adopts the solar-calibrated envelope overshoot parameters as documented in \citet{Choi2016}.
+The figure (\ref{fig:li_var_vel_0g}) clearly allows to appreciate how the Li abundance in the star surface decreases over time for all simulated models.
+The solid black line represents the reference model that adopts the solar-calibrated envelope overshooting parameters as documented in \citet{Choi2016}.
 In general, all models burn too much Li early on and do not match with the Pleiades surface Li abundance.
-Later, the fidicual model (black line) does not deplete lithium efficiently on the MS and fail (again) to match the current solar surface Li abundance.
+Later, the reference model (black line) does not deplete Li efficiently on the MS and fail (again) to match the current solar surface Li abundance.
 On the other hand, the other models which include rotation during the PMS with values of $v/v_{crit}$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but it rotational velocity is much higher (see fig.
@@
 
-Figure (\ref{fig:rot_vel_0g}) shows the photosphere rotation profile at different ages for the different rotating models.
-At the beginning of the PMS phase, the star is fully convective and rotates as a solid body.
-As soon as the radiative core develops, differential rotation takes place in the radiative zone mainly as a result of the star contracting.
-This effects can be clearly recognized in the figure (\ref{fig:rot_vel_cz_0g}) which shows, in addition to the angular velocity of the star surface, the velocity corresponding to the upper and lower limits of the convective zone.
-Differential rotation in radiative zone is directly influenced by the initial angular velocity and leads to a steep rotation profile at the base of the convective zone; the higher the initial velocity, the bigger the velocity gradient between the bottom and top limits of the convective zone.
-As a consequence, the turbulence strength located at the bottom of the convective envelope also increases so that the resulting mixing can transport Li to deeper and hotter regions, where it is finally destroyed in a more efficient way (see figure (\ref{fig:li_var_vel_0g})).On the other hand, the quantitative impact of rotation on the PMS Li abundance of solar-type stars seems not to be very sensitive to the initial angular velocity until the radiative core develops.\par
+For much of the PMS the star rotates as a solid body (see fig.
+(\ref{fig:rot_vel_0g})) and this is because the star has a completely convective interior.
+It is not until the end of Hayashi track that the star begins to develop a radiative nucleus due to the increase in temperature that occurs inside the star during the contraction process.
+It is in this stage of evolution of the star when a difference of angular velocity appears between the upper and lower limits of the radiative and convective zones respectively.
+This effects can be clearly recognized in the figure (\ref{fig:rot_vel_cz_0g}) which shows, in addition to star surface $\Omega$, the velocity corresponding to the upper and lower limits of the CZ.
+The degree of differential rotation in radiative zone is directly influenced by the initial $\Omega$.
+As the models are initialized with a higher angular velocity, the difference in velocity between the BCZ and the star surface is accentuated in a directly proportional mode; the higher the initial velocity, the bigger the velocity gradient between the bottom and top limits of the CZ.
+As a consequence, the turbulence strength located at the BCZ increases so that Li can reach regions with temperatures about $T_{Li}$ where it is finally burned and destroyed (see fig.
+(\ref{fig:li_var_vel_0g})).
+Also remarkable from the results offered by the simulations is the fact that there are barely any differences between the different models in terms of the abundance of Li during the PMS (see fig.
+(\ref{fig:li_var_vel_0g_z1})).\par
 
@@
 
-The evolutionary tracks in the HR diagram are shown for the models in figure (\ref{fig:hr_var_vel_0g}).
-A well known structural effect of rotation is the decrease of the effective temperature and to a less extent of stellar luminosity.
+Another well known structural effect of rotation is the decrease of the effective temperature ($T_{eff}$) and to a less extent of stellar luminosity.
+Both effects can be observed graphically in the HR diagram of figure (\ref{fig:hr_var_vel_0g}).
 These results are in line with those of previous studies (see e.g.
 \citet{Eggenberger2012}, \citet{Piau2001}, \citet{Pinsonneault1989}).
-Figure (\ref{fig:li_var_vel_0g_z1}) shows a zoomed-in view of evolution tracks from the ZAMS till the TAMS in which the aforementioned effects caused by rotation are easily recognized.
-As depicted in figure (\ref{fig:li_var_vel_0g_z1}) at the end of the PMS, the rotating models exhibits a lower surface Li abundance than the non-rotating one.
-We thus see that the inclusion of rotation results in a global increase of the Li depletion during the PMS.
-We also note that the Li depletion occurs earlier for the rotating model than for the non-rotating.\par
+The figure (\ref{fig:hr_var_vel_0g_z1}) shows a zoomed-in view of evolution tracks from the ZAMS till the TAMS in which the aforementioned effects caused by rotation are even more easily recognizable.
+For example, if we compare the non-rotating model with the rotating ones we can recognize that at the end of the PMS, the latter reach before the ZAMS and exhibit a lower $T_{eff}$ than the non-rotating one.\par
 
@@
 If we concentrate our attention on figure (\ref{fig:li_var_vel_4_0g}) and compare it with figure (\ref{fig:li_var_vel_0g}) we will see how the profiles of Li abundance are already altered during PMS and MS.
-In the first phase we can describe the effect as modest, somewhat expected and in line with the fact that the loss of angular momentum caused by magnetic braking (see equation (\ref{eq:j_dot})) depends directly on the loss of mass.
-If we take into account that for solar type stars the models predict a modest total loss of mass, this is even much lower during PMS.
-On the contrary, during the MS it is observed in a much more evident way how the loss of angular momentum is much more significant, causing the star to rotate more slowly and as a result a smaller amount of Li to be destroyed.\par
+In the first phase we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see equation (\ref{eq:j_dot})) depends directly on the loss of mass.
+If we take into account that for solar-type stars the models predict a modest total loss of mass, this is even much lower during PMS.
+On the contrary, during the MS it is observed in a much more evident way how the AML is much more significant, causing the star to rotate more slowly and as a result a smaller amount of Li to be destroyed.\par
 
@@
 
-The effect of the magnetic braking routine can be seen even more clearly in figure (\ref{fig:rot_vel_4g}).
+The effect of the MB routine can be seen even more clearly in figure (\ref{fig:rot_vel_4g}).
 Here we represent the rotation profiles on the surface of the stars for different rotating models.
 As it can be observed, and in a similar way to the evolution profiles of the Li on the surface commented in the previous paragraph, the effect of the routine is much more accentuated once the ZAMS is reached.
-If the evolution of the curves presented here with those of figure (\ref{fig:rot_vel_0g}) we see how the star, instead of continuing to increase its angular velocity, begins to slow down after having reached its maximum in the passage through the ZAMS.\par
+If we compare the evolution of the curves presented here with those of figure (\ref{fig:rot_vel_0g}) we see how the star, instead of continuing to increase its $\Omega$, begins to slow down after having reached its maximum in the passage through the ZAMS.\par
 
-We also find it interesting to highlight the behaviour of angular velocity profiles when approaching TAMS.
-As can be seen in figure (\ref{fig:rot_vel_4g}) and in more detail in (\ref{fig:rot_vel_4g_z1}), we find that the evolution of angular velocity suffers from a series of ups and downs that do not appear previously.
-Everything seems to indicate that this behavior is due to a combined effect of the magnetic braking routine and the time-step (probably to big) used by MESA during the last part of the simulation.
+We also find interesting to highlight the behaviour of $\Omega$ profiles when approaching TAMS.
+As can be seen in figure (\ref{fig:rot_vel_4g}) and in more detail in (\ref{fig:rot_vel_4g_z1}), we find that the evolution of $\Omega$ suffers from a series of ups and downs that do not appear previously.
+Everything seems to indicate that this behavior is due to a combined effect of the MB routine and the time-step (probably to big) used by MESA during the last part of the simulation.
 We based this affirmation on the fact that this effect is not observed in the series of simulations in which the routine is not included (figs.
@@
 
-It's also remarkable to highlight the lower angular velocity observed in figure (\ref{fig:rot_vel_cz_4g} at the star surface in comparison with the upper limit of the convective layer.
+It's also remarkable to highlight the lower $\Omega$ observed in figure (\ref{fig:rot_vel_cz_4g} at the star surface in comparison with the upper limit of the convective layer.
 This fact is most probably a consequence of the expansion of the superficial layers after the mass loss produced by stellar winds.\par
 
-As mentioned above, the magnetic braking routine distributes the total amount of AML calculated according to equation (\ref{eq:k_jdot}) among the different layers that compose it.
-In figure (\ref{fig:cz_vc_028_var_b}) we can observe the evolution of the most external convective zone normalized with respect to the radius of the star.
-In accordance with the established models of stellar evolution, in a solar-type star the convective zone covers practically all of it for a large part of the PMS.
+As mentioned above, the MB routine distributes the total amount of AML calculated according to equation (\ref{eq:k_jdot}) among the different layers that compose it.
+In figure (\ref{fig:cz_vc_028_var_b}) we can observe the evolution of the most external CZ normalized with respect to the radius of the star.
+In accordance with the established models of stellar evolution, in a solar-type star the CZ covers practically all of it for a large part of the PMS.
 As it approaches ZAMS it is decreasing, as a consequence of the appearance of a radiative core, maintaining an approximately constant radius until the final stage of the MS, where it increases significantly as a response to the generalized increase of the star's radius.
-Regarding the effect of magnetic braking on the size of the convective zone, we observe that as the intensity of the magnetic field increases, the size of the convective zone decreases.
-According to \citet{Jeffries2004} this direct consequence is expected because the magnetic fields in the convective zone raise the adiabatic temperature gradient and accelerate the formation of a radiative core.
-This radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the BCZ drops below the one required for burning Li.
+Regarding the effect of MB on the size of the CZ, we observe that as the intensity of the magnetic field increases, the size of the CZ decreases.
+According to \citet{Jeffries2004} this direct consequence is expected because the magnetic fields in the CZ raise the adiabatic temperature gradient and accelerate the formation of a radiative core.
+This radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base drops below the one required for burning Li.
 This effect is most evident during MS (fig.
 (\ref{fig:cz_vc_028_var_b_z1})).
-The decrease in the size of the convective zone is in line with the fact that less Li is destroyed by causing less star material to reach areas with temperatures above $T_{Li}$.
-\par
+The decrease in the size of the CZ is in line with the fact that less Li is destroyed by causing less star material to reach areas with temperatures above $T_{Li}$.\par
 
-Similar to the evolution of the angular velocity towards the end of the MS discussed in the previous paragraph, we observe that the size of the convective zone also shows a series of ups and downs.
+Similar to the evolution of the $\Omega$ towards the end of the MS discussed in the previous paragraph, we observe that the size of the CZ also shows a series of ups and downs.
 Again we think that this effect is a direct consequence of the temporal resolution used by MESA in the stellar evolution equations.
-The specific origin of these fluctuations is a subject that we will deal with in more depth and detail the origin of these fluctuations in subsequent works to the current one.\par
+The specific origin of these fluctuations is a subject that we will deal with in more depth and detail in subsequent works.\par
 
@@
 In the first stages of PMS is where the greatest loss of mass is concentrated and this diminishes as it approaches the ZAMS.
-If we add to the scenario the fact that the star also decreases its radius during the PMS we have as a result that the star increases $\Omega$ obeying the principle of conservation of angular momentum.
-When reaching the ZAMS, the radius of the star remains more or less stable for much of the MS (except for its final stage) but continues to lose mass so it continues to increase its angular velocity, although in a less aggressive way if we compare it with the PMS.
-However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a convective zone (see fig.
-(\ref{fig:mb_act_var_vel_vc_028})), the magnetic braking routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
+If we add to the scenario the fact that the star also decreases its radius during the PMS we have as a result that the star increases $\Omega$ obeying the principle of conservation of AM.
+When reaching the ZAMS, the radius of the star remains more or less stable for much of the MS (except for its final stage) but continues to lose mass so it continues to increase its $\Omega$, although in a less aggressive way if we compare it with the PMS.
+However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see fig.
+(\ref{fig:mb_act_var_vel_vc_028})), the MB routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
 The more intense the magnetic field, the greater the braking effect.\par
@@
 
-\begin{figure*}
-    \centering
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_0_0g.eps}
-    \label{fig:subim1}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_3_5g.eps}
-    \label{fig:subim2}
-    \end{subfigure}
+\section{Conclusions and future works}
+Throughout this work we are confident that we have demonstrated that both rotation and MB induced effects must be taken into account in the theoretical models for a proper understanding and modelling of the evolution stars.
+The MB effect is of major importance for stellar evolution, both for the transport of AM and for the transport of chemical elements.
+We are also well aware that we are still far from understanding the exact physical mechanisms that govern these processes, so it is necessary to continue to delve into these areas of study.
+Of particular interest for us is the study of the evolution of the magnetic fields during the PMS and MS and their impact on AML.
+Future and improved implementations of our routine will be made to study the results on tracks, surface Li composition, rotation, etc...
+It is likely that in a rotating star calculated with MB from the beginning, the differential rotation is very much reduced and Li abundances observed in young stars could be explained.
+It is equally important to better understand the general role of mass loss in AML and particularly in colder, low-mass stars, since for the time being, among other factors, it is not easy to determine the terminal velocity of the stellar wind for this type of stars.\par
 
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_4_0g.eps}
-    \label{fig:subim3}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_4_5g.eps}
-    \label{fig:subim4}
-    \end{subfigure}
+It is also important to emphasize that our models fail to match the observed solar Li abundance and $\Omega$.
+We cannot, therefore, ascertain to have correctly modeled the rotational history of the Sun.
+In addition, we cannot (yet) properly explain the ups and downs observed in the $\Omega$ at the end of the MS.
+Therefore, in view of these shortcomings in our models, we must analyse the results obtained with caution and not draw any premature conclusions.
+Having said that, we are convinced that our research is on the right track and we do believe that the following conclusions, based on the results described in previous sections, have been sufficiently supported in the course of this work:
+\begin{enumerate}
+    \item The $T_{eff}$ from standard evolutionary tracks represent upper limits since these models do not include the magnetic field effect on the thermal structure.
+    \item The combination of rotation during the PMS and MB effect during MS produces different, potentially more promising behavior than those produced by standard models.
+    \item Inclusion of the magnetic field leads to cooler models and lower Li depletions in the MS.
+    \item MB during PMS seems to be also required for explaining Li abundances in young clusters.
+\end{enumerate}
 
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_5_0g.eps}
-    \label{fig:subim5}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_5_5g.eps}
-    \label{fig:subim6}
-    \end{subfigure}
-\caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
-The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
-The purple star is the surface Li abundance for the present-day Sun \citep{Asplund2009}.
-The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
-\label{fig:image2}
-\end{figure*}
-\par
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_vc_0084_var_b.eps}
-    \label{fig:subim21}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_vc_014_var_b.eps}
-    \label{fig:subim22}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_vc_0196_var_b.eps}
-    \label{fig:subim23}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_vc_028_var_b.eps}
-    \label{fig:subim24}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_vc_0336_var_b.eps}
-    \label{fig:subim25}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/blank.eps}
-    \label{fig:subim26}
-    \end{subfigure}
-\caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
-The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
-The purple star is the surface Li abundance for the present-day Sun \citep{Asplund2009}.
-The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
-\label{fig:image22}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/mb_act_3_5g.eps}
-    \label{fig:subim31}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/mb_act_4_0g.eps}
-    \label{fig:subim32}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/mb_act_4_5g.eps}
-    \label{fig:subim33}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/mb_act_5_0g.eps}
-    \label{fig:subim34}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/mb_act_5_5g.eps}
-    \label{fig:subim35}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/blank.eps}
-    \label{fig:subim36}
-    \end{subfigure}
-\caption{The evolution of magnetic braking activation, as a function of time and the existence of a radiative core for several 1 $M_{\sun}$ models.
-The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
-The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
-\label{fig:image32}
-\end{figure*}
-
-\begin{figure}
-	\includegraphics[width=\columnwidth]{figures/li_var_vel_3_5g.eps}
-    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
-The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
-The purple star is the surface Li abundance for the present-day Sun \citep{Asplund2009}.
-The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
-    \label{fig:li_var_vel_3_5g}
-\end{figure}
-
-In Figs. 2 and 3, we notice several interesting new facts for models with a magnetic field:
-- the models have almost an internal solid body rotation, with a difference of only a few percent between the core and the bulk of the envelope, with a short transition at their inter- face.
-The differential rotation parameter q is very small, but still different from zero.
-The relative constancy of $\Omega$ is the result of the high value of the transport coefficient ν by the magnetic field (see below Fig.
-8).
-We notice that these results are in agreement with those of Heger et al.
-(2004), who also find essentially solid body rotation until the end of the MS phase.
-Then in later phases, differential rotation appears together with a slowing down of the core;
-- there is a general slowing down of rotation during evolution due to two effects: the mass loss which removes angular momentum and the expansion of the external layers.
-The strong internal coupling of rotation transports this slowing down to the core;
-- models M 1 and M 5, although not strictly identical (see Fig. 4), exhibit similar $\Omega$ distributions;
-- we notice a small decrease of $\Omega$ near the surface. This is likely resulting from the expansion of the superficial layers after the removal of some mass by stellar winds at the surface.
-
-red lines are models with less and more efficient overshoot, resulting in reduced and enhanced lithium depletion, respectively. The three dot-dashed lines show additional variations in input
-physics: the red lines correspond to a MLT = 1.7 , while the orange and green lines are models that include PMS rotation with v v crit = 0.01 and 0.10, respectively.
-As expected, a lower a MLT produces a puffier, cooler star, resulting in less lithium depletion.
-The inclusion of rotation during the PMS produces very different, potentially more promising behavior, though the current absence of magnetic braking (see Sec-
-tions 3.6.5 and 3.5) in MESA results in an unrealistically high rotation speed ( v v crit between 0.1 and 1) on the MS.
-The models presented here fail to simultaneously match both the young and present-day solar surface lithium abundances.
-
-\section{Conclusions}
-The inclusion of rotation during the PMS and magnetic braking during MS produces very different, potentially more promising behavio, though magnetic braking during PMS seems to be also required for explaining Li abundances in young clusters.
-
-The Teff’s from standard evolutionary tracks represent upper limits since these models do not include the magnetic field effect on the thermal structure.
-Inclusion of the magnetic field leads to cooler models and lower Li depletions in the PMS.
-
-¿Para la cabecera?
-This generally neglected physical input ( thermal modifications that are due to the magnetic field) affects the location of the PMS: tracks that do not allow for the effect of the magnetic field on the convective Teff gradients provide an upper limit to the.
-This new feature in MESA is not equivalent to an artificial parametric change of the convection efficiency.
-We also point out that more complete models are needed, including a self-consistency model between the dynamo
-magnetic field and the evolution of stellar rotation.
-
-Mass loss by stellar winds drastically reduces vrot during the evolution (Packet
-et al 1980; Langer \& Heger 1998; Figs.
-3 and 4).
-Even if isotropic, the stellar winds carry away quite a lot of angular momentum, and this is even more important in the case of equatorial mass loss.
-The new surface layers then have a lower vrot as a result of expansion and redistribution.
-
-Indicar que se hace un planteamiento relativamente sencillo para la existencia de un campo magnético: existencia de núcleo radiativo y envelope convectivo.
-Para un tratamiento más avanzado se puede recurrir al documento (pag 10) "Stellar Evolution with rotation and magnetic fields" (Maeder \& Meynet)
-
-Página 6 de "On the revolution ofrotational velocity..." hay un planteamiento teórico para hacer dependiente la intensidad del campo magnético.
-Utilizar esto para futuros planteamientos.
-En la página 16 deriva la fórmula para el cálculo de la pérdida de momento angular.
-Utilizarla para enlazar con la versión simplificada ofrecida por Cantiello.
-
-The last numbered section should briefly summarise what has been done, and describe
-the final conclusions which the authors draw from their work.
-
-A T Tauri star is a type of pre–main sequence star that is being heated through gravitational contraction and has not yet begun to burn hydrogen at its core.
-They are variable stars that are magnetically active.
-The magnetic field of these stars is thought to interact with its strong stellar wind, transferring angular momentum to the surrounding protoplanetary disk.
-This allows the star to brake its rotation rate as it collapses.
-PODRÍAMOS INDICAR QUE YA QUE NO ACTIVAMOS EL FRENADO MAGNÉTICO DURANTE ESTA FASE SERÍA FACTIBLE QUE LA CONCENTRACIÓN DE LI FUERA SUPERIOR A LA QUE TENEMOS EN EL MODELO, ACORDE CON PLEIADES, Y QUE EL FRENADO MAGNÉTICO DURANTE LA FASE T-TAURI PERMITIESE QUE SE LLEGASE A ESE VALOR Y LUEGO FUERA REDUCIENDO LA VELOCIDAD PARA QUE NO TERMINARA DESTRUYENDO MÁS DE LA CUENTA.
+In the near future, with the new version of our code including magnetic field evolution in MESA, we will focus our efforts on understanding how magnetic fields are linked to rotation (e.g.
+\citep{Maeder2005}, \citep{Morin2012}, \citep{Gondoin2018}), how their topologies are and how they evolves in time (e.g.
+\citep{Keppens1995}, \citep{Maeder2003a}, \citep{Vidotto2014}) in order to go beyond the treatment on the MB routine that has been done in this work.\par
 
 \section*{Acknowledgements}
-
-The Acknowledgements section is not numbered. Here you can thank helpful
-colleagues, acknowledge funding agencies, telescopes and facilities used etc.
-Try to keep it short. xxxx
+I am pleased to acknowledge Jieun Choi her kindness in responding to the different questions about her work necessary to reproduce the results obtained as far as Li is concerned, as well as sharing the MESA files that she used.
+I wish also to thank Matteo Cantiello and Bill Paxton for their very valuable support in the development of the magnetic braking routine in MESA.
+Finally, I wish to acknowledge the generous work and help offered by the MESA community.
 
@@
 
-%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-
-\section{Some extra material}
-
-If you want to present additional material which would interrupt the flow of the main paper,
-it can be placed in an Appendix which appears after the list of references.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 % Don't change these lines

### 2019-06-23 project/overleaf/paper.tex
Paper v2.1

@@
 \begin{abstract}
-\textit{Goals.} The study the influence of rotation and magnetic braking (MB) on lithium (Li) depletion during pre-main sequence (PMS) and main sequence (MS) for solar-type stars.
+\textit{Goals.} The study of the influence of rotation and magnetic braking (MB) on lithium (Li) depletion during pre-main sequence (PMS) and main sequence (MS) for solar-type stars.
 \newline\textit{Methods.} The effects of rotational mixing and of the hydrostatic effects of rotation on Li abundances are studied by computing rotating PMS \& MS models.
@@
 \section{Introduction}
-Despite decades of theoretical efforts, a coherent explanation cannot be found from the models for the discrepancies found in the comparisons between the abundance of Li for stars belonging to clusters of different ages and that are also in the same evolutionary state either in the PMS or in the MS.
+Despite decades of theoretical efforts, a coherent explanation cannot be found from the models for the discrepancies found in the comparisons between the abundance of Li for stars belonging to clusters of different ages and that are also in the same evolutionary state, either in the PMS or in the MS.
 Additionally, these theoretical models are not able to explain the abundances detected in the late stages of the MS \citep{Tschape2001}.\par
@@
 Li burns at a temperature $T_{Li} \approx 2.5 x 10^6\; K$, and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value.
-solar-type stars are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $T_{Li}$ \citep{Iben1965}.
-This depletion stops during the approach to the zero-age MS (ZAMS), when the convection zone retreats and the temperature at BCZ is cooler that $T_{Li}$.
+Solar-type stars are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $T_{Li}$ \citep{Iben1965}.
+This depletion stops during the approach to the zero-age MS (ZAMS), when the convection zone retreats and the temperature at BCZ is cooler than $T_{Li}$.
 In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances, and remain nearly fixed until the terminal-age MS (TAMS).
@@
 Of particular interest for this work is the impact caused by the magnetic field of the star in the AML, more specifically the effect produced on it by MB as a consequence of the solidary rotation that produces on the particles that constitute the solar wind and the torque derived from this coupling.
-These magnetic fields would originate in the limits between the radiative and CZ's of the star, in the so-called tachoclines, due to the rotational difference between both zones \citep{Guerrero2016}.
+These magnetic fields would originate in the limits between the radiative and convective zones of the star, in the so-called tachoclines, due to the rotational difference between both zones \citep{Guerrero2016}.
 A proper consideration of the interaction between rotation and magnetic fields is therefore essential.
-We have adapted our one-dimensional stellar rotation code to model MB and include its effect on the equations for the evolution of the AM distribution and stellar structure.
+We have adapted the one-dimensional stellar evolution code used in this work to model MB and include its effect on the equations for the evolution of the AM distribution and stellar structure.
 This produces a more complete, though still reasonably simple, model for AM evolution.
@@
 
-We define some conventions and assumptions adopted in the paper.
+We define here also some conventions and assumptions adopted in the paper.
 For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity, not the current photospheric metallicity, as the reference value.
@@
 It is also responsible for inducing through hydrodynamic instabilities the mixing of the chemical elements that make it up.
-On the other hand, in order to be able to correctly model the effects induced in the mixing by rotation and for the results obtained from the theoretical models that implement this physical effect to be consistent with the observational data (see \citet{Pinsonneault1997}, \citet{Maeder2000}), it is necessary to analyze and understand more precisely the mechanisms that govern the distribution of angular momentum, since in this area the theoretical models also offer deficiencies to explain the observations \citep{Denissenkov2007}.
+On the other hand, in order to be able to correctly model the effects induced in the mixing by rotation and for the results obtained from the theoretical models that implement this physical effect to be consistent with the observational data (see \citet{Pinsonneault1997}, \citet{Maeder2000}), it is necessary to analyze and understand more precisely the mechanisms that govern the distribution of AM, since in this area the theoretical models also offer deficiencies to explain the observations \citep{Denissenkov2007}.
 \par
@@
 However, it has been proven sufficient to solve the equations in only one dimension, extrapolating the results to the other two, if it is fulfilled that the angular velocity is constant in regions subjected to the same pressure (isobaric).
-This is what is known as shellular approximation \citet{Meynet1997} and is based on calculating the impact of the centrifugal acceleration caused by the rotation in the stellar evolution equations \citep{Endal1976}.
+This is what is known as shellular approximation \citep{Meynet1997} and is based on calculating the impact of the centrifugal acceleration caused by the rotation in the stellar evolution equations \citep{Endal1976}.
 MESA uses this simplification to accelerate computational calculations.\par
@@
     \item The centrifugal force causes the local effective gravity ($g_{eff}$) associated with a point of the star not located on the axis of rotation to be reduced.
-    \item In general, the centrifugal force will not be parallel to the gravitational force and as a consequence of this the equipotential layers will cease to be spherical.
-    \item As the von Zeipel theorem states, that the radiative flux in a uniformly rotating star is proportional to the $g_{eff}$, therefore the radiative flux is not constant on an equipotential surface.
+    \item In general, the centrifugal force will not be parallel to the gravitational force and, as a consequence of this, the equipotential layers will cease to be spherical.
+    \item As the von Zeipel theorem states, the radiative flux in a uniformly rotating star is proportional to the $g_{eff}$, therefore the radiative flux is not constant on an equipotential surface.
     \item Rotation can suppress certain mechanisms that trigger convection, indirectly affecting mixing processes.
@@
 Observational data support the fact that stars are born with different angular velocities even if they come from the same interstellar cloud and that throughout evolution they undergo a deceleration process \citep{Skumanich1972} in the time period between 0.5 and 5 Gyr \citep{Somers2014a}.
-This range of different velocities implies that the mixing mechanisms induced by rotation vary their efficiency and effectiveness and, therefore, have a direct influence on the abundances of the different chemical elements observed on the surface of stars.
-This fact, together with the differential degree of rotation to which the limits between the radiative and convective zones of the star are subjected, is of particular interest for our study since Li, as mentioned in the introduction, is destroyed at a low temperature, so variations in the intensity of the mixing processes can cause convective zones to extend to interior zones with temperatures higher than $T_{Li}$ in which Li is destroyed.
+This range of different velocities implies that the mixing mechanisms induced by rotation vary their efficiency and effectiveness and therefore have a direct influence on the abundances of the different chemical elements observed on the surface of stars.
+This fact, together with the differential degree of rotation to which the limits between the radiative and convective zones of the star are subjected, is of particular interest for our study since Li, as mentioned in the introduction, is destroyed at a low temperature, so variations in the intensity of the mixing processes can cause convective zones to extend to interior zones with temperatures higher than $T_{Li}$.
 The deceleration process to which a star is subjected throughout its evolution and the direct link that the rotation has with the mixing processes also offer a natural and consistent explanation for the fact that the destruction of Li is more accentuated or efficient in young stars than in old stars \citep{Sestito2005}.
@@
 
-Additionally, and according to the prevalent theory of rotational mixing, meridional circulation and internal shears leads to a homogenization of the material depleted of Li found within the CZ, causing the inner layers where there is a temperature higher than $T_{Li}$, and therefore where the Li is destroyed, been mixed with those cooler above them.
+Additionally, and according to the prevalent theory of rotational mixing, southern circulation and internal shears leads to a homogenization of the material depleted of Li found within the CZ, causing that the inner layers where there is a temperature higher than $T_{Li}$ and where Li is destroyed, been mixed with those cooler above them.
 These mechanism causes a positive feedback loop in which the cooler material sinks, is heated again and the remaining Li is destroyed, totally or partially, before proceeding to rise once more into cooler layers.
@@
 A key question is to understand how these young stars can accrete large amounts of material from their circumstellar disk with high specific angular momentum and at the same time they're able to keep slow rotation periods, between 7-10 days \citep{Hussain2014}.
-One plausible explanation for this behaviour considers the loss of angular momentum produced by the interaction between the magnetic field and the particles that form its circumstellar disk \citep{Zanni2012}.
+One plausible explanation for this behaviour considers the AML produced by the interaction between the magnetic field and the particles that form its circumstellar disk \citep{Zanni2012}.
 Once accretion stops or becomes less efficient then stars are free to spin up as they contract towards the MS.\par
@@
 a dynamo type, continued in time to explain its existence.
-The way in which this dynamo process is able to generate magnetic fields depends on different parameters and to this day is still not fully accepted and is subject of controversy (e.g.
-\citep{Charbonneau2010}).\par
+The way in which this dynamo process is able to generate magnetic fields depends on different parameters and to this day this mechanism is still not fully accepted and is subject of controversy, e.g.
+\citep{Charbonneau2010}.\par
 
@@
 The dashed blue line marks the position of the birthline.
-All stars with $1.0 \; M_{\sun}$ will undergo a stage along their pre-main sequence evolution in which they have a fully convective interior near the birthline, to developing a radiative core when it's approaching the ZAMS (black dot-dashed line) \citet{Hussain2014}}
+All stars with $1.0 \; M_{\sun}$ will undergo a stage along their PMS evolution in which they have a fully convective interior near the birthline, to developing a radiative core when it's approaching the ZAMS (black dot-dashed line) \citet{Hussain2014}}
     \label{fig:hr_hussain}
@@
 From this moment on, the MB routine will be activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participate in the star AML.
-Additionally, based on our simple approach, we assume that the magnetic field does not vary its intensity throughout the evolution of the star until it reaches TAMS.\par
+Additionally, we assume that the magnetic field does not vary its intensity throughout the evolution of the star until it reaches TAMS.\par
 
-Our main objective for this paper is to contrast, in a first approximation, the degree of validity of an approach that includes MB as a complementary mechanism of AML that can contribute to explain the evolution of Li in solar-type stars.
+Our main objective for this paper is to check, in a first approximation, the degree of validity of an approach that includes MB as a complementary mechanism of AML that can contribute to explain the evolution of Li in solar-type stars.
 More advanced models of the arrangement of the magnetic field as well as the evolution of its intensity throughout the life of the star are left for later work.\par
@@
 According to the theoretical models this magnetic field is generated by a dynamo type effect in the tachocline, the intermediate zone between the radiative core and the CZ \citep{Aschwanden2014}, and has an intensity of approximately $B\approx10^5\; G$.
-Additionally, these models predict that because of the differential rotation existing between the core and CZ, the lines of the magnetic field that at the moment of their generation have a poloidal orientation will be affected by the differential rotation and forced to adopt a toroidal typology; the $\omega$-effect.
-On the other hand, we have that the opposite case can occur, that is to say, starting from a toroidal magnetic field to obtain a poloidal configuration.
+Additionally, these models predict that because of the differential rotation existing between the core and CZ, the lines of the magnetic field that at the moment of their generation have a poloidal orientation will be affected by the differential rotation and forced to adopt a toroidal typology.
+This change of typology in the magnetic field, from poloidal to toroidal, is known as the $\omega$-effect.
+On the other hand, we have that the opposite case can also occur, that is to say, starting from a toroidal magnetic field to obtain a poloidal configuration.
 This would be possible, on the one hand, thanks to the convective movements of material that are produced in the CZ of the star and, on the other, to assume that this material behaves as a perfect conductor.
-This second condition allows us to assume that the magnetic field lines are "frozen" in the conductive fluid and therefore must move in solidarity with it.
+This second condition allows to assume that the magnetic field lines are "frozen" in the conductive fluid and therefore must move in solidarity with it.
 Under these circumstances, the magnetic field lines will be pushed to higher areas of the CZ as the material moves vertically and, due to the combined effect of stellar rotation plus Coriolis forces, the magnetic lines will be twisted and give rise to a poloidal magnetic field.
@@
 As mentioned in the previous section, the dynamo model establishes that the magnetic field produced inside the star reaches the surface with the help of convective processes taken place in the CZ and gives rise to magnetically active regions on they reach the solar surface.
-Among them, we can find sunspots dominated by magnetic fields with an approximate an intensity of $B\approx10^3\; G$ and coronal loops with field strengths of $B\approx10^2\; G$ at the photosphere, and $B\approx10 G$ in larger coronal heights \citep{Aschwanden2014}.
-Additionally, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism (see, e.g.
+Among them, we can find sunspots dominated by magnetic fields with an approximate intensity of $B\approx10^3\; G$ and coronal loops with field strengths of $B\approx10^2\; G$ at the photosphere, and $B\approx10 G$ in larger coronal heights \citep{Aschwanden2014}.
+Besides, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism (see, e.g.
 \citet{Brandenburg2004}, \citet{Charbonneau2010} or \citet{Brun2017}).
-Whatever the origin of surface magnetic fields, these are expected to couple to the wind mass-loss and, if this is strong enough, to produce MB \citet{UdDoula2002}, \citet{Ud-Doula2007}, \citet{Ud-Doula2008} \citet{Meynet2010}.\par
+Whatever the origins of surface magnetic fields are, it's expected to couple to the wind mass-loss and, if this is strong enough, to produce MB \citet{UdDoula2002}, \citet{Ud-Doula2007}, \citet{Ud-Doula2008} \citet{Meynet2010}.\par
 
@@
 It has been also observed that strongly magnetic intermediate-mass stars typically have rotation rates much slower than other stars in their parent population \citep{Mathys2006}.
-In those stars, the presence of such magnetic field will interact with the mass loss and the Alfv\'{e}n radius $R_{A}$ play an important role in this interaction.
+In those stars, the presence of such magnetic field will interact with the mass loss and the Alfv\'{e}n radius $R_{A}$ plays an important role in this interaction.
 The $R_{A}$ is defined as the point in which the magnetic field energy density $B^{2}/8\upi$ and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$ of the stellar wind are balanced, that is, where the coefficient between both is equal to 1.
@@
 This energy ratio (eq.
-\ref{eq:wind_conf}) denominated wind confinement magnetic parameter $\eta_*$, defines a characteristic parameter for the relative effectiveness of the magnetic fields in circumscribing and/or channeling the wind outflow \citep{UdDoula2002}.
-In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval $\Dot{M}$, and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity $\nu_{esc}$.\par
+\ref{eq:wind_conf}) is denominated wind confinement magnetic parameter ($\eta_*$) and defines a characteristic parameter for the relative effectiveness of the magnetic fields in circumscribing and/or channeling the wind outflow \citep{UdDoula2002}.
+In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval ($\Dot{M}$), and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity ($\nu_{esc}$).\par
 
@@
 
-In general, a magnetically channeled outflow will have a complex stream geometry, but for convenience, the second equality in equation simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate $\Dot{M}=4\upi r^{2}\rho\nu$.
-$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law $ \nu(r) = \nu_\infty (1-R_*/r)$, where $\nu_\infty$ is the terminal wind velocity defined as the velocity that the wind (the out-flowing matter) reaches at large distance from the central star, where it is not accelerated anymore by the wind driving force but its deceleration due to interaction with the interstellar medium (ISM) is negligible \citep{Niedzielski2002}.\par
+In general, a magnetically channeled outflow will have a complex stream geometry but for convenience, the expression (\ref{eq:mass_loss} simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate.
+$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law
+\begin{ceqn}
+\begin{equation}
+    \nu(r) = \nu_\infty (1-R_*/r)
+\end{equation}
+\end{ceqn}
+where $\nu_\infty$ is the terminal wind velocity defined as the velocity that the wind, the outflowing matter, reaches at large distance from the central star, where it is not accelerated anymore by the wind driving force but its deceleration due to interaction with the interstellar medium (ISM) is negligible \citep{Niedzielski2002}.\par
 
-The line-driven winds of massive OB stars have terminal velocities (see e.g. Lamers \& Cassinelli 2000) that scale with the photospheric escape velocity 
+The line-driven winds of massive OB stars have terminal velocities that scale with the photospheric escape velocity \citep{Lamers2000}
 \begin{ceqn}
@@
 
-Given that the version of MESA used in this work does include neither the treatment of magnetic fields and, therefore, nor for magnetic braking effects, it is necessary to extend the simulator to include this phenomenon.
-In the present work we have followed the theoretical work on magnetic braking in the Sun carried out by \citet{Weber1967}, making use of the expression (\ref{eq:j_dot}) for the calculation of the AML.
+Given that the version of MESA used in this work does include neither the treatment of magnetic fields and therefore, nor for magnetic braking effects, it is necessary to extend the simulator to include this phenomenon.
+In the present work we have followed the theoretical work on MB in the Sun carried out by \citet{Weber1967}, making use of the expression (\ref{eq:j_dot}) for the calculation of the AML.
 \begin{ceqn}
@@
 With regards to the MB routine implementation, we've based on \citet{Ud-Doula2007}.
-A set of different simulation scenarios with magnetic field strength ranging from between 3 and 10G have been exercised.
-We'd like to highlight (again) that the magnetic field remains constant during the PMS \& MS phases.
-We compute the evolution of $1M\sun$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.\par
+A set of different simulation scenarios with magnetic field strength ranging from between 3 and 10 G have been exercised.
+We'd like to stress again that the magnetic field remains constant during the PMS \& MS phases.
+We compute the evolution of $1M_{\sun}$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.\par
 
 \subsection{Evolving the angular velocity}
-As stated in the work of \citet{Paxton2015} makes its models evolve by default without taking into account the rotation but offers options to initialize models with a certain angular speed $\Omega$.
+As stated in the work of \citet{Paxton2015}, MESA makes its models evolve by default without taking into account the rotation but offers options to initialize them with a user defined angular speed $\Omega$.
 The initial value for $\Omega$ can be indicated as a function of the rotation speed of the star surface expressed in $km\,s^{-1}$ or using the critical rotation speed as a reference (option used in our case).
-Once the model has been initialized with a specific $\Omega$, it is recalculated as the star evolves as a function of the mass, radius and AML that the simulator determines in each time step.
+Once the model has been initialized with a specific $\Omega$, it is recalculated as the star evolves as a function of its mass, radius and AML that the simulator determines in each time step.
 \par
@@
 The number of cells does not remain fixed throughout the simulation but varies according to certain evolutionary parameters (see \citet{Paxton2015} for more details).
-The $\Omega$ for each cell $k$ ($\Omega_k$) is adjusted so that the resulting angular momentum is retained after adjusting the new mass of the cell ($m_k$) according to the loss or gain suffered by the solar wind and calculate the distance from the cell to the center of the star ($r_k$).\par
+The $\Omega$ for each cell $k$ ($\Omega_k$) is adjusted so that the resulting angular momentum is retained after adjusting the new mass of the cell ($m_k$), according to the loss or gain suffered by the solar wind, and the distance from the cell to the center of the star ($r_k$).\par
 
@@
 It's in this point when our MB routine plugs in and allows us to influence the internally calculated value by providing an additional contribution calculated by the user routine.
-Our implementation of the MB routine calculates the additional AML ($\Dot{J}_{k}$), derived from the external torque exerted by the magnetic field, to be taken into account and distributes it through the different layers that make up the CZ, according to the expression (\ref{eq:j_dot}.\par
+Our implementation of the MB routine calculates the additional AML ($\Dot{J}_{k}$), derived from the external torque exerted by the magnetic field and distributes it through the different layers that make up the CZ, according to the expression (\ref{eq:j_dot}).\par
 
@@
 To calibrate this law, Skumanich relied on data obtained from type G stars found in the MS \citep{Skumanich}.
-The influence of the MB on the evolution of the star is closely related to the transport of the AM and, by extension, influenced by how the chemical elements are transported within it.
-According to this approach \citep{Meynet2010}, two main mechanisms are distinguished, attending to how the transport of the AM is produced, which condition the mixing of these elements:
+The influence of the MB on the evolution of the star is closely related to the transport of the AM and by extension, influenced by how the chemical elements are transported within it.
+According to this approach \citep{Meynet2010}, two main mechanisms are distinguished attending to how the transport of the AM is produced which influence the mixing of these elements:
 
@@
 On the other hand, if we consider that the convective movements that take place in the stellar interior together with the presence of differential rotation can give rise to the generation of magnetic fields that end up reaching the surface of the star \citep{Langer2012}, the AML can be increased in several orders of magnitude as a consequence of the magnetic field that forces the ionized particles that compose the stellar wind to turn with the same angular speed and whose influence on the stellar wind extends a distance several times the radius of the star (see \citet{UdDoula2002}, \citet{Ud-Doula2007} and \citet{Ud-Doula2008} for more details).
-This distance is directly influenced by the radius of alven (see eq.
-(\ref{eq:j_dot})) or alternatively by the confinement parameter (see eq.
+This distance is directly influenced by the $R_A$ (see eq.
+(\ref{eq:j_dot})) or alternatively by the $\eta_*$ (see eq.
 (\ref{eq:j_dot_mesa})).
@@
 
-On the other hand, the initial abundance of $A(\isotope[7]{Li}) = 3.34 \, dex$ is obtained from meteorite measurements \citet{Randich2006}.
+On the other hand, the initial abundance of $A(\isotope[7]{Li}) = 3.34 \, dex$ is obtained from meteorite measurements \citep{Randich2006}.
 According to the theory for newly born stars, we have that the initial abundance of Li can be estimated fairly accurately from photospherical measurements on T-Tauri type stars, or from the hottest F stars forming part of slightly older clusters.
 For the latter, the current theory of stellar evolution suggests that Li must not yet be exhausted.
-The results obtained from the measurements on both types of stars allow us to fix the initial abundance of Li in the interval $3.0 \, dex < A(\isotope[7]{Li}) < 3.4 \, dex$ \citet{Randich2006}.\par
+The results obtained from the measurements on both types of stars allow us to fix the initial abundance of Li in the interval $3.0 \, dex < A(\isotope[7]{Li}) < 3.4 \, dex$ \citep{Randich2006}.\par
 
@@
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
-The solid black line represents the fiducial model according to \citet{Choi2016}.
+The solid black line represents the reference model according to \citet{Choi2016}.
 The rest of lines are models which include PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
@@
 Figure (\ref{fig:li_var_vel_0g}) shows the evolution of surface Li abundance relative to H as a function of time and for several $1M_{\odot}$ models which take into account the effects of rotation and AML caused by stellar winds but not MB.
-The purple star and square are surface Li abundance for the present-day Sun \citet{Asplund2009} and Pleiades surface Li abundance \citep{Sestito2005}.\par
+The purple star and square are surface Li abundances for the present-day Sun \citep{Asplund2009} and Pleiades \citep{Sestito2005}.\par
 
@@
 On the other hand, the other models which include rotation during the PMS with values of $v/v_{crit}$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but it rotational velocity is much higher (see fig.
-\ref{fig:rot_vel_0g}) than the $2 kms^{-1}$ of the Sun \citep{Gill2012} \par
+(\ref{fig:rot_vel_0g})) than the $2\,kms^{-1}$ of the Sun \citep{Gill2012}.
+\par
 
@@
 (\ref{fig:rot_vel_0g})) and this is because the star has a completely convective interior.
-It is not until the end of Hayashi track that the star begins to develop a radiative nucleus due to the increase in temperature that occurs inside the star during the contraction process.
-It is in this stage of evolution of the star when a difference of angular velocity appears between the upper and lower limits of the radiative and convective zones respectively.
-This effects can be clearly recognized in the figure (\ref{fig:rot_vel_cz_0g}) which shows, in addition to star surface $\Omega$, the velocity corresponding to the upper and lower limits of the CZ.
-The degree of differential rotation in radiative zone is directly influenced by the initial $\Omega$.
+It is not until the end of Hayashi track that the star begins to develop a radiative core due to the increase in temperature that occurs inside the star during the contraction process.
+It is in this stage of evolution of the star when a difference in angular velocity appears between the upper and lower limits of the radiative and convective zones respectively.
+This effects can be clearly recognized in the figure (\ref{fig:rot_vel_cz_0g}) which shows, in addition to star surface $\Omega$, the angular velocity corresponding to the upper and lower limits of the CZ.
+The degree of differential rotation is directly influenced by the initial $\Omega$.
 As the models are initialized with a higher angular velocity, the difference in velocity between the BCZ and the star surface is accentuated in a directly proportional mode; the higher the initial velocity, the bigger the velocity gradient between the bottom and top limits of the CZ.
@@
     \caption {The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models in the approach to the ZAMS (dashed lines).
-The solid black line represents the fiducial model according to \citet{Choi2016}.
+The solid black line represents the reference model according to \citet{Choi2016}.
 The rest of lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.}
@@
 The figure (\ref{fig:hr_var_vel_0g_z1}) shows a zoomed-in view of evolution tracks from the ZAMS till the TAMS in which the aforementioned effects caused by rotation are even more easily recognizable.
-For example, if we compare the non-rotating model with the rotating ones we can recognize that at the end of the PMS, the latter reach before the ZAMS and exhibit a lower $T_{eff}$ than the non-rotating one.\par
+For example, if we compare the non-rotating model with the rotating ones we can recognize that at the end of the PMS, the latter reach before the ZAMS and exhibit a lower $T_{eff}$ than the non-rotating one (solid black line).\par
 
@@
 
-It's also remarkable to highlight the lower $\Omega$ observed in figure (\ref{fig:rot_vel_cz_4g} at the star surface in comparison with the upper limit of the convective layer.
+It's also remarkable to highlight the lower $\Omega$ observed in figure (\ref{fig:rot_vel_cz_4g}) at the star surface in comparison with the upper limit of the convective layer.
 This fact is most probably a consequence of the expansion of the superficial layers after the mass loss produced by stellar winds.\par
@@
 According to \citet{Jeffries2004} this direct consequence is expected because the magnetic fields in the CZ raise the adiabatic temperature gradient and accelerate the formation of a radiative core.
-This radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base drops below the one required for burning Li.
+This radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base drops below $T_{Li}$.
 This effect is most evident during MS (fig.
@@
 	\includegraphics[width=\columnwidth]{figures/mb_act_var_vel_vc_028.eps}
-    \caption{The evolution of convective zone size, as a function of time for several 1 $M_{\sun}$ models.
+    \caption{The activation of the magnetic braking routine as a function of the presence of a radiative core.
 The models include a magnetic field with an intensity of 4G and PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
@@
 The horizontal dashed lines inform about the star's core nature: radiative (rad) or convective (conv).
+By implementation decision, once the routine is activated, it remains on even if the star's core nature change to convective.
 The vertical dashed lines make reference to the ZAMS.}
@@
 It is likely that in a rotating star calculated with MB from the beginning, the differential rotation is very much reduced and Li abundances observed in young stars could be explained.
-It is equally important to better understand the general role of mass loss in AML and particularly in colder, low-mass stars, since for the time being, among other factors, it is not easy to determine the terminal velocity of the stellar wind for this type of stars.\par
+It is equally important to better understand the general role of mass loss in AML and particularly in colder, low-mass stars, since for the time being and among other factors, it is not easy to determine the terminal velocity of the stellar wind for this type of stars.\par
 
@@
 \begin{enumerate}
-    \item The $T_{eff}$ from standard evolutionary tracks represent upper limits since these models do not include the magnetic field effect on the thermal structure.
+    \item The $T_{eff}$ from standard evolutionary tracks represent upper limits since these models do not take into consideration the magnetic braking effect.
     \item The combination of rotation during the PMS and MB effect during MS produces different, potentially more promising behavior than those produced by standard models.

### 2019-06-23 project/octave/plotting_routines.m

@@ -748,7 +748,7 @@ function plot_0G_var_vel()
   global gauss_fields;
   global rotational_vels; 
 
-  age_vs_li_plots(gauss_fields(1,:), rotational_vels(2:6,:), 'A(Li7) - 0G & variable rotational velocity', 0.5, [1.0e5,1.0e10,0,4.5]);
+  age_vs_li_plots(gauss_fields(1,:), rotational_vels(1:6,:), 'A(Li7) - 0G & variable rotational velocity', 0.5, [1.0e5,1.0e10,0,4.5]);
 end
 
 
@@ -756,7 +756,7 @@ function plot_0G_var_vel_z1()
   global gauss_fields;
   global rotational_vels; 
 
-  age_vs_li_plots(gauss_fields(1,:), rotational_vels(2:6,:), 'A(Li7) - 0G & variable rotational velocity', 0.1, [1.0e7, 1.0e8, 2.0, 2.5]);
+  age_vs_li_plots(gauss_fields(1,:), rotational_vels(1:6,:), 'A(Li7) - 0G & variable rotational velocity', 0.1, [1.0e7, 1.0e8, 2.0, 2.5]);
 end
 
 
@@ -1084,7 +1084,7 @@ function main()
   %plot_age_vs_mb_activation_028vc();
   
   
-  %plot_0G_var_vel();
+  plot_0G_var_vel();
   %plot_0G_var_vel_z1();
   %plot_3_5G_var_vel();  
   %plot_4_0G_var_vel(); 
@@ -1101,7 +1101,7 @@ function main()
   
   %plot_vel_rot_0G_var_vel();
   %plot_vel_rot_3_5G_var_vel();
-  plot_vel_rot_4G_var_vel();
+  %plot_vel_rot_4G_var_vel();
   %plot_vel_rot_4G_var_vel_z1();
   %plot_vel_rot_4_5G_var_vel();
   %plot_vel_rot_5G_var_vel();

### 2019-06-23 project/overleaf/paper.tex
Paper v2.2
Add grid images

@@
 
+\begin{figure*}
+    \centering
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_var_vel_0_0g.eps}
+    \label{fig:subim1}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_var_vel_3_5g.eps}
+    \label{fig:subim2}
+    \end{subfigure}
+
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_var_vel_4_0g.eps}
+    \label{fig:subim3}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_var_vel_4_5g.eps}
+    \label{fig:subim4}
+    \end{subfigure}
+
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_var_vel_5_0g.eps}
+    \label{fig:subim5}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_var_vel_5_5g.eps}
+    \label{fig:subim6}
+    \end{subfigure}
+\caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
+Each figure shows a set of models in which the magnetic field with intensity has been fixed and $v/v_{crit}$ varies between 0.0084 and 0.0336, respectively.
+The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
+The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
+\label{fig:image2}
+\end{figure*}
+\par
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_vc_0084_var_b.eps}
+    \label{fig:subim21}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_vc_014_var_b.eps}
+    \label{fig:subim22}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_vc_0196_var_b.eps}
+    \label{fig:subim23}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_vc_028_var_b.eps}
+    \label{fig:subim24}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/li_vc_0336_var_b.eps}
+    \label{fig:subim25}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/blank.eps}
+    \label{fig:subim26}
+    \end{subfigure}
+\caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
+Each figure shows a set of models in which $v/v_{crit}$ has been fixed and the magnetic field with intensity varies between 3.5 and 5.5 G, respectively.
+The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
+The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
+\label{fig:image22}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/rot_vel_0g.eps}
+    \label{fig:subim41}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/rot_vel_3_5g.eps}
+    \label{fig:subim42}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/rot_vel_4g.eps}
+    \label{fig:subim43}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/rot_vel_4_5g.eps}
+    \label{fig:subim44}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/rot_vel_5g.eps}
+    \label{fig:subim45}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/rot_vel_5_5g.eps}
+    \label{fig:subim46}
+    \end{subfigure}
+\caption{Grid showing the evolution of the evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
+Each figure shows a set of models in which the magnetic field with intensity has been fixed and $v/v_{crit}$ varies between 0.0084 and 0.0336.
+The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
+The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
+\label{fig:image24}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/mb_act_3_5g.eps}
+    \label{fig:subim31}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/mb_act_4_0g.eps}
+    \label{fig:subim32}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/mb_act_4_5g.eps}
+    \label{fig:subim33}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/mb_act_5_0g.eps}
+    \label{fig:subim34}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/mb_act_5_5g.eps}
+    \label{fig:subim35}
+    \end{subfigure}
+    \begin{subfigure}[h]{0.47\textwidth}
+    \includegraphics[width=\textwidth]{figures/blank.eps}
+    \label{fig:subim36}
+    \end{subfigure}
+\caption{Grid showing the evolution of magnetic braking activation, as a function of time and the existence of a radiative core for several 1 $M_{\sun}$ models.
+Each figure shows a set of models in which the magnetic field with intensity has been fixed and $v/v_{crit}$ varies between 0.0084 and 0.0336.
+The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
+\label{fig:image32}
+\end{figure*}
+
 \section*{Acknowledgements}

### 2019-06-24 project/overleaf/paper.tex
Paper v2.3

@@
 \textit{Goals.} The study of the influence of rotation and magnetic braking (MB) on lithium (Li) depletion during pre-main sequence (PMS) and main sequence (MS) for solar-type stars.
-\newline\textit{Methods.} The effects of rotational mixing and of the hydrostatic effects of rotation on Li abundances are studied by computing rotating PMS \& MS models.
-The impact of the MB effect is then analyzed by comparing the Li abundances of those models experiencing different magnetic field intensities ranging between 3.5 and 5.5 G.
-\newline\textit{Results.} The surface Li abundance at the end of the PMS and throughout the MS is decreased when rotational effects are included.
-This results in an increment in the Li depletion rate for rotating models compared to non-rotating ones.
-This effect is attenuated when the braking effect produced by a magnetic field is added.
+\newline\textit{Methods.} The effects of rotational mixing and of the hydrostatic effects of rotation on Li abundances are studied by simulating several grids of PMS \& MS rotating and non-rotating models.
+The previous effects are combined with the additional impact of the MB one (with magnetic field intensities ranging between 3.5 and 5.5 G) and the data obtained from the two set of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
+\newline\textit{Results.} The surface Li abundance at the end of the PMS and throughout the MS is decreased when rotational effects are included, i.e.
+the Li depletion rate for rotating models is higher than for non-rotating ones.
+However, this effect is attenuated when the braking effect produced by a magnetic field is present.
 This physical phenomena, generally neglected in star evolutionary models, impacts also the effective temperature ($T_{eff}$) of the star and therefore its location in the HR diagram.
 The impact of MB in Li depletion is sensitive to the field intensity: the higher it is, the lower the Li destruction.
-A relationship between the magnetic fields and the convective zone (CZ) size is observed: stronger magnetic fields produce shallower CZ's.
+A direct link between the magnetic fields and the convective zone (CZ) size is observed: stronger magnetic fields produce shallower CZ's.
 Another important result of this study is that MB effect must be taken into consideration during PMS must if we aim to reproduce Li abundances in young clusters.
@@
 The denominated standard models, models that include convection only as a mixing process and do not consider other transport options like diffusion or angular momentum loss (AML), have been mainly involved in the elaboration of these predictions \citep{Sestito2005}.
-Li burns at a temperature $T_{Li} \approx 2.5 x 10^6\; K$, and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value.
+Li is destroyed at a temperature $T_{Li} \approx 2.5 x 10^6\; K$, and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value.
 Solar-type stars are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $T_{Li}$ \citep{Iben1965}.
 This depletion stops during the approach to the zero-age MS (ZAMS), when the convection zone retreats and the temperature at BCZ is cooler than $T_{Li}$.
-In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances, and remain nearly fixed until the terminal-age MS (TAMS).
+In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances, in addition to showing a very similar Li evolution until their approach to the terminal-age MS (TAMS).
 During this same period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the convective envelope, from its lower limit to the star surface.
@@
 
-The impact of rotation on PMS Li depletion of solar-type stars has already been discussed (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}).
+The impact of rotation on PMS and on Li depletion for solar-type stars has already been discussed (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}).
 These previous studies focused mainly on the hydrostatic effects and didn't consider the influence of a coupling between the star magnetic field and its possible spin-down effect.
@@
 We define here also some conventions and assumptions adopted in the paper.
-For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity, not the current photospheric metallicity, as the reference value.
-In other words, [Z/H] is computed with respect to $Z = Z_{\odot, protosolar} = 0.0142$, not $Z = Z_{\odot, photosphere} = 0.0134$.
-Lastly, in accordance with the XXIXth IAU resolutions B3 \citep{Mamajek2015} we adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10} cm$ and $M_{\odot} = 1.988x10^{33} g$.
+For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, proto} = 0.0142$, not the current photospheric metallicity $Z = Z_{\odot, photosphere} = 0.0134$, as the reference value.
+We also adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10} cm$ and $M_{\odot} = 1.988x10^{33} g$ which are consistent with IAU resolutions \citep{Mamajek2015} .
 For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}.
@@
 
-The paper is organized as follows.
-Section 2 gives an overview of rotation and magnetic fields physical processes and how do they impact on the star evolution.
+This paper is structured as follows: Section 2 gives an overview of rotation and magnetic fields physical processes and how do they impact on the star evolution.
 We also summarized the most relevant parameters configured in MESA for the model simulations.
@@
 \begin{enumerate}
-    \item AM transport from the radiative interior to the convective envelope in response to the rotational deceleration of the stellar surface layers\label{itm:1}.
+    \item AM transport from the radiative interior to the convective envelope in response to the rotational deceleration of the star surface layers\label{itm:1}.
     \item AM redistribution associated with changes in internal structure during the process of contraction to the MS\label{itm:2}.
@@
 
-With regards to MESA, the transport of AM and chemical elements due to rotationally induced instabilities is implemented in a diffusion approximation \citet{Endal1978}.
-MESA calculates diffusion coefficients for five different rotationally induced mixing processes: dynamical shear instability, Solberg-H{\o}iland instability, secular shear instability, Eddington-Sweet circulation, and the Goldreich-Schubert-Fricke instability \citep{Paxton2013}.
+In reference to MESA and as stated in \citet{Paxton2013}: "the transport of AM and chemical elements due to rotationally induced instabilities is implemented in a diffusion approximation \citet{Endal1978}.
+MESA calculates diffusion coefficients for five different rotationally induced mixing processes: dynamical shear instability, Solberg-H{\o}iland instability, secular shear instability, Eddington-Sweet circulation, and the Goldreich-Schubert-Fricke instability".
 
@@
 The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields (see \citet{Dudorov2014} for details about the Theory of fossil magnetic field), or fields produced through a dynamo mechanism \citep{Cantiello2009}.
-Besides that, it has long been a requirement for T Tauri stars to host kG magnetic fields \citep{Hussain2014} in order to explain several key observational characteristics in young, accreting PMS systems \citep{Johns-Krull2007}.
+Besides that, the presence of magnetic fields with intensities of the order of kG \citep{Hussain2014} has been a sine qua non condition for dealing with certain data observed in T Tauri type stars in young, accreting PMS systems \citep{Johns-Krull2007}.
 One of this features is the long rotating periods  which is much longer than predicted from AM conservation.
@@
 One plausible explanation for this behaviour considers the AML produced by the interaction between the magnetic field and the particles that form its circumstellar disk \citep{Zanni2012}.
-Once accretion stops or becomes less efficient then stars are free to spin up as they contract towards the MS.\par
+Once the accretion process stops or becomes less efficient then the star is able to spin up as it contracts towards the ZAMS.\par
 
@@
 	\includegraphics[width=\columnwidth]{figures/hussain_gaitee_fig1.png}
-    \caption{H-R diagram showing Behrend \& Maeder (2001) pre-main sequence evolutionary tracks for stellar masses up to $4.0 \; M_{\sun}$.
+    \caption{H-R diagram showing PMS evolutionary tracks for stars with masses up to $4.0 \; M_{\sun}$.
 The dashed blue line marks the position of the birthline.
-All stars with $1.0 \; M_{\sun}$ will undergo a stage along their PMS evolution in which they have a fully convective interior near the birthline, to developing a radiative core when it's approaching the ZAMS (black dot-dashed line) \citet{Hussain2014}}
+The stars with $1.0 \; M_{\sun}$ which have a fully convective interior near the Hayashi track, they will to develop a radiative core when they're approaching the ZAMS (black dot-dashed line).
+Image copied from \citet{Hussain2014}}
     \label{fig:hr_hussain}
@@
 \begin{enumerate}
-    \item The $T_{eff}$ from standard evolutionary tracks represent upper limits since these models do not take into consideration the magnetic braking effect.
-    \item The combination of rotation during the PMS and MB effect during MS produces different, potentially more promising behavior than those produced by standard models.
     \item Inclusion of the magnetic field leads to cooler models and lower Li depletions in the MS.
+    \item A combination of rotation during the PMS and MB effect during MS produces different, potentially more promising behavior than those produced by standard models.
+    \item $T_{eff}$ from standard evolutionary tracks represent upper limits since these models do not take into consideration the magnetic braking effect.
+    \item The convective zone extension decreases when the intensity of the magnetic field increases.
     \item MB during PMS seems to be also required for explaining Li abundances in young clusters.

### 2019-07-09 project/octave/plotting_routines.m
Increase font sizes in graphs

@@ -35,6 +35,8 @@ global sun_gauss_field = 1.0; %G
 global sun_age = 4.57e9; %years
 global sun_A_Li7 = 1.1;
 global sun_vel_rot = 2.0; %km/s
+global sun_T_eff = 5772; %K
+global sun_L = 3.828e26; %W
 
 %Pleiades constants
 global pleiades_age = 1.0e8; %years
@@ -45,9 +47,14 @@ global pleiades_A_Li7 = 2.95;
 global delta_h1_center = 0.0015;
 global ratio_log_LH_vs_log_L = 0.99;
 
-%Graphic size in points
-global W=12;
-global H=7;
+%Graphic size in inches
+global line_width = 3;
+global W = 18;
+global H = 14;
+global tick_font_size = 16;
+global title_font_size = 20;
+global axis_font_size = 18;
+global legend_font_size = 18;
 
 
 
@@ -167,6 +174,8 @@ function result = calculate_ZAMS(full_path)
 end
 
 function plot_ZAMS(A, color, width, ytick, axis_limits)
+  global tick_font_size;
+  
   %Plot values
   plot(A(:,1), A(:,2) .- A(:,3), color, 'linewidth', width);
   
@@ -184,6 +193,7 @@ function plot_ZAMS(A, color, width, ytick, axis_limits)
   yticks = get (gca, "ytick"); 
   ylabels = arrayfun (@(x) sprintf ("%1.2f", x), yticks, "uniformoutput", false); 
   set (gca, "yticklabel", ylabels);
+  set(gca, 'fontsize', tick_font_size);
 
 end
 
@@ -210,6 +220,8 @@ end
 
 
 function plot_A_Li7(A, B, color, width, ytick, axis_limits)
+  global tick_font_size
+  
   %Plot values
   plot(A(:,1), B(:,1), color, 'linewidth', width);
 
@@ -227,31 +239,45 @@ function plot_A_Li7(A, B, color, width, ytick, axis_limits)
   yticks = get (gca, "ytick"); 
   ylabels = arrayfun (@(x) sprintf ("%2.1f", x), yticks, "uniformoutput", false); 
   set (gca, "yticklabel", ylabels);
+  set(gca, 'fontsize', tick_font_size);
 
 end
 
-function plot_hr(A, color, width, ytick, axis_limits)
+function plot_hr(A, color, width, xy_ticks, axis_limits)
+  global tick_font_size;
   %Plot values
[…]

### 2019-07-09 project/overleaf/paper.tex
Paper v3.0
Improve graphs quality

@@
 In turn, the presence of this convective region conditions the evolution of magnetic fields.
-As depicted in figure (\ref{fig:hr_hussain}), massive stars do not have external convective zones although they can expose a convective core.
+Massive stars do not have external convective zones although they can expose a convective core.
 In these cases, the dominant type of energy transport in the outer zones of the star is radiative.
@@
 \citep{Charbonneau2010}.\par
-
-\begin{figure}
-	\includegraphics[width=\columnwidth]{figures/hussain_gaitee_fig1.png}
-    \caption{H-R diagram showing PMS evolutionary tracks for stars with masses up to $4.0 \; M_{\sun}$.
-The dashed blue line marks the position of the birthline.
-The stars with $1.0 \; M_{\sun}$ which have a fully convective interior near the Hayashi track, they will to develop a radiative core when they're approaching the ZAMS (black dot-dashed line).
-Image copied from \citet{Hussain2014}}
-    \label{fig:hr_hussain}
-\end{figure}
 
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/li_var_vel_0_0g.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip, width=\columnwidth]{figures/li_var_vel_0_0g.eps}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
@@
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
+The dashed lines make reference to the ZAMS.}
     \label{fig:li_var_vel_0g}
@@
 The solid black line represents the reference model that adopts the solar-calibrated envelope overshooting parameters as documented in \citet{Choi2016}.
-In general, all models burn too much Li early on and do not match with the Pleiades surface Li abundance.
+In general, all models burn too much Li before the ZAMS \footnote{Defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.} and do not match with the Pleiades surface Li abundance.
 Later, the reference model (black line) does not deplete Li efficiently on the MS and fail (again) to match the current solar surface Li abundance.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
     \caption {The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models in the approach to the ZAMS (dashed lines).
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/rot_vel_0g.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
     \caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/rot_vel_cz_0g.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
     \caption{The evolution of angular velocity at surface and limits of the uppermost convective zone, as a function of time for several 1 $M_{\sun}$ models.
@@
 \end{figure}
+%\includegraphics[trim = 35mm 95mm 40mm 100mm, clip, width=0.475\textwidth]
 
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/hr_var_vel_0g.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/hr_vc_0336_var_g_z1.eps}
     \caption{An example solar 1$M_{\sun}$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities. The rotation is activated in the models in the PMS.}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/hr_var_vel_0g_z1.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/hr_var_vel_0_0g_z1.eps}
     \caption{Similar to figure (\ref{fig:hr_var_vel_0g}) but now showing in detail the effects of rotation on the evolutionary tracks.}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/li_var_vel_4_0g.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_4_0g.eps}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/rot_vel_4g.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g.eps}
     \caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/rot_vel_cz_4g.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g.eps}
     \caption{The evolution of angular velocity at surface and limits of the uppermost convective zone, as a function of time for several 1 $M_{\sun}$ models.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/rot_vel_4g_z1.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g_z1.eps}
     \caption{Similar to figure (\ref{fig:rot_vel_4g}) but now showing in detail the surface rotational velocity as the star approaches TAMS.}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/cz_vc_028_var_b.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g.eps}
     \caption{The evolution of convective zone size, as a function of time for several 1 $M_{\sun}$ models.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/cz_vc_028_var_b_z1.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g_z1.eps}
     \caption{Similar to figure (\ref{fig:cz_vc_028_var_b}) but now showing in detail the convection zone size from the near ZAMS till the TAMS.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/mdot_vc_028_var_b.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g.eps}
     \caption{The evolution of mass loss $\Dot{M}$ , as a function of time for several 1 $M_{\sun}$ models.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/mdot_vc_028_var_b_z1.eps}
+	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g_z1.eps}
     \caption{The evolution of mass loss $\Dot{M}$ as the star is approaching TAMS, as a function of time for several 1 $M_{\sun}$ models.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/mb_act_var_vel_vc_028.eps}
+	\includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\columnwidth]{figures/mb_act_vc_028_var_g.eps}
     \caption{The activation of the magnetic braking routine as a function of the presence of a radiative core.
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_0_0g.eps}
+    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_0_0g.eps}
     \label{fig:subim1}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_3_5g.eps}
+    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_3_5g.eps}
     \label{fig:subim2}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_4_0g.eps}
+    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_0g.eps}
     \label{fig:subim3}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_4_5g.eps}
+    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_5g.eps}
     \label{fig:subim4}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_5_0g.eps}
+    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_5_0g.eps}
     \label{fig:subim5}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_var_vel_5_5g.eps}
+    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_5_5g.eps}
     \label{fig:subim6}
@@
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
+The dashed lines make reference to the ZAMS.}
 \label{fig:image2}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_vc_0084_var_b.eps}
+    \includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0084_var_g.eps}
     \label{fig:subim21}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_vc_014_var_b.eps}
+    \includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_014_var_g.eps}
     \label{fig:subim22}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_vc_0196_var_b.eps}
+    \includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0196_var_g.eps}
     \label{fig:subim23}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_vc_028_var_b.eps}
+    \includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_028_var_g.eps}
     \label{fig:subim24}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/li_vc_0336_var_b.eps}
+    \includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0336_var_g.eps}
     \label{fig:subim25}
@@
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
+The dashed lines make reference to the ZAMS.}
 \label{fig:image22}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/rot_vel_0g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_0_0g.eps}
     \label{fig:subim41}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/rot_vel_3_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_3_5g.eps}
     \label{fig:subim42}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/rot_vel_4g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_0g.eps}
     \label{fig:subim43}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/rot_vel_4_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_5g.eps}
     \label{fig:subim44}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/rot_vel_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_5_0g.eps}
     \label{fig:subim45}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/rot_vel_5_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_5_5g.eps}
     \label{fig:subim46}
@@
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
+The dashed lines make reference to the ZAMS.}
 \label{fig:image24}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/mb_act_3_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_3_5g.eps}
     \label{fig:subim31}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/mb_act_4_0g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_4_0g.eps}
     \label{fig:subim32}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/mb_act_4_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_4_5g.eps}
     \label{fig:subim33}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/mb_act_5_0g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_5_0g.eps}
     \label{fig:subim34}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/mb_act_5_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_5_5g.eps}
     \label{fig:subim35}
@@
 Each figure shows a set of models in which the magnetic field with intensity has been fixed and $v/v_{crit}$ varies between 0.0084 and 0.0336.
-The dashed lines make reference to the ZAMS defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.}
+The dashed lines make reference to the ZAMS.}
 \label{fig:image32}

### 2019-07-11 project/overleaf/paper.tex
Paper v3.1

@@
 \usepackage{multicol}
+\usepackage[firstpage]{draftwatermark}
+\SetWatermarkText{DRAFT}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{abstract}
-\textit{Goals.} The study of the influence of rotation and magnetic braking (MB) on lithium (Li) depletion during pre-main sequence (PMS) and main sequence (MS) for solar-type stars.
+\textit{Goals.} The study of lithium (Li) surface abundance in the Sun and the influence of rotation and magnetic braking (MB) on its depletion during pre-main sequence (PMS) and main sequence (MS).
 \newline\textit{Methods.} The effects of rotational mixing and of the hydrostatic effects of rotation on Li abundances are studied by simulating several grids of PMS \& MS rotating and non-rotating models.
 The previous effects are combined with the additional impact of the MB one (with magnetic field intensities ranging between 3.5 and 5.5 G) and the data obtained from the two set of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
-\newline\textit{Results.} The surface Li abundance at the end of the PMS and throughout the MS is decreased when rotational effects are included, i.e.
+\newline\textit{Results.} The surface Li abundance for the Sun like models at the end of the PMS and throughout the MS is decreased when rotational effects are included, i.e.
 the Li depletion rate for rotating models is higher than for non-rotating ones.
@@
 Li is destroyed at a temperature $T_{Li} \approx 2.5 x 10^6\; K$, and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value.
-Solar-type stars are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $T_{Li}$ \citep{Iben1965}.
-This depletion stops during the approach to the zero-age MS (ZAMS), when the convection zone retreats and the temperature at BCZ is cooler than $T_{Li}$.
+The Sun in particular and the solar-type stars in general are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $T_{Li}$ \citep{Iben1965}.
+This depletion stops during the approach to the zero-age MS (ZAMS) when the convection zone retreats and the temperature at BCZ is cooler than $T_{Li}$.
 In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances, in addition to showing a very similar Li evolution until their approach to the terminal-age MS (TAMS).
 During this same period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the convective envelope, from its lower limit to the star surface.
-However different abundances of Li have been observed for different star populations (\citet{Somers2014} and references therein).\par
+However, different abundances of Li have been observed for different star populations (\citet{Somers2014} and references therein).\par
 
@@
 The studies of Eddington (1925) and Vogt (1925) had revealed the fact that the southern circulation must necessarily arise in those stars in rotation as a consequence of the von Zeipel paradox: a star cannot be simultaneously in thermal and hydrostatic equilibrium if its rotation speed depends exclusively on its radius.
-The solution to this depends on the existence of meridian circulations.
+The solution to this depends on the existence of southern circulations.
 At the same time, this type of instability implies an impact on the convection mechanisms not considered by the standard models.
 Today it is known that in combination with this type of instability exists a catalogue of others that appear as a consequence of assuming models in rotation \citep{Maeder2003a}, especially the differential rotation as a function of the radius of the star or the deviation of the spherical symmetry.
-Although initially these effects could be candidates to explain the observations, the fact is that the Sun is a slow rotator, meaning slow rotation that one for which the maximum centrifugal force is much smaller than the force of gravity, its deviation of the measured spherical symmetry is of the order of $10^{-5}$ and the time scale for the southern circulation is of the order of $10^{12}$ years, a time period well above the nuclear combustion time scale \citep{Pinsonneault1997}.
+Although initially these effects could be candidates to explain the observations, the fact is that the Sun is a slow rotator, meaning slow rotation that one for which the maximum centrifugal force is much smaller than the force of gravity.
+Furthermore, its deviation of the measured spherical symmetry is of the order of $10^{-5}$ and the time scale for the southern circulation is of the order of $10^{12}$ years, a time period well above the nuclear combustion time scale \citep{Pinsonneault1997}.
 In view of these findings, it seems evident that there are still open questions to be resolved on the subject of the processes that participate, directly or indirectly, in the mixing mechanisms that are not incorporated in the standard stellar models.\par
 
-The impact of rotation on PMS and on Li depletion for solar-type stars has already been discussed (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}).
-These previous studies focused mainly on the hydrostatic effects and didn't consider the influence of a coupling between the star magnetic field and its possible spin-down effect.
-As previously mentioned, it's thought that AML has a direct influence in the mixing processes and transport of the momentum could be produced by a number of mechanisms that could be important: mass loss, magnetic fields and waves (g-modes).
+The impact of rotation on PMS and on Li depletion for solar-type stars has already extensively debated in the past (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}).
+However, these previous studies focused mainly on the hydrostatic effects and didn't consider the influence of a coupling between the star magnetic field and its possible spin-down effect.
+It's thought that AML has a direct influence in the mixing processes and that transport of the momentum could be produced by a number of relevant mechanisms: mass loss, magnetic fields and waves (g-modes).
 Concerning the two latter, waves and magnetic fields have the property that they generally transmit angular momentum (AM) much more effectively than they induce mixing \citep{Denissenkov2007}.
@@
 A proper consideration of the interaction between rotation and magnetic fields is therefore essential.
-We have adapted the one-dimensional stellar evolution code used in this work to model MB and include its effect on the equations for the evolution of the AM distribution and stellar structure.
+We have adapted the one-dimensional stellar evolution code used in this work to model MB and included its effect on the equations for the evolution of the AM distribution and stellar structure.
 This produces a more complete, though still reasonably simple, model for AM evolution.
-In this study we are now focused in investigating not only the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.\par
+It's worth to mention that in this study we don't only focused in investigating the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.\par
+
+\subsection{Effects of rotation}
+Rotation may affect the equations of stellar structure in four different ways according to \citet{Endal1976}:
+%\renewcommand{\theenumi}{\arabic{enumi}.}
+\begin{enumerate}
+    \item The centrifugal force causes the local effective gravity ($g_{eff}$) associated with a point of the star not located on the axis of rotation to be reduced.
+    \item In general, the centrifugal force will not be parallel to the gravitational force and, as a consequence of this, the equipotential layers will cease to be spherical.
+    \item As the von Zeipel theorem states, the radiative flux in a uniformly rotating star is proportional to the $g_{eff}$, therefore the radiative flux is not constant on an equipotential surface.
+    \item Rotation can suppress certain mechanisms that trigger convection, indirectly affecting mixing processes.
+\end{enumerate}
+
+Observational data support the fact that stars are born with different angular velocities even if they come from the same interstellar cloud and that throughout evolution they undergo a deceleration process \citep{Skumanich1972} in a time period between 0.5 and 5 Gyr \citep{Somers2014a}.
+This range of different velocities implies that the mixing mechanisms induced by rotation vary their efficiency and effectiveness and therefore have a direct influence on the abundances of the different chemical elements observed on the surface of stars.
+This fact, together with the differential degree of rotation to which the limits between the radiative and convective zones of the star are subjected, is of particular interest for our study since Li, as mentioned in the introduction, is destroyed at a low temperature.
+Therefore, variations in the intensity of the mixing processes can cause convective zones to extend to interior zones with temperatures higher than $T_{Li}$.
+The deceleration process to which a star is subjected throughout its evolution and the direct link that the rotation has with the mixing processes also offer a natural and consistent explanation for the fact that the destruction of Li is more accentuated or efficient in young stars than in old stars \citep{Sestito2005}.
+In a similar way, stellar rotation rates show a large dispersion at ZAMS \citep{Stauffer1984}, providing in this way a consistent argument for explaining the variance of Li abundances among stars which have a similar mass and were formed within the same cloud.
+Although this approach is not a novelty at all, it is no less true that in those studies based on it the effect caused by MB over the rotation on the PMS and MS is often overlooked.
+\par
+
+Additionally, and according to the prevalent theory of rotational mixing, southern circulation and internal shears leads to a homogenization of the material depleted of Li found within the CZ, causing that the inner layers where there is a temperature higher than $T_{Li}$ and where Li is destroyed, been mixed with those cooler above them.
+These mechanism causes a positive feedback loop in which the cooler material sinks, is heated again and the remaining Li is destroyed, totally or partially, before proceeding to rise once more into cooler layers.
+\par
 
 In the remainder of this paper, we describe a physically sound yet computationally simple model extension to Modules for Experiments in Stellar Astrophysics star evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar wind.
-In its simplest form, implementation of the model requires specification of both the surface magnetic field strength $B$ and the initial  rotation rate $\Omega$ at star surface.
-Our simulations trace the rotational history and A(Li) of a $1 M_{\sun}$ star for a variety of initial values for $B$ and $\Omega$.\par
+In its simplest form, the implementation of the model requires the specification of both the surface magnetic field strength ($B$) and the initial  rotation rate ($\Omega$) at star surface.
+Our simulations trace the rotational history and A(Li) of a $1\, M_{\sun}$ star for a variety of initial values for $B$ and $\Omega$.\par
 
 We define here also some conventions and assumptions adopted in the paper.
-For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, proto} = 0.0142$, not the current photospheric metallicity $Z = Z_{\odot, photosphere} = 0.0134$, as the reference value.
-We also adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10} cm$ and $M_{\odot} = 1.988x10^{33} g$ which are consistent with IAU resolutions \citep{Mamajek2015} .
+For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, proto} = 0.0142$) not the current photospheric metallicity ($Z = Z_{\odot, photosphere} = 0.0134$) as the reference value.
+We also adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10}\, cm$ and $M_{\odot} = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
 For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}.
-This work has been used as a starting point for our work regarding the parameterization of the MESA project calibrated to reproduce the measured values on the solar surface, in particular the section 4.1 of the aforementioned paper.\par
+That work, in particular its section 4.1, has been used as a starting point for ours regarding the parameterization of the MESA project which is calibrated to reproduce the measured values on the solar surface.\par
+
+\subsection{Effects of magnetic fields}
+The origin of magnetic fields is still unknown.
+The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields (see \citet{Dudorov2014} for details about the theory of fossil magnetic field), or fields produced through a dynamo mechanism \citep{Cantiello2009}.
+Besides that, the presence of magnetic fields with intensities of the order of kG \citep{Hussain2014} has been a sine qua non condition for dealing with certain data observed in T Tauri type stars in young, accreting PMS systems \citep{Johns-Krull2007}.
+One of this features is the long rotating periods  which is much longer than predicted from AM conservation.
+A key question is to understand how these young stars can accrete large amounts of material from their circumstellar disk with high specific angular momentum and at the same time they're able to keep slow rotation periods, between 7-10 days \citep{Hussain2014}.
+One plausible explanation for this behaviour considers the AML produced by the interaction between the magnetic field and the particles that form its circumstellar disk \citep{Zanni2012}.
+Once the accretion process stops or becomes less efficient then the star is able to spin up as it contracts towards the ZAMS.\par
+
+The mass of stars plays a determining role in the existence or not of an external convective envelope.
+In turn, the presence of this convective region conditions the evolution of magnetic fields.
+Massive stars do not have external convective zones although they can expose a convective core.
+In these cases, the dominant type of energy transport in the outer zones of the star is radiative.
+On the other hand, the less massive stars, such as the Sun, have convective covers and a radiative core.
+In both types of stars strong magnetic fields can be generated in the tachocline but with notable differences.
+In the case of massive stars, these magnetic fields can survive for a long time but not reach the surface of the star.
+On the other hand in less massive stars these magnetic fields, although they are able to reach the surface of the star, decay quickly in time, in a few hundred years or even only in a few decades \citep{Chabrier2006}, so it is necessary a mechanism for generating magnetic field, e.g.
+a dynamo type, continued in time to explain its existence.
+The way in which this dynamo process is able to generate magnetic fields depends on different parameters and to this day this mechanism is still not fully accepted and is subject of controversy, e.g.
+\citep{Charbonneau2010}.\par
+
+\subsubsection{Internal Magnetic Fields}
+Our Sun is an active star as far as the magnetic field is concerned.
+According to the theoretical models this magnetic field is generated by a dynamo type effect in the tachocline, the intermediate zone between the radiative core and the CZ \citep{Aschwanden2014}, and has an intensity of approximately $B\approx10^5\, G$.
+Additionally, these models predict that because of the differential rotation existing between the core and CZ, the lines of the magnetic field that at the moment of their generation have a poloidal orientation will be affected by the differential rotation and forced to adopt a toroidal typology.
+This change of typology in the magnetic field, from poloidal to toroidal, is known as the $\omega$-effect.
+On the other hand, we have that the opposite case can also occur, that is to say, starting from a toroidal magnetic field to obtain a poloidal configuration.
+This would be possible, on the one hand, thanks to the convective movements of material that are produced in the CZ of the star and, on the other, to assume that this material behaves as a perfect conductor.
+This second condition allows to assume that the magnetic field lines are "frozen" in the conductive fluid and therefore must move in solidarity with it.
+Under these circumstances the magnetic field lines will be pushed to higher areas of the CZ as the material moves vertically and, due to the combined effect of stellar rotation plus Coriolis forces, the magnetic lines will be twisted and give rise to a poloidal magnetic field.
+This is what is known as the $\alpha$-effect.\par
+
+A number of theoretical models have been proposed based on $\alpha$ and $\omega$ effects as an explanation to the generation of magnetic fields, among which is the so-called $\alpha\omega$ or also known as dynamo effect which was proposed by \citet{Spruit2002}.
+By means of this model it is possible to explain the change in polarity of the magnetic field that occurs in the Sun every 11 years, completing a full cycle every 22 years.
+It's also worth to mention that a theoretical debate on this topic is still ongoing \citep{Denissenkov2007} and with regards to MESA, an implementation according to the work of Spruit-Tayler has been implemented \citep{Paxton2013}.\par
+
+\subsubsection{Surface Magnetic Fields} \label{surf_mf}
+The intensity of the magnetic field measured on the Sun's surface is about $B\approx1\, G$, twice the value of the Earth's one.
+Additionally its topology is varied and complex.
+As mentioned in the previous section, the dynamo model establishes that the magnetic field produced inside the star reaches the surface with the help of convective processes taken place in the CZ and gives rise to magnetically active regions on they reach the solar surface.
+Among them, we can find sunspots dominated by magnetic fields with an approximate intensity of $B\approx10^3\, G$ and coronal loops with field strengths of $B\approx10^2\, G$ at the photosphere, and $B\approx 10\, G$ in larger coronal heights \citep{Aschwanden2014}.
+Besides, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism (see, e.g.
+\citet{Brandenburg2004}, \citet{Charbonneau2010} or \citet{Brun2017}).
+Whatever the origins of surface magnetic fields are, it's expected to couple to the wind mass-loss and, if this is strong enough, to produce MB \citet{UdDoula2002}, \citet{Ud-Doula2007}, \citet{Ud-Doula2008} \citet{Meynet2010}.\par
+
+\subsection{Effects of magnetic braking}
+In its most widely accepted approach, the MB is linked to Skumanich's work in which he develops an empirical law of evolution of the MB according to the age of the star.
+To calibrate this law, Skumanich relied on data obtained from type G stars found in the MS \citep{Skumanich1972}.
+The influence of the MB on the evolution of the star is closely related to the transport of the AM and by extension, influenced by how the chemical elements are transported within it.
+According to this approach \citep{Meynet2010}, two main mechanisms are distinguished attending to how the transport of the AM is produced which influence the mixing of these elements:
+
+\begin{enumerate}
+    \item Differential rotation: the AM transport is pushed by southern currents and shear instabilities.
+    \item Solid body rotation: when the transport of AM is very efficient, the the solid body rotation is the prevailing mechanism maintained throughout the entire MS phase and the chemical elements are transported by the southern currents.
+\end{enumerate}
+
+Returning to section \ref{surf_mf} on the estimated mass loss ratio for a solar-type star, approximately $10^{-13}M_{\odot} \; yr^{-1}$, and using this value to estimate $\Dot{J}$, the result obtained can be considered relatively modest as to decisively influence the evolution of the star.
+On the other hand, if we consider that the convective movements that take place in the stellar interior together with the presence of differential rotation can give rise to the generation of magnetic fields that end up reaching the surface of the star \citep{Langer2012} and, the AML can be increased in several orders of magnitude as a consequence of the magnetic field that forces the ionized particles that compose the stellar wind to turn with the same angular speed and whose influence on the stellar wind extends a distance several times the radius of the star (see \citet{UdDoula2002}, \citet{Ud-Doula2007} and \citet{Ud-Doula2008} for more details).
 
 This paper is structured as follows: Section 2 gives an overview of rotation and magnetic fields physical processes and how do they impact on the star evolution.
-We also summarized the most relevant parameters configured in MESA for the model simulations.
+We also summarized here the most relevant parameters configured in MESA for the model simulations.
 In Section 3, the model outputs are presented and discussed in detail.
@@
 
+Keeping in mind this approach, our major objective is aimed to check the degree of validity of the resulting model after including the MB effect as a complementary mechanism of AML that can contribute to explain the evolution of Li in the Sun and other solar-type stars.
+A set of different simulation scenarios with magnetic field strengths ranging from between 3 and 5.5 G have been exercised.
+We'd like to stress again that the magnetic field remains constant during the PMS \& MS phases.
+We compute the evolution of $1M_{\sun}$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.
+More advanced models for more complex arrangements of the magnetic field as well as the evolution of its intensity throughout the life of the star are left for later work.\par
+
+The rest of this section is devoted to describe the most relevant aspects and assumptions made in modelling the evolution of rotation, magnetic braking and angular momentum in MESA.\par
+
 \subsection{Modelling rotation}
@@
 This is what is known as shellular approximation \citep{Meynet1997} and is based on calculating the impact of the centrifugal acceleration caused by the rotation in the stellar evolution equations \citep{Endal1976}.
-MESA uses this simplification to accelerate computational calculations.\par
-
-\subsection{Effects of rotation}
-Rotation may affect the equations of stellar structure in four ways according to \citet{Endal1976}:
-%\renewcommand{\theenumi}{\arabic{enumi}.}
-\begin{enumerate}
-    \item The centrifugal force causes the local effective gravity ($g_{eff}$) associated with a point of the star not located on the axis of rotation to be reduced.
-    \item In general, the centrifugal force will not be parallel to the gravitational force and, as a consequence of this, the equipotential layers will cease to be spherical.
-    \item As the von Zeipel theorem states, the radiative flux in a uniformly rotating star is proportional to the $g_{eff}$, therefore the radiative flux is not constant on an equipotential surface.
-    \item Rotation can suppress certain mechanisms that trigger convection, indirectly affecting mixing processes.
-\end{enumerate}
-
-Observational data support the fact that stars are born with different angular velocities even if they come from the same interstellar cloud and that throughout evolution they undergo a deceleration process \citep{Skumanich1972} in the time period between 0.5 and 5 Gyr \citep{Somers2014a}.
-This range of different velocities implies that the mixing mechanisms induced by rotation vary their efficiency and effectiveness and therefore have a direct influence on the abundances of the different chemical elements observed on the surface of stars.
-This fact, together with the differential degree of rotation to which the limits between the radiative and convective zones of the star are subjected, is of particular interest for our study since Li, as mentioned in the introduction, is destroyed at a low temperature, so variations in the intensity of the mixing processes can cause convective zones to extend to interior zones with temperatures higher than $T_{Li}$.
-The deceleration process to which a star is subjected throughout its evolution and the direct link that the rotation has with the mixing processes also offer a natural and consistent explanation for the fact that the destruction of Li is more accentuated or efficient in young stars than in old stars \citep{Sestito2005}.
-In a similar way, stellar rotation rates show a large dispersion at ZAMS \citep{Stauffer1984}, providing in this way a consistent argument for explaining the variance of Li abundances among stars which have a similar mass and were formed within the same cloud.
-Although this approach is not a novelty at all, it is no less true that in those studies based on it the effect caused by MB over the rotation on the PMS and MS is often overlooked.
-\par
-
-Additionally, and according to the prevalent theory of rotational mixing, southern circulation and internal shears leads to a homogenization of the material depleted of Li found within the CZ, causing that the inner layers where there is a temperature higher than $T_{Li}$ and where Li is destroyed, been mixed with those cooler above them.
-These mechanism causes a positive feedback loop in which the cooler material sinks, is heated again and the remaining Li is destroyed, totally or partially, before proceeding to rise once more into cooler layers.
-\par
-
-In reference to MESA and as stated in \citet{Paxton2013}: "the transport of AM and chemical elements due to rotationally induced instabilities is implemented in a diffusion approximation \citet{Endal1978}.
-MESA calculates diffusion coefficients for five different rotationally induced mixing processes: dynamical shear instability, Solberg-H{\o}iland instability, secular shear instability, Eddington-Sweet circulation, and the Goldreich-Schubert-Fricke instability".
+MESA uses this simplification to accelerate computational calculations.
+Additionally, as stated in \citet{Paxton2013}: "the transport of AM and chemical elements due to rotationally induced instabilities is implemented in a diffusion approximation \citet{Endal1978}.
+MESA calculates diffusion coefficients for five different rotationally induced mixing processes: dynamical shear instability, Solberg-H{\o}iland instability, secular shear instability, Eddington-Sweet circulation, and the Goldreich-Schubert-Fricke instability".\par
 
 \subsection{Modelling magnetic braking}
-The origin of magnetic fields is still unknown.
-The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields (see \citet{Dudorov2014} for details about the Theory of fossil magnetic field), or fields produced through a dynamo mechanism \citep{Cantiello2009}.
-Besides that, the presence of magnetic fields with intensities of the order of kG \citep{Hussain2014} has been a sine qua non condition for dealing with certain data observed in T Tauri type stars in young, accreting PMS systems \citep{Johns-Krull2007}.
-One of this features is the long rotating periods  which is much longer than predicted from AM conservation.
-A key question is to understand how these young stars can accrete large amounts of material from their circumstellar disk with high specific angular momentum and at the same time they're able to keep slow rotation periods, between 7-10 days \citep{Hussain2014}.
-One plausible explanation for this behaviour considers the AML produced by the interaction between the magnetic field and the particles that form its circumstellar disk \citep{Zanni2012}.
-Once the accretion process stops or becomes less efficient then the star is able to spin up as it contracts towards the ZAMS.\par
-
-The mass of stars plays a determining role in the existence or not of an external convective envelope.
-In turn, the presence of this convective region conditions the evolution of magnetic fields.
-Massive stars do not have external convective zones although they can expose a convective core.
-In these cases, the dominant type of energy transport in the outer zones of the star is radiative.
-On the other hand, the less massive stars, such as those of the solar-type, have convective covers and radiative core.
-In both types of stars strong magnetic fields can be generated in the limits of both zones (the so-called tachocline) but with notable differences.
-In the case of massive stars, these magnetic fields can survive for a long time but not reach the surface of the star.
-On the other hand in less massive stars these magnetic fields, although they are able to reach the surface of the star, decay quickly in time, in a few hundred years or even only in a few decades \citep{Chabrier2006}, so it is necessary a mechanism for generating magnetic field, e.g.
-a dynamo type, continued in time to explain its existence.
-The way in which this dynamo process is able to generate magnetic fields depends on different parameters and to this day this mechanism is still not fully accepted and is subject of controversy, e.g.
-\citep{Charbonneau2010}.\par
-
-In the present work we do not take into account the existence of possible magnetic fields during the T-Tauri phase and we adopt a simple theoretical approach to establish their existence during the ZAMS approximation phase.
-The adopted criterion is based on the simultaneous existence of an extensive convective layer and a radiative core.
-From this moment on, the MB routine will be activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participate in the star AML.
-Additionally, we assume that the magnetic field does not vary its intensity throughout the evolution of the star until it reaches TAMS.\par
-
-Our main objective for this paper is to check, in a first approximation, the degree of validity of an approach that includes MB as a complementary mechanism of AML that can contribute to explain the evolution of Li in solar-type stars.
-More advanced models of the arrangement of the magnetic field as well as the evolution of its intensity throughout the life of the star are left for later work.\par
-
-\subsubsection{Internal Magnetic Fields}
-Our Sun is an active star as far as the magnetic field is concerned.
-According to the theoretical models this magnetic field is generated by a dynamo type effect in the tachocline, the intermediate zone between the radiative core and the CZ \citep{Aschwanden2014}, and has an intensity of approximately $B\approx10^5\; G$.
-Additionally, these models predict that because of the differential rotation existing between the core and CZ, the lines of the magnetic field that at the moment of their generation have a poloidal orientation will be affected by the differential rotation and forced to adopt a toroidal typology.
-This change of typology in the magnetic field, from poloidal to toroidal, is known as the $\omega$-effect.
-On the other hand, we have that the opposite case can also occur, that is to say, starting from a toroidal magnetic field to obtain a poloidal configuration.
-This would be possible, on the one hand, thanks to the convective movements of material that are produced in the CZ of the star and, on the other, to assume that this material behaves as a perfect conductor.
-This second condition allows to assume that the magnetic field lines are "frozen" in the conductive fluid and therefore must move in solidarity with it.
-Under these circumstances, the magnetic field lines will be pushed to higher areas of the CZ as the material moves vertically and, due to the combined effect of stellar rotation plus Coriolis forces, the magnetic lines will be twisted and give rise to a poloidal magnetic field.
-This is what is known as the $\alpha$-effect.\par
-
-A number of theoretical models have been proposed based on $\alpha$ and $\omega$ effects as an explanation to the generation of magnetic fields, among which is the so-called $\alpha\omega$ or also known as dynamo effect which was proposed by \citet{Spruit2002}.
-By means of this model it is possible to explain the change in polarity of the magnetic field that occurs in the Sun every 11 years, completing a full cycle every 22 years.
-It's also worth to mention that a theoretical debate on this topic is still ongoing \citep{Denissenkov2007} and with regards to MESA, an implementation according to the work of Spruit-Tayler has been implemented \citep{Paxton2013}.\par
-
-\subsubsection{Surface Magnetic Fields} \label{surf_mf}
-The intensity of the magnetic field measured on the Sun's surface is about $B\approx1\; G$, twice the value of the Earth's one.
-Additionally its topology is varied and complex.
-As mentioned in the previous section, the dynamo model establishes that the magnetic field produced inside the star reaches the surface with the help of convective processes taken place in the CZ and gives rise to magnetically active regions on they reach the solar surface.
-Among them, we can find sunspots dominated by magnetic fields with an approximate intensity of $B\approx10^3\; G$ and coronal loops with field strengths of $B\approx10^2\; G$ at the photosphere, and $B\approx10 G$ in larger coronal heights \citep{Aschwanden2014}.
-Besides, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism (see, e.g.
-\citet{Brandenburg2004}, \citet{Charbonneau2010} or \citet{Brun2017}).
-Whatever the origins of surface magnetic fields are, it's expected to couple to the wind mass-loss and, if this is strong enough, to produce MB \citet{UdDoula2002}, \citet{Ud-Doula2007}, \citet{Ud-Doula2008} \citet{Meynet2010}.\par
-
 In the presence of mass-loss, AM is removed by the material in the stellar wind, resulting in a spin-down effect.
@@
 It has been also observed that strongly magnetic intermediate-mass stars typically have rotation rates much slower than other stars in their parent population \citep{Mathys2006}.
-In those stars, the presence of such magnetic field will interact with the mass loss and the Alfv\'{e}n radius $R_{A}$ plays an important role in this interaction.
+In those stars, the presence of such magnetic field will interact with the mass loss and the Alfv\'{e}n radius ($R_{A}$) plays an important role in this interaction.
 The $R_{A}$ is defined as the point in which the magnetic field energy density $B^{2}/8\upi$ and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$ of the stellar wind are balanced, that is, where the coefficient between both is equal to 1.
@@
 As a consequence, the material will leave the stellar surface with a higher specific AM, as the co-rotation radius has increased and it will roughly correspond to the $R_{A}$.
-This is the effect known as MB.\par
+This effect is known as MB.\par
 
-This energy ratio (eq.
+In the present work we do not take into account either the influence of internal magnetic fields or their existence during the T-Tauri phase.
+We adopt a simple and pragmatic theoretical approach to establish their occurrence when the star is reaching the ZAMS phase.
+The adopted criterion is based on the simultaneous existence of an extensive convective layer and a radiative core.
+From this moment on, the MB routine will be activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participates in the star AML.
+Additionally, we assume that the magnetic field does not vary its intensity throughout the evolution of the star until it reaches the TAMS.\par
+
+The energy ratio (eq.
 \ref{eq:wind_conf}) is denominated wind confinement magnetic parameter ($\eta_*$) and defines a characteristic parameter for the relative effectiveness of the magnetic fields in circumscribing and/or channeling the wind outflow \citep{UdDoula2002}.
-In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval ($\Dot{M}$), and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity ($\nu_{esc}$).\par
+In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval, and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity ($\nu_{esc}$).\par
 
@@
 
-Each of these factors has a decisive influence on the future of the star.
+Each of these factors has a decisive influence on the future of the star and therefore in our routine implementation.
 On the one hand, stars with similar initial masses but different mass loss ratios will end up evolving very differently.
-On the other hand, the ionized particles carried by the solar wind not only contribute to the loss of mass but also to the loss of kinetic energy that is deposited in the interstellar medium.
-Given a star with a spherically symmetric wind, the mass loss rate and the kinetic energy are characterized by the following expressions:
+On the other hand, the ionized particles carried by the solar wind not only contribute to the loss of mass but also to the loss of kinetic ($K_e$) energy that is deposited in the interstellar medium.
+Given a star with a spherically symmetric wind, the mass loss rate ($\Dot{M}$) and $K_e$ are characterized and related by the following expressions:
 
@@
 
-Using (\ref{eq:wind_conf}) and (\ref{eq:mass_loss}) $\eta_*$ can be approximated by: 
+Using (\ref{eq:wind_conf}) and (\ref{eq:mass_loss}), $\eta_*$ can be approximated by: 
 \begin{ceqn}
@@
 
-In general, a magnetically channeled outflow will have a complex stream geometry but for convenience, the expression (\ref{eq:mass_loss} simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate.
-$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law
+In general, a magnetically channeled outflow will have a complex stream geometry but for convenience, the expression (\ref{eq:mass_loss}) simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate.
+$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law:
 \begin{ceqn}
@@
 
-The line-driven winds of massive OB stars have terminal velocities that scale with the photospheric escape velocity \citep{Lamers2000}
+The line-driven winds of massive OB stars have terminal velocities that scale with the photospheric escape velocity \citep{Lamers2000} according to:
 \begin{ceqn}
@@
 \end{ceqn}
-where (\ref{eq:vesc}) is the Newtonian escape velocity from the stellar surface and $G$ is the gravitational constant and (\ref{eq:vinf}) the terminal velocity expressed in terms of escape velocity, radius and mass of the Sun.\par
+where (\ref{eq:vesc}) is the Newtonian escape velocity from the stellar surface, $G$ is the gravitational constant and expression (\ref{eq:vinf}) represents the terminal velocity in terms of escape velocity, radius and mass of the Sun.\par
 
@@
 \end{ceqn}
-which is a more convenient expression to be implemented in MESA because is based on values directly exposed during the simulations.
+which is a more convenient expression to be implemented in MESA because is based on values directly exposed during the simulations and convenient for the approach described in \citet{Ud-Doula2007}.
 
-As reported in table (\ref{tab:phy_mesa}), for the set of simulations performed by MESA and in particular for the mass loss estimation, the empirical formula developed by Reimers for stars in the asymptotic giant branch (AGB) is used.
-For a solar-type star the loss of mass during MS is relatively small, about  $10^{-13}M_{\odot} \; yr^{-1}$.
-With regards to the MB routine implementation, we've based on \citet{Ud-Doula2007}.
-A set of different simulation scenarios with magnetic field strength ranging from between 3 and 10 G have been exercised.
-We'd like to stress again that the magnetic field remains constant during the PMS \& MS phases.
-We compute the evolution of $1M_{\sun}$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.\par
+The AML is directly influenced by the $R_A$ (or alternatively by the $\eta_*$), $\Omega$ and $\Dot{M}$.
+If the value of $\eta_*$ is large, the obtained AML will be also large and the star undergoes a considerable deceleration.
+In our proposal, the value of $\eta_*$ can be directly influenced by establishing in the different simulations the value of the magnetic field ($B$) on the surface or the rotation speed of the star ($\Omega$).
+With regards to $\Dot{M}$ and as reported in table (\ref{tab:phy_mesa}), for the set of simulations performed by MESA the empirical formula developed by Reimers for stars in the asymptotic giant branch (AGB) is used for calculating its value.
+For a solar-type star the $\Dot{M}$ during MS is relatively small, about  $10^{-13}M_{\odot} \, yr^{-1}$.
+Finally, $\Omega$ can be influenced in our models specifying the ratio $\Omega / \Omega_{crit}$ as another free parameter.
 
-\subsection{Evolving the angular velocity}
-As stated in the work of \citet{Paxton2015}, MESA makes its models evolve by default without taking into account the rotation but offers options to initialize them with a user defined angular speed $\Omega$.
+\subsection{Modelling angular velocity and angular momentum evolution} \label{mod_ang_vel_ang_mom}
+As stated in the work of \citet{Paxton2015}, MESA makes its models evolve by default without taking into account the rotation but offers options to initialize them with a user defined angular speed.
 The initial value for $\Omega$ can be indicated as a function of the rotation speed of the star surface expressed in $km\,s^{-1}$ or using the critical rotation speed as a reference (option used in our case).
@@
 The number of cells does not remain fixed throughout the simulation but varies according to certain evolutionary parameters (see \citet{Paxton2015} for more details).
-The $\Omega$ for each cell $k$ ($\Omega_k$) is adjusted so that the resulting angular momentum is retained after adjusting the new mass of the cell ($m_k$), according to the loss or gain suffered by the solar wind, and the distance from the cell to the center of the star ($r_k$).\par
-
-\subsection{Evolving the angular momentum}
-Similar to the situation with $\Omega$, MESA also assigns to each cell an AM value ($J_k$) after resolving the new interior structure of the star.
-The evolution of the AM for each of the cells is performed after knowing the new radius of the star and the distance at which each cell is from the star center.
-Known $m_k$ and $r_k$ proceed to calculate the AM of specific inertia ($I_k$), and from $I_k$ and using $J_k$ of the previous computational step, derives the new value of $\Omega_k$ by means of $J_k/I_k$ and thus manages to conserve the AM.
-It is only after these calculations that he proceeds to calculate the new value of $J_k$ (see \citet{Paxton2015} for more details).
+The $\Omega$ for each cell ($\Omega_k$) is adjusted so that the resulting angular momentum is retained after adjusting the new mass of the cell ($m_k$), according to the loss or gain suffered by the solar wind, and the distance from the cell to the center of the star ($r_k$).
+After resolving the new interior structure of the star, an AM value is assigned to each cell $k$ ($J_k$).
+The evolution of the AM for each of these cells is performed after knowing the new radius of the star and, the mass and distance at which each cell is from the star center.
+Known $m_k$ and $r_k$ proceeds to calculate the AM of specific inertia ($I_k$), and from $I_k$ and using $J_k$ of the previous computational step, derives the new value of $\Omega_k$ by means of $J_k/I_k$ and thus manages to conserve the AM.
+It is only after these calculations that MESA derives the new value of $J_k$ (see \citet{Paxton2015} for more details).
 It's in this point when our MB routine plugs in and allows us to influence the internally calculated value by providing an additional contribution calculated by the user routine.
-Our implementation of the MB routine calculates the additional AML ($\Dot{J}_{k}$), derived from the external torque exerted by the magnetic field and distributes it through the different layers that make up the CZ, according to the expression (\ref{eq:j_dot}).\par
+Our implementation of the MB routine calculates the additional AML ($\Dot{J}_{k}$), derived from the external torque exerted by the magnetic field and distributes it through the different layers that make up the CZ, according to the expression (\ref{eq:k_jdot}):\par
 
@@
 
-\subsection{Effects of magnetic braking}
-In its most widely accepted approach, the MB is linked to Skumanich's work in which he develops an empirical law of evolution of the MB according to the age of the star.
-To calibrate this law, Skumanich relied on data obtained from type G stars found in the MS \citep{Skumanich}.
-The influence of the MB on the evolution of the star is closely related to the transport of the AM and by extension, influenced by how the chemical elements are transported within it.
-According to this approach \citep{Meynet2010}, two main mechanisms are distinguished attending to how the transport of the AM is produced which influence the mixing of these elements:
-
-\begin{enumerate}
-    \item Differential rotation: the AM transport is pushed by southern currents and shear instabilities.
-    \item Solid body rotation: when the transport of AM is very efficient, the the solid body rotation is the prevailing mechanism maintained throughout the entire MS phase and the chemical elements are transported by the southern currents.
-\end{enumerate}
-
-Returning to section \ref{surf_mf} on the estimated mass loss ratio for a solar-type star, approximately $10^{-13}M_{\odot} \; yr^{-1}$, and using this value to estimate $\Dot{J}$, the result obtained can be considered relatively modest as to decisively influence the evolution of the star.
-On the other hand, if we consider that the convective movements that take place in the stellar interior together with the presence of differential rotation can give rise to the generation of magnetic fields that end up reaching the surface of the star \citep{Langer2012}, the AML can be increased in several orders of magnitude as a consequence of the magnetic field that forces the ionized particles that compose the stellar wind to turn with the same angular speed and whose influence on the stellar wind extends a distance several times the radius of the star (see \citet{UdDoula2002}, \citet{Ud-Doula2007} and \citet{Ud-Doula2008} for more details).
-This distance is directly influenced by the $R_A$ (see eq.
-(\ref{eq:j_dot})) or alternatively by the $\eta_*$ (see eq.
-(\ref{eq:j_dot_mesa})).
-As a result of this approach, if the value of $\eta_*$ is large, the obtained AML will be also large and the star undergoes a considerable deceleration.
-In our proposal, the value of $\eta_*$ can be directly influenced by establishing in the different simulations the value of the magnetic field ($B$) on the surface or the rotation speed of the star ($\Omega$).
-
 \section{Results}
-\subsection{Angular velocity and Li evolution}
+\subsection{Angular velocity, Li abundance and effective temperature}
 The destruction of Li occurs when a Li atom collides with a proton producing two He atoms, something that occurs during proton-proton type II reactions (P-P II) and at a temperature lower than that required for H fusion, as low as $2.5\, x\, 10^6\, K$.
@@
 The solid black line represents the reference model according to \citet{Choi2016}.
-The rest of lines are models which include PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
+The rest of lines are models which include PMS rotation with $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336, respectively.
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
@@
 In general, all models burn too much Li before the ZAMS \footnote{Defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.} and do not match with the Pleiades surface Li abundance.
-Later, the reference model (black line) does not deplete Li efficiently on the MS and fail (again) to match the current solar surface Li abundance.
-On the other hand, the other models which include rotation during the PMS with values of $v/v_{crit}$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but it rotational velocity is much higher (see fig.
-(\ref{fig:rot_vel_0g})) than the $2\,kms^{-1}$ of the Sun \citep{Gill2012}.
+Later, the reference model does not deplete Li efficiently on the MS and fail (again) to match the current solar surface Li abundance.
+On the other hand, the other models which include rotation during the PMS with values of $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but it rotational velocity is much higher (see figure (\ref{fig:rot_vel_0g})) than the $2\,kms^{-1}$ of the Sun \citep{Gill2012}.
 \par
 
-For much of the PMS the star rotates as a solid body (see fig.
-(\ref{fig:rot_vel_0g})) and this is because the star has a completely convective interior.
+For much of the PMS the star rotates as a solid body (see figure (\ref{fig:rot_vel_0g})) and this is because the star has a completely convective interior.
 It is not until the end of Hayashi track that the star begins to develop a radiative core due to the increase in temperature that occurs inside the star during the contraction process.
 It is in this stage of evolution of the star when a difference in angular velocity appears between the upper and lower limits of the radiative and convective zones respectively.
-This effects can be clearly recognized in the figure (\ref{fig:rot_vel_cz_0g}) which shows, in addition to star surface $\Omega$, the angular velocity corresponding to the upper and lower limits of the CZ.
 The degree of differential rotation is directly influenced by the initial $\Omega$.
 As the models are initialized with a higher angular velocity, the difference in velocity between the BCZ and the star surface is accentuated in a directly proportional mode; the higher the initial velocity, the bigger the velocity gradient between the bottom and top limits of the CZ.
-As a consequence, the turbulence strength located at the BCZ increases so that Li can reach regions with temperatures about $T_{Li}$ where it is finally burned and destroyed (see fig.
-(\ref{fig:li_var_vel_0g})).
-Also remarkable from the results offered by the simulations is the fact that there are barely any differences between the different models in terms of the abundance of Li during the PMS (see fig.
-(\ref{fig:li_var_vel_0g_z1})).\par
+As a consequence, the turbulence strength located at the BCZ increases so that Li can reach regions with temperatures about $T_{Li}$ where it is finally burned and destroyed (see figure (\ref{fig:li_var_vel_0g})).
+Also remarkable from the results offered by the simulations is the fact that there are barely any differences between the different models in terms of the abundance of Li for much of the PMS.
+It is not until about $10^6$ years later that the differences begin to be most noticeable.
+(see figure (\ref{fig:li_var_vel_0g_z1})).\par
 
@@
 	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
-    \caption {The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models in the approach to the ZAMS (dashed lines).
-The solid black line represents the reference model according to \citet{Choi2016}.
-The rest of lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.}
+    \caption {Similar to figure (\ref{fig:li_var_vel_0g}) but zooming in on the ZAMS.
+The models with a higher initial rotational velocity reach already the ZAMS with a lower amount of Li measure on the star surface.}
     \label{fig:li_var_vel_0g_z1}
-\end{figure}
-
-Another well known structural effect of rotation is the decrease of the effective temperature ($T_{eff}$) and to a less extent of stellar luminosity.
-Both effects can be observed graphically in the HR diagram of figure (\ref{fig:hr_var_vel_0g}).
-These results are in line with those of previous studies (see e.g.
-\citet{Eggenberger2012}, \citet{Piau2001}, \citet{Pinsonneault1989}).
-The figure (\ref{fig:hr_var_vel_0g_z1}) shows a zoomed-in view of evolution tracks from the ZAMS till the TAMS in which the aforementioned effects caused by rotation are even more easily recognizable.
-For example, if we compare the non-rotating model with the rotating ones we can recognize that at the end of the PMS, the latter reach before the ZAMS and exhibit a lower $T_{eff}$ than the non-rotating one (solid black line).\par
-
-\begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
-    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
-The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
-The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The dashed lines make reference to the ZAMS.}
-    \label{fig:rot_vel_0g}
 \end{figure}
@@
 	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
-    \caption{The evolution of angular velocity at surface and limits of the uppermost convective zone, as a function of time for several 1 $M_{\sun}$ models.
-The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
+    \caption{The evolution of angular velocity at surface (solid line) and the top (dashed line) and bottom (dotted line) limits of the uppermost convective zone, as a function of time for several 1 $M_{\sun}$ models.
+The models includes PMS rotation with $\Omega / \Omega_{crit}$ values between 0.0084 and 0.0336.
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The dashed lines make reference to the ZAMS.}
-    \label{fig:rot_vel_cz_0g}
+The dashed vertical lines make reference to the ZAMS.}
+    \label{fig:rot_vel_0g}
 \end{figure}
-%\includegraphics[trim = 35mm 95mm 40mm 100mm, clip, width=0.475\textwidth]
+
+Another well known structural effect of rotation is the decrease of the effective temperature ($T_{eff}$) and to a less extent of stellar luminosity.
+Both effects can be observed graphically in the HR diagram of figure (\ref{fig:hr_var_vel_0g}) which shows a zoomed-in view of evolution tracks from the ZAMS till the TAMS.
+If we compare the non-rotating model (black solid line on the left side) with the rotating ones we can recognize that at the end of the PMS, the latter reach before the ZAMS and exhibit a lower $T_{eff}$ than the former.
+These results are in line with those of previous studies (see e.g.
+\citet{Eggenberger2012}, \citet{Piau2001}, \citet{Pinsonneault1989}).\par
 
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/hr_vc_0336_var_g_z1.eps}
-    \caption{An example solar 1$M_{\sun}$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities. The rotation is activated in the models in the PMS.}
+	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/hr_var_vel_0_0g_z1.eps}
+    \caption{An example solar 1$M_{\sun}$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities.
+The rotation is activated in the models in the PMS and those models reach before the ZAMS and at a lower $T_{eff}$ than the non-rotating one (solid black line)}
     \label{fig:hr_var_vel_0g}
@@
 
-\begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/hr_var_vel_0_0g_z1.eps}
-    \caption{Similar to figure (\ref{fig:hr_var_vel_0g}) but now showing in detail the effects of rotation on the evolutionary tracks.}
-    \label{fig:hr_var_vel_0g_z1}
-\end{figure}
-
-\subsection{Magnetic braking and Li evolution}
+\subsection{Magnetic braking, Li abundance and effective temperature}
 In the rest of this document we are going to focus, unless otherwise indicated, on models that include rotation from the PMS since they are the ones that can offer a greater approximation to the observations.
@@
 
-If we concentrate our attention on figure (\ref{fig:li_var_vel_4_0g}) and compare it with figure (\ref{fig:li_var_vel_0g}) we will see how the profiles of Li abundance are already altered during PMS and MS.
+If we concentrate our attention on figure (\ref{fig:li_var_vel_4_0g}) and compare it with figure (\ref{fig:li_var_vel_0g}) we will see how the profiles of Li abundance are altered during PMS and MS.
 In the first phase we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see equation (\ref{eq:j_dot})) depends directly on the loss of mass.
-If we take into account that for solar-type stars the models predict a modest total loss of mass, this is even much lower during PMS.
-On the contrary, during the MS it is observed in a much more evident way how the AML is much more significant, causing the star to rotate more slowly and as a result a smaller amount of Li to be destroyed.\par
+If we take into account that for solar-type stars the models predict a modest total loss of mass, that value is even much lower during PMS.
+On the contrary, during the MS it is observed that the AML is much more significant, causing the star to rotate more slowly and as a result a smaller amount of Li to be destroyed.\par
 
@@
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
-The models include a magnetic field with an intensity of 4G and PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
+The models include a magnetic field with an intensity of 4G and PMS rotation with $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336, respectively.
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
@@
 Here we represent the rotation profiles on the surface of the stars for different rotating models.
-As it can be observed, and in a similar way to the evolution profiles of the Li on the surface commented in the previous paragraph, the effect of the routine is much more accentuated once the ZAMS is reached.
+Similarly to the evolution profiles of the Li on the surface commented in the previous paragraph, the effect of the routine is much more accentuated once the ZAMS is reached.
 If we compare the evolution of the curves presented here with those of figure (\ref{fig:rot_vel_0g}) we see how the star, instead of continuing to increase its $\Omega$, begins to slow down after having reached its maximum in the passage through the ZAMS.\par
@@
 We also find interesting to highlight the behaviour of $\Omega$ profiles when approaching TAMS.
-As can be seen in figure (\ref{fig:rot_vel_4g}) and in more detail in (\ref{fig:rot_vel_4g_z1}), we find that the evolution of $\Omega$ suffers from a series of ups and downs that do not appear previously.
+It can be recognized in figure (\ref{fig:rot_vel_4g}) and in more detail in (\ref{fig:rot_vel_4g_z1}), we find that the evolution of $\Omega$ suffers from a series of ups and downs that do not appear previously.
 Everything seems to indicate that this behavior is due to a combined effect of the MB routine and the time-step (probably to big) used by MESA during the last part of the simulation.
-We based this affirmation on the fact that this effect is not observed in the series of simulations in which the routine is not included (figs.
-(\ref{fig:rot_vel_0g}) and (\ref{fig:rot_vel_cz_0g})).\par
+We based this affirmation on the fact that this effect is not observed in the series of simulations in which the routine is not included (see figure (\ref{fig:rot_vel_0g})).\par
 
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g.eps}
+	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g.eps}
     \caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
-The models include a magnetic field with an intensity of 4G and PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
+The models include a magnetic field with an intensity of 4G and PMS rotation with $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336, respectively.
 Additionally it's simulated the magnetic braking effect of magnetic field with an intensity of 4G.
@@
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g.eps}
-    \caption{The evolution of angular velocity at surface and limits of the uppermost convective zone, as a function of time for several 1 $M_{\sun}$ models.
-The lines are models with includes PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
-The dashed lines make reference to the ZAMS.}
-    \label{fig:rot_vel_cz_4g}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g_z1.eps}
+	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g_z1.eps}
     \caption{Similar to figure (\ref{fig:rot_vel_4g}) but now showing in detail the surface rotational velocity as the star approaches TAMS.}
@@
 
-It's also remarkable to highlight the lower $\Omega$ observed in figure (\ref{fig:rot_vel_cz_4g}) at the star surface in comparison with the upper limit of the convective layer.
+It's also remarkable to observed, for some of the models depicted figure (\ref{fig:rot_vel_4g}), the lower $\Omega$ at the star surface (solid line) in comparison with the upper limit of the convective layer (dashed line).
 This fact is most probably a consequence of the expansion of the superficial layers after the mass loss produced by stellar winds.\par
 
-As mentioned above, the MB routine distributes the total amount of AML calculated according to equation (\ref{eq:k_jdot}) among the different layers that compose it.
+The MB also leaves its mark on the HR diagram by significantly affecting the $T_{eff}$ of the star.
+To visualize this effect we will take as reference figure (\ref{hr_vc_0336_var_g_z1}) in which all models were started with the same value $\Omega / \Omega_{crit}=0.0336$.
+In particular, let us pay attention to the evolutionary trace in which a magnetic field was not simulated (black solid line) and to the rest of the evolutionary traces in which the magnetic field was simulated.
+We have that the group of simulations with presence of a magnetic field produces hotter stars due to the influence of the magnetic braking.
+The lower speed with which the star rotates due to the MB effect causes the increase of the $T_{eff}$, being this difference of practically $95\,K$ between the simulated models with $0\,G$ and $5.5\,G$ respectively.
+
+\begin{figure}
+	\includegraphics[trim = 30mm 15mm 15mm 15mm, clip,width=\columnwidth]{figures/hr_vc_0336_var_g_z1.eps}
+    \caption{Similar to figure (\ref{fig:hr_var_vel_0g}) but now showing in detail the effects of magnetic braking on the evolutionary tracks for different magnetic field strengths and $\Omega / \Omega_{crit}=0.0336$.
+The presence of a magnetic field produces hotter stars due to the influence of the magnetic braking on the rotational velocity of the star.}
+    \label{hr_vc_0336_var_g_z1}
+\end{figure}
+
+As described in section \ref{mod_ang_vel_ang_mom}, the MB routine distributes the total amount of AML calculated according to equation (\ref{eq:k_jdot}) among the different layers that compose it.
 In figure (\ref{fig:cz_vc_028_var_b}) we can observe the evolution of the most external CZ normalized with respect to the radius of the star.
@@
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g.eps}
+	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g.eps}
     \caption{The evolution of convective zone size, as a function of time for several 1 $M_{\sun}$ models.
-The models include a magnetic field with an intensity of 4G and PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
+The models include a magnetic field with an intensity of 4G and PMS rotation with $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336, respectively.
 Additionally it's simulated the magnetic braking effect of magnetic field with an intensity of 4G.The dashed lines make reference to the ZAMS.}
@@
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g_z1.eps}
+	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g_z1.eps}
     \caption{Similar to figure (\ref{fig:cz_vc_028_var_b}) but now showing in detail the convection zone size from the near ZAMS till the TAMS.
@@
 If we add to the scenario the fact that the star also decreases its radius during the PMS we have as a result that the star increases $\Omega$ obeying the principle of conservation of AM.
-When reaching the ZAMS, the radius of the star remains more or less stable for much of the MS (except for its final stage) but continues to lose mass so it continues to increase its $\Omega$, although in a less aggressive way if we compare it with the PMS.
-However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see fig.
-(\ref{fig:mb_act_var_vel_vc_028})), the MB routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
+When reaching the ZAMS, the radius of the star remains more or less stable for much of the MS (except for its final stage) but continues to lose mass so it also increases its $\Omega$, although in a less aggressive way if we compare it with the PMS.
+However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see figure (\ref{fig:mb_act_var_vel_vc_028})), the MB routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
 The more intense the magnetic field, the greater the braking effect.\par
@@
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g.eps}
+	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g.eps}
     \caption{The evolution of mass loss $\Dot{M}$ , as a function of time for several 1 $M_{\sun}$ models.
-The models include a variable magnetic field intensity between 0G and 5.5G and PMS rotation with $v/v_{crit}=0.028$.
+The models include a variable magnetic field intensity between $0\,G$ and $5.5\,G$ and PMS rotation with $\Omega / \Omega_{crit}=0.028$.
 The dashed lines make reference to the ZAMS.}
@@
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g_z1.eps}
+	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g_z1.eps}
     \caption{The evolution of mass loss $\Dot{M}$ as the star is approaching TAMS, as a function of time for several 1 $M_{\sun}$ models.
-The models include a variable magnetic field intensity between 0G and 5.5G and PMS rotation with $v/v_{crit}=0.028$.
+The models include a variable magnetic field intensity between $0\,G$ and $5.5\,G$ and PMS rotation with $\Omega / \Omega_{crit}=0.028$.
 The stronger the magnetic field, the lower both the angular velocity and $\Dot{M}$.
@@
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\columnwidth]{figures/mb_act_vc_028_var_g.eps}
+	\includegraphics[trim = 30mm 15mm 15mm 15mm, clip,width=\columnwidth]{figures/mb_act_vc_028_var_g.eps}
     \caption{The activation of the magnetic braking routine as a function of the presence of a radiative core.
-The models include a magnetic field with an intensity of 4G and PMS rotation with $v/v_{crit}$ between 0.0084 and 0.0336, respectively.
-Additionally it's simulated the magnetic braking effect of magnetic field with an intensity of 4G.
+The models include a magnetic field with an intensity ranging between $3.5\,G$ and $5.5\,G$ and PMS rotation with $\Omega / \Omega_{crit}=0.0228$.
 The solid lines signal the magnetic braking routine activation (on) and deactivation (off).
@@
 Throughout this work we are confident that we have demonstrated that both rotation and MB induced effects must be taken into account in the theoretical models for a proper understanding and modelling of the evolution stars.
-The MB effect is of major importance for stellar evolution, both for the transport of AM and for the transport of chemical elements.
+The latter is of major importance for stellar evolution, both for the transport of AM and for the transport of chemical elements.
 We are also well aware that we are still far from understanding the exact physical mechanisms that govern these processes, so it is necessary to continue to delve into these areas of study.
@@
 It is likely that in a rotating star calculated with MB from the beginning, the differential rotation is very much reduced and Li abundances observed in young stars could be explained.
-It is equally important to better understand the general role of mass loss in AML and particularly in colder, low-mass stars, since for the time being and among other factors, it is not easy to determine the terminal velocity of the stellar wind for this type of stars.\par
+It is equally important to better understand the general role of mass loss in AML and particularly in cooler, low-mass stars, since for the time being and among other factors, it is not easy to determine the terminal velocity of the stellar wind for this type of stars.\par
 
-It is also important to emphasize that our models fail to match the observed solar Li abundance and $\Omega$.
+We emphasize as well that our models fail to match the observed solar Li abundance and $\Omega$.
 We cannot, therefore, ascertain to have correctly modeled the rotational history of the Sun.
@@
 
-In the near future, with the new version of our code including magnetic field evolution in MESA, we will focus our efforts on understanding how magnetic fields are linked to rotation (e.g.
-\citep{Maeder2005}, \citep{Morin2012}, \citep{Gondoin2018}), how their topologies are and how they evolves in time (e.g.
-\citep{Keppens1995}, \citep{Maeder2003a}, \citep{Vidotto2014}) in order to go beyond the treatment on the MB routine that has been done in this work.\par
+In the near future, with the new version of our code including magnetic field evolution in MESA, we will focus our efforts on understanding how magnetic fields are linked to rotation, how their topologies are and how they evolves in time in order to go beyond the treatment on the MB routine that has been done in this work.\par
 
@@
 \caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
-Each figure shows a set of models in which the magnetic field with intensity has been fixed and $v/v_{crit}$ varies between 0.0084 and 0.0336, respectively.
+Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\Omega / \Omega_{crit}$ varies between 0.0084 and 0.0336, respectively.
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
@@
 \caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
-Each figure shows a set of models in which $v/v_{crit}$ has been fixed and the magnetic field with intensity varies between 3.5 and 5.5 G, respectively.
+Each figure shows a set of models in which $\Omega / \Omega_{crit}$ has been fixed and the magnetic field with intensity varies between $3.5\,G$ and $5.5\,G$, respectively.
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
@@
 \caption{Grid showing the evolution of the evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
-Each figure shows a set of models in which the magnetic field with intensity has been fixed and $v/v_{crit}$ varies between 0.0084 and 0.0336.
+Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\Omega / \Omega_{crit}$ varies between 0.0084 and 0.0336.
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
@@
 \caption{Grid showing the evolution of magnetic braking activation, as a function of time and the existence of a radiative core for several 1 $M_{\sun}$ models.
-Each figure shows a set of models in which the magnetic field with intensity has been fixed and $v/v_{crit}$ varies between 0.0084 and 0.0336.
+Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\Omega / \Omega_{crit}$ varies between 0.0084 and 0.0336.
 The dashed lines make reference to the ZAMS.}
@@
 \section*{Acknowledgements}
-I am pleased to acknowledge Jieun Choi her kindness in responding to the different questions about her work necessary to reproduce the results obtained as far as Li is concerned, as well as sharing the MESA files that she used.
+I am pleased to acknowledge Jieun Choi her kindness in responding to the different questions about her work which were aimed to reproduce the results obtained as far as Li is concerned, and also for allowing us access to the MESA files she used.
 I wish also to thank Matteo Cantiello and Bill Paxton for their very valuable support in the development of the magnetic braking routine in MESA.

### 2019-07-11 project/octave/plotting_routines.m

@@ -698,7 +698,7 @@ end
 
 function save_figure(f, title)  
   print(f,'-deps','-color',[title,'.eps']);
-  close;
+  %close;
   %print(f,'-dpng','-color',[title,'.png']);
 end
 
@@ -1275,14 +1275,14 @@ function main()
   %plot_hr_3_5G_var_vel();
   %plot_hr_5_0G_var_vel();
   %plot_hr_0336vc_var_g();
-  %plot_hr_0336vc_var_g_z1();
+  plot_hr_0336vc_var_g_z1();
   %plot_hr_0G_var_vel();
   %plot_hr_0G_var_vel_z1();
   %plot_hr_3_5G_var_vel_z_1();
   %filename  = "/home/rcaballeron/MESA/workspace/sun-jupiter-system/Docs/runs/run_paper/4g_12kms/1M_photosphere_history.data";
   %calculate_ZAMS(filename);
   
-  paper1();
+  %paper1();
 end
 
 

### 2019-08-21 project/overleaf/paper.tex
Paper v3.2

@@
 \textit{Goals.} The study of lithium (Li) surface abundance in the Sun and the influence of rotation and magnetic braking (MB) on its depletion during pre-main sequence (PMS) and main sequence (MS).
-\newline\textit{Methods.} The effects of rotational mixing and of the hydrostatic effects of rotation on Li abundances are studied by simulating several grids of PMS \& MS rotating and non-rotating models.
-The previous effects are combined with the additional impact of the MB one (with magnetic field intensities ranging between 3.5 and 5.5 G) and the data obtained from the two set of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
+\newline\textit{Methods.} The effects of rotational mixing and of the rotational hydrostatic effects on Li abundances are studied by simulating several grids of PMS \& MS rotating and non-rotating models.
+The previous effects are combined with the additional impact of the MB (with magnetic field intensities ranging between 3.5 and 5.5 G) and the data obtained from the two set of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
 \newline\textit{Results.} The surface Li abundance for the Sun like models at the end of the PMS and throughout the MS is decreased when rotational effects are included, i.e.
@@
 This physical phenomena, generally neglected in star evolutionary models, impacts also the effective temperature ($T_{eff}$) of the star and therefore its location in the HR diagram.
-The impact of MB in Li depletion is sensitive to the field intensity: the higher it is, the lower the Li destruction.
+The impact of MB in Li depletion is sensitive to the magnetic field intensity: the higher it is, the lower the Li destruction.
 A direct link between the magnetic fields and the convective zone (CZ) size is observed: stronger magnetic fields produce shallower CZ's.
@@
 
-\section{Introduction}
-Despite decades of theoretical efforts, a coherent explanation cannot be found from the models for the discrepancies found in the comparisons between the abundance of Li for stars belonging to clusters of different ages and that are also in the same evolutionary state, either in the PMS or in the MS.
+\section{Introduction} \label{sec_1}
+Despite decades of theoretical efforts, a coherent explanation derived from models cannot be found for the Li abundance discrepancies detected in stars belonging to clusters of different ages and which are in the same evolutionary state, either in the PMS or in the MS.
 Additionally, these theoretical models are not able to explain the abundances detected in the late stages of the MS \citep{Tschape2001}.\par
@@
 This depletion stops during the approach to the zero-age MS (ZAMS) when the convection zone retreats and the temperature at BCZ is cooler than $T_{Li}$.
-In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances, in addition to showing a very similar Li evolution until their approach to the terminal-age MS (TAMS).
-During this same period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the convective envelope, from its lower limit to the star surface.
+In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances.
+Furthermore, they should also show a very similar Li evolution until their approach to the terminal-age MS (TAMS).
+During this same period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the convective envelope, from its lower limit till the star surface.
 However, different abundances of Li have been observed for different star populations (\citet{Somers2014} and references therein).\par
@@
 At the same time, this type of instability implies an impact on the convection mechanisms not considered by the standard models.
-Today it is known that in combination with this type of instability exists a catalogue of others that appear as a consequence of assuming models in rotation \citep{Maeder2003a}, especially the differential rotation as a function of the radius of the star or the deviation of the spherical symmetry.
+Today it is known that in combination with this type of instability exists a catalogue of others that appear as a consequence of assuming models in rotation \citep{Maeder2003a}, particularly the differential rotation as a function of the radius of the star or the deviation of the spherical symmetry.
 Although initially these effects could be candidates to explain the observations, the fact is that the Sun is a slow rotator, meaning slow rotation that one for which the maximum centrifugal force is much smaller than the force of gravity.
@@
 
-The impact of rotation on PMS and on Li depletion for solar-type stars has already extensively debated in the past (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}).
+The impact of rotation both on PMS and Li depletion for solar-type stars has already extensively debated in the past (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}) and revised more recently on the basis of the availability of more accurate measures (\citet{Bouvier2016}).
 However, these previous studies focused mainly on the hydrostatic effects and didn't consider the influence of a coupling between the star magnetic field and its possible spin-down effect.
-It's thought that AML has a direct influence in the mixing processes and that transport of the momentum could be produced by a number of relevant mechanisms: mass loss, magnetic fields and waves (g-modes).
-Concerning the two latter, waves and magnetic fields have the property that they generally transmit angular momentum (AM) much more effectively than they induce mixing \citep{Denissenkov2007}.
-As result, the efficiency of these processes is most easily tested by their impact on AM evolution.
-Of particular interest for this work is the impact caused by the magnetic field of the star in the AML, more specifically the effect produced on it by MB as a consequence of the solidary rotation that produces on the particles that constitute the solar wind and the torque derived from this coupling.
+It's thought that AML has a direct influence in the mixing processes and that transport of the momentum could be produced by a number of relevant mechanisms: mass loss, magnetic fields and gravity waves (aka, g-modes).
+Concerning the two latter, gravity waves \citep{Charbonnel2005} and magnetic fields \citep{Eggenberger2009} have the property of transmiting angular momentum (AM) much more effectively than inducing mixing \citep{Denissenkov2007}.
+As a consequence of this increment in the AM transport efficiency, the amount of differential rotation between the radiative and the convective zones of the star is reduced (a solid-body rotation is fostered), as are the induced rotational instabilities.
+Some relevant efforts have already been undertaken aimed to model their link with AM distribution (\citet{Eggenberger2008}, \citet{Fuller2014}).
+As result, the efficiency of these processes is most easily tested by their influence on AM evolution.
+Of particular interest for this work is the impact caused by the magnetic field of the star in the AML, more specifically the effect produced on the solar wind particles that are forced to rotate in sync with it.
+As a consequence of this coupling a torque is induced which slows down the star, i.e.
+the magnetic braking effect.
 These magnetic fields would originate in the limits between the radiative and convective zones of the star, in the so-called tachoclines, due to the rotational difference between both zones \citep{Guerrero2016}.
-A proper consideration of the interaction between rotation and magnetic fields is therefore essential.
-We have adapted the one-dimensional stellar evolution code used in this work to model MB and included its effect on the equations for the evolution of the AM distribution and stellar structure.
-This produces a more complete, though still reasonably simple, model for AM evolution.
+\par
+
+Therefore, it's essential a proper consideration of the interactions between rotation and magnetic fields when it comes to study the AM distribution.
+We have adapted the one-dimensional stellar evolution code used in this work to model MB and included its effect on the equations which govern both AM distribution and stellar structure.
+This approach produces a more complete, though still reasonably simple, model for AM evolution.
 It's worth to mention that in this study we don't only focused in investigating the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.\par
@@
     \item The centrifugal force causes the local effective gravity ($g_{eff}$) associated with a point of the star not located on the axis of rotation to be reduced.
-    \item In general, the centrifugal force will not be parallel to the gravitational force and, as a consequence of this, the equipotential layers will cease to be spherical.
+    \item In general, the centrifugal force will not be parallel to the gravitational force so that the equipotential layers will cease to be spherical.
     \item As the von Zeipel theorem states, the radiative flux in a uniformly rotating star is proportional to the $g_{eff}$, therefore the radiative flux is not constant on an equipotential surface.
@@
 
-Observational data support the fact that stars are born with different angular velocities even if they come from the same interstellar cloud and that throughout evolution they undergo a deceleration process \citep{Skumanich1972} in a time period between 0.5 and 5 Gyr \citep{Somers2014a}.
-This range of different velocities implies that the mixing mechanisms induced by rotation vary their efficiency and effectiveness and therefore have a direct influence on the abundances of the different chemical elements observed on the surface of stars.
-This fact, together with the differential degree of rotation to which the limits between the radiative and convective zones of the star are subjected, is of particular interest for our study since Li, as mentioned in the introduction, is destroyed at a low temperature.
+Observational data support the fact that stars are born with different angular velocities even if they come from the same interstellar cloud.
+They also back up that they undergo a deceleration process throughout their evolution \citep{Skumanich1972} in a time period between 0.5 and 5 Gyr \citep{Somers2014a}.
+This range of different velocities implies that the mixing mechanisms induced by rotation vary their efficiency and effectiveness, and therefore have a direct influence on the abundances of the different chemical elements observed on the surface of stars.
+This fact, together with the differential degree of rotation to which the limits between the radiative and convective zones of the star are subjected, is specially interesting for our study since Li, as mentioned in the introduction, is destroyed at a low temperature.
 Therefore, variations in the intensity of the mixing processes can cause convective zones to extend to interior zones with temperatures higher than $T_{Li}$.
@@
 
-In the remainder of this paper, we describe a physically sound yet computationally simple model extension to Modules for Experiments in Stellar Astrophysics star evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar wind.
+In the remainder of this paper, we describe a physically sound yet computationally simple model extension to Modules for Experiments in Stellar Astrophysics star evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.
 In its simplest form, the implementation of the model requires the specification of both the surface magnetic field strength ($B$) and the initial rotation rate ($\Omega$) at star surface.
-Our simulations trace the rotational history and A(Li) of a $1\, M_{\sun}$ star for a variety of initial values for $B$ and $\Omega$.\par
+The models include already rotation during the PMS as there evidences that advocate for a strong established relationship between Li destruction and rotation on that phase (\citeauthor{Bouvier2016} \citeyear{Bouvier2016}, \citeyear{Bouvier2018}), and the numerical simulations trace the rotational history and A(Li) of a $1\, M_{\sun}$ star for a variety of initial values for $B$ and $\Omega$.\par
 
@@
 For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}.
-That work, in particular its section 4.1, has been used as a starting point for ours regarding the parameterization of the MESA project which is calibrated to reproduce the measured values on the solar surface.\par
+That work, in particular its section 4.1, has been used as a starting point for ours regarding the parameterization of the MESA project which is calibrated to reproduce the measured element abundances on the solar surface.\par
 
@@
 In these cases, the dominant type of energy transport in the outer zones of the star is radiative.
-On the other hand, the less massive stars, such as the Sun, have convective covers and a radiative core.
+On the other hand, the less massive stars such as the Sun, have convective covers and a radiative core.
 In both types of stars strong magnetic fields can be generated in the tachocline but with notable differences.
-In the case of massive stars, these magnetic fields can survive for a long time but not reach the surface of the star.
-On the other hand in less massive stars these magnetic fields, although they are able to reach the surface of the star, decay quickly in time, in a few hundred years or even only in a few decades \citep{Chabrier2006}, so it is necessary a mechanism for generating magnetic field, e.g.
+For the former, these magnetic fields can survive for a long time but not reach the surface of the star.
+For the latter, these magnetic fields, although they are able to reach the surface of the star, decay quickly in time, in a few hundred years or even only in a few decades \citep{Chabrier2006}.
+It's therefore necessary a mechanism for generating magnetic field, e.g.
 a dynamo type, continued in time to explain its existence.
 The way in which this dynamo process is able to generate magnetic fields depends on different parameters and to this day this mechanism is still not fully accepted and is subject of controversy, e.g.
-\citep{Charbonneau2010}.\par
+\citet{Charbonneau2010}.\par
 
@@
 On the other hand, we have that the opposite case can also occur, that is to say, starting from a toroidal magnetic field to obtain a poloidal configuration.
-This would be possible, on the one hand, thanks to the convective movements of material that are produced in the CZ of the star and, on the other, to assume that this material behaves as a perfect conductor.
+This would be possible on the one hand, thanks to the convective movements of material that are produced in the CZ of the star and on the other, to assume that this material behaves as a perfect conductor.
 This second condition allows to assume that the magnetic field lines are "frozen" in the conductive fluid and therefore must move in solidarity with it.
@@
 By means of this model it is possible to explain the change in polarity of the magnetic field that occurs in the Sun every 11 years, completing a full cycle every 22 years.
-It's also worth to mention that a theoretical debate on this topic is still ongoing \citep{Denissenkov2007} and with regards to MESA, an implementation according to the work of Spruit-Tayler has been implemented \citep{Paxton2013}.\par
+It's also worth to mention that a theoretical debate on this topic is still ongoing \citep{Denissenkov2007} and with regards to MESA, an implementation according to the work of Spruit-Tayler has been coded \citep{Paxton2013}.\par
 
@@
 Among them, we can find sunspots dominated by magnetic fields with an approximate intensity of $B\approx10^3\, G$ and coronal loops with field strengths of $B\approx10^2\, G$ at the photosphere, and $B\approx 10\, G$ in larger coronal heights \citep{Aschwanden2014}.
-Besides, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism (see, e.g.
+Besides, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism (e.g.
 \citet{Brandenburg2004}, \citet{Charbonneau2010} or \citet{Brun2017}).
-Whatever the origins of surface magnetic fields are, it's expected to couple to the wind mass-loss and, if this is strong enough, to produce MB \citet{UdDoula2002}, \citet{Ud-Doula2007}, \citet{Ud-Doula2008} \citet{Meynet2010}.\par
+Whatever the origins of surface magnetic fields are, it's expected to couple to the wind mass-loss and, if this is strong enough, to produce MB (e.g.
+\citet{UdDoula2002}, \citet{Ud-Doula2007}, \citet{Ud-Doula2008} \citet{Meynet2010}) .\par
 
@@
 
-Returning to section \ref{surf_mf} on the estimated mass loss ratio for a solar-type star, approximately $10^{-13}M_{\odot} \; yr^{-1}$, and using this value to estimate $\Dot{J}$, the result obtained can be considered relatively modest as to decisively influence the evolution of the star.
-On the other hand, if we consider that the convective movements that take place in the stellar interior together with the presence of differential rotation can give rise to the generation of magnetic fields that end up reaching the surface of the star \citep{Langer2012} and, the AML can be increased in several orders of magnitude as a consequence of the magnetic field that forces the ionized particles that compose the stellar wind to turn with the same angular speed and whose influence on the stellar wind extends a distance several times the radius of the star (see \citet{UdDoula2002}, \citet{Ud-Doula2007} and \citet{Ud-Doula2008} for more details).
+The AML loss caused by MB depends directly on the amount of mass lost by the star due to the stellar winds.
+The estimated mass loss ratio for a solar-type star, approximately $10^{-13}M_{\odot} \; yr^{-1}$, and the resulting AML derived from that value (refer to \ref{mod_mb} for details) can be considered relatively modest as to decisively influence the evolution of the star.
+Furthermore, the combined effect of convective movements and differential rotation lead to the generation of magnetic fields.
+Those magnetic fields end up reaching the surface of the star \citep{Langer2012}, so that the AML can be increased in several orders of magnitude.
+That increment is as a consequence of the magnetic field that forces the stellar wind ionized particles to turn with the same angular speed, winds which extends a distance several times the radius of the star (see \citet{UdDoula2002}, \citet{Ud-Doula2007} and \citet{Ud-Doula2008} for more details).
 
-This paper is structured as follows: Section 2 gives an overview of rotation and magnetic fields physical processes and how do they impact on the star evolution.
+This paper is structured as follows: Section \ref{sec_2} gives an overview of rotation and magnetic fields physical processes and how do they impact on the star evolution.
 We also summarized here the most relevant parameters configured in MESA for the model simulations.
-In Section 3, the model outputs are presented and discussed in detail.
-Finally, Section 4 finishes this paper with a discussion of open points and future works.\par
+In Section \ref{sec_3}, the model outputs are presented and discussed in detail.
+Finally, Section \ref{sec_4} finishes this paper with a discussion about open points and future works.\par
 
-\section{Method}
+\section{Method} \label{sec_2}
 \begin{table}
@@
 
-In this section we review the core physic aspects simulated in MESA and the most relevant parameters adopted in the models simulations.
+In this section we review the core physic aspects simulated in MESA and the most relevant parameters adopted in the numerical simulations.
 In its simplest version, by activating rotation in the model, the evolutionary computational code MESA takes into account the following effects:
@@
 A set of different simulation scenarios with magnetic field strengths ranging from between 3 and 5.5 G have been exercised.
-We'd like to stress again that the magnetic field remains constant during the PMS \& MS phases.
+We'd like to stress that the magnetic field remains constant during the PMS \& MS phases.
 We compute the evolution of $1M_{\sun}$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.
@@
 It is also responsible for inducing through hydrodynamic instabilities the mixing of the chemical elements that make it up.
-On the other hand, in order to be able to correctly model the effects induced in the mixing by rotation and for the results obtained from the theoretical models that implement this physical effect to be consistent with the observational data (see \citet{Pinsonneault1997}, \citet{Maeder2000}), it is necessary to analyze and understand more precisely the mechanisms that govern the distribution of AM, since in this area the theoretical models also offer deficiencies to explain the observations \citep{Denissenkov2007}.
+In order to be able to correctly model those induces mixing effects consistently with the observational data, it is necessary to analyze and understand more precisely the mechanisms that govern the distribution of AM (\citet{Pinsonneault1997}, \citet{Maeder2000}), since in this area the theoretical models also offer deficiencies to explain the observations \citep{Denissenkov2007}.
 \par
@@
 
-\subsection{Modelling magnetic braking}
+\subsection{Modelling magnetic braking} \label{mod_mb}
 In the presence of mass-loss, AM is removed by the material in the stellar wind, resulting in a spin-down effect.
@@
 \ref{eq:wind_conf}) is denominated wind confinement magnetic parameter ($\eta_*$) and defines a characteristic parameter for the relative effectiveness of the magnetic fields in circumscribing and/or channeling the wind outflow \citep{UdDoula2002}.
-In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval, and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity ($\nu_{esc}$).\par
+In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval, and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity ($\nu_\infty$).\par
 
@@
 
-Given that the version of MESA used in this work does include neither the treatment of magnetic fields and therefore, nor for magnetic braking effects, it is necessary to extend the simulator to include this phenomenon.
+Given that the version of MESA used in this work does include neither the treatment of surface magnetic fields (it does account for transport by internal magnetic fields of AM and chemical elements) and therefore, nor for magnetic braking effects, it is necessary to extend the simulator to include this phenomenon.
 In the present work we have followed the theoretical work on MB in the Sun carried out by \citet{Weber1967}, making use of the expression (\ref{eq:j_dot}) for the calculation of the AML.
@@
 MESA assigns an $\Omega$ value to each of the cells that make up the star model.
-The number of cells does not remain fixed throughout the simulation but varies according to certain evolutionary parameters (see \citet{Paxton2015} for more details).
-The $\Omega$ for each cell ($\Omega_k$) is adjusted so that the resulting angular momentum is retained after adjusting the new mass of the cell ($m_k$), according to the loss or gain suffered by the solar wind, and the distance from the cell to the center of the star ($r_k$).
+The number of cells does not remain fixed throughout the simulation but varies according to certain evolutionary parameters.
+The $\Omega$ for each cell $k$ ($\Omega_k$) is adjusted so that the resulting angular momentum is retained after adjusting the new mass of the cell $k$ ($m_k$), according to the loss or gain suffered by the solar wind, and the distance from that particular cell to the center of the star ($r_k$).
 After resolving the new interior structure of the star, an AM value is assigned to each cell $k$ ($J_k$).
@@
 It's in this point when our MB routine plugs in and allows us to influence the internally calculated value by providing an additional contribution calculated by the user routine.
-Our implementation of the MB routine calculates the additional AML ($\Dot{J}_{k}$), derived from the external torque exerted by the magnetic field and distributes it through the different layers that make up the CZ, according to the expression (\ref{eq:k_jdot}):\par
+Our implementation of the MB routine calculates the additional AML ($\Dot{J}_{k}$), derived from the external torque exerted by the magnetic field and distributes it through the different layers that make up the CZ according to the expression (\ref{eq:k_jdot}):\par
 
@@
 
-\section{Results}
+\section{Results} \label{sec_3}
 \subsection{Angular velocity, Li abundance and effective temperature}
-The destruction of Li occurs when a Li atom collides with a proton producing two He atoms, something that occurs during proton-proton type II reactions (P-P II) and at a temperature lower than that required for H fusion, as low as $2.5\, x\, 10^6\, K$.
+The destruction of Li occurs when a Li atom collides with a proton producing two He atoms, something that takes place during proton-proton type II reactions (P-P II) and at a temperature lower than that required for H fusion, as low as $2.5\, x\, 10^6\, K$.
 Those stars without a convective zone can still destroy Li by non-convective mixing processes which can take place in radiative regions and are driven by AML.
 Fast rotating ZAMS stars have suffered little AML and so would have the highest Li abundances.
-Slow rotators may have suffered low AML if they started with a less AM, or a lot if they were magnetically coupled to a circumstellar disk for an extended period of time, and therefore may have a range of Li abundances.
+Slow rotators may have suffered low AML if they started with a less AM, or a lot if they were coupled to a circumstellar disk for an extended period of time \citep{Eggenberger2010}, and therefore may have a range of Li abundances.
 \par
@@
 Today, standard models are not able to replicate in a satisfactory way the observed values of Li abundance on star surfaces with the predictions that these models yield.
-This lack of agreement makes us think that certain physical mechanisms that influence the destruction of Li are either being modelled correctly or simply not being taken into account, e.g.
+This lack of agreement makes us think that certain physical mechanisms that influence the destruction of Li are either being modelled improperly or simply not being taken into account, e.g.
 the magnetic braking.
@@
 The solid black line represents the reference model that adopts the solar-calibrated envelope overshooting parameters as documented in \citet{Choi2016}.
-In general, all models burn too much Li before the ZAMS \footnote{Defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.} and do not match with the Pleiades surface Li abundance.
+In general, all models burn too much Li before the ZAMS \footnote{Defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.} and therefore do not match with the Pleiades surface Li abundance.
 Later, the reference model does not deplete Li efficiently on the MS and fail (again) to match the current solar surface Li abundance.
@@
 As a consequence, the turbulence strength located at the BCZ increases so that Li can reach regions with temperatures about $T_{Li}$ where it is finally burned and destroyed (see figure (\ref{fig:li_var_vel_0g})).
-Also remarkable from the results offered by the simulations is the fact that there are barely any differences between the different models in terms of the abundance of Li for much of the PMS.
-It is not until about $10^6$ years later that the differences begin to be most noticeable.
-(see figure (\ref{fig:li_var_vel_0g_z1})).\par
+
+We would also like to point out that there are other investigations in which the conclusions obtained seem to point to a tendency diametrically opposed to the one exposed here, that is to say, the faster the rotation speed, the greater the abundance of Li on the surface of the star.
+An example of this is \citet{Bouvier2018} work in which the authors analyse the possible connection between the Li and the periods of rotation of FGK type stars present in the Pleiades cluster.
+Particularly interesting is the relationship found of a greater abundance of Li in the faster stars and a smaller one in the slower ones.
+A possible hypothesis proposed to explain these observations would suggest that the effects of rotation in those stars that rotate faster makes the overshooting effect less efficient and therefore less material of the convective envelope reaches the radiative core where there is enough temperature to destroy the Li \citep{Baraffe2017}.
+Another possibility exposes that stars with slow rotation develop a higher level of differential rotation, so that a higher Li destruction is expected \citep{Bouvier2008}.
+Also remarkable from the results offered by the numerical simulations is the fact that there are barely any differences between the different models in terms of the abundance of Li for much of the PMS.
+It is not until about $10^7$ years that the differences begin to be most noticeable (see figure (\ref{fig:li_var_vel_0g_z1})).\par
 
@@
 
-Another well known structural effect of rotation is the decrease of the effective temperature ($T_{eff}$) and to a less extent of stellar luminosity.
+Another well known structural effect of rotation is the decrease of the effective temperature ($T_{eff}$) and to a less extent of stellar luminosity ($L$).
 Both effects can be observed graphically in the HR diagram of figure (\ref{fig:hr_var_vel_0g}) which shows a zoomed-in view of evolution tracks from the ZAMS till the TAMS.
@@
     \caption{An example solar 1$M_{\sun}$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities.
-The rotation is activated in the models in the PMS and those models reach before the ZAMS and at a lower $T_{eff}$ than the non-rotating one (solid black line)}
+The rotation is activated in the models in the PMS and those models reach before the ZAMS and at a lower $T_{eff}$ than the non-rotating one (solid black line).
+The luminosity is expressed in terms of $L_{\sun}$}.
     \label{fig:hr_var_vel_0g}
@@
 
-The effect of the MB routine can be seen even more clearly in figure (\ref{fig:rot_vel_4g}).
+The effect of the MB routine can be seen even more clearly in figures (\ref{fig:rot_vel_4g}) \& (\ref{fig:grid_rot_vel}).
 Here we represent the rotation profiles on the surface of the stars for different rotating models.
 Similarly to the evolution profiles of the Li on the surface commented in the previous paragraph, the effect of the routine is much more accentuated once the ZAMS is reached.
-If we compare the evolution of the curves presented here with those of figure (\ref{fig:rot_vel_0g}) we see how the star, instead of continuing to increase its $\Omega$, begins to slow down after having reached its maximum in the passage through the ZAMS.\par
+If we compare the evolution of the curves presented here with those of figure (\ref{fig:rot_vel_0g}) we see how the star, instead of continuing to increase its $\Omega$, begins to slow down after having reached its maximum in the passage through the ZAMS.
+These results are also consistent with those obtained by \citet{Eggenberger2010} as far as the effect of the magnetic field, in particular its influence on the loss of angular momentum, has on the rotational velocity of the star.
+In our models we have also chosen to include the rotation effects offered by MESA but not those of Tayler-Spruit dynamo on the diffusion of elements.
+In a similar way we also observe that the models with lower angular rotation generally end up exhibiting higher values in the abundance of Li on the surface (see figures (\ref{fig:li_var_vel_4_0g}), (\ref{fig:grid_li_var_vel}) \& (\ref{fig:grid_li_var_g})).
+In none of those cases we do obtain values of Li on the surface higher than those shown by the model without rotation.\par
 
 We also find interesting to highlight the behaviour of $\Omega$ profiles when approaching TAMS.
-It can be recognized in figure (\ref{fig:rot_vel_4g}) and in more detail in (\ref{fig:rot_vel_4g_z1}), we find that the evolution of $\Omega$ suffers from a series of ups and downs that do not appear previously.
-Everything seems to indicate that this behavior is due to a combined effect of the MB routine and the time-step (probably to big) used by MESA during the last part of the simulation.
+It can be recognized in figure (\ref{fig:rot_vel_4g}) and in more detail in (\ref{fig:rot_vel_4g_z1}), that its evolution suffers from a series of ups and downs that do not appear previously.
+Everything seems to indicate that this behavior is due to a combined effect of the MB routine and the time-step resolution (probably to big) used by MESA during the last part of the simulation.
 We based this affirmation on the fact that this effect is not observed in the series of simulations in which the routine is not included (see figure (\ref{fig:rot_vel_0g})).\par
@@
 
-It's also remarkable to observed, for some of the models depicted figure (\ref{fig:rot_vel_4g}), the lower $\Omega$ at the star surface (solid line) in comparison with the upper limit of the convective layer (dashed line).
+It's also remarkable to observed, for some of the models depicted in figure (\ref{fig:rot_vel_4g}), the lower $\Omega$ at the star surface (solid line) in comparison with the upper limit of the convective layer (dashed line).
 This fact is most probably a consequence of the expansion of the superficial layers after the mass loss produced by stellar winds.\par
@@
 To visualize this effect we will take as reference figure (\ref{hr_vc_0336_var_g_z1}) in which all models were started with the same value $\Omega / \Omega_{crit}=0.0336$.
-In particular, let us pay attention to the evolutionary trace in which a magnetic field was not simulated (black solid line) and to the rest of the evolutionary traces in which the magnetic field was simulated.
-We have that the group of simulations with presence of a magnetic field produces hotter stars due to the influence of the magnetic braking.
-The lower speed with which the star rotates due to the MB effect causes the increase of the $T_{eff}$, being this difference of practically $95\,K$ between the simulated models with $0\,G$ and $5.5\,G$ respectively.
+In particular, let us pay attention to the evolutionary trace in which a magnetic field was not simulated (black solid line) and to the rest of the evolutionary traces in which it was.
+We have that the group of simulations with presence of a magnetic field produces hotter stars due to the MB influence.
+The lower speed with which the star rotates due to the MB effect causes the increase of the $T_{eff}$, being this difference of practically $95\,K$ between the simulated models with $0\,G$ and $5.5\,G$ respectively for $log(L/L_{\sun})=0$.
 
@@
 In accordance with the established models of stellar evolution, in a solar-type star the CZ covers practically all of it for a large part of the PMS.
-As it approaches ZAMS it is decreasing, as a consequence of the appearance of a radiative core, maintaining an approximately constant radius until the final stage of the MS, where it increases significantly as a response to the generalized increase of the star's radius.
+As it approaches the ZAMS, the CZ is decreasing as a consequence of the appearance of a radiative core and maintains an approximately constant radius until the final stage of the MS.
+In this point it increases significantly as a response to the generalized expansion of the star's radius.
 Regarding the effect of MB on the size of the CZ, we observe that as the intensity of the magnetic field increases, the size of the CZ decreases.
@@
 This radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base drops below $T_{Li}$.
-This effect is most evident during MS (fig.
-(\ref{fig:cz_vc_028_var_b_z1})).
+This effect is most evident during MS (figure (\ref{fig:cz_vc_028_var_b_z1})).
 The decrease in the size of the CZ is in line with the fact that less Li is destroyed by causing less star material to reach areas with temperatures above $T_{Li}$.\par
 
-Similar to the evolution of the $\Omega$ towards the end of the MS discussed in the previous paragraph, we observe that the size of the CZ also shows a series of ups and downs.
+Similar to the evolution of the $\Omega$ towards the end of the MS, we observe that the size of the CZ also shows a series of ups and downs.
 Again we think that this effect is a direct consequence of the temporal resolution used by MESA in the stellar evolution equations.
@@
 In the first stages of PMS is where the greatest loss of mass is concentrated and this diminishes as it approaches the ZAMS.
-If we add to the scenario the fact that the star also decreases its radius during the PMS we have as a result that the star increases $\Omega$ obeying the principle of conservation of AM.
+If we add to this scenario the fact that the star also decreases its radius during the PMS we have as result that the star increases $\Omega$ obeying the principle of conservation of AM.
 When reaching the ZAMS, the radius of the star remains more or less stable for much of the MS (except for its final stage) but continues to lose mass so it also increases its $\Omega$, although in a less aggressive way if we compare it with the PMS.
-However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see figure (\ref{fig:mb_act_var_vel_vc_028})), the MB routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
+However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see figures (\ref{fig:mb_act_var_vel_vc_028}) \& (\ref{fig:grid_mb_act})), the MB routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
 The more intense the magnetic field, the greater the braking effect.\par
@@
 The models include a variable magnetic field intensity between $0\,G$ and $5.5\,G$ and PMS rotation with $\Omega / \Omega_{crit}=0.028$.
-The stronger the magnetic field, the lower both the angular velocity and $\Dot{M}$.
-The dashed lines make reference to the ZAMS.}
+The stronger the magnetic field, the lower $\Dot{M}$.}
     \label{fig:mdot_vc_028_var_b_z1}
@@
 
-\section{Conclusions and future works}
-Throughout this work we are confident that we have demonstrated that both rotation and MB induced effects must be taken into account in the theoretical models for a proper understanding and modelling of the evolution stars.
-The latter is of major importance for stellar evolution, both for the transport of AM and for the transport of chemical elements.
+\section{Conclusions and future works} \label{sec_4}
+Throughout this work we are confident that we have demonstrated that both rotation and MB induced effects must be taken into account in the theoretical models for a proper understanding and modelling of the evolution of stars.
+The latter is of major importance both for the transport of AM and for the transport of chemical elements.
 We are also well aware that we are still far from understanding the exact physical mechanisms that govern these processes, so it is necessary to continue to delve into these areas of study.
-Of particular interest for us is the study of the evolution of the magnetic fields during the PMS and MS and their impact on AML.
+Particularly the study of the evolution of the magnetic fields during the PMS and MS and their impact on AML.
 Future and improved implementations of our routine will be made to study the results on tracks, surface Li composition, rotation, etc...
-It is likely that in a rotating star calculated with MB from the beginning, the differential rotation is very much reduced and Li abundances observed in young stars could be explained.
-It is equally important to better understand the general role of mass loss in AML and particularly in cooler, low-mass stars, since for the time being and among other factors, it is not easy to determine the terminal velocity of the stellar wind for this type of stars.\par
+It is likely that in a rotating star model simulated with MB from the beginning, the differential rotation is very much reduced and therefore Li abundances observed in young stars could be properly explained.
+It is equally important to understand better the general role of mass loss in AML and mostly in cooler, low-mass stars.
+Nowadays it's challenging to determine the terminal velocity of the stellar wind for this type of stars which, as documented previously, plays a key role on the AML.\par
 
@@
 The dashed lines make reference to the ZAMS.}
-\label{fig:image2}
+\label{fig:grid_li_var_vel}
 \end{figure*}
@@
 The dashed lines make reference to the ZAMS.}
-\label{fig:image22}
+\label{fig:grid_li_var_g}
 \end{figure*}
@@
 The dashed lines make reference to the ZAMS.}
-\label{fig:image24}
+\label{fig:grid_rot_vel}
 \end{figure*}
@@
 The dashed lines make reference to the ZAMS.}
-\label{fig:image32}
+\label{fig:grid_mb_act}
 \end{figure*}
@@
 \section*{Acknowledgements}
-I am pleased to acknowledge Jieun Choi her kindness in responding to the different questions about her work which were aimed to reproduce the results obtained as far as Li is concerned, and also for allowing us access to the MESA files she used.
-I wish also to thank Matteo Cantiello and Bill Paxton for their very valuable support in the development of the magnetic braking routine in MESA.
-Finally, I wish to acknowledge the generous work and help offered by the MESA community.
+We are pleased to acknowledge Jieun Choi her kindness in responding to the different questions about her work which were aimed to reproduce the results obtained as far as Li is concerned, and also for allowing us access to the MESA files she used.
+Also we very much appreciate the expert support provided by Elisa Delgado-Mena during the process review of this document.
+Similarly, we wish also to thank Matteo Cantiello and Bill Paxton for their very valuable support in the development of the magnetic braking routine in MESA.
+Finally, we wish to acknowledge the generous work and help offered by the MESA community.
 

### 2019-08-21 project/octave/plotting_routines.m

@@ -698,7 +698,7 @@ end
 
 function save_figure(f, title)  
   print(f,'-deps','-color',[title,'.eps']);
-  %close;
+  close;
   %print(f,'-dpng','-color',[title,'.png']);
 end
 
@@ -872,6 +872,20 @@ function plot_4_0G_var_vel()
   age_vs_li_plots(gauss_fields(3,:), rotational_vels(2:6,:), true, 0.5, [1.0e5,1.0e10,0,4.5], 'A(Li7) - 4.0G & var. rotational velocity', 'li_var_vel_4_0g');
 end
 
+function plot_4_0G_var_vel_st()
+  global gauss_fields;
+  rotational_vels = ['0196crit'; '0196crit_st'];
+  age_vs_li_plots(gauss_fields(3,:), rotational_vels(1:2,:), true, 0.5, [1.0e5,1.0e10,0,4.5], 'A(Li7) - 4.0G & var. rotational velocity', 'li_var_vel_4_0g');
+end
+
+function plot_4_0G_var_vel_vc5()
+  global gauss_fields;
+  rotational_vels = ['0196crit'; '0196crit_vc5'; '0196crit_vc5_md5'];
+  age_vs_li_plots(gauss_fields(3,:), rotational_vels(1:3,:), true, 0.5, [1.0e5,1.0e10,0,4.5], 'A(Li7) - 4.0G & var. rotational velocity', 'li_var_vel_4_0g');
+end
+
+
+
 function plot_4_5G_var_vel()
   global gauss_fields;
   global rotational_vels; 
@@ -952,11 +966,26 @@ function plot_vel_rot_4G_var_vel()
   age_vs_vel_plots(gauss_fields(3,:), rotational_vels(2:6,:), true, 10, [1.0e5,1.0e10], 'Rotational velocity - 4.0G & var. rotational velocity', 'rot_vel_var_vel_4_0g');
 end
 
+function plot_vel_rot_4G_var_vel_vc5_md5()
+  global gauss_fields;
+  rotational_vels = ['0196crit'; '0196crit_vc5'; '0196crit_vc5_md5'];
+  
+  age_vs_vel_plots(gauss_fields(3,:), rotational_vels(1:3,:), true, 10, [1.0e5,1.0e10], 'Rotational velocity - 4.0G & var. rotational velocity', 'rot_vel_var_vel_4_0g');
+end
+
+
[…]

### 2019-08-21 project/overleaf/paper.tex
Paper v3.2.1
Increment graphic font sizes

@@
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip, width=\columnwidth]{figures/li_var_vel_0_0g.eps}
+	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip, width=\columnwidth]{figures/li_var_vel_0_0g.eps}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
@@
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
+	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
     \caption {Similar to figure (\ref{fig:li_var_vel_0g}) but zooming in on the ZAMS.
@@
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
+	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
     \caption{The evolution of angular velocity at surface (solid line) and the top (dashed line) and bottom (dotted line) limits of the uppermost convective zone, as a function of time for several 1 $M_{\sun}$ models.
@@
 The rotation is activated in the models in the PMS and those models reach before the ZAMS and at a lower $T_{eff}$ than the non-rotating one (solid black line).
-The luminosity is expressed in terms of $L_{\sun}$}.
+The luminosity is expressed in terms of $L_{\sun}$.}
     \label{fig:hr_var_vel_0g}
@@
 \begin{figure}
-	\includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_4_0g.eps}
+	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_4_0g.eps}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_0_0g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_0_0g.eps}
     \label{fig:subim1}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_3_5g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_3_5g.eps}
     \label{fig:subim2}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_0g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_0g.eps}
     \label{fig:subim3}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_5g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_5g.eps}
     \label{fig:subim4}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_5_0g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_5_0g.eps}
     \label{fig:subim5}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_5_5g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_5_5g.eps}
     \label{fig:subim6}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0084_var_g.eps}
+    \includegraphics[trim = 35mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0084_var_g.eps}
     \label{fig:subim21}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_014_var_g.eps}
+    \includegraphics[trim = 35mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_014_var_g.eps}
     \label{fig:subim22}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0196_var_g.eps}
+    \includegraphics[trim = 35mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0196_var_g.eps}
     \label{fig:subim23}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_028_var_g.eps}
+    \includegraphics[trim = 35mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_028_var_g.eps}
     \label{fig:subim24}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 40mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0336_var_g.eps}
+    \includegraphics[trim = 35mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0336_var_g.eps}
     \label{fig:subim25}

### 2019-08-21 project/octave/plotting_routines.m

@@ -51,10 +51,10 @@ global ratio_log_LH_vs_log_L = 0.99;
 global line_width = 3;
 global W = 18;
 global H = 14;
-global tick_font_size = 16;
-global title_font_size = 20;
-global axis_font_size = 18;
-global legend_font_size = 18;
+global tick_font_size = 18;
+global title_font_size = 22;
+global axis_font_size = 20;
+global legend_font_size = 20;
 
 
 

### 2019-10-20 project/overleaf/paper.tex
Paper v3.3.0
Incorporando cambios tras revisión de Antonio y Adrián

@@
 A direct link between the magnetic fields and the convective zone (CZ) size is observed: stronger magnetic fields produce shallower CZ's.
-Another important result of this study is that MB effect must be taken into consideration during PMS must if we aim to reproduce Li abundances in young clusters.
+Another important result of this study is that MB effect must be taken into consideration during PMS if we aim to reproduce Li abundances in young clusters.
 \end{abstract}
@@
 The denominated standard models, models that include convection only as a mixing process and do not consider other transport options like diffusion or angular momentum loss (AML), have been mainly involved in the elaboration of these predictions \citep{Sestito2005}.
-Li is destroyed at a temperature $T_{Li} \approx 2.5 x 10^6\; K$, and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value.
+Li is destroyed at a temperature $T_{Li} \approx 2.5 x 10^6\; K$ when a Li atom collides with a proton producing two He atoms, something that takes place during proton-proton type II reactions (P-P II), and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value.
 The Sun in particular and the solar-type stars in general are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $T_{Li}$ \citep{Iben1965}.
@@
 During this same period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the convective envelope, from its lower limit till the star surface.
-However, different abundances of Li have been observed for different star populations (\citet{Somers2014} and references therein).\par
+However, different abundances of Li have been observed for different star populations \citep[][and references therein]{Somers2014}.\par
 
-From approximately the 1920s of the last century it has been accepted that rotation is capable of inducing the mixing of the elements present inside the stars.
-The studies of Eddington (1925) and Vogt (1925) had revealed the fact that the southern circulation must necessarily arise in those stars in rotation as a consequence of the von Zeipel paradox: a star cannot be simultaneously in thermal and hydrostatic equilibrium if its rotation speed depends exclusively on its radius.
-The solution to this depends on the existence of southern circulations.
-At the same time, this type of instability implies an impact on the convection mechanisms not considered by the standard models.
-Today it is known that in combination with this type of instability exists a catalogue of others that appear as a consequence of assuming models in rotation \citep{Maeder2003a}, particularly the differential rotation as a function of the radius of the star or the deviation of the spherical symmetry.
-Although initially these effects could be candidates to explain the observations, the fact is that the Sun is a slow rotator, meaning slow rotation that one for which the maximum centrifugal force is much smaller than the force of gravity.
-Furthermore, its deviation of the measured spherical symmetry is of the order of $10^{-5}$ and the time scale for the southern circulation is of the order of $10^{12}$ years, a time period well above the nuclear combustion time scale \citep{Pinsonneault1997}.
-In view of these findings, it seems evident that there are still open questions to be resolved on the subject of the processes that participate, directly or indirectly, in the mixing mechanisms that are not incorporated in the standard stellar models.\par
+Theoretical models of stars don't inform about the initial amount of Li that a star has, they only describe how it is exhausted.
+Therefore, in order to make an accurate estimate of the initial abundance of Li, it is a prerequisite to be able to compare observations and models beforehand.
+Our Sun represents a unique exception, since it allows us to know the current abundance of this element in its photosphere, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex$ \citep{Jeffries2004}, where $A(\isotope[7]{Li})$ is defined according to eq.
+(\ref{eq:A_Li})\par
 
-The impact of rotation both on PMS and Li depletion for solar-type stars has already extensively debated in the past (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}) and revised more recently on the basis of the availability of more accurate measures (\citet{Bouvier2016}).
+\begin{ceqn}
+\begin{align}
+    A(\isotope[7]{Li}) &= log(N_{\isotope[7]{Li}} / N_{\isotope[1]{H}}) + 12
+    \label{eq:A_Li}
+\end{align}
+\end{ceqn}
+
+On the other hand, the initial abundance of $A(\isotope[7]{Li}) = 3.34 \, dex$ is obtained from meteorite measurements \citep{Randich2006}.
+According to the theory for newly born stars, we have that the initial abundance of Li can be estimated fairly accurately from photospherical measurements on T-Tauri type stars, or from the hottest F stars forming part of slightly older clusters.
+For the latter, the current theory of stellar evolution suggests that Li must not yet be exhausted.
+The results obtained from the measurements on both types of stars allow us to fix the initial abundance of Li in the interval $3.0 \, dex < A(\isotope[7]{Li}) < 3.4 \, dex$ \citep{Randich2006}.\par
+
+The impact of rotation both on PMS and Li depletion for solar-type stars has already extensively debated in the past \citep{Pinsonneault1997,Jeffries2004,Somers2014} and revised more recently on the basis of the availability of more accurate measures \citep{Bouvier2016}.
 However, these previous studies focused mainly on the hydrostatic effects and didn't consider the influence of a coupling between the star magnetic field and its possible spin-down effect.
-It's thought that AML has a direct influence in the mixing processes and that transport of the momentum could be produced by a number of relevant mechanisms: mass loss, magnetic fields and gravity waves (aka, g-modes).
+It's thought that AML has a direct influence in the mixing processes and that transport of the momentum could be produced by a number of relevant mechanisms: mass loss, magnetic fields and gravity waves (g-modes).
 Concerning the two latter, gravity waves \citep{Charbonnel2005} and magnetic fields \citep{Eggenberger2009} have the property of transmiting angular momentum (AM) much more effectively than inducing mixing \citep{Denissenkov2007}.
 As a consequence of this increment in the AM transport efficiency, the amount of differential rotation between the radiative and the convective zones of the star is reduced (a solid-body rotation is fostered), as are the induced rotational instabilities.
-Some relevant efforts have already been undertaken aimed to model their link with AM distribution (\citet{Eggenberger2008}, \citet{Fuller2014}).
-As result, the efficiency of these processes is most easily tested by their influence on AM evolution.
 Of particular interest for this work is the impact caused by the magnetic field of the star in the AML, more specifically the effect produced on the solar wind particles that are forced to rotate in sync with it.
@@
 
+Today, standard models are not able to replicate in a satisfactory way the observed values of Li abundance on star surfaces with the predictions that these models yield.
+This lack of agreement makes us think that certain physical mechanisms that influence the destruction of Li are either being modelled improperly or simply not being taken into account, e.g.
+the magnetic braking.
+In view of these findings, it seems evident that there are still open questions to be resolved on the subject of the processes that participate, directly or indirectly, in the mixing mechanisms that are not incorporated in the standard stellar models.\par
+
 Therefore, it's essential a proper consideration of the interactions between rotation and magnetic fields when it comes to study the AM distribution.
@@
 
-\subsection{Effects of rotation}
-Rotation may affect the equations of stellar structure in four different ways according to \citet{Endal1976}:
-%\renewcommand{\theenumi}{\arabic{enumi}.}
-\begin{enumerate}
-    \item The centrifugal force causes the local effective gravity ($g_{eff}$) associated with a point of the star not located on the axis of rotation to be reduced.
-    \item In general, the centrifugal force will not be parallel to the gravitational force so that the equipotential layers will cease to be spherical.
-    \item As the von Zeipel theorem states, the radiative flux in a uniformly rotating star is proportional to the $g_{eff}$, therefore the radiative flux is not constant on an equipotential surface.
-    \item Rotation can suppress certain mechanisms that trigger convection, indirectly affecting mixing processes.
-\end{enumerate}
-
-Observational data support the fact that stars are born with different angular velocities even if they come from the same interstellar cloud.
-They also back up that they undergo a deceleration process throughout their evolution \citep{Skumanich1972} in a time period between 0.5 and 5 Gyr \citep{Somers2014a}.
-This range of different velocities implies that the mixing mechanisms induced by rotation vary their efficiency and effectiveness, and therefore have a direct influence on the abundances of the different chemical elements observed on the surface of stars.
-This fact, together with the differential degree of rotation to which the limits between the radiative and convective zones of the star are subjected, is specially interesting for our study since Li, as mentioned in the introduction, is destroyed at a low temperature.
-Therefore, variations in the intensity of the mixing processes can cause convective zones to extend to interior zones with temperatures higher than $T_{Li}$.
-The deceleration process to which a star is subjected throughout its evolution and the direct link that the rotation has with the mixing processes also offer a natural and consistent explanation for the fact that the destruction of Li is more accentuated or efficient in young stars than in old stars \citep{Sestito2005}.
-In a similar way, stellar rotation rates show a large dispersion at ZAMS \citep{Stauffer1984}, providing in this way a consistent argument for explaining the variance of Li abundances among stars which have a similar mass and were formed within the same cloud.
-Although this approach is not a novelty at all, it is no less true that in those studies based on it the effect caused by MB over the rotation on the PMS and MS is often overlooked.
-\par
-
-Additionally, and according to the prevalent theory of rotational mixing, southern circulation and internal shears leads to a homogenization of the material depleted of Li found within the CZ, causing that the inner layers where there is a temperature higher than $T_{Li}$ and where Li is destroyed, been mixed with those cooler above them.
+According to the prevalent theory of rotational mixing, southern circulation and internal shears leads to a homogenization of the material depleted of Li found within the CZ, causing that the inner layers where there is a temperature higher than $T_{Li}$ and where Li is destroyed, been mixed with those cooler above them.
 These mechanism causes a positive feedback loop in which the cooler material sinks, is heated again and the remaining Li is destroyed, totally or partially, before proceeding to rise once more into cooler layers.
@@
 
-We define here also some conventions and assumptions adopted in the paper.
-For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, proto} = 0.0142$) not the current photospheric metallicity ($Z = Z_{\odot, photosphere} = 0.0134$) as the reference value.
-We also adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10}\, cm$ and $M_{\odot} = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
-For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}.
-That work, in particular its section 4.1, has been used as a starting point for ours regarding the parameterization of the MESA project which is calibrated to reproduce the measured element abundances on the solar surface.\par
-
 \subsection{Effects of magnetic fields}
 The origin of magnetic fields is still unknown.
-The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields (see \citet{Dudorov2014} for details about the theory of fossil magnetic field), or fields produced through a dynamo mechanism \citep{Cantiello2009}.
-Besides that, the presence of magnetic fields with intensities of the order of kG \citep{Hussain2014} has been a sine qua non condition for dealing with certain data observed in T Tauri type stars in young, accreting PMS systems \citep{Johns-Krull2007}.
-One of this features is the long rotating periods  which is much longer than predicted from AM conservation.
-A key question is to understand how these young stars can accrete large amounts of material from their circumstellar disk with high specific angular momentum and at the same time they're able to keep slow rotation periods, between 7-10 days \citep{Hussain2014}.
-One plausible explanation for this behaviour considers the AML produced by the interaction between the magnetic field and the particles that form its circumstellar disk \citep{Zanni2012}.
-Once the accretion process stops or becomes less efficient then the star is able to spin up as it contracts towards the ZAMS.\par
+The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields \citep[see][for details about the theory of fossil magnetic field]{Dudorov2014}, or fields produced through a dynamo mechanism \citep{Cantiello2009}.
+Besides that, the presence of magnetic fields with intensities of the order of kG \citep{Hussain2014} has been a sine qua non condition for dealing with certain data observed in T Tauri type stars in young, accreting PMS systems \citep{Johns-Krull2007}.\par
 
@@
 Massive stars do not have external convective zones although they can expose a convective core.
-In these cases, the dominant type of energy transport in the outer zones of the star is radiative.
 On the other hand, the less massive stars such as the Sun, have convective covers and a radiative core.
-In both types of stars strong magnetic fields can be generated in the tachocline but with notable differences.
-For the former, these magnetic fields can survive for a long time but not reach the surface of the star.
-For the latter, these magnetic fields, although they are able to reach the surface of the star, decay quickly in time, in a few hundred years or even only in a few decades \citep{Chabrier2006}.
-It's therefore necessary a mechanism for generating magnetic field, e.g.
-a dynamo type, continued in time to explain its existence.
-The way in which this dynamo process is able to generate magnetic fields depends on different parameters and to this day this mechanism is still not fully accepted and is subject of controversy, e.g.
-\citet{Charbonneau2010}.\par
+In both types of stars strong magnetic fields can be generated in the tachocline but with notable differences (see \citep{Chabrier2006} \citet{Charbonneau2010}\ for details).\par
 
@@
 According to the theoretical models this magnetic field is generated by a dynamo type effect in the tachocline, the intermediate zone between the radiative core and the CZ \citep{Aschwanden2014}, and has an intensity of approximately $B\approx10^5\, G$.
-Additionally, these models predict that because of the differential rotation existing between the core and CZ, the lines of the magnetic field that at the moment of their generation have a poloidal orientation will be affected by the differential rotation and forced to adopt a toroidal typology.
-This change of typology in the magnetic field, from poloidal to toroidal, is known as the $\omega$-effect.
+Additionally, these models predict that because of the differential rotation existing between the core and CZ, the lines of the magnetic field that at the moment of their generation have a poloidal orientation will be affected by the differential rotation and forced to adopt a toroidal typology (the $\omega$-effect).
 On the other hand, we have that the opposite case can also occur, that is to say, starting from a toroidal magnetic field to obtain a poloidal configuration.
 This would be possible on the one hand, thanks to the convective movements of material that are produced in the CZ of the star and on the other, to assume that this material behaves as a perfect conductor.
-This second condition allows to assume that the magnetic field lines are "frozen" in the conductive fluid and therefore must move in solidarity with it.
-Under these circumstances the magnetic field lines will be pushed to higher areas of the CZ as the material moves vertically and, due to the combined effect of stellar rotation plus Coriolis forces, the magnetic lines will be twisted and give rise to a poloidal magnetic field.
-This is what is known as the $\alpha$-effect.\par
-
-A number of theoretical models have been proposed based on $\alpha$ and $\omega$ effects as an explanation to the generation of magnetic fields, among which is the so-called $\alpha\omega$ or also known as dynamo effect which was proposed by \citet{Spruit2002}.
-By means of this model it is possible to explain the change in polarity of the magnetic field that occurs in the Sun every 11 years, completing a full cycle every 22 years.
-It's also worth to mention that a theoretical debate on this topic is still ongoing \citep{Denissenkov2007} and with regards to MESA, an implementation according to the work of Spruit-Tayler has been coded \citep{Paxton2013}.\par
+This second condition allows to assume that the magnetic field lines are "frozen" in the conductive fluid and therefore must move in solidarity with it ($\alpha$-effect).\par
 
@@
 Additionally its topology is varied and complex.
-As mentioned in the previous section, the dynamo model establishes that the magnetic field produced inside the star reaches the surface with the help of convective processes taken place in the CZ and gives rise to magnetically active regions on they reach the solar surface.
-Among them, we can find sunspots dominated by magnetic fields with an approximate intensity of $B\approx10^3\, G$ and coronal loops with field strengths of $B\approx10^2\, G$ at the photosphere, and $B\approx 10\, G$ in larger coronal heights \citep{Aschwanden2014}.
-Besides, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism (e.g.
-\citet{Brandenburg2004}, \citet{Charbonneau2010} or \citet{Brun2017}).
-Whatever the origins of surface magnetic fields are, it's expected to couple to the wind mass-loss and, if this is strong enough, to produce MB (e.g.
-\citet{UdDoula2002}, \citet{Ud-Doula2007}, \citet{Ud-Doula2008} \citet{Meynet2010}) .\par
+The dynamo model establishes that the magnetic field produced inside the star reaches the surface with the help of convective processes taken place in the CZ and gives rise to magnetically active regions on they reach the solar surface.
+Besides, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism \citep[e.g.][]{Brandenburg2004,Charbonneau2010,Brun2017}.
+Whatever the origins of surface magnetic fields are, it's expected to couple to the wind mass-loss and, if this is strong enough, to produce MB \citep[e.g.][]{UdDoula2002,Ud-Doula2007,Ud-Doula2008,Meynet2010}.\par
 
@@
     \item Differential rotation: the AM transport is pushed by southern currents and shear instabilities.
-    \item Solid body rotation: when the transport of AM is very efficient, the the solid body rotation is the prevailing mechanism maintained throughout the entire MS phase and the chemical elements are transported by the southern currents.
+    \item Solid body rotation: when the transport of AM is very efficient, the solid body rotation is the prevailing mechanism maintained throughout the entire MS phase and the chemical elements are transported by the southern currents.
 \end{enumerate}
 
-The AML loss caused by MB depends directly on the amount of mass lost by the star due to the stellar winds.
+The AML caused by MB depends directly on the amount of mass lost by the star due to the stellar winds.
 The estimated mass loss ratio for a solar-type star, approximately $10^{-13}M_{\odot} \; yr^{-1}$, and the resulting AML derived from that value (refer to \ref{mod_mb} for details) can be considered relatively modest as to decisively influence the evolution of the star.
@@
 Those magnetic fields end up reaching the surface of the star \citep{Langer2012}, so that the AML can be increased in several orders of magnitude.
-That increment is as a consequence of the magnetic field that forces the stellar wind ionized particles to turn with the same angular speed, winds which extends a distance several times the radius of the star (see \citet{UdDoula2002}, \citet{Ud-Doula2007} and \citet{Ud-Doula2008} for more details).
+That increment is as a consequence of the magnetic field that forces the stellar wind ionized particles to turn with the same angular speed, winds which extends a distance several times the radius of the star \citep[see][for more details]{UdDoula2002,Ud-Doula2007,Ud-Doula2008}.
 
@@
 	\centering
-	\caption{Summary of adopted physics in MESA (based on \citet{Choi2016})}
+	\caption{Summary of adopted physics in MESA \citep[based on][]{Choi2016}.}
 	\label{tab:phy_mesa}
@@
 
+We define here also some conventions and assumptions adopted in the paper.
+For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, proto} = 0.0142$) not the current photospheric metallicity ($Z = Z_{\odot, photosphere} = 0.0134$) as the reference value.
+We also adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10}\, cm$ and $M_{\odot} = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
+For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}.
+That work, in particular its section 4.1, has been used as a starting point for ours regarding the parameterization of the MESA project which is calibrated to reproduce the measured element abundances on the solar surface.\par
+
 In this section we review the core physic aspects simulated in MESA and the most relevant parameters adopted in the numerical simulations.
@@
 
-Keeping in mind this approach, our major objective is aimed to check the degree of validity of the resulting model after including the MB effect as a complementary mechanism of AML that can contribute to explain the evolution of Li in the Sun and other solar-type stars.
-A set of different simulation scenarios with magnetic field strengths ranging from between 3 and 5.5 G have been exercised.
-We'd like to stress that the magnetic field remains constant during the PMS \& MS phases.
-We compute the evolution of $1M_{\sun}$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.
-More advanced models for more complex arrangements of the magnetic field as well as the evolution of its intensity throughout the life of the star are left for later work.\par
-
 The rest of this section is devoted to describe the most relevant aspects and assumptions made in modelling the evolution of rotation, magnetic braking and angular momentum in MESA.\par
@@
 It is also responsible for inducing through hydrodynamic instabilities the mixing of the chemical elements that make it up.
-In order to be able to correctly model those induces mixing effects consistently with the observational data, it is necessary to analyze and understand more precisely the mechanisms that govern the distribution of AM (\citet{Pinsonneault1997}, \citet{Maeder2000}), since in this area the theoretical models also offer deficiencies to explain the observations \citep{Denissenkov2007}.
+In order to be able to correctly model those induces mixing effects consistently with the observational data, it is necessary to analyze and understand more precisely the mechanisms that govern the distribution of AM \citep{Pinsonneault1997,Maeder2000}, since in this area the theoretical models also offer deficiencies to explain the observations \citep{Denissenkov2007}.
 \par
 
-Angular velocity, or rather the way in which it evolves, also conditions the mechanism by which theoretical models solve the equations of stellar structure.
-It is evident that stars have a three-dimensional structure and for this reason the equations that model their evolution must take this fact into account.
-However, it has been proven sufficient to solve the equations in only one dimension, extrapolating the results to the other two, if it is fulfilled that the angular velocity is constant in regions subjected to the same pressure (isobaric).
-This is what is known as shellular approximation \citep{Meynet1997} and is based on calculating the impact of the centrifugal acceleration caused by the rotation in the stellar evolution equations \citep{Endal1976}.
-MESA uses this simplification to accelerate computational calculations.
-Additionally, as stated in \citet{Paxton2013}: "the transport of AM and chemical elements due to rotationally induced instabilities is implemented in a diffusion approximation \citet{Endal1978}.
-MESA calculates diffusion coefficients for five different rotationally induced mixing processes: dynamical shear instability, Solberg-H{\o}iland instability, secular shear instability, Eddington-Sweet circulation, and the Goldreich-Schubert-Fricke instability".\par
+Stars have a three-dimensional structure and the equations that model their evolution must take this fact into account.
+However, it has been proven sufficient to solve the equations in only one dimension, extrapolating the results to the other two (shellular approximation \citep{Meynet1997}) MESA uses this simplification to accelerate computational calculations \citet{Paxton2013}.\par
 
 \subsection{Modelling magnetic braking} \label{mod_mb}
+In this section, we describe a physically sound yet computationally simple model extension to Modules for Experiments in Stellar Astrophysics star evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.
+In its simplest form, the implementation of the model requires the specification of both the surface magnetic field strength ($B$) and the initial rotation rate ($\Omega$) at star surface.
+The models include already rotation during the PMS as there evidences that advocate for a strong established relationship between Li destruction and rotation on that phase (\citeauthor{Bouvier2016} \citeyear{Bouvier2016}, \citeyear{Bouvier2018}), and the numerical simulations trace the rotational history and A(Li) of a $1\, M_{\sun}$ star for a variety of initial values for $B$ and $\Omega$.\par
+
 In the presence of mass-loss, AM is removed by the material in the stellar wind, resulting in a spin-down effect.
@@
 In those stars, the presence of such magnetic field will interact with the mass loss and the Alfv\'{e}n radius ($R_{A}$) plays an important role in this interaction.
-The $R_{A}$ is defined as the point in which the magnetic field energy density $B^{2}/8\upi$ and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$ of the stellar wind are balanced, that is, where the coefficient between both is equal to 1.
+The $R_{A}$ is defined as the point in which the magnetic field energy density $B^{2}/8\upi$, marked by magnetic field strength ($B$), and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$, defined by the mass density ($\rho$) and velocity of the stellar wind $(\nu)$, are balanced.
+That is, where the coefficient between both is equal to 1.
 In case that $R_{A}$ is greater than the star radius, then the wind flow will have to follow the magnetic field.
@@
 The energy ratio (eq.
-\ref{eq:wind_conf}) is denominated wind confinement magnetic parameter ($\eta_*$) and defines a characteristic parameter for the relative effectiveness of the magnetic fields in circumscribing and/or channeling the wind outflow \citep{UdDoula2002}.
-In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval, and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity ($\nu_\infty$).\par
+\ref{eq:wind_conf}) is denominated wind confinement magnetic parameter ($\eta_*$) and defines a characteristic parameter for the relative effectiveness of the magnetic fields in circumscribing and/or channeling the wind outflow \citep{UdDoula2002}.\par
 
@@
 Each of these factors has a decisive influence on the future of the star and therefore in our routine implementation.
-On the one hand, stars with similar initial masses but different mass loss ratios will end up evolving very differently.
+On the one hand, stars with similar initial masses but different mass loss ($\Dot{M}$) ratios will end up evolving very differently.
 On the other hand, the ionized particles carried by the solar wind not only contribute to the loss of mass but also to the loss of kinetic ($K_e$) energy that is deposited in the interstellar medium.
-Given a star with a spherically symmetric wind, the mass loss rate ($\Dot{M}$) and $K_e$ are characterized and related by the following expressions:
+Given a star with a spherically symmetric wind, $\Dot{M}$ and $K_e$ are characterized and related by the following expressions:
 
@@
 
+In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval, and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity ($\nu_\infty$).
 In general, a magnetically channeled outflow will have a complex stream geometry but for convenience, the expression (\ref{eq:mass_loss}) simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate.
-$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law:
+$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law, where $r$ represents the distance from the center of the star to the point where the velocity is to be measured and $R_*$ is the radius of the star:
 \begin{ceqn}
@@
 where (\ref{eq:vesc}) is the Newtonian escape velocity from the stellar surface, $G$ is the gravitational constant and expression (\ref{eq:vinf}) represents the terminal velocity in terms of escape velocity, radius and mass of the Sun.\par
+
+\begin{figure}
+	\includegraphics[width=\columnwidth]{figures/terminal_vel_vs_escape_vel.png}
+    \caption {Terminal vel vs escape vel.}
+    \label{fig:terminal_vel_vs_escape_vel}
+\end{figure}
 
@@
 
+\section{Results} \label{sec_3}
+
+\subsection{Mesh Models}
+We define here also some conventions and assumptions adopted in the paper.
+For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, proto} = 0.0142$) not the current photospheric metallicity ($Z = Z_{\odot, photosphere} = 0.0134$) as the reference value.
+We also adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10}\, cm$ and $M_{\odot} = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
+For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}.
+That work, in particular its section 4.1, has been used as a starting point for ours regarding the parameterization of the MESA project which is calibrated to reproduce the measured element abundances on the solar surface.\par
+
+Keeping in mind this approach, our major objective is aimed to check the degree of validity of the resulting model after including the MB effect as a complementary mechanism of AML that can contribute to explain the evolution of Li in the Sun and other solar-type stars.
+A set of different simulation scenarios with magnetic field strengths ranging from between 3 and 5.5 G have been exercised.
+We'd like to stress that the magnetic field remains constant during the PMS \& MS phases.
+We compute the evolution of $1M_{\sun}$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.
+More advanced models for more complex arrangements of the magnetic field as well as the evolution of its intensity throughout the life of the star are left for later work.\par
+
 The AML is directly influenced by the $R_A$ (or alternatively by the $\eta_*$), $\Omega$ and $\Dot{M}$.
@@
 In our proposal, the value of $\eta_*$ can be directly influenced by establishing in the different simulations the value of the magnetic field ($B$) on the surface or the rotation speed of the star ($\Omega$).
-With regards to $\Dot{M}$ and as reported in table (\ref{tab:phy_mesa}), for the set of simulations performed by MESA the empirical formula developed by Reimers for stars in the asymptotic giant branch (AGB) is used for calculating its value.
-For a solar-type star the $\Dot{M}$ during MS is relatively small, about  $10^{-13}M_{\odot} \, yr^{-1}$.
-Finally, $\Omega$ can be influenced in our models specifying the ratio $\Omega / \Omega_{crit}$ as another free parameter.
+With regards to $\Dot{M}$ and as reported in table (\ref{tab:phy_mesa}), for the set of simulations performed by MESA the empirical formula developed by Reimers \citet{Reimers1975} for stars in the asymptotic giant branch (AGB) is used for calculating its value.
+For a solar-type star the $\Dot{M}$ during MS is relatively small, about  $10^{-14}M_{\odot} \, yr^{-1}$ \citet{Noerdlinger2008}.
+Finally, $\Omega$ can be influenced in our models specifying the ratio $\Omega / \Omega_{crit}$ as another free parameter, where  $\Omega_{crit}$ represents the star surface velocity at which the centrifugal force balances  the gravitational force.
+Beyond this point, the star would begin to expels matter along the equator.\par
 
-\subsection{Modelling angular velocity and angular momentum evolution} \label{mod_ang_vel_ang_mom}
-As stated in the work of \citet{Paxton2015}, MESA makes its models evolve by default without taking into account the rotation but offers options to initialize them with a user defined angular speed.
-The initial value for $\Omega$ can be indicated as a function of the rotation speed of the star surface expressed in $km\,s^{-1}$ or using the critical rotation speed as a reference (option used in our case).
-Once the model has been initialized with a specific $\Omega$, it is recalculated as the star evolves as a function of its mass, radius and AML that the simulator determines in each time step.
-\par
-
-MESA assigns an $\Omega$ value to each of the cells that make up the star model.
-The number of cells does not remain fixed throughout the simulation but varies according to certain evolutionary parameters.
-The $\Omega$ for each cell $k$ ($\Omega_k$) is adjusted so that the resulting angular momentum is retained after adjusting the new mass of the cell $k$ ($m_k$), according to the loss or gain suffered by the solar wind, and the distance from that particular cell to the center of the star ($r_k$).
-After resolving the new interior structure of the star, an AM value is assigned to each cell $k$ ($J_k$).
-The evolution of the AM for each of these cells is performed after knowing the new radius of the star and, the mass and distance at which each cell is from the star center.
-Known $m_k$ and $r_k$ proceeds to calculate the AM of specific inertia ($I_k$), and from $I_k$ and using $J_k$ of the previous computational step, derives the new value of $\Omega_k$ by means of $J_k/I_k$ and thus manages to conserve the AM.
-It is only after these calculations that MESA derives the new value of $J_k$ (see \citet{Paxton2015} for more details).
-It's in this point when our MB routine plugs in and allows us to influence the internally calculated value by providing an additional contribution calculated by the user routine.
-Our implementation of the MB routine calculates the additional AML ($\Dot{J}_{k}$), derived from the external torque exerted by the magnetic field and distributes it through the different layers that make up the CZ according to the expression (\ref{eq:k_jdot}):\par
+MESA assigns an $\Omega$ value for each cell $k$ ($\Omega_k$) which is adjusted so that the resulting angular momentum is retained after adjusting the new mass of the cell $k$ ($m_k$) and its distance to the center of the star ($r_k$).
+After that, an AM value is assigned to each cell $k$ ($J_k$).
+It's in this point when our MB routine plugs in and influences the internally calculated value by providing an additional contribution ($\Dot{J}_{k}$) which is the result of the external torque exerted by the magnetic field once it has been distributed among the different layers that make up the CZ (\ref{eq:k_jdot}):\par
 
@@
 
-\section{Results} \label{sec_3}
 \subsection{Angular velocity, Li abundance and effective temperature}
-The destruction of Li occurs when a Li atom collides with a proton producing two He atoms, something that takes place during proton-proton type II reactions (P-P II) and at a temperature lower than that required for H fusion, as low as $2.5\, x\, 10^6\, K$.
-Those stars without a convective zone can still destroy Li by non-convective mixing processes which can take place in radiative regions and are driven by AML.
-Fast rotating ZAMS stars have suffered little AML and so would have the highest Li abundances.
-Slow rotators may have suffered low AML if they started with a less AM, or a lot if they were coupled to a circumstellar disk for an extended period of time \citep{Eggenberger2010}, and therefore may have a range of Li abundances.
-\par
-
-Theoretical models of stars don't inform about the initial amount of Li that a star has, they only describe how it is exhausted.
-Therefore, in order to make an accurate estimate of the initial abundance of Li, it is a prerequisite to be able to compare observations and models beforehand.
-Our Sun represents a unique exception, since it allows us to know the current abundance of this element in its photosphere, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex$ \citep{Jeffries2004}, where $A(\isotope[7]{Li})$ is defined according to eq.
-(\ref{eq:A_Li})\par
-
-\begin{ceqn}
-\begin{align}
-    A(\isotope[7]{Li}) &= log(N_{\isotope[7]{Li}} / N_{\isotope[1]{H}}) + 12
-    \label{eq:A_Li}
-\end{align}
-\end{ceqn}
-
-On the other hand, the initial abundance of $A(\isotope[7]{Li}) = 3.34 \, dex$ is obtained from meteorite measurements \citep{Randich2006}.
-According to the theory for newly born stars, we have that the initial abundance of Li can be estimated fairly accurately from photospherical measurements on T-Tauri type stars, or from the hottest F stars forming part of slightly older clusters.
-For the latter, the current theory of stellar evolution suggests that Li must not yet be exhausted.
-The results obtained from the measurements on both types of stars allow us to fix the initial abundance of Li in the interval $3.0 \, dex < A(\isotope[7]{Li}) < 3.4 \, dex$ \citep{Randich2006}.\par
 
@@
 
-Today, standard models are not able to replicate in a satisfactory way the observed values of Li abundance on star surfaces with the predictions that these models yield.
-This lack of agreement makes us think that certain physical mechanisms that influence the destruction of Li are either being modelled improperly or simply not being taken into account, e.g.
-the magnetic braking.
 Figure (\ref{fig:li_var_vel_0g}) shows the evolution of surface Li abundance relative to H as a function of time and for several $1M_{\odot}$ models which take into account the effects of rotation and AML caused by stellar winds but not MB.
@@
 The solid black line represents the reference model that adopts the solar-calibrated envelope overshooting parameters as documented in \citet{Choi2016}.
-In general, all models burn too much Li before the ZAMS \footnote{Defined as the temporal simulation instant closest to the timestep in which both the central H mass fraction has been reduced by 0.0015 from its initial value and the model first $L_H/L_{phot} \geq 0.99$, where $L_{H}$ is luminosity produced by the H burning power at the star core and $L_{phot}$ represents the star luminosity in the photosphere.} and therefore do not match with the Pleiades surface Li abundance.
+In general, all models burn too much Li before the ZAMS and therefore do not match with the Pleiades average surface Li abundance.
+Also remarkable is the fact that there are barely any differences between the different models in terms of the abundance of Li for much of the PMS.
+Only after one million years is Li destroyed to an accentuated degree, since the necessary temperature is not reached at BCZ before.
+Afterwards the different models destroy the Li in a very similar way due to two main reasons.
+On the one hand, the convective zones that the models develop have very similar sizes, so that the temperature at the BCZ is practically the same.
+On the other hand, the differential rotation between the core and the convective zone does not begin to develop until it reaches $10^6$ years, achieving its maximum difference in the ZAMS about  $10^7$ years.
+It is at this moment that the combined effects of turbulence, rotational difference and magnetic braking should increase, leading to notable differences in the evolution of A(Li) (see figure (\ref{fig:li_var_vel_0g_z1})).
 Later, the reference model does not deplete Li efficiently on the MS and fail (again) to match the current solar surface Li abundance.
-On the other hand, the other models which include rotation during the PMS with values of $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but it rotational velocity is much higher (see figure (\ref{fig:rot_vel_0g})) than the $2\,kms^{-1}$ of the Sun \citep{Gill2012}.
+On the other hand, the other models which include rotation during the PMS with values of $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but its rotational velocity is much higher (see figure \ref{fig:rot_vel_0g}) than the $2\,kms^{-1}$ of the Sun \citep{Gill2012}.
 \par
@@
 As a consequence, the turbulence strength located at the BCZ increases so that Li can reach regions with temperatures about $T_{Li}$ where it is finally burned and destroyed (see figure (\ref{fig:li_var_vel_0g})).
-
-We would also like to point out that there are other investigations in which the conclusions obtained seem to point to a tendency diametrically opposed to the one exposed here, that is to say, the faster the rotation speed, the greater the abundance of Li on the surface of the star.
-An example of this is \citet{Bouvier2018} work in which the authors analyse the possible connection between the Li and the periods of rotation of FGK type stars present in the Pleiades cluster.
-Particularly interesting is the relationship found of a greater abundance of Li in the faster stars and a smaller one in the slower ones.
-A possible hypothesis proposed to explain these observations would suggest that the effects of rotation in those stars that rotate faster makes the overshooting effect less efficient and therefore less material of the convective envelope reaches the radiative core where there is enough temperature to destroy the Li \citep{Baraffe2017}.
-Another possibility exposes that stars with slow rotation develop a higher level of differential rotation, so that a higher Li destruction is expected \citep{Bouvier2008}.
-Also remarkable from the results offered by the numerical simulations is the fact that there are barely any differences between the different models in terms of the abundance of Li for much of the PMS.
-It is not until about $10^7$ years that the differences begin to be most noticeable (see figure (\ref{fig:li_var_vel_0g_z1})).\par
+Other investigations \citep{Bouvier2018, Baraffe2017}  point to a tendency diametrically opposed to the one exposed here, the faster the rotation speed, the greater the abundance of Li on the surface of the star.
+\par
 
@@
 It can be recognized in figure (\ref{fig:rot_vel_4g}) and in more detail in (\ref{fig:rot_vel_4g_z1}), that its evolution suffers from a series of ups and downs that do not appear previously.
-Everything seems to indicate that this behavior is due to a combined effect of the MB routine and the time-step resolution (probably to big) used by MESA during the last part of the simulation.
+Everything seems to indicate that this behavior is due to a combined effect of the MB routine and the time-step resolution (probably too big) used by MESA during the last part of the simulation.
 We based this affirmation on the fact that this effect is not observed in the series of simulations in which the routine is not included (see figure (\ref{fig:rot_vel_0g})).\par
@@
 
-Before ending this section we would also like to comment on the other factor that directly affects AML; the loss of mass, $\Dot{M}$.
+Before ending this section we would also like to comment on the other factor that directly affects AML; the mass loss, $\Dot{M}$.
 In figure (\ref{fig:mdot_vc_028_var_b}) we can see the evolution of $\Dot{M}$ during PMS and MS.

### 2019-11-17 project/overleaf/paper.tex
Paper v3.3.1
Después de reunión 15.11.19

@@
 
-% MNRAS is set in Times font. If you don't have this installed (most LaTeX
+% MNRAS is set in Times font. If you do not have this installed (most LaTeX
 % installations will be fine) or prefer the old Computer Modern fonts, comment
@@
 
+\input definitions.tex
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % Keep the title short and informative.
-\title[Rotation, magnetic braking \& Li abundances]{Effects of rotation and magnetic braking on Li abundances for solar-type stars}
+\title[Rotation, magnetic braking \& Li abundances]{The Sun lithium problem and how magnetic braking could help to solve it}
 
@@
 \begin{abstract}
-\textit{Goals.} The study of lithium (Li) surface abundance in the Sun and the influence of rotation and magnetic braking (MB) on its depletion during pre-main sequence (PMS) and main sequence (MS).
+\textit{Goals.} The study of lithium (Li) surface abundance in the Sun and young stellar globular clusters which are seemingly anomalous in present-day scenarios, as well as the influence of rotation and magnetic braking (MB) on its depletion during pre-main sequence (PMS) and main sequence (MS).
 \newline\textit{Methods.} The effects of rotational mixing and of the rotational hydrostatic effects on Li abundances are studied by simulating several grids of PMS \& MS rotating and non-rotating models.
-The previous effects are combined with the additional impact of the MB (with magnetic field intensities ranging between 3.5 and 5.5 G) and the data obtained from the two set of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
+The previous effects are combined with the additional impact of the MB (with magnetic field intensities ranging between 3.0 and 5.0 G) and the data obtained from the two sets of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
 \newline\textit{Results.} The surface Li abundance for the Sun like models at the end of the PMS and throughout the MS is decreased when rotational effects are included, i.e.
@@
 However, this effect is attenuated when the braking effect produced by a magnetic field is present.
-This physical phenomena, generally neglected in star evolutionary models, impacts also the effective temperature ($T_{eff}$) of the star and therefore its location in the HR diagram.
+This physical phenomena, generally neglected in stellar evolutionary models, impacts also the effective temperature ($\teff$) of the star and therefore its location in the HR diagram.
 The impact of MB in Li depletion is sensitive to the magnetic field intensity: the higher it is, the lower the Li destruction.
@@
 The denominated standard models, models that include convection only as a mixing process and do not consider other transport options like diffusion or angular momentum loss (AML), have been mainly involved in the elaboration of these predictions \citep{Sestito2005}.
-Li is destroyed at a temperature $T_{Li} \approx 2.5 x 10^6\; K$ when a Li atom collides with a proton producing two He atoms, something that takes place during proton-proton type II reactions (P-P II), and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value.
-The Sun in particular and the solar-type stars in general are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $T_{Li}$ \citep{Iben1965}.
-This depletion stops during the approach to the zero-age MS (ZAMS) when the convection zone retreats and the temperature at BCZ is cooler than $T_{Li}$.
-In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances.
+Li is destroyed at a temperature $\tli \approx 2.5 x 10^6\; K$ when a Li atom collides with a proton producing two He atoms, something that takes place during proton-proton type II reactions (P-P II), and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value.
+The Sun in particular and the solar-type stars in general are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $\tli$ \citep{Iben1965}.
+This depletion stops during the approach to the zero-age MS (ZAMS) when the convection zone retreats and the temperature at BCZ is cooler than $\tli$.
+In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $\tli$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances.
 Furthermore, they should also show a very similar Li evolution until their approach to the terminal-age MS (TAMS).
@@
 
-Theoretical models of stars don't inform about the initial amount of Li that a star has, they only describe how it is exhausted.
+Theoretical models of stars do not inform about the initial amount of Li that a star has, they only describe how it is exhausted.
 Therefore, in order to make an accurate estimate of the initial abundance of Li, it is a prerequisite to be able to compare observations and models beforehand.
-Our Sun represents a unique exception, since it allows us to know the current abundance of this element in its photosphere, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex$ \citep{Jeffries2004}, where $A(\isotope[7]{Li})$ is defined according to eq.
-(\ref{eq:A_Li})\par
+Our Sun represents a unique exception, since it allows us to know the current abundance of this element in its photosphere, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, ex$ \citep{Jeffries2004}, where $A(\isotope[7]{Li})$ is defined according to Eq.~\ref{eq:A_Li}\par
 
@@
 On the other hand, the initial abundance of $A(\isotope[7]{Li}) = 3.34 \, dex$ is obtained from meteorite measurements \citep{Randich2006}.
-According to the theory for newly born stars, we have that the initial abundance of Li can be estimated fairly accurately from photospherical measurements on T-Tauri type stars, or from the hottest F stars forming part of slightly older clusters.
+According to the theory for newly born stars, we have that the initial abundance of Li can be estimated fairly accurately from photospheric measurements on T-Tauri type stars, or from the hottest F stars forming part of slightly older clusters.
 For the latter, the current theory of stellar evolution suggests that Li must not yet be exhausted.
@@
 The impact of rotation both on PMS and Li depletion for solar-type stars has already extensively debated in the past \citep{Pinsonneault1997,Jeffries2004,Somers2014} and revised more recently on the basis of the availability of more accurate measures \citep{Bouvier2016}.
-However, these previous studies focused mainly on the hydrostatic effects and didn't consider the influence of a coupling between the star magnetic field and its possible spin-down effect.
-It's thought that AML has a direct influence in the mixing processes and that transport of the momentum could be produced by a number of relevant mechanisms: mass loss, magnetic fields and gravity waves (g-modes).
-Concerning the two latter, gravity waves \citep{Charbonnel2005} and magnetic fields \citep{Eggenberger2009} have the property of transmiting angular momentum (AM) much more effectively than inducing mixing \citep{Denissenkov2007}.
+However, these previous studies focused mainly on the hydrostatic effects and did not consider the influence of a coupling between the star magnetic field and its possible spin-down effect.
+It is thought that AML has a direct influence in the mixing processes and that transport of the momentum could be produced by a number of relevant mechanisms: mass loss, magnetic fields and gravity waves (g-modes).
+Concerning the two latter, gravity waves \citep{Charbonnel2005} and magnetic fields \citep{Eggenberger2009} have the property of transmitting angular momentum (AM) much more effectively than inducing mixing \citep{Denissenkov2007}.
 As a consequence of this increment in the AM transport efficiency, the amount of differential rotation between the radiative and the convective zones of the star is reduced (a solid-body rotation is fostered), as are the induced rotational instabilities.
-Of particular interest for this work is the impact caused by the magnetic field of the star in the AML, more specifically the effect produced on the solar wind particles that are forced to rotate in sync with it.
-As a consequence of this coupling a torque is induced which slows down the star, i.e.
-the magnetic braking effect.
-These magnetic fields would originate in the limits between the radiative and convective zones of the star, in the so-called tachoclines, due to the rotational difference between both zones \citep{Guerrero2016}.
+Additionally, magnetic fields would also originate in the limits between these zones, in the so-called tachoclines \citep{Aschwanden2014, Guerrero2016}, which interact with the solar wind particles that are forced to rotate in sync with it and therefore slow down the star, i.e.
+the magnetic braking (MB) effect.
 \par
@@
 the magnetic braking.
-In view of these findings, it seems evident that there are still open questions to be resolved on the subject of the processes that participate, directly or indirectly, in the mixing mechanisms that are not incorporated in the standard stellar models.\par
-
-Therefore, it's essential a proper consideration of the interactions between rotation and magnetic fields when it comes to study the AM distribution.
-We have adapted the one-dimensional stellar evolution code used in this work to model MB and included its effect on the equations which govern both AM distribution and stellar structure.
-This approach produces a more complete, though still reasonably simple, model for AM evolution.
-It's worth to mention that in this study we don't only focused in investigating the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.\par
-
-According to the prevalent theory of rotational mixing, southern circulation and internal shears leads to a homogenization of the material depleted of Li found within the CZ, causing that the inner layers where there is a temperature higher than $T_{Li}$ and where Li is destroyed, been mixed with those cooler above them.
-These mechanism causes a positive feedback loop in which the cooler material sinks, is heated again and the remaining Li is destroyed, totally or partially, before proceeding to rise once more into cooler layers.
+In view of these findings, it seems evident that there are still open questions to be resolved on the subject of the processes that participate, directly or indirectly, in the mixing mechanisms that are not incorporated in the standard stellar models.
+Therefore, it is essential a proper consideration of the interactions between rotation and magnetic fields when it comes to study the AM distribution.
 \par
 
-In the remainder of this paper, we describe a physically sound yet computationally simple model extension to Modules for Experiments in Stellar Astrophysics star evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.
-In its simplest form, the implementation of the model requires the specification of both the surface magnetic field strength ($B$) and the initial rotation rate ($\Omega$) at star surface.
-The models include already rotation during the PMS as there evidences that advocate for a strong established relationship between Li destruction and rotation on that phase (\citeauthor{Bouvier2016} \citeyear{Bouvier2016}, \citeyear{Bouvier2018}), and the numerical simulations trace the rotational history and A(Li) of a $1\, M_{\sun}$ star for a variety of initial values for $B$ and $\Omega$.\par
-
 \subsection{Effects of magnetic fields}
-The origin of magnetic fields is still unknown.
-The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields \citep[see][for details about the theory of fossil magnetic field]{Dudorov2014}, or fields produced through a dynamo mechanism \citep{Cantiello2009}.
+The origin of magnetic fields in the stars is still unknown.
+The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields \citep[see][for details ]{Dudorov2014}, or fields produced through a dynamo mechanism \citep{Cantiello2009}.
 Besides that, the presence of magnetic fields with intensities of the order of kG \citep{Hussain2014} has been a sine qua non condition for dealing with certain data observed in T Tauri type stars in young, accreting PMS systems \citep{Johns-Krull2007}.\par
@@
 Massive stars do not have external convective zones although they can expose a convective core.
-On the other hand, the less massive stars such as the Sun, have convective covers and a radiative core.
-In both types of stars strong magnetic fields can be generated in the tachocline but with notable differences (see \citep{Chabrier2006} \citet{Charbonneau2010}\ for details).\par
+On the other hand the less massive stars, such as the Sun, have convective covers and a radiative core.
+In both types of stars strong magnetic fields can be generated in the tachocline but with notable differences \citep[e.g.][for details]{Chabrier2006,Charbonneau2010}.\par
 
@@
 Our Sun is an active star as far as the magnetic field is concerned.
-According to the theoretical models this magnetic field is generated by a dynamo type effect in the tachocline, the intermediate zone between the radiative core and the CZ \citep{Aschwanden2014}, and has an intensity of approximately $B\approx10^5\, G$.
-Additionally, these models predict that because of the differential rotation existing between the core and CZ, the lines of the magnetic field that at the moment of their generation have a poloidal orientation will be affected by the differential rotation and forced to adopt a toroidal typology (the $\omega$-effect).
-On the other hand, we have that the opposite case can also occur, that is to say, starting from a toroidal magnetic field to obtain a poloidal configuration.
-This would be possible on the one hand, thanks to the convective movements of material that are produced in the CZ of the star and on the other, to assume that this material behaves as a perfect conductor.
+According to the theoretical models this magnetic field is generated by a dynamo in the tachocline and with an intensity about $B\approx10^{5}\,\Gauss$ \citep{Maeder2003a,Dudorov2014}.
+At the moment of their generation, the lines of that magnetic field will have a poloidal orientation and will be affected by the differential rotation and forced to adopt a toroidal typology (the $\omega$-effect).
+On the other hand, the opposite case can also occur, starting from a toroidal magnetic field to obtain a poloidal configuration.
+This would be possible thanks to the convective movements of material that are produced in the CZ of the star and to assume that this material behaves as a perfect conductor.
 This second condition allows to assume that the magnetic field lines are "frozen" in the conductive fluid and therefore must move in solidarity with it ($\alpha$-effect).\par
@@
 \subsubsection{Surface Magnetic Fields} \label{surf_mf}
-The intensity of the magnetic field measured on the Sun's surface is about $B\approx1\, G$, twice the value of the Earth's one.
+The intensity of the magnetic field measured on the Sun's surface is about $B\approx1\,\Gauss$ \citep{Weber1967,DAntona2000,Morin2012}, twice the value of the Earth's one.
 Additionally its topology is varied and complex.
-The dynamo model establishes that the magnetic field produced inside the star reaches the surface with the help of convective processes taken place in the CZ and gives rise to magnetically active regions on they reach the solar surface.
+The dynamo model establishes that the magnetic field produced inside the star reaches the surface with the help of convective processes taken place in the CZ and gives rise to magnetically active regions on the solar surface.
 Besides, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism \citep[e.g.][]{Brandenburg2004,Charbonneau2010,Brun2017}.
-Whatever the origins of surface magnetic fields are, it's expected to couple to the wind mass-loss and, if this is strong enough, to produce MB \citep[e.g.][]{UdDoula2002,Ud-Doula2007,Ud-Doula2008,Meynet2010}.\par
+Whatever the origins of surface magnetic fields are, it is expected to couple to the wind mass-loss and, if this is strong enough, to produce MB \citep[e.g.][]{UdDoula2002,Ud-Doula2007,Ud-Doula2008,Meynet2010}.\par
 
 \subsection{Effects of magnetic braking}
-In its most widely accepted approach, the MB is linked to Skumanich's work in which he develops an empirical law of evolution of the MB according to the age of the star.
-To calibrate this law, Skumanich relied on data obtained from type G stars found in the MS \citep{Skumanich1972}.
-The influence of the MB on the evolution of the star is closely related to the transport of the AM and by extension, influenced by how the chemical elements are transported within it.
-According to this approach \citep{Meynet2010}, two main mechanisms are distinguished attending to how the transport of the AM is produced which influence the mixing of these elements:
+In its most widely accepted approach, the MB is linked to Skumanich's work \citep{Skumanich1972} in which he develops an empirical law of its evolution in parallel to the age of the star.
+To calibrate this law, Skumanich relied on data obtained from type G stars found in the MS.
+The influence of the MB on the evolution of the star is closely related to the transport of the AM.
+According to \citet{Meynet2010}, two main mechanisms are distinguished attending to how the transport of the AM is produced:
 
@@
     \item Differential rotation: the AM transport is pushed by southern currents and shear instabilities.
-    \item Solid body rotation: when the transport of AM is very efficient, the solid body rotation is the prevailing mechanism maintained throughout the entire MS phase and the chemical elements are transported by the southern currents.
+    \item Solid body rotation: when the transport of AM is very efficient, solid body rotation is 
+    maintained throughout the entire MS phase.
 \end{enumerate}
@@
 The AML caused by MB depends directly on the amount of mass lost by the star due to the stellar winds.
-The estimated mass loss ratio for a solar-type star, approximately $10^{-13}M_{\odot} \; yr^{-1}$, and the resulting AML derived from that value (refer to \ref{mod_mb} for details) can be considered relatively modest as to decisively influence the evolution of the star.
+The estimated mass loss ratio for a solar-type star, approximately $10^{-14}\msun \; yr^{-1}$ \citep{Noerdlinger2008}, and the resulting AML derived from that value can be considered relatively modest as to decisively influence the evolution of the star.
 Furthermore, the combined effect of convective movements and differential rotation lead to the generation of magnetic fields.
-Those magnetic fields end up reaching the surface of the star \citep{Langer2012}, so that the AML can be increased in several orders of magnitude.
-That increment is as a consequence of the magnetic field that forces the stellar wind ionized particles to turn with the same angular speed, winds which extends a distance several times the radius of the star \citep[see][for more details]{UdDoula2002,Ud-Doula2007,Ud-Doula2008}.
-
-This paper is structured as follows: Section \ref{sec_2} gives an overview of rotation and magnetic fields physical processes and how do they impact on the star evolution.
-We also summarized here the most relevant parameters configured in MESA for the model simulations.
-In Section \ref{sec_3}, the model outputs are presented and discussed in detail.
-Finally, Section \ref{sec_4} finishes this paper with a discussion about open points and future works.\par
+Those magnetic fields end up reaching the surface of the star, so that the AML can be increased in several orders of magnitude \citep{Langer2012}.
+That increment is a consequence of the magnetic field that forces the stellar wind ionized particles to turn with the same angular speed, winds which extends a distance several times the radius of the star \citep[see][for more details]{UdDoula2002,Ud-Doula2007,Ud-Doula2008}.
 
 \section{Method} \label{sec_2}
+
+In the remainder of this paper, we describe a physically sound yet computationally simple model which will be used as an extension to Modules for Experiments in Stellar Astrophysics stellar evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.
+
+\subsection{Modelling magnetic braking} \label{mod_mb}
+We'll start enumerating the most relevant aspects and assumptions made in modelling the evolution of rotation, magnetic braking and angular momentum in MESA for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.\par
+
+In the presence of mass-loss, AM is removed by the material in the stellar wind, resulting in a spin-down effect.
+This is quite important for example in massive stars, that are known to be rapidly rotating and have strong stellar winds.
+It has been also observed that strongly magnetic intermediate-mass stars typically have rotation rates much slower than other stars in their parent population \citep{Mathys2006}.
+In those stars, the presence of such magnetic field will interact with the mass loss and the Alfv\'{e}n radius ($R_{A}$) plays an important role in this interaction.\par
+
+A relevant parameter to characterize the influence of a given magnetic field on the stellar wind is the denominated wind confinement magnetic parameter ($\eta_*$).
+It represents a energy ratio (Eq.~\ref{eq:wind_conf}) and defines a characteristic parameter for the relative effectiveness of the magnetic fields in circumscribing and/or channeling the wind outflow \citep{UdDoula2002}.\par
+
+\begin{ceqn}
+\begin{equation}
+    \eta_* = \frac{B^{2}/8\upi}{\rho\nu^2/2} \label{eq:wind_conf}
+\end{equation}
+\end{ceqn}
+
+Making using of ($\eta_*$), the $R_{A}$ is defined as the point in which the magnetic field energy density $B^{2}/8\upi$, marked by magnetic field strength ($B$), and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$, defined by the mass density ($\rho$) and velocity of the stellar wind $(\nu)$, are balanced.
+That is, where the coefficient between both is equal to 1.
+In case that $R_{A}$ is greater than the star radius, then the wind flow will have to follow the magnetic field.
+As a consequence, the material will leave the stellar surface with a higher specific AM, as the co-rotation radius has increased and it will roughly correspond to the $R_{A}$.
+This effect is known as MB.\par
+
+Each of these factors has a decisive influence on the future of the star and therefore in our routine implementation.
+On the one hand, stars with similar initial masses but different mass loss ($\Dot{M}$) ratios will end up evolving very differently.
+On the other hand, the ionized particles carried by the solar wind not only contribute to the loss of mass but also to the loss of kinetic ($K_e$) energy that is deposited in the interstellar medium.
+Given a star with a spherically symmetric wind, $\Dot{M}$ and $K_e$ are characterized and related by the following expressions:
+
+\begin{ceqn}
+\begin{align}
+    \Dot{M} &= 4\upi r^2\rho\nu \label{eq:mass_loss}\\
+    K_e &= 0.5\Dot{M}\nu_{esc}
+\end{align}
+\end{ceqn}
+
+Using (\ref{eq:wind_conf}) and (\ref{eq:mass_loss}), $\eta_*$ can be approximated by: 
+\begin{ceqn}
+\begin{equation}
+    \eta_* = \frac{B^{2}r^{2}}{\Dot{M}\nu} \label{eq:wind_conf2}
+\end{equation}
+\end{ceqn}
+
+In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval, and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity ($\nu_\infty$).
+In general, a magnetically channeled outflow will have a complex stream geometry but for convenience, the Eq.~\ref{eq:mass_loss} simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate.
+$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law, where $r$ represents the distance from the center of the star to the point where the velocity is to be measured, and $R_*$ is the radius of the star:
+\begin{ceqn}
+\begin{equation}
+    \nu(r) = \nu_\infty (1-R_*/r)
+\end{equation}
+\end{ceqn}
+where $\nu_\infty$ is the terminal wind velocity defined as the velocity that the wind, the outflowing matter, reaches at large distance from the central star, where it is not accelerated anymore by the wind driving force but its deceleration due to interaction with the interstellar medium (ISM) is negligible \citep{Niedzielski2002}.\par
+
+The line-driven winds of massive OB stars have terminal velocities that scale with the photospheric escape velocity \citep{Lamers2000} according to:
+\begin{ceqn}
+\begin{align}
+\nu_{esc} &= \sqrt{\frac{2GM_*}{R_*}} \label{eq:vesc} \\
+\nu_\infty &\simeq 1.92 \;\nu_{esc}\\
+\nu_\infty &\simeq 1.92 \; x \; 618 \; \Bigg(\sqrt{\frac{R_{\sun}}{R_*}\frac{M_*}{\msun}} \;\Bigg) \label{eq:vinf}
+\end{align}
+\end{ceqn}
+where Eq.~\ref{eq:vesc} describes the Newtonian escape velocity from the stellar surface, $G$ is the gravitational constant and Eq.~\ref{eq:vinf} represents the terminal velocity in terms of escape velocity, radius and mass of the Sun.\par
+
+Given that the version of MESA used in this work does include neither the treatment of surface magnetic fields (it does account for transport by internal magnetic fields of AM and chemical elements) and therefore, nor for magnetic braking effects, it is necessary to extend the simulator to include this phenomenon.
+In the present work we have followed the theoretical work on MB in the Sun carried out by \citet{Weber1967}, making use of the expression \ref{eq:j_dot} for the calculation of the AML.
+\begin{ceqn}
+\begin{equation}
+ \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A} \label{eq:j_dot}
+\end{equation}
+\end{ceqn}
+where $\Dot{M}$ is the mass loss rate, $\Omega$ the angular velocity at star surface and $R_A$ the Alfv\'{e}n radius. \par
+
+The above expression can be rewritten so that it depends on the $\eta_*$ instead of the $R_A$ as follows:
+\begin{ceqn}
+\begin{equation}
+ \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{*}\eta_* \label{eq:j_dot_mesa}
+\end{equation}
+\end{ceqn}
+which is a more convenient expression to be implemented in MESA because is based on values directly exposed during the simulations and convenient for the approach described in \citet{Ud-Doula2007}.
+
+\section{Results} \label{sec_3}
+
+\subsection{Mesh Models} \label{sec_mesh}
+
+In its simplest form the implementation of the model requires the specification of both the surface magnetic field strength ($B$) and the initial rotation rate ($\Omega$) at star surface.
+The numerical simulations trace the rotational history and A(Li) of a $1\, \msun$ star for a variety of initial values for $B$ and $\Omega$.
+Although stars have a three-dimensional structure and the equations that model their evolution must take this fact into account,it has been proven sufficient to solve the equations in only one dimension, extrapolating the results to the other two \citep[shellular approximation,][]{Meynet1997}.
+MESA uses this simplification to accelerate computational calculations \citet{Paxton2013}.\par
+
 \begin{table}
@@
 		Magnetic Effects & Magnetic braking based on idealized \\ & monopole field\\
-		Magnetic Field & B(G) variable between [3.5 - 5.5]\\
-		Mass Loss & activated, $\Dot{M}_{max} = 10^{-3} \: M_{\sun} \: yr^{-1}$\\
+		Magnetic Field & B(G) variable between [3.0 - 5.0]\\
+		Mass Loss & activated, $\Dot{M}_{max} = 10^{-3} \: \msun \: yr^{-1}$\\
 		Angular Moment Loss & activated, $\Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A}$\\
@@
 
-We define here also some conventions and assumptions adopted in the paper.
-For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, proto} = 0.0142$) not the current photospheric metallicity ($Z = Z_{\odot, photosphere} = 0.0134$) as the reference value.
-We also adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10}\, cm$ and $M_{\odot} = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
+We define here also additional conventions and assumptions adopted in the simulations.
+For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, \mathrm{pr} = 0.0142$) not the current photospheric metallicity ($Z = Z_{\odot, \mathrm{ph} = 0.0134$) as the reference value.
+We also adopt the following nominal values to express stellar properties values in SI units $\rsun = 6.957x10^{10}\, cm$ and $\msun = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
 For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}.
@@
 
-In this section we review the core physic aspects simulated in MESA and the most relevant parameters adopted in the numerical simulations.
-In its simplest version, by activating rotation in the model, the evolutionary computational code MESA takes into account the following effects:
-%\vspace{-1pc}
+The models include already rotation during the PMS as there evidences that advocate for a strong established relationship between Li destruction and rotation on that phase (\citeauthor{Bouvier2016} \citeyear{Bouvier2016}, \citeyear{Bouvier2018}), and by activating it in the MESA models, the evolutionary computational code takes into account the following effects:
+
 \begin{enumerate}
@@
 The enumerated effects \ref{itm:1} and \ref{itm:2} are standard features offered by MESA.
-On the contrary, the effect \ref{itm:3} isn't available as an out-of-the-box feature during the elaboration of this paper and represents our extension to MESA.
-
-The rest of this section is devoted to describe the most relevant aspects and assumptions made in modelling the evolution of rotation, magnetic braking and angular momentum in MESA.\par
-
-\subsection{Modelling rotation}
-The rotation to which a star is subjected has profound implications for its inner structure and, therefore, for the way in which it evolves throughout its life.
-It is also responsible for inducing through hydrodynamic instabilities the mixing of the chemical elements that make it up.
-In order to be able to correctly model those induces mixing effects consistently with the observational data, it is necessary to analyze and understand more precisely the mechanisms that govern the distribution of AM \citep{Pinsonneault1997,Maeder2000}, since in this area the theoretical models also offer deficiencies to explain the observations \citep{Denissenkov2007}.
-\par
-
-Stars have a three-dimensional structure and the equations that model their evolution must take this fact into account.
-However, it has been proven sufficient to solve the equations in only one dimension, extrapolating the results to the other two (shellular approximation \citep{Meynet1997}) MESA uses this simplification to accelerate computational calculations \citet{Paxton2013}.\par
-
-\subsection{Modelling magnetic braking} \label{mod_mb}
-In this section, we describe a physically sound yet computationally simple model extension to Modules for Experiments in Stellar Astrophysics star evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.
-In its simplest form, the implementation of the model requires the specification of both the surface magnetic field strength ($B$) and the initial rotation rate ($\Omega$) at star surface.
-The models include already rotation during the PMS as there evidences that advocate for a strong established relationship between Li destruction and rotation on that phase (\citeauthor{Bouvier2016} \citeyear{Bouvier2016}, \citeyear{Bouvier2018}), and the numerical simulations trace the rotational history and A(Li) of a $1\, M_{\sun}$ star for a variety of initial values for $B$ and $\Omega$.\par
-
-In the presence of mass-loss, AM is removed by the material in the stellar wind, resulting in a spin-down effect.
-This is quite important for example in massive stars, that are known to be rapidly rotating and have strong stellar winds.
-It has been also observed that strongly magnetic intermediate-mass stars typically have rotation rates much slower than other stars in their parent population \citep{Mathys2006}.
-In those stars, the presence of such magnetic field will interact with the mass loss and the Alfv\'{e}n radius ($R_{A}$) plays an important role in this interaction.
-The $R_{A}$ is defined as the point in which the magnetic field energy density $B^{2}/8\upi$, marked by magnetic field strength ($B$), and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$, defined by the mass density ($\rho$) and velocity of the stellar wind $(\nu)$, are balanced.
-That is, where the coefficient between both is equal to 1.
-In case that $R_{A}$ is greater than the star radius, then the wind flow will have to follow the magnetic field.
-As a consequence, the material will leave the stellar surface with a higher specific AM, as the co-rotation radius has increased and it will roughly correspond to the $R_{A}$.
-This effect is known as MB.\par
-
-In the present work we do not take into account either the influence of internal magnetic fields or their existence during the T-Tauri phase.
+On the contrary, the effect \ref{itm:3} is not available as an out-of-the-box feature during the elaboration of this paper and represents our extension to MESA.
+Additionally, we do not take into account either the influence of internal magnetic fields or their existence during the T-Tauri phase.
 We adopt a simple and pragmatic theoretical approach to establish their occurrence when the star is reaching the ZAMS phase.
 The adopted criterion is based on the simultaneous existence of an extensive convective layer and a radiative core.
-From this moment on, the MB routine will be activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participates in the star AML.
+From this moment on, the MB routine will be activated acting as an additional mechanism to those existing in the MESA evolutionary code that participates in the star AML.
 Additionally, we assume that the magnetic field does not vary its intensity throughout the evolution of the star until it reaches the TAMS.\par
 
-The energy ratio (eq.
-\ref{eq:wind_conf}) is denominated wind confinement magnetic parameter ($\eta_*$) and defines a characteristic parameter for the relative effectiveness of the magnetic fields in circumscribing and/or channeling the wind outflow \citep{UdDoula2002}.\par
-
-\begin{ceqn}
-\begin{equation}
-    \eta_* = \frac{B^{2}/8\upi}{\rho\nu^2/2} \label{eq:wind_conf}
-\end{equation}
-\end{ceqn}
-
-Each of these factors has a decisive influence on the future of the star and therefore in our routine implementation.
-On the one hand, stars with similar initial masses but different mass loss ($\Dot{M}$) ratios will end up evolving very differently.
-On the other hand, the ionized particles carried by the solar wind not only contribute to the loss of mass but also to the loss of kinetic ($K_e$) energy that is deposited in the interstellar medium.
-Given a star with a spherically symmetric wind, $\Dot{M}$ and $K_e$ are characterized and related by the following expressions:
-
-\begin{ceqn}
-\begin{align}
-    \Dot{M} &= 4\upi r^2\rho\nu \label{eq:mass_loss}\\
-    K_e &= 0.5\Dot{M}\nu_{esc}
-\end{align}
-\end{ceqn}
-
-Using (\ref{eq:wind_conf}) and (\ref{eq:mass_loss}), $\eta_*$ can be approximated by: 
-\begin{ceqn}
-\begin{equation}
-    \eta_* = \frac{B^{2}r^{2}}{\Dot{M}\nu} \label{eq:wind_conf2}
-\end{equation}
-\end{ceqn}
-
-In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval, and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity ($\nu_\infty$).
-In general, a magnetically channeled outflow will have a complex stream geometry but for convenience, the expression (\ref{eq:mass_loss}) simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate.
-$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law, where $r$ represents the distance from the center of the star to the point where the velocity is to be measured and $R_*$ is the radius of the star:
-\begin{ceqn}
-\begin{equation}
-    \nu(r) = \nu_\infty (1-R_*/r)
-\end{equation}
-\end{ceqn}
-where $\nu_\infty$ is the terminal wind velocity defined as the velocity that the wind, the outflowing matter, reaches at large distance from the central star, where it is not accelerated anymore by the wind driving force but its deceleration due to interaction with the interstellar medium (ISM) is negligible \citep{Niedzielski2002}.\par
-
-The line-driven winds of massive OB stars have terminal velocities that scale with the photospheric escape velocity \citep{Lamers2000} according to:
-\begin{ceqn}
-\begin{align}
-\nu_{esc} &= \sqrt{\frac{2GM_*}{R_*}} \label{eq:vesc} \\
-\nu_\infty &\simeq 1.92 \;\nu_{esc}\\
-\nu_\infty &\simeq 1.92 \; x \; 618 \; \Bigg(\sqrt{\frac{R_{\sun}}{R_*}\frac{M_*}{M_{\sun}}} \;\Bigg) \label{eq:vinf}
-\end{align}
-\end{ceqn}
-where (\ref{eq:vesc}) is the Newtonian escape velocity from the stellar surface, $G$ is the gravitational constant and expression (\ref{eq:vinf}) represents the terminal velocity in terms of escape velocity, radius and mass of the Sun.\par
-
-\begin{figure}
-	\includegraphics[width=\columnwidth]{figures/terminal_vel_vs_escape_vel.png}
-    \caption {Terminal vel vs escape vel.}
-    \label{fig:terminal_vel_vs_escape_vel}
-\end{figure}
-
-Given that the version of MESA used in this work does include neither the treatment of surface magnetic fields (it does account for transport by internal magnetic fields of AM and chemical elements) and therefore, nor for magnetic braking effects, it is necessary to extend the simulator to include this phenomenon.
-In the present work we have followed the theoretical work on MB in the Sun carried out by \citet{Weber1967}, making use of the expression (\ref{eq:j_dot}) for the calculation of the AML.
-\begin{ceqn}
-\begin{equation}
- \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A} \label{eq:j_dot}
-\end{equation}
-\end{ceqn}
-where $\Dot{M}$ is the mass loss rate and $R_A$ the Alfv\'{e}n radius. \par
-
-The above expression can be rewritten so that it depends on the $\eta_*$ instead of the $R_A$ as follows:
-\begin{ceqn}
-\begin{equation}
- \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{*}\eta_* \label{eq:j_dot_mesa}
-\end{equation}
-\end{ceqn}
-which is a more convenient expression to be implemented in MESA because is based on values directly exposed during the simulations and convenient for the approach described in \citet{Ud-Doula2007}.
-
-\section{Results} \label{sec_3}
-
-\subsection{Mesh Models}
-We define here also some conventions and assumptions adopted in the paper.
-For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, proto} = 0.0142$) not the current photospheric metallicity ($Z = Z_{\odot, photosphere} = 0.0134$) as the reference value.
-We also adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10}\, cm$ and $M_{\odot} = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
-For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}.
-That work, in particular its section 4.1, has been used as a starting point for ours regarding the parameterization of the MESA project which is calibrated to reproduce the measured element abundances on the solar surface.\par
-
+In this study we do not only focused in investigating the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.
 Keeping in mind this approach, our major objective is aimed to check the degree of validity of the resulting model after including the MB effect as a complementary mechanism of AML that can contribute to explain the evolution of Li in the Sun and other solar-type stars.
-A set of different simulation scenarios with magnetic field strengths ranging from between 3 and 5.5 G have been exercised.
+A set of different simulation scenarios with magnetic field strengths ranging from between 3.0 and 5.0 G have been exercised.
 We'd like to stress that the magnetic field remains constant during the PMS \& MS phases.
-We compute the evolution of $1M_{\sun}$ stellar models at solar metallicity with $\Omega / \Omega_{crit}$ varying between $0.0084$ and $0.0336$.
+We compute the evolution of $1\msun$ stellar models at solar metallicity with $\oomegac$ varying between $0.0084$ and $0.0336$.
 More advanced models for more complex arrangements of the magnetic field as well as the evolution of its intensity throughout the life of the star are left for later work.\par
 
-The AML is directly influenced by the $R_A$ (or alternatively by the $\eta_*$), $\Omega$ and $\Dot{M}$.
+As expressed by Eq.~\ref{eq:j_dot_mesa}, the amount of AML is directly influenced by the $R_A$ (or alternatively by the $\eta_*$), $\Omega$ and $\Dot{M}$.
 If the value of $\eta_*$ is large, the obtained AML will be also large and the star undergoes a considerable deceleration.
 In our proposal, the value of $\eta_*$ can be directly influenced by establishing in the different simulations the value of the magnetic field ($B$) on the surface or the rotation speed of the star ($\Omega$).
-With regards to $\Dot{M}$ and as reported in table (\ref{tab:phy_mesa}), for the set of simulations performed by MESA the empirical formula developed by Reimers \citet{Reimers1975} for stars in the asymptotic giant branch (AGB) is used for calculating its value.
-For a solar-type star the $\Dot{M}$ during MS is relatively small, about  $10^{-14}M_{\odot} \, yr^{-1}$ \citet{Noerdlinger2008}.
-Finally, $\Omega$ can be influenced in our models specifying the ratio $\Omega / \Omega_{crit}$ as another free parameter, where  $\Omega_{crit}$ represents the star surface velocity at which the centrifugal force balances  the gravitational force.
+With regards to $\Dot{M}$ and as reported in table (\ref{tab:phy_mesa}), for the set of simulations performed by MESA the empirical formula developed by Reimers \citep{Reimers1975} for stars in the asymptotic giant branch (AGB) is used for calculating its value.
+For a solar-type star the $\Dot{M}$ during MS is relatively small, about  $10^{-14}\msun \, yr^{-1}$ \citep{Noerdlinger2008}.
+Finally, $\Omega$ can be influenced in our models specifying the ratio $\oomegac$ as another free parameter, where  $\omegac$ represents the star surface velocity at which the centrifugal force balances  the gravitational force.
 Beyond this point, the star would begin to expels matter along the equator.\par
@@
 After that, an AM value is assigned to each cell $k$ ($J_k$).
-It's in this point when our MB routine plugs in and influences the internally calculated value by providing an additional contribution ($\Dot{J}_{k}$) which is the result of the external torque exerted by the magnetic field once it has been distributed among the different layers that make up the CZ (\ref{eq:k_jdot}):\par
+It is in this point when our MB routine plugs in and influences the internally calculated value by providing an additional contribution ($\Dot{J}_{k}$) which is the result of the external torque exerted by the magnetic field once it has been distributed among the different layers that make up the CZ as dictated by Eq.~\ref{eq:k_jdot}:\par
 
@@
 
-\subsection{Angular velocity, Li abundance and effective temperature}
+\subsection{Li evolution without MB}
 
@@
 	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip, width=\columnwidth]{figures/li_var_vel_0_0g.eps}
-    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
+    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
 The solid black line represents the reference model according to \citet{Choi2016}.
-The rest of lines are models which include PMS rotation with $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336, respectively.
-The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
+The rest of lines are models which include PMS rotation with $\oomegac$ between 0.0084 and 0.0336, respectively.
+The purple star and square are surface Li abundances for the present-day Sun \citep{Asplund2009} and the average for the Pleiades cluster \citep{Sestito2005} respectively.
 The dashed lines make reference to the ZAMS.}
@@
 
-Figure (\ref{fig:li_var_vel_0g}) shows the evolution of surface Li abundance relative to H as a function of time and for several $1M_{\odot}$ models which take into account the effects of rotation and AML caused by stellar winds but not MB.
+Figure~\ref{fig:li_var_vel_0g} shows the evolution of surface Li abundance relative to H as a function of time and for several $1\msun$ models initialized with different rotational velocities.
+The simulations take into account the effects of rotation and AML caused by stellar winds but not those of MB.
 The purple star and square are surface Li abundances for the present-day Sun \citep{Asplund2009} and Pleiades \citep{Sestito2005}.\par
 
-The figure (\ref{fig:li_var_vel_0g}) clearly allows to appreciate how the Li abundance in the star surface decreases over time for all simulated models.
+Additionally, notice how the Li abundance in the star surface decreases over time for all simulated models.
 The solid black line represents the reference model that adopts the solar-calibrated envelope overshooting parameters as documented in \citet{Choi2016}.
-In general, all models burn too much Li before the ZAMS and therefore do not match with the Pleiades average surface Li abundance.
+All models burn too much Li before the ZAMS and therefore do not match with the Pleiades average surface Li abundance.
 Also remarkable is the fact that there are barely any differences between the different models in terms of the abundance of Li for much of the PMS.
@@
 On the one hand, the convective zones that the models develop have very similar sizes, so that the temperature at the BCZ is practically the same.
-On the other hand, the differential rotation between the core and the convective zone does not begin to develop until it reaches $10^6$ years, achieving its maximum difference in the ZAMS about  $10^7$ years.
-It is at this moment that the combined effects of turbulence, rotational difference and magnetic braking should increase, leading to notable differences in the evolution of A(Li) (see figure (\ref{fig:li_var_vel_0g_z1})).
+On the other hand, the differential rotation between the core and the convective zone does not begin to develop until it reaches $ \approx 10^6$ years, achieving its maximum difference in the ZAMS about $ \approx 10^7$ years.
+It is at this moment that the combined effects of turbulence and rotational difference should increase, leading to notable differences in the evolution of A(Li) (see Figure~\ref{fig:li_var_vel_0g_z1}).
 Later, the reference model does not deplete Li efficiently on the MS and fail (again) to match the current solar surface Li abundance.
-On the other hand, the other models which include rotation during the PMS with values of $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but its rotational velocity is much higher (see figure \ref{fig:rot_vel_0g}) than the $2\,kms^{-1}$ of the Sun \citep{Gill2012}.
+On the other hand, the other models which include rotation during the PMS with values of $\oomegac$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but its rotational velocity is much higher (see Figure~\ref{fig:rot_vel_0g}) than the $2\,\kms$ of the Sun \citep{Gill2012}.
 \par
 
-For much of the PMS the star rotates as a solid body (see figure (\ref{fig:rot_vel_0g})) and this is because the star has a completely convective interior.
+For much of the PMS the star rotates as a solid body (see Figure~\ref{fig:rot_vel_0g}) and this is because the star has a completely convective interior.
 It is not until the end of Hayashi track that the star begins to develop a radiative core due to the increase in temperature that occurs inside the star during the contraction process.
@@
 As the models are initialized with a higher angular velocity, the difference in velocity between the BCZ and the star surface is accentuated in a directly proportional mode; the higher the initial velocity, the bigger the velocity gradient between the bottom and top limits of the CZ.
-As a consequence, the turbulence strength located at the BCZ increases so that Li can reach regions with temperatures about $T_{Li}$ where it is finally burned and destroyed (see figure (\ref{fig:li_var_vel_0g})).
-Other investigations \citep{Bouvier2018, Baraffe2017}  point to a tendency diametrically opposed to the one exposed here, the faster the rotation speed, the greater the abundance of Li on the surface of the star.
+As a consequence, the turbulence strength located at the BCZ increases so that Li can reach regions with temperatures about $\tli$ where it is finally burned and destroyed (see figure~ \ref{fig:li_var_vel_0g}).
+Other investigations \citep{Bouvier2018, Baraffe2017}  point to a tendency diametrically opposed to the one exposed here, i.e.
+the faster the rotation speed, the greater the abundance of Li on the surface of the star.
 \par
@@
 	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
-    \caption {Similar to figure (\ref{fig:li_var_vel_0g}) but zooming in on the ZAMS.
+    \caption {Similar to Figure~\ref{fig:li_var_vel_0g} but zooming in on the ZAMS.
 The models with a higher initial rotational velocity reach already the ZAMS with a lower amount of Li measure on the star surface.}
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
-    \caption{The evolution of angular velocity at surface (solid line) and the top (dashed line) and bottom (dotted line) limits of the uppermost convective zone, as a function of time for several 1 $M_{\sun}$ models.
-The models includes PMS rotation with $\Omega / \Omega_{crit}$ values between 0.0084 and 0.0336.
+    \caption{The evolution of angular velocity at surface (solid line) and the top (dashed line) and bottom (dotted line) limits of the uppermost convective zone, as a function of time for several 1 $\msun$ models.
+The models includes PMS rotation with $\oomegac$ values between 0.0084 and 0.0336.
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
@@
 
-Another well known structural effect of rotation is the decrease of the effective temperature ($T_{eff}$) and to a less extent of stellar luminosity ($L$).
-Both effects can be observed graphically in the HR diagram of figure (\ref{fig:hr_var_vel_0g}) which shows a zoomed-in view of evolution tracks from the ZAMS till the TAMS.
-If we compare the non-rotating model (black solid line on the left side) with the rotating ones we can recognize that at the end of the PMS, the latter reach before the ZAMS and exhibit a lower $T_{eff}$ than the former.
-These results are in line with those of previous studies (see e.g.
-\citet{Eggenberger2012}, \citet{Piau2001}, \citet{Pinsonneault1989}).\par
+Another well known structural effect of rotation is the decrease of the effective temperature ($\teff$) and to a less extent of stellar luminosity ($L$).
+Both effects can be observed graphically in the HR diagram of Figure~\ref{fig:hr_var_vel_0g} which shows a zoomed-in view of evolutionary tracks from the ZAMS until the TAMS for several $1\msun$ models initialized with different rotational velocities.
+If we compare the non-rotating model (black solid line on the left side) with the rotating ones we can recognize that at the end of the PMS, the latter reach before the ZAMS and exhibit a lower $\teff$ than the former.
+These results are in line with those of previous studies \citep[see e.g.
+][]{Eggenberger2012,Piau2001,Pinsonneault1989}.\par
+%(see e.g. \citet{Eggenberger2012}, \citet{Piau2001}, \citet{Pinsonneault1989}).\par
 
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/hr_var_vel_0_0g_z1.eps}
-    \caption{An example solar 1$M_{\sun}$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities.
-The rotation is activated in the models in the PMS and those models reach before the ZAMS and at a lower $T_{eff}$ than the non-rotating one (solid black line).
-The luminosity is expressed in terms of $L_{\sun}$.}
+    \caption{An example solar 1$\msun$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities.
+The rotation is activated in the models in the PMS and those models reach before the ZAMS and at a lower $\teff$ than the non-rotating one (solid black line).
+The luminosity is expressed in terms of $\lsun$.}
     \label{fig:hr_var_vel_0g}
@@
 
-\subsection{Magnetic braking, Li abundance and effective temperature}
-In the rest of this document we are going to focus, unless otherwise indicated, on models that include rotation from the PMS since they are the ones that can offer a greater approximation to the observations.
-In turn, these models will serve as a basis for others in which our magnetic braking routine is used by simulating magnetic fields of different intensity, varying from 3.5G to 5.5G in 0.5G increments.\par
+\subsection{Li evolution with MB}
+In the rest of this document we are going to focus, unless otherwise indicated, on models that include rotation from the PMS since they are the ones that can offer a greater approximation to the observations.\par
 
-If we concentrate our attention on figure (\ref{fig:li_var_vel_4_0g}) and compare it with figure (\ref{fig:li_var_vel_0g}) we will see how the profiles of Li abundance are altered during PMS and MS.
-In the first phase we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see equation (\ref{eq:j_dot})) depends directly on the loss of mass.
-If we take into account that for solar-type stars the models predict a modest total loss of mass, that value is even much lower during PMS.
+Figure \ref{fig:li_var_vel_4_0g} shows the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
+Those models were initialized with different rotational velocities and took into consideration both the effects of AML and the MB caused by a magnetic field of intensity 4G.
+If we compare it with Figure~\ref{fig:li_var_vel_0g} in which the effects of MB are neglected, we notice how the profiles of Li abundance are altered during PMS and MS.
+In the first phase we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see Eq.~\ref{eq:j_dot}) depends directly on the mass loss rate.
+If we take into account that for solar-type stars the models predict a modest total mass loss rate, that value is even much lower during PMS.
+On the contrary, during the MS it is observed that the AML is much more significant, causing the star to rotate more slowly and as a result a smaller amount of Li to be destroyed.\par
+
+If we concentrate our attention on Figure~\ref{fig:li_var_vel_4_0g} which represents the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models initialized with different rotational velocities and considers the effects of AML and the MB caused by a magnetic field of intensity 4G, and compare it with Figure~\ref{fig:li_var_vel_0g} in which the effects of MB are neglected, we will see how the profiles of Li abundance are altered during PMS and MS.
+In the first phase we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see Eq.~\ref{eq:j_dot}) depends directly on the mass loss rate.
+If we take into account that for solar-type stars the models predict a modest total mass loss rate, that value is even much lower during PMS.
 On the contrary, during the MS it is observed that the AML is much more significant, causing the star to rotate more slowly and as a result a smaller amount of Li to be destroyed.\par
@@
 	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_4_0g.eps}
-    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
-The models include a magnetic field with an intensity of 4G and PMS rotation with $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336, respectively.
+    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
+The models include a magnetic field with an intensity of 4G and PMS rotation with $\oomegac$ between 0.0084 and 0.0336, respectively.
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
@@
 
-The effect of the MB routine can be seen even more clearly in figures (\ref{fig:rot_vel_4g}) \& (\ref{fig:grid_rot_vel}).
-Here we represent the rotation profiles on the surface of the stars for different rotating models.
-Similarly to the evolution profiles of the Li on the surface commented in the previous paragraph, the effect of the routine is much more accentuated once the ZAMS is reached.
-If we compare the evolution of the curves presented here with those of figure (\ref{fig:rot_vel_0g}) we see how the star, instead of continuing to increase its $\Omega$, begins to slow down after having reached its maximum in the passage through the ZAMS.
+The effect of the MB routine can be seen even more clearly in figures \ref{fig:rot_vel_4g}, \ref{fig:rot_vel_4g_z1} \& \ref{fig:grid_rot_vel}.
+In those figures we represent the rotation profiles for the surface of the stars and for the limits of the convective envelope for a number of 1 $\msun$ models initialized with different rotational velocities and considering the influence of MB.
+Similarly to the evolution profiles of the Li on the surface commented in the previous paragraph, the effect of the routine is visible once the ZAMS is reached.
+If we compare the evolution of the curves presented here with those of Figure~\ref{fig:rot_vel_0g} we see how the star, instead of continuing to increase its $\Omega$, begins to slow down after having reached its maximum in the passage through the ZAMS.
+It is also at this stage when the angular velocities at the surface of the star and at the BCZ reach their maximum difference.
+On the other hand, the effect of magnetic braking causes the angular velocities between both zones of the star to decrease until, for an age close to that of the Sun (see Figure~\ref{fig:rot_vel_4g_z1}), the star practically rotates like a rigid solid.
 These results are also consistent with those obtained by \citet{Eggenberger2010} as far as the effect of the magnetic field, in particular its influence on the loss of angular momentum, has on the rotational velocity of the star.
-In our models we have also chosen to include the rotation effects offered by MESA but not those of Tayler-Spruit dynamo on the diffusion of elements.
-In a similar way we also observe that the models with lower angular rotation generally end up exhibiting higher values in the abundance of Li on the surface (see figures (\ref{fig:li_var_vel_4_0g}), (\ref{fig:grid_li_var_vel}) \& (\ref{fig:grid_li_var_g})).
+In our models we have also chosen to include the rotation effects offered by MESA but not those of Spruit-Tayler (ST) dynamo on the diffusion of elements.
+The reason is that the implementation of MESA that makes of the ST dynamo is to simulate internal magnetic fields and not for surface magnetic fields \citep{Paxton2013}.
+
+In a similar way we also observe that the models with lower angular rotation generally end up exhibiting higher values in the abundance of Li on the surface (see figures \ref{fig:li_var_vel_4_0g}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).
 In none of those cases we do obtain values of Li on the surface higher than those shown by the model without rotation.\par
-
-We also find interesting to highlight the behaviour of $\Omega$ profiles when approaching TAMS.
-It can be recognized in figure (\ref{fig:rot_vel_4g}) and in more detail in (\ref{fig:rot_vel_4g_z1}), that its evolution suffers from a series of ups and downs that do not appear previously.
-Everything seems to indicate that this behavior is due to a combined effect of the MB routine and the time-step resolution (probably too big) used by MESA during the last part of the simulation.
-We based this affirmation on the fact that this effect is not observed in the series of simulations in which the routine is not included (see figure (\ref{fig:rot_vel_0g})).\par
 
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g.eps}
-    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
-The models include a magnetic field with an intensity of 4G and PMS rotation with $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336, respectively.
-Additionally it's simulated the magnetic braking effect of magnetic field with an intensity of 4G.
+    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
+The models include a magnetic field with an intensity of 4\,G, PMS rotation with $\oomegac$ between 0.0084 and 0.0336, respectively and MB.
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g_z1.eps}
-    \caption{Similar to figure (\ref{fig:rot_vel_4g}) but now showing in detail the surface rotational velocity as the star approaches TAMS.}
+    \caption{Similar to Figure~\ref{fig:rot_vel_4g} but now showing in detail the surface rotational velocity as the star approaches TAMS.}
     \label{fig:rot_vel_4g_z1}
@@
 
-It's also remarkable to observed, for some of the models depicted in figure (\ref{fig:rot_vel_4g}), the lower $\Omega$ at the star surface (solid line) in comparison with the upper limit of the convective layer (dashed line).
+It is also remarkable to observe that, for some of the models depicted in Figure~\ref{fig:rot_vel_4g}, their $\Omega$ at the star surface (solid line) is slower than the one of the upper limit of their convective layer (dashed line).
 This fact is most probably a consequence of the expansion of the superficial layers after the mass loss produced by stellar winds.\par
 
-The MB also leaves its mark on the HR diagram by significantly affecting the $T_{eff}$ of the star.
-To visualize this effect we will take as reference figure (\ref{hr_vc_0336_var_g_z1}) in which all models were started with the same value $\Omega / \Omega_{crit}=0.0336$.
+The MB also leaves its mark on the HR diagram by significantly affecting the $\teff$ of the star.
+To visualize this effect we will take as reference Figure~\ref{hr_vc_0336_var_g_z1} in which all models were started with the same value $\oomegac=0.0336$ but the intensity of the simulated magnetic field was different.
 In particular, let us pay attention to the evolutionary trace in which a magnetic field was not simulated (black solid line) and to the rest of the evolutionary traces in which it was.
 We have that the group of simulations with presence of a magnetic field produces hotter stars due to the MB influence.
-The lower speed with which the star rotates due to the MB effect causes the increase of the $T_{eff}$, being this difference of practically $95\,K$ between the simulated models with $0\,G$ and $5.5\,G$ respectively for $log(L/L_{\sun})=0$.
+The lower speed with which the star rotates due to the MB effect causes the increase of the $\teff$, being this difference of practically $95\,\Kelvin$ between the simulated models with $0\,\Gauss$ and $5.0\,\Gauss$ respectively for $log(\llsun)=0$.
 
@@
 	\includegraphics[trim = 30mm 15mm 15mm 15mm, clip,width=\columnwidth]{figures/hr_vc_0336_var_g_z1.eps}
-    \caption{Similar to figure (\ref{fig:hr_var_vel_0g}) but now showing in detail the effects of magnetic braking on the evolutionary tracks for different magnetic field strengths and $\Omega / \Omega_{crit}=0.0336$.
+    \caption{Similar to Figure~\ref{fig:hr_var_vel_0g} but now showing in detail the effects of magnetic braking on the evolutionary tracks for different magnetic field strengths and $\oomegac=0.0336$.
 The presence of a magnetic field produces hotter stars due to the influence of the magnetic braking on the rotational velocity of the star.}
@@
 
-As described in section \ref{mod_ang_vel_ang_mom}, the MB routine distributes the total amount of AML calculated according to equation (\ref{eq:k_jdot}) among the different layers that compose it.
-In figure (\ref{fig:cz_vc_028_var_b}) we can observe the evolution of the most external CZ normalized with respect to the radius of the star.
+As described in section \ref{sec_mesh}, the MB routine distributes the total amount of AML calculated according to Eq.~\ref{eq:k_jdot} among the different layers that compose it.
+In Figure~\ref{fig:cz_vc_028_var_b} we can observe the evolution of the most external CZ normalized with respect to the radius of the star for several 1 $\msun$ models.
+The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths varying between $0\,\Gauss$ and $5.0\,\Gauss$.
 In accordance with the established models of stellar evolution, in a solar-type star the CZ covers practically all of it for a large part of the PMS.
@@
 In this point it increases significantly as a response to the generalized expansion of the star's radius.
-Regarding the effect of MB on the size of the CZ, we observe that as the intensity of the magnetic field increases, the size of the CZ decreases.
+Regarding the effect of MB on the size of the CZ, we observe that as the intensity of the magnetic field increases, the size of the CZ decreases (see \ref{fig:cz_vc_028_var_b}).
 According to \citet{Jeffries2004} this direct consequence is expected because the magnetic fields in the CZ raise the adiabatic temperature gradient and accelerate the formation of a radiative core.
-This radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base drops below $T_{Li}$.
-This effect is most evident during MS (figure (\ref{fig:cz_vc_028_var_b_z1})).
-The decrease in the size of the CZ is in line with the fact that less Li is destroyed by causing less star material to reach areas with temperatures above $T_{Li}$.\par
-
-Similar to the evolution of the $\Omega$ towards the end of the MS, we observe that the size of the CZ also shows a series of ups and downs.
-Again we think that this effect is a direct consequence of the temporal resolution used by MESA in the stellar evolution equations.
-The specific origin of these fluctuations is a subject that we will deal with in more depth and detail in subsequent works.\par
+This radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base drops below $\tli$.
+This effect is most evident during MS (see Figure~\ref{fig:cz_vc_028_var_b_z1}).
+The decrease in the size of the CZ is in line with the fact that less Li is destroyed by causing less star material to reach areas with temperatures above $\tli$.\par
 
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g.eps}
-    \caption{The evolution of convective zone size, as a function of time for several 1 $M_{\sun}$ models.
-The models include a magnetic field with an intensity of 4G and PMS rotation with $\Omega / \Omega_{crit}$ between 0.0084 and 0.0336, respectively.
-Additionally it's simulated the magnetic braking effect of magnetic field with an intensity of 4G.The dashed lines make reference to the ZAMS.}
+    \caption{The evolution of convective zone size, as a function of time for several 1 $\msun$ models.
+The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths vary between $0\,\Gauss$ and $5\,\Gauss$.
+The dashed lines make reference to the ZAMS.}
     \label{fig:cz_vc_028_var_b}
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g_z1.eps}
-    \caption{Similar to figure (\ref{fig:cz_vc_028_var_b}) but now showing in detail the convection zone size from the near ZAMS till the TAMS.
+    \caption{Similar to Figure~\ref{fig:cz_vc_028_var_b} but now showing in detail the convection zone size from the near ZAMS till the TAMS.
 As the intensity of the magnetic field increases, the size of the convective zone decreases.}
@@
 
-Before ending this section we would also like to comment on the other factor that directly affects AML; the mass loss, $\Dot{M}$.
-In figure (\ref{fig:mdot_vc_028_var_b}) we can see the evolution of $\Dot{M}$ during PMS and MS.
+Before ending this section we would also like to comment on the other factor that directly affects AML; the mass loss rate, $\Dot{M}$.
+In Figure~\ref{fig:mdot_vc_028_var_b} we can see the evolution of $\Dot{M}$ during PMS and MS for several 1 $\msun$ models.
+The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths varying between $0\,\Gauss$ and $5.0\,\Gauss$.
 In the first stages of PMS is where the greatest loss of mass is concentrated and this diminishes as it approaches the ZAMS.
@@
 When reaching the ZAMS, the radius of the star remains more or less stable for much of the MS (except for its final stage) but continues to lose mass so it also increases its $\Omega$, although in a less aggressive way if we compare it with the PMS.
-However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see figures (\ref{fig:mb_act_var_vel_vc_028}) \& (\ref{fig:grid_mb_act})), the MB routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
+However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see figures \ref{fig:mb_act_var_vel_vc_028} \& \ref{fig:grid_mb_act}), the MB routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
 The more intense the magnetic field, the greater the braking effect.\par
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g.eps}
-    \caption{The evolution of mass loss $\Dot{M}$ , as a function of time for several 1 $M_{\sun}$ models.
-The models include a variable magnetic field intensity between $0\,G$ and $5.5\,G$ and PMS rotation with $\Omega / \Omega_{crit}=0.028$.
+    \caption{The evolution of mass loss $\Dot{M}$ , as a function of time for several 1 $\msun$ models.
+The models include different magnetic field intensity between $0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.028$.
 The dashed lines make reference to the ZAMS.}
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g_z1.eps}
-    \caption{The evolution of mass loss $\Dot{M}$ as the star is approaching TAMS, as a function of time for several 1 $M_{\sun}$ models.
-The models include a variable magnetic field intensity between $0\,G$ and $5.5\,G$ and PMS rotation with $\Omega / \Omega_{crit}=0.028$.
+    \caption{The evolution of mass loss $\Dot{M}$ as the star is approaching TAMS, as a function of time for several 1 $\msun$ models.
+The models include a variable magnetic field intensity between $0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.028$.
 The stronger the magnetic field, the lower $\Dot{M}$.}
@@
     \caption{The activation of the magnetic braking routine as a function of the presence of a radiative core.
-The models include a magnetic field with an intensity ranging between $3.5\,G$ and $5.5\,G$ and PMS rotation with $\Omega / \Omega_{crit}=0.0228$.
+The models include a magnetic field with an intensity ranging between $3.5\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.0228$.
 The solid lines signal the magnetic braking routine activation (on) and deactivation (off).
@@
 \section{Conclusions and future works} \label{sec_4}
-Throughout this work we are confident that we have demonstrated that both rotation and MB induced effects must be taken into account in the theoretical models for a proper understanding and modelling of the evolution of stars.
+We have demonstrated that both rotation and MB induced effects must be taken into account in the theoretical models for a proper understanding and modelling of the evolution of stars.
 The latter is of major importance both for the transport of AM and for the transport of chemical elements.
@@
 Particularly the study of the evolution of the magnetic fields during the PMS and MS and their impact on AML.
-Future and improved implementations of our routine will be made to study the results on tracks, surface Li composition, rotation, etc...
-It is likely that in a rotating star model simulated with MB from the beginning, the differential rotation is very much reduced and therefore Li abundances observed in young stars could be properly explained.
+Future and improved implementations of our routine will be made to study the results on tracks, surface Li composition, rotation, etc.
+It is likely that in a rotating stellar model simulated with MB from the beginning, the differential rotation is very much reduced and therefore Li abundances observed in young stars could be properly explained.
 It is equally important to understand better the general role of mass loss in AML and mostly in cooler, low-mass stars.
-Nowadays it's challenging to determine the terminal velocity of the stellar wind for this type of stars which, as documented previously, plays a key role on the AML.\par
+Nowadays it is challenging to determine the terminal velocity of the stellar wind for this type of stars which, as documented previously, plays a key role on the AML.\par
 
@@
 We cannot, therefore, ascertain to have correctly modeled the rotational history of the Sun.
-In addition, we cannot (yet) properly explain the ups and downs observed in the $\Omega$ at the end of the MS.
 Therefore, in view of these shortcomings in our models, we must analyse the results obtained with caution and not draw any premature conclusions.
@@
 \begin{enumerate}
-    \item Inclusion of the magnetic field leads to cooler models and lower Li depletions in the MS.
+    \item Inclusion of the magnetic field leads to cooler models and lower Li depletion in the MS.
     \item A combination of rotation during the PMS and MB effect during MS produces different, potentially more promising behavior than those produced by standard models.
-    \item $T_{eff}$ from standard evolutionary tracks represent upper limits since these models do not take into consideration the magnetic braking effect.
+    \item $\teff$ from standard evolutionary tracks represent upper limits since these models do not take into consideration the magnetic braking effect nor rotation.
     \item The convective zone extension decreases when the intensity of the magnetic field increases.
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_3_5g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_3_0g.eps}
     \label{fig:subim2}
     \end{subfigure}
-
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_0g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_3_5g.eps}
     \label{fig:subim3}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_5g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_0g.eps}
     \label{fig:subim4}
     \end{subfigure}
-
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_5_0g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_5g.eps}
     \label{fig:subim5}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_5_5g.eps}
+    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_5_0g.eps}
     \label{fig:subim6}
     \end{subfigure}
-\caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
-Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\Omega / \Omega_{crit}$ varies between 0.0084 and 0.0336, respectively.
+\caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
+Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\oomegac$ varies between 0.0084 and 0.0336, respectively.
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
@@
     \end{subfigure}
-\caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $M_{\sun}$ models.
-Each figure shows a set of models in which $\Omega / \Omega_{crit}$ has been fixed and the magnetic field with intensity varies between $3.5\,G$ and $5.5\,G$, respectively.
+\caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
+Each figure shows a set of models in which $\oomegac$ has been fixed and the magnetic field with intensity varies between $0.0\,\Gauss$ and $5.0\,\Gauss$, respectively.
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_3_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_3_0g.eps}
     \label{fig:subim42}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_0g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_3_5g.eps}
     \label{fig:subim43}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_0g.eps}
     \label{fig:subim44}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_5_0g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_5g.eps}
     \label{fig:subim45}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_5_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_5_0g.eps}
     \label{fig:subim46}
     \end{subfigure}
-\caption{Grid showing the evolution of the evolution of surface rotational velocity, as a function of time for several 1 $M_{\sun}$ models.
-Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\Omega / \Omega_{crit}$ varies between 0.0084 and 0.0336.
+\caption{Grid showing the evolution of the evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
+Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\oomegac$ varies between 0.0084 and 0.0336.
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_3_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_3_0g.eps}
     \label{fig:subim31}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_4_0g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_3_5g.eps}
     \label{fig:subim32}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_4_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_4_0g.eps}
     \label{fig:subim33}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_5_0g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_4_5g.eps}
     \label{fig:subim34}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_5_5g.eps}
+    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_5_0g.eps}
     \label{fig:subim35}
@@
     \end{subfigure}
-\caption{Grid showing the evolution of magnetic braking activation, as a function of time and the existence of a radiative core for several 1 $M_{\sun}$ models.
-Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\Omega / \Omega_{crit}$ varies between 0.0084 and 0.0336.
+\caption{Grid showing the evolution of magnetic braking activation, as a function of time and the existence of a radiative core for several 1 $\msun$ models.
+Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\oomegac$ varies between 0.0084 and 0.0336.
 The dashed lines make reference to the ZAMS.}
@@
 Finally, we wish to acknowledge the generous work and help offered by the MESA community.
+\textbf{Authors acknowledges funding support from Spanish public funds (including FEDER fonds) for research under projects ESP2017-87676-C5-2-R and ESP2017-87676-C5-5-R.
+JCS also acknowledges support from project RYC-2012-09913 under the 'Ram\'on y Cajal' program of the Spanish Ministry of Science and Education.}
+
+\section*{Software}
+\textbf{The construction of the Depth-wise CNN was fully developed in R language version 3.6.1 \citep{R}, with RStudio 1.2.5001, the R libraries dplyr 0.8.3 \citep{dplyr2019} and ggplot2 3.2.1  \citep{ggplot2019} and TensorFlow 1.11.0 \citep{Abadi2016TensorFlow:Learning}.
+Evolutionary models were computed with \cestam\ revision 286:ce6c386387d  \citep{Marques2012}.
+Oscillation frequencies were computed with the code \filou\ v3.17.0 \citep{Suarez2008}.}
 

### 2019-11-17 project/octave/plotting_routines.m

@@ -5,6 +5,7 @@ global header_lines       = 6;
 
 %cols position in output files
 global star_age_col       = 3;
+global star_mass_col      = 5;
 global log_abs_m_dot_col  = 7;
 global mass_conv_core_col = 8;
 global log_LH_col         = 17;
@@ -26,9 +27,31 @@ global sz_bot_vrot_col    = 86;
 
 global data_parent_folder = '/home/rcaballeron/MESA/workspace/sun-jupiter-system/Docs/runs/run_paper';
 global filename = '1M_photosphere_history.data';
-global gauss_fields = ['0g'; '3.5g'; '4g'; '4.5g'; '5g'; '5.5g'];
+global gauss_fields = ['0g'; '2g'; '2.5g'; '3g'; '3.3g'; '3.5g'; '4g'; '4.3g'; '4.5g'; '5g'; '5.5g'];
+%global gauss_fields = ['0g'; '3g'; '3.5g'; '4g'; '4.5g'; '5g'];
 global rotational_vels = ['0crit';'0084crit'; '014crit'; '0196crit'; '028crit'; '0336crit';];
+global rotational_vels2 = ['na/035';'na/037'; 'na/039'; 'na/0395';'na/040'; 'na/041'; ];
 global colors = ['k'; 'r'; 'g'; 'b'; 'y'; 'm'; 'c'];
+
+%Array indexes
+global idx_0_0G  = 1;
+global idx_2_0G  = 2;
+global idx_2_5G  = 3;
+global idx_3_0G  = 4;
+global idx_3_3G  = 5;
+global idx_3_5G  = 6;
+global idx_4_0G  = 7;
+global idx_4_3G  = 8;
+global idx_4_5G  = 9;
+global idx_5_0G  = 10;
+global idx_5_5G  = 11;
+
+global idx_0crit    = 1;
+global idx_0084crit = 2;
+global idx_014crit  = 3;
+global idx_0196crit = 4;
+global idx_028crit  = 5;
+global idx_0336crit = 6;
 
 %Sun constants
 global sun_gauss_field = 1.0; %G
@@ -52,9 +75,9 @@ global line_width = 3;
 global W = 18;
 global H = 14;
 global tick_font_size = 18;
-global title_font_size = 22;
+global title_font_size = 24;
 global axis_font_size = 20;
-global legend_font_size = 20;
+global legend_font_size = 22;
 
 
 
@@ -350,6 +373,56 @@ function plot_m_dot(A, color, width, ytick, axis_limits)
 
 end
 
[…]

### 2019-11-17 project/src/run_star_extras.f

@@ -262,7 +262,7 @@
                   end do
 
                   if (debug_reset_other_torque) then
-                        !Just only who the values, don't pass them back to MESA
+                        !Just only print out the values, don't pass them back to MESA
                         s% extra_jdot(:) = 0.0
                   end if
             end if

### 2019-11-30 project/overleaf/paper.tex
Paper v3.5.0
Version sent to MNRAS

@@
 \usepackage{multicol}
-\usepackage[firstpage]{draftwatermark}
-\SetWatermarkText{DRAFT}
+%\usepackage[firstpage]{draftwatermark}
+%\SetWatermarkText{DRAFT}
 
@@
 % Keep the title short and informative.
-\title[Rotation, magnetic braking \& Li abundances]{The Sun lithium problem and how magnetic braking could help to solve it}
+\title[Rotation, magnetic braking \& Li abundances]{Study of the effects of magnetic braking on the lithium abundances of the Sun and solar-type stars}
 
@@
 Therefore, in order to make an accurate estimate of the initial abundance of Li, it is a prerequisite to be able to compare observations and models beforehand.
-Our Sun represents a unique exception, since it allows us to know the current abundance of this element in its photosphere, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, ex$ \citep{Jeffries2004}, where $A(\isotope[7]{Li})$ is defined according to Eq.~\ref{eq:A_Li}\par
+Our Sun represents a unique exception, since it allows us to know the current abundance of this element in its photosphere, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex$ \citep{Jeffries2004}, where $A(\isotope[7]{Li})$ is defined according to Eq.~\ref{eq:A_Li}\par
 
@@
 
-Making using of ($\eta_*$), the $R_{A}$ is defined as the point in which the magnetic field energy density $B^{2}/8\upi$, marked by magnetic field strength ($B$), and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$, defined by the mass density ($\rho$) and velocity of the stellar wind $(\nu)$, are balanced.
+Making use of $\eta_*$, the $R_{A}$ is defined as the point in which the magnetic field energy density $B^{2}/8\upi$, marked by magnetic field strength ($B$), and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$, defined by the mass density ($\rho$) and velocity of the stellar wind $(\nu)$, are balanced.
 That is, where the coefficient between both is equal to 1.
@@
 On the one hand, stars with similar initial masses but different mass loss ($\Dot{M}$) ratios will end up evolving very differently.
-On the other hand, the ionized particles carried by the solar wind not only contribute to the loss of mass but also to the loss of kinetic ($K_e$) energy that is deposited in the interstellar medium.
+On the other hand, the ionized particles carried by the solar wind not only contribute to the mass loss but also to the loss of kinetic ($K_e$) energy that is deposited in the interstellar medium.
 Given a star with a spherically symmetric wind, $\Dot{M}$ and $K_e$ are characterized and related by the following expressions:
@@
 In general, a magnetically channeled outflow will have a complex stream geometry but for convenience, the Eq.~\ref{eq:mass_loss} simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate.
-$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law, where $r$ represents the distance from the center of the star to the point where the velocity is to be measured, and $R_*$ is the radius of the star:
+$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law (Eq.~\ref{eq:vel_law}), where $r$ represents the distance from the center of the star to the point where the velocity is to be measured, and $R_*$ is the radius of the star:
 \begin{ceqn}
 \begin{equation}
-    \nu(r) = \nu_\infty (1-R_*/r)
+    \nu(r) = \nu_\infty (1-R_*/r) \label{eq:vel_law}
 \end{equation}
@@
 Given that the version of MESA used in this work does include neither the treatment of surface magnetic fields (it does account for transport by internal magnetic fields of AM and chemical elements) and therefore, nor for magnetic braking effects, it is necessary to extend the simulator to include this phenomenon.
-In the present work we have followed the theoretical work on MB in the Sun carried out by \citet{Weber1967}, making use of the expression \ref{eq:j_dot} for the calculation of the AML.
+In the present work we have followed the theoretical work on MB in the Sun carried out by \citet{Weber1967}, making use of Eq.~ \ref{eq:j_dot} for the calculation of the AML.
 \begin{ceqn}
@@
 \end{ceqn}
-which is a more convenient expression to be implemented in MESA because is based on values directly exposed during the simulations and convenient for the approach described in \citet{Ud-Doula2007}.
+which is a more convenient expression to be implemented in MESA because is based on values directly exposed during the simulations and suitable for the approach described in \citet{Ud-Doula2007}.
 
@@
 The numerical simulations trace the rotational history and A(Li) of a $1\, \msun$ star for a variety of initial values for $B$ and $\Omega$.
-Although stars have a three-dimensional structure and the equations that model their evolution must take this fact into account,it has been proven sufficient to solve the equations in only one dimension, extrapolating the results to the other two \citep[shellular approximation,][]{Meynet1997}.
-MESA uses this simplification to accelerate computational calculations \citet{Paxton2013}.\par
+Although stars have a three-dimensional structure and the equations that model their evolution must take this fact into account, it has been proven sufficient to solve the equations in only one dimension, extrapolating the results to the other two \citep[shellular approximation,][]{Meynet1997}.
+MESA uses this simplification to accelerate computational calculations \citep{Paxton2013}.\par
 
@@
 We define here also additional conventions and assumptions adopted in the simulations.
-For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, \mathrm{pr} = 0.0142$) not the current photospheric metallicity ($Z = Z_{\odot, \mathrm{ph} = 0.0134$) as the reference value.
+For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, \mathrm{pr} = 0.0142}$) not the current photospheric metallicity ($Z = Z_{\odot, \mathrm{ph} = 0.0134}$) as the reference value.
 We also adopt the following nominal values to express stellar properties values in SI units $\rsun = 6.957x10^{10}\, cm$ and $\msun = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
@@
 The adopted criterion is based on the simultaneous existence of an extensive convective layer and a radiative core.
-From this moment on, the MB routine will be activated acting as an additional mechanism to those existing in the MESA evolutionary code that participates in the star AML.
+From this moment on, the MB routine will be activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participates in the star AML.
 Additionally, we assume that the magnetic field does not vary its intensity throughout the evolution of the star until it reaches the TAMS.\par
 
-In this study we do not only focused in investigating the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.
+In this study we do not only focus on investigating the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.
 Keeping in mind this approach, our major objective is aimed to check the degree of validity of the resulting model after including the MB effect as a complementary mechanism of AML that can contribute to explain the evolution of Li in the Sun and other solar-type stars.
-A set of different simulation scenarios with magnetic field strengths ranging from between 3.0 and 5.0 G have been exercised.
-We'd like to stress that the magnetic field remains constant during the PMS \& MS phases.
+A set of different simulation scenarios with magnetic field strengths ranging between 3.0 and 5.0 G have been exercised.
+We would like to stress that the magnetic field remains constant during the PMS \& MS phases.
 We compute the evolution of $1\msun$ stellar models at solar metallicity with $\oomegac$ varying between $0.0084$ and $0.0336$.
@@
 
-Figure~\ref{fig:li_var_vel_0g} shows the evolution of surface Li abundance relative to H as a function of time and for several $1\msun$ models initialized with different rotational velocities.
+Figure \ref{fig:li_var_vel_0g} shows the evolution of surface Li abundance relative to H as a function of time and for several $1\msun$ models initialized with different rotational velocities.
 The simulations take into account the effects of rotation and AML caused by stellar winds but not those of MB.
-The purple star and square are surface Li abundances for the present-day Sun \citep{Asplund2009} and Pleiades \citep{Sestito2005}.\par
+The purple star and square are the surface Li abundances for the present-day Sun \citep{Asplund2009} and for the Pleiades, respectively.
+\citep{Sestito2005}.\par
 
@@
 On the other hand, the differential rotation between the core and the convective zone does not begin to develop until it reaches $ \approx 10^6$ years, achieving its maximum difference in the ZAMS about $ \approx 10^7$ years.
-It is at this moment that the combined effects of turbulence and rotational difference should increase, leading to notable differences in the evolution of A(Li) (see Figure~\ref{fig:li_var_vel_0g_z1}).
+It is at this moment that the combined effects of turbulence and rotational difference should increase, leading to notable differences in the evolution of A(Li) (see Figure \ref{fig:li_var_vel_0g_z1}).
 Later, the reference model does not deplete Li efficiently on the MS and fail (again) to match the current solar surface Li abundance.
-On the other hand, the other models which include rotation during the PMS with values of $\oomegac$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but its rotational velocity is much higher (see Figure~\ref{fig:rot_vel_0g}) than the $2\,\kms$ of the Sun \citep{Gill2012}.
+On the other hand, the other models which include rotation during the PMS with values of $\oomegac$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but its rotational velocity is much higher (see Figure \ref{fig:rot_vel_0g}) than the $2\,\kms$ of the Sun \citep{Gill2012}.
 \par
 
-For much of the PMS the star rotates as a solid body (see Figure~\ref{fig:rot_vel_0g}) and this is because the star has a completely convective interior.
+For much of the PMS the star rotates as a solid body (see Figure \ref{fig:rot_vel_0g}) and this is because the star has a completely convective interior.
 It is not until the end of Hayashi track that the star begins to develop a radiative core due to the increase in temperature that occurs inside the star during the contraction process.
@@
 As the models are initialized with a higher angular velocity, the difference in velocity between the BCZ and the star surface is accentuated in a directly proportional mode; the higher the initial velocity, the bigger the velocity gradient between the bottom and top limits of the CZ.
-As a consequence, the turbulence strength located at the BCZ increases so that Li can reach regions with temperatures about $\tli$ where it is finally burned and destroyed (see figure~ \ref{fig:li_var_vel_0g}).
+As a consequence, the turbulence strength located at the BCZ increases so that Li can reach regions with temperatures about $\tli$ where it is finally burned and destroyed (see Figure  \ref{fig:li_var_vel_0g}).
 Other investigations \citep{Bouvier2018, Baraffe2017}  point to a tendency diametrically opposed to the one exposed here, i.e.
@@
 	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
-    \caption {Similar to Figure~\ref{fig:li_var_vel_0g} but zooming in on the ZAMS.
+    \caption {Similar to Figure \ref{fig:li_var_vel_0g} but zooming in on the ZAMS.
 The models with a higher initial rotational velocity reach already the ZAMS with a lower amount of Li measure on the star surface.}
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
-    \caption{The evolution of angular velocity at surface (solid line) and the top (dashed line) and bottom (dotted line) limits of the uppermost convective zone, as a function of time for several 1 $\msun$ models.
+    \caption{The evolution of angular velocity at surface (solid line) and at the bottom (dotted line) limit of the uppermost convective zone, as a function of time for several 1 $\msun$ models.
 The models includes PMS rotation with $\oomegac$ values between 0.0084 and 0.0336.
@@
 Another well known structural effect of rotation is the decrease of the effective temperature ($\teff$) and to a less extent of stellar luminosity ($L$).
-Both effects can be observed graphically in the HR diagram of Figure~\ref{fig:hr_var_vel_0g} which shows a zoomed-in view of evolutionary tracks from the ZAMS until the TAMS for several $1\msun$ models initialized with different rotational velocities.
+Both effects can be observed graphically in the HR diagram of Figure \ref{fig:hr_var_vel_0g} which shows a zoomed-in view of evolutionary tracks from the ZAMS until the TAMS for several $1\msun$ models initialized with different rotational velocities.
 If we compare the non-rotating model (black solid line on the left side) with the rotating ones we can recognize that at the end of the PMS, the latter reach before the ZAMS and exhibit a lower $\teff$ than the former.
@@
 Those models were initialized with different rotational velocities and took into consideration both the effects of AML and the MB caused by a magnetic field of intensity 4G.
-If we compare it with Figure~\ref{fig:li_var_vel_0g} in which the effects of MB are neglected, we notice how the profiles of Li abundance are altered during PMS and MS.
-In the first phase we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see Eq.~\ref{eq:j_dot}) depends directly on the mass loss rate.
-If we take into account that for solar-type stars the models predict a modest total mass loss rate, that value is even much lower during PMS.
-On the contrary, during the MS it is observed that the AML is much more significant, causing the star to rotate more slowly and as a result a smaller amount of Li to be destroyed.\par
-
-If we concentrate our attention on Figure~\ref{fig:li_var_vel_4_0g} which represents the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models initialized with different rotational velocities and considers the effects of AML and the MB caused by a magnetic field of intensity 4G, and compare it with Figure~\ref{fig:li_var_vel_0g} in which the effects of MB are neglected, we will see how the profiles of Li abundance are altered during PMS and MS.
+If we compare it with Figure \ref{fig:li_var_vel_0g} in which the effects of MB are neglected, we notice how the profiles of Li abundance are altered during PMS and MS.
 In the first phase we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see Eq.~\ref{eq:j_dot}) depends directly on the mass loss rate.
@@
 
-The effect of the MB routine can be seen even more clearly in figures \ref{fig:rot_vel_4g}, \ref{fig:rot_vel_4g_z1} \& \ref{fig:grid_rot_vel}.
-In those figures we represent the rotation profiles for the surface of the stars and for the limits of the convective envelope for a number of 1 $\msun$ models initialized with different rotational velocities and considering the influence of MB.
+The effect of the MB routine can be seen even more clearly in Figures~\ref{fig:rot_vel_4g}, \ref{fig:rot_vel_4g_z1} \& Appendixes \footnote{The appendices comprise a series of grids as a function of time and for several 1 $\msun$ models in which on the one hand, the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H} for both variable magnetic field intensities and angular velocities and on the other hand, the evolution of surface rotational velocity are shown.}.
+In those figures we represent the rotation profiles for the surface of the stars and for the bottom of the convective envelope for a number of 1 $\msun$ models initialized with different rotational velocities and considering the influence of MB.
 Similarly to the evolution profiles of the Li on the surface commented in the previous paragraph, the effect of the routine is visible once the ZAMS is reached.
-If we compare the evolution of the curves presented here with those of Figure~\ref{fig:rot_vel_0g} we see how the star, instead of continuing to increase its $\Omega$, begins to slow down after having reached its maximum in the passage through the ZAMS.
+If we compare the evolution of the curves presented here with those of Figure \ref{fig:rot_vel_0g} we see how the star, instead of continuing to increase its $\Omega$, begins to slow down after having reached its maximum in the passage through the ZAMS.
 It is also at this stage when the angular velocities at the surface of the star and at the BCZ reach their maximum difference.
-On the other hand, the effect of magnetic braking causes the angular velocities between both zones of the star to decrease until, for an age close to that of the Sun (see Figure~\ref{fig:rot_vel_4g_z1}), the star practically rotates like a rigid solid.
+On the other hand, the effect of magnetic braking causes the angular velocities between both zones of the star to decrease until, for an age close to that of the Sun (see Figure \ref{fig:rot_vel_4g_z1}), the star practically rotates like a rigid solid.
 These results are also consistent with those obtained by \citet{Eggenberger2010} as far as the effect of the magnetic field, in particular its influence on the loss of angular momentum, has on the rotational velocity of the star.
@@
 
-In a similar way we also observe that the models with lower angular rotation generally end up exhibiting higher values in the abundance of Li on the surface (see figures \ref{fig:li_var_vel_4_0g}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).
+In a similar way we also observe that the models with lower angular velocity generally end up exhibiting higher values in the abundance of Li on the surface (see Figures~ \ref{fig:li_var_vel_4_0g}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).
 In none of those cases we do obtain values of Li on the surface higher than those shown by the model without rotation.\par
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g_z1.eps}
-    \caption{Similar to Figure~\ref{fig:rot_vel_4g} but now showing in detail the surface rotational velocity as the star approaches TAMS.}
+    \caption{Similar to Figure \ref{fig:rot_vel_4g} but now showing in detail the surface rotational velocity as the star approaches TAMS.}
     \label{fig:rot_vel_4g_z1}
@@
 
-It is also remarkable to observe that, for some of the models depicted in Figure~\ref{fig:rot_vel_4g}, their $\Omega$ at the star surface (solid line) is slower than the one of the upper limit of their convective layer (dashed line).
-This fact is most probably a consequence of the expansion of the superficial layers after the mass loss produced by stellar winds.\par
-
 The MB also leaves its mark on the HR diagram by significantly affecting the $\teff$ of the star.
-To visualize this effect we will take as reference Figure~\ref{hr_vc_0336_var_g_z1} in which all models were started with the same value $\oomegac=0.0336$ but the intensity of the simulated magnetic field was different.
+To visualize this effect we will take as reference Figure \ref{hr_vc_0336_var_g_z1} in which all models were started with the same value $\oomegac=0.0336$ but the intensity of the simulated magnetic field was different.
 In particular, let us pay attention to the evolutionary trace in which a magnetic field was not simulated (black solid line) and to the rest of the evolutionary traces in which it was.
 We have that the group of simulations with presence of a magnetic field produces hotter stars due to the MB influence.
-The lower speed with which the star rotates due to the MB effect causes the increase of the $\teff$, being this difference of practically $95\,\Kelvin$ between the simulated models with $0\,\Gauss$ and $5.0\,\Gauss$ respectively for $log(\llsun)=0$.
+The lower speed with which the star rotates due to the MB effect causes the increase of the $\teff$, being this difference of practically $95\,\Kelvin$ between the simulated models with $0.0\,\Gauss$ and $5.0\,\Gauss$ respectively for $log(\llsun)=0$.
 
@@
 	\includegraphics[trim = 30mm 15mm 15mm 15mm, clip,width=\columnwidth]{figures/hr_vc_0336_var_g_z1.eps}
-    \caption{Similar to Figure~\ref{fig:hr_var_vel_0g} but now showing in detail the effects of magnetic braking on the evolutionary tracks for different magnetic field strengths and $\oomegac=0.0336$.
+    \caption{Similar to Figure \ref{fig:hr_var_vel_0g} but now showing in detail the effects of magnetic braking on the evolutionary tracks for different magnetic field strengths and $\oomegac=0.0336$.
 The presence of a magnetic field produces hotter stars due to the influence of the magnetic braking on the rotational velocity of the star.}
@@
 
-As described in section \ref{sec_mesh}, the MB routine distributes the total amount of AML calculated according to Eq.~\ref{eq:k_jdot} among the different layers that compose it.
-In Figure~\ref{fig:cz_vc_028_var_b} we can observe the evolution of the most external CZ normalized with respect to the radius of the star for several 1 $\msun$ models.
-The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths varying between $0\,\Gauss$ and $5.0\,\Gauss$.
+As described in section \ref{sec_mesh}, the MB routine distributes the total amount of AML calculated according to Eq.~\ref{eq:k_jdot} among the different layers that compose the CZ.
+In Figure \ref{fig:cz_vc_028_var_b} we can observe the evolution of the most external CZ normalized with respect to the radius of the star for several 1 $\msun$ models.
+The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths varying between $0.0\,\Gauss$ and $5.0\,\Gauss$.
 In accordance with the established models of stellar evolution, in a solar-type star the CZ covers practically all of it for a large part of the PMS.
@@
 In this point it increases significantly as a response to the generalized expansion of the star's radius.
-Regarding the effect of MB on the size of the CZ, we observe that as the intensity of the magnetic field increases, the size of the CZ decreases (see \ref{fig:cz_vc_028_var_b}).
-According to \citet{Jeffries2004} this direct consequence is expected because the magnetic fields in the CZ raise the adiabatic temperature gradient and accelerate the formation of a radiative core.
-This radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base drops below $\tli$.
-This effect is most evident during MS (see Figure~\ref{fig:cz_vc_028_var_b_z1}).
+Regarding the effect of MB on the size of the CZ, we observe that as the intensity of the magnetic field increases, the size of the CZ decreases (see Figure \ref{fig:cz_vc_028_var_b}).
+The radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base drops below $\tli$.
+This effect is most evident during MS (see Figure \ref{fig:cz_vc_028_var_b_z1}).
 The decrease in the size of the CZ is in line with the fact that less Li is destroyed by causing less star material to reach areas with temperatures above $\tli$.\par
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g.eps}
-    \caption{The evolution of convective zone size, as a function of time for several 1 $\msun$ models.
-The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths vary between $0\,\Gauss$ and $5\,\Gauss$.
+    \caption{The evolution of convective zone size as a function of time for several 1 $\msun$ models.
+The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths vary between $0.0\,\Gauss$ and $5.0\,\Gauss$.
 The dashed lines make reference to the ZAMS.}
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g_z1.eps}
-    \caption{Similar to Figure~\ref{fig:cz_vc_028_var_b} but now showing in detail the convection zone size from the near ZAMS till the TAMS.
+    \caption{Similar to Figure \ref{fig:cz_vc_028_var_b} but now showing in detail the convection zone size from the near ZAMS till the TAMS.
 As the intensity of the magnetic field increases, the size of the convective zone decreases.}
@@
 
-Before ending this section we would also like to comment on the other factor that directly affects AML; the mass loss rate, $\Dot{M}$.
-In Figure~\ref{fig:mdot_vc_028_var_b} we can see the evolution of $\Dot{M}$ during PMS and MS for several 1 $\msun$ models.
-The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths varying between $0\,\Gauss$ and $5.0\,\Gauss$.
-In the first stages of PMS is where the greatest loss of mass is concentrated and this diminishes as it approaches the ZAMS.
+In Figure \ref{fig:mdot_vc_028_var_b} we can see the evolution of $\Dot{M}$ during PMS and MS for several 1 $\msun$ models.
+The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths varying between $0.0\,\Gauss$ and $5.0\,\Gauss$.
+In the first stages of PMS is where the greatest mass loss is concentrated and this diminishes as it approaches the ZAMS.
 If we add to this scenario the fact that the star also decreases its radius during the PMS we have as result that the star increases $\Omega$ obeying the principle of conservation of AM.
 When reaching the ZAMS, the radius of the star remains more or less stable for much of the MS (except for its final stage) but continues to lose mass so it also increases its $\Omega$, although in a less aggressive way if we compare it with the PMS.
-However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see figures \ref{fig:mb_act_var_vel_vc_028} \& \ref{fig:grid_mb_act}), the MB routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
+However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see Figure \ref{fig:mb_act_var_vel_vc_028}), the MB routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
 The more intense the magnetic field, the greater the braking effect.\par
@@
 	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g.eps}
-    \caption{The evolution of mass loss $\Dot{M}$ , as a function of time for several 1 $\msun$ models.
-The models include different magnetic field intensity between $0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.028$.
+    \caption{The evolution of mass loss $\Dot{M}$ as a function of time for several 1 $\msun$ models.
+The models include different magnetic field intensity between $0.0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.028$.
 The dashed lines make reference to the ZAMS.}
@@
     \caption{The evolution of mass loss $\Dot{M}$ as the star is approaching TAMS, as a function of time for several 1 $\msun$ models.
-The models include a variable magnetic field intensity between $0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.028$.
+The models include a variable magnetic field intensity between $0.0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.028$.
 The stronger the magnetic field, the lower $\Dot{M}$.}
@@
     \caption{The activation of the magnetic braking routine as a function of the presence of a radiative core.
-The models include a magnetic field with an intensity ranging between $3.5\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.0228$.
+The models include a magnetic field with an intensity ranging between $3.0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.0228$.
 The solid lines signal the magnetic braking routine activation (on) and deactivation (off).
@@
 \end{figure}
+
+\subsection{Alternative Li evolution with MB}
+Up to this point, the simulations of the different models have been based on the parameterization gathered in table \ref{tab:phy_mesa}, which in turn has been adopted from \citet{Choi2016}.
+If we recall the evolution of the Li shown in Figures~\ref{fig:li_var_vel_0g} and \ref{fig:li_var_vel_4_0g}, we highlighted that on both cases too much Li was burned before reaching the ZAMS and therefore do not match with the Pleiades average surface Li abundance.
+On the other than, using a parameterization slightly different from the one used until now in which the parameters of convection and overshooting have been readjusted according to table \ref{tab:phy_alt_mesa}, the simulations can reproduce more faithfully both the Li abundance of the Pleiades cluster and of the Sun (see Figure \ref{fig:li_3_0g_0314vc}).\par
+
+\begin{table}
+	\centering
+	\caption{Alternative MTL and overshooting parameters.}
+	\label{tab:phy_alt_mesa}
+	\begin{tabular}{ll} 
+		\hline
+		Parameter & Adopted prescriptions and values\\
+		\hline
+		Convection & MLT + Ledoux, $\alpha_\mathrm{MLT}$ = 1.70\\
+		Overshoot & time-dependent, diffusive, $f_\mathrm{ov,core}=0.016$, \\ & $f_\mathrm{ov,sh}=0.002$\\
+		\hline
+	\end{tabular}
+\end{table}
+
+\begin{figure}
+	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_3_0g_0314vc.eps}
+    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for a 1 $\msun$ model.
+The model include a magnetic field with an intensity of 3\,G and PMS rotation with $\oomegac= 0.0314$, respectively.
+MLT and overshooting parameters have been set to $\alpha_\mathrm{MLT}=1.70$, $f_\mathrm{ov,core}=0.016$, $f_\mathrm{ov,sh}=0.002$.
+The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
+The dashed lines make reference to the ZAMS.}
+    \label{fig:li_3_0g_0314vc}
+\end{figure}
+
+Similarly, the evolution of the angular velocity for this new configuration is shown in Figure \ref{fig:rot_vel_var_vel_mlt_3_0g}.
+Here, on the contrary, we have to note that with the new simulation we still do not reproduce correctly the angular velocity of the Sun.
+Having said this and based on the results obtained in this last section, it becomes evident that the influence of the free, relatively arbitrary, parameters associated with Mixing Lenght Theory (MLT) significantly condition the evolution of the abundance of Li.
+This opens the door for other solar-compatible parameterizations to offer results in line with the observations in young stellar clusters and for the Sun.
+\par
+
+\begin{figure}
+	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_3_0g_0314vc.eps}
+    \caption{The evolution of surface rotational velocity, as a function of time for a 1 $\msun$ model.
+The model include a magnetic field with an intensity of 3\,G, PMS rotation with $\oomegac=0.0314$, respectively and MB.
+MLT and overshooting parameters have been set to $\alpha_\mathrm{MLT}=1.70$, $f_\mathrm{ov,core}=0.016$, $f_\mathrm{ov,sh}=0.002$.
+The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
+The dashed lines make reference to the ZAMS.}
+    \label{fig:rot_vel_var_vel_mlt_3_0g}
+\end{figure}
+
+We propose that this adjustment could also potentially be achieved by a law of interdependence between $\Omega$ and $B$.
+So that, during the PMS, when the star rotates faster, the MB effect is more efficient.
+On the other hand, during the MS, when the start slows down, the MB will become less intensive.
+This approach would establish a self-regulating mechanism over the angular velocity of the star that would end up directly influencing the evolution of A(Li).
+This is the line of work in which we will continue our research.\par
 
@@
 Therefore, in view of these shortcomings in our models, we must analyse the results obtained with caution and not draw any premature conclusions.
-Having said that, we are convinced that our research is on the right track and we do believe that the following conclusions, based on the results described in previous sections, have been sufficiently supported in the course of this work:
+Having said that, we have also shown how an alternative MLT parametrizations could produce results in line with the observations.
+We are convinced that our research is on the right track and we do believe that the following conclusions, based on the results described in previous sections, have been sufficiently supported in the course of this work:
 \begin{enumerate}
@@
     \item The convective zone extension decreases when the intensity of the magnetic field increases.
-    \item MB during PMS seems to be also required for explaining Li abundances in young clusters.
+    \item MB during PMS and/or adjustment of MLT overshooting and $\amlt$ free parameters seems to be also required for explaining Li abundances in young clusters.
 \end{enumerate}
 
-In the near future, with the new version of our code including magnetic field evolution in MESA, we will focus our efforts on understanding how magnetic fields are linked to rotation, how their topologies are and how they evolves in time in order to go beyond the treatment on the MB routine that has been done in this work.\par
+In the near future, with the new version of our code including magnetic field evolution in MESA, we will focus our efforts on understanding how magnetic fields are linked to rotation, how their topologies are and how they evolve in time in order to go beyond the treatment on the MB routine that has been done in this work.\par
+
+\section*{Acknowledgements}
+We are pleased to acknowledge Jieun Choi her kindness in responding to the different questions about her work which were aimed to reproduce the results obtained as far as Li is concerned, and also for allowing us access to the MESA files she used.
+Also we very much appreciate the expert support provided by Elisa Delgado-Mena during the process review of this document.
+Similarly, we wish also to thank Matteo Cantiello and Bill Paxton for their very valuable support in the development of the magnetic braking routine in MESA.
+Finally, we wish to acknowledge the generous work and help offered by the MESA community.
+Authors acknowledges funding support from Spanish public funds (including FEDER fonds) for research under projects ESP2017-87676-C5-2-R and ESP2017-87676-C5-5-R.
+JCS also acknowledges support from project RYC-2012-09913 under the 'Ram\'on y Cajal' program of the Spanish Ministry of Science and Education.
+
+\section*{Software}
+The simulations used in this paper been executed with the MESA release 10398 \footnote{\url{http://mesa.sourceforge.net/release/2018/03/21/r10398.html}}.
+The different figures have been generated making use of GNU Octave 5.1.0.
+\footnote{\url{https://www.gnu.org/software/octave/}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%% REFERENCES %%%%%%%%%%%%%%%%%%
+
+% The best way to enter references is to use BibTeX:
+
+\bibliographystyle{mnras}
+%\bibliography{magbrlitium} % if your bibtex file is called example.bib
+\bibliography{mblithium}
+
+% Alternatively you could enter them by hand, like this:
+% This method is tedious and prone to error if you have lots of references
+%\begin{thebibliography}{99}
+%\bibitem[\protect\citeauthoryear{Author}{2012}]{Author2012}
+%Author A.~N., 2013, Journal of Improbable Astronomy, 1, 1
+%\bibitem[\protect\citeauthoryear{Others}{2013}]{Others2013}
+%Others S., 2012, Journal of Interesting Stuff, 17, 198
+%\end{thebibliography}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%
+
+\appendix
+
+\section{Mesh models visualization}
+The following figures comprise a series of grids as a function of time and for several 1 $\msun$ models in which on the one hand, the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H} for both variable magnetic field intensities and angular velocities and on the other hand, the evolution of surface rotational velocity are shown.
 
@@
     \end{subfigure}
-\caption{Grid showing the evolution of the evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
+\caption{Grid showing of the evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
 Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\oomegac$ varies between 0.0084 and 0.0336.
@@
 \end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_3_0g.eps}
-    \label{fig:subim31}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_3_5g.eps}
-    \label{fig:subim32}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_4_0g.eps}
-    \label{fig:subim33}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_4_5g.eps}
-    \label{fig:subim34}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/mb_act_var_vel_5_0g.eps}
-    \label{fig:subim35}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/blank.eps}
-    \label{fig:subim36}
-    \end{subfigure}
-\caption{Grid showing the evolution of magnetic braking activation, as a function of time and the existence of a radiative core for several 1 $\msun$ models.
-Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\oomegac$ varies between 0.0084 and 0.0336.
-The dashed lines make reference to the ZAMS.}
-\label{fig:grid_mb_act}
-\end{figure*}
-
-\section*{Acknowledgements}
-We are pleased to acknowledge Jieun Choi her kindness in responding to the different questions about her work which were aimed to reproduce the results obtained as far as Li is concerned, and also for allowing us access to the MESA files she used.
-Also we very much appreciate the expert support provided by Elisa Delgado-Mena during the process review of this document.
-Similarly, we wish also to thank Matteo Cantiello and Bill Paxton for their very valuable support in the development of the magnetic braking routine in MESA.
-Finally, we wish to acknowledge the generous work and help offered by the MESA community.
-\textbf{Authors acknowledges funding support from Spanish public funds (including FEDER fonds) for research under projects ESP2017-87676-C5-2-R and ESP2017-87676-C5-5-R.
-JCS also acknowledges support from project RYC-2012-09913 under the 'Ram\'on y Cajal' program of the Spanish Ministry of Science and Education.}
-
-\section*{Software}
-\textbf{The construction of the Depth-wise CNN was fully developed in R language version 3.6.1 \citep{R}, with RStudio 1.2.5001, the R libraries dplyr 0.8.3 \citep{dplyr2019} and ggplot2 3.2.1  \citep{ggplot2019} and TensorFlow 1.11.0 \citep{Abadi2016TensorFlow:Learning}.
-Evolutionary models were computed with \cestam\ revision 286:ce6c386387d  \citep{Marques2012}.
-Oscillation frequencies were computed with the code \filou\ v3.17.0 \citep{Suarez2008}.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%% REFERENCES %%%%%%%%%%%%%%%%%%
-
-% The best way to enter references is to use BibTeX:
-
-\bibliographystyle{mnras}
-%\bibliography{magbrlitium} % if your bibtex file is called example.bib
-\bibliography{mblithium}
-
-% Alternatively you could enter them by hand, like this:
-% This method is tedious and prone to error if you have lots of references
-%\begin{thebibliography}{99}
-%\bibitem[\protect\citeauthoryear{Author}{2012}]{Author2012}
-%Author A.~N., 2013, Journal of Improbable Astronomy, 1, 1
-%\bibitem[\protect\citeauthoryear{Others}{2013}]{Others2013}
-%Others S., 2012, Journal of Interesting Stuff, 17, 198
-%\end{thebibliography}
 

### 2019-11-30 project/octave/plotting_routines.m

@@ -29,8 +29,7 @@ global data_parent_folder = '/home/rcaballeron/MESA/workspace/sun-jupiter-system
 global filename = '1M_photosphere_history.data';
 global gauss_fields = ['0g'; '2g'; '2.5g'; '3g'; '3.3g'; '3.5g'; '4g'; '4.3g'; '4.5g'; '5g'; '5.5g'];
 %global gauss_fields = ['0g'; '3g'; '3.5g'; '4g'; '4.5g'; '5g'];
-global rotational_vels = ['0crit';'0084crit'; '014crit'; '0196crit'; '028crit'; '0336crit';];
-global rotational_vels2 = ['na/035';'na/037'; 'na/039'; 'na/0395';'na/040'; 'na/041'; ];
+global rotational_vels = ['0crit';'0084crit'; '014crit'; '0196crit'; '028crit'; '0336crit';'029crit';'030crit';'031crit';'0312crit';'0314crit';'032crit';];
 global colors = ['k'; 'r'; 'g'; 'b'; 'y'; 'm'; 'c'];
 
 %Array indexes
@@ -46,12 +45,21 @@ global idx_4_5G  = 9;
 global idx_5_0G  = 10;
 global idx_5_5G  = 11;
 
+%The following are with mlt=1.82
 global idx_0crit    = 1;
 global idx_0084crit = 2;
 global idx_014crit  = 3;
 global idx_0196crit = 4;
 global idx_028crit  = 5;
 global idx_0336crit = 6;
+%The following are with mlt=1.7
+global idx_029crit  = 7;
+global idx_030crit  = 8;
+global idx_031crit  = 9;
+global idx_0312crit  = 10;
+global idx_0314crit  = 11;
+global idx_032crit  = 12;
+
 
 %Sun constants
 global sun_gauss_field = 1.0; %G
@@ -303,17 +311,22 @@ function plot_vel_rot(A, color, width, ytick, axis_limits)
   
   %Plot values
   plot(A(:,1), A(:,2), color, 'linewidth', width);
-  plot(A(:,1), A(:,3), color, 'linewidth', width, 'linestyle', '--');
+  %Don't show sup limit of CZ
+  %plot(A(:,1), A(:,3), color, 'linewidth', width, 'linestyle', '--');
   plot(A(:,1), A(:,4), color, 'linewidth', width, 'linestyle', ':');
 
   %Axis scales
   set(gca, 'XScale', 'log');  
[…]

### 2020-05-17 project/overleaf/paper.tex
MNRAS revision 2

@@
 \usepackage{multicol}
+\usepackage[normalem]{ulem}
 %\usepackage[firstpage]{draftwatermark}
@@
 \textit{Goals.} The study of lithium (Li) surface abundance in the Sun and young stellar globular clusters which are seemingly anomalous in present-day scenarios, as well as the influence of rotation and magnetic braking (MB) on its depletion during pre-main sequence (PMS) and main sequence (MS).
-\newline\textit{Methods.} The effects of rotational mixing and of the rotational hydrostatic effects on Li abundances are studied by simulating several grids of PMS \& MS rotating and non-rotating models.
-The previous effects are combined with the additional impact of the MB (with magnetic field intensities ranging between 3.0 and 5.0 G) and the data obtained from the two sets of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
+\newline\textit{Methods.} The effects of rotational mixing and of the rotational hydrostatic effects on Li abundances are studied by simulating several grids of PMS and MS rotating and non-rotating models.
+The previous effects are combined with the additional impact of the MB, (with magnetic field intensities ranging between 3.0 and 5.0 G).
+The data obtained from the two sets of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
 \newline\textit{Results.} The surface Li abundance for the Sun like models at the end of the PMS and throughout the MS is decreased when rotational effects are included, i.e.
@@
 This physical phenomena, generally neglected in stellar evolutionary models, impacts also the effective temperature ($\teff$) of the star and therefore its location in the HR diagram.
-The impact of MB in Li depletion is sensitive to the magnetic field intensity: the higher it is, the lower the Li destruction.
+The impact of MB in Li depletion is sensitive to the angular speed and this in turn on the magnetic field intensity: the higher the latter, the lower the Li destruction.
 A direct link between the magnetic fields and the convective zone (CZ) size is observed: stronger magnetic fields produce shallower CZ's.
-Another important result of this study is that MB effect must be taken into consideration during PMS if we aim to reproduce Li abundances in young clusters.
+This result suggests that MB effect must be taken into consideration during PMS if we aim to reproduce Li abundances in young clusters.
 \end{abstract}
@@
 
-\section{Introduction} \label{sec_1}
-Despite decades of theoretical efforts, a coherent explanation derived from models cannot be found for the Li abundance discrepancies detected in stars belonging to clusters of different ages and which are in the same evolutionary state, either in the PMS or in the MS.
+\section{Introduction} \label{sec_intro}
+Despite decades of theoretical efforts, a coherent explanation for the Li abundance discrepancies detected in star clusters cannot be found.
+In young and old clusters these differences have been documented for stars in both PMS and MS  evolutionary stages.
 Additionally, these theoretical models are not able to explain the abundances detected in the late stages of the MS \citep{Tschape2001}.\par
 
-The denominated standard models, models that include convection only as a mixing process and do not consider other transport options like diffusion or angular momentum loss (AML), have been mainly involved in the elaboration of these predictions \citep{Sestito2005}.
-Li is destroyed at a temperature $\tli \approx 2.5 x 10^6\; K$ when a Li atom collides with a proton producing two He atoms, something that takes place during proton-proton type II reactions (P-P II), and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value.
+Standard models, models that include convection only as a mixing process and do not consider other transport options like diffusion or angular momentum loss (AML), have been mainly involved in the elaboration of these predictions \citep{Sestito2005}.
+Li is destroyed at a temperature $\tli \approx 2.5 x 10^6\; K$ when a Li atom collides with a proton producing two He atoms.
+That takes place during proton-proton type II reactions (P-P II), and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches  $\tli$.
 The Sun in particular and the solar-type stars in general are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $\tli$ \citep{Iben1965}.
 This depletion stops during the approach to the zero-age MS (ZAMS) when the convection zone retreats and the temperature at BCZ is cooler than $\tli$.
-In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $\tli$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances.
+In the standard models only the mass and the initial chemical composition determine at what distance from the stellar surface $\tli$ temperature is reached.
+Therefore stars with similar mass belonging clusters are expected to reach the ZAMS with equal surface Li abundances.
 Furthermore, they should also show a very similar Li evolution until their approach to the terminal-age MS (TAMS).
-During this same period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the convective envelope, from its lower limit till the star surface.
-However, different abundances of Li have been observed for different star populations \citep[][and references therein]{Somers2014}.\par
+During this period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the envelope, from its lower limit till the star surface.
+However, different abundances of Li have been observed for different stellar populations \citep[see][and references therein]{Somers2014}.\par
 
-Theoretical models of stars do not inform about the initial amount of Li that a star has, they only describe how it is exhausted.
+Theoretical models take initial Li abundance as an input parameter and only describe how it is exhausted.
 Therefore, in order to make an accurate estimate of the initial abundance of Li, it is a prerequisite to be able to compare observations and models beforehand.
-Our Sun represents a unique exception, since it allows us to know the current abundance of this element in its photosphere, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex$ \citep{Jeffries2004}, where $A(\isotope[7]{Li})$ is defined according to Eq.~\ref{eq:A_Li}\par
+Our Sun represents a unique exception, since it allows us to know the current abundance of this element in its photosphere, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex$ \citep{Jeffries2004}, where $A(\isotope[7]{Li})$ is defined
 
@@
 
-On the other hand, the initial abundance of $A(\isotope[7]{Li}) = 3.34 \, dex$ is obtained from meteorite measurements \citep{Randich2006}.
-According to the theory for newly born stars, we have that the initial abundance of Li can be estimated fairly accurately from photospheric measurements on T-Tauri type stars, or from the hottest F stars forming part of slightly older clusters.
-For the latter, the current theory of stellar evolution suggests that Li must not yet be exhausted.
-The results obtained from the measurements on both types of stars allow us to fix the initial abundance of Li in the interval $3.0 \, dex < A(\isotope[7]{Li}) < 3.4 \, dex$ \citep{Randich2006}.\par
+and its initial abundance of $A(\isotope[7]{Li}) = 3.34 \, dex$ is obtained from meteorite measurements \citep{Randich2006}.
+For other stars the initial Li abundance is found to be in the interval $3.0 \, dex < A(\isotope[7]{Li}) < 3.4 \, dex$.
+This has been estimated from  \citet{Randich2006}.\par
 
 The impact of rotation both on PMS and Li depletion for solar-type stars has already extensively debated in the past \citep{Pinsonneault1997,Jeffries2004,Somers2014} and revised more recently on the basis of the availability of more accurate measures \citep{Bouvier2016}.
-However, these previous studies focused mainly on the hydrostatic effects and did not consider the influence of a coupling between the star magnetic field and its possible spin-down effect.
-It is thought that AML has a direct influence in the mixing processes and that transport of the momentum could be produced by a number of relevant mechanisms: mass loss, magnetic fields and gravity waves (g-modes).
-Concerning the two latter, gravity waves \citep{Charbonnel2005} and magnetic fields \citep{Eggenberger2009} have the property of transmitting angular momentum (AM) much more effectively than inducing mixing \citep{Denissenkov2007}.
-As a consequence of this increment in the AM transport efficiency, the amount of differential rotation between the radiative and the convective zones of the star is reduced (a solid-body rotation is fostered), as are the induced rotational instabilities.
-Additionally, magnetic fields would also originate in the limits between these zones, in the so-called tachoclines \citep{Aschwanden2014, Guerrero2016}, which interact with the solar wind particles that are forced to rotate in sync with it and therefore slow down the star, i.e.
-the magnetic braking (MB) effect.
+However, these previous studies focused mainly on the hydrostatic effects and did not consider the influence of a coupling between the stellar magnetic field and its possible spin-down effect.
+AML has a direct influence in the mixing processes.
+It can be produced by a number of relevant mechanisms: mass loss, magnetic fields and gravity waves (g-modes).
+Gravity waves \citep{Charbonnel2005} and magnetic fields \citep{Eggenberger2009} transmit angular momentum (AM) much more effectively than inducing mixing \citep{Denissenkov2007}.
+As a consequence, the amount of differential rotation between the radiative and the convective zones of the star is reduced.
+Additionally, magnetic fields would also originate in the limits between these zones, in the so-called tachoclines \citep{Aschwanden2014, Guerrero2016}, which interact with the solar wind particles that are forced to co-rotate with it.
+This slows down the star, effect which is known as magnetic braking (MB).
 \par
 
-Today, standard models are not able to replicate in a satisfactory way the observed values of Li abundance on star surfaces with the predictions that these models yield.
-This lack of agreement makes us think that certain physical mechanisms that influence the destruction of Li are either being modelled improperly or simply not being taken into account, e.g.
+Today, standard models are not able to replicate the observed values of Li abundance on stellar surfaces.
+This might happen because certain physical mechanisms that influence the destruction of Li are either being modelled improperly or simply not being taken into account, e.g.
 the magnetic braking.
-In view of these findings, it seems evident that there are still open questions to be resolved on the subject of the processes that participate, directly or indirectly, in the mixing mechanisms that are not incorporated in the standard stellar models.
-Therefore, it is essential a proper consideration of the interactions between rotation and magnetic fields when it comes to study the AM distribution.
+In this paper we study the interactions between rotation and magnetic fields when it comes to discuss the AM distribution.
 \par
 
-\subsection{Effects of magnetic fields}
-The origin of magnetic fields in the stars is still unknown.
-The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields \citep[see][for details ]{Dudorov2014}, or fields produced through a dynamo mechanism \citep{Cantiello2009}.
-Besides that, the presence of magnetic fields with intensities of the order of kG \citep{Hussain2014} has been a sine qua non condition for dealing with certain data observed in T Tauri type stars in young, accreting PMS systems \citep{Johns-Krull2007}.\par
+\section{Method} \label{sec_method}
 
-The mass of stars plays a determining role in the existence or not of an external convective envelope.
-In turn, the presence of this convective region conditions the evolution of magnetic fields.
-Massive stars do not have external convective zones although they can expose a convective core.
-On the other hand the less massive stars, such as the Sun, have convective covers and a radiative core.
-In both types of stars strong magnetic fields can be generated in the tachocline but with notable differences \citep[e.g.][for details]{Chabrier2006,Charbonneau2010}.\par
-
-\subsubsection{Internal Magnetic Fields}
-Our Sun is an active star as far as the magnetic field is concerned.
-According to the theoretical models this magnetic field is generated by a dynamo in the tachocline and with an intensity about $B\approx10^{5}\,\Gauss$ \citep{Maeder2003a,Dudorov2014}.
-At the moment of their generation, the lines of that magnetic field will have a poloidal orientation and will be affected by the differential rotation and forced to adopt a toroidal typology (the $\omega$-effect).
-On the other hand, the opposite case can also occur, starting from a toroidal magnetic field to obtain a poloidal configuration.
-This would be possible thanks to the convective movements of material that are produced in the CZ of the star and to assume that this material behaves as a perfect conductor.
-This second condition allows to assume that the magnetic field lines are "frozen" in the conductive fluid and therefore must move in solidarity with it ($\alpha$-effect).\par
-
-\subsubsection{Surface Magnetic Fields} \label{surf_mf}
-The intensity of the magnetic field measured on the Sun's surface is about $B\approx1\,\Gauss$ \citep{Weber1967,DAntona2000,Morin2012}, twice the value of the Earth's one.
-Additionally its topology is varied and complex.
-The dynamo model establishes that the magnetic field produced inside the star reaches the surface with the help of convective processes taken place in the CZ and gives rise to magnetically active regions on the solar surface.
-Besides, rotating stars that expose a significant outer CZ can produce surface magnetic fields through a dynamo mechanism \citep[e.g.][]{Brandenburg2004,Charbonneau2010,Brun2017}.
-Whatever the origins of surface magnetic fields are, it is expected to couple to the wind mass-loss and, if this is strong enough, to produce MB \citep[e.g.][]{UdDoula2002,Ud-Doula2007,Ud-Doula2008,Meynet2010}.\par
-
-\subsection{Effects of magnetic braking}
-In its most widely accepted approach, the MB is linked to Skumanich's work \citep{Skumanich1972} in which he develops an empirical law of its evolution in parallel to the age of the star.
-To calibrate this law, Skumanich relied on data obtained from type G stars found in the MS.
-The influence of the MB on the evolution of the star is closely related to the transport of the AM.
-According to \citet{Meynet2010}, two main mechanisms are distinguished attending to how the transport of the AM is produced:
-
-\begin{enumerate}
-    \item Differential rotation: the AM transport is pushed by southern currents and shear instabilities.
-    \item Solid body rotation: when the transport of AM is very efficient, solid body rotation is 
-    maintained throughout the entire MS phase.
-\end{enumerate}
-
-The AML caused by MB depends directly on the amount of mass lost by the star due to the stellar winds.
-The estimated mass loss ratio for a solar-type star, approximately $10^{-14}\msun \; yr^{-1}$ \citep{Noerdlinger2008}, and the resulting AML derived from that value can be considered relatively modest as to decisively influence the evolution of the star.
-Furthermore, the combined effect of convective movements and differential rotation lead to the generation of magnetic fields.
-Those magnetic fields end up reaching the surface of the star, so that the AML can be increased in several orders of magnitude \citep{Langer2012}.
-That increment is a consequence of the magnetic field that forces the stellar wind ionized particles to turn with the same angular speed, winds which extends a distance several times the radius of the star \citep[see][for more details]{UdDoula2002,Ud-Doula2007,Ud-Doula2008}.
-
-\section{Method} \label{sec_2}
-
-In the remainder of this paper, we describe a physically sound yet computationally simple model which will be used as an extension to Modules for Experiments in Stellar Astrophysics stellar evolution code (MESA; \citeauthor{Paxton2011} \citeyear{Paxton2011}, \citeyear{Paxton2013}, \citeyear{Paxton2015}, \citeyear{Paxton2018}, \citeyear{Paxton2019}) for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.
+In the remainder of this paper, we describe a semi-empirical approach for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.
+This has been implemented as an extension to Modules for Experiments in Stellar Astrophysics stellar evolution code \citep[MESA; ][]{Paxton2011, Paxton2013,Paxton2015, Paxton2018, Paxton2019}.
 
 \subsection{Modelling magnetic braking} \label{mod_mb}
-We'll start enumerating the most relevant aspects and assumptions made in modelling the evolution of rotation, magnetic braking and angular momentum in MESA for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.\par
-
-In the presence of mass-loss, AM is removed by the material in the stellar wind, resulting in a spin-down effect.
-This is quite important for example in massive stars, that are known to be rapidly rotating and have strong stellar winds.
-It has been also observed that strongly magnetic intermediate-mass stars typically have rotation rates much slower than other stars in their parent population \citep{Mathys2006}.
-In those stars, the presence of such magnetic field will interact with the mass loss and the Alfv\'{e}n radius ($R_{A}$) plays an important role in this interaction.\par
+We start by enumerating the most relevant aspects and assumptions made in modelling the evolution of rotation, magnetic braking and angular momentum in MESA.
+These will be considered for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.\par
 
@@
 
-Making use of $\eta_*$, the $R_{A}$ is defined as the point in which the magnetic field energy density $B^{2}/8\upi$, marked by magnetic field strength ($B$), and the kinetic energy density $\frac{1}{2}\rho\nu^{2}$, defined by the mass density ($\rho$) and velocity of the stellar wind $(\nu)$, are balanced.
-That is, where the coefficient between both is equal to 1.
-In case that $R_{A}$ is greater than the star radius, then the wind flow will have to follow the magnetic field.
-As a consequence, the material will leave the stellar surface with a higher specific AM, as the co-rotation radius has increased and it will roughly correspond to the $R_{A}$.
-This effect is known as MB.\par
+where $B^{2}/8\upi$ is the magnetic field energy density, $\frac{1}{2}\rho\nu^{2}$ the kinetic energy density, $B$ the magnetic field intensity, $\rho$ the mass density, and $\nu$ the stellar wind velocity.\par
 
-Each of these factors has a decisive influence on the future of the star and therefore in our routine implementation.
-On the one hand, stars with similar initial masses but different mass loss ($\Dot{M}$) ratios will end up evolving very differently.
-On the other hand, the ionized particles carried by the solar wind not only contribute to the mass loss but also to the loss of kinetic ($K_e$) energy that is deposited in the interstellar medium.
-Given a star with a spherically symmetric wind, $\Dot{M}$ and $K_e$ are characterized and related by the following expressions:
+Stars with similar initial masses but different mass loss ($\Dot{M}$) ratios will end up evolving very differently.
+The ionized particles carried by the solar wind not only contribute to the mass loss but also to the loss of kinetic ($K_e$) energy that is deposited in the interstellar medium.
+Given a star with a spherically symmetric wind, $\Dot{M}$ is characterized by the following expression:
 
 \begin{ceqn}
-\begin{align}
-    \Dot{M} &= 4\upi r^2\rho\nu \label{eq:mass_loss}\\
-    K_e &= 0.5\Dot{M}\nu_{esc}
-\end{align}
+\begin{equation}
+    \Dot{M} = 4\upi r^2\rho\nu \label{eq:mass_loss}
+\end{equation}
 \end{ceqn}
@@
 
-In turn, of the observable characteristics of the solar wind, we especially care about the parameters related to the amount of mass that is able to pull out of the outer layers of the star in a given time interval, and the speed that reaches the wind itself measured at a great distance from the star, the terminal velocity ($\nu_\infty$).
-In general, a magnetically channeled outflow will have a complex stream geometry but for convenience, the Eq.~\ref{eq:mass_loss} simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate.
-$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law (Eq.~\ref{eq:vel_law}), where $r$ represents the distance from the center of the star to the point where the velocity is to be measured, and $R_*$ is the radius of the star:
+In general, a magnetically channeled outflow has a complex stream geometry but for convenience, the Eq.~\ref{eq:mass_loss} simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate.
+$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law (Eq.~\ref{eq:vel_law})
 \begin{ceqn}
 \begin{equation}
-    \nu(r) = \nu_\infty (1-R_*/r) \label{eq:vel_law}
+    \nu(r) = \nu_\infty (1-\rstar/r) \label{eq:vel_law}
 \end{equation}
 \end{ceqn}
-where $\nu_\infty$ is the terminal wind velocity defined as the velocity that the wind, the outflowing matter, reaches at large distance from the central star, where it is not accelerated anymore by the wind driving force but its deceleration due to interaction with the interstellar medium (ISM) is negligible \citep{Niedzielski2002}.\par
+where $r$ represents the distance from the center of the star to the point where the velocity is to be measured, $R_*$ is the radius of the star, and $\nu_\infty$ is the terminal wind velocity.
+$\nu_\infty$ is defined as the velocity that the wind reaches at large distance from the central star, where it is not accelerated anymore by the wind driving force but its deceleration due to interaction with the interstellar medium (ISM) is negligible \citep{Niedzielski2002}.\par
 
-The line-driven winds of massive OB stars have terminal velocities that scale with the photospheric escape velocity \citep{Lamers2000} according to:
+The line-driven winds of OB stars have terminal velocities that scale with the photospheric escape velocity \citep{Lamers2000} according to:
 \begin{ceqn}
 \begin{align}
-\nu_{esc} &= \sqrt{\frac{2GM_*}{R_*}} \label{eq:vesc} \\
-\nu_\infty &\simeq 1.92 \;\nu_{esc}\\
-\nu_\infty &\simeq 1.92 \; x \; 618 \; \Bigg(\sqrt{\frac{R_{\sun}}{R_*}\frac{M_*}{\msun}} \;\Bigg) \label{eq:vinf}
+    \nu_\infty &\simeq 1.92 \;\nu_{{\rm esc}} \label{eq:vinf}\\
+    \nu_{{\rm esc}} &= \sqrt{\frac{2\,G\,\mstar}{\rstar}} \label{eq:vesc}
 \end{align}
 \end{ceqn}
-where Eq.~\ref{eq:vesc} describes the Newtonian escape velocity from the stellar surface, $G$ is the gravitational constant and Eq.~\ref{eq:vinf} represents the terminal velocity in terms of escape velocity, radius and mass of the Sun.\par
+where Eq.~\ref{eq:vesc} describes the Newtonian escape velocity from the stellar surface, where $G$ is the gravitational constant,   ${\rstar}$ the radius, and ${\mstar}$ the mass of the star.
 
-Given that the version of MESA used in this work does include neither the treatment of surface magnetic fields (it does account for transport by internal magnetic fields of AM and chemical elements) and therefore, nor for magnetic braking effects, it is necessary to extend the simulator to include this phenomenon.
-In the present work we have followed the theoretical work on MB in the Sun carried out by \citet{Weber1967}, making use of Eq.~ \ref{eq:j_dot} for the calculation of the AML.
+Making use of (\ref{eq:vesc}), $\nu_\infty$ can be reformulated as:
+
+\begin{ceqn}
+\begin{equation}
+    \nu_\infty \simeq 1.92 \; x \; 618 \; \Bigg(\sqrt{\frac{\rsun}{\rstar}\frac{\mstar}{\msun}} \;\Bigg) \label{eq:vinf}
+\end{equation}
+\end{ceqn}
+where ${\rsun}$ is the radius, and ${\msun}$ the mass of the Sun.
+
+It has been observed that strongly magnetic intermediate-mass stars typically have rotation rates much slower than other stars in their parent population \citep{Mathys2006}.
+In those stars, the magnetic fields interacts with the mass loss, where the Alfv\'{e}n radius ($R_{A}$) plays an important role.
+Making use of $\eta_*$ (Eq.
+\ref{eq:wind_conf}), $R_{A}$ is defined as the point in which the magnetic field energy density and the kinetic energy density are balanced.
+In case that $R_{A}$ is greater than the stellar radius, then the wind flow will have to follow the magnetic field.
+As a consequence, the material leaves the stellar surface with a higher specific AM, as the co-rotation radius has increased and it roughly corresponds to $R_{A}$.
+Following \citet{Weber1967} the AML can be calculated by:
 \begin{ceqn}
@@
 
-The above expression can be rewritten so that it depends on the $\eta_*$ instead of the $R_A$ as follows:
+The above expression can be rewritten according to \citet{Ud-Doula2008} so that it depends on the $\eta_*$ instead of the $R_A$ as follows:
 \begin{ceqn}
@@
 \end{ceqn}
-which is a more convenient expression to be implemented in MESA because is based on values directly exposed during the simulations and suitable for the approach described in \citet{Ud-Doula2007}.
+which is a more convenient expression to be implemented in MESA because is based on values directly exposed during the simulations.
 
@@
 \subsection{Mesh Models} \label{sec_mesh}
-
-In its simplest form the implementation of the model requires the specification of both the surface magnetic field strength ($B$) and the initial rotation rate ($\Omega$) at star surface.
-The numerical simulations trace the rotational history and A(Li) of a $1\, \msun$ star for a variety of initial values for $B$ and $\Omega$.
-Although stars have a three-dimensional structure and the equations that model their evolution must take this fact into account, it has been proven sufficient to solve the equations in only one dimension, extrapolating the results to the other two \citep[shellular approximation,][]{Meynet1997}.
-MESA uses this simplification to accelerate computational calculations \citep{Paxton2013}.\par
+According to the formulation in the previous sections, the only two free parameters of our implementation are $B$ and $\Omega$.
+The numerical simulations traced the rotational history and $A(\isotope[7]{Li})$ of a $1\, \msun$ star for a variety of initial values for $B$ and $\Omega$ (see also Table \ref{tab:phy_mesa}).
+MESA uses the shellular approach \citep{Meynet1997} to account for hydrostatic effects of rotation into 1D stellar models.\par
 
@@
 		Rotation & Differential rotation at PMS \& MS\\
-		Convection & MLT + Ledoux, $\alpha_{MLT}$ = 1.82\\
-		Overshoot & time-dependent, diffusive, \\ & $f_{ov,core}=0.0160$,\\ 
-		& $f_{ov,sh}=0.0174$\\
-		Semiconvection & $\alpha_{sc}=0.1$\\
-		Thermohaline & $\alpha_{th}=666$\\
+		Convection & MLT + Ledoux, $\alpha_{{\rm MLT}}$ = 1.82\\
+		Overshoot & time-dependent, diffusive, \\ & $f_{{\rm ov,core}}=0.0160$,\\ 
+		& $f_{{\rm ov,sh}}=0.0174$\\
+		Semiconvection & $\alpha_{{\rm sc}}=0.1$\\
+		Thermohaline & $\alpha_{{\rm th}}=666$\\
 		Rotational Mixing & Include SH, ES, GSF, SSI \& DSI\\
@@
 		Magnetic Field & B(G) variable between [3.0 - 5.0]\\
-		Mass Loss & activated, $\Dot{M}_{max} = 10^{-3} \: \msun \: yr^{-1}$\\
+		Mass Loss & activated, $\Dot{M}_{{\rm max}} = 10^{-3} \: \msun \: yr^{-1}$\\
 		Angular Moment Loss & activated, $\Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A}$\\
@@
 
-We define here also additional conventions and assumptions adopted in the simulations.
-For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, \mathrm{pr} = 0.0142}$) not the current photospheric metallicity ($Z = Z_{\odot, \mathrm{ph} = 0.0134}$) as the reference value.
-We also adopt the following nominal values to express stellar properties values in SI units $\rsun = 6.957x10^{10}\, cm$ and $\msun = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
-For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}.
-That work, in particular its section 4.1, has been used as a starting point for ours regarding the parameterization of the MESA project which is calibrated to reproduce the measured element abundances on the solar surface.\par
+We adopted solar-scaled abundances and assume \citet{Asplund2009} solar initial abundance ($Z = Z_{\odot, \mathrm{pr} = 0.0142}$).
+We also adopted the following nominal values to express stellar properties values in SI units, $\rsun = 6.957x10^{10}\, cm$ and $\msun = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
+For a detailed description of the physics adopted in this paper, we refer the reader to \citet{Choi2016}.
+That work was used as starting point for ours regarding the parameterization of the MESA project which was calibrated to reproduce the measured element abundances on the solar surface.\par
 
-The models include already rotation during the PMS as there evidences that advocate for a strong established relationship between Li destruction and rotation on that phase (\citeauthor{Bouvier2016} \citeyear{Bouvier2016}, \citeyear{Bouvier2018}), and by activating it in the MESA models, the evolutionary computational code takes into account the following effects:
+The models included rotation during the PMS as there are evidences that advocate for a strong established relationship between Li destruction and rotation on that phase (\citeauthor{Bouvier2016} \citeyear{Bouvier2016}, \citeyear{Bouvier2018}).
+We modeled rotation such as AML is computed as a result of the torque applied to the convection zone by a magnetically-coupled wind.
+Additionally, we did not take into account either the influence of internal magnetic fields or their existence during the T-Tauri phase.
+We adopted a simple and pragmatic theoretical approach to establish when the star is reaching the ZAMS phase.
+The selected criterion was based on the simultaneous existence of an extensive convective layer and a radiative core.
+From this moment on, the MB routine was activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participated in the star AML.
+We assumed that the magnetic field did not vary its intensity throughout the evolution of the star until it reached the TAMS.
+In addition, other rotation effects were also considered during the evolution of the models: the AM transport from the radiative interior to the convective envelope and the AM redistribution associated with changes in the internal structure during the process of contraction to the MS.\par
 
-\begin{enumerate}
-    \item AM transport from the radiative interior to the convective envelope in response to the rotational deceleration of the star surface layers\label{itm:1}.
-    \item AM redistribution associated with changes in internal structure during the process of contraction to the MS\label{itm:2}.
-    \item AML as a result of the torque applied to the convection zone by a magnetically coupled wind\label{itm:3}.
-\end{enumerate}
+In this study we focused on the indirect role played by the MB on the Li destruction.
+This is a consequence of its influence on the rotational history of solar-type stars.
+Our major objective was to check how MB could contribute to explain the evolution of Li in the Sun and other solar-type stars.
+A set of different scenarios with magnetic field strengths ranging between 3.0 and 5.0 G were exercised.
+We computed the evolution of $1\,\msun$ stellar models at solar initial metallicity with $\oomegac$ varying between $0.0084$ and $0.0336$.
+More advanced models for more complex arrangements of the magnetic field as well as the evolution of its intensity throughout the life of the star were left for later work.\par
 
-The enumerated effects \ref{itm:1} and \ref{itm:2} are standard features offered by MESA.
-On the contrary, the effect \ref{itm:3} is not available as an out-of-the-box feature during the elaboration of this paper and represents our extension to MESA.
-Additionally, we do not take into account either the influence of internal magnetic fields or their existence during the T-Tauri phase.
-We adopt a simple and pragmatic theoretical approach to establish their occurrence when the star is reaching the ZAMS phase.
-The adopted criterion is based on the simultaneous existence of an extensive convective layer and a radiative core.
-From this moment on, the MB routine will be activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participates in the star AML.
-Additionally, we assume that the magnetic field does not vary its intensity throughout the evolution of the star until it reaches the TAMS.\par
+As expressed by Eq.~\ref{eq:j_dot_mesa}, the amount of AML depends on $R_A$ (or alternatively on $\eta_*$), $\Omega$ and $\Dot{M}$.
+For large $\eta_*$ value, the star undergoes a significant deceleration.
+We controlled the value of $\eta_*$ in the models by varying the magnetic field ($B$) and the stellar rotation speed ($\Omega$).
+With regards to $\Dot{M}$ and as reported in Table \ref{tab:phy_mesa}, the empirical formula developed by Reimers \citep{Reimers1975} for stars in the asymptotic giant branch (AGB) was used for calculating the mass-loss.
+For a solar-type star the $\Dot{M}$ during MS is relatively small, about  $10^{-14}\msun \, yr^{-1}$ \citep{Noerdlinger2008}.
+\par
 
-In this study we do not only focus on investigating the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.
-Keeping in mind this approach, our major objective is aimed to check the degree of validity of the resulting model after including the MB effect as a complementary mechanism of AML that can contribute to explain the evolution of Li in the Sun and other solar-type stars.
-A set of different simulation scenarios with magnetic field strengths ranging between 3.0 and 5.0 G have been exercised.
-We would like to stress that the magnetic field remains constant during the PMS \& MS phases.
-We compute the evolution of $1\msun$ stellar models at solar metallicity with $\oomegac$ varying between $0.0084$ and $0.0336$.
-More advanced models for more complex arrangements of the magnetic field as well as the evolution of its intensity throughout the life of the star are left for later work.\par
-
-As expressed by Eq.~\ref{eq:j_dot_mesa}, the amount of AML is directly influenced by the $R_A$ (or alternatively by the $\eta_*$), $\Omega$ and $\Dot{M}$.
-If the value of $\eta_*$ is large, the obtained AML will be also large and the star undergoes a considerable deceleration.
-In our proposal, the value of $\eta_*$ can be directly influenced by establishing in the different simulations the value of the magnetic field ($B$) on the surface or the rotation speed of the star ($\Omega$).
-With regards to $\Dot{M}$ and as reported in table (\ref{tab:phy_mesa}), for the set of simulations performed by MESA the empirical formula developed by Reimers \citep{Reimers1975} for stars in the asymptotic giant branch (AGB) is used for calculating its value.
-For a solar-type star the $\Dot{M}$ during MS is relatively small, about  $10^{-14}\msun \, yr^{-1}$ \citep{Noerdlinger2008}.
-Finally, $\Omega$ can be influenced in our models specifying the ratio $\oomegac$ as another free parameter, where  $\omegac$ represents the star surface velocity at which the centrifugal force balances  the gravitational force.
-Beyond this point, the star would begin to expels matter along the equator.\par
-
-MESA assigns an $\Omega$ value for each cell $k$ ($\Omega_k$) which is adjusted so that the resulting angular momentum is retained after adjusting the new mass of the cell $k$ ($m_k$) and its distance to the center of the star ($r_k$).
+MESA assigns an $\Omega$ value for each cell $k$ ($\Omega_k$) which is adjusted so that the resulting angular momentum is retained after calculating the new mass of the cell $k$ ($m_k$) and its distance to the center of the star ($r_k$).
 After that, an AM value is assigned to each cell $k$ ($J_k$).
-It is in this point when our MB routine plugs in and influences the internally calculated value by providing an additional contribution ($\Dot{J}_{k}$) which is the result of the external torque exerted by the magnetic field once it has been distributed among the different layers that make up the CZ as dictated by Eq.~\ref{eq:k_jdot}:\par
+At this point, our MB turns on, modifying $J_k$.
+This was done by providing an additional contribution ($\Dot{J}_{k}$).
+This contribution is the result of the external torque exerted by the magnetic field once it has been distributed among the different layers that make up the CZ as dictated by Eq.~\ref{eq:k_jdot}:\par
 
@@
 \begin{figure}
-	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip, width=\columnwidth]{figures/li_var_vel_0_0g.eps}
+	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip, width=\columnwidth]{figures/li_var_vel_0_0g.eps}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
@@
 The purple star and square are surface Li abundances for the present-day Sun \citep{Asplund2009} and the average for the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
     \label{fig:li_var_vel_0g}
@@
 
-Figure \ref{fig:li_var_vel_0g} shows the evolution of surface Li abundance relative to H as a function of time and for several $1\msun$ models initialized with different rotational velocities.
-The simulations take into account the effects of rotation and AML caused by stellar winds but not those of MB.
-The purple star and square are the surface Li abundances for the present-day Sun \citep{Asplund2009} and for the Pleiades, respectively.
-\citep{Sestito2005}.\par
+Figure \ref{fig:li_var_vel_0g} shows the temporal evolution of surface Li abundance for several $1\msun$ models initialized with different rotational velocities.
+The simulations took into account the effects of rotation and AML caused by stellar winds but not those of MB.
+The purple star and square are the surface Li abundances for the present-day Sun \citep{Asplund2009} and for the Pleiades, respectively \citep{Sestito2005}.\par
 
-Additionally, notice how the Li abundance in the star surface decreases over time for all simulated models.
+Notice how the Li abundance in the star surface decreased over time for all simulated models.
 The solid black line represents the reference model that adopts the solar-calibrated envelope overshooting parameters as documented in \citet{Choi2016}.
-All models burn too much Li before the ZAMS and therefore do not match with the Pleiades average surface Li abundance.
-Also remarkable is the fact that there are barely any differences between the different models in terms of the abundance of Li for much of the PMS.
-Only after one million years is Li destroyed to an accentuated degree, since the necessary temperature is not reached at BCZ before.
-Afterwards the different models destroy the Li in a very similar way due to two main reasons.
-On the one hand, the convective zones that the models develop have very similar sizes, so that the temperature at the BCZ is practically the same.
-On the other hand, the differential rotation between the core and the convective zone does not begin to develop until it reaches $ \approx 10^6$ years, achieving its maximum difference in the ZAMS about $ \approx 10^7$ years.
-It is at this moment that the combined effects of turbulence and rotational difference should increase, leading to notable differences in the evolution of A(Li) (see Figure \ref{fig:li_var_vel_0g_z1}).
-Later, the reference model does not deplete Li efficiently on the MS and fail (again) to match the current solar surface Li abundance.
-On the other hand, the other models which include rotation during the PMS with values of $\oomegac$ between 0.0084 and 0.0336 are able to burn Li in a more realistic way although only one of them (green line) is close to match the present-day Li abundance of the Sun but its rotational velocity is much higher (see Figure \ref{fig:rot_vel_0g}) than the $2\,\kms$ of the Sun \citep{Gill2012}.
+All models burned too much Li before the ZAMS and therefore did not match with the Pleiades average surface Li abundance.
+Also remarkable was the fact that there were barely any differences between the different models in terms of the abundance of Li for much of the PMS.
+Only after one million years was Li destroyed to an accentuated degree, since the necessary temperature was not reached at BCZ before.
+Afterwards the different models destroyed the Li in a very similar way due to two main reasons.
+On the one hand, the convective zones that the models developed had very similar sizes, so that the temperature at the BCZ was practically the same.
+On the other hand, the differential rotation between the core and the convective zone did not begin to develop until it reached $ \approx 10^6$ years, achieving its maximum difference in the ZAMS about $ \approx 10^7$ years.
+It was at this moment that the combined effects of turbulence and rotational difference should increase, leading to notable differences in the evolution of A(Li) (see Figure \ref{fig:li_var_vel_0g_z1}).
+Later, the reference model (black line) did not deplete Li efficiently on the MS and failed (again) to match the current solar surface Li abundance.
+The other models which included rotation during the PMS with values of $\oomegac$ between 0.0084 and 0.0336 were able to burn Li in a more realistic way.
+Among them only one (green line) was close to match the present-day Li abundance of the Sun but its rotational velocity was much higher (see Figure \ref{fig:rot_vel_0g}) than the $2\,\kms$ of the Sun \citep{Gill2012}.
 \par
 
-For much of the PMS the star rotates as a solid body (see Figure \ref{fig:rot_vel_0g}) and this is because the star has a completely convective interior.
-It is not until the end of Hayashi track that the star begins to develop a radiative core due to the increase in temperature that occurs inside the star during the contraction process.
-It is in this stage of evolution of the star when a difference in angular velocity appears between the upper and lower limits of the radiative and convective zones respectively.
-The degree of differential rotation is directly influenced by the initial $\Omega$.
-As the models are initialized with a higher angular velocity, the difference in velocity between the BCZ and the star surface is accentuated in a directly proportional mode; the higher the initial velocity, the bigger the velocity gradient between the bottom and top limits of the CZ.
-As a consequence, the turbulence strength located at the BCZ increases so that Li can reach regions with temperatures about $\tli$ where it is finally burned and destroyed (see Figure  \ref{fig:li_var_vel_0g}).
+For much of the PMS the star rotated as a solid body (see Figure \ref{fig:rot_vel_0g}) and this was because the star had a full convective interior.
+It was not until the end of the Hayashi track that the star began to develop a radiative core.
+It was in this stage when a difference in angular velocity appeared between the upper and lower limits of the radiative and convective zones respectively.
+The degree of differential rotation was directly influenced by the initial $\Omega$.
+As the models were initialized with a higher angular velocity, the difference in velocity between the BCZ and the star surface was accentuated; the higher the initial velocity, the bigger the velocity gradient between the bottom and top limits of the CZ.
+As a consequence, the turbulence strength located at the BCZ increased so that Li could reach regions with temperatures about $\tli$ where it was finally burned and destroyed (see Figure  \ref{fig:li_var_vel_0g}).
 Other investigations \citep{Bouvier2018, Baraffe2017}  point to a tendency diametrically opposed to the one exposed here, i.e.
@@
 \begin{figure}
-	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
+	\includegraphics[trim = 25mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
     \caption {Similar to Figure \ref{fig:li_var_vel_0g} but zooming in on the ZAMS.
@@
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
     \caption{The evolution of angular velocity at surface (solid line) and at the bottom (dotted line) limit of the uppermost convective zone, as a function of time for several 1 $\msun$ models.
@@
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The dashed vertical lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
     \label{fig:rot_vel_0g}
@@
 
-Another well known structural effect of rotation is the decrease of the effective temperature ($\teff$) and to a less extent of stellar luminosity ($L$).
+Other well-known structural effects of rotation are the decrease of the effective temperature ($\teff$) and to a less extent of stellar luminosity ($L$).
 Both effects can be observed graphically in the HR diagram of Figure \ref{fig:hr_var_vel_0g} which shows a zoomed-in view of evolutionary tracks from the ZAMS until the TAMS for several $1\msun$ models initialized with different rotational velocities.
-If we compare the non-rotating model (black solid line on the left side) with the rotating ones we can recognize that at the end of the PMS, the latter reach before the ZAMS and exhibit a lower $\teff$ than the former.
+If we compare the non-rotating model (black solid line) with the rotating ones we can recognize that at the end of the PMS, the latter reach the ZAMS with a lower $\teff$ than the former.
 These results are in line with those of previous studies \citep[see e.g.
@@
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/hr_var_vel_0_0g_z1.eps}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/hr_var_vel_0_0g_z1.eps}
     \caption{An example solar 1$\msun$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities.
@@
 \subsection{Li evolution with MB}
-In the rest of this document we are going to focus, unless otherwise indicated, on models that include rotation from the PMS since they are the ones that can offer a greater approximation to the observations.\par
-
-Figure \ref{fig:li_var_vel_4_0g} shows the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
-Those models were initialized with different rotational velocities and took into consideration both the effects of AML and the MB caused by a magnetic field of intensity 4G.
-If we compare it with Figure \ref{fig:li_var_vel_0g} in which the effects of MB are neglected, we notice how the profiles of Li abundance are altered during PMS and MS.
-In the first phase we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see Eq.~\ref{eq:j_dot}) depends directly on the mass loss rate.
-If we take into account that for solar-type stars the models predict a modest total mass loss rate, that value is even much lower during PMS.
-On the contrary, during the MS it is observed that the AML is much more significant, causing the star to rotate more slowly and as a result a smaller amount of Li to be destroyed.\par
+Figure \ref{fig:li_var_vel_4_0g} shows the temporal evolution of surface Li abundance for several 1 $\msun$ models.
+Those models were initialized with different rotational velocities and took into consideration the effects of MB caused by a magnetic field of intensity 4G.
+If we compare it with Figure \ref{fig:li_var_vel_0g} in which the effects of MB were neglected, we notice how the profiles of Li abundance were altered during PMS and MS.
+During the PMS we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see Eq.~\ref{eq:j_dot}) depends directly on the mass loss rate.
+If we take into account that for solar-type stars the models predict a modest total mass loss rate, that value is even much lower in this phase.
+On the contrary, during the MS it is observed that the AML is much more significant, causing a smaller amount of Li to be destroyed.\par
 
 \begin{figure}
-	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_var_vel_4_0g.eps}
+	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/li_var_vel_4_0g.eps}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
@@
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
     \label{fig:li_var_vel_4_0g}
@@
 The effect of the MB routine can be seen even more clearly in Figures~\ref{fig:rot_vel_4g}, \ref{fig:rot_vel_4g_z1} \& Appendixes \footnote{The appendices comprise a series of grids as a function of time and for several 1 $\msun$ models in which on the one hand, the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H} for both variable magnetic field intensities and angular velocities and on the other hand, the evolution of surface rotational velocity are shown.}.
-In those figures we represent the rotation profiles for the surface of the stars and for the bottom of the convective envelope for a number of 1 $\msun$ models initialized with different rotational velocities and considering the influence of MB.
-Similarly to the evolution profiles of the Li on the surface commented in the previous paragraph, the effect of the routine is visible once the ZAMS is reached.
-If we compare the evolution of the curves presented here with those of Figure \ref{fig:rot_vel_0g} we see how the star, instead of continuing to increase its $\Omega$, begins to slow down after having reached its maximum in the passage through the ZAMS.
-It is also at this stage when the angular velocities at the surface of the star and at the BCZ reach their maximum difference.
-On the other hand, the effect of magnetic braking causes the angular velocities between both zones of the star to decrease until, for an age close to that of the Sun (see Figure \ref{fig:rot_vel_4g_z1}), the star practically rotates like a rigid solid.
-These results are also consistent with those obtained by \citet{Eggenberger2010} as far as the effect of the magnetic field, in particular its influence on the loss of angular momentum, has on the rotational velocity of the star.
-In our models we have also chosen to include the rotation effects offered by MESA but not those of Spruit-Tayler (ST) dynamo on the diffusion of elements.
-The reason is that the implementation of MESA that makes of the ST dynamo is to simulate internal magnetic fields and not for surface magnetic fields \citep{Paxton2013}.
+In those figures we represented the rotation profiles for the surface of the stars and for the bottom of the convective envelope.
+They are 1 $\msun$ models initialized with different rotational velocities and considering the influence of MB.
+Similarly to the $A(\isotope[7]{Li})$ evolution profiles commented in the previous paragraph, the effect of the routine were visible once the ZAMS was reached.
+If we compare the evolution of the curves presented here with those of Figure \ref{fig:rot_vel_0g} we see how the star, instead of keep increasing $\Omega$, began to slow down after having reached its maximum in the passage through the ZAMS.
+Notice that at this point the angular velocities at the surface of the star and at the BCZ reached their maximum difference.
+On the other hand, the MB effect caused the angular velocities between both zones of the star to decrease until, for an age close to that of the Sun (see Figure \ref{fig:rot_vel_4g_z1}), the star practically rotated like a rigid solid.
+These results were also consistent with those obtained by \citet{Eggenberger2010} as far as the effect of the magnetic field, in particular its influence on the loss of angular momentum, has on the rotational velocity of the star.\par
 
-In a similar way we also observe that the models with lower angular velocity generally end up exhibiting higher values in the abundance of Li on the surface (see Figures~ \ref{fig:li_var_vel_4_0g}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).
-In none of those cases we do obtain values of Li on the surface higher than those shown by the model without rotation.\par
+In a similar way we also observed that the models with lower angular velocity generally ended up exhibiting higher values for the Li abundance on the surface (see Figures~ \ref{fig:li_var_vel_4_0g}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).
+In none of those cases we did obtain values of Li on the surface higher than those shown by the model without rotation.\par
 
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g.eps}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g.eps}
     \caption{The evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
@@
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
     \label{fig:rot_vel_4g}
@@
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g_z1.eps}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g_z1.eps}
     \caption{Similar to Figure \ref{fig:rot_vel_4g} but now showing in detail the surface rotational velocity as the star approaches TAMS.}
@@
 
-The MB also leaves its mark on the HR diagram by significantly affecting the $\teff$ of the star.
-To visualize this effect we will take as reference Figure \ref{hr_vc_0336_var_g_z1} in which all models were started with the same value $\oomegac=0.0336$ but the intensity of the simulated magnetic field was different.
-In particular, let us pay attention to the evolutionary trace in which a magnetic field was not simulated (black solid line) and to the rest of the evolutionary traces in which it was.
-We have that the group of simulations with presence of a magnetic field produces hotter stars due to the MB influence.
-The lower speed with which the star rotates due to the MB effect causes the increase of the $\teff$, being this difference of practically $95\,\Kelvin$ between the simulated models with $0.0\,\Gauss$ and $5.0\,\Gauss$ respectively for $log(\llsun)=0$.
+The MB also left its mark on the HR diagram by significantly affecting the $\teff$ of the star.
+To visualize this effect we take as reference Figure \ref{hr_vc_0336_var_g_z1} in which all models were started with the same value $\oomegac=0.0336$ but the intensity of the simulated magnetic field was different.
+Notice that the models with MB produced hotter stars due to its influence.
+The lower speed with which the star rotated due to the MB effect caused the increase of the $\teff$, being this difference of practically $95\,\Kelvin$ between the simulated models with $0.0\,\Gauss$ and $5.0\,\Gauss$ respectively for $log(\llsun)=0$.
 
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 15mm 15mm, clip,width=\columnwidth]{figures/hr_vc_0336_var_g_z1.eps}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/hr_vc_0336_var_g_z1.eps}
     \caption{Similar to Figure \ref{fig:hr_var_vel_0g} but now showing in detail the effects of magnetic braking on the evolutionary tracks for different magnetic field strengths and $\oomegac=0.0336$.
@@
 
-As described in section \ref{sec_mesh}, the MB routine distributes the total amount of AML calculated according to Eq.~\ref{eq:k_jdot} among the different layers that compose the CZ.
+As described in section \ref{sec_mesh}, the MB routine distributed the total amount of AML calculated according to Eq.~\ref{eq:k_jdot} among the different layers that composed the CZ.
 In Figure \ref{fig:cz_vc_028_var_b} we can observe the evolution of the most external CZ normalized with respect to the radius of the star for several 1 $\msun$ models.
@@
 In this point it increases significantly as a response to the generalized expansion of the star's radius.
-Regarding the effect of MB on the size of the CZ, we observe that as the intensity of the magnetic field increases, the size of the CZ decreases (see Figure \ref{fig:cz_vc_028_var_b}).
-The radiative core pushes outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base drops below $\tli$.
-This effect is most evident during MS (see Figure \ref{fig:cz_vc_028_var_b_z1}).
-The decrease in the size of the CZ is in line with the fact that less Li is destroyed by causing less star material to reach areas with temperatures above $\tli$.\par
+Regarding the effect of MB on the size of the CZ, we observe that as the intensity of the magnetic field increased, the size of the CZ decreased (see Figure \ref{fig:cz_vc_028_var_b_z1_new}).
+The radiative core pushed outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base dropped below $\tli$.
+This effect was most evident during MS.
+The decrease in the size of the CZ was in line with the fact that less Li was destroyed by causing less stellar material to reach areas with temperatures above $\tli$.\par
+
+\textbf{Notice the jagged appearance in Figure~\ref{fig:cz_vc_028_var_b_z1_new}.
+We suspected a numerical origin for the wriggles.
+In order to check it out, we computed models for all the physical options considered with higher temporal (time step) and spatial (number of shells) resolutions.
+For the sake of visual clarity, we illustrated the impact on the numerical solutions only for the 5G case, although the effects are the same in all cases.
+These are indicated as dotted lines in the figure, shifted-down just to become visible.
+Note that the larger the spatial resolution the smoother is the curve.
+The effect is less significant for the temporal resolution.
+The curve with 2x in both spatial and temporal resolution shows no difference with the case of 2x in only the number of shells.
+The 4x in spatial resolution points to the most appropriate number of shells to fine-track the evolution of the convective zone.
+However such high resolution implies to significantly increase the computation time, with no effect in the final conclusions of this work.
+Indeed, the size of the zigzag features is around $10^{-3}$,  which is far below the convection timescale and Li abundances measured in the stellar surface.
+It is thus save to neglect such effects by using the original time/space resolution.}
+
+\par
 
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g.eps}
+	\includegraphics[trim = 15mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g.eps}
     \caption{The evolution of convective zone size as a function of time for several 1 $\msun$ models.
 The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths vary between $0.0\,\Gauss$ and $5.0\,\Gauss$.
-The dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
     \label{fig:cz_vc_028_var_b}
@@
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g_z1.eps}
-    \caption{Similar to Figure \ref{fig:cz_vc_028_var_b} but now showing in detail the convection zone size from the near ZAMS till the TAMS.
-As the intensity of the magnetic field increases, the size of the convective zone decreases.}
-    \label{fig:cz_vc_028_var_b_z1}
+	\includegraphics[trim = 15mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g_z1_new.eps}
+    \caption{\textbf{Similar to Figure \ref{fig:cz_vc_028_var_b} but now showing in detail the size of the convection zone in the models from ZAMS up to TAMS.
+As the intensity of the magnetic field increases, the size of the convective zone decreases.
+The first three pointed curves (below 0.26), have been artificially shifted downwards to illustrate the impact of the temporal (2x in time step and number of shells, red-dotted line) and spatial (2x and 4x number of shells, black and green-dotted lines, respectively) resolution on the numerical solution for the 5G case.}}
+    \label{fig:cz_vc_028_var_b_z1_new}
 \end{figure}
@@
 The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths varying between $0.0\,\Gauss$ and $5.0\,\Gauss$.
-In the first stages of PMS is where the greatest mass loss is concentrated and this diminishes as it approaches the ZAMS.
-If we add to this scenario the fact that the star also decreases its radius during the PMS we have as result that the star increases $\Omega$ obeying the principle of conservation of AM.
-When reaching the ZAMS, the radius of the star remains more or less stable for much of the MS (except for its final stage) but continues to lose mass so it also increases its $\Omega$, although in a less aggressive way if we compare it with the PMS.
-However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see Figure \ref{fig:mb_act_var_vel_vc_028}), the MB routine is activated, causing the angular velocity of the star to begin to decrease along the entire MS.
+In the first stages of PMS was where the greatest mass loss was concentrated and this diminished as it approached the ZAMS.
+Because the star also decreased its radius during the PMS, it increased $\Omega$ obeying the principle of conservation of AM.
+When reaching the ZAMS, the stellar radius remained more or less stable for much of the MS (except for its final stage, see Figure \ref{fig:mdot_vc_028_var_b_z1}) but continued to lose mass whilst increasing $\Omega$, although in a less aggressive way if we compare it with the PMS.
+However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see Figure \ref{fig:mb_act_var_vel_vc_028}), the MB routine was activated, causing the angular velocity of the star to begin to decrease along the entire MS.
 The more intense the magnetic field, the greater the braking effect.\par
@@
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g.eps}
+	\includegraphics[trim = 5mm 10mm 20mm 10mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g.eps}
     \caption{The evolution of mass loss $\Dot{M}$ as a function of time for several 1 $\msun$ models.
 The models include different magnetic field intensity between $0.0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.028$.
-The dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
     \label{fig:mdot_vc_028_var_b}
@@
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g_z1.eps}
+	\includegraphics[trim = 5mm 10mm 20mm 10mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g_z1.eps}
     \caption{The evolution of mass loss $\Dot{M}$ as the star is approaching TAMS, as a function of time for several 1 $\msun$ models.
@@
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 15mm 15mm, clip,width=\columnwidth]{figures/mb_act_vc_028_var_g.eps}
+	\includegraphics[trim = 15mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/mb_act_vc_028_var_g.eps}
     \caption{The activation of the magnetic braking routine as a function of the presence of a radiative core.
@@
 By implementation decision, once the routine is activated, it remains on even if the star's core nature change to convective.
-The vertical dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
     \label{fig:mb_act_var_vel_vc_028}
@@
 \subsection{Alternative Li evolution with MB}
-Up to this point, the simulations of the different models have been based on the parameterization gathered in table \ref{tab:phy_mesa}, which in turn has been adopted from \citet{Choi2016}.
-If we recall the evolution of the Li shown in Figures~\ref{fig:li_var_vel_0g} and \ref{fig:li_var_vel_4_0g}, we highlighted that on both cases too much Li was burned before reaching the ZAMS and therefore do not match with the Pleiades average surface Li abundance.
-On the other than, using a parameterization slightly different from the one used until now in which the parameters of convection and overshooting have been readjusted according to table \ref{tab:phy_alt_mesa}, the simulations can reproduce more faithfully both the Li abundance of the Pleiades cluster and of the Sun (see Figure \ref{fig:li_3_0g_0314vc}).\par
+Up to this point, the simulations of the different models were based on the parameterization gathered in Table \ref{tab:phy_mesa}, which in turn was adopted from \citet{Choi2016}.
+If we recall the evolution of the Li shown in Figures~\ref{fig:li_var_vel_0g} and \ref{fig:li_var_vel_4_0g}, we highlighted that on both cases too much Li was burned before reaching the ZAMS and therefore did not match with the Pleiades average surface Li abundance.
+On the other hand, using a parameterization slightly different from the one used until now in which the parameters of convection and overshooting have been readjusted according to Table \ref{tab:phy_alt_mesa}, the simulations could reproduce more faithfully both the Pleiades cluster and the Sun Li abundances (see Figure \ref{fig:li_3_0g_0314vc}).\par
 
@@
 \begin{figure}
-	\includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/li_3_0g_0314vc.eps}
+	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/li_3_0g_0314vc.eps}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for a 1 $\msun$ model.
@@
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
     \label{fig:li_3_0g_0314vc}
@@
 Similarly, the evolution of the angular velocity for this new configuration is shown in Figure \ref{fig:rot_vel_var_vel_mlt_3_0g}.
-Here, on the contrary, we have to note that with the new simulation we still do not reproduce correctly the angular velocity of the Sun.
-Having said this and based on the results obtained in this last section, it becomes evident that the influence of the free, relatively arbitrary, parameters associated with Mixing Lenght Theory (MLT) significantly condition the evolution of the abundance of Li.
-This opens the door for other solar-compatible parameterizations to offer results in line with the observations in young stellar clusters and for the Sun.
+Here we did not yet reproduce correctly the angular velocity of the Sun.
+It became evident that the influence of the free, relatively arbitrary, parameters associated with MLT significantly conditioned the evolution of the Li abundance.
+However, without the inclusion of MB it was not possible to fit it for the Pleiades and the Sun with the same evolutionary track.
+This fact opens the door for other solar-compatible parameterizations to reproduce the observations in young stellar clusters and for the Sun.
 \par
@@
 \begin{figure}
-	\includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\columnwidth]{figures/rot_vel_3_0g_0314vc.eps}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/rot_vel_3_0g_0314vc.eps}
     \caption{The evolution of surface rotational velocity, as a function of time for a 1 $\msun$ model.
-The model include a magnetic field with an intensity of 3\,G, PMS rotation with $\oomegac=0.0314$, respectively and MB.
-MLT and overshooting parameters have been set to $\alpha_\mathrm{MLT}=1.70$, $f_\mathrm{ov,core}=0.016$, $f_\mathrm{ov,sh}=0.002$.
+The model include a magnetic field with an intensity of 3\,G, PMS rotation with $\oomegac=0.0314$, parameters in Table \ref{tab:phy_alt_mesa} and MB.
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
     \label{fig:rot_vel_var_vel_mlt_3_0g}
@@
 
-We propose that this adjustment could also potentially be achieved by a law of interdependence between $\Omega$ and $B$.
-So that, during the PMS, when the star rotates faster, the MB effect is more efficient.
-On the other hand, during the MS, when the start slows down, the MB will become less intensive.
-This approach would establish a self-regulating mechanism over the angular velocity of the star that would end up directly influencing the evolution of A(Li).
-This is the line of work in which we will continue our research.\par
-
-\section{Conclusions and future works} \label{sec_4}
-We have demonstrated that both rotation and MB induced effects must be taken into account in the theoretical models for a proper understanding and modelling of the evolution of stars.
+\section{Conclusions and future works} \label{sec_conclusions}
+We have shown, through the different simulated stellar models, that the effects induced by the combination of both rotation and magnetic braking mechanisms offer a plausible way to reconcile the observational data with the theoretical models.
 The latter is of major importance both for the transport of AM and for the transport of chemical elements.
@@
 Particularly the study of the evolution of the magnetic fields during the PMS and MS and their impact on AML.
+\par
+
 Future and improved implementations of our routine will be made to study the results on tracks, surface Li composition, rotation, etc.
 It is likely that in a rotating stellar model simulated with MB from the beginning, the differential rotation is very much reduced and therefore Li abundances observed in young stars could be properly explained.
+We propose that this result could be achieved by a law of interdependence between $\Omega$ and $B$.
+So that, during the PMS, when the star rotates faster, the MB effect is more efficient.
+On the other hand, during the MS, when the start slows down, the MB will become less intensive.
+This approach would establish a self-regulating mechanism over the angular velocity of the star that would end up directly influencing the Li evolution.
 It is equally important to understand better the general role of mass loss in AML and mostly in cooler, low-mass stars.
-Nowadays it is challenging to determine the terminal velocity of the stellar wind for this type of stars which, as documented previously, plays a key role on the AML.\par
+Nowadays it is challenging to determine the terminal velocity of the stellar wind for this type of stars which plays a key role on the AML.\par
 
-We emphasize as well that our models fail to match the observed solar Li abundance and $\Omega$.
-We cannot, therefore, ascertain to have correctly modeled the rotational history of the Sun.
-Therefore, in view of these shortcomings in our models, we must analyse the results obtained with caution and not draw any premature conclusions.
-Having said that, we have also shown how an alternative MLT parametrizations could produce results in line with the observations.
-We are convinced that our research is on the right track and we do believe that the following conclusions, based on the results described in previous sections, have been sufficiently supported in the course of this work:
+We emphasize as well that our models failed to match at the same time the observed solar Li abundance and $\Omega$.
+We could not, therefore, ascertain to have correctly modeled the rotational history of the Sun.
+In view of these shortcomings in our models, we must analyse the results obtained with caution and not draw any premature conclusions.
+We have also shown how an alternative MLT parametrizations could produce results in line with the observations.
+The following conclusions have been supported in the course of this work:
 \begin{enumerate}
@@
     \item A combination of rotation during the PMS and MB effect during MS produces different, potentially more promising behavior than those produced by standard models.
+Thus, our approach points to reproduce the observed A(Li) and the solar rotational velocity at the same time.
     \item $\teff$ from standard evolutionary tracks represent upper limits since these models do not take into consideration the magnetic braking effect nor rotation.
@@
 
-In the near future, with the new version of our code including magnetic field evolution in MESA, we will focus our efforts on understanding how magnetic fields are linked to rotation, how their topologies are and how they evolve in time in order to go beyond the treatment on the MB routine that has been done in this work.\par
+The next steps will be focused on understanding how magnetic fields are linked to rotation, how their topologies are and how they evolve in time in order to go beyond the treatment on the MB routine that has been done in this work.\par
 
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_0_0g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_0_0g.eps}
     \label{fig:subim1}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_3_0g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_3_0g.eps}
     \label{fig:subim2}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_3_5g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_3_5g.eps}
     \label{fig:subim3}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_0g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_4_0g.eps}
     \label{fig:subim4}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_4_5g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_4_5g.eps}
     \label{fig:subim5}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/li_var_vel_5_0g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_5_0g.eps}
     \label{fig:subim6}
@@
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
 \label{fig:grid_li_var_vel}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0084_var_g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_vc_0084_var_g.eps}
     \label{fig:subim21}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_014_var_g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_vc_014_var_g.eps}
     \label{fig:subim22}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0196_var_g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_vc_0196_var_g.eps}
     \label{fig:subim23}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_028_var_g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_vc_028_var_g.eps}
     \label{fig:subim24}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 35mm 15mm 15mm 15mm, clip,width=\textwidth]{figures/li_vc_0336_var_g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_vc_0336_var_g.eps}
     \label{fig:subim25}
@@
 The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
 \label{fig:grid_li_var_g}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_0_0g.eps}
+    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_0_0g.eps}
     \label{fig:subim41}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_3_0g.eps}
+    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_3_0g.eps}
     \label{fig:subim42}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_3_5g.eps}
+    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_3_5g.eps}
     \label{fig:subim43}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_0g.eps}
+    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_0g.eps}
     \label{fig:subim44}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_5g.eps}
+    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_5g.eps}
     \label{fig:subim45}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 30mm 15mm 20mm 15mm, clip,width=\textwidth]{figures/rot_vel_var_vel_5_0g.eps}
+    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_5_0g.eps}
     \label{fig:subim46}
@@
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The dashed lines make reference to the ZAMS.}
+The dashed vertical line makes reference to the ZAMS.}
 \label{fig:grid_rot_vel}

### 2020-05-17 project/octave/plotting_routines.m

@@ -27,6 +27,8 @@ global sz_bot_vrot_col    = 86;
 
 global data_parent_folder = '/home/rcaballeron/MESA/workspace/sun-jupiter-system/Docs/runs/run_paper';
 global filename = '1M_photosphere_history.data';
+%global filename = 'annexB/1M_photosphere_history.data';
+%global filename = 'annex2B/1M_photosphere_history.data';
 global gauss_fields = ['0g'; '2g'; '2.5g'; '3g'; '3.3g'; '3.5g'; '4g'; '4.3g'; '4.5g'; '5g'; '5.5g'];
 %global gauss_fields = ['0g'; '3g'; '3.5g'; '4g'; '4.5g'; '5g'];
 global rotational_vels = ['0crit';'0084crit'; '014crit'; '0196crit'; '028crit'; '0336crit';'029crit';'030crit';'031crit';'0312crit';'0314crit';'032crit';];
@@ -344,10 +346,15 @@ function plot_vel_rot(A, color, width, ytick, axis_limits)
 end
 
 function plot_size_cz(A, color, width, ytick, axis_limits)
+  plot_size_cz2(A, color, width, ytick, axis_limits, 0, '-');
+end
+
+function plot_size_cz2(A, color, width, ytick, axis_limits, offset, linestyle)
   global tick_font_size;
   
   %Plot values
-  plot(A(:,1), A(:,3) .- A(:,4), color, 'linewidth', width);
+  %plot(A(:,1), A(:,3) .- A(:,4), color, 'linewidth', width);
+  plot(A(:,1), A(:,3) .- A(:,4) .+ offset , color, 'linewidth', width, 'linestyle', linestyle);
   %plot(A(:,1), 10 .** A(:,2), color, 'linewidth', width, 'linestyle', '--');
   
   %Axis scales
@@ -359,8 +366,10 @@ function plot_size_cz(A, color, width, ytick, axis_limits)
   %Axis ticks
   axis(axis_limits);
   xticks = get (gca, "xtick"); 
-  %xlabels = arrayfun (@(x) sprintf ("%.2e", x), xticks, "uniformoutput", false); 
-  xlabels = ['1.00e+02';' ';'1.00e+04';' ';'1.00e+06';' ';'1.00e+08';' ';'1.00e+10'];
+  xlabels = arrayfun (@(x) sprintf ("%.2e", x), xticks, "uniformoutput", false); 
[…]

### 2022-12-07 project/src/run_star_extras.f
Improve the angular moment lost routine

@@ -85,9 +85,10 @@
          call star_ptr(id, s, ierr)
          if (ierr /= 0) return
 
-         write(*,*) '*******  Version: 2.3.0'
+         write(*,*) '*******  Version: 2.4.0'
          write(*,*) '*******  09.06.2020: Disk locking & Magnetic braking routines'
-         write(*,*) '*******  11.11.2021: Varaible MLT alpha & magnetic filed strengh'
+         write(*,*) '*******  11.11.2021: Variable MLT alpha & magnetic filed strengh'
+         write(*,*) '*******  28.09.2022: Improve the angular moment lost routine'
          write(*,*) '*******  Roque Caballero'
          
          original_diffusion_dt_limit = s% diffusion_dt_limit
@@ -297,7 +298,8 @@
 
             !Distribute the loss of angular momentum
             !call distribute_j_dot(s, j_dot, sz_info_ptr, mag_brk_jdot)
-            call distribute_with_residual_j_dot(s, j_dot, sz_info_ptr, mag_brk_jdot)            
+            call distribute_with_residual_j_dot(s, j_dot, sz_info_ptr, mag_brk_jdot)   
+            !call distribute_all_zones_with_residual_j_dot(s, j_dot, mag_brk_jdot)
 
             !It happens that s% extra_jdot is longer than s% nz but mag_brk_jdot is just defined
             !for s% nz elements
@@ -584,11 +586,13 @@
          type (star_zone_info), pointer, intent(in) :: sz_info
          real(dp), dimension(:), pointer, intent(out) :: mb_jdot_list
          integer :: k, j
-         real(dp) :: residual_jdot
+         real(dp) :: residual_jdot, final_residual_jdot, delta_residual_jdot
 
          !By default, no lost of angular moment
          mb_jdot_list(:) = 0.0
          residual_jdot = 0d0
+         final_residual_jdot = 0d0
+         delta_residual_jdot = 0d0
       
          do k = sz_info% top_zone, sz_info% bot_zone, 1
             !Here the jdot distribution strategy is defined
@@ -602,21 +606,36 @@
                   (sz_info% d_mass * sz_info% d_radius**2)) / s% dm(k)
 
[…]

### 2022-12-08 project/overleaf/paper2/mblithium.tex
WIP paper II

--- 2020-05-17 project/overleaf/paper.tex
@@
 % Keep the title short and informative.
-\title[Rotation, magnetic braking \& Li abundances]{Study of the effects of magnetic braking on the lithium abundances of the Sun and solar-type stars}
+\title[Rotation, magnetic braking \& Li abundances II]{Study of the effects of magnetic braking on the lithium abundances of the Sun and solar-type stars II}
 
@@
 A. Garc\'ia Hern\'andez,$^{1,2}$\thanks{E-mail: agh@ugr.es}
-A. Ayala$^{2},$\thanks{E-mail: aayala@ugr.es}
 J.~C. Su\'arez$^{1,2}$\thanks{E-mail: jcsuarez@ugr.es}
@@
 % These dates will be filled out by the publisher
-\date{Accepted XXX. Received YYY; in original form ZZZ}
+\date{Accepted 2020 May 29. Received 2020 April 17; in original form 2019 November 30}
 
@@
 \begin{abstract}
-\textit{Goals.} The study of lithium (Li) surface abundance in the Sun and young stellar globular clusters which are seemingly anomalous in present-day scenarios, as well as the influence of rotation and magnetic braking (MB) on its depletion during pre-main sequence (PMS) and main sequence (MS).
-\newline\textit{Methods.} The effects of rotational mixing and of the rotational hydrostatic effects on Li abundances are studied by simulating several grids of PMS and MS rotating and non-rotating models.
-The previous effects are combined with the additional impact of the MB, (with magnetic field intensities ranging between 3.0 and 5.0 G).
-The data obtained from the two sets of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
-\newline\textit{Results.} The surface Li abundance for the Sun like models at the end of the PMS and throughout the MS is decreased when rotational effects are included, i.e.
+The study of lithium (Li) surface abundance in the Sun and young stellar globular clusters which are seemingly anomalous in present-day scenarios, as well as the influence of rotation and magnetic braking (MB) on its depletion during pre-main sequence (PMS) and main sequence (MS).
+In this work, the effects of rotational mixing and of the rotational hydrostatic effects on Li abundances are studied by simulating several grids of PMS and MS rotating and non-rotating models.
+Those effects are combined with the additional impact of the MB (with magnetic field intensities ranging between 3.0 and 5.0 G).
+The data obtained from simulations are confronted by comparing different stellar parameters.
+The results show that the surface Li abundance for the Sun like models at the end of the PMS and throughout the MS decreases when rotational effects are included, i.e.
 the Li depletion rate for rotating models is higher than for non-rotating ones.
-However, this effect is attenuated when the braking effect produced by a magnetic field is present.
-This physical phenomena, generally neglected in stellar evolutionary models, impacts also the effective temperature ($\teff$) of the star and therefore its location in the HR diagram.
-The impact of MB in Li depletion is sensitive to the angular speed and this in turn on the magnetic field intensity: the higher the latter, the lower the Li destruction.
+This effect is attenuated when the MB produced by a magnetic field is present.
+This physical phenomena impacts also the star effective temperature ($\teff$) and its location in the HR diagram.
+The impact of MB in Li depletion is sensitive to the magnetic field intensity: the higher it is, the lower the Li destruction.
 A direct link between the magnetic fields and the convective zone (CZ) size is observed: stronger magnetic fields produce shallower CZ's.
@@
 \section{Introduction} \label{sec_intro}
+Over the last few years, decades, data from different observational campaigns have allowed studies of the rotational evolution of low and medium-mass stars (REFERENCES).
+Among them we find studies (REFERENCES) that focus on modelling the effects of magnetic braking and how its coupling with stellar winds impact the rotational evolution of the star.
+These models prefix a number of parameters in a fixed way, among them the magnetic field strength B (we will come back to this point later).
+This approach is not sufficient to explain the observational data, and interaction with other processes in the angular velocity evolution of low-mass stars is necessary: the distribution of angular momentum loss in the interior of the star along the pre-main sequence (PMS; REFERENCES), and main sequence (MS; REFERENCES), and the star - protostellar disc interaction during the PMS (REFERENCES).
+
 Despite decades of theoretical efforts, a coherent explanation for the Li abundance discrepancies detected in star clusters cannot be found.
@@
 \par
+---NEW---
+In the first paper of this series \citep{Navarro2020}, we developed a model to account for the angular momentum evolution of solar-mass stars.
+The parametric model included a magnetic braking law, and mass-loss prescriptions that relied on \citet{Ud-Doula2008}, and core-envelope differential rotation at PMS and MS.
+The aim of the present study is to extend that model to with a law of interdependence between $\Omega$ and $B$.
+So that, during the PMS, when the star rotates faster, the MB effect is more efficient.
+On the other hand, during the MS, when the start slows down, the MB will become less intensive.
+This approach would establish a self-regulating mechanism over the angular velocity of the star that
+would end up directly influencing the Li evolution.\par
+
+Following a similar approach, we aim the model the evolution of alpha MLT with another law of interdependence between $Teff$ and surface gravity, both of them directly depending on $\Omega$.\par
+
+---NEW---
+
+\section{Data} \label{sec_method}
+Mirar Gaia-ESO Survey: Role of magnetic activity and starspots on pre-main-sequence lithium evolution
 
@@
 This has been implemented as an extension to Modules for Experiments in Stellar Astrophysics stellar evolution code \citep[MESA; ][]{Paxton2011, Paxton2013,Paxton2015, Paxton2018, Paxton2019}.
+
+\subsection{Modelling magnetic intensity} \label{mod_mbi}
 
@@
 
+\subsection{Modelling MLT $\alpha$} \label{mod_mltalpha}
+
 \section{Results} \label{sec_3}
@@
 	\centering
-	\caption{Summary of adopted physics in MESA \citep[based on][]{Choi2016}.}
+	\caption{Summary of adopted physics in MESA \citep[based on][]{Choi2016,Navarro2020}.}
 	\label{tab:phy_mesa}
@@
 		Boundary Conditions & ATLAS12; $\tau$=100 tables + photosphere\\
-		Difussion & Track \isotope[1]{H}, \isotope[2]{He}, \isotope[7]{Li}, \isotope[7]{Be}\\
+		Diffusion & Track \isotope[1]{H}, \isotope[2]{He}, \isotope[7]{Li}, \isotope[7]{Be}\\
 		Rotation & Differential rotation at PMS \& MS\\
-		Convection & MLT + Ledoux, $\alpha_{{\rm MLT}}$ = 1.82\\
-		Overshoot & time-dependent, diffusive, \\ & $f_{{\rm ov,core}}=0.0160$,\\ 
-		& $f_{{\rm ov,sh}}=0.0174$\\
+		Convection & MLT + Ledoux, variable $\alpha_{{\rm MLT}}$\\
+		Overshoot deactivated & $f_{{\rm ov,core}}=0.0$, \& $f_{{\rm ov,sh}}=0.0$\\
 		Semiconvection & $\alpha_{{\rm sc}}=0.1$\\
@@
 		Rotational Mixing & Include SH, ES, GSF, SSI \& DSI\\
+            Mixing Length Theory & $\alpha_{{\rm MLT}}$ variable depending on $\teff$\\ & \& effective surface gravity\\
 		Magnetic Effects & Magnetic braking based on idealized \\ & monopole field\\
-		Magnetic Field & B(G) variable between [3.0 - 5.0]\\
+		Magnetic Field & B(G) variable, depending on $\teff$ \&  $\Omega$\\
 		Mass Loss & activated, $\Dot{M}_{{\rm max}} = 10^{-3} \: \msun \: yr^{-1}$\\
@@
 
-\textbf{Notice the jagged appearance in Figure~\ref{fig:cz_vc_028_var_b_z1_new}.
+Notice the jagged appearance in Figure~\ref{fig:cz_vc_028_var_b_z1_new}.
 We suspected a numerical origin for the wriggles.
@@
 Indeed, the size of the zigzag features is around $10^{-3}$,  which is far below the convection timescale and Li abundances measured in the stellar surface.
-It is thus save to neglect such effects by using the original time/space resolution.}
+It is thus save to neglect such effects by using the original time/space resolution.
 
@@
 	\includegraphics[trim = 15mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g_z1_new.eps}
-    \caption{\textbf{Similar to Figure \ref{fig:cz_vc_028_var_b} but now showing in detail the size of the convection zone in the models from ZAMS up to TAMS.
+    \caption{Similar to Figure \ref{fig:cz_vc_028_var_b} but now showing in detail the size of the convection zone in the models from ZAMS up to TAMS.
 As the intensity of the magnetic field increases, the size of the convective zone decreases.
-The first three pointed curves (below 0.26), have been artificially shifted downwards to illustrate the impact of the temporal (2x in time step and number of shells, red-dotted line) and spatial (2x and 4x number of shells, black and green-dotted lines, respectively) resolution on the numerical solution for the 5G case.}}
+The first three pointed curves (below 0.26), have been artificially shifted downwards to illustrate the impact of the temporal (2x in time step and number of shells, red-dotted line) and spatial (2x and 4x number of shells, black and green-dotted lines, respectively) resolution on the numerical solution for the 5G case.}
     \label{fig:cz_vc_028_var_b_z1_new}

### 2022-12-08 project/octave/plotting_routines.m
Adjustments for paper II simulation and visualization

@@ -60,6 +60,7 @@ global clusters_log_g_col = 4;
 global clusters_e_log_g_col = 5;
 global clusters_fe_h_col = 6;
 global clusters_li_col = 8;
+global clusters_e_li_col = 9;
 global clusters_age_col = 10;
 global clusters_err_up_age_col = 7;
 global clusters_err_down_age_col = 8;
@@ -149,7 +150,10 @@ global gauss_fields = ['0g'; '2g'; '2.5g'; '3g'; '3.3g'; '3.5g'; '4g'; '4.3g'; '
 global rotational_vels = ['0crit';'0084crit'; '014crit'; '0196crit'; '028crit'; 
   '0336crit';'029crit';'030crit';'031crit';'0312crit';'0314crit';'032crit';'0336crit_alpha';
   '11crit';'105crit';'1075crit';'1125crit';'1025crit';
-  '115crit';'10crit';'0975crit';'095crit';'0925crit'];
+  '115crit';'10crit';'0975crit';'095crit';'0925crit';'125crit';
+  '135crit';'1175crit';'12crit';'1225crit';'1275crit';
+  '13crit';'1325crit';'1375crit';'14crit';'1425crit';
+  '145crit';'1475crit';'15crit';'1525crit';'155crit'];
 %dl -> disk locking
 global dl_rotational_vels = ['0crit_dl';'0084crit_dl'; '014crit_dl'; '0196crit_dl'; 
   '028crit_dl'; '0336crit_dl';'029crit_dl';'030crit_dl';'031crit_dl';
@@ -195,7 +199,22 @@ global idx_10crit  = 20;
 global idx_0975crit  = 21;
 global idx_095crit  = 22;
 global idx_0925crit  = 23;
-
+global idx_125crit  = 24;
+global idx_135crit  = 25;
+global idx_1175crit  = 26;
+global idx_12crit  = 27;
+global idx_1225crit  = 28;
+global idx_1275crit  = 29;
+global idx_13crit  = 30;
+global idx_1325crit  = 31;
+global idx_1375crit  = 32;
+global idx_14crit  = 33;
+global idx_1425crit  = 34;
+global idx_145crit  = 35;
+global idx_1475crit  = 36;
+global idx_15crit  = 37;
+global idx_1525crit  = 38;
+global idx_155crit  = 39;
 
 
 %The following with mlt=var and disk locking
[…]

### 2023-04-01 project/octave/plotting_routines.m
Minor changes

@@ -2127,7 +2127,7 @@ function age_vs_alpha_mlt(gauss_fields, rotational_vels, is_var_vel, ytick, axis
   full_path = strcat(data_parent_folder, '/', sub_folder, '/', filename);
   zams = calculate_ZAMS(full_path);
   line("xdata",[zams,zams], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 3, "linestyle", "--", "color", "k");
-
+  
   grid on;
   l = legend(labels, "location", leg_loc);
 
@@ -2165,7 +2165,9 @@ function omegs_vs_mag_field(gauss_fields, rotational_vels, show_limits, ytick, a
   global bf_max_col;
   global bf_star_col;
   global alfven_r_col;
- 
+  global sun_vel_rot;
+  global sun_gauss_field;  
+  global sun_age;
   
   global header_lines;
   global colors;
@@ -2206,6 +2208,11 @@ function omegs_vs_mag_field(gauss_fields, rotational_vels, show_limits, ytick, a
   full_path = strcat(data_parent_folder, '/', sub_folder, '/', filename);
   zams = calculate_ZAMS(full_path);
   line("xdata",[zams,zams], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 3, "linestyle", "--", "color", "k");
+  
+  % Plot sun reference
+  plot(sun_age, sun_vel_rot, '*', 'markersize', 15, 'linewidth', 2, 'color', [0.5,0.1,0.8]);
+  plot(sun_age, sun_gauss_field, 's', 'markersize', 10, 'color', [0.5,0.1,0.8], 'markerfacecolor', [0.5,0.1,0.8]);
+  
 
   grid on;
   l = legend(labels, "location", leg_loc);
@@ -3201,7 +3208,7 @@ function plot_omega_vs_mag_field_XG(rot_vels, show_limits, idx)
   global idx_X_G;
   
   %omegs_vs_mag_field_plots(gauss_fields(idx_X_G,:), rot_vels, true, 10, [0.5, 50, 1, 1000], 'northwest', 'Magnetic field intensity  & var. rotational velocity', 'mag_field_var_vel_g');
[…]

### 2023-04-01 project/overleaf/paper2/mblithium.tex
Add paper 2

--- 2020-05-17 project/overleaf/paper.tex
@@
 % These dates will be filled out by the publisher
-\date{Accepted XXX. Received YYY; in original form ZZZ}
+\date{Accepted 2023 XXX XX. Received 2023 XXXX XX; in original form 2023 XXXX XX}
 
 % Enter the current year, for the copyright statements etc.
-\pubyear{2019}
+\pubyear{2023}
 
@@
 \begin{abstract}
-\textit{Goals.} The study of lithium (Li) surface abundance in the Sun and young stellar globular clusters which are seemingly anomalous in present-day scenarios, as well as the influence of rotation and magnetic braking (MB) on its depletion during pre-main sequence (PMS) and main sequence (MS).
-\newline\textit{Methods.} The effects of rotational mixing and of the rotational hydrostatic effects on Li abundances are studied by simulating several grids of PMS and MS rotating and non-rotating models.
-The previous effects are combined with the additional impact of the MB, (with magnetic field intensities ranging between 3.0 and 5.0 G).
-The data obtained from the two sets of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
-\newline\textit{Results.} The surface Li abundance for the Sun like models at the end of the PMS and throughout the MS is decreased when rotational effects are included, i.e.
-the Li depletion rate for rotating models is higher than for non-rotating ones.
-However, this effect is attenuated when the braking effect produced by a magnetic field is present.
-This physical phenomena, generally neglected in stellar evolutionary models, impacts also the effective temperature ($\teff$) of the star and therefore its location in the HR diagram.
-The impact of MB in Li depletion is sensitive to the angular speed and this in turn on the magnetic field intensity: the higher the latter, the lower the Li destruction.
-A direct link between the magnetic fields and the convective zone (CZ) size is observed: stronger magnetic fields produce shallower CZ's.
-This result suggests that MB effect must be taken into consideration during PMS if we aim to reproduce Li abundances in young clusters.
+The study of lithium (Li) surface abundance in the Sun and young stellar globular clusters which are seemingly anomalous in present-day scenarios, as well as the influence of rotation and magnetic braking (MB) on its depletion during pre-main sequence (PMS) and main sequence (MS).
+In this work, the effects of rotational mixing and of the rotational hydrostatic effects on Li abundances are studied by simulating several grids of PMS and MS rotating models.
+Those effects are combined with the additional impact caused by the MB effect.
+Its intensity evolves as the star does i.e.
+not remaining fixed through the simulations.
+The data obtained from simulations are confronted with observational data measured on young open clusters collected by the Gaia-ESO Survey (GES).
+The results show how faster rotating models destroy more Li in the late PMS and during the MS, and how the effects of magnetic braking mitigate this destruction.
+The interconnection between the star's effective temperature, surface gravity and rotational velocity, as well as its chemical composition, determine the strength of the magnetic field and impact the effectiveness of the convective motions involved in the destruction of Li.
+The present work documents an approach in which neither the magnetic field strength nor the MLT parametrisation remains fixed throughout the evolution of the star, and which ends up providing results that are compatible with observational data for Li abundances of solar siblings hosted in young clusters.
+
 \end{abstract}
@@
 \section{Introduction} \label{sec_intro}
-Despite decades of theoretical efforts, a coherent explanation for the Li abundance discrepancies detected in star clusters cannot be found.
-In young and old clusters these differences have been documented for stars in both PMS and MS  evolutionary stages.
-Additionally, these theoretical models are not able to explain the abundances detected in the late stages of the MS \citep{Tschape2001}.\par
+The angular momentum loss (AML) is considered to be governed by magnetically-coupled stellar winds which exert a braking force on the star that eventually causes a loss of angular momentum \citep{UdDoula2002}.
 
-Standard models, models that include convection only as a mixing process and do not consider other transport options like diffusion or angular momentum loss (AML), have been mainly involved in the elaboration of these predictions \citep{Sestito2005}.
-Li is destroyed at a temperature $\tli \approx 2.5 x 10^6\; K$ when a Li atom collides with a proton producing two He atoms.
-That takes place during proton-proton type II reactions (P-P II), and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches  $\tli$.
-The Sun in particular and the solar-type stars in general are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $\tli$ \citep{Iben1965}.
-This depletion stops during the approach to the zero-age MS (ZAMS) when the convection zone retreats and the temperature at BCZ is cooler than $\tli$.
-In the standard models only the mass and the initial chemical composition determine at what distance from the stellar surface $\tli$ temperature is reached.
-Therefore stars with similar mass belonging clusters are expected to reach the ZAMS with equal surface Li abundances.
-Furthermore, they should also show a very similar Li evolution until their approach to the terminal-age MS (TAMS).
-During this period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the envelope, from its lower limit till the star surface.
-However, different abundances of Li have been observed for different stellar populations \citep[see][and references therein]{Somers2014}.\par
+Over the last few years, decades, data from different observational campaigns have allowed studies of the rotational evolution of low and medium-mass stars \citep{Hartman2010,Gallet2013,Bouvier2016}. 
+Among them we find studies \citep{Ud-Doula2008,Cranmer2011,Gallet2015,Amard2016} that focus on modelling the effects of magnetic braking (MB) and how its coupling with stellar winds impact the rotational evolution of the star.
+These models rely on a number of parameters which are fixed, among them the magnetic field strength B (we will come back to this point later).
+This approach is not sufficient to explain the observational data, and interaction with other processes in the angular velocity evolution of low-mass stars is necessary: the distribution of angular momentum loss in the interior of the star along the pre-main sequence (PMS) and main sequence (MS) \citep{Charbonnel2005, Eggenberger2008, Eggenberger2009, Navarro2020}, and the star - protostellar disc interaction during the PMS \citep{Bouvier2008,Gallet2013,Eggenberger2012,Zanni2012,Bouvier2016}.\par
 
-Theoretical models take initial Li abundance as an input parameter and only describe how it is exhausted.
-Therefore, in order to make an accurate estimate of the initial abundance of Li, it is a prerequisite to be able to compare observations and models beforehand.
-Our Sun represents a unique exception, since it allows us to know the current abundance of this element in its photosphere, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex$ \citep{Jeffries2004}, where $A(\isotope[7]{Li})$ is defined
+As noted in our earlier work, discrepancies in Li abundances have been detected in young cluster for which there is as yet no coherent explanation \citep[see][and references therein]{Navarro2020}.
+Li is destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches $\tli \approx 2.5 x 10^6\; K$.
+The solar-type stars are known to have a convection zone (CZ) that covers much of the stellar radius during the PMS.
+In this way, the convective movements that occur in its interior manage to drag the Li to areas in which the BCZ limit exceeds the $\tli$ and Li is burned \citep{Iben1965}.
+This process slows down as the star approaches the zero-age MS (ZAMS) when the convection zone retreats and moves away from areas where the internal temperature is high enough to destroy the Li.\par
+
+The impact of rotation both on PMS and Li depletion for solar-type stars has already extensively debated in the past \citep{Pinsonneault1997,Jeffries2004,Somers2014} and revised more recently on the basis of the availability of more accurate measures \citep{Gallet2013,Bouvier2016, Bouvier2018, Franciosini2022}.
+Of particular interest for this paper are the Li observational data measured on young open clusters.
+The Gaia-ESO Survey (GES) \citep{Gilmore2012,Randich2013,Randich2022} has produced very high quality spectroscopy records of around 110 000 Milky Way stars, in the field and in open clusters, down to magnitude 19, It has delivered good-quality radial velocities and stellar parameters for a large fraction of its unique target stars.
+Elemental abundances were derived for up to 31 elements for targets observed with the Fibre Large Array Multi Element Spectrograph (FLAMES) and the UV-Visual Echelle Spectrograph (UVES).
+Lithium abundances are delivered for about 1/3 of the sample.\par
+
+As indicated by its name, GES intends to complement the data about parallaxes and proper motions from the Gaia satellite \citep{Mignard2005} with extremely precise information on radial velocities (RVs), Li, and chemistry in general.
+The first Gaia data release (GDR1) \citep{Brown2016} dated on 2016 and contained information on the first 14 months of operation.
+Gaia data release 2 (GDR2) \citep{Brown2018}, released on 2018, contained positions, parallaxes, and proper motions for about 1.3 billion sources, together with photometry.
+The third data release of Gaia has been divided in two parts: the Gaia early data release 3 (GEDR3) \citep{Brown2021} on 2020, and the Gaia data release 3 (GDR3) \citep{Brown2022} on 2022.
+They convey updated and more precise positions, parallaxes, proper motions, and photometry, spectra, classification and astrophysical parameters.
+The Gaia-ESO Survey will supplement the Gaia data for a significant subset of its targets, so that Gaia precision astrometry can be coupled with Gaia-ESO precision radial velocities and chemical abundances.
+In addition, a large fraction of the Gaia-ESO initiative is dedicated to the study of a large sample of open clusters (OCs).\par
+
+GES collected data are very handy for our research because they conform a large data set of stellar parameters and Li abundances for several OCs, covering a large age range, from a few million to several billion years \citep{Dumont2021}.
+The analysis of surface Li abundances in young OCs of different ages can provide important constraints on stellar evolutionary models.
+As mentioned, GES enhances Gaia observation and several data releases has been made available.
+Of particular interests for this paper, it is the  latest release (DR5.0).
+The DR5.0 release includes all the astrophysical parameters derived by the Gaia-ESO consortium for 114 324 stars.
+In detail the catalogue provides \citep{Gilmore2022}:
+\begin{itemize}
+    \item The radial and projected RVs.
+    \item Stellar parameters (effective temperature, surface gravity and metallicity).
+    \item Abundances of several elements, among them Li.
+    \item Specific parameters for tracing accretion and activity in young stars.
+    \item Cluster probability membership.  
+\end{itemize}
+
+\section{Data} \label{sec_method}
+The datasets provided by the Gaia mission in GEDR3 and GDR3, as well as those published by GES in DR5.0 form the main references used in this work.
+In general, we focus on the observations made on OCs, and particularly on those analysed in the \cite{Bragaglia2022} and \cite{Randich2022} papers.
+On the basis of the results presented in these papers, a set of 65 OCs are selected as the reference dataset for our work.
+
+\begin{table}
+	\centering
+	\caption{Information on the pre-selected OCs.}
+	\label{tab:oc_list}
+	\begin{tabular}{l l l l} 
+		\hline
+		Cluster & GES\_FLD & age (Gyr) & [Fe/H]\\
+		\hline
+            25 Ori & 25\_Ori & 0.013 & 0 \\
+            ASCC 50 & Assc50 & 0.011 & 0.02 \\
+            Berkeley 21 & Br21 & 2.138 & -0.21 \\
+            Berkeley 22 & Br22 & 2.455 & -0.26 \\
+            Berkeley 25 & Br25 & 2.455 & -0.25 \\
+            Berkeley 30 & Br30 & 0.295 & -0.13 \\
+            Berkeley 31 & Br31 & 2.818 & -0.29 \\
+            Berkeley 32 & Br32 & 4.898 & -0.31 \\
+            Berkeley 36 & Br36 & 6.761 & -0.15 \\
+            Berkeley 39 & Br39 & 5.623 & -0.14 \\
+            Berkeley 44 & Br44 & 1.445 & 0.22 \\
+            Berkeley 73 & Br73 & 1.413 & -0.26 \\
+            Berkeley 75 & Br75 & 1.738 & -0.34 \\
+            Berkeley 81 & Br81 & 1.148 & 0.22 \\
+            Blanco 1 & Blanco1 & 0.105 & -0.03 \\
+            Chamaeleon I & Cha\_I & 0.001 & -0.03 \\
+            Collinder 197 & Col197 & 0.014 & 0.03 \\
+            Collinder 69 & lam\_Ori & 0.013 & -0.09 \\
+            Czernik 24 & Cz24 & 2.692 & -0.11 \\
+            Czernik 30 & Cz30 & 2.884 & -0.31 \\
+            ESO 92-05 & ESO92\_05 & 4.467 & -0.29 \\
+            Haffner 10 & Haf10 & 3.802 & -0.1 \\
+            IC 2391 & IC2391 & 0.029 & -0.06 \\
+            IC 2602 & IC2602 & 0.036 & -0.06 \\
+            IC 4665 & IC4665 & 0.033 & 0.01 \\
+            Loden 165 & Loden165 & 3 &  \\
+            Messier 67 & M67 & 3.981 & -0.02 \\
+            NGC 2141 & NGC2141 & 1.862 & -0.04 \\
+            NGC 2158 & NGC2158 & 1.549 & -0.15 \\
+            NGC 2232 & NGC2232 & 0.018 & -0.03 \\
+            NGC 2243 & NGC2243 & 4.365 & -0.45 \\
+            NGC 2244 & NGC2244 & 0.004 & -0.04 \\
+            NGC 2264 & NGC2264 & 0.003 & -0.1 \\
+            NGC 2355 & NGC2355 & 1 & -0.13 \\
+            NGC 2420 & NGC2420 & 1.698 & -0.15 \\
+            NGC 2425 & NGC2425 & 2.399 & -0.13 \\
+            NGC 2451A & NGC2451 & 0.035 & -0.08 \\
+            NGC 2451B & NGC2451 & 0.041 & -0.02 \\
+            NGC 2516 & NGC2516 & 0.24 & -0.04 \\
+            NGC 2547 & NGC2457 & 0.032 & -0.03 \\
+            NGC 3293 & NGC3293 & 0.01 & -0.08 \\
+            NGC 3532 & NGC3532 & 0.398 & -0.01 \\
+            NGC 3766 & NGC3766 & 0.023 & -0.12 \\
+            NGC 4815 & NGC4815 & 0.372 & 0.08 \\
+            NGC 6005 & NGC6005 & 1.259 & 0.22 \\
+            NGC 6067 & NGC6067 & 0.126 & 0.03 \\
+            NGC 6253 & NGC6253 & 3.246 & 0.33 \\
+            NGC 6259 & NGC6259 & 0.269 & 0.18 \\
+            NGC 6281 & NGC6281 & 0.513 & -0.04 \\
+            NGC 6405 & NGC6405 & 0.035 & -0.02 \\
+            NGC 6530 & NGC6530 & 0.002 & -0.02 \\
+            NGC 6633 & NGC6633 & 0.692 & -0.03 \\
+            NGC 6649 & NGC6649 & 0.071 & -0.08 \\
+            NGC 6705 & NGC6705 & 0.309 & 0.03 \\
+            NGC 6709 & NGC6709 & 0.191 & -0.02 \\
+            NGC 6802 & NGC6802 & 0.661 & 0.14 \\
+            Pismis 15 & Pismis15 & 0.871 & 0.02 \\
+            Pismis 18 & Pismis18 & 0.575 & 0.14 \\
+            Ruprecht 134 & Rup134 & 1.66 & 0.27 \\
+            Trumpler 14 & Trumpler14 & 0.003 & -0.01 \\
+            Trumpler 20 & Trumpler20 & 1.862 & 0.13 \\
+            Trumpler 23 & Trumpler23 & 0.708 & 0.2 \\
+            Trumpler 5 & Trumpler5 & 4.266 & -0.35 \\
+            Gamma Vel & gamma2\_vel & 0.02 & -0.02 \\
+            Rho Oph & Rho\_Oph & 0.001 & 0.03 \\
+		\hline
+	\end{tabular}
+\end{table}
+
+\subsection{The selection of clusters}
+GES DR5.0 provides information on 114 324 stars, not all of which correspond to OCs, see \cite{Gilmore2022} for details.
+Of interest for the present work are those identified as belonging to OCs.
+For them we proceed to perform a first filtering on the GES\_TYPE field to identify those records that meet the condition:
+\begin{itemize}
+    \item GE\_CL: observed by GES, OC programme field
+    \item GE\_SD\_OC: observed by GES, standard field OC
+    \item AR\_CL: ESO Archive observation, OC programme field
+    \item AR\_SD\_OC: Archive observation, standard field OC
+\end{itemize}
+
+As a result of this first filtering, the number of selected records is reduced to 43 299.
+On this subset of data we perform another filtering, this time aimed at selecting those that offer information on the following attributes:
+\begin{itemize}
+    \item Metallicity
+    \item Li abundance
+    \item GES field identifier
+    \item Effective temperature
+    \item Surface gravity
+\end{itemize}
+
+This second filtering reduces the population of stars to be considered to a total of 21 125 records.
+Finally, a third filtering is carried out to select those stars belonging to the OCs listed in Table \ref{tab:oc_list}, giving us a final study population of 20 913 stars distributed in 65 OCs.\par
+
+\subsection{The selection of closest solar siblings}
+Once we have pre-selected the stars belonging to the OCs listed in the previous chapter, our interest focus on identifying those that are most similar to our Sun i.e.
+solar siblings.
+Finding such solar twins is a difficult task since they are disseminated across our Galaxy.
+Finding them would help us to figure out the origin of our Sun and and gain information about what conditions prevailed in the OCs in which they were born.\par
+
+Several attempts have been made to find solar siblings in the past \citep[see][and references therein]{Adibekyan2018}.
+In most of them, the search method is oriented to search the candidates based on its kinematic properties and then their metallicities and chemical abundances are compared with those of our Sun.\par
+
+In our case we extend the selection criteria by adding additional constraints: the surface gravity and age of the star.
+These two new parameters represent a fine-tuning of the search process for sibling candidates to our Sun.
+We use the age of the OCs reported in Table \ref{tab:oc_list} and the data obtained from the set of simulations performed using Modules for Experiments in Stellar Astrophysics stellar evolution code \citep[MESA; ][]{Paxton2011, Paxton2013,Paxton2015, Paxton2018, Paxton2019}.
+Our models extend the investigations documented in \cite{Navarro2020} on the influence of rotation and magnetic braking on the star Li depletion during the pre-main sequence (PMS) and the main sequence (MS).
+The data obtained from the simulations are confronted by comparing different stellar parameters obtained by Gaia and GES.
+This allows us to analyse, validate and refine our models against a very large set of candidates.
+The end result of this process is to obtain a subset of candidate sibling stars of our Sun that help us to support (or refute) our line of investigation of how the Li abundance behaves throughout the evolution of solar-type stars.\par
+
+From the approximately 21 000 identified stars, we proceed to select those with metallicity ($\feh$), effective temperature ($\teff$), and surface gravity ($\gsurf$) within the following limits:
+\begin{itemize}
+    \item $0.05 \, \dex <= \feh <= 0.05 \, \dex$
+    \item $50.0 \, \Kelvin <= \teff <= 50.0 \, \Kelvin$
+    \item $0.05 \, \dex <= \gsurf <= 0.05 \, \dex$
+\end{itemize}
+
+\section{Method} \label{sec_method}
+The AML is considered to be governed by magnetically-coupled stellar winds which exert a braking force on the star that eventually causes a loss of angular momentum.
+In addition, throughout the star's evolution, directly observable parameters such as its rotational velocity, effective temperature, surface gravity or abundance of chemical elements undergo variations.
+These parameters have an influence, either directly or indirectly, on the angular momentum loss caused by the star's magnetic field.
+Therefore, assuming that the magnetic field strength of a star remains constant throughout its evolution is a necessary simplification in some scenarios but does not correspond to observations.\par
+
+In the remainder of this paper, we describe a semi-empirical approach for the calculation of the AML as a result of the braking caused by a magnetic field of varying strength.
+The magnetic field evolves according to the $\teff$ and angular velocity ($\Omega$) of the star.
+This has been implemented as an extension to MESA.
+
+\subsection{Modelling magnetic intensity} \label{mod_mbi}
+The variable magnetic field strength modelling approach is supported by the papers \cite{Cranmer2011} and \cite{Gallet2013}.
+We start by enumerating the most relevant aspects and assumptions made in modelling the evolution of magnetic field intensity.\par
+
+A relevant parameter to characterize the influence of a given magnetic field on the stellar wind is $\Omega$.
+Assuming that the magnetic field is generated by a dynamo, its strength is proportional to some power of $\Omega$,
+\begin{ceqn}
+\begin{equation}
+    f_*B_* \propto \Omega_*^b \label{eq:mf_strenght}
+\end{equation}
+\end{ceqn}
+where $b$ is the dynamo exponent, $B_*$ is the strength of the magnetic filed, and $f_*$ is the filling factor, i.e. the fraction of the star surface which is magnetized \citep{Gallet2013}.
+
+On the other hand, $B_*$ is proportional to the equipartition magnetic field strength,
+\begin{ceqn}
+\begin{equation}
+    B_* \approx 1.13 \Beq \label{eq:mf_bstar}
+\end{equation}
+\end{ceqn}
+where $\Beq$ is defined as 
+\begin{ceqn}
+\begin{equation}
+    \Beq = \sqrt{8\upi P_*} = \sqrt{\frac{8\upi\rho_* \boltzmann \teff}{\mu\massH}}\label{eq:mf_beq}    
+\end{equation}
+\end{ceqn}
+where $\rho_*$ is the photospheric density, $\boltzmann$ is Boltzmann's constant, $\massH$ is the mass of a hydrogen atom, and $\mu$ the mean atomic weight \citep{Cranmer2011}.
+
+As indicated in \cite{Cranmer2011} $\mu$ can be estimated using the OPAL plasma equations of state\footnote{\url{https://opalopacity.llnl.gov/EOS_2005/}},
+\begin{ceqn}
+\begin{equation}
+    \mu \approx \frac{7}{4} + \frac{1}{2} \; {\rm tanh}\Bigg(\frac{3500-\teff}{600}\;\Bigg) \label{eq:mf_atom_weight}
+\end{equation}
+\end{ceqn}
+Once we are in a position to calculate $\Beq$, we can calculate $B_*$.\par
+
+\cite{Cranmer2011} found that the magnetic field strength is mainly influenced by $f_*$ rather than by the rotation period of the star.
+That is, variations in the star's angular velocity do not significantly alter its magnetic field strength.
+In addition, $f_*$ has a strong dependence on the Rossby number ($\rossby$).
+In order to calculate $\rossby$ of a given star, it is necessary to know the turn over time ($\turnover$),
+\begin{ceqn}
+\begin{equation}
+    \rossby = \frac{\prot}{\turnover} \label{eq:mf_rossby}
+\end{equation}
+\begin{ceqn}
+\begin{equation}
+    \turnover = 314.24\;exp\left[-\Bigg(\frac{\teff}{1952.5}\Bigg)-\Bigg(\frac{\teff}{6250}\Bigg)^{18} \;\right]+0.002 \label{eq:mf_turnover}
+\end{equation}
+
+\end{ceqn}
+\cite{Cranmer2011} Crammer provides two different settings for $f_*$ which define respectively the upper and lower limits for the mean surface magnetic field strength ($f_*B_*$),
 
@@
 \begin{align}
-    A(\isotope[7]{Li}) &= log(N_{\isotope[7]{Li}} / N_{\isotope[1]{H}}) + 12
-    \label{eq:A_Li}
+     \fmin &= \frac{0.5}{(1+(x/0.16)^{2.6})^{1.3}} \label{eq:fmin}\\
+     \fmax &= \frac{1}{1+(x/0.31)^{2.5}} \label{eq:fmax}
 \end{align}
 \end{ceqn}
+where x = $\rossby$/$\rossbysun$, and $\rossbysun$ = 1.96.
 
-and its initial abundance of $A(\isotope[7]{Li}) = 3.34 \, dex$ is obtained from meteorite measurements \citep{Randich2006}.
-For other stars the initial Li abundance is found to be in the interval $3.0 \, dex < A(\isotope[7]{Li}) < 3.4 \, dex$.
-This has been estimated from  \citet{Randich2006}.\par
+Finally, we are able to calculate the filling factor for a star by applying the adjustment made by \cite{Gallet2013},
+\begin{ceqn}
+\begin{equation}
+     \fstar = \frac{0.55}{(1+(x/0.16)^{2.3})^{1.22}} \label{eq:fstart}
+\end{equation}
+\end{ceqn}
 
-The impact of rotation both on PMS and Li depletion for solar-type stars has already extensively debated in the past \citep{Pinsonneault1997,Jeffries2004,Somers2014} and revised more recently on the basis of the availability of more accurate measures \citep{Bouvier2016}.
-However, these previous studies focused mainly on the hydrostatic effects and did not consider the influence of a coupling between the stellar magnetic field and its possible spin-down effect.
-AML has a direct influence in the mixing processes.
-It can be produced by a number of relevant mechanisms: mass loss, magnetic fields and gravity waves (g-modes).
-Gravity waves \citep{Charbonnel2005} and magnetic fields \citep{Eggenberger2009} transmit angular momentum (AM) much more effectively than inducing mixing \citep{Denissenkov2007}.
-As a consequence, the amount of differential rotation between the radiative and the convective zones of the star is reduced.
-Additionally, magnetic fields would also originate in the limits between these zones, in the so-called tachoclines \citep{Aschwanden2014, Guerrero2016}, which interact with the solar wind particles that are forced to co-rotate with it.
-This slows down the star, effect which is known as magnetic braking (MB).
-\par
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/mag_field_limits_var_vel_g3.pdf}
+    \caption{The evolution of magnetic field intensity and its upper and lower limits, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
+The model include PMS rotation with $\oomegac$ = 0.1425.
+The solid lines represent the magnetic field strength, while the dotted lines represent the angular evolution of the star.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:mag_field_limits_var_vel_g3}
+\end{figure}
 
-Today, standard models are not able to replicate the observed values of Li abundance on stellar surfaces.
-This might happen because certain physical mechanisms that influence the destruction of Li are either being modelled improperly or simply not being taken into account, e.g.
-the magnetic braking.
-In this paper we study the interactions between rotation and magnetic fields when it comes to discuss the AM distribution.
-\par
-
-\section{Method} \label{sec_method}
-
-In the remainder of this paper, we describe a semi-empirical approach for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.
-This has been implemented as an extension to Modules for Experiments in Stellar Astrophysics stellar evolution code \citep[MESA; ][]{Paxton2011, Paxton2013,Paxton2015, Paxton2018, Paxton2019}.
+With this formalism we are able to calculate the mean magnetic field as a time-dependent function of $\rho$, $\teff$, and $\Omega$.
+The time values for these three variables will be obtained from the different simulations carried out with the help of MESA.
 
 \subsection{Modelling magnetic braking} \label{mod_mb}
-We start by enumerating the most relevant aspects and assumptions made in modelling the evolution of rotation, magnetic braking and angular momentum in MESA.
-These will be considered for the calculation of the AML as a result of the torque applied by a magnetically-coupled stellar winds.\par
+Once we have characterised how the magnetic field strength evolves, we need to model how the loss of angular momentum occurs.
+In \cite{Navarro2020} we use a model applicable to magnetic fields of non-variable intensity and of few Gauss ($1.0\,\Gauss \leq B \leq 10.0\,\Gauss$).
+As shown in Figure \ref{fig:mag_field_limits_var_vel_g3}, the simulations yield magnetic field values that show maximum intensities of around $1.0K\,\Gauss$, making the use of the model employed previously non viable.
 
-A relevant parameter to characterize the influence of a given magnetic field on the stellar wind is the denominated wind confinement magnetic parameter ($\eta_*$).
-It represents a energy ratio (Eq.~\ref{eq:wind_conf}) and defines a characteristic parameter for the relative effectiveness of the magnetic fields in circumscribing and/or channeling the wind outflow \citep{UdDoula2002}.\par
+Let's start by enumerating the most relevant aspects and assumptions made in modelling the evolution of rotation, magnetic braking and angular momentum.
+If we assume a spherical outflow, the torque applied by a magnetically-coupled stellar winds $\torquewind$ is dictated by,
+\begin{ceqn}
+\begin{equation}
+    \torquewind \propto \Omega_* \; \mwind \; \ralfven^{2} \label{eq:mb_torque}
+\end{equation}
+\end{ceqn}
+where $\mwind$ is the mass loss rate, and $\ralfven$ the averaged value of the Alfvén radius.\par
+Stars with similar initial masses but different mass loss ($\Dot{M}$) ratios will end up evolving very differently.
+The ionized particles carried by the solar wind not only contribute to the mass loss but also to the loss of kinetic energy that is deposited in the interstellar medium.
+Given a star with a spherically symmetric wind, $\mwind$ is characterized by the following expression:
 
@@
 \begin{equation}
-    \eta_* = \frac{B^{2}/8\upi}{\rho\nu^2/2} \label{eq:wind_conf}
+    \mwind = 4\upi r^2\rho\nu \label{eq:mass_loss}
+\end{equation}
+\end{ceqn}
+where $r$ is the stellar radius, $\rho$ the mass density, and $\nu$ the stellar wind velocity.
+
+It has been observed that strongly magnetic intermediate-mass stars typically have rotation rates much slower than other stars in their parent population \citep{Mathys2006}.
+In those stars, the magnetic fields interacts with the mass loss, where the Alfv\'{e}n radius plays an important role.
+$\ralfven$ is defined as the point in which the magnetic field energy density and the kinetic energy density are balanced.
+In case that $\ralfven$ is greater than the stellar radius, then the wind flow will have to follow the magnetic field.
+As a consequence, the material leaves the stellar surface with a higher specific AM, as the co-rotation radius has increased and it roughly corresponds to $\ralfven$ which can be expressed as \citep{Matt2012}
+\begin{ceqn}
+\begin{equation}
+    \ralfven = K_1\left[\frac{\Bp^{2}\;R_*^{2}}{\mwind\;\sqrt{K_2^2\vesc^2 + \Omega_*^2R_*^{2}}\ }\right]^{m}R_*  \label{eq:mb_ralfven}
+\end{equation}
+\end{ceqn}
+where $\Bp$ is the magnetic field intensity at the stellar surface, $\vesc$ is the escape velocity, $m = 0.1675$, $K_1 = 1.30$, and $K_2 = 0.0506$ \citep{Gallet2013}.
+\begin{ceqn}
+\begin{equation}
+\vesc = \sqrt{\frac{2\,G\,\mstar}{\rstar}} \label{eq:vesc}
 \end{equation}
@@
 
-where $B^{2}/8\upi$ is the magnetic field energy density, $\frac{1}{2}\rho\nu^{2}$ the kinetic energy density, $B$ the magnetic field intensity, $\rho$ the mass density, and $\nu$ the stellar wind velocity.\par
+Finally, we have that the AML can be calculated as follows:
+\begin{ceqn}
+\begin{equation}
+ \Dot{J} = \Omega_* \; \mwind \; \ralfven^{2} \label{eq:j_dot}
+\end{equation}
+\end{ceqn}
+where $\mwind$ is the mass loss rate. This expression is implemented in MESA based on values directly exposed during the simulations.
 
-Stars with similar initial masses but different mass loss ($\Dot{M}$) ratios will end up evolving very differently.
-The ionized particles carried by the solar wind not only contribute to the mass loss but also to the loss of kinetic ($K_e$) energy that is deposited in the interstellar medium.
-Given a star with a spherically symmetric wind, $\Dot{M}$ is characterized by the following expression:
+\subsection{Modelling MLT $\alpha$} \label{mod_mltalpha}
+The mixing-length theory (MLT) introduced by Böhm-Vitense (BV) has been commonly adopted to model stellar convection in stellar evolutionary codes, including MESA.
+The most important free parameter in this theory is the mixing length $l$ which is defined as
+\begin{ceqn}
+\begin{equation}
+ l = \alpha H_p\label{eq:mixlength}
+\end{equation}
+\end{ceqn}
+where $H_p$ is the pressure scale height, and $\alpha$ a free parameter which is determined in advance and keep fixed during the stellar evolution simulations.\par
+
+ It became evident that the influence of the free, relatively arbitrary, parameters associated with MLT significantly conditioned the evolution of the Li abundance.
+Moreover, there is no strong arguments which suggest that the mixing-length parameter is the same in all stars and at all evolutionary phases \citep{Pasetto2014}.
+In \cite{Navarro2020} we showed how an alternative MLT parameterization could produce results in line with the observations.
+Adjustment of MLT over-shooting and $\amlt$ free parameter seems to be required for explaining Li abundances in OC's.\par
+
+In \cite{Sonoi2018}, the authors introduced a way for calibrating $\amlt$ for solar-like stars in which its value increases with increasing $\gsurf$ or decreasing $\teff$,
 
@@
 \begin{equation}
-    \Dot{M} = 4\upi r^2\rho\nu \label{eq:mass_loss}
+ f(x,y) = a_0 + (a_1 + (a_3 + a_5x +a_6y)x + a_4y)x + a_2y\label{eq:alpha_ml}
 \end{equation}
 \end{ceqn}
-
-Using (\ref{eq:wind_conf}) and (\ref{eq:mass_loss}), $\eta_*$ can be approximated by: 
-\begin{ceqn}
-\begin{equation}
-    \eta_* = \frac{B^{2}r^{2}}{\Dot{M}\nu} \label{eq:wind_conf2}
-\end{equation}
-\end{ceqn}
-
-In general, a magnetically channeled outflow has a complex stream geometry but for convenience, the Eq.~\ref{eq:mass_loss} simply characterizes the wind strength in terms of a spherically symmetric mass-loss rate.
-$\nu$ can be characterized by the radial variation of outflow velocity in terms of the velocity law (Eq.~\ref{eq:vel_law})
-\begin{ceqn}
-\begin{equation}
-    \nu(r) = \nu_\infty (1-\rstar/r) \label{eq:vel_law}
-\end{equation}
-\end{ceqn}
-where $r$ represents the distance from the center of the star to the point where the velocity is to be measured, $R_*$ is the radius of the star, and $\nu_\infty$ is the terminal wind velocity.
-$\nu_\infty$ is defined as the velocity that the wind reaches at large distance from the central star, where it is not accelerated anymore by the wind driving force but its deceleration due to interaction with the interstellar medium (ISM) is negligible \citep{Niedzielski2002}.\par
-
-The line-driven winds of OB stars have terminal velocities that scale with the photospheric escape velocity \citep{Lamers2000} according to:
+where
 \begin{ceqn}
 \begin{align}
-    \nu_\infty &\simeq 1.92 \;\nu_{{\rm esc}} \label{eq:vinf}\\
-    \nu_{{\rm esc}} &= \sqrt{\frac{2\,G\,\mstar}{\rstar}} \label{eq:vesc}
+     x &= \frac{\teff-5777}{1000} \label{eq:eq:alpha_x}\\
+     y &= \gsurf-4.44 \label{eq:eq:alpha_y}
 \end{align}
 \end{ceqn}
-where Eq.~\ref{eq:vesc} describes the Newtonian escape velocity from the stellar surface, where $G$ is the gravitational constant,   ${\rstar}$ the radius, and ${\mstar}$ the mass of the star.
-
-Making use of (\ref{eq:vesc}), $\nu_\infty$ can be reformulated as:
-
-\begin{ceqn}
-\begin{equation}
-    \nu_\infty \simeq 1.92 \; x \; 618 \; \Bigg(\sqrt{\frac{\rsun}{\rstar}\frac{\mstar}{\msun}} \;\Bigg) \label{eq:vinf}
-\end{equation}
-\end{ceqn}
-where ${\rsun}$ is the radius, and ${\msun}$ the mass of the Sun.
-
-It has been observed that strongly magnetic intermediate-mass stars typically have rotation rates much slower than other stars in their parent population \citep{Mathys2006}.
-In those stars, the magnetic fields interacts with the mass loss, where the Alfv\'{e}n radius ($R_{A}$) plays an important role.
-Making use of $\eta_*$ (Eq.
-\ref{eq:wind_conf}), $R_{A}$ is defined as the point in which the magnetic field energy density and the kinetic energy density are balanced.
-In case that $R_{A}$ is greater than the stellar radius, then the wind flow will have to follow the magnetic field.
-As a consequence, the material leaves the stellar surface with a higher specific AM, as the co-rotation radius has increased and it roughly corresponds to $R_{A}$.
-Following \citet{Weber1967} the AML can be calculated by:
-\begin{ceqn}
-\begin{equation}
- \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A} \label{eq:j_dot}
-\end{equation}
-\end{ceqn}
-where $\Dot{M}$ is the mass loss rate, $\Omega$ the angular velocity at star surface and $R_A$ the Alfv\'{e}n radius. \par
-
-The above expression can be rewritten according to \citet{Ud-Doula2008} so that it depends on the $\eta_*$ instead of the $R_A$ as follows:
-\begin{ceqn}
-\begin{equation}
- \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{*}\eta_* \label{eq:j_dot_mesa}
-\end{equation}
-\end{ceqn}
-which is a more convenient expression to be implemented in MESA because is based on values directly exposed during the simulations.
+and $a_i$ are the resulting coefficients of the fitting function \ref{eq:alpha_ml} to the calibrated $\alpha$ values for the MLT-BV convection model \citep{Sonoi2018}: $a_0=1.790295$, $a_1=-0.14954$, $a_2=0.069574$, $a_3=-0.00829$, $a_4=0.013165$, $a_5=0.080333$, $a_6=-0.03306$.\par
 
@@
 \subsection{Mesh Models} \label{sec_mesh}
-According to the formulation in the previous sections, the only two free parameters of our implementation are $B$ and $\Omega$.
-The numerical simulations traced the rotational history and $A(\isotope[7]{Li})$ of a $1\, \msun$ star for a variety of initial values for $B$ and $\Omega$ (see also Table \ref{tab:phy_mesa}).
-MESA uses the shellular approach \citep{Meynet1997} to account for hydrostatic effects of rotation into 1D stellar models.\par
+In the current work we have replaced two pre-set parameters (alpha and intensity) for the different simulations of our model.
+Not only have we eliminated the need to assign a value to them, but we allow their values to vary as the simulations evolve over time.
+The single free parameter of our implementation is $\Omega$.
+The numerical simulations traced the rotational history and $A(\isotope[7]{Li})$ of a $1\, \msun$ star for a variety of initial values for $\Omega$ (see also Table \ref{tab:phy_mesa}).\par
 
@@
 		Rotational Mixing & Include SH, ES, GSF, SSI \& DSI\\
-		Magnetic Effects & Magnetic braking based on idealized \\ & monopole field\\
-		Magnetic Field & B(G) variable between [3.0 - 5.0]\\
+            Mixing Length Theory & $\alpha_{{\rm MLT}}$ variable depending on $\teff$\\ & \& $\gsurf$\\
+		Magnetic Effects & Magnetic braking based on idealized \\ & dipole field\\
+		Magnetic Field & B(G) variable, depending on $\rho$, $\teff$ \&  $\Omega$\\\
 		Mass Loss & activated, $\Dot{M}_{{\rm max}} = 10^{-3} \: \msun \: yr^{-1}$\\
-		Angular Moment Loss & activated, $\Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A}$\\
+		Angular Moment Loss & activated, $\Dot{J} = \Omega_* \; \mwind \; \ralfven^2$\\
 		\hline
@@
 
-We adopted solar-scaled abundances and assume \citet{Asplund2009} solar initial abundance ($Z = Z_{\odot, \mathrm{pr} = 0.0142}$).
-We also adopted the following nominal values to express stellar properties values in SI units, $\rsun = 6.957x10^{10}\, cm$ and $\msun = 1.988x10^{33}\, g$ which are consistent with IAU resolutions \citep{Mamajek2015}.
-For a detailed description of the physics adopted in this paper, we refer the reader to \citet{Choi2016}.
-That work was used as starting point for ours regarding the parameterization of the MESA project which was calibrated to reproduce the measured element abundances on the solar surface.\par
+Most of the configuration used in the simulations remain the same as in our previous work, with the exception of the overshooting (deactivated), the $\amlt$ pre-set value, and the formalism used for calculating $\Dot{J}$ (see \cite{Navarro2020} for further details).\par
 
-The models included rotation during the PMS as there are evidences that advocate for a strong established relationship between Li destruction and rotation on that phase (\citeauthor{Bouvier2016} \citeyear{Bouvier2016}, \citeyear{Bouvier2018}).
-We modeled rotation such as AML is computed as a result of the torque applied to the convection zone by a magnetically-coupled wind.
-Additionally, we did not take into account either the influence of internal magnetic fields or their existence during the T-Tauri phase.
-We adopted a simple and pragmatic theoretical approach to establish when the star is reaching the ZAMS phase.
-The selected criterion was based on the simultaneous existence of an extensive convective layer and a radiative core.
+The models continue to including rotation during the PMS.
+We adopted a simple and pragmatic theoretical approach to establish when the star is reaching the ZAMS phase: the simultaneous existence of an extensive convective layer and a radiative core.
 From this moment on, the MB routine was activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participated in the star AML.
-We assumed that the magnetic field did not vary its intensity throughout the evolution of the star until it reached the TAMS.
-In addition, other rotation effects were also considered during the evolution of the models: the AM transport from the radiative interior to the convective envelope and the AM redistribution associated with changes in the internal structure during the process of contraction to the MS.\par
+This time, the restriction of a magnetic field with a constant intensity throughout the evolution of the star until it reached the TAMS is removed in favor of variable one.
+Something similar happens with $\amlt$ which indeed is  a key factor to determine the near-surface structure of stellar models.
 
-In this study we focused on the indirect role played by the MB on the Li destruction.
-This is a consequence of its influence on the rotational history of solar-type stars.
-Our major objective was to check how MB could contribute to explain the evolution of Li in the Sun and other solar-type stars.
-A set of different scenarios with magnetic field strengths ranging between 3.0 and 5.0 G were exercised.
-We computed the evolution of $1\,\msun$ stellar models at solar initial metallicity with $\oomegac$ varying between $0.0084$ and $0.0336$.
-More advanced models for more complex arrangements of the magnetic field as well as the evolution of its intensity throughout the life of the star were left for later work.\par
+In this new paper we continue to focusing on the indirect role played by the MB on the Li destruction.
+The main objective continues to be to check how MB could contribute to explain the evolution of Li in the Sun and other solar-type stars.
+We computed the evolution of $1\,\msun$ stellar models at solar initial metallicity with $\oomegac$ varying between $0.12$ and $0.1425$.
+The results are then compared with the data collected by the Gaia mission and GES for the stars belonging to the OC's referenced in Table \ref{tab:oc_list}.\par
 
-As expressed by Eq.~\ref{eq:j_dot_mesa}, the amount of AML depends on $R_A$ (or alternatively on $\eta_*$), $\Omega$ and $\Dot{M}$.
-For large $\eta_*$ value, the star undergoes a significant deceleration.
-We controlled the value of $\eta_*$ in the models by varying the magnetic field ($B$) and the stellar rotation speed ($\Omega$).
+As expressed by Eq.~\ref{eq:mb_torque}, the amount of AML depends on $\ralfven$, $\Omega$ and $\Dot{M}$.
+For large $\ralfven$ values, the star undergoes a significant deceleration.
+The value of $\ralfven$ depends on the following varying parameters: magnetic field ($B$) intensity, $\teff$ and$\Omega$.
 With regards to $\Dot{M}$ and as reported in Table \ref{tab:phy_mesa}, the empirical formula developed by Reimers \citep{Reimers1975} for stars in the asymptotic giant branch (AGB) was used for calculating the mass-loss.
@@
 
+The distribution is performed in two iterations.
+In the first iteration, each cell $k$ is assigned the maximum amount of $\Dot{J}_{k}$ that it can accommodate in the time step established by the simulation.
+The amounts of $\Dot{J}$ not accommodated by the cells are accumulated as residual angular moment ($\jres$).
+In a second pass, the aggregated residual $\jres$ value collected in the previous iteration is redistributed among those cells that can accommodate more torque.
+In this way we manage to conserve the angular momentum.\par
+
 \subsection{Li evolution without MB}
+As a baseline, we borrow from \cite{Navarro2020} the Figure \ref{fig:li_var_vel_0g} which shows the temporal evolution of surface Li abundance for several $1\msun$ models initialized with different rotational velocities.
+The simulations took into account the effects of rotation and AML caused by stellar winds but not those of MB.
+The purple star and square are the surface Li abundances for the present-day Sun \citep{Asplund2009} and for the Pleiades, respectively \citep{Sestito2005}.
+For more details, refer to the cited work.\par
 
 \begin{figure}
-	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip, width=\columnwidth]{figures/li_var_vel_0_0g.eps}
+	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip, width=\columnwidth]{figures/paper1/li_var_vel_0_0g.eps}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
@@
 
-Figure \ref{fig:li_var_vel_0g} shows the temporal evolution of surface Li abundance for several $1\msun$ models initialized with different rotational velocities.
-The simulations took into account the effects of rotation and AML caused by stellar winds but not those of MB.
-The purple star and square are the surface Li abundances for the present-day Sun \citep{Asplund2009} and for the Pleiades, respectively \citep{Sestito2005}.\par
-
-Notice how the Li abundance in the star surface decreased over time for all simulated models.
+The Li abundance in the star surface decreased over time for all simulated models.
 The solid black line represents the reference model that adopts the solar-calibrated envelope overshooting parameters as documented in \citet{Choi2016}.
 All models burned too much Li before the ZAMS and therefore did not match with the Pleiades average surface Li abundance.
-Also remarkable was the fact that there were barely any differences between the different models in terms of the abundance of Li for much of the PMS.
+There were barely any differences between the different models in terms of the abundance of Li for much of the PMS.
 Only after one million years was Li destroyed to an accentuated degree, since the necessary temperature was not reached at BCZ before.
 Afterwards the different models destroyed the Li in a very similar way due to two main reasons.
-On the one hand, the convective zones that the models developed had very similar sizes, so that the temperature at the BCZ was practically the same.
+On the one hand, the convective zones that the models developed had very similar sizes.
 On the other hand, the differential rotation between the core and the convective zone did not begin to develop until it reached $ \approx 10^6$ years, achieving its maximum difference in the ZAMS about $ \approx 10^7$ years.
-It was at this moment that the combined effects of turbulence and rotational difference should increase, leading to notable differences in the evolution of A(Li) (see Figure \ref{fig:li_var_vel_0g_z1}).
-Later, the reference model (black line) did not deplete Li efficiently on the MS and failed (again) to match the current solar surface Li abundance.
-The other models which included rotation during the PMS with values of $\oomegac$ between 0.0084 and 0.0336 were able to burn Li in a more realistic way.
+It was at this moment that the combined effects of turbulence and rotational difference increase, leading to notable differences in the evolution of A(Li) (see Figure \ref{fig:li_var_vel_0g_z1}).
+Later, the reference model (black line) did not deplete Li efficiently on the MS.
+The other models were able to burn Li in a more realistic way.
 Among them only one (green line) was close to match the present-day Li abundance of the Sun but its rotational velocity was much higher (see Figure \ref{fig:rot_vel_0g}) than the $2\,\kms$ of the Sun \citep{Gill2012}.
@@
 
-For much of the PMS the star rotated as a solid body (see Figure \ref{fig:rot_vel_0g}) and this was because the star had a full convective interior.
-It was not until the end of the Hayashi track that the star began to develop a radiative core.
-It was in this stage when a difference in angular velocity appeared between the upper and lower limits of the radiative and convective zones respectively.
-The degree of differential rotation was directly influenced by the initial $\Omega$.
-As the models were initialized with a higher angular velocity, the difference in velocity between the BCZ and the star surface was accentuated; the higher the initial velocity, the bigger the velocity gradient between the bottom and top limits of the CZ.
-As a consequence, the turbulence strength located at the BCZ increased so that Li could reach regions with temperatures about $\tli$ where it was finally burned and destroyed (see Figure  \ref{fig:li_var_vel_0g}).
-Other investigations \citep{Bouvier2018, Baraffe2017}  point to a tendency diametrically opposed to the one exposed here, i.e.
-the faster the rotation speed, the greater the abundance of Li on the surface of the star.
-\par
-
 \begin{figure}
-	\includegraphics[trim = 25mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/li_var_vel_0_0g_z1.eps}
+	\includegraphics[trim = 25mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/paper1/li_var_vel_0_0g_z1.eps}
     \caption {Similar to Figure \ref{fig:li_var_vel_0g} but zooming in on the ZAMS.
@@
 \begin{figure}
-	\includegraphics[trim = 10mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/rot_vel_var_vel_0_0g.eps}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/paper1/rot_vel_var_vel_0_0g.eps}
     \caption{The evolution of angular velocity at surface (solid line) and at the bottom (dotted line) limit of the uppermost convective zone, as a function of time for several 1 $\msun$ models.
@@
 \end{figure}
-
-Other well-known structural effects of rotation are the decrease of the effective temperature ($\teff$) and to a less extent of stellar luminosity ($L$).
-Both effects can be observed graphically in the HR diagram of Figure \ref{fig:hr_var_vel_0g} which shows a zoomed-in view of evolutionary tracks from the ZAMS until the TAMS for several $1\msun$ models initialized with different rotational velocities.
-If we compare the non-rotating model (black solid line) with the rotating ones we can recognize that at the end of the PMS, the latter reach the ZAMS with a lower $\teff$ than the former.
-These results are in line with those of previous studies \citep[see e.g.
-][]{Eggenberger2012,Piau2001,Pinsonneault1989}.\par
-%(see e.g. \citet{Eggenberger2012}, \citet{Piau2001}, \citet{Pinsonneault1989}).\par
 
@@
 \subsection{Li evolution with MB}
-Figure \ref{fig:li_var_vel_4_0g} shows the temporal evolution of surface Li abundance for several 1 $\msun$ models.
-Those models were initialized with different rotational velocities and took into consideration the effects of MB caused by a magnetic field of intensity 4G.
+Figure \ref{fig:li_var_vel_var_g_3} shows the temporal evolution of surface Li abundance for several 1 $\msun$ models.
+Those models were initialized with different rotational velocities and took into consideration the effects of MB caused by a variable magnetic field which its intensity evolution is shown in Figure \ref{fig:mag_field_var_vel_g3}.
 If we compare it with Figure \ref{fig:li_var_vel_0g} in which the effects of MB were neglected, we notice how the profiles of Li abundance were altered during PMS and MS.
-During the PMS we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see Eq.~\ref{eq:j_dot}) depends directly on the mass loss rate.
+During the PMS we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see Eq.~\ref{eq:j_dot}) depends directly on the mass loss rate (see Figure \ref{fig:mdot_var_vel_g3}) and the $\Omega$ (see Figure \ref{fig:rot_vel_var_vel_var_g3}) evolution.
 If we take into account that for solar-type stars the models predict a modest total mass loss rate, that value is even much lower in this phase.
-On the contrary, during the MS it is observed that the AML is much more significant, causing a smaller amount of Li to be destroyed.\par
+It is in the approach phase to the TAMS that the angular velocity reaches its maximum.
+This, according to our model, plays a crucial role in both the mass loss and the magnetic field strength.
+The higher the $\Omega$, the higher the $\Dot{M}$, and therefore the higher the magnetic field strength.
+These effects combine and lead to an accentuation of the magnetic braking effect.
+As a consequence, there is a slowing down of Li destruction once the models enter the MS.\par
 
 \begin{figure}
-	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/li_var_vel_4_0g.eps}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/li_var_vel_var_g_3.pdf}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
-The models include a magnetic field with an intensity of 4G and PMS rotation with $\oomegac$ between 0.0084 and 0.0336, respectively.
-The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:li_var_vel_4_0g}
-\end{figure}
-
-The effect of the MB routine can be seen even more clearly in Figures~\ref{fig:rot_vel_4g}, \ref{fig:rot_vel_4g_z1} \& Appendixes \footnote{The appendices comprise a series of grids as a function of time and for several 1 $\msun$ models in which on the one hand, the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H} for both variable magnetic field intensities and angular velocities and on the other hand, the evolution of surface rotational velocity are shown.}.
-In those figures we represented the rotation profiles for the surface of the stars and for the bottom of the convective envelope.
-They are 1 $\msun$ models initialized with different rotational velocities and considering the influence of MB.
-Similarly to the $A(\isotope[7]{Li})$ evolution profiles commented in the previous paragraph, the effect of the routine were visible once the ZAMS was reached.
-If we compare the evolution of the curves presented here with those of Figure \ref{fig:rot_vel_0g} we see how the star, instead of keep increasing $\Omega$, began to slow down after having reached its maximum in the passage through the ZAMS.
-Notice that at this point the angular velocities at the surface of the star and at the BCZ reached their maximum difference.
-On the other hand, the MB effect caused the angular velocities between both zones of the star to decrease until, for an age close to that of the Sun (see Figure \ref{fig:rot_vel_4g_z1}), the star practically rotated like a rigid solid.
-These results were also consistent with those obtained by \citet{Eggenberger2010} as far as the effect of the magnetic field, in particular its influence on the loss of angular momentum, has on the rotational velocity of the star.\par
-
-In a similar way we also observed that the models with lower angular velocity generally ended up exhibiting higher values for the Li abundance on the surface (see Figures~ \ref{fig:li_var_vel_4_0g}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).
-In none of those cases we did obtain values of Li on the surface higher than those shown by the model without rotation.\par
-
-\begin{figure}
-	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g.eps}
-    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
-The models include a magnetic field with an intensity of 4\,G, PMS rotation with $\oomegac$ between 0.0084 and 0.0336, respectively and MB.
+The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425, respectively and MB.
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:rot_vel_4g}
+The rest of the dots in different colours represent surface \isotope[7]{Li} abundances of stars with similar parameters to the evolution curve represented with the same colour.The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:li_var_vel_var_g_3}
 \end{figure}
@@
 \begin{figure}
-	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/rot_vel_var_vel_4_0g_z1.eps}
-    \caption{Similar to Figure \ref{fig:rot_vel_4g} but now showing in detail the surface rotational velocity as the star approaches TAMS.}
-    \label{fig:rot_vel_4g_z1}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/mag_field_var_vel_g3.pdf}
+    \caption{The evolution of magnetic field intensity, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
+The models include PMS rotation with $\oomegac$ between 0.12 and 0.1425.
+The solid lines represent the magnetic field strength, while the dotted lines represent the angular evolution of the star.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:mag_field_var_vel_g3}
 \end{figure}
 
-The MB also left its mark on the HR diagram by significantly affecting the $\teff$ of the star.
-To visualize this effect we take as reference Figure \ref{hr_vc_0336_var_g_z1} in which all models were started with the same value $\oomegac=0.0336$ but the intensity of the simulated magnetic field was different.
-Notice that the models with MB produced hotter stars due to its influence.
-The lower speed with which the star rotated due to the MB effect caused the increase of the $\teff$, being this difference of practically $95\,\Kelvin$ between the simulated models with $0.0\,\Gauss$ and $5.0\,\Gauss$ respectively for $log(\llsun)=0$.
+In Figure \ref{fig:rot_vel_var_vel_var_g3} we depict the rotation profiles for the surface of the stars and for the bottom of the convective envelop.
+It shows how the angular velocity increases in value as the star decreases in radius along the PMS.
+During this stage of its evolution, the star has a nearly convective structure, as can be seen in Figure \ref{fig:cz_var_vel_var_g3}.
+As it approaches the ZAMS at $\approx 10^6$ years, the convective zone of the star begins to shrink in size and the core of the star reaches the temperature, pressure and density conditions necessary to develop a radiative core.
+In turn, the difference in the angular velocities of the different models becomes more evident, being higher in those that were initialised with a higher initial angular velocity.
+This fact implies that the mass loss is also higher for those stars that rotate faster.
+As we have described above, a larger mass loss also implies a larger angular momentum loss.
+The combined effects of the increases in effective temperature, angular velocity and mass loss cause the magnetic field strength to reach its maximum in the approach phase of the ZAMS (see Figure \ref{fig:mag_field_var_vel_g3}).
+We observed that the models with lower angular velocity generally ended up exhibiting higher values for the Li abundance on the surface (see Figures~ \ref{fig:li_var_vel_var_g_3}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).\par
 
 \begin{figure}
-	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/hr_vc_0336_var_g_z1.eps}
-    \caption{Similar to Figure \ref{fig:hr_var_vel_0g} but now showing in detail the effects of magnetic braking on the evolutionary tracks for different magnetic field strengths and $\oomegac=0.0336$.
-The presence of a magnetic field produces hotter stars due to the influence of the magnetic braking on the rotational velocity of the star.}
-    \label{hr_vc_0336_var_g_z1}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/rot_vel_var_vel_var_g3.pdf}
+    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
+The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425, respectively and MB.
+The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:rot_vel_var_vel_var_g3}
 \end{figure}
 
-As described in section \ref{sec_mesh}, the MB routine distributed the total amount of AML calculated according to Eq.~\ref{eq:k_jdot} among the different layers that composed the CZ.
-In Figure \ref{fig:cz_vc_028_var_b} we can observe the evolution of the most external CZ normalized with respect to the radius of the star for several 1 $\msun$ models.
-The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths varying between $0.0\,\Gauss$ and $5.0\,\Gauss$.
+As described in section \ref{mod_mb}, the MB routine distributed the total amount of AML calculated according to Eq.~\ref{eq:k_jdot} among the different layers that composed the CZ.
+In Figure \ref{fig:cz_var_vel_var_g3} we can observe the evolution of the most external CZ normalized with respect to the radius of the star for several 1 $\msun$ models.
+The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425.
 In accordance with the established models of stellar evolution, in a solar-type star the CZ covers practically all of it for a large part of the PMS.
-As it approaches the ZAMS, the CZ is decreasing as a consequence of the appearance of a radiative core and maintains an approximately constant radius until the final stage of the MS.
-In this point it increases significantly as a response to the generalized expansion of the star's radius.
-Regarding the effect of MB on the size of the CZ, we observe that as the intensity of the magnetic field increased, the size of the CZ decreased (see Figure \ref{fig:cz_vc_028_var_b_z1_new}).
-The radiative core pushed outward to include a rapidly increasing fraction of the stellar mass, making that the temperature at the CZ base dropped below $\tli$.
-This effect was most evident during MS.
-The decrease in the size of the CZ was in line with the fact that less Li was destroyed by causing less stellar material to reach areas with temperatures above $\tli$.\par
-
-\textbf{Notice the jagged appearance in Figure~\ref{fig:cz_vc_028_var_b_z1_new}.
-We suspected a numerical origin for the wriggles.
-In order to check it out, we computed models for all the physical options considered with higher temporal (time step) and spatial (number of shells) resolutions.
-For the sake of visual clarity, we illustrated the impact on the numerical solutions only for the 5G case, although the effects are the same in all cases.
-These are indicated as dotted lines in the figure, shifted-down just to become visible.
-Note that the larger the spatial resolution the smoother is the curve.
-The effect is less significant for the temporal resolution.
-The curve with 2x in both spatial and temporal resolution shows no difference with the case of 2x in only the number of shells.
-The 4x in spatial resolution points to the most appropriate number of shells to fine-track the evolution of the convective zone.
-However such high resolution implies to significantly increase the computation time, with no effect in the final conclusions of this work.
-Indeed, the size of the zigzag features is around $10^{-3}$,  which is far below the convection timescale and Li abundances measured in the stellar surface.
-It is thus save to neglect such effects by using the original time/space resolution.}
-
-\par
+From the ZAMS on-wards, its size starts to decrease gradually in a first phase and then becomes more pronounced at around $4.0x10^8$ years of age.
+This change in the size of the CZ is due to magnetic braking.
+As can be seen in Figure \ref{fig:rot_vel_var_vel_var_g3}, the star, after reaching its maximum angular velocity when passing the ZAMS, starts to decelerate due to magnetic braking.
+In our models the magnetic braking routine is activated when the star develops a radiative core and the existence of a convective zone above it (see Figure \ref{fig:mb_act_var_vel_g3}).
+It is the presence of the MB that prevents the star from further increasing its angular velocity, as in the models without MB (see Figure \ref{fig:rot_vel_0g}).
+In this way, the Li destruction is attenuated by a lower rotational velocity.
+The models start to decelerate gradually once the ZAMS is passed and during the initial period of their stay in the MS.
+The magnetic field strength remains close to its maximum, decreasing slightly and then decreasing more sharply.
+Its behaviour reflects that of the evolution of the angular velocity, a consequence of the effects of the MB.
+The deceleration process continues progressively until, around the present age of the Sun, we obtain angular velocity values very similar to the Sun.
+The deceleration leads to a reduced effect of centrifugal forces, which allows for a contraction of the star's radius, and thus a smaller size of the CZ.
+Additionally, a lower angular velocity has an effect on the magnetic field strength, as can be seen in Figure \ref{fig:mag_field_var_vel_g3}.
+The coupling between angular velocity, magnetic field strength, and star radius is evident and consistent.\par
 
 \begin{figure}
-	\includegraphics[trim = 15mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g.eps}
-    \caption{The evolution of convective zone size as a function of time for several 1 $\msun$ models.
-The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths vary between $0.0\,\Gauss$ and $5.0\,\Gauss$.
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/cz_var_vel_var_g3.pdf}
+    \caption{The evolution of radiative and convective zones size as a function of time for several 1 $\msun$ models.
+The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425.
 The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:cz_vc_028_var_b}
+    \label{fig:cz_var_vel_var_g3}
 \end{figure}
@@
 \begin{figure}
-	\includegraphics[trim = 15mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/cz_vc_028_var_g_z1_new.eps}
-    \caption{\textbf{Similar to Figure \ref{fig:cz_vc_028_var_b} but now showing in detail the size of the convection zone in the models from ZAMS up to TAMS.
-As the intensity of the magnetic field increases, the size of the convective zone decreases.
-The first three pointed curves (below 0.26), have been artificially shifted downwards to illustrate the impact of the temporal (2x in time step and number of shells, red-dotted line) and spatial (2x and 4x number of shells, black and green-dotted lines, respectively) resolution on the numerical solution for the 5G case.}}
-    \label{fig:cz_vc_028_var_b_z1_new}
-\end{figure}
-
-In Figure \ref{fig:mdot_vc_028_var_b} we can see the evolution of $\Dot{M}$ during PMS and MS for several 1 $\msun$ models.
-The models were all initialized with $\oomegac=0.0336$ and magnetic field strengths varying between $0.0\,\Gauss$ and $5.0\,\Gauss$.
-In the first stages of PMS was where the greatest mass loss was concentrated and this diminished as it approached the ZAMS.
-Because the star also decreased its radius during the PMS, it increased $\Omega$ obeying the principle of conservation of AM.
-When reaching the ZAMS, the stellar radius remained more or less stable for much of the MS (except for its final stage, see Figure \ref{fig:mdot_vc_028_var_b_z1}) but continued to lose mass whilst increasing $\Omega$, although in a less aggressive way if we compare it with the PMS.
-However, as a consequence of both the appearance of the radiative core during Henyey track and the existence of a CZ (see Figure \ref{fig:mb_act_var_vel_vc_028}), the MB routine was activated, causing the angular velocity of the star to begin to decrease along the entire MS.
-The more intense the magnetic field, the greater the braking effect.\par
-
-\begin{figure}
-	\includegraphics[trim = 5mm 10mm 20mm 10mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g.eps}
-    \caption{The evolution of mass loss $\Dot{M}$ as a function of time for several 1 $\msun$ models.
-The models include different magnetic field intensity between $0.0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.028$.
-The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:mdot_vc_028_var_b}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[trim = 5mm 10mm 20mm 10mm, clip,width=\columnwidth]{figures/mdot_vc_028_var_g_z1.eps}
-    \caption{The evolution of mass loss $\Dot{M}$ as the star is approaching TAMS, as a function of time for several 1 $\msun$ models.
-The models include a variable magnetic field intensity between $0.0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.028$.
-The stronger the magnetic field, the lower $\Dot{M}$.}
-    \label{fig:mdot_vc_028_var_b_z1}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[trim = 15mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/mb_act_vc_028_var_g.eps}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/mb_act_var_vel_g3.pdf}
     \caption{The activation of the magnetic braking routine as a function of the presence of a radiative core.
-The models include a magnetic field with an intensity ranging between $3.0\,\Gauss$ and $5.0\,\Gauss$ and PMS rotation with $\oomegac=0.0228$.
+The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425.
 The solid lines signal the magnetic braking routine activation (on) and deactivation (off).
@@
 The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:mb_act_var_vel_vc_028}
+    \label{fig:mb_act_var_vel_g3}
 \end{figure}
 
-\subsection{Alternative Li evolution with MB}
-Up to this point, the simulations of the different models were based on the parameterization gathered in Table \ref{tab:phy_mesa}, which in turn was adopted from \citet{Choi2016}.
-If we recall the evolution of the Li shown in Figures~\ref{fig:li_var_vel_0g} and \ref{fig:li_var_vel_4_0g}, we highlighted that on both cases too much Li was burned before reaching the ZAMS and therefore did not match with the Pleiades average surface Li abundance.
-On the other hand, using a parameterization slightly different from the one used until now in which the parameters of convection and overshooting have been readjusted according to Table \ref{tab:phy_alt_mesa}, the simulations could reproduce more faithfully both the Pleiades cluster and the Sun Li abundances (see Figure \ref{fig:li_3_0g_0314vc}).\par
-
-\begin{table}
-	\centering
-	\caption{Alternative MTL and overshooting parameters.}
-	\label{tab:phy_alt_mesa}
-	\begin{tabular}{ll} 
-		\hline
-		Parameter & Adopted prescriptions and values\\
-		\hline
-		Convection & MLT + Ledoux, $\alpha_\mathrm{MLT}$ = 1.70\\
-		Overshoot & time-dependent, diffusive, $f_\mathrm{ov,core}=0.016$, \\ & $f_\mathrm{ov,sh}=0.002$\\
-		\hline
-	\end{tabular}
-\end{table}
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/lograd_var_vel_g3.pdf}
+    \caption{The evolution of the stellar radius, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
+The models include PMS rotation with $\oomegac$ between 0.12 and 0.1425.
+The purple star is the radius for the present-day Sun \citep{Gill2012}.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:lograd_var_vel_g3}
+\end{figure}
 
 \begin{figure}
-	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/li_3_0g_0314vc.eps}
-    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for a 1 $\msun$ model.
-The model include a magnetic field with an intensity of 3\,G and PMS rotation with $\oomegac= 0.0314$, respectively.
-MLT and overshooting parameters have been set to $\alpha_\mathrm{MLT}=1.70$, $f_\mathrm{ov,core}=0.016$, $f_\mathrm{ov,sh}=0.002$.
-The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:li_3_0g_0314vc}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/lograd_var_vel_g_z13.pdf}
+    \caption{Similar to \ref{fig:lograd_var_vel_g3} but zooming in on the ZAMS. The fastest model exhibits the smallest radius from around $4.5x10^{7}$ till $11.0x10^{7}$ years.}
+    \label{fig:lograd_var_vel_g_z13}
 \end{figure}
 
-Similarly, the evolution of the angular velocity for this new configuration is shown in Figure \ref{fig:rot_vel_var_vel_mlt_3_0g}.
-Here we did not yet reproduce correctly the angular velocity of the Sun.
-It became evident that the influence of the free, relatively arbitrary, parameters associated with MLT significantly conditioned the evolution of the Li abundance.
-However, without the inclusion of MB it was not possible to fit it for the Pleiades and the Sun with the same evolutionary track.
-This fact opens the door for other solar-compatible parameterizations to reproduce the observations in young stellar clusters and for the Sun.
-\par
+The effects of rotation and magnetic braking can also be seen in the H-R diagram and in the evolution of the $\amlt$.
+The inclusion of rotation implies the appearance of centrifugal forces that affect both the structure of the star, the chemical composition of the different strata that compose it, as well as its temperature and luminosity.
+The differential rotation between the boundaries of the radiative core and the convective layers leads to mixing effects in the so-called tachocline.
+The effects of this mixing and of its hydrostatic effects are governed by the MLT in which the $\amlt$ plays a major role in it.
+As aforementioned, MLT suffers from the arbitrariness of the value of $\amlt$, and it was necessary to establish its value based on physical constraints occurring in the interior of the star.
+Figure \ref{fig:alpha_mlt_var_vel_g3} shows how the $\amlt$ parameter evolves over time.\par
 
 \begin{figure}
-	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/rot_vel_3_0g_0314vc.eps}
-    \caption{The evolution of surface rotational velocity, as a function of time for a 1 $\msun$ model.
-The model include a magnetic field with an intensity of 3\,G, PMS rotation with $\oomegac=0.0314$, parameters in Table \ref{tab:phy_alt_mesa} and MB.
-The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/alpha_mlt_var_vel_g3.pdf}
+    \caption{The evolution of $\amlt$, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
+The models include PMS rotation with $\oomegac$ between 0.12 and 0.1425.
 The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:rot_vel_var_vel_mlt_3_0g}
+    \label{fig:alpha_mlt_var_vel_g3}
+\end{figure}
+
+Additionally, these same centrifugal forces, accentuated in stars with a higher angular velocity, reinforce the gravity darkening \citep[see e.g.
+][]{Gossage2021,Paxton2019,Eggenberger2012} as well.
+The centrifugal force causes the star to shift its mass outwards with respect to the rotation axis, the effect being more pronounced in the equatorial regions than at the poles.
+The consequence is that the pressure ($\gsurf$) to which the gas is subjected is lower in the former than in the latter.
+This effect causes those stars that rotate fast enough to appear less dense ($\rho$), less luminous ($L$) and therefore with a lower effective temperature ($\teff$) as they approach the ZAMS (see Figure \ref{fig:hr_var_vel_var_g_z13}).
+If we compare the non-rotating model (black solid line in Figure \ref{fig:rot_vel_0g}) with the rotating ones we can recognize that at the end of the PMS, the latter reach the ZAMS with a lower $\teff$ than the former.
+This also has an impact on the evolution of $\amlt$ because, as discussed above, in our model its value depends proportionally on $\teff$ and $\gsurf$.
+Both parameters are comparatively smaller for stars that rotate faster than for those that rotate more slowly.
+As a consequence $\amlt$ yields a lower value in its time evolution in those phases in which the star rotates faster, being this effect more accentuated in the ZAMS approach.\par
+
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/hr_var_vel_var_g_z13.pdf}
+    \caption{Similar to Figure \ref{fig:hr_var_vel_0g} but now showing in detail the combined effects of the gravity darkening and magnetic braking on the evolutionary tracks for several 1 $\msun$ models.
+The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425.
+The presence of a magnetic field produces hotter stars due to the influence of the magnetic braking on the rotational velocity of the star.
+$\lsun$ = 3.761.}
+    \label{fig:hr_var_vel_var_g_z13}
+\end{figure}
+
+Figures \ref{fig:teff_logg_var_vel_g3} and \ref{fig:teff_logg_var_vel_g_z13} show how $\teff$ and $\gsurf$ behave over the evolution of the models.
+For those with a higher rotation speed we observe that both their temperature and surface gravity are lower than in the slower rotating models, in line with the gravity darkening.
+It is worth noting that in the ZAMS approach phase we observe for the faster rotating model (yellow) that its surface gravity is higher than that of the rest of the slower models.
+This can be explained by looking at Figure \ref{fig:mdot_var_vel_g3}, which shows the mass loss of the star.
+Around the interval $2.5x10^{7}$ and $3.5x10^{7}$ years the mass loss of the fastest model is smaller than that of the rest.
+Since the surface gravity is directly proportional to the amount of mass of the star, if the mass of the star decreases, the surface gravity decreases as well.
+During this period the faster model does so at a lower rate and so we have a higher surface gravity.
+This "anomaly" disappears as soon as the mass loss becomes more pronounced and becomes noticeably larger between $3.5x10^{7}$ and $4.5x10^{7}$ years, shortly before reaching ZAMS.
+In this phase the surface gravity of the fastest rotator is lower than that of the other models.
+We have a similar situation past ZAMS, from around $4.5x10^{7}$ till $11.0x10^{7}$ years.
+In this case the explanation for the higher surface gravity of the model that rotates faster than the others is due to the size of the star, how its radius behaves.
+Figures \ref{fig:lograd_var_vel_g3} and \ref{fig:lograd_var_vel_g_z13} show the time evolution of the radius for the different models.
+In them we can see how for the fastest rotator, in the above-mentioned time interval, its size is smaller.
+As it decreases more than the others, and although it also continues to lose more mass, its $\gsurf$ is greater.
+Let us remember that the radius has an inversely quadratic influence on the value of the $\gsurf$.\par
+
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/teff_logg_var_vel_g3.pdf}
+    \caption{The evolution of $\teff$ and $\gsurf$, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
+The models include PMS rotation with $\oomegac$ between 0.12 and 0.1425.
+The purple star is the $\teff$, and the purple square is the $\gsurf$ for the present-day Sun \citep{Gill2012}.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:teff_logg_var_vel_g3}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/teff_logg_var_vel_g_z13.pdf}
+    \caption{Similar to \ref{fig:teff_logg_var_vel_g3} but zooming in on the ZAMS.
+The models with a higher initial rotational velocity reach already the ZAMS with a lower $\teff$ and higher $\gsurf$.}
+    \label{fig:teff_logg_var_vel_g_z13}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/mdot_var_vel_g3.pdf}
+    \caption{The evolution of mass loss $\Dot{M}$ as a function of time for several 1 $\msun$ models.
+The models include variable magnetic field intensity and PMS rotation with $\oomegac$ between 0.12 and 0.1425.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:mdot_var_vel_g3}
 \end{figure}
@@
 We have shown, through the different simulated stellar models, that the effects induced by the combination of both rotation and magnetic braking mechanisms offer a plausible way to reconcile the observational data with the theoretical models.
-The latter is of major importance both for the transport of AM and for the transport of chemical elements.
-We are also well aware that we are still far from understanding the exact physical mechanisms that govern these processes, so it is necessary to continue to delve into these areas of study.
-Particularly the study of the evolution of the magnetic fields during the PMS and MS and their impact on AML.
+We propose that this result could be achieved by a law of interdependence between $\Omega$ and $B$.
+In contrast to our previous work, these new models remove the restrictions of pre-setting initial and fixed values along their time evolution for the magnetic field strength and the $\amlt$ parameter.
+We cross-checked the model results with measurements obtained from the Gaia and GES mission for stars with similar characteristics to our Sun, focusing on their Li abundances.
+Our models seem to suggest a lower limit to the surface Li abundances for Sun-like stars, and that the Sun indeed appears to have a lower surface Li abundance than the rest.
+They also suggest that the Sun had a much higher angular velocity than it currently shows and that as a result it effectively destroyed its Li reservoirs.
+Coupled with this higher angular velocity, strong magnetic fields developed, triggering an intensive magnetic braking that slowed its angular velocity to its present value.\par
+
+Yet our models still seem to be destroying too much Li when comparing the Li abundances obtained with those provided by the Gaia and GES data.
+Taking as a starting point that faster rotating stars destroy more Li, other mechanisms in addition to magnetic braking are necessary to explain the observational data.
+Among them, disk locking to the protostellar disk contributes to counteract the acceleration of the star during PMS.
+Thus, we would have two braking mechanisms acting at different ages of the star.
+The disk locking would act during the PMS and the magnetic braking mainly during the MS.
+This approach would establish a self-regulating mechanism over the angular velocity of the star that would end up directly influencing the Li evolution.\par
+
+We emphasize as well that our models are close to match at the same time the observed solar Li abundance and $\Omega$, but they are not yet accurate enough.
+Therefore, we must analyse the results obtained with caution and not draw any premature conclusions.
+In addition, it should be noted that our models fail to replicate the average magnetic field strength of the Sun of $1.0\,\Gauss$.
+We have also shown how both an evolving magnetic field strength and, a MLT parametrization could produce results in line with the observations.
+The following conclusions have been supported in the course of this work:
+
+\begin{enumerate}
+    \item A magnetic field with varying intensity is a valid alternative to explain the rotational history of Sun-like stars.
+    \item It is a plausible line of research to provide an answer to the problem of surface Li on the Sun.
+    \item A variable $\amlt$ eliminates the need to pre-fix its value in advance for the entire evolution of the star.
+    \item Our models seem to suggest a lower limit to the observations provided by Gaia and GES for solar sibling stars.
+    \item The strength of the magnetic field regulates the size of the star's convection zone.
+The magnetic field has a direct implication on the rotational history of the star, and the speed with which the star rotates impacts the size of the convective zone.
+    \item Inclusion of the magnetic field leads to cooler models and lower Li depletion in the MS.
+    \item Disk-locking during PMS seem to be also required for explaining Li abundances in young clusters.
+\end{enumerate}
 \par
 
-Future and improved implementations of our routine will be made to study the results on tracks, surface Li composition, rotation, etc.
-It is likely that in a rotating stellar model simulated with MB from the beginning, the differential rotation is very much reduced and therefore Li abundances observed in young stars could be properly explained.
-We propose that this result could be achieved by a law of interdependence between $\Omega$ and $B$.
-So that, during the PMS, when the star rotates faster, the MB effect is more efficient.
-On the other hand, during the MS, when the start slows down, the MB will become less intensive.
-This approach would establish a self-regulating mechanism over the angular velocity of the star that would end up directly influencing the Li evolution.
-It is equally important to understand better the general role of mass loss in AML and mostly in cooler, low-mass stars.
-Nowadays it is challenging to determine the terminal velocity of the stellar wind for this type of stars which plays a key role on the AML.\par
-
-We emphasize as well that our models failed to match at the same time the observed solar Li abundance and $\Omega$.
-We could not, therefore, ascertain to have correctly modeled the rotational history of the Sun.
-In view of these shortcomings in our models, we must analyse the results obtained with caution and not draw any premature conclusions.
-We have also shown how an alternative MLT parametrizations could produce results in line with the observations.
-The following conclusions have been supported in the course of this work:
-\begin{enumerate}
-    \item Inclusion of the magnetic field leads to cooler models and lower Li depletion in the MS.
-    \item A combination of rotation during the PMS and MB effect during MS produces different, potentially more promising behavior than those produced by standard models.
-Thus, our approach points to reproduce the observed A(Li) and the solar rotational velocity at the same time.
-    \item $\teff$ from standard evolutionary tracks represent upper limits since these models do not take into consideration the magnetic braking effect nor rotation.
-    \item The convective zone extension decreases when the intensity of the magnetic field increases.
-    \item MB during PMS and/or adjustment of MLT overshooting and $\amlt$ free parameters seems to be also required for explaining Li abundances in young clusters.
-\end{enumerate}
-
-The next steps will be focused on understanding how magnetic fields are linked to rotation, how their topologies are and how they evolve in time in order to go beyond the treatment on the MB routine that has been done in this work.\par
-
 \section*{Acknowledgements}
-We are pleased to acknowledge Jieun Choi her kindness in responding to the different questions about her work which were aimed to reproduce the results obtained as far as Li is concerned, and also for allowing us access to the MESA files she used.
-Also we very much appreciate the expert support provided by Elisa Delgado-Mena during the process review of this document.
-Similarly, we wish also to thank Matteo Cantiello and Bill Paxton for their very valuable support in the development of the magnetic braking routine in MESA.
-Finally, we wish to acknowledge the generous work and help offered by the MESA community.
+We wish to acknowledge the generous work and help offered by the MESA and TOPCAT communities.
 Authors acknowledges funding support from Spanish public funds (including FEDER fonds) for research under projects ESP2017-87676-C5-2-R and ESP2017-87676-C5-5-R.
@@
 The different figures have been generated making use of GNU Octave 5.1.0.
-\footnote{\url{https://www.gnu.org/software/octave/}}
+\footnote{\url{https://www.gnu.org/software/octave/}}.
+The Gaia and GES data sets have been analyzed using TOPCAT 4.8-7 \footnote{\url{https://www.g-vo.org/topcat/topcat/}}.
 
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_0_0g.eps}
-    \label{fig:subim1}
+    \includegraphics[clip,width=\textwidth]{figures/paper2/li_var_vel_var_g_1.pdf}
+    \label{fig:subim11}
     \end{subfigure}
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_3_0g.eps}
-    \label{fig:subim2}
+    \includegraphics[clip,width=\textwidth]{figures/paper2/li_var_vel_var_g_3.pdf}
+    \label{fig:subim12}
     \end{subfigure}
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_3_5g.eps}
-    \label{fig:subim3}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/paper1/li_var_vel_0_0g.eps}
+    \label{fig:subim13}
     \end{subfigure}
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_4_0g.eps}
-    \label{fig:subim4}
+    \includegraphics[width=\textwidth]{figures/blank.eps}
+    \label{fig:subim14}
     \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_4_5g.eps}
-    \label{fig:subim5}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_var_vel_5_0g.eps}
-    \label{fig:subim6}
-    \end{subfigure}
+
 \caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_vc_0084_var_g.eps}
+    \includegraphics[clip,width=\textwidth]{figures/paper2/rot_vel_var_vel_var_g1.pdf}
     \label{fig:subim21}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_vc_014_var_g.eps}
+    \includegraphics[clip,width=\textwidth]{figures/paper2/rot_vel_var_vel_var_g3.pdf}
     \label{fig:subim22}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_vc_0196_var_g.eps}
+    \includegraphics[clip,width=\textwidth]{figures/paper1/rot_vel_var_vel_0_0g.eps}
     \label{fig:subim23}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_vc_028_var_g.eps}
+    \includegraphics[width=\textwidth]{figures/blank.eps}
     \label{fig:subim24}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/li_vc_0336_var_g.eps}
-    \label{fig:subim25}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[width=\textwidth]{figures/blank.eps}
-    \label{fig:subim26}
     \end{subfigure}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_0_0g.eps}
+    \includegraphics[clip,width=\textwidth]{figures/paper2/alpha_mlt_var_vel_g1.pdf}
     \label{fig:subim41}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_3_0g.eps}
+    \includegraphics[clip,width=\textwidth]{figures/paper2/alpha_mlt_var_vel_g3.pdf}
     \label{fig:subim42}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_3_5g.eps}
-    \label{fig:subim43}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_0g.eps}
-    \label{fig:subim44}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_4_5g.eps}
-    \label{fig:subim45}
-    \end{subfigure}
-    \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/rot_vel_var_vel_5_0g.eps}
-    \label{fig:subim46}
     \end{subfigure}

### 2023-04-25 project/overleaf/paper2/mblithium.tex

@@
 \usepackage[normalem]{ulem}
+\usepackage{array}
+\usepackage[table]{xcolor}
+\usepackage{longtable}
 %\usepackage[firstpage]{draftwatermark}
@@
 In general, we focus on the observations made on OCs, and particularly on those analysed in the \cite{Bragaglia2022} and \cite{Randich2022} papers.
-On the basis of the results presented in these papers, a set of 65 OCs are selected as the reference dataset for our work.
+On the basis of the results presented in these papers, a set of 65 OCs (see Table \ref{tab:oc_list}) are selected as the reference dataset for our work.
 
-\begin{table}
+\begin{table*}
 	\centering
-	\caption{Information on the pre-selected OCs.}
-	\label{tab:oc_list}
-	\begin{tabular}{l l l l} 
+	\begin{tabular}{|l l l l l || c c c c c | c c c c c|} 
 		\hline
-		Cluster & GES\_FLD & age (Gyr) & [Fe/H]\\
+             & & & & & & & $\oomegac$ & & & & & $\oomegac$ & & \\
+		Open Cluster & GES\_FLD & Age (Gyr) & [Fe/H] & $N_*$ & 0.095 & 0.10 & 0.105 & 0.11 & 0.115 & 0.12 & 0.125 & 0.13 & 0.14 & 0.1425\\
 		\hline
-            25 Ori & 25\_Ori & 0.013 & 0 \\
-            ASCC 50 & Assc50 & 0.011 & 0.02 \\
-            Berkeley 21 & Br21 & 2.138 & -0.21 \\
-            Berkeley 22 & Br22 & 2.455 & -0.26 \\
-            Berkeley 25 & Br25 & 2.455 & -0.25 \\
-            Berkeley 30 & Br30 & 0.295 & -0.13 \\
-            Berkeley 31 & Br31 & 2.818 & -0.29 \\
-            Berkeley 32 & Br32 & 4.898 & -0.31 \\
-            Berkeley 36 & Br36 & 6.761 & -0.15 \\
-            Berkeley 39 & Br39 & 5.623 & -0.14 \\
-            Berkeley 44 & Br44 & 1.445 & 0.22 \\
-            Berkeley 73 & Br73 & 1.413 & -0.26 \\
-            Berkeley 75 & Br75 & 1.738 & -0.34 \\
-            Berkeley 81 & Br81 & 1.148 & 0.22 \\
-            Blanco 1 & Blanco1 & 0.105 & -0.03 \\
-            Chamaeleon I & Cha\_I & 0.001 & -0.03 \\
-            Collinder 197 & Col197 & 0.014 & 0.03 \\
-            Collinder 69 & lam\_Ori & 0.013 & -0.09 \\
-            Czernik 24 & Cz24 & 2.692 & -0.11 \\
-            Czernik 30 & Cz30 & 2.884 & -0.31 \\
-            ESO 92-05 & ESO92\_05 & 4.467 & -0.29 \\
-            Haffner 10 & Haf10 & 3.802 & -0.1 \\
-            IC 2391 & IC2391 & 0.029 & -0.06 \\
-            IC 2602 & IC2602 & 0.036 & -0.06 \\
-            IC 4665 & IC4665 & 0.033 & 0.01 \\
-            Loden 165 & Loden165 & 3 &  \\
-            Messier 67 & M67 & 3.981 & -0.02 \\
-            NGC 2141 & NGC2141 & 1.862 & -0.04 \\
-            NGC 2158 & NGC2158 & 1.549 & -0.15 \\
-            NGC 2232 & NGC2232 & 0.018 & -0.03 \\
-            NGC 2243 & NGC2243 & 4.365 & -0.45 \\
-            NGC 2244 & NGC2244 & 0.004 & -0.04 \\
-            NGC 2264 & NGC2264 & 0.003 & -0.1 \\
-            NGC 2355 & NGC2355 & 1 & -0.13 \\
-            NGC 2420 & NGC2420 & 1.698 & -0.15 \\
-            NGC 2425 & NGC2425 & 2.399 & -0.13 \\
-            NGC 2451A & NGC2451 & 0.035 & -0.08 \\
-            NGC 2451B & NGC2451 & 0.041 & -0.02 \\
-            NGC 2516 & NGC2516 & 0.24 & -0.04 \\
-            NGC 2547 & NGC2457 & 0.032 & -0.03 \\
-            NGC 3293 & NGC3293 & 0.01 & -0.08 \\
-            NGC 3532 & NGC3532 & 0.398 & -0.01 \\
-            NGC 3766 & NGC3766 & 0.023 & -0.12 \\
-            NGC 4815 & NGC4815 & 0.372 & 0.08 \\
-            NGC 6005 & NGC6005 & 1.259 & 0.22 \\
-            NGC 6067 & NGC6067 & 0.126 & 0.03 \\
-            NGC 6253 & NGC6253 & 3.246 & 0.33 \\
-            NGC 6259 & NGC6259 & 0.269 & 0.18 \\
-            NGC 6281 & NGC6281 & 0.513 & -0.04 \\
-            NGC 6405 & NGC6405 & 0.035 & -0.02 \\
-            NGC 6530 & NGC6530 & 0.002 & -0.02 \\
-            NGC 6633 & NGC6633 & 0.692 & -0.03 \\
-            NGC 6649 & NGC6649 & 0.071 & -0.08 \\
-            NGC 6705 & NGC6705 & 0.309 & 0.03 \\
-            NGC 6709 & NGC6709 & 0.191 & -0.02 \\
-            NGC 6802 & NGC6802 & 0.661 & 0.14 \\
-            Pismis 15 & Pismis15 & 0.871 & 0.02 \\
-            Pismis 18 & Pismis18 & 0.575 & 0.14 \\
-            Ruprecht 134 & Rup134 & 1.66 & 0.27 \\
-            Trumpler 14 & Trumpler14 & 0.003 & -0.01 \\
-            Trumpler 20 & Trumpler20 & 1.862 & 0.13 \\
-            Trumpler 23 & Trumpler23 & 0.708 & 0.2 \\
-            Trumpler 5 & Trumpler5 & 4.266 & -0.35 \\
-            Gamma Vel & gamma2\_vel & 0.02 & -0.02 \\
-            Rho Oph & Rho\_Oph & 0.001 & 0.03 \\
+            25 Ori & 25\_Ori & 0.013 & 0 & 294 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            ASCC 50 & Assc50 & 0.011 & 0.02 & 1225 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            Berkeley 21 & Br21 & 2.138 & -0.21 & 744 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1 & 1 & 1\\
+            Berkeley 22 & Br22 & 2.455 & -0.26 & 395 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 25 & Br25 & 2.455 & -0.25 & 81 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 30 & Br30 & 0.295 & -0.13 & 332 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 31 & Br31 & 2.818 & -0.29 & 616 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 32 & Br32 & 4.898 & -0.31 & 438 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 36 & Br36 & 6.761 & -0.15 & 739 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            Berkeley 39 & Br39 & 5.623 & -0.14 & 899 & 1 & 1 & 1 & 1 & 1 & 1 & 2 & 2 & 2 & 2\\
+            Berkeley 44 & Br44 & 1.445 & 0.22 & 93 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 73 & Br73 & 1.413 & -0.26 & 77 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 75 & Br75 & 1.738 & -0.34 & 75 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 81 & Br81 & 1.148 & 0.22 & 279 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Blanco 1 & Blanco1 & 0.105 & -0.03 & 463 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Chamaeleon I & Cha\_I & 0.001 & -0.03 & 709 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Collinder 197 & Col197 & 0.014 & 0.03 & 409 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Collinder 69 & lam\_Ori & 0.013 & -0.09 & 836 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Czernik 24 & Cz24 & 2.692 & -0.11 & 346 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Czernik 30 & Cz30 & 2.884 & -0.31 & 226 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            ESO 92-05 & ESO92\_05 & 4.467 & -0.29 & 212 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Haffner 10 & Haf10 & 3.802 & -0.1 & 562 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            IC 2391 & IC2391 & 0.029 & -0.06 & 434 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            IC 2602 & IC2602 & 0.036 & -0.06 & 1836 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0\\
+            \rowcolor{lightgray}
+            IC 4665 & IC4665 & 0.033 & 0.01 & 567 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0 & 0 & 0\\
+            Loden 165 & Loden165 & 3 &  & 388 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            Messier 67 & M67 & 3.981 & -0.02 & 131 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6\\
+            \rowcolor{lightgray}
+            NGC 2141 & NGC2141 & 1.862 & -0.04 & 853 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            NGC 2158 & NGC2158 & 1.549 & -0.15 & 616 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 2232 & NGC2232 & 0.018 & -0.03 & 1866 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 2243 & NGC2243 & 4.365 & -0.45 & 661 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 2244 & NGC2244 & 0.004 & -0.04 & 452 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 2264 & NGC2264 & 0.003 & -0.1 & 1857 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 2355 & NGC2355 & 1 & -0.13 & 208 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            \rowcolor{lightgray}
+            NGC 2420 & NGC2420 & 1.698 & -0.15 & 562 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            \rowcolor{lightgray}
+            NGC 2425 & NGC2425 & 2.399 & -0.13 & 528 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            \rowcolor{lightgray}
+            NGC 2451 & NGC2451 & 0.035 & -0.08 & 1656 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 2 & 1\\
+            \rowcolor{lightgray}
+            NGC 2516 & NGC2516 & 0.24 & -0.04 & 759 & 0 & 0 & 0 & 1 & 1 & 1 & 3 & 4 & 4 & 3\\
+            NGC 2547 & NGC2457 & 0.032 & -0.03 & 477 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 3293 & NGC3293 & 0.01 & -0.08 & 584 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 3532 & NGC3532 & 0.398 & -0.01 & 1145 & 1 & 1 & 0 & 1 & 1 & 1 & 1 & 1 & 0 & 1\\
+            NGC 3766 & NGC3766 & 0.023 & -0.12 & 399 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 4815 & NGC4815 & 0.372 & 0.08 & 218 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 6005 & NGC6005 & 1.259 & 0.22 & 560 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1\\
+            NGC 6067 & NGC6067 & 0.126 & 0.03 & 780 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 6253 & NGC6253 & 3.246 & 0.33 & 646 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 6259 & NGC6259 & 0.269 & 0.18 & 494 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1\\
+            \rowcolor{lightgray}
+            NGC 6281 & NGC6281 & 0.513 & -0.04 & 320 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
+            \rowcolor{lightgray}
+            NGC 6405 & NGC6405 & 0.035 & -0.02 & 701 & 1 & 1 & 1 & 0 & 1 & 3 & 2 & 0 & 0 & 0\\
+            NGC 6530 & NGC6530 & 0.002 & -0.02 & 1983 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 6633 & NGC6633 & 0.692 & -0.03 & 1662 & 2 & 2 & 2 & 1 & 1 & 0 & 0 & 0 & 1 & 1\\
+            NGC 6649 & NGC6649 & 0.071 & -0.08 & 283 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 6705 & NGC6705 & 0.309 & 0.03 & 1066 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 6709 & NGC6709 & 0.191 & -0.02 & 730 & 2 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
+            NGC 6802 & NGC6802 & 0.661 & 0.14 & 197 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Pismis 15 & Pismis15 & 0.871 & 0.02 & 333 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Pismis 18 & Pismis18 & 0.575 & 0.14 & 142 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Ruprecht 134 & Rup134 & 1.66 & 0.27 & 680 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Trumpler 14 & Trumpler14 & 0.003 & -0.01 & 1902 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            Trumpler 20 & Trumpler20 & 1.862 & 0.13 & 1213 & 3 & 3 & 3 & 2 & 2 & 2 & 2 & 2 & 2 & 2\\
+            Trumpler 23 & Trumpler23 & 0.708 & 0.2 & 165 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Trumpler 5 & Trumpler5 & 4.266 & -0.35 & 1138 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Gamma Vel & gamma2\_vel & 0.02 & -0.02 & 1269 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Rho Oph & Rho\_Oph & 0.001 & 0.03 & 311 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\           
             \hline
 	\end{tabular}
-\end{table}
+ 	\caption{List of selected OCs.
+For each OC, name, GES denomination, estimated age, metallicity and number of components are listed.
+The number of OC components selected by the different simulations are shown separately.
+This information is split into two groups, each corresponding to the set of simulations used in the figures shown in the following sections.
+Finally, those OCs that have been selected in any of the simulations are highlighted in grey.}
+  	\label{tab:oc_list}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{l l l l l l l} 
+		\hline
+            ObjectId & $\teff$(K) & $\gsurf$(dex) & FeH(dex) & ALi(dex) & eALi(dex) & Age(Gyr)\\
+		\hline
+            8510914\_1157003 & 5843 & 4.42 & -0.03 & 1.8 & 0.08 & 3.981\\ 
+            8510991\_1146169 & 5797 & 4.4 & -0.01 & 1.65 & 0.08 & 3.981\\ 
+            8512177\_1144050 & 5818 & 4.46 & 0 & 1.99 & 0.06 & 3.981\\ 
+            8513941\_1200571 & 5792 & 4.41 & -0.02 & 1.49 & 0.13 & 3.981\\ 
+            8515559\_1148381 & 5794 & 4.43 & -0.04 & 1.56 & 0.1 & 3.981\\ 
+            8520350\_1147480 & 5875 & 4.47 & -0.03 & 1.21 & 0.24 & 3.981\\ 
+            \hline
+	\end{tabular}
+ 	\caption{M64 - Filter setup: $\teff$ low=5789.184, $\teff$ top=5889.6435, $\gsurf$ low=4.3945, $\gsurf$ top=4.4939, $\feh$ low=-0.05, $\feh$ top=0.05, Age low=3.977, Age top=3.985}
+	\label{tab:oc_m64}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{l l l l l l l} 
+		\hline
+            ObjectId & $\teff$(K) & $\gsurf$(dex) & FeH(dex) & ALi(dex) & eALi(dex) & Age(Gyr)\\
+		\hline
+            12392824-6041134 & 5831 & 4.54 & 0.01 & 2.66 & 0.08 & 1.862\\ 
+            12401318-6037507 & 5810 & 4.54 & 0 & 2.35 & 0.12 & 1.862\\ 
+            12402039-6042428 & 5824 & 4.48 & 0.02 & 2.05 & 0 & 1.862\\ 
+            \hline
+	\end{tabular}
+ 	\caption{Trumpler20 - Filter setup: $\teff$ low=5731.2886, $\teff$ top=5831.2886, $\gsurf$ low=4.4518, $\gsurf$ top=4.5518, $\feh$ low=-0.05, $\feh$ top=0.05, Age low=1.8601, Age top=1.8639}
+	\label{tab:oc_trumpler20}
+
+\end{table*}
 
@@
 \end{itemize}
+
+Tables \ref{tab:oc_m64} and \ref{tab:oc_trumpler20} show the subset of components of OC M67 and Trumpler 20 selected during the simulations for models initialised with $\oomegac$ = 0.14.
+Each of the tables lists the GES identifier of the star and the $\teff$, $\gsurf$, $\feh$, A(Li) and estimated age ranges used in the simulation.
 
@@
 \end{equation}
-\begin{ceqn}
 \begin{equation}
@@
 \end{equation}
-
 \end{ceqn}
@@
 \begin{figure}
-	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/hr_var_vel_0_0g_z1.eps}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/paper1/hr_var_vel_0_0g_z1.eps}
     \caption{An example solar 1$\msun$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities.
@@
 \appendix
-
+\clearpage
 \section{Mesh models visualization}
@@
 
+\appendix
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-07-24 project/overleaf/paper2/mblithium.tex
WIP paper II

@@
 \usepackage{longtable}
+\usepackage{threeparttable}
 %\usepackage[firstpage]{draftwatermark}
@@
 \section{Data} \label{sec_method}
-The datasets provided by the Gaia mission in GEDR3 and GDR3, as well as those published by GES in DR5.0 form the main references used in this work.
+The data sets provided by the Gaia mission in GEDR3 and GDR3, as well as those published by GES in DR5.0 form the main references used in this work.
 In general, we focus on the observations made on OCs, and particularly on those analysed in the \cite{Bragaglia2022} and \cite{Randich2022} papers.
-On the basis of the results presented in these papers, a set of 65 OCs (see Table \ref{tab:oc_list}) are selected as the reference dataset for our work.
+On the basis of the results presented in these papers, a set of 64 OCs (see Table \ref{tab:oc_full_list}) are selected as the reference dataset for our work.
 
@@
 		\hline
-            25 Ori & 25\_Ori & 0.013 & 0 & 294 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            ASCC 50 & Assc50 & 0.011 & 0.02 & 1225 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             Berkeley 21 & Br21 & 2.138 & -0.21 & 744 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1 & 1 & 1\\
-            Berkeley 22 & Br22 & 2.455 & -0.26 & 395 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Berkeley 25 & Br25 & 2.455 & -0.25 & 81 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Berkeley 30 & Br30 & 0.295 & -0.13 & 332 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Berkeley 31 & Br31 & 2.818 & -0.29 & 616 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Berkeley 32 & Br32 & 4.898 & -0.31 & 438 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Berkeley 36 & Br36 & 6.761 & -0.15 & 739 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             Berkeley 39 & Br39 & 5.623 & -0.14 & 899 & 1 & 1 & 1 & 1 & 1 & 1 & 2 & 2 & 2 & 2\\
-            Berkeley 44 & Br44 & 1.445 & 0.22 & 93 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Berkeley 73 & Br73 & 1.413 & -0.26 & 77 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Berkeley 75 & Br75 & 1.738 & -0.34 & 75 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Berkeley 81 & Br81 & 1.148 & 0.22 & 279 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Blanco 1 & Blanco1 & 0.105 & -0.03 & 463 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Chamaeleon I & Cha\_I & 0.001 & -0.03 & 709 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Collinder 197 & Col197 & 0.014 & 0.03 & 409 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Collinder 69 & lam\_Ori & 0.013 & -0.09 & 836 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Czernik 24 & Cz24 & 2.692 & -0.11 & 346 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Czernik 30 & Cz30 & 2.884 & -0.31 & 226 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            ESO 92-05 & ESO92\_05 & 4.467 & -0.29 & 212 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Haffner 10 & Haf10 & 3.802 & -0.1 & 562 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            IC 2391 & IC2391 & 0.029 & -0.06 & 434 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             IC 2602 & IC2602 & 0.036 & -0.06 & 1836 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0\\
-            \rowcolor{lightgray}
             IC 4665 & IC4665 & 0.033 & 0.01 & 567 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0 & 0 & 0\\
-            Loden 165 & Loden165 & 3 &  & 388 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             Messier 67 & M67 & 3.981 & -0.02 & 131 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6\\
-            \rowcolor{lightgray}
             NGC 2141 & NGC2141 & 1.862 & -0.04 & 853 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
-            NGC 2158 & NGC2158 & 1.549 & -0.15 & 616 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            NGC 2232 & NGC2232 & 0.018 & -0.03 & 1866 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            NGC 2243 & NGC2243 & 4.365 & -0.45 & 661 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            NGC 2244 & NGC2244 & 0.004 & -0.04 & 452 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            NGC 2264 & NGC2264 & 0.003 & -0.1 & 1857 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             NGC 2355 & NGC2355 & 1 & -0.13 & 208 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
-            \rowcolor{lightgray}
             NGC 2420 & NGC2420 & 1.698 & -0.15 & 562 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
-            \rowcolor{lightgray}
             NGC 2425 & NGC2425 & 2.399 & -0.13 & 528 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
-            \rowcolor{lightgray}
             NGC 2451 & NGC2451 & 0.035 & -0.08 & 1656 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 2 & 1\\
-            \rowcolor{lightgray}
             NGC 2516 & NGC2516 & 0.24 & -0.04 & 759 & 0 & 0 & 0 & 1 & 1 & 1 & 3 & 4 & 4 & 3\\
-            NGC 2547 & NGC2457 & 0.032 & -0.03 & 477 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            NGC 3293 & NGC3293 & 0.01 & -0.08 & 584 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             NGC 3532 & NGC3532 & 0.398 & -0.01 & 1145 & 1 & 1 & 0 & 1 & 1 & 1 & 1 & 1 & 0 & 1\\
-            NGC 3766 & NGC3766 & 0.023 & -0.12 & 399 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            NGC 4815 & NGC4815 & 0.372 & 0.08 & 218 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             NGC 6005 & NGC6005 & 1.259 & 0.22 & 560 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1\\
-            NGC 6067 & NGC6067 & 0.126 & 0.03 & 780 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            NGC 6253 & NGC6253 & 3.246 & 0.33 & 646 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             NGC 6259 & NGC6259 & 0.269 & 0.18 & 494 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1\\
-            \rowcolor{lightgray}
             NGC 6281 & NGC6281 & 0.513 & -0.04 & 320 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
-            \rowcolor{lightgray}
             NGC 6405 & NGC6405 & 0.035 & -0.02 & 701 & 1 & 1 & 1 & 0 & 1 & 3 & 2 & 0 & 0 & 0\\
-            NGC 6530 & NGC6530 & 0.002 & -0.02 & 1983 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             NGC 6633 & NGC6633 & 0.692 & -0.03 & 1662 & 2 & 2 & 2 & 1 & 1 & 0 & 0 & 0 & 1 & 1\\
-            NGC 6649 & NGC6649 & 0.071 & -0.08 & 283 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            NGC 6705 & NGC6705 & 0.309 & 0.03 & 1066 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             NGC 6709 & NGC6709 & 0.191 & -0.02 & 730 & 2 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
-            NGC 6802 & NGC6802 & 0.661 & 0.14 & 197 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Pismis 15 & Pismis15 & 0.871 & 0.02 & 333 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Pismis 18 & Pismis18 & 0.575 & 0.14 & 142 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Ruprecht 134 & Rup134 & 1.66 & 0.27 & 680 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Trumpler 14 & Trumpler14 & 0.003 & -0.01 & 1902 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            \rowcolor{lightgray}
             Trumpler 20 & Trumpler20 & 1.862 & 0.13 & 1213 & 3 & 3 & 3 & 2 & 2 & 2 & 2 & 2 & 2 & 2\\
-            Trumpler 23 & Trumpler23 & 0.708 & 0.2 & 165 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Trumpler 5 & Trumpler5 & 4.266 & -0.35 & 1138 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Gamma Vel & gamma2\_vel & 0.02 & -0.02 & 1269 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
-            Rho Oph & Rho\_Oph & 0.001 & 0.03 & 311 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\           
             \hline
@@
 The number of OC components selected by the different simulations are shown separately.
-This information is split into two groups, each corresponding to the set of simulations used in the figures shown in the following sections.
-Finally, those OCs that have been selected in any of the simulations are highlighted in grey.}
-  	\label{tab:oc_list}
+This information is split into two groups, each corresponding to the set of simulations used in the figures shown in the following sections.}
+  	\label{tab:oc_reduced_list}
 \end{table*}
@@
 	\end{tabular}
- 	\caption{M64 - Filter setup: $\teff$ low=5789.184, $\teff$ top=5889.6435, $\gsurf$ low=4.3945, $\gsurf$ top=4.4939, $\feh$ low=-0.05, $\feh$ top=0.05, Age low=3.977, Age top=3.985}
-	\label{tab:oc_m64}
+ 	\caption{M67 - Filter setup: $\teff$(K) [5787.4892, 5887.6823], $\gsurf$(dex) [4.3979, 4.4977], $\feh$(dex) [-0.05, 0.05], Age(Gyr) [3.977, 3.985]}
+	\label{tab:oc_m67}
 \end{table*}
@@
 		\hline
-            12392824-6041134 & 5831 & 4.54 & 0.01 & 2.66 & 0.08 & 1.862\\ 
-            12401318-6037507 & 5810 & 4.54 & 0 & 2.35 & 0.12 & 1.862\\ 
-            12402039-6042428 & 5824 & 4.48 & 0.02 & 2.05 & 0 & 1.862\\ 
+            7571111-6048156 & 5084 & 4.54 & 0 & 2.29 & 0.08 & 0.24\\ 
+            7595031-6044149 & 5108 & 4.52 & -0.05 & 1.92 & 0.08 & 0.24\\ 
+            7595280-6032498 & 5093 & 4.49 & -0.03 & 2.29 & 0.08 & 0.24\\ 
             \hline
 	\end{tabular}
- 	\caption{Trumpler20 - Filter setup: $\teff$ low=5731.2886, $\teff$ top=5831.2886, $\gsurf$ low=4.4518, $\gsurf$ top=4.5518, $\feh$ low=-0.05, $\feh$ top=0.05, Age low=1.8601, Age top=1.8639}
-	\label{tab:oc_trumpler20}
+ 	\caption{NGC2516 - Filter setup: $\teff$(K) [5010.0324, 5111.0095], $\gsurf$(dex) [4.4405, 4.5405], $\feh$(dex) [-0.05, 0.05], Age(Gyr) [0.23964, 0.24036]}
+	\label{tab:oc_ngc2516}
 
@@
 As a result of this first filtering, the number of selected records is reduced to 43 299.
-On this subset of data we perform another filtering, this time aimed at selecting those that offer information on the following attributes:
+On this subset of data we perform another filtering, this time aimed at selecting those that offer not null information on the following attributes:
 \begin{itemize}
@@
     \item Surface gravity
+    \item Cluster membership probability (>= 0.95)
 \end{itemize}
 
-This second filtering reduces the population of stars to be considered to a total of 21 125 records.
-Finally, a third filtering is carried out to select those stars belonging to the OCs listed in Table \ref{tab:oc_list}, giving us a final study population of 20 913 stars distributed in 65 OCs.\par
+This second filtering reduces the population of stars to be considered to a total of 5 895 records which are distributed among the 58 OCs listed in Table \ref{tab:oc_full_list}.\par
 
@@
 These two new parameters represent a fine-tuning of the search process for sibling candidates to our Sun.
-We use the age of the OCs reported in Table \ref{tab:oc_list} and the data obtained from the set of simulations performed using Modules for Experiments in Stellar Astrophysics stellar evolution code \citep[MESA; ][]{Paxton2011, Paxton2013,Paxton2015, Paxton2018, Paxton2019}.
+We use the age of the OCs reported in Table \ref{tab:oc_reduced_list} and the data obtained from the set of simulations performed using Modules for Experiments in Stellar Astrophysics stellar evolution code \citep[MESA; ][]{Paxton2011, Paxton2013,Paxton2015, Paxton2018, Paxton2019}.
 Our models extend the investigations documented in \cite{Navarro2020} on the influence of rotation and magnetic braking on the star Li depletion during the pre-main sequence (PMS) and the main sequence (MS).
@@
 
-From the approximately 21 000 identified stars, we proceed to select those with metallicity ($\feh$), effective temperature ($\teff$), and surface gravity ($\gsurf$) within the following limits:
+From the approximately 5 900 identified stars, we proceed to select those with metallicity ($\feh$), effective temperature ($\teff$), and surface gravity ($\gsurf$) within the following limits:
 \begin{itemize}
@@
 
-Tables \ref{tab:oc_m64} and \ref{tab:oc_trumpler20} show the subset of components of OC M67 and Trumpler 20 selected during the simulations for models initialised with $\oomegac$ = 0.14.
+The method of selection is as follows: Based on the values for $\teff$, $\gsurf$ and $\feh$ given by our simulations, we proceed to select in the different OCs those stars that have values within the established ranges.
+That is, for the estimated age of a particular cluster, we calculate an interval of $\pm$0.1 for young OCs (age < 1Gyr) and $\pm$0.15 for old OCs (age >= 1Gyr) \citep[see][and reference values therein]{Cantat-Gaudin2020}.
+This age interval is used to select in our simulations the time step that is closest to the lower and upper age limits.
+Once these two time steps are found, the associated limit values for $\teff$, $\gsurf$ and $\feh$ are gathered.
+With these value ranges we proceed then to select in each of the OCs those stars that have values for these same parameters within them.
+In this way, we finally obtain the candidate stars for relevant for being compared with the simulation results.\par
+
+Tables \ref{tab:oc_m67} and \ref{tab:oc_ngc2516} summarize the subset of components of OC M67 and NGC2516 selected during the simulations for models initialised with $\oomegac$ = 0.14.
 Each of the tables lists the GES identifier of the star and the $\teff$, $\gsurf$, $\feh$, A(Li) and estimated age ranges used in the simulation.
@@
 
+In \cite{Gossage2021}, the authors also approach the problem of AML from a similar starting point to ours, but with some notable differences.
+In the Table \ref{tab:gassage_vs_navarro} we compare the most notable similarities and differences between the two studies.
+In both works the models consider rotation not only in the MS but also in the PMS, with the difference that in Gossage's work they take into account disk-locking.
+This aspect is not so important in our simulations since in our case we have focused on the A(Li) profiles.
+In contrast, Gossage's work focuses mainly on modelling rotation profiles of stars, which are significantly affected by disk-locking.
+Similarly, the effects of magnetic braking on angular momentum loss are also considered.
+In our case for stars with $\mstar$=1$\msun$ (solar-twins) and in their with 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$.
+Both models also use MLT but in our case the MLT alpha value is not pre-fixed but varies throughout the simulation.
+These two works complement each other as they are based on different formalisms.
+On the one hand, our work is based on the proposal of \cite{Gallet2013} whose most relevant parameters are Ro and the presence of a magnetic field of varying strength (B), and on the other hand Gossage presents two formalisms: \cite{Matt2015} and \cite{Garraffo2018}.
+The former is mainly characterised by $\rossby$ and the $\mstar$, while the latter by $\rossby$ and the complexity of the magnetic field ($\magfc$).
+As far as OCs observations are concerned, in this work we have had access to a larger number of them, and covered a wider range of ages [0.001-6.761] Gyr.
+Finally, both papers simulate their respective models in MESA.\par
+
+\begin{table}
+	\centering
+	\caption{Comparison with \cite{Gossage2021} work.}
+	\label{tab:gassage_vs_navarro}
+        \begin{threeparttable}
+	\begin{tabular}{lll} 
+		\hline
+		Parameter & This work & \cite{Gossage2021}\\
+		\hline
+
+		Rotation & At PMS \& MS & At PMS \& MS\\
+  		Solar Mass & $\mstar$=1$\msun$ & 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$ \\
+		Convection & MLT variable $\alpha_{{\rm MLT}}$ & MLT fixed $\alpha_{{\rm MLT}}$\\
+		Magnetic Braking & \cite{Gallet2013} & \cite{Matt2015}, \\ Formalism & & \cite{Garraffo2018}\\
+            Magnetic Braking & $\rossbystar$ \& $\bstar$ & $\rossbystar$ \& $\mstar$, \\ Parameters & & $\rossbystar$ \& $\magfc$\tnote{(1)}\\
+		Magnetic Field & B(G) variable & Not used\\ Strength & & \\
+            Disk locking & Not present & Present\\
+            Number of OC's & 64 & 11 \\
+            Source of data & GDR3.0 GES DR5.0 & Several works \\
+            MESA version & r10398 & r11701\\
+		\hline
+	\end{tabular}
+        \end{threeparttable}
+        \begin{tablenotes}\footnotesize
+        \item (1) Magnetic field complexity
+        \end{tablenotes}
+\end{table}
+
 In the remainder of this paper, we describe a semi-empirical approach for the calculation of the AML as a result of the braking caused by a magnetic field of varying strength.
 The magnetic field evolves according to the $\teff$ and angular velocity ($\Omega$) of the star.
-This has been implemented as an extension to MESA.
+This has been implemented as an extension to MESA.\par
 
 \subsection{Modelling magnetic intensity} \label{mod_mbi}
-The variable magnetic field strength modelling approach is supported by the papers \cite{Cranmer2011} and \cite{Gallet2013}.
-We start by enumerating the most relevant aspects and assumptions made in modelling the evolution of magnetic field intensity.\par
+In our models we make the case that the magnetic field strength should vary over the lifetime of the star rather than remain fixed throughout its evolution.
+For this we rely on a formalism \citep{Gallet2013} that allows us to calculate a variable magnetic field strength.
+This model adopts a dynamo-type prescription that allows a relationship to be established between the $\Omega$ , $\teff$, and the strength of the magnetic field ($B$), as well as with the mass-loss ($\mwind$) rate caused by the stellar wind.
+We start by enumerating the most relevant aspects and assumptions made in modelling the evolution of magnetic field intensity.
+\\par
 
@@
 \end{ceqn}
-where $b$ is the dynamo exponent, $B_*$ is the strength of the magnetic filed, and $f_*$ is the filling factor, i.e. the fraction of the star surface which is magnetized \citep{Gallet2013}.
-
-On the other hand, $B_*$ is proportional to the equipartition magnetic field strength,
+where $f_*$ is the filling factor, i.e.
+the fraction of the star surface which is magnetized, $B_*$ is the strength of the magnetic filed, $\Omega_*$ is the star angular velocity at stellar surface, and $b$ is the dynamo exponent \citep{Gallet2013}.
+The multiplication of $f_*$ by $B_*$ will allow us to obtain the magnetic field strength.
+\par
+
+According to this approach, the magnetic field strength will vary proportionally to the effective temperature ($\teff$) of the star as follows: 
 \begin{ceqn}
@@
 \end{ceqn}
-where $\Beq$ is defined as 
+$B_*$ is proportional to the equipartition magnetic field strength, being $\Beq$ is defined as 
 \begin{ceqn}
@@
 \end{ceqn}
-where $\rho_*$ is the photospheric density, $\boltzmann$ is Boltzmann's constant, $\massH$ is the mass of a hydrogen atom, and $\mu$ the mean atomic weight \citep{Cranmer2011}.
+where $P_*$ is the photospheric gas pressure, $\rho_*$ is the photospheric density, $\boltzmann$ is Boltzmann's constant, $\massH$ is the mass of a hydrogen atom, and $\mu$ the mean atomic weight \citep{Cranmer2011}.
 
@@
 
-\cite{Cranmer2011} found that the magnetic field strength is mainly influenced by $f_*$ rather than by the rotation period of the star.
+On the other hand, \cite{Cranmer2011} found that the magnetic field strength is mainly influenced by $f_*$ rather than by the rotation period of the star.
 That is, variations in the star's angular velocity do not significantly alter its magnetic field strength.
@@
 
-Finally, we are able to calculate the filling factor for a star by applying the adjustment made by \cite{Gallet2013},
+Finally, we are able to calculate the filling factor for a star by applying the adjustment made by \cite{Gallet2013} which more closely reproduces the average filling factor of the present Sun ($\fsun$ = 0.001-0.01),
 \begin{ceqn}
@@
 As shown in Figure \ref{fig:mag_field_limits_var_vel_g3}, the simulations yield magnetic field values that show maximum intensities of around $1.0K\,\Gauss$, making the use of the model employed previously non viable.
+\par
 
@@
 
-\subsection{Mesh Models} \label{sec_mesh}
+\subsection{Grid Models} \label{sec_grid}
 In the current work we have replaced two pre-set parameters (alpha and intensity) for the different simulations of our model.
@@
 
-The distribution is performed in two iterations.
-In the first iteration, each cell $k$ is assigned the maximum amount of $\Dot{J}_{k}$ that it can accommodate in the time step established by the simulation.
-The amounts of $\Dot{J}$ not accommodated by the cells are accumulated as residual angular moment ($\jres$).
-In a second pass, the aggregated residual $\jres$ value collected in the previous iteration is redistributed among those cells that can accommodate more torque.
-In this way we manage to conserve the angular momentum.\par
-
 \subsection{Li evolution without MB}
@@
 \begin{figure}
-	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip, width=\columnwidth]{figures/paper1/li_var_vel_0_0g.eps}
+	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip, width=\columnwidth]{figures/paper1/li_var_vel_0_0g.pdf}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
@@
 \begin{figure}
-	\includegraphics[trim = 25mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/paper1/li_var_vel_0_0g_z1.eps}
+	\includegraphics[trim = 25mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/paper1/li_var_vel_0_0g_z1.pdf}
     \caption {Similar to Figure \ref{fig:li_var_vel_0g} but zooming in on the ZAMS.
@@
 \begin{figure}
-	\includegraphics[trim = 10mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/paper1/rot_vel_var_vel_0_0g.eps}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/paper1/rot_vel_var_vel_0_0g.pdf}
     \caption{The evolution of angular velocity at surface (solid line) and at the bottom (dotted line) limit of the uppermost convective zone, as a function of time for several 1 $\msun$ models.
@@
 \begin{figure}
-	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/paper1/hr_var_vel_0_0g_z1.eps}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/paper1/hr_var_vel_0_0g_z1.pdf}
     \caption{An example solar 1$\msun$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities.
@@
 %%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%
-
 \appendix
+\section{Full list of OC's}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|l l l l l || c c c c c | c c c c c|} 
+		\hline
+             & & & & & & & $\oomegac$ & & & & & $\oomegac$ & & \\
+		Open Cluster & GES\_FLD & Age (Gyr) & [Fe/H] & $N_*$ & 0.095 & 0.10 & 0.105 & 0.11 & 0.115 & 0.12 & 0.125 & 0.13 & 0.14 & 0.1425\\
+		\hline
+            25 Ori & 25\_Ori & 0.013 & 0 & 294 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            ASCC 50 & Assc50 & 0.011 & 0.02 & 1225 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            Berkeley 21 & Br21 & 2.138 & -0.21 & 744 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1 & 1 & 1\\
+            Berkeley 22 & Br22 & 2.455 & -0.26 & 395 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 25 & Br25 & 2.455 & -0.25 & 81 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 30 & Br30 & 0.295 & -0.13 & 332 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 31 & Br31 & 2.818 & -0.29 & 616 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 32 & Br32 & 4.898 & -0.31 & 438 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 36 & Br36 & 6.761 & -0.15 & 739 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            Berkeley 39 & Br39 & 5.623 & -0.14 & 899 & 1 & 1 & 1 & 1 & 1 & 1 & 2 & 2 & 2 & 2\\
+            Berkeley 44 & Br44 & 1.445 & 0.22 & 93 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 73 & Br73 & 1.413 & -0.26 & 77 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 75 & Br75 & 1.738 & -0.34 & 75 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Berkeley 81 & Br81 & 1.148 & 0.22 & 279 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Blanco 1 & Blanco1 & 0.105 & -0.03 & 463 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Chamaeleon I & Cha\_I & 0.001 & -0.03 & 709 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Collinder 197 & Col197 & 0.014 & 0.03 & 409 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Collinder 69 & lam\_Ori & 0.013 & -0.09 & 836 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Czernik 24 & Cz24 & 2.692 & -0.11 & 346 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Czernik 30 & Cz30 & 2.884 & -0.31 & 226 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            ESO 92-05 & ESO92\_05 & 4.467 & -0.29 & 212 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Haffner 10 & Haf10 & 3.802 & -0.1 & 562 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            IC 2391 & IC2391 & 0.029 & -0.06 & 434 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            IC 2602 & IC2602 & 0.036 & -0.06 & 1836 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0\\
+            \rowcolor{lightgray}
+            IC 4665 & IC4665 & 0.033 & 0.01 & 567 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0 & 0 & 0\\
+            Loden 165 & Loden165 & 3 &  & 388 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            Messier 67 & M67 & 3.981 & -0.02 & 131 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6\\
+            \rowcolor{lightgray}
+            NGC 2141 & NGC2141 & 1.862 & -0.04 & 853 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            NGC 2158 & NGC2158 & 1.549 & -0.15 & 616 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 2232 & NGC2232 & 0.018 & -0.03 & 1866 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 2243 & NGC2243 & 4.365 & -0.45 & 661 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 2244 & NGC2244 & 0.004 & -0.04 & 452 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 2264 & NGC2264 & 0.003 & -0.1 & 1857 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 2355 & NGC2355 & 1 & -0.13 & 208 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            \rowcolor{lightgray}
+            NGC 2420 & NGC2420 & 1.698 & -0.15 & 562 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            \rowcolor{lightgray}
+            NGC 2425 & NGC2425 & 2.399 & -0.13 & 528 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            \rowcolor{lightgray}
+            NGC 2451 & NGC2451 & 0.035 & -0.08 & 1656 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 2 & 1\\
+            \rowcolor{lightgray}
+            NGC 2516 & NGC2516 & 0.24 & -0.04 & 759 & 0 & 0 & 0 & 1 & 1 & 1 & 3 & 4 & 4 & 3\\
+            NGC 2547 & NGC2457 & 0.032 & -0.03 & 477 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 3293 & NGC3293 & 0.01 & -0.08 & 584 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 3532 & NGC3532 & 0.398 & -0.01 & 1145 & 1 & 1 & 0 & 1 & 1 & 1 & 1 & 1 & 0 & 1\\
+            NGC 3766 & NGC3766 & 0.023 & -0.12 & 399 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 4815 & NGC4815 & 0.372 & 0.08 & 218 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 6005 & NGC6005 & 1.259 & 0.22 & 560 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1\\
+            NGC 6067 & NGC6067 & 0.126 & 0.03 & 780 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 6253 & NGC6253 & 3.246 & 0.33 & 646 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 6259 & NGC6259 & 0.269 & 0.18 & 494 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1\\
+            \rowcolor{lightgray}
+            NGC 6281 & NGC6281 & 0.513 & -0.04 & 320 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
+            \rowcolor{lightgray}
+            NGC 6405 & NGC6405 & 0.035 & -0.02 & 701 & 1 & 1 & 1 & 0 & 1 & 3 & 2 & 0 & 0 & 0\\
+            NGC 6530 & NGC6530 & 0.002 & -0.02 & 1983 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 6633 & NGC6633 & 0.692 & -0.03 & 1662 & 2 & 2 & 2 & 1 & 1 & 0 & 0 & 0 & 1 & 1\\
+            NGC 6649 & NGC6649 & 0.071 & -0.08 & 283 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            NGC 6705 & NGC6705 & 0.309 & 0.03 & 1066 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            NGC 6709 & NGC6709 & 0.191 & -0.02 & 730 & 2 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
+            NGC 6802 & NGC6802 & 0.661 & 0.14 & 197 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Pismis 15 & Pismis15 & 0.871 & 0.02 & 333 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Pismis 18 & Pismis18 & 0.575 & 0.14 & 142 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Ruprecht 134 & Rup134 & 1.66 & 0.27 & 680 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Trumpler 14 & Trumpler14 & 0.003 & -0.01 & 1902 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            \rowcolor{lightgray}
+            Trumpler 20 & Trumpler20 & 1.862 & 0.13 & 1213 & 3 & 3 & 3 & 2 & 2 & 2 & 2 & 2 & 2 & 2\\
+            Trumpler 23 & Trumpler23 & 0.708 & 0.2 & 165 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Trumpler 5 & Trumpler5 & 4.266 & -0.35 & 1138 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Gamma Vel & gamma2\_vel & 0.02 & -0.02 & 1269 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
+            Rho Oph & Rho\_Oph & 0.001 & 0.03 & 311 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\           
+            \hline
+	\end{tabular}
+ 	\caption{List of selected OCs.
+For each OC, name, GES denomination, estimated age, metallicity and number of components are listed.
+The number of OC components selected by the different simulations are shown separately.
+This information is split into two groups, each corresponding to the set of simulations used in the figures shown in the following sections.
+Finally, those OCs that have been selected in any of the simulations are highlighted in grey.}
+  	\label{tab:oc_full_list}
+\end{table*}
+
 \clearpage
-\section{Mesh models visualization}
+\section{Grid models visualization}
 The following figures comprise a series of grids as a function of time and for several 1 $\msun$ models in which on the one hand, the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H} for both variable magnetic field intensities and angular velocities and on the other hand, the evolution of surface rotational velocity are shown.
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/paper1/li_var_vel_0_0g.eps}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/paper1/li_var_vel_0_0g.pdf}
     \label{fig:subim13}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[clip,width=\textwidth]{figures/paper1/rot_vel_var_vel_0_0g.eps}
+    \includegraphics[clip,width=\textwidth]{figures/paper1/rot_vel_var_vel_0_0g.pdf}
     \label{fig:subim23}

### 2023-07-24 project/octave/plotting_routines.m

@@ -74,7 +74,7 @@ global ngc_2516_table_filename = 'NGC_2516_J_A+A_659_A85_tableb6';
 global ges_dr5_all_oc_table_filename = 'ges_dr5_oc_li1_feh_teff_logg_age2.csv';
 global br_21 ='Br21';
 global br_22 ='Br22';
-global br_25 ='Br25';
+#global br_25 ='Br25';
 global br_30 ='Br30';
 global br_31 ='Br31';
 global br_32 ='Br32';
@@ -96,7 +96,7 @@ global haf_10 = 'Haf10';
 global ic_2391 = 'IC2391';
 global ic_2602 = 'IC2602';
 global ic_4665 = 'IC4665';
-global loden165 = 'Loden165';
+#global loden165 = 'Loden165';
 global m_67 = 'M67';
 global ngc_2141 = 'NGC2141';
 global ngc_2158 = 'NGC2158';
@@ -125,7 +125,7 @@ global ngc_6709 = 'NGC6709';
 global ngc_6802 = 'NGC6802';
 global pismis_15 = 'Pismis15';
 global pismis_18 = 'Pismis18';
-global rho_oph = 'Rho_Oph';
+#global rho_oph = 'Rho_Oph';
 global rup_134 = 'Rup134';
 global trumpler_14 = 'Trumpler14';
 global trumpler_20 = 'Trumpler20';
@@ -402,7 +402,7 @@ function [plot_path] = plot_clusters(A, color, width, ytick, axis_limits, aidx, 
   global ges_dr5_all_oc_table_filename;
   global br_21;
   global br_22;
-  global br_25;
+  #global br_25;
   global br_30;
   global br_31;
   global br_32;
@@ -424,7 +424,7 @@ function [plot_path] = plot_clusters(A, color, width, ytick, axis_limits, aidx, 
   global ic_2391;
   global ic_2602;
   global ic_4665;
-  global loden165;
+  #global loden165;
   global m_67;
   global ngc_2141;
   global ngc_2158;
@@ -453,7 +453,7 @@ function [plot_path] = plot_clusters(A, color, width, ytick, axis_limits, aidx, 
   global ngc_6802;
   global pismis_15;
   global pismis_18;
-  global rho_oph;
+  #global rho_oph;
   global rup_134;
[…]

### 2023-09-03 project/octave/plotting_routines.m
Adjust grid lines in plots

@@ -1,7 +1,9 @@
 format longEng;
+% grid on by default. Each plot decides if any axis should be hidden
+%grid on;
 clear global;
 
-global header_lines       = 6; 
+global header_lines       = 6;
 global table_header_lines = 1;
 
 %cols position in output MESA files
@@ -147,7 +149,7 @@ global filename = '1M_photosphere_history.data';
 %global filename = 'annex2B/1M_photosphere_history.data';
 global gauss_fields = ['0g'; '2g'; '2.5g'; '3g'; '3.3g'; '3.5g'; '4g'; '4.3g'; '4.5g'; '5g'; '5.5g'; 'Xg'];
 %global gauss_fields = ['0g'; '3g'; '3.5g'; '4g'; '4.5g'; '5g'];
-global rotational_vels = ['0crit';'0084crit'; '014crit'; '0196crit'; '028crit'; 
+global rotational_vels = ['0crit';'0084crit'; '014crit'; '0196crit'; '028crit';
   '0336crit';'029crit';'030crit';'031crit';'0312crit';'0314crit';'032crit';'0336crit_alpha';
   '11crit';'105crit';'1075crit';'1125crit';'1025crit';
   '115crit';'10crit';'0975crit';'095crit';'0925crit';'125crit';
@@ -155,7 +157,7 @@ global rotational_vels = ['0crit';'0084crit'; '014crit'; '0196crit'; '028crit';
   '13crit';'1325crit';'1375crit';'14crit';'1425crit';
   '145crit';'1475crit';'15crit';'1525crit';'155crit'];
 %dl -> disk locking
-global dl_rotational_vels = ['0crit_dl';'0084crit_dl'; '014crit_dl'; '0196crit_dl'; 
+global dl_rotational_vels = ['0crit_dl';'0084crit_dl'; '014crit_dl'; '0196crit_dl';
   '028crit_dl'; '0336crit_dl';'029crit_dl';'030crit_dl';'031crit_dl';
   '0312crit_dl';'0314crit_dl';'032crit_dl';'9_090256e-6_dl'];
 global colors = ['k'; 'r'; 'g'; 'b'; 'y'; 'm'; 'c'];
@@ -226,6 +228,7 @@ global idx_9_090256e_6_dl  = 13;
 global sun_gauss_field = 1.0; %G
 global sun_age = 4.57e9; %years
 global sun_A_Li7 = 1.1;
+global sun_A_eLi7 = 0.1;
[…]

### 2023-09-03 project/overleaf/paper2/mblithium.tex
WIP paper II

@@
 % Keep the title short and informative.
-\title[Rotation, magnetic braking \& Li abundances II]{Study of the effects of magnetic braking on the lithium abundances of the Sun and solar-type stars II}
+\title[Rotation, magnetic braking \& Li abundances II]{Variable magnetic field and mixing-length convection parameter to reproduce Li abundances and rotation period of the Sun and solar-type stars}
 
@@
 \begin{abstract}
-The study of lithium (Li) surface abundance in the Sun and young stellar globular clusters which are seemingly anomalous in present-day scenarios, as well as the influence of rotation and magnetic braking (MB) on its depletion during pre-main sequence (PMS) and main sequence (MS).
-In this work, the effects of rotational mixing and of the rotational hydrostatic effects on Li abundances are studied by simulating several grids of PMS and MS rotating models.
-Those effects are combined with the additional impact caused by the MB effect.
-Its intensity evolves as the star does i.e.
-not remaining fixed through the simulations.
+Finding the reasons why lithium (Li) surface abundance in the Sun and young stellar globular clusters are seemingly anomalous in present-day scenarios, will deliver a better understanding about the mechanisms which influence the Li depletion during the stellar evolution, in particular during the pre-main sequence (PMS) and main sequence (MS).\par
+
+The present work documents an approach in which neither the magnetic field strength nor the MLT parametrisation remains fixed throughout the evolution of the star.
+The influence of free parameters associated with $\amlt$ and $B$ significantly condition the the Li abundance.
+We document an alternative to prefixing values to these parameters which ends up providing results that are compatible with observational data for Li abundances of solar siblings hosted in young clusters.\par
+
+We explore, by simulating several grids of PMS and MS rotating models, how the effects of rotational mixing and of the rotational hydrostatic effects governs the Li abundances.
+The combined effects of the MLT and MB constraint the Li evolution of the solar models.
+We calculate accurate solar models for which the $\amlt$ and $B$ values are variable over time, i.e.
+not remaining fixed through the simulations, and make them dependent on important solar parameters.
 The data obtained from simulations are confronted with observational data measured on young open clusters collected by the Gaia-ESO Survey (GES).
 The results show how faster rotating models destroy more Li in the late PMS and during the MS, and how the effects of magnetic braking mitigate this destruction.
-The interconnection between the star's effective temperature, surface gravity and rotational velocity, as well as its chemical composition, determine the strength of the magnetic field and impact the effectiveness of the convective motions involved in the destruction of Li.
-The present work documents an approach in which neither the magnetic field strength nor the MLT parametrisation remains fixed throughout the evolution of the star, and which ends up providing results that are compatible with observational data for Li abundances of solar siblings hosted in young clusters.
+The interconnection between the star's internal convective motions, effective temperature, surface gravity and rotational velocity, as well as its chemical composition, determine the strength of the magnetic field and impact the effectiveness of the convective motions involved in the destruction of Li.
 
@@
 
-As noted in our earlier work, discrepancies in Li abundances have been detected in young cluster for which there is as yet no coherent explanation \citep[see][and references therein]{Navarro2020}.
+Discrepancies in Li abundances have been detected in young cluster for which there is as yet no coherent explanation \citep[see][and references therein]{Navarro2020}.
 Li is destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches $\tli \approx 2.5 x 10^6\; K$.
@@
 Gaia data release 2 (GDR2) \citep{Brown2018}, released on 2018, contained positions, parallaxes, and proper motions for about 1.3 billion sources, together with photometry.
-The third data release of Gaia has been divided in two parts: the Gaia early data release 3 (GEDR3) \citep{Brown2021} on 2020, and the Gaia data release 3 (GDR3) \citep{Brown2022} on 2022.
-They convey updated and more precise positions, parallaxes, proper motions, and photometry, spectra, classification and astrophysical parameters.
-The Gaia-ESO Survey will supplement the Gaia data for a significant subset of its targets, so that Gaia precision astrometry can be coupled with Gaia-ESO precision radial velocities and chemical abundances.
-In addition, a large fraction of the Gaia-ESO initiative is dedicated to the study of a large sample of open clusters (OCs).\par
-
-GES collected data are very handy for our research because they conform a large data set of stellar parameters and Li abundances for several OCs, covering a large age range, from a few million to several billion years \citep{Dumont2021}.
-The analysis of surface Li abundances in young OCs of different ages can provide important constraints on stellar evolutionary models.
-As mentioned, GES enhances Gaia observation and several data releases has been made available.
-Of particular interests for this paper, it is the  latest release (DR5.0).
-The DR5.0 release includes all the astrophysical parameters derived by the Gaia-ESO consortium for 114 324 stars.
-In detail the catalogue provides \citep{Gilmore2022}:
+The third data release of Gaia, the Gaia early data release 3 (GEDR3) \citep{Brown2021} and the Gaia data release 3 (GDR3) \citep{Brown2022}, convey updated and more precise measures.
+GES enhances them and in its latest release (DR5.0) includes all the astrophysical parameters derived by the Gaia-ESO consortium for 114 324 stars \citep{Gilmore2022}.
+The interest for the present work are:
 \begin{itemize}
@@
 \subsection{The selection of clusters}
-GES DR5.0 provides information on 114 324 stars, not all of which correspond to OCs, see \cite{Gilmore2022} for details.
-Of interest for the present work are those identified as belonging to OCs.
-For them we proceed to perform a first filtering on the GES\_TYPE field to identify those records that meet the condition:
+GES DR5.0 provides information on 114 324 stars corresponding to OCs, see \cite{Gilmore2022} for details. Of interest for the present work are those records that meet the condition:
 \begin{itemize}
@@
 
+\section{Method} \label{sec_method}
+The AML is considered to be governed by magnetically-coupled stellar winds which exert a braking force on the star that eventually causes a loss of angular momentum.
+In addition, throughout the star's evolution, directly observable parameters such as its rotational velocity, effective temperature, surface gravity or abundance of chemical elements undergo variations.
+These parameters have an influence, either directly or indirectly, on the angular momentum loss caused by the star's magnetic field.
+On the other hand, we have stellar convection which in most stellar evolution models is prescribed by MLT.
+That makes use of the mixing-length scale to express which in turn is proportional to the height of the local pressure scale, and to $\amlt$.
+$\amlt$ is considered a free-parameter and must be determined by comparing the stellar models with some calibrator, typically our Sun, remaining fixed.
+Therefore, assuming that the magnetic field strength ($B$) and $\amlt$ remain constant throughout the stellar evolution is a necessary simplification in some scenarios but might not correspond with reality.\par
+
+The method of work in the present work has as its initial premise the development of evolutionary stellar models in which both $B$, as well as $\amlt$ are not fixed as free parameters but their values vary over time as a function of certain stellar parameters.
+As benchmark data we take a subset of the observations obtained by GES.
+This subset consists of those stars that belong to OCs, and among them are "hidden" potential siblings of our Sun.
+To identify the latter, an additional screening is performed, this time using the values given by our models during the simulations for metallicity ($\feh$), effective temperature ($\teff$), and surface gravity ($\gsurf$).
+Finally we are left with those stars for which the GES observations have measured values congruent with our model results and from which we obtain their Li abundances.\par
+
+In the remainder of this paper we present in detail how we select those GES stars most similar to our Sun, and describe semi-empirical approximations for the calculation of the AML as a result of the braking caused by a magnetic field of varying strength, and to make the $\amlt$ value variable.
+This has been implemented as an extension to MESA.\par
+
 \subsection{The selection of closest solar siblings}
@@
 
-From the approximately 5 900 identified stars, we proceed to select those with metallicity ($\feh$), effective temperature ($\teff$), and surface gravity ($\gsurf$) within the following limits:
+From the approximately 5 900 identified stars, we proceed to select those whose $\feh$, $\teff$, and $\gsurf$ values are within the following limits:
 \begin{itemize}
@@
 
-The method of selection is as follows: Based on the values for $\teff$, $\gsurf$ and $\feh$ given by our simulations, we proceed to select in the different OCs those stars that have values within the established ranges.
+The method of selection is as follows: Based on the values given by our simulations for the for the aforementioned, we proceed to select in the different OCs those stars that have values within the established ranges.
 That is, for the estimated age of a particular cluster, we calculate an interval of $\pm$0.1 for young OCs (age < 1Gyr) and $\pm$0.15 for old OCs (age >= 1Gyr) \citep[see][and reference values therein]{Cantat-Gaudin2020}.
@@
 Each of the tables lists the GES identifier of the star and the $\teff$, $\gsurf$, $\feh$, A(Li) and estimated age ranges used in the simulation.
-
-\section{Method} \label{sec_method}
-The AML is considered to be governed by magnetically-coupled stellar winds which exert a braking force on the star that eventually causes a loss of angular momentum.
-In addition, throughout the star's evolution, directly observable parameters such as its rotational velocity, effective temperature, surface gravity or abundance of chemical elements undergo variations.
-These parameters have an influence, either directly or indirectly, on the angular momentum loss caused by the star's magnetic field.
-Therefore, assuming that the magnetic field strength of a star remains constant throughout its evolution is a necessary simplification in some scenarios but does not correspond to observations.\par
-
-In \cite{Gossage2021}, the authors also approach the problem of AML from a similar starting point to ours, but with some notable differences.
-In the Table \ref{tab:gassage_vs_navarro} we compare the most notable similarities and differences between the two studies.
-In both works the models consider rotation not only in the MS but also in the PMS, with the difference that in Gossage's work they take into account disk-locking.
-This aspect is not so important in our simulations since in our case we have focused on the A(Li) profiles.
-In contrast, Gossage's work focuses mainly on modelling rotation profiles of stars, which are significantly affected by disk-locking.
-Similarly, the effects of magnetic braking on angular momentum loss are also considered.
-In our case for stars with $\mstar$=1$\msun$ (solar-twins) and in their with 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$.
-Both models also use MLT but in our case the MLT alpha value is not pre-fixed but varies throughout the simulation.
-These two works complement each other as they are based on different formalisms.
-On the one hand, our work is based on the proposal of \cite{Gallet2013} whose most relevant parameters are Ro and the presence of a magnetic field of varying strength (B), and on the other hand Gossage presents two formalisms: \cite{Matt2015} and \cite{Garraffo2018}.
-The former is mainly characterised by $\rossby$ and the $\mstar$, while the latter by $\rossby$ and the complexity of the magnetic field ($\magfc$).
-As far as OCs observations are concerned, in this work we have had access to a larger number of them, and covered a wider range of ages [0.001-6.761] Gyr.
-Finally, both papers simulate their respective models in MESA.\par
-
-\begin{table}
-	\centering
-	\caption{Comparison with \cite{Gossage2021} work.}
-	\label{tab:gassage_vs_navarro}
-        \begin{threeparttable}
-	\begin{tabular}{lll} 
-		\hline
-		Parameter & This work & \cite{Gossage2021}\\
-		\hline
-
-		Rotation & At PMS \& MS & At PMS \& MS\\
-  		Solar Mass & $\mstar$=1$\msun$ & 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$ \\
-		Convection & MLT variable $\alpha_{{\rm MLT}}$ & MLT fixed $\alpha_{{\rm MLT}}$\\
-		Magnetic Braking & \cite{Gallet2013} & \cite{Matt2015}, \\ Formalism & & \cite{Garraffo2018}\\
-            Magnetic Braking & $\rossbystar$ \& $\bstar$ & $\rossbystar$ \& $\mstar$, \\ Parameters & & $\rossbystar$ \& $\magfc$\tnote{(1)}\\
-		Magnetic Field & B(G) variable & Not used\\ Strength & & \\
-            Disk locking & Not present & Present\\
-            Number of OC's & 64 & 11 \\
-            Source of data & GDR3.0 GES DR5.0 & Several works \\
-            MESA version & r10398 & r11701\\
-		\hline
-	\end{tabular}
-        \end{threeparttable}
-        \begin{tablenotes}\footnotesize
-        \item (1) Magnetic field complexity
-        \end{tablenotes}
-\end{table}
-
-In the remainder of this paper, we describe a semi-empirical approach for the calculation of the AML as a result of the braking caused by a magnetic field of varying strength.
-The magnetic field evolves according to the $\teff$ and angular velocity ($\Omega$) of the star.
-This has been implemented as an extension to MESA.\par
 
@@
 
-\subsection{Li evolution without MB}
-As a baseline, we borrow from \cite{Navarro2020} the Figure \ref{fig:li_var_vel_0g} which shows the temporal evolution of surface Li abundance for several $1\msun$ models initialized with different rotational velocities.
-The simulations took into account the effects of rotation and AML caused by stellar winds but not those of MB.
-The purple star and square are the surface Li abundances for the present-day Sun \citep{Asplund2009} and for the Pleiades, respectively \citep{Sestito2005}.
-For more details, refer to the cited work.\par
-
-\begin{figure}
-	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip, width=\columnwidth]{figures/paper1/li_var_vel_0_0g.pdf}
-    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
-The solid black line represents the reference model according to \citet{Choi2016}.
-The rest of lines are models which include PMS rotation with $\oomegac$ between 0.0084 and 0.0336, respectively.
-The purple star and square are surface Li abundances for the present-day Sun \citep{Asplund2009} and the average for the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:li_var_vel_0g}
-\end{figure}
-
-The Li abundance in the star surface decreased over time for all simulated models.
-The solid black line represents the reference model that adopts the solar-calibrated envelope overshooting parameters as documented in \citet{Choi2016}.
-All models burned too much Li before the ZAMS and therefore did not match with the Pleiades average surface Li abundance.
-There were barely any differences between the different models in terms of the abundance of Li for much of the PMS.
-Only after one million years was Li destroyed to an accentuated degree, since the necessary temperature was not reached at BCZ before.
-Afterwards the different models destroyed the Li in a very similar way due to two main reasons.
-On the one hand, the convective zones that the models developed had very similar sizes.
-On the other hand, the differential rotation between the core and the convective zone did not begin to develop until it reached $ \approx 10^6$ years, achieving its maximum difference in the ZAMS about $ \approx 10^7$ years.
-It was at this moment that the combined effects of turbulence and rotational difference increase, leading to notable differences in the evolution of A(Li) (see Figure \ref{fig:li_var_vel_0g_z1}).
-Later, the reference model (black line) did not deplete Li efficiently on the MS.
-The other models were able to burn Li in a more realistic way.
-Among them only one (green line) was close to match the present-day Li abundance of the Sun but its rotational velocity was much higher (see Figure \ref{fig:rot_vel_0g}) than the $2\,\kms$ of the Sun \citep{Gill2012}.
-\par
-
-\begin{figure}
-	\includegraphics[trim = 25mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/paper1/li_var_vel_0_0g_z1.pdf}
-    \caption {Similar to Figure \ref{fig:li_var_vel_0g} but zooming in on the ZAMS.
-The models with a higher initial rotational velocity reach already the ZAMS with a lower amount of Li measure on the star surface.}
-    \label{fig:li_var_vel_0g_z1}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[trim = 10mm 10mm 15mm 10mm clip,width=\columnwidth]{figures/paper1/rot_vel_var_vel_0_0g.pdf}
-    \caption{The evolution of angular velocity at surface (solid line) and at the bottom (dotted line) limit of the uppermost convective zone, as a function of time for several 1 $\msun$ models.
-The models includes PMS rotation with $\oomegac$ values between 0.0084 and 0.0336.
-The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:rot_vel_0g}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/paper1/hr_var_vel_0_0g_z1.pdf}
-    \caption{An example solar 1$\msun$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities.
-The rotation is activated in the models in the PMS and those models reach before the ZAMS and at a lower $\teff$ than the non-rotating one (solid black line).
-The luminosity is expressed in terms of $\lsun$.}
-    \label{fig:hr_var_vel_0g}
-\end{figure}
-
 \subsection{Li evolution with MB}
@@
 These effects combine and lead to an accentuation of the magnetic braking effect.
-As a consequence, there is a slowing down of Li destruction once the models enter the MS.\par
-
-\begin{figure}
-	\includegraphics[clip,width=\columnwidth]{figures/paper2/li_var_vel_var_g_3.pdf}
+As a consequence, there is a slowing down of Li destruction once the models enter the MS.
+The results of our simulations for those models respectively initialized to $\oomegac$ = 0.14 and 0.1425 reproduce A(Li) in line with that of the Sun surface \isotope[7]{Li} abundance  (1.1 $\pm$ 0.1 dex).
+The latter gives a value of 1.133 dex, which would represent a deviation of about 3\% from the nominal value.\par
+
+\begin{figure}
+	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip, width=\columnwidth]{figures/paper2/li_var_vel_0_0g_0.pdf}
     \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
-The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425, respectively and MB.
-The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The rest of the dots in different colours represent surface \isotope[7]{Li} abundances of stars with similar parameters to the evolution curve represented with the same colour.The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:li_var_vel_var_g_3}
+The solid black line represents the reference model according to \citet{Choi2016}.
+The rest of lines are models which include PMS rotation with $\oomegac$ between 0.0084 and 0.0336, respectively.
+The purple star and square are surface Li abundances for the present-day Sun \citep{Asplund2009} and the average for the Pleiades cluster \citep{Sestito2005} respectively.
+The dashed vertical line makes reference to the ZAMS.
+(Figure taken from \citet{Navarro2020}.)}
+    \label{fig:li_var_vel_0g}
 \end{figure}
@@
 The combined effects of the increases in effective temperature, angular velocity and mass loss cause the magnetic field strength to reach its maximum in the approach phase of the ZAMS (see Figure \ref{fig:mag_field_var_vel_g3}).
-We observed that the models with lower angular velocity generally ended up exhibiting higher values for the Li abundance on the surface (see Figures~ \ref{fig:li_var_vel_var_g_3}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).\par
-
-\begin{figure}
-	\includegraphics[clip,width=\columnwidth]{figures/paper2/rot_vel_var_vel_var_g3.pdf}
-    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
+We observed that the models with lower angular velocity generally ended up exhibiting higher values for the Li abundance on the surface (see Figures~ \ref{fig:li_var_vel_var_g_3}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).
+For these same simulated models initialized to $\oomegac$ = 0.14 and 0.1425, the values obtained for the $\Omega$ and the magnetic field strength are far from the measurements obtained for the Sun.
+$\osun$ is 2km/s with a minimum error of 3m/s, and the mean magnetic field strength is 1G.
+For the simulation with $\oomegac$=0.1425, we obtain a rotational velocity of 4.72km/s, which represents a 235\% deviation, and for the average magnetic field strength we have 36.9G, far away from the reference value.\par
+
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/li_var_vel_var_g_3.pdf}
+    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
 The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425, respectively and MB.
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:rot_vel_var_vel_var_g3}
+The rest of the dots in different colours represent surface \isotope[7]{Li} abundances of stars with similar parameters to the evolution curve represented with the same colour.The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:li_var_vel_var_g_3}
 \end{figure}
@@
 In our models the magnetic braking routine is activated when the star develops a radiative core and the existence of a convective zone above it (see Figure \ref{fig:mb_act_var_vel_g3}).
-It is the presence of the MB that prevents the star from further increasing its angular velocity, as in the models without MB (see Figure \ref{fig:rot_vel_0g}).
+It is the presence of the MB that prevents the star from further increasing its angular velocity.
 In this way, the Li destruction is attenuated by a lower rotational velocity.
@@
 \begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/rot_vel_var_vel_var_g3.pdf}
+    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
+The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425, respectively and MB.
+The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:rot_vel_var_vel_var_g3}
+\end{figure}
+
+\begin{figure}
 	\includegraphics[clip,width=\columnwidth]{figures/paper2/mb_act_var_vel_g3.pdf}
@@
 
-\begin{figure}
-	\includegraphics[clip,width=\columnwidth]{figures/paper2/lograd_var_vel_g3.pdf}
-    \caption{The evolution of the stellar radius, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
-The models include PMS rotation with $\oomegac$ between 0.12 and 0.1425.
-The purple star is the radius for the present-day Sun \citep{Gill2012}.
-The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:lograd_var_vel_g3}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[clip,width=\columnwidth]{figures/paper2/lograd_var_vel_g_z13.pdf}
-    \caption{Similar to \ref{fig:lograd_var_vel_g3} but zooming in on the ZAMS. The fastest model exhibits the smallest radius from around $4.5x10^{7}$ till $11.0x10^{7}$ years.}
-    \label{fig:lograd_var_vel_g_z13}
-\end{figure}
-
 The effects of rotation and magnetic braking can also be seen in the H-R diagram and in the evolution of the $\amlt$.
@@
 This effect causes those stars that rotate fast enough to appear less dense ($\rho$), less luminous ($L$) and therefore with a lower effective temperature ($\teff$) as they approach the ZAMS (see Figure \ref{fig:hr_var_vel_var_g_z13}).
-If we compare the non-rotating model (black solid line in Figure \ref{fig:rot_vel_0g}) with the rotating ones we can recognize that at the end of the PMS, the latter reach the ZAMS with a lower $\teff$ than the former.
+If we compare the non-rotating model (black solid line in Figure \ref{fig:hr_var_vel_0g}) with the rotating ones we can recognize that at the end of the PMS, the latter reach the ZAMS with a lower $\teff$ than the former.
 This also has an impact on the evolution of $\amlt$ because, as discussed above, in our model its value depends proportionally on $\teff$ and $\gsurf$.
@@
     \label{fig:hr_var_vel_var_g_z13}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/paper2/hr_var_vel_0_0g_z10.pdf}
+    \caption{An example solar 1$\msun$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities.
+The rotation is activated in the models in the PMS and those models reach before the ZAMS and at a lower $\teff$ than the non-rotating one (solid black line).
+The luminosity is expressed in terms of $\lsun$.
+(Figure taken from \citet{Navarro2020}.)}
+    \label{fig:hr_var_vel_0g}
 \end{figure}
@@
 
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/lograd_var_vel_g3.pdf}
+    \caption{The evolution of the stellar radius, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
+The models include PMS rotation with $\oomegac$ between 0.12 and 0.1425.
+The purple star is the radius for the present-day Sun \citep{Gill2012}.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:lograd_var_vel_g3}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/lograd_var_vel_g_z13.pdf}
+    \caption{Similar to Figure \ref{fig:lograd_var_vel_g3} but zooming in on the ZAMS. The fastest model exhibits the smallest radius from around $4.5x10^{7}$ till $11.0x10^{7}$ years.}
+    \label{fig:lograd_var_vel_g_z13}
+\end{figure}
+
+\subsection{Gossage et al. (2021) Comparison}
+In \cite{Gossage2021}, the authors also approach the problem of AML from a similar starting point to ours, but with some notable differences.
+In the Table \ref{tab:gassage_vs_navarro} we compare the most notable similarities and differences between the two studies.
+In both works the models consider rotation not only in the MS but also in the PMS, with the difference that in Gossage's work they take into account disk-locking.
+This aspect is not so important in our simulations since in our case we have focused on the A(Li) profiles.
+In contrast, Gossage's work focuses mainly on modelling rotation profiles of stars, which are significantly affected by disk-locking.
+Similarly, the effects of magnetic braking on angular momentum loss are also considered.
+In our case for stars with 1$\msun$ (solar-twins) and in their with 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$.
+Both models also use MLT but in our case the MLT alpha value is not pre-fixed but varies throughout the simulation.
+These two works complement each other as they are based on different formalisms.
+On the one hand, our work is based on the proposal of \cite{Gallet2013} whose most relevant parameters are Ro and the presence of a magnetic field of varying strength (B), and on the other hand Gossage presents two formalisms: \cite{Matt2015} and \cite{Garraffo2018}.
+The former is mainly characterised by $\rossby$ and the $\mstar$, while the latter by $\rossby$ and the complexity of the magnetic field ($\magfc$).
+As far as OCs observations are concerned, in this work we have had access to a larger number of them, and covered a wider range of ages [0.001-6.761] Gyr.
+Finally, both papers simulate their respective models in MESA.\par
+
+\begin{table}
+	\centering
+	\caption{Comparison with \cite{Gossage2021} work.}
+	\label{tab:gassage_vs_navarro}
+        \begin{threeparttable}
+	\begin{tabular}{lll} 
+		\hline
+		Parameter & This work & \cite{Gossage2021}\\
+		\hline
+
+		Rotation & At PMS \& MS & At PMS \& MS\\
+  		Solar Mass & $\mstar$=1$\msun$ & 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$ \\
+		Convection & MLT variable $\alpha_{{\rm MLT}}$ & MLT fixed $\alpha_{{\rm MLT}}$\\
+		Magnetic Braking & \cite{Gallet2013} & \cite{Matt2015}, \\ Formalism & & \cite{Garraffo2018}\\
+            Magnetic Braking & $\rossbystar$ \& $\bstar$ & $\rossbystar$ \& $\mstar$, \\ Parameters & & $\rossbystar$ \& $\magfc$\tnote{(1)}\\
+		Magnetic Field & B(G) variable & Not used\\ Strength & & \\
+            Disk locking & Not present & Present\\
+            Number of OC's & 64 & 11 \\
+            Source of data & GDR3.0 GES DR5.0 & Several works \\
+            MESA version & r10398 & r11701\\
+		\hline
+	\end{tabular}
+        \end{threeparttable}
+        \begin{tablenotes}\footnotesize
+        \item (1) Magnetic field complexity
+        \end{tablenotes}
+\end{table}
+
 \section{Conclusions and future works} \label{sec_conclusions}
@@
 We cross-checked the model results with measurements obtained from the Gaia and GES mission for stars with similar characteristics to our Sun, focusing on their Li abundances.
-Our models seem to suggest a lower limit to the surface Li abundances for Sun-like stars, and that the Sun indeed appears to have a lower surface Li abundance than the rest.
-They also suggest that the Sun had a much higher angular velocity than it currently shows and that as a result it effectively destroyed its Li reservoirs.
-Coupled with this higher angular velocity, strong magnetic fields developed, triggering an intensive magnetic braking that slowed its angular velocity to its present value.\par
-
-Yet our models still seem to be destroying too much Li when comparing the Li abundances obtained with those provided by the Gaia and GES data.
-Taking as a starting point that faster rotating stars destroy more Li, other mechanisms in addition to magnetic braking are necessary to explain the observational data.
-Among them, disk locking to the protostellar disk contributes to counteract the acceleration of the star during PMS.
-Thus, we would have two braking mechanisms acting at different ages of the star.
-The disk locking would act during the PMS and the magnetic braking mainly during the MS.
-This approach would establish a self-regulating mechanism over the angular velocity of the star that would end up directly influencing the Li evolution.\par
-
-We emphasize as well that our models are close to match at the same time the observed solar Li abundance and $\Omega$, but they are not yet accurate enough.
-Therefore, we must analyse the results obtained with caution and not draw any premature conclusions.
-In addition, it should be noted that our models fail to replicate the average magnetic field strength of the Sun of $1.0\,\Gauss$.
-We have also shown how both an evolving magnetic field strength and, a MLT parametrization could produce results in line with the observations.
-The following conclusions have been supported in the course of this work:
+
+Our models seem to suggest a lower limit to the surface Li abundances of 1.133 dex for Sun-like stars, value which is within the margin of error obtained for the Sun (1.1 $\pm$ 0.1 dex).
+On the other hand, they predict a higher angular velocity (4.72 km/s) than the Sun (2.0 km/s) so that they could burn more Li and finally reproduced the Sun Li existing reservoirs.
+Coupled with this higher angular velocity, strong magnetic fields developed, triggering an intensive magnetic braking that slowed its angular velocity but not enough to match the present $\Omega$ value of the Sun.
+In addition, it should be noted that our models fail to replicate the average magnetic field strength of the Sun.\par
+
+As far as Li abundance and $\Omega$ are concerned, our models, while not exactly reproducing the values of the Sun, show promising results.
+Although we have shown how both an evolving magnetic field strength and, a MLT parametrization could produce results in line with the observations, we must analyse them results obtained with caution and not draw any premature conclusions.
+Nonetheless, we can draw the following finding from the present work:
 
@@
     \item A magnetic field with varying intensity is a valid alternative to explain the rotational history of Sun-like stars.
-    \item It is a plausible line of research to provide an answer to the problem of surface Li on the Sun.
     \item A variable $\amlt$ eliminates the need to pre-fix its value in advance for the entire evolution of the star.
@@
 \end{enumerate}
-\par
+
+We would like to conclude with the following thoughts for the next steps.
+It is a fact that our models still destroy too much Li when comparing the produced Li abundances obtained with those from Gaia and GES.
+Taking as starting point that faster rotating stars destroy more Li, other mechanisms, in addition to magnetic braking, are necessary to explain the observational data.
+Among them, disk locking to the protostellar disk could contribute to counteract the acceleration of the star during PMS.
+Thus, we would have two braking mechanisms acting at different ages of the star.
+The disk locking would act during the PMS and the magnetic braking mainly during the MS.
+This approach would establish a self-regulating mechanism over the angular velocity of the star that would end up directly influencing the both Li evolution and the magnetic field strength.
+For the latter, lower intensities are expected which could finally converge with that of our Sun.
+On the other hand, a lower rotational speed also means less Li destruction for our models.
+One way to compensate for this lower Li remaining is to resort to over-shooting (see \cite{Navarro2020}) an effect that has not been simulated in our models.
+By activating this mechanism, we would allow the Li to be reached by bubbles of material hot enough to destroy it, thus contributing to decrease its abundance.
+Ideally, as we have done with $\amlt$, overshooting should not be introduced as a free parameter but dependent on other fundamental stellar parameters.\par
 
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/paper1/li_var_vel_0_0g.pdf}
+    \includegraphics[trim = 25mm 10mm 15mm 10mm, clip,width=\textwidth]{figures/paper2/li_var_vel_0_0g_0.pdf}
     \label{fig:subim13}
@@
     \begin{subfigure}[h]{0.47\textwidth}
-    \includegraphics[clip,width=\textwidth]{figures/paper1/rot_vel_var_vel_0_0g.pdf}
+    \includegraphics[clip,width=\textwidth]{figures/paper2/rot_vel_var_vel_0_0g_0.pdf}
     \label{fig:subim23}

### 2024-01-29 project/overleaf/Tesis III/capitulos/capitulo01.tex
Add Tesis project

% !TeX root = ../libro.tex
% !TeX encoding = utf8

\chapter{¿Por qué el Litio?}\label{ch:primer-capitulo}

\section{Introducción}
El estudio de las abundancias de Litio en las estrellas, particularmente aquellas similares al Sol, es crucial por varias razones. En primer lugar, el Litio es uno de los pocos elementos producidos en la nucleosíntesis del Big Bang. Su abundancia proporciona una prueba crítica para la teoría del Big Bang y nos permite sondear las condiciones del universo primitivo. Además, el Litio es sensible a las temperaturas estelares, y su tasa de destrucción aumenta rápidamente a temperaturas superiores a 2.5 millones de Kelvin, que son típicas para los interiores estelares.\par

En segundo lugar, la abundancia de Litio puede arrojar luz sobre la estructura interna y los procesos de mezcla dentro de las estrellas. En los modelos estelares estándar, se espera que el Litio se agote en el sobre convectivo exterior de una estrella debido a su transporte a capas más profundas y calientes donde se destruye. Sin embargo, las observaciones a menudo muestran más agotamiento de Litio del que predicen estos modelos, lo que sugiere procesos de mezcla adicionales. Estudiar las abundancias de Litio puede ayudar a refinar nuestra comprensión de estos procesos y mejorar los modelos estelares.\par

El estudio de las abundancias de Litio es particularmente relevante al examinar cúmulos abiertos. Los cúmulos abiertos son grupos de estrellas que se han formado a partir de la misma nube molecular gigante, lo que significa que comparten una edad y composición química inicial comunes. Esto los convierte en excelentes laboratorios para estudiar la evolución estelar y la nucleosíntesis. El contenido de Litio en estas estrellas puede proporcionar valiosos conocimientos sobre los procesos de mezcla interna y la edad del cúmulo. A medida que las estrellas de un cúmulo envejecen, su abundancia de Litio superficial disminuye debido a la mezcla y la quema. Al comparar las abundancias de Litio observadas en las estrellas de un cúmulo abierto con modelos teóricos, podemos estimar la edad del cúmulo y obtener conocimientos sobre la eficiencia de los procesos de mezcla.\par

Además, dado que todas las estrellas en un cúmulo inicialmente tienen la misma abundancia de Litio, cualquier diferencia observada debe ser debido a procesos que ocurren dentro de las estrellas. Esto nos permite investigar cómo factores como la masa, la temperatura, la rotación y la presencia de campos magnéticos afectan a la destrucción de Litio, contribuyendo a nuestra comprensión de los interiores estelares. Por lo tanto, el estudio de las abundancias de Litio en cúmulos abiertos juega un papel crucial en el avance de nuestro conocimiento de las estrellas y su evolución.\par

aaaa \gls{lcm} \acrshort{gcd} \Gls{latex}

\endinput
%--------------------------------------------------------------------
% FIN DEL CAPÍTULO. 
%--------------------------------------------------------------------

### 2024-01-29 project/overleaf/Tesis III/capitulos/capitulo02.tex
Add Tesis project

% !TeX root = ../libro.tex
% !TeX encoding = utf8

\chapter{El ciclo del litio}\label{ch:segundo-capitulo}

\section{Introducción}
La abundancia de Li observada en la fotosfera de las estrellas es un indicador de su composición interior y de los procesos de mezclado que en su él tienen lugar. Adicionalmente, estas abundancias (además de otras métricas) se utilizan para comprobar la validez de los modelos estelares. Para que esto sea posible hay que tomar como premisa inicial que la abundancia de Li generada en la nucleosíntesis asociada al Big Bang es conocida y este elemento solo se destruye a través de reacciones nucleares. A pesar de décadas de esfuerzos teóricos, se sigue sin encontrar una explicación coherente con los modelos para las discrepancias encontradas en las comparaciones sobre la abundancia de Li para estrellas pertenecientes a cúmulos de diferentes edades y que se encuentran en el mismo estado evolutivo, o bien en la presecuencia principal (Pre-Main Sequence, PMS), o bien en la secuencia principal (Main Sequence, MS). Los modelos no son capaces de explicar las abundancias detectadas en las etapas tardías de la MS \cite{Tschape2001}.\par

Es sabido que parte de la pérdida de Li se produce durante la PMS y que además ésta se acentúa según decrece la masa de la estrella y, a igualdad de masa, según aumenta la metalicidad de la misma. Para reproducir la dependencia entre la edad y masa de las estrellas con la merma de las concentraciones de Li, estos modelos necesitan de una combinación de procesos de mezclado cada vez más complejos, como overshooting, mezclado debido a procesos rotacionales o difusión microscópica, o la presencia de campos magnéticos.A pesar de décadas de esfuerzos teóricos, aún no se ha podido encontrar una explicación coherente derivada de modelos para las discrepancias de abundancia de Li detectadas en estrellas pertenecientes a cúmulos de diferentes edades y que se encuentran en el mismo estado evolutivo, ya sea en la PMS o en la MS. Además, estos modelos teóricos no son capaces de explicar las abundancias de elementos químicos detectadas en las etapas tardías de la MS \cite{Tschape2001}.\par

En la presente tesis estamos interesados en investigar cómo la presencia de campos magnéticos pueden influir en las abundancias de Li detectadas en las atmósferas de las estrellas de tipo solar. El Li es un elemento que se destruye fácilmente en las capas interiores de este tipo de estrellas. Este proceso ocurre principalmente durante la PMS en mayor medida, aunque puede darse también durante la MS. Adicionalmente, cabe la posibilidad de que se dé en las capas exteriores si existe un proceso eficiente de mezclado en la estrella, en el que puede influir la presencia de campos magnéticos. Por tanto, el estudio de la abundancia de Li en la superficie de la estrella puede ser clave para entender el proceso de evolución del momento angular de la misma.\par

Los denominados modelos estándar, modelos que incluyen la convección sólo como proceso de mezcla y no consideran otras opciones de transporte como la difusión o la pérdida de momento angular (Angular Moment Loss, AML), han sido los principales implicados en la elaboración de estas predicciones \cite{Sestito2005}. El Li se destruye a una temperatura $\tli \approx 2.5 x 10^6\; K$ cuando un átomo de Li colisiona con un protón produciendo dos átomos de He, algo que tiene lugar durante las reacciones protón-protón tipo II (P-P II), y por tanto se destruye directamente en las envolturas estelares cuando la temperatura en la base de la zona de convección (Base Convective Zone, BCZ) alcanza ese valor. El Sol en particular, y las estrellas de tipo solar en general, se caracterizan por tener una CZ que cubre gran parte del radio estelar durante la PMS lo que hace que su límite inferior supere $\tli$ \cite{Iben1965}. Este agotamiento se detiene durante la aproximación al inicio de la MS (Zero-Age Main Sequence, ZAMS) cuando la zona de convección retrocede y la temperatura en la BCZ es más fría que $\tli$. En los modelos estelares estándar sólo la masa y la composición química inicial determinan a qué distancia de la superficie estelar se alcanza la temperatura $\tli$, por lo que se espera que estrellas dentro de cúmulos con masa similar alcancen la ZAMS con iguales abundancias superficiales de Li. Además, también deberían mostrar una evolución del Li muy similar hasta su aproximación a la edad terminal de la MS (Terminal-Age Main Sequence, TAMS). Durante este mismo periodo los mecanismos de convección desencadenan un proceso de mezcla que homogeneiza la composición química de la envoltura convectiva, desde su límite inferior hasta la superficie estelar. Sin embargo, se han observado diferentes abundancias de Li en distintas poblaciones estelares \cite[][y referencias en las mismas]{Somers2014}.\par

\section{El Li primordial}
Los modelos teóricos de estrellas no informan sobre la cantidad inicial de Li que tiene una estrella, sólo describen cómo se agota. Por tanto, para hacer una estimación precisa de la abundancia inicial de Li, es un requisito poder comparar previamente las observaciones y los modelos. Nuestro Sol representa una excepción única, ya que nos permite conocer la abundancia actual de este elemento en su fotosfera, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex \footnote{Contracción procedente del inglés para decimal exponent.}$ \cite{Jeffries2004}, donde $A(\isotope[7]{Li})$ se define según la Ec.~\ref{eq:A_Li}\par

\begin{ceqn}
	\begin{equation}
		A(\isotope[7]{Li}) = log\frac{N(Li)}{N(H)} + 12
		\label{eq:A_Li}
	\end{equation}
\end{ceqn}
donde $N$ es el número de átomos del elemento en cuestión por unidad de volumen y 12 representa la abundancia del H, el cual se toma como punto de referencia para evitar que aparezcan números negativos.\par

Por otro lado, la abundancia inicial de $A(\isotope[7]{Li}) = 3.34 \, dex$ se obtiene a partir de mediciones de meteoritos \cite{Randich2006}. Según la teoría establecida para las estrellas recién nacidas, tenemos que la abundancia inicial de Li puede estimarse con bastante precisión a partir de mediciones fotosféricas en estrellas de tipo T-Tauri, o de las estrellas F más calientes que forman parte de cúmulos algo más antiguos. Para estas últimas, la teoría actual de la evolución estelar sugiere que el Li aún no debe estar agotado. Los resultados obtenidos a partir de las medidas en ambos tipos de estrellas nos permiten fijar la abundancia inicial de Li en el intervalo $3.0 \, dex < A(\isotope[7]{Li}) < 3.4 \, dex$ \cite{Randich2006}.\par

\section{Evolución del Li en la presecuencia principal (PMS)}
La PMS es la continuación directa de la fase protoestelar. Inicialmente, a medida que el objeto estelar joven evoluciona a lo largo del tiempo, su luminosidad es fundamentalmente consecuencia directa de un proceso de acreción suficientemente fuerte que es capaz de mantener la fusión activa de deuterio \cite{Stahler1983}. Una vez que cesa la acumulación de material, la protoestrella, rodeada sólo por un disco residual de material, desciende a lo largo de su traza de Hayashi de una forma casi vertical. Esta fase de la evolución estelar se localiza en el diagrama Hertzspring-Russell \footnote{Diagrama en el que se enfrenta la luminosidad de la estrella contra su temperatura} y en la misma la estrella es totalmente convectiva. A medida que la estrella desciende por la traza de Hayashi se van contrayendo y su radio disminuye, a la par que lo hace su luminosidad. Su temperatura superficial ($\teff$)tiende a mantenerse  en torno $3000..5000 K$ aunque la interior va en aumento a consecuencia de la contracción del gas que conforma la estrella y el aumento de presión. Llegado un punto con suficiente temperatura en el interior de la estrella, ésta desarrolla una zona convectiva y la estrella entra en la traza de Heney, donde se producen reacciones nucleares, marcando así la entrada en la MS. \par

Durante la evolución estelar en esta fase, el Li se quema a temperaturas relativamente bajas (2.5 - 3.0 $10^{6} K$) y, en las estrellas de baja masa (< 1.2 $\msun$), el mezclado que se produce en la zona convectiva de la estrella puede hacer que el material ya agotado de Li alcance en relativamente poco tiempo la fotosfera de la estrella. Por esta razón, las mediciones de abundancia de Li fotosférica nos brindan uno de los pocos métodos de sondeo de interiores estelares y, a su vez, representan un importante banco de pruebas contra el que enfrentar los resultados que se obtienen a partir de los modelos evolutivos para PMS. Entender cómo se produce el agotamiento del Li durante la PMS también nos ofrece la posibilidad de poder estimar la edad de las estrellas más jóvenes y, por supuesto, es un punto de partida para poder cuantificar cualquier agotamiento subsiguiente de Li que se produzca a lo largo de la MS.\par

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{img/tesis/isochrones.pdf}
	\caption {Secuencias evolutivas (en $\msun$) e isócronas (en Myr) para estrellas de baja masa comprendidas entre 0.4 y 1.2 $\msun$ y metalicidad Z=0.02. Se indican las franjas temporales en las que se produce el agotamiento fotosférico de Li y el desarrollo de un núcleo radiativo. Los números a la derecha de los intervalos de destrucción de Li indican la fracción fotosférica de éste que queda en el punto donde se desarrolla el núcleo radiativo y al final de su combustión, figura de Jeffries (2004).}
	\label{fig:isocrhones}
\end{figure}

La Figura \ref{fig:isocrhones} muestra la fase de PMS para modelos de distintas masas iniciales. Asimismo, se indican la fase de quema de Li y el momento de aparición del núcleo radiativo. Como se puede ver, las estrellas con M < 0.35 $\msun$ que se encuentran en la PMS tienen una estructura relativamente sencilla: son totalmente convectivas hasta que consiguen alcanzar la Termination Age Main Sequence (TAMS). A medida que la estrella va reduciendo su volumen según desciende a lo largo de la traza de Hayashi, su núcleo se va calentando, pero el gradiente de temperatura se mantiene muy cerca del gradiente adiabático, el cual delimita la frontera entre un núcleo radiativo o convectivo. Esta situación se mantiene para la práctica totalidad de la estrella, con la excepción de las capas superficiales. El Li comienza a arder en reacciones de captura protónica cuando la temperatura central, Tc=2.5 $10^{6} K$ y, debido a que la tasa de reacción es tan sensible a la temperatura para densidades típicas de PMS \cite{Randich2006} y a que el mezclado convectivo se produce de forma tan rápida, todo el Li se quema en una pequeña fracción de tiempo. Por otra parte, se ha contrastado que la edad a la que se produce el agotamiento de Li es inversamente proporcional a la masa de la estrella, es decir, aumenta según disminuye la masa hasta que se alcanza el límite de M < 0.06 $\msun$, ya que en este tipo de estrellas nunca llegará a alcanzarse una temperatura lo suficientemente alta como para poder quemar Li.\par

En las estrellas M > 0.35 $\msun$, el proceso de agotamiento del Li es bastante más complejo. Tienen densidades centrales más bajas y a medida que la Tc aumenta durante la contracción del PMS, la opacidad cae lo suficiente como para que el gradiente de temperatura se vuelva subadiabático, es decir, inhibe el proceso de mezclado y se forma un núcleo radiativo que empuja hacia afuera para incluir una fracción rápidamente creciente de la masa estelar.  Para M <  $\msun$ solo existe una pequeña ventana de oportunidad en la que poder quemar un poco de Li antes de que se desarrolle el núcleo radiativo (aproximadamente a 2 Myr para 1 $\msun$). Como se puede observar en la Figura \ref{fig:isocrhones}, para M < 0.6 $\msun$ el núcleo radiativo se desarrolla antes de que la quema de Li esté completa. A falta de mezcla por convección, el material agotado de Li no puede llegar a la fotosfera y una vez que la temperatura en la base de la zona de convección cae por debajo del umbral de combustión de Li, el agotamiento de Li fotosférico se detiene. Para una estrella de 1 $\msun$, el agotamiento de Li fotosférico comienza aproximadamente a los 2 Myr y debería de terminar al alcanzar los 15 Myr.\par

Los estudios sobre el Li en los cúmulos muy jóvenes indican que las estrellas de tipo solar sufren muy poco (si es que lo hay) agotamiento de Li durante las fases de la PMS  \cite{Jeffries2004}. Al mismo tiempo, se ha dedicado una gran cantidad de esfuerzo observacional al muestreo de abundancia de Li y también del Be \cite{Mena2011, DelgadoMena2014} con el objetivo de poner restricciones empíricas más severas a los mecanismos propuestos.

\section{Evolución del Li en la secuencia principal (MS)}
Como ya mencionó en el trabajo \cite{Zappala1972} hace algo más de cuatro décadas, la medición de Li en cúmulos más antiguos que los Hyades es una herramienta clave para investigar el agotamiento que sufre el Li en la MS, además de las escalas de tiempo en las que ocurre este proceso. Ahora, más de 40 años después, se han llevado nuevas mediciones de Li para varios cúmulos antiguos, incluyendo NGC 752 y M 67 \cite{Sestito2006} que han permitido caracterizar de manera más completa la dispersión de Li en estrellas de tipo solar y con metalicidad y edad similar al Sol \cite{Sestito2005}.\par

Basándonos en los resultados de los modelos de evolución estándar que solo tienen en cuenta los procesos de convección como los mecanismos a través de los cuales se produce el mezclado, solo debería agotarse el Li en la PMS, ya que es en esta fase cuando las estrellas exhiben una envolvente convectiva profunda con una base lo suficientemente caliente como para quemar el Li. Por otra parte, estos mismos modelos predicen que este proceso se detiene en la MS para estrellas de tipo solar o aquéllas más calientes. Sin embargo, los estudios llevados a cabo con cúmulos abiertos han revelado A(Li) 10 veces inferior respecto a los valores esperados en estrellas en su MS \cite{Sestito2005}, y de hecho, nuestro propio Sol presenta un agotamiento una A(Li) 100 veces inferior si comparamos con la abundancia de Li detectada en los meteoritos \cite{Lodders2003}.\par

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{img/tesis/li_abundances_vs_age.pdf}
	\caption{Abundancia media de Li en función de la edad. Las masas y edades han sido estimadas a partir de las temperaturas efectivas hacienda uso de isócronas. Las abundancias de Li se denotan mediante la notación log n(Li) = N (Li)/N (H) + 12. Los diferentes símbolos indican estrellas con diferentes masas, a saber: 1$\pm$0.02 $\msun$ (círculos), 1.05$\pm$0.02 $\msun$ (cuadrados) y 1.1 $\pm$0.02 $\msun$ (triángulos). El Sol ($\odot$) también se muestra. La línea horizontal indica la abundancia primigenia de Li, figura de \cite{Randich2006}.}
	\label{fig:li_abundances_vs_age}
\end{figure}

La Figura \ref{fig:li_abundances_vs_age} muestra que la evolución del Li desde aproximadamente 100 Myr hasta 6-8 Gyr es similar para los 3 rangos de masas. Se observa que los valores log n(Li) declinan a una ratio casi constante a medida que la estrella envejece. Esto ocurre así hasta aproximadamente los 2 Gyr, momento a partir del cual el agotamiento del Li parece virtualmente detenerse y las abundancias tienden a converger hacia los valores del Spite plateau, la línea de base en la A(Li) encontrada para estrellas antiguas (o de población II) y que por tanto se formaron a partir de material no alterado por otros procesos, que orbitan el halo galáctico. \par

\section{Mecanismos de destrucción de Li}
En este apartado procedemos a enumerar las diferentes opciones que actualmente defienden las principales líneas de investigación para intentar establecer una relación causa-efecto para la discrepancia de la abundancia de Li detectada en las fotosferas de estrellas de tipo solar.\par

\subsection{Li y metalicidad}
Durante más de tres décadas, las estrellas con baja metalicidad ([Fe/H] < -1.0 dex), también denominadas de Población II, se han caracterizado por exhibir una relación bastante estable entre la abundancia de Li y su metalicidad \cite{Guiglion2016}. Este nivel quedó inicialmente vinculado a la abundancia primordial asociada al Li, aunque esta situación ha cambiado a la luz de los resultados obtenidos por la misión Wilkinson Microwave Anisotropy Probe (WMAP), que han propiciado que la abundancia de este elemento predicha por el modelo Estándar de Nucleosíntesis para el Big Bang (SBBN) haya pasado a estimarse en 2.6 dex \cite{Spergel2003}. Este dato, sin embargo, entra en contradicción con la abundancia de Li medida en enanas viejas de baja metalicidad  (A(Li) $\simeq$ 2.2 dex) situadas en el halo de la galaxia conocido como el Spite plateau \cite{Spite1982}. Además, para estrellas de baja metalicidad (\ref{fig:li_abundances_sbbn}) se ha encontrado que la abundancia de Li no parece guardar relación con la metalicidad de la estrella \cite{Fu2015}.\par

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{img/tesis/li_abundances_sbbn.pdf}
	\caption{La abundancia de Li derivada para estrellas de baja metalicidad situadas en la secuencia principal es aproximadamente tres veces menor que el valor de Li primordial predicho por la SBBN, figura de \cite{Fu2015}.}
	\label{fig:li_abundances_sbbn}
\end{figure}

Se han realizado investigaciones para intentar vislumbrar si existe una relación entre las concentraciones de Li y la metalicidad de la nube de la que surgen las estrellas \cite{Pinsonneault1997, BarradoyNavascues2001}). Parece que esto no es así y el agotamiento del Li observado no sugiere que dependa de manera concluyente del contenido en elementos metálicos presente en esa nube primordial \cite{Guiglion2016}. Se ha contrastado que tanto para los cúmulos jóvenes (alta metalicidad) como para los antiguos (baja metalicidad), cuyos niveles de metalicidad varían aproximadamente $\pm$0.2 dex con respecto a la metalicidad de nuestro Sol, parecen compartir una distribución de Li muy similar \cite{Sestito2005}. Adicionalmente, se ha medido que la abundancia de Li medida en el Sol es mucho más baja que la que presentada por otras estrellas similares situadas en su vecindad \cite{Reddy2003}. Esta variación es más acusada que la detectada para otros tipos de elementos presentes en él.\par

Por otro lado, otros autores \cite{Sestito2005} defienden que la amplia gama de abundancias de Li observadas en estrellas similares al Sol y cercanas a éste es más fácilmente explicable si se establece una dependencia entre la abundancia de Li, la edad y masa de la estrella.\par

\subsection{Li y convección}
Una gran cantidad de trabajo teórico y observacional se ha dedicado a la comprensión de Li y su evolución, las evidencias acumuladas no concuerdan con las predicciones de los modelos estándar \cite{Sestito2005}; entendiendo por estándar a aquellos modelos que incluyen sólo los mecanismos de convección como proceso de mezcla y no tienen en cuenta fenómenos que podrían influir también como el transporte, como la difusión, las ondas de gravedad (nos referimos a ondas en el fluido de la estrella inducidas por la gravedad), la pérdida de momento angular o la presencia de planetas.\par

En las estrellas de tipo solar, la quema de Li se produce a una temperatura aproximada de 2.5 $10^{6} K$ mediante las reacciones de captura protónica que acontecen en interiores estelares. Por lo tanto, si se mantiene de manera continuada en el tiempo el mecanismo que transporta el Li entre la zona de convección exterior químicamente mixta y las regiones más profundas que poseen temperaturas lo suficientemente altas, éste acabará siendo destruido y su abundancia fotosférica disminuirá, llegando en algunos casos prácticamente a desparecer de la superficie estelar \cite{DelgadoMena2014}.\par

A su vez, la abundancia de Li medida en la fotosfera nos sirve también para realizar la calibración de modelos de evolución estelar. Sin embargo, es preciso mencionar que para que este planteamiento sea coherente, hay que suponer que la abundancia inicial de Li es conocida y que éste es única y exclusivamente destruido mediante reacciones nucleares.\par

\subsection{Li y difusión}
Un factor determinante para conocer la fase evolutiva en la que se encuentra una estrella es su composición química interna. Los modelos estándar consideran, como hemos apuntado anteriormente, a los procesos desencadenados por reacciones nucleares y por convección como los únicos que pueden modificar este perfil químico. Sin embargo, la difusión química de elementos también puede alterar este perfil y puede considerarse como otro mecanismo que contribuye a la disminución de la abundancia de litio en la superficie que tiene lugar a lo largo de MS. Este sería otro factor a tener en cuenta que contribuiría a poder explicar por qué la abundancia solar fotosférica es mucho menor que la meteórica. Sin embargo, este proceso estelar a largo plazo no puede ser observado directamente en las estrellas.\par

En el trabajo de \cite{Richard2004}, mostraron que la menor abundancia de Li observada en las estrellas de tipo Población II es el resultado del agotamiento de Li producido por el efecto de difusión atómica que entra en competencia con los procesos de mezclado en las zonas radiativas de estas estrellas.\par

La difusión microscópica acorta la vida útil de la secuencia principal y conduce a un agotamiento de los elementos superficiales. El modelo solar requiere la inclusión de difusión microscópica, de lo contrario ni siquiera la edad del Sol se podría producir correctamente \cite{Thoul1993}.\par

\subsection{Li y exoplanetas}
El descubrimiento de un gran número de exoplanetas durante las dos últimas décadas \cite{Mayor1995, Bonfils2018} ha significado un importante empuje para la comunidad científica en el campo de la astrofísica. Aparte de las consecuencias directas derivadas de poder estudiar los nuevos planetas descubiertos más allá de nuestro sistema solar, lo que ya es de por sí extremadamente interesante, no lo es menos la observación de sus estrellas anfitrionas, ya que éstas pueden llegar a aportar información muy valiosa acerca de las características globales, composición y formación de los sistemas planetarios extrasolares que albergan.\par

La hipótesis de que la variabilidad de concentración de Li en las estrellas puede ser debida a una posible correlación entre la presencia de planetas es una línea de trabajo de diversos estudios \cite{Israelian2009, DelgadoMena2014, Figueira2014}. Estas investigaciones apuntan a que esta presencia podría ser el desencadenante de la discrepancia en la abundancia de Li medidas en estrellas de tipo solar. Para nuestro Sol, ésta es unas 140 veces inferior a la que le correspondería a su modelo estelar \cite{Israelian2009}, aunque la temperatura en la BCZ no es lo suficientemente alta como para quemarlo.\par

En un primer momento, esta línea de investigación no se encontraba suficientemente fundamentada en evidencias que apoyasen este planteamiento, y esto se debía a la escasa disponibilidad de datos sobre estrellas que tuviesen sistemas planetarios asociados. En los últimos años, gracias a satélites como CoRoT \cite{Baglin2006} y Kepler \cite{Borucki2010}, dedicados a la búsqueda de planetas extrasolares, se han podido obtener datos de un gran número de estrellas con sistemas planetarios asociados. Esta nueva situación ha derivado en la realización de nuevos trabajos, o en la revisión de otros anteriores, en los que se han podido realizar nuevos estudios comparativos más amplios y sin sesgo en lo que se refiere a poder comparar estrellas con y sin sistemas planetarios. Estos nuevos trabajos han arrojado evidencias, tanto a favor como en contra, sobre la existencia de una relación directa entre la menor abundancia de Li en las estrellas con planetas con respecto a las que no tienen. Entre los primeros, tenemos el trabajo de \cite{Israelian2009}, en el que sus autores afirman que tras el análisis comparativo de un total de 133 estrellas en el rango de temperaturas que va de los $5600..5900 K$, de las que 30 de ellas albergan planetas y 103 no, encontraron que la gran mayoría de aquéllas que poseían planetas presentaron una disminución acusada en la abundancia de Li respecto de las que no tenían planetas (\ref{fig:li_abundances_planets}). También indican que por encima y por debajo de ese rango de temperaturas no parece existir relación alguna con la presencia o no de planetas, extremo que justifican al alegar que para estrellas con una temperatura superficial por encima de los $5850 K$, las capas convectivas de las estrellas más masivas que el Sol no son lo suficientemente profundas y además se encuentran demasiado alejadas como para alcanzar las regiones interiores de las estrella donde se alcanza una temperatura suficiente para quemar el Li.\par

Incluso su trabajo va más allá al afirmar que la menor abundancia de Li en estrellas con una temperatura efectiva restringida al rango $\tsun \pm 80$ K ($\tsun = 5777$ K) es una constante en las estrellas con planetas que se encuentran en su MS y que este resultado es independiente de la temperatura efectiva o la masa de las mismas. Por otro lado, el Li apenas se ve afectado en las estrellas menos evolucionadas, más calientes ($\teff > 6000$ K) y masivas, ya que éstas presentan una capa convectiva delgada. Por tanto, en este tipo de estrellas no existe una relación directa entre la temperatura de las mismas y la concentración de Li.\par

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{img/tesis/li_abundances_planets.pdf}
	\caption{Abundancia de Li vs temperatura efectiva en estrellas de tipo solar con y sin planetas. Las estrellas con planetas vienen representadas por los círculos rojos y las sin planetas por los círculos vacíos. El Sol queda representado por el punto negro rodeado de un círculo rojo, figura de \cite{Israelian2009}.}
	\label{fig:li_abundances_planets}
\end{figure}

Recientemente, se ha postulado otra alternativa por la que la presencia de planetas alrededor de una estrella podría afectar la evolución de la abundancia de Li fotosférico \cite{Bouvier2008}. El autor mantiene que, en el contexto de las estrellas de tipo solar en las que se han detectado exoplanetas respecto de las que no los albergan, la menor concentración de Li en las primeras puede ser el resultado de su historia rotacional. Afirma que, mediante el desarrollo de modelos de evolución de tipo solar incluyendo rotación, tanto para rotadores lentos (unas 5 veces mayor la velocidad de rotación del Sol) como rápidos (unas 70 veces mayor), que hacen evolucionar desde la PMS hasta la edad actual de nuestro Sol, han conseguido encontrar pruebas de que las estrellas con una rotación menor desarrollan un alto grado de rotación diferencial entre el núcleo radiativo y la envolvente convectiva, mientras que los rotadores rápidos evolucionan con poco desacoplamiento entre núcleo y envoltura. A raíz de esta fuerte rotación diferencial en la base de la envoltura convectiva, en los rotadores lentos el proceso de destrucción del Li se vuelve más eficiente, concluyendo que las estrellas que albergan exoplanetas y presentan un agotamiento del Li debieron de ser rotadores lentos durante ZAMS y que esto se debió a una interacción duradera entre la estrella y el disco plotoplanetario durante la PMS. De este planteamiento también se desprende que la vida de los discos debe ser más longeva (>5 Myr) de lo esperado y, posiblemente, también sea una condición necesaria para que se produzca la formación de planetas masivos y su posterior migración hacia zonas más exteriores del sistema planetario.\par

Asimismo, no faltan tampoco los referentes bibliográficos que defienden que no existe relación directa alguna entre la menor abundancia de Li detectada en la atmósfera de las estrellas que albergan planetas y las que no los tienen. En el trabajo de \cite{Baumann2010} se procede a analizar 117 estrellas con propiedades semejantes a nuestro Sol. Sus autores sostienen que las evidencias encontradas por otros investigadores en la dirección de la existencia de una relación entre las abundancias de Li detectadas y las estrellas que albergan planetas, se deben única y exclusivamente a consideraciones sesgadas y errores sistemáticos producidos durante el análisis de los datos recopilados. Por el contrario, estos mismos autores sostienen que sí existen evidencias claras entre la abundancia de Li de una estrella con su edad y la metalicidad de la misma. Por otra parte, la ingestión de un planeta “fallido” podría ser la causa por la que se produce un aumento de la abundancia de este elemento en la superficie de las estrellas.\par

\subsection{Li y rotación}
El impacto de la rotación tanto en la PMS como en el agotamiento del Li para las estrellas de tipo solar ha sido ampliamente debatido en el pasado \cite{Pinsonneault1997,Jeffries2004,Somers2014} y revisado más recientemente sobre la base de la disponibilidad de medidas más precisas \cite{Bouvier2016}. Los resultados sugieren una conexión entre la velocidad de rotación de las estrellas y la A(Li) detectada en sus atmósferas estelares, de modo que las estrellas que giran más rápido destruyen menos Li que las que lo hacen de manera más lenta. De alguna manera los resultados publicados por \cite{Bouvier2016} son inesperados ya que apuntan en una dirección diametralmente opuesta a los de trabajos anteriores, en los que se predecía que las estrellas que rotan más rápidamente deberían de destruir una mayor cantidad de Li.\par

Para explicar esta tendencia, es necesario recurrir a mecanismos adicionales que vinculen el agotamiento del Li y la rotación. Esta aparente contradicción de los resultados de diferentes trabajos no tiene por qué se tal. De hecho se plantean diferentes ideas que pueden ayudar a entender las discrepancias, entre las que se encuentra el acoplamiento de las estrellas con sus disco proto-estelar, que influiría en los mecanismos de mezclado ocasionados por la rotación \cite{Bouvier2008, Eggenberger2012}, la influencia de campos magnéticos \cite{Eggenberger2009} que tienen la propiedad de transmitir el momento angular (AM) de forma mucho más eficaz que inducir la mezcla \cite{Denissenkov2007}. Como consecuencia de este incremento en la eficiencia del transporte del AM, se reduce la cantidad de rotación diferencial entre las zonas radiativa y convectiva de la estrella (se fomenta una rotación de cuerpo sólido), así como las inestabilidades rotacionales inducidas. La acreción de material \cite{Baraffe2010} también se postula como una posibilidad para explicar la menor A(Li) para estrellas en el mismo rango de masas y edad. Episodios periódicos de acreción de material por parte de la estrella conducen a desarrollar temperaturas centrales significativamente más altas. A consecuencia de ello, una mayor cantidad de Li puede ser destruido.\par

\subsection{A qué se debe esta merma de Li}
Como hemos enumerado en los puntos anteriores, existe una propuesta amplia de potenciales mecanismos que podrían, por una parte, gobernar el ciclo de vida del Li, y por otra, ayudar a entender a qué se deben las discrepancias entre las observaciones y los resultados arrojados por los modelos. Sirva como ejemplo la línea de trabajo de \cite{Ramirez2012} en el que sus autores miden las abundancias de Li de una gran muestra de estrellas para encontrar correlaciones entre éstas y otras propiedades. Tales correlaciones deberían poder ayudar a determinar exactamente cómo se destruye el Li en las estrellas y cómo la abundancia galáctica de este elemento ha cambiado durante la vida de la Vía Láctea.\par

De manera genérica, estos mecanismos pueden agruparse en dos ideas principales. Por un lado, los que abogan por la propuesta de que la abundancia de este elemento en la superficie de las estrellas podría ser menor de la que los modelos predicen debido a que el Li podría haberse transformado en elementos más pesados mediante reacciones nucleares. El mayor inconveniente de este planteamiento es que la temperatura requerida para fusionar el Li y crear otros elementos es normalmente más elevada que la que puede ser alcanzada por las capas externas convectivas presentes en estrellas similares a nuestro Sol. Por esa razón, se ha propuesto una segunda explicación alternativa que plantea que el Li podría haberse mezclado en las capas más profundas de la estrella y de este modo, haber desaparecido de las zonas exteriores \cite{Pinsonneault1997}.\par

\section{El litio y las reacciones nucleares} \label{sec:li_reac_nuc}
Hasta aquí hemos presentado los posibles mecanismos que influyen de manera más o menos directa en la destrucción del Li, pero finalmente éste se destruye porque se produce una "quema" del mismo cuando las condiciones en el interior estelar son las adecuadas para que se produzcan una serie de reacciones nucleares. Como consecuencia, para poder obtener las abundancias de Li aL final de la TAMS, primeramente necesitamos conocer qué reacciones nucleares son las responsables de su creación. De acuerdo a Cox y Giuli \cite{Cox1968}, en las condiciones de temperatura y densidad que se dan en el interior de una estrella de tipo solar (1.0 $\msun$), el tiempo de vida de los elementos más ligeros: $\isotope[2]{H}$, $\isotope[3]{H}$, $\isotope[6]{Li}$, $\isotope[9]{B}$, $\isotope[10]{B}$ y $\isotope[11]{B}$ y los resultantes de combinarse con el H a través de las reacciones de cadena protón-protón en sus variantes I, II y III, tienen un tiempo de vida que va desde los pocos segundos hasta los aproximadamente $10^4$ años. Por lo tanto, todos estos elementos deberían de consumirse en las etapas iniciales de vida de la estrella, tan pronto su interior ronde los $10^7$ K.\par

La cadena de reacción protón-protón I es la cadena de reacciones dominantes en estrellas de tamaño solar o menor. El primer paso conduce a la fusión de dos núcleos de hidrógeno $\isotope[1]{H}$ (protones) a deuterio $\isotope[2]{H}$, liberando un positrón y un neutrino al transformar un protón en un neutrón. El positrón resultante de dicha reacción se aniquila inmediatamente con un electrón y su masa se convierte en energía liberada a través de dos fotones gamma. Tras esta reacción, el deuterio producido en el primer paso se puede fusionar con otro hidrógeno para producir un isótopo ligero de helio $\isotope[3]{He}$. A partir de este punto la reacción se subdivide en tres ramas diferentes que desembocan todas en la generación de un núcleo $\isotope[4]{He}$. En la variante I (Figura \ref{fig:pp-I}), éste se produce por la fusión de dos núcleos de $\isotope[3]{He}$.\par

\begin{figure}
	\centering
	\includegraphics[width=0.4\textwidth]{img/tesis/pp-I.png}
	\caption {Cadena de reacciones protón-protón I (pp-I)}
	\label{fig:pp-I}
\end{figure}

En las otras dos variantes, II (\ref{fig:pp-II}) y III (\ref{fig:pp-III}), se requiere del $\isotope[4]{He}$ previamente producido y ambas cadenas surgen a consecuencia de los dos caminos que el $\isotope[7]{Be}$ puede tomar. Como podemos observar un primer paso en común para las tres cadenas es la fusión de dos protones en deuterio.

\begin{figure}
	\centering
	\includegraphics[width=0.4\textwidth]{img/tesis/pp-II.png}
	\caption {Cadena de reacciones protón-protón II (pp-II)}
	\label{fig:pp-II}
\end{figure}

De particular interés para este trabajo es la variante II, ya que en ella es donde se produce $\isotope[7]{Li}$. En esta variante de la reacción se requieren temperaturas del orden de $1.4x10^7..2.3x10^7$ K. Bajo estas condiciones los isótopos estables de $\isotope[3]{He}$ se fusionan con núcleos de $\isotope[4]{He}$ para dar lugar a $\isotope[7]{Be}$, que tras capturar un electrón forma $\isotope[7]{Li}$. Finalmente, $\isotope[7]{Li}$ se fusiona con un protón para dar lugar a dos núcleos de $\isotope[4]{He}$.\par

\begin{figure}
	\centering
	\includegraphics[width=0.4\textwidth]{img/tesis/pp-III.png}
	\caption {Cadena de reacciones protón-protón III (pp-III)}
	\label{fig:pp-III}
\end{figure}

La variante III es la rama dominante cuando se alcanzan temperaturas superiores a los $2.3x10^7$ K. En esta rama, el $\isotope[7]{Li}$ se fusiona con un protón para formar un núcleo de $\isotope[7]{Li}$, que se descompone en un núcleo de $\isotope[8]{Be}$ y un positrón. Luego, el $\isotope[8]{Be}$ se descompone en dos núcleos de $\isotope[4]{He}$.\par

Los tres tipos de variantes ocurren de manera simultánea en el interior de las estrellas, pero unas son más frecuentes que otras. De acuerdo a las investigaciones teóricas, parcialmente soportadas por las medidas de neutrinos procedentes del Sol, se calcula que el 86\% de las reacciones se corresponden con las de tipo I, un 14\% con la de tipo II, y solo un 0.02\% con las de tipo III \cite{Scholz2018}.

De manera similar, la vida media de los isótopos involucrados en el ciclo CNO (Figura \ref{fig:ciclo-cno}) al recombinarse con el H dominante en esas primeras etapas debería de rondar los $10^6$-$10^7$ años. Sin embargo, este ciclo de fusión a través del cual las estrellas también generan He a partir del H es más representativo de estrellas con una masa superior a 1.3 $\msun$.

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{img/tesis/ciclo CNO.png}
	\caption {Cadena de reacciones del ciclo CNO}
	\label{fig:ciclo-cno}
\end{figure}

\endinput
%--------------------------------------------------------------------
% FIN DEL CAPÍTULO. 
%--------------------------------------------------------------------

### 2024-01-29 project/overleaf/Tesis III/capitulos/capitulo03.tex
Add Tesis project

% !TeX root = ../libro.tex
% !TeX encoding = utf8

\chapter{Rotación y campos magnéticos de intensidad constante}\label{ch:tercer-capitulo}

\section{Introducción}

El impacto de la rotación tanto en la PMS como en el agotamiento del Li para las estrellas de tipo solar ha sido ampliamente debatido en el pasado \cite{Pinsonneault1997,Jeffries2004,Somers2014} y revisado más recientemente sobre la base de la disponibilidad de medidas más precisas \cite{Bouvier2016}. Sin embargo, estos estudios previos se centraron principalmente en los efectos hidrostáticos y no consideraron la influencia de un acoplamiento entre el campo magnético estelar y su posible efecto de desacelaración. Se piensa que la AML tiene una influencia directa en los procesos de mezcla y que el transporte del momento podría producirse por una serie de mecanismos tales como: pérdida de masa, campos magnéticos y ondas gravitatorias (modos g). Respecto a estos dos últimos, las ondas gravitatorias \cite{Charbonnel2005, Pincon2016} y los campos magnéticos \cite{Eggenberger2009} tienen la propiedad de transmitir el momento angular (AM) de forma mucho más eficaz que inducir la mezcla \cite{Denissenkov2007}. Como consecuencia de este incremento en la eficiencia del transporte del AM, se reduce la cantidad de rotación diferencial entre las zonas radiativa y convectiva de la estrella (se fomenta una rotación de cuerpo sólido), así como las inestabilidades rotacionales inducidas. Además, también se originarían campos magnéticos en los límites entre estas zonas, en las llamadas tacoclinas \cite{Aschwanden2014, Guerrero2016}, que interaccionarían con las partículas del viento solar que se verían obligadas a rotar en sincronía con el campo magnético imperante y por tanto, frenarían la rotación estrella. Esto es lo que se conoce como el efecto de frenado magnético (Magnetic Braking, MB). \par

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{img/tesis/gravitational_waves.jpg}
    \caption{Vista esquemática de la estrella. Los penachos convectivos se producen en las capas superiores de la estrella y se adentran en la región convectiva. Crecen por arrastre turbulento de materia en sus bordes y alcanzan la parte superior de la zona radiativa. Allí, cada una de ellas, caracterizada por su posición angular ($\theta_i$,$\phi_i$), libera una parte de su energía cinética y genera ondas internas que pueden propagarse hacia el centro. (Figura tomada de \cite{Pincon2016})}
    \label{fig:grav_waves}
\end{figure}

A día de hoy, los modelos estándar no son capaces de replicar de forma satisfactoria los valores observados de abundancia de Li en las superficies estelares con las predicciones que estos modelos arrojan. Esta falta de concordancia nos hace pensar que ciertos mecanismos físicos que influyen en la destrucción del Li están siendo modelizados de forma inadecuada o simplemente no se tienen en cuenta, como por ejemplo el frenado magnético. A la vista de estos hallazgos, parece evidente que aún quedan cuestiones abiertas por resolver sobre el tema de los procesos que participan, directa o indirectamente, en los mecanismos de mezcla que no están incorporados en los modelos estelares estándar. Por ello, es imprescindible una adecuada consideración de las interacciones entre rotación y campos magnéticos a la hora de estudiar la distribución del AM. \par

La abundancia de Li observada en la fotosfera de las estrellas es un indicador de su composición interior y de los procesos de mezclado que en su interior tienen lugar. Adicionalmente, estas abundancias (además de otras métricas) se utilizan para comprobar la validez de los modelos estelares. Para que esto sea posible hay que tomar como premisa inicial que la abundancia de Li generada en la nucleosíntesis asociada al Big Bang es conocida y este elemento solo se destruye a través de reacciones nucleares. A pesar de décadas de esfuerzos teóricos, se sigue sin encontrar una explicación coherente con los modelos para las discrepancias encontradas en las comparaciones sobre la abundancia de Li para estrellas pertenecientes a cúmulos de diferentes edades y que se encuentran en el mismo estado evolutivo, o bien en la PMS, o bien en MS. Los modelos no son capaces de explicar las abundancias detectadas en las etapas tardías de la MS \cite{Tschape2001}.\par

Es sabido que parte de la pérdida de Li se produce durante la PMS y que además ésta se acentúa según decrece la masa de la estrella y, a igualdad de masa, según aumenta la metalicidad de la misma. Para reproducir la dependencia entre la edad y masa de las estrellas con la merma de las concentraciones de Li, estos modelos necesitan de una combinación de procesos de mezclado cada vez más complejos, como overshooting, mezclado debido a procesos rotacionales o difusión microscópica.
Para las estrellas de tipo solar, nos encontramos con dos factores que tienen que ser explicados desde el punto de vista de la abundancia de Li:

\begin{enumerate}
    \item La deficiencia generalizada de Li en el Sol y en estrellas de tipo tardío presentes en cúmulos a diferentes edades.
    \item La gran dispersión en abundancia de Li que existe en estrellas con una determinada temperatura efectiva detectada en los cúmulos abiertos.
\end{enumerate}

Existen datos observacionales que sugieren que las estrellas que giran más rápido preservan el Li mejor que las que lo hacen más lentamente. Este hecho se puede deber a que por encima de cierto umbral de velocidad de rotación, ésta impide de manera drástica la penetración vertical de penachos generados por movimientos convectivos que, cuando están presentes, contribuyen a hacer más eficiente el proceso de destrucción de Li \cite{Baraffe2017}.\par

Adicionalmente y según el modelo estándar de nucleosíntesis para el Big Bang, la abundancia de Li original puede estimarse en 2.6 dex \cite{Spergel2003} pero la abundancia de litio detectadas en meteoritos se conoce que está en el orden A(Li) = [3.26-3.34] dex \cite{Randich2006, Grevesse2007}, lo que apunta hacia un enriquecimiento desde que se produjo el Big Bang. Por otro lado, la abundancia medida en las estrellas de tipo solar es del orden de ALi = 1.05 dex \cite{Grevesse2007}, lo que implica la existencia de procesos de destrucción del Li en su interior.  Esta discrepancia entre las abundancias medidas en meteoritos y en estrellas de tipo solar es lo que ha pasado a denominarse el “problema del litio”.\par

Continuando con la importancia del Li y teniendo en cuenta que tanto éste, como los demás elementos ligeros berilio y boro, se queman a temperaturas relativamente bajas en interiores estelares, tenemos que estos elementos sólo sobreviven en las capas más externas de una estrella, donde la temperatura es menor. Adicionalmente, su presencia es un trazador muy potente relacionado con los mecanismos de mezcla en funcionamiento en estructuras estelares, ya que nos puede aportar información acerca de qué tal eficientes son a la hora de transportar estos elementos a zonas más interiores donde serían destruidos.\par

\section{El Sol y la rotación}
Nuestro Sol, como estrella que es y como ocurre con el resto de los cuerpos celestes, gira en torno a su eje. El tiempo que tarda en completar una revolución completa sobre este es de aproximadamente 26 días. Decimos aproximadamente porque la duración del día solar es diferente dependiendo de la latitud que se tome como referencia. Si comparamos las velocidades de rotación solar y la terrestre con respecto al diámetro de los correspondiente cuerpos celestes, se obtiene que un punto situado sobre la superficie solar girar aproximadamente cuatro veces más rápido de lo que lo haría sobre la Tierra.

Esto ocurre porque, a diferencia de la Tierra, el Sol no es un cuerpo rígido. Por lo tanto, al estudiar la rotación del Sol, no se puede considerar como una estructura compacta. El Sol es en realidad una enorme esfera de plasma, más parecida a una enorme bola de gas que a una estructura sólida rígida.\par

Si tomamos como referencia medidas de la rotación solar a lo largo de un período largo de tiempo podemos observar que, dependiendo de la latitud que tomemos como referencia, la velocidad de rotación se va incrementando conforme la distancia al ecuador solar se reduce. Es decir, el Sol gira más rápidamente en el ecuador que en sus polos \ref{fig:rot_solar_vs_latitud}. Este comportamiento que se conoce como rotación diferencial fue descrito por Richard Carrington (1826-1875) tras analizar el movimiento de las manchas solares \cite{Carrington1863}. Carrington dedujo la siguiente ley empírica mediante donde $\Omega_*$ es la cantidad angular de la rotación diaria y $\phi$ es la latitud heliográfica, que corresponde a la latitud en la Tierra  \cite{Banisch2009}.\par

\begin{ceqn}
\begin{equation}
    \Omega_* = 14.4 - 2.8\sin^2\phi \label{eq:ang_rot_sun}
\end{equation}
\end{ceqn}

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{img/tesis/sun_1.png}
    \caption{Rotación diferencial presente en el Sol. Tomando como referencia un meridiano solar y situando sobre este diferentes puntos de referencias a diferentes latitudes, se observa que la velocidad con la que estos giran es mayor a medida la distancia al ecuador solar disminuye. Crédito McGraw-Hill.}
    \label{fig:rot_solar_vs_latitud}
\end{figure}

El período de rotación $d_*$ en días se calcula como:

\begin{equation}\label{eq:dia_solar}
    d_* = \frac{360}{\Omega_*}
\end{equation}

En la región ecuatorial la rotación dura solo 25 días, mientras que en los polos el Sol gira más lentamente \ref{fig:ang_solar_vs_latitud}. Aquí el día solar tiene una duración aproximada de 30 días. A pesar de esta rotación diferencial entre regiones situadas en latitudes diferentes, no se produce un aplanamiento reseñable del disco solar ya que la velocidad de rotación es pequeña, de media 2km/s \cite{Gill2012}.\par

\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{img/tesis/sun_3.png}
    \caption{Valor predicho teóricamente para la velocidad de rotación del Sol en función de la latitud. Sobre el ecuador solar se obtiene la máxima velocidad angular y esta va disminuyendo a medida que aumenta la latitud. Crédito CESAR ESAC.}
    \label{fig:ang_solar_vs_latitud}
\end{figure}

\section{Efectos de los campos magnéticos}
Aún se desconoce el origen de los campos magnéticos en las estrellas. Los campos magnéticos organizados detectados en algunas estrellas O \cite{Wade2010} podrían ser campos fósiles \cite[ver][para más detalles ]{Dudorov2014}, o campos producidos mediante un mecanismo de dinamo \cite{Cantiello2009}. Además, la presencia de campos magnéticos con intensidades del orden de kG \cite{Hussain2014} ha sido una condición sine qua non para tratar ciertos datos observados en estrellas de tipo T Tauri en sistemas PMS jóvenes en acreción \cite{Johns-Krull2007}.\par

La masa de las estrellas juega un papel determinante en la existencia o no de una envoltura convectiva externa. A su vez, la presencia de esta región convectiva condiciona la evolución de los campos magnéticos. Las estrellas masivas no tienen zonas convectivas externas aunque pueden exponer un núcleo convectivo. En cambio, las estrellas menos masivas, como el Sol, tienen envolturas convectivas y un núcleo radiativo. En ambos tipos de estrellas se pueden generar fuertes campos magnéticos en la tacoclina pero con notables diferencias \cite[v.g.][para más detalles]{Chabrier2006,Charbonneau2010}.\par

En el Sol, la materia que lo compone se encuentra en su mayoría presente en forma de plasma. Es este estado, la envoltura electrónica se encuentra desacoplada del núcleo atómico y bajo ciertas condiciones físicas pueden moverse libremente, por lo que pueden poner en movimiento una carga electrónica. Bajo estas condiciones, la materia solar puede alcanzar la conductividad del cobre \cite{Banisch2009}.\par

\subsubsection{El mecanismo de dinamo}
Una dinamo es el proceso por el cual la energía mecánica se convierte
convierte en energía electromagnética por inducción. La energía mecánica se debe al movimiento de los fluidos y la energía electromagnética resultante produce los campos magnéticos observados. Para que una estrella tenga un campo magnético generado por dinamo, debe contener una región fluida conductora de electricidad que experimente movimientos para generar inducción. Sin embargo, existen otras condiciones necesarias para el vigor y la morfología de los movimientos que se derivan del hecho de que el campo magnético no debe decaer debido a la disipación óhmica. La necesidad básica para mantener un mecanismo de dinamo es una región fluida eléctricamente conductora en el interior de la estrella.

El Sol se rige por un fuerte campo magnético que se genera con una intensidad de campo magnético de B z 10 5 G \cite{Aschwanden2014} en la tacoclina, la delgada capa de cizalla intercalada entre la zona radiativa y la convectiva. Los tubos de flujo magnético flotante se elevan a través de la zona de convección (debido a la inestabilidad convectiva que obedece a la ley de Schwarz).
que obedece al criterio de Schwarzschild) y emergen en la superficie solar en las regiones activas, donde forman manchas solares con intensidades de campo magnético de B z 10 3 G y bucles coronales con intensidades de campo de B z 10 2 G en los puntos de apoyo fotosféricos, y de B z 10 G en las mayores alturas coronales. Se cree que la rotación diferencial en la superficie solar enrolla el campo magnético superficial, que se fragmenta bajo la tensión magnética, circula meridionalmente hacia los polos y se reorienta desde el estado de tensión toroidal (con líneas de campo orientadas en dirección este-oeste) en el máximo solar a un campo dipolar poloidal (que conecta el polo norte con el polo sur) en el mínimo solar.
mínimo solar.

\subsubsection{Campos magnéticos internos}
Nuestro Sol es una estrella activa en lo que se refiere al campo magnético. Según los modelos teóricos este campo magnético es generado por una dinamo en la tacoclina y con una intensidad en torno a $B\approx10^{5}\,\Gauss$ \cite{Maeder2003a,Dudorov2014}. En el momento de su generación, las líneas de ese campo magnético tendrán una orientación poloidal y se verán afectadas por la rotación diferencial y obligadas a adoptar una tipología toroidal (el efecto $\omega$). Por otro lado, también puede darse el caso contrario, partiendo de un campo magnético toroidal obtener una configuración poloidal. Esto sería posible gracias a los movimientos convectivos de material que se producen en el CZ de la estrella y suponer que este material se comporta como un conductor perfecto. Esta segunda condición permite suponer que las líneas de campo magnético están "congeladas" en el fluido conductor y por tanto deben moverse solidariamente con él (efecto $\alpha$) \cite{Stanley2014}.\par

\begin{figure}
    \centering
    \includegraphics[width=0.6\textwidth]{img/tesis/alpha_omega_effect.png}
    \caption{Esquema de los mecanismos canónicos de generación de dinamo. Las líneas de campo magnético están en rojo y los campos de velocidad en azul. \textbf{(a)} En el efecto, un flujo helicoidal estira ($t_2$) y luego retuerce ($t_3$) el campo magnético. Una pequeña cantidad de difusión en el punto de torsión ($t_4$) puede generar un bucle de campo magnético ortogonal al campo original. \textbf{(b)} En el efecto $\omega$, una rotación diferencial a gran escala afecta una línea de campo poloidal y la estira en dirección zonal ($t_1$ - $t_3$) generando bucles magnéticos toroidales a partir de la línea de campo poloidal. El círculo negro indica el límite exterior de la región de la dinamo. \cite{Stanley2014}}
    \label{fig:alpha_omega_effect}
\end{figure}

\subsubsection{Campos magnéticos superficiales} \label{surf_mf}
La intensidad del campo magnético medida en la superficie del Sol es de aproximadamente $B\approx1\,\Gauss$ \cite{Weber1967,DAntona2000,Morin2012}, el doble del valor del de la Tierra. Además su topología es variada y compleja. El modelo de dinamo establece que el campo magnético producido en el interior de la estrella alcanza la superficie con la ayuda de procesos convectivos que tienen lugar en la CZ y da lugar a regiones magnéticamente activas en la superficie solar. Además, las estrellas en rotación que exponen una CZ exterior significativa pueden producir campos magnéticos superficiales a través de un mecanismo de dinamo \cite[v.g.][]{Brandenburg2004,Charbonneau2010,Brun2017}. Sea cual sea el origen de los campos magnéticos superficiales, se predice un acoplamiento con el viento estelar, provocando así una pérdida de masa y, si ésta es lo suficientemente fuerte, produzcan MB \cite[v.g.][]{UdDoula2002,Ud-Doula2007,Ud-Doula2008,Meynet2010}.\par

\section{Efectos del frenado magnético}
En su enfoque más aceptado, el MB está vinculado al trabajo de Skumanich \cite{Skumanich1972} en el que desarrolla una ley empírica de su evolución en paralelo a la edad de la estrella. Para calibrar esta ley, Skumanich se basó en datos obtenidos de estrellas de tipo G encontradas en la MS. La influencia del MB en la evolución de la estrella está estrechamente relacionada con el transporte del AM. Según \cite{Meynet2010}, se distinguen dos mecanismos principales atendiendo a cómo se produce el transporte del AM:

\begin{enumerate}
    \item Rotación diferencial: el transporte de AM es impulsado por corrientes meridionales e inestabilidades de cizalla.
    \item Rotación de cuerpo sólido: cuando el transporte de AM es muy eficiente, la rotación de cuerpo sólido se mantiene durante toda la MS.
\end{enumerate}

El AML causado por MB depende directamente de la cantidad de masa perdida por la estrella debido a los vientos estelares. El coeficiente de pérdida de masa estimado para una estrella de tipo solar, aproximadamente $10^{-14}\msun \; yr^{-1}$ \cite{Noerdlinger2008}, y el AML resultante derivado de ese valor pueden considerarse relativamente modesto como para influir decisivamente en la evolución de la estrella. Además, el efecto combinado de los movimientos convectivos y la rotación diferencial conducen a la generación de campos magnéticos. Esos campos magnéticos acaban alcanzando la superficie de la estrella, de modo que el AML puede incrementarse en varios órdenes de magnitud \cite{Langer2012}. Ese incremento es consecuencia del campo magnético que obliga a las partículas ionizadas del viento estelar a girar con la misma velocidad angular, vientos que se extienden una distancia varias veces superior al radio de la estrella \cite[ver][para más detalles]{UdDoula2002,Ud-Doula2007,Ud-Doula2008}.

Además, a lo largo de la evolución de la estrella, parámetros directamente observables como su velocidad de rotación, temperatura efectiva, gravedad superficial o abundancia de elementos químicos sufren variaciones. Estos parámetros influyen, directa o indirectamente, en la pérdida de momento angular causada por el campo magnético de la estrella.\par

\subsection{Formalismos de frenado magnético} \label{mod_mb}
Comenzamos enumerando los aspectos más relevantes y las suposiciones realizadas en la modelización de la evolución de la rotación, el frenado magnético y el momento angular en MESA. Estos serán considerados para el cálculo del AML como resultado del par aplicado por un viento estelar acoplado magnéticamente.\par

En presencia de pérdida de masa, el material del viento estelar elimina el AM, lo que provoca un efecto de giro a la baja. Esto es muy importante, por ejemplo, en las estrellas masivas, que se sabe que rotan rápidamente y tienen fuertes vientos estelares. También se ha observado que las estrellas de masa intermedia fuertemente magnéticas suelen tener tasas de rotación mucho más lentas que otras estrellas de su población parental \cite{Mathys2006}. En esas estrellas, la presencia de dicho campo magnético interactuará con la pérdida de masa y el radio Alfv'{e}n ($R_{A}$) juega un papel importante en esta interacción.\par

En el resto de este trabajo, describimos dos enfoques semi-empíricos para el cálculo del AML como resultado del par aplicado por un viento estelar acoplado magnéticamente. Esto se ha implementado como una extensión del código de evolución estelar Modules for Experiments in Stellar Astrophysics \cite[MESA; ][]{Paxton2011, Paxton2013,Paxton2015, Paxton2018, Paxton2019}.

\subsubsection{Frenado magnético de intesidad fija según Ud-Doula \& Owocki (2002)}
Un parámetro relevante para caracterizar la influencia de un campo magnético dado sobre el viento estelar es el denominado parámetro magnético de confinamiento del viento ($\eta_*$). Representa una relación energética (Ec.~\ref{eq:wind_conf}) y define un parámetro característico de la eficacia relativa de los campos magnéticos para circunscribir y/o canalizar el flujo de salida del viento \cite{UdDoula2002}.\par

\begin{ceqn}
\begin{equation}
    \eta_* = \frac{B^{2}/8 \pi}{3} \label{eq:wind_conf2}
\end{equation}
\end{ceqn}

\begin{ceqn}
\begin{equation}
    \eta_* = \frac{B^{2}/8\pi}{\rho\nu^2/2} \label{eq:wind_conf}
\end{equation}
\end{ceqn}

donde $B^{2}/8\pi$ es la densidad de energía del campo magnético, $\frac{1}{2}\rho\nu^{2}$ la densidad de energía cinética, $B$ la intensidad del campo magnético, $\rho$ la densidad de masa, y $\nu$ la velocidad del viento estelar.\par

Haciendo uso de $\eta_*$, el $R_{A}$ se define como el punto en el que se equilibran la densidad de energía del campo magnético, marcada $B$, y la densidad de energía cinética, definida por $\rho$ y $(\nu)$. Es decir, donde el coeficiente entre ambos es igual a 1. En caso de que $R_{A}$ sea mayor que el radio de la estrella, entonces el flujo del viento tendrá que seguir el campo magnético. Como consecuencia, el material saldrá de la superficie estelar con un AM específico mayor, ya que el radio de co-rotación ha aumentado y corresponderá aproximadamente al $R_{A}$. Este efecto se conoce como MB.\par

Las estrellas con masas iniciales similares pero diferentes relaciones de pérdida de masa ($\Dot{M}$) acabarán evolucionando de forma muy diferente. Las partículas ionizadas transportadas por el viento solar no sólo contribuyen a la pérdida de masa, sino también a la pérdida de energía cinética que se deposita en el medio interestelar. Dada una estrella con un viento esféricamente simétrico, $\Dot{M}$ se caracteriza por la siguiente expresión:

\begin{ceqn}
\begin{equation}
    \Dot{M} = 4\pi r^2\rho\nu \label{eq:mass_loss}
\end{equation}
\end{ceqn}

Usando (\ref{eq:wind_conf}) and (\ref{eq:mass_loss}), $\eta_*$ puede ser aproximado por: 
\begin{ceqn}
\begin{equation}
    \eta_* = \frac{B^{2}r^{2}}{\Dot{M}\nu} \label{eq:wind_conf2}
\end{equation}
\end{ceqn}

\begin{ceqn}
\begin{align}
    \Dot{M} &= 4\pi r^2\rho\nu \label{eq:mass_loss}\\
    K_e &= 0.5\Dot{M}\nu_{esc}
\end{align}
\end{ceqn}

\begin{ceqn}
\begin{equation}
    \eta_* = \frac{B^{2}r^{2}}{\Dot{M}\nu} \label{eq:wind_conf2}
\end{equation}
\end{ceqn}

\begin{equation}\label{eq:identidad-pitagorica2}
  \cos^2 x + \sin^2 x = 1.
\end{equation}

A su vez, de las características observables del viento solar, nos importan especialmente los parámetros relacionados con la cantidad de masa que es capaz de arrancar de las capas externas de la estrella en un intervalo de tiempo dado, y la velocidad que alcanza el propio viento medida a gran distancia de la estrella, la velocidad terminal ($\nu_\infty$). En general, un flujo de salida canalizado magnéticamente tendrá una geometría de corriente compleja pero, por conveniencia, la Ec.~\ref{eq:mass_loss} simplemente caracteriza la fuerza del viento en términos de una tasa de pérdida de masa distribuida simétricamente sobre una esfera. $\nu$ puede caracterizarse por la variación radial de la velocidad del flujo saliente en términos de la ley de velocidad (Ec.~\ref{eq:vel_law}), donde $r$ representa la distancia desde el centro de la estrella hasta el punto donde se va a medir la velocidad, y $R_*$ es el radio de la estrella:\par

\begin{ceqn}
\begin{equation}
    \nu(r) = \nu_\infty (1-R_*/r) \label{eq:vel_law}
\end{equation}
\end{ceqn}
donde $\nu_\infty$ es la velocidad terminal del viento definida como la velocidad que alcanza el viento, la materia que fluye a gran distancia de la estrella central, donde ya no es acelerada por la fuerza impulsora del viento pero su desaceleración debida a la interacción con el medio interestelar (ISM) es despreciable \cite{Niedzielski2002}.\par

Los vientos impulsados por líneas de las estrellas OB masivas tienen velocidades terminales que son directamente proporcionales a la velocidad de escape fotosférica \cite{Lamers2000} según:\par

\begin{ceqn}
\begin{align}
    \nu_\infty &\simeq 1.92 \;\nu_{\textrm{esc}} \label{eq:vinf}\\
    \nu_{\textrm{esc}} &= \sqrt{\frac{2\,G\,\mstar}{\rstar}} \label{eq:vesc}
\end{align}
\end{ceqn}

donde la Ec.~\ref{eq:vesc} describe la velocidad de escape newtoniana desde la superficie estelar, donde $G$ es la constante gravitatoria, ${\rstar}$ el radio, y ${\mstar}$ la masa de la estrella.

Haciendo uso de (\ref{eq:vesc}), $\nu_\infty$ puede reformularse como:
\begin{ceqn}
\begin{equation}
    \nu_\infty \simeq 1.92 \; x \; 618 \; \Bigg(\sqrt{\frac{\rsun}{\rstar}\frac{\mstar}{\msun}} \;\Bigg) \label{eq:vinf}
\end{equation}
\end{ceqn}
donde ${\rsun}$ es el radio, y ${\msun}$ la masa del Sol.

Dado que la versión de MESA utilizada en este trabajo no incluye el tratamiento de los campos magnéticos superficiales (sí tiene en cuenta el transporte por campos magnéticos internos de momento angular y elementos químicos) y, por tanto, tampoco los efectos de frenado magnético, es necesario extender el simulador para incluir este fenómeno. En la presente tesis hemos seguido el trabajo teórico sobre el AM en el Sol realizado por \cite{Weber1967}, haciendo uso de la Ec.~ \ref{eq:j_dot} para el cálculo del AML.\par

Se ha observado que las estrellas de masa intermedia fuertemente magnéticas suelen tener tasas de rotación mucho más lentas que otras estrellas de su población parental \cite{Mathys2006}. En esas estrellas, los campos magnéticos interactúan con la pérdida de masa, donde el radio Alfv'{e}n ($R_{A}$) juega un papel importante. Haciendo uso de $\eta_*$ (Ec. \ref{eq:wind_conf}), $R_{A}$ se define como el punto en el que la densidad de energía del campo magnético y la densidad de energía cinética se equilibran. En caso de que $R_{A}$ sea mayor que el radio estelar, entonces el flujo del viento tendrá que seguir el campo magnético. Como consecuencia, el material abandona la superficie estelar con un AM específico mayor, ya que el radio de co-rotación ha aumentado y corresponde aproximadamente a $R_{A}$. Siguiendo a \cite{Weber1967} el AML puede calcularse mediante:

\begin{ceqn}
\begin{equation}
 \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A} \label{eq:j_dot}
\end{equation}
\end{ceqn}

donde $\Dot{M}$ es la tasa de pérdida de masa, $\Omega$ la velocidad angular en la superficie de la estrella y $R_A$ el radio Alfv\'{e}n. \par

La expresión anterior se puede reescribir según \cite{Ud-Doula2008} de modo que dependa de $\eta_*$ en lugar de $R_A$ como sigue:

\begin{ceqn}
\begin{equation}
 \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{*}\eta_* \label{eq:j_dot_mesa}
\end{equation}

\end{ceqn}
que es una expresión más conveniente para ser implementada en MESA porque se basa en valores directamente expuestos durante las simulaciones.

\section{Efectos de la distrubución de AML}
\subsection{Formalismo para la distribución de AML} \label{mod_aml}
Como expresa la Eq.~\ref{eq:mb_torque}, la cantidad de AML depende de $\ralfven$, $\Omega$ y $\Dot{M}$. Para valores grandes de $\ralfven$, la estrella sufre una desaceleración significativa. El valor de $\ralfven$ depende de los siguientes parámetros variables: intensidad del campo magnético ($B$), $\teff$ y $\Omega$. Con respecto a $\Dot{M}$ y como se indica en la tabla \ref{tab:phy_mesa}, para calcular la pérdida de masa se utilizó la fórmula empírica desarrollada por Reimers \cite{Reimers1975} para estrellas en la rama gigante asintótica (AGB). Para una estrella de tipo solar el $\Dot{M}$ durante la EM es relativamente pequeño, alrededor de $10^{-14}\msun \, yr^{-1}$ \cite{Noerdlinger2008}. \par

MESA asigna un valor $\Omega$ para cada celda $k$ ($\Omega_k$) que se ajusta para que el momento angular resultante se conserve tras calcular la nueva masa de la celda $k$ ($m_k$) y su distancia al centro de la estrella ($r_k$). A continuación, se asigna un valor de AM a cada celda $k$ ($J_k$). En este punto, nuestro MB se activa, modificando $J_k$. Para ello, se proporciona una contribución adicional ($\Dot{J}_{k}$). Esta contribución es el resultado del par externo ejercido por el campo magnético una vez que se ha distribuido entre las diferentes capas que componen la CZ según dicta la Eq.~\ref{eq:k_jdot}:\par

\begin{ceqn}
\begin{align}
    \Dot{J}_{k} &= \Dot{J}_*\;\frac{m^{}_{k} r^2_{k}}{m^{}_* r_*^2} \label{eq:k_jdot}
\end{align}
\end{ceqn}

La distribución se realiza en dos iteraciones. En la primera iteración, a cada celda $k$ se le asigna la cantidad máxima de $\Dot{J}_{k}$ que puede acomodar en el paso de tiempo establecido por la simulación. Las cantidades de $\Dot{J}_{k}$ no acomodadas por las celdas se acumulan como momento angular residual ($\jres$). En una segunda pasada, el valor $\jres$ residual acumulado recogido en la iteración anterior se redistribuye entre aquellas celdas que pueden acomodar más par. De esta forma conseguimos conservar el momento angular.\par

\endinput
%--------------------------------------------------------------------
% FIN DEL CAPÍTULO. 
%--------------------------------------------------------------------

### 2024-01-29 project/overleaf/Tesis III/capitulos/capitulo04.tex
Add Tesis project

% !TeX root = ../libro.tex
% !TeX encoding = utf8
\chapter{Modules for Experiments in Stellar Astrophysics - MESA}\label{ch:cuarto-capitulo}
\section{Introducción}
Para la realización de este tesis doctoral nos hemos apoyado en la herramienta de evolución estelar \textit{Modules for Experiments in Stellar Astrophysics} (MESA). Este simulador incorpora módulos que se actualizan de manera periódica en base a los avances derivados de los trabajos más novedosos sobre ecuaciones de estado, opacidad, velocidades de reacción nuclear, datos de difusión de elementos y condiciones de límite atmosférico. Es una herramienta muy útil para poder contrastar mediciones obtenidas con los resultados arrojados por un determinado modelo, o para hacer predicciones a partir de modelos y contrastar los resultados que obtenemos de estos con las mediciones disponibles en otros trabajos de investigación.\par

En el caso particular de esta tesis doctoral estamos interesados en estudiar los mecanismos que intervienen en la destrucción de Li en estrellas similares a nuestro Sol, y particularmente en ella. Como hemos comentado a día de hoy no existe una opinión unánime en la comunidad científica, más allá de que los modelos de evolución estelar no producen resultados coherentes con las observaciones, de qué procesos gobiernan el proceso de agotamiento del Li estelar y de cuándo se ponen en marcha o se detienen. Existen diversos planteamientos de procesos que intentan dar una respuesta a este enigma. En este trabajo nos vamos a centrar en cómo influyen en modelos de estrellas de tipo solar, con 1.0 $\msun$ y con una metalicidad (Z) similar al Sol, los efectos de la rotación y de la pérdida de momento angular causada por la presencia de campos magnéticos que inducen un efecto de frenado magnético. Estos campos magnéticos tendrán, en una primera aproximación, una intensidad fija a lo largo de toda la evolución temporal del modelo. Posteriormente se extenderá el modelo para simular campos magnéticos de intensidad variable. Adicionalmente, los modelos simulados se basan en la teoría de la longitud de mezcla para modelar la convección estelar. Este formalismo depende del parámetro libre longitud de mezcla ($\amlt$), parámetro que mayoritariamente tiene un valor preestablecido y fijo durante toda la simulación. En nuestro trabajo, realizaremos simulaciones manteniendo estable el valor de $\amlt$ pero también introduciremos la posibilidad de hacerlo evolucionar temporalmente en base a otros parámetros estelares.\par

En estas circunstancias, nos encontramos en que el primer problema a solventar es el de cómo incorporar estas extensiones al simulador MESA, al mismo tiempo que se respeta su estructura modular y, sobre todo, sin provocar efectos indeseables en el resto de parámetros de la simulación. Para dar solución a este punto tenemos que basarnos en los planteamientos teóricos documentados en el Capítulo \ref{ch:tercer-capitulo} e implementarlos, en forma de rutinas en código Fortran, que puedan ser utilizadas en conjunto con el resto de módulos de MESA.\par

Por otra parte, se hace también necesario el comprender los diferentes parámetros y procesos que MESA utiliza a la hora de calcular las abundancias de los diferentes elementos que podemos encontrar en una estrella a lo largo de su evolución, tanto en su interior como en su atmósfera. Por tanto, aquí tenemos otra área de estudio que pasa por conocer las diferentes capacidades ya disponibles en el simulador que tengan impacto sobre el agotamiento del Li, entender cuándo pasan a ser relevantes en el proceso evolutivo de la estrella, cuáles son los parámetros que gobiernan su funcionamiento, las relaciones que guardan entre ellos y, por último, qué valores son los más recomendables para dar soporte al escenario de simulación que acabamos de plantear.\par

Una vez tengamos la parametrización adecuada para nuestro planteamiento teórico, pasaremos a simular de manera sistemática, haciendo uso de nuestras nuevas rutinas, un conjunto de modelos con diferentes valores para los parámetros relevantes de nuestros modelos: velocidad angular, intensidad del campo magnético y valor de $\amlt$. Los resultados de abundancia de Li que obtengamos en cada uno de los escenarios planteados los enfrentaremos con los que arroja MESA sin hacer uso de nuestra rutina. Esta comparación de resultados nos deberá de dar un primer indicio de si nuestro planteamiento realmente tiene algún efecto notable sobre la abundancia Li detectado en la superficie estelar. Finalmente, en base a los resultados obtenidos, analizaremos las consecuencias que nuestras rutinas han tenido sobre ellos y presentaremos una serie de conclusiones de por qué obtenemos estos datos y cómo se interpretan los mismos en base a los estudios teóricos y observaciones obtenidas por otras líneas de investigación.\par

\subsection{El proyecto MESA}
Como indica su contribuidor y desarrollador principal \author{QUEPONERAQUÍ??} \cite[xxxxxx]{Pasetto2014} (Paxton, et al., 2011, 2013, 2015) MESA es un conjunto de librerías de código abierto, robustas y eficientes escritas en el lenguaje de programación Fortran 95 que es susceptible de ser utilizado en una amplia gama de aplicaciones en astrofísica estelar computacional. Está categorizado dentro de los denominados códigos de evolución estelar unidimensional (1-D) y combina un gran número de módulos numéricos y físicos que le permiten simular una amplia gama de escenarios de evolución estelar, que van desde los que incluyen estrellas de muy baja masa hasta lo que caracterizan estrellas masivas, además de tener en cuenta fases avanzadas de evolución, como el flash del helio, pulsos térmicos o la rama asintótica gigante (AGB). Utiliza un modelo de malla o zonas adaptables, que representa las diferentes capas y celdas en las que se estructura el modelo estelar, y emplea sofisticados controles de evolución temporal.\par

MESA se caracteriza por resolver las ecuaciones de estructura y composición de forma simultánea. Incluye módulos con el último “estado del arte” capaces de resolver las ecuaciones de estado, opacidad, velocidades de reacción nuclear, difusión de elementos y condiciones de límite atmosférico necesarias para realizar la evolución estelar. Cada módulo está construido a partir de una biblioteca desarrollada en Fortran 95 y diseñada de forma modular, diferenciando entre una interfaz pública destinada a ser utilizada por otros módulos y una parte privada donde se implementa la lógica necesaria que permanece inaccesible al resto de módulos. Esta buena práctica permite desarrollar nuevos módulos o revisar los existentes de manera independiente.\par

Por un lado, MESA aborda la física estelar, su estructura y su evolución con métodos numéricos modernos y sofisticados. Por otro, utiliza una física actualizada que le confiere una amplia gama de aplicaciones en diferentes escenarios. Los métodos numéricos y computacionales empleados por MESA le permiten evolucionar de manera consistentemente a través de fases exigentes planteadas por los modelos estelares.\par

MESA se basa en el principio de código abierto: cualquiera puede descargar el código fuente en el que se basa, compilarlo y ejecutarlo para sus propias necesidades de investigación, además de poder ampliar su funcionalidad, como se mostrará a continuación en este trabajo. El propósito de ser distribuido como código abierto es el de involucrar y llegar al mayor número de miembros de la comunidad astrofísica para que hagan uso de la herramienta, y al mismo tiempo alentarla para que hagan contribuciones al proyecto, ya sea en forma de pruebas, detección y corrección de errores o añadiendo nuevas capacidades al simulador.\par

\section{¿Cómo trabajar con MESA?}
Cuando se descarga e instala MESA (en el presente trabajo no entramos al detalle de cómo hacer esto) veremos que bajo el directorio de instalación existe una serie amplia de subdirectorios adicionales. La mayoría de ellos contienen el código para los diferentes módulos que componen MESA y que proporciona alguna funcionalidad específica. De todos ellos, el módulo más importante es \textit{star}, ya que es éste el módulo que interacciona y controla la ejecución de los demás. Adicionalmente es también el módulo encargado gestionar el estado interno de la estrella y de calcular el paso temporal a utilizar en cada iteración durante el proceso de simulación.\par

Por otra parte, MESA se apoya en ficheros de proyecto en los que se fijan los diferentes parámetros iniciales, bajo qué condiciones tiene que evolucionar la estrella, opciones de paradas y representación gráfica que deben utilizarse durante la simulación. MESA ofrece en su distribución un buen número de proyectos pre configurados para diferentes escenarios de evolución estelar. 
A continuación, pasamos a comentar en más detalle el contenido del fichero de configuración de proyecto que en MESA se denomina \textit{inlist}.\par

\subsubsection{Fichero de configuración}
MESA utiliza el fichero \textit{inlist} como el punto de configuración y entrada para las simulaciones. Este fichero está compuesto de tres secciones principales. Cada una de ellas contiene un conjunto de opciones que controlan diferentes aspectos de MESA:
\begin{itemize}
    \item \textit{star\_job} - aquí encontramos las opciones que controlan cómo evoluciona la estrella
    \item \textit{controls} - opciones para el módulo star de MESA
    \item \textit{pgstar} - opciones para la salida gráfica por pantalla
\end{itemize}

La diferencia entre las secciones \textit{star\_job} y controls puede llegar a ser muy sutil. Aun así, intentaremos dar unas reglas generales para tratar de averiguar en qué sección esperaríamos encontrar determinadas opciones de configuración.

La sección \textit{star\_job} contiene opciones relacionadas con la respuesta a preguntas como las siguientes:
\begin{itemize}
    \item ¿Cómo debería MESA obtener un modelo inicial a partir del que realizar la simulación?
    \item ¿Existe algún tipo de ajuste que MESA debería realizar sobre el modelo inicial?
    \item ¿Qué datos aplicables a la microfísica del simulador debe leer MESA?
    \item ¿Dónde debe almacenarse la salida de MESA?
\end{itemize}

Por otra parte, la sección controls es la adecuada para opciones de configuración encaminadas a responder a las siguientes cuestiones:
\begin{itemize}
    \item ¿Cuándo se da por finalizada la simulación?
    \item ¿Qué procesos de transporte debe tener en cuenta MESA?
    \item ¿Qué tolerancias numéricas debe aplicar los métodos de resolución numérica que incorpora MESA?
\end{itemize}

Para conocer en más detalle las opciones, lo mejor es remitirse a la documentación estándar de MESA disponible en su sitio oficial\footnote{http://mesa.sourceforge.net}. Otra opción disponible, es la de consultar los ficheros de código donde se encuentran los valores por defecto asignados a cada una de las opciones del simulador. Para cada una de las secciones comentadas anteriormente, MESA dispone de un fichero específico (\$MESA\_DIR hace referencia al directorio donde se encuentra instalado el software):
\begin{itemize}
    \item \$MESA\_DIR/star/defaults/star\_job.defaults
    \item \$MESA\_DIR/star/defaults/controls.defaults
    \item \$MESA\_DIR/star/defaults/pgstar.defaults
\end{itemize}

Las opciones tienden a estar agrupadas según la funcionalidad del simulador que regulan. Es recomendable que cuando se esté buscando una determinada opción, antes de entrar al detalle de cada una de las entradas en los ficheros, se intente identificar mediante las preguntas listadas anteriormente en qué fichero podríamos encontrarla y, una vez accedamos al contenido del mismo, revisemos las diferentes cabeceras disponibles a modo de comentarios que en ellos se encuentra.

\subsubsection{Controlando la salida}
MESA genera por defecto una cantidad ingente de información. Existen dos tipos de ficheros de salida con información de la simulación: uno con información histórica sobre cantidades globales (masa, luminosidad) almacenadas a diferentes intervalos de tiempo y otro tipo con perfiles del interior estelar que almacena información con cantidades que varían de forma espacial (densidad, presión) en un determinado instante de tiempo. El contenido de estos ficheros no se encuentra directamente accesible en el fichero inlist, sino en los siguientes ficheros específicos para cada uno de ellos:

\begin{itemize}
    \item \$MESA\_DIR/star/defaults/history\_columns.list
    \item \$MESA\_DIR/star/defaults/profile\_columns.list
\end{itemize}

Los ficheros anteriores contienen las configuraciones por defecto que ofrece MESA. Si necesitamos modificar alguna de las opciones que aparece en ellos, lo recomendable es no hacerlo directamente sobre estos ficheros, sino hacer una copia de los mismos en nuestro directorio de trabajo y realizar las modificaciones oportunas en ellos.\par

Adicionalmente, si se diese la situación (como ha sido en este trabajo) de que se quieren mostrar valores de propiedades diferentes a los que aparecen por defecto en alguno de estos dos ficheros, cabe la posibilidad de añadirlos haciendo uso, como veremos más adelante, del fichero \textit{run\_star\_extras.f}.

\subsection{¿Cómo extender MESA?}
A veces se hace necesario extender la funcionalidad de MESA porque la que viene por defecto en el simulador no cumple con las necesidades del problema que queremos abordar. Para estas situaciones, MESA ofrece la opción de extender su funcionalidad de un modo relativamente sencillo y elegante y sin necesidad de tener que modificar el código fuente del simulador en sí. Con esto se garantiza que el simulador no queda alterado en su funcionamiento estándar a raíz de una alteración del código base y permite que otros usuarios de MESA no tengan que alterar sus instalaciones para poder ejecutar simulaciones que hacen uso de funcionalidades no estándar. \par

MESA proporciona el archivo \textit{run\_star\_extras.f} que ofrece una variedad de puntos de extensión (hooks) destinados a implementar acciones que no pueden hacerse únicamente mediante el uso del fichero de proyecto \textit{inlists}. Entre estas acciones adicionales tenemos la posibilidad de cambiar parámetros en cada paso de la simulación o la de reemplazar muchas de las rutinas de física predeterminadas que trae el simulador.\par

\subsubsection{Activación de capacidades extras}
El primer paso para hacer uso de las capacidades extendidas de MESA es el de proceder a activarlas. Para ello debemos editar el fichero \textit{run\_star\_extras.f}. Veremos que el contenido del mismo es bastante simple, se limita a incluir otro fichero que es el que contiene el conjunto de rutinas predeterminadas que se entrega con MESA. Las rutinas definidas en ese fichero que se incorpora son las que queremos personalizar, pero sin afectar a las versiones originales. Para ello, procederemos a reemplazar esta declaración de inclusión por el contenido del archivo referenciado, así conseguimos tener nuestra propia copia local y personal.\par

\subsubsection{Extensión mediante hooks}
MESA proporciona una forma de sustituir la mayoría de las rutinas físicas que ofrece por defecto sin la necesidad de, como apuntábamos anteriormente, tener que modificar el código original del programa mediante el uso de los denominados \textit{hooks} o puntos de extensión. Para ello se necesitan seguir los siguientes pasos:
\begin{enumerate}
    \item Activar el hook deseado en el fichero \textit{run\_star\_extra.f} situado en el directorio \textit{src} del proyecto en el que estamos trabajando
    \item Indicar nuestra versión del módulo de física a sustituir/ampliar
\end{enumerate}

Seguidamente debemos localizar qué \textit{hook} nos interesa modificar de entre todos los que ofrece MESA. Lo primero que debemos hacer es conocer los que MESA pone a nuestra disposición. Un \textit{hook} no deja de ser más que un fichero escrito en Fortran que implementa un conjunto de rutinas numéricas relacionadas con un determinado proceso físico que se ejecuta durante la evolución de una estrella. Cada una de estas rutinas es susceptible de ser sustituida por una versión propia con nuestro planteamiento físico.\par

Los dos conceptos más importantes que se necesita saber a la hora de poder usar \textit{run\_star\_extras.f} de una manera efectiva son:
\begin{itemize}
    \item El flujo de control de un paso de ejecución de MESA
    \item El contenido de la estructura \textit{star\_info}
\end{itemize}

Las diferentes rutinas presentes en \textit{run\_star\_extras.f} se invocan en diferentes puntos durante la ejecución de MESA. En el Listado \ref{lst:bucle_ctrl}, y a modo de pseudo\-código en Fortran, podemos encontrar un aspecto del bucle principal de control sobre el que itera MESA.\par
\begin{lstlisting}[language=Fortran, caption={Bucle de control principal sobre el que itera MESA en cada paso temporal}, label={lst:bucle_ctrl}]
subroutine run1_star(...)
  ! star is initialized here

  ! before evolve loop calls:
  !   extras_controls
  !   extras_startup
  call before_evolve_loop(...)

  ! evolve one step per loop
  evolve_loop: do while(continue_evolve_loop)
    call before_step_loop(...)

    ! may need to repeat this loop
    step_loop: do 
      if (stop_is_requested(s)) then
        continue_evolve_loop = .false.
        result = terminate
        exit
      end if

      result = star_evolve_step(...)
      if (result == keep_going) 
        result = star_check_model(...)
      if (result == keep_going) 
        result = extras_check_model(...)
      if (result == keep_going)
        result = star_pick_next_timestep(...)
      if (result == keep_going) 
        exit step_loop

      ! redo, retry, or backup must be done 
      ! inside the step_loop
      if (result == redo) then
        result = star_prepare_to_redo(...)
      end if
      if (result == retry) then
        result = star_prepare_to_retry(...)
      end if
      if (result == backup) then
        result = star_do1_backup(...)
        just_did_backup = .true.
      else
        just_did_backup = .false.
      end if
      if (result == terminate) then
        continue_evolve_loop = .false.
        exit step_loop
      end if
    end do step_loop

    ! once we get here, the only options
    ! are keep_going or terminate.

    ! after_step_loop calls:
    !   extras_finish_step
    call after_step_loop(...)

    if (result /= keep_going) then
       exit evolve_loop
    end if

    ! write out data
    !
    ! do_saves calls:
    !   how_many_extra_history_columns
    !   data_for_extra_history_columns
    !   how_many_extra_profile_columns
    !   data_for_extra_profile_columns
    call do_saves(...)

   end do evolve_loop

   ! after_evolve_loop calls:
   !   extras_after_evolve
   call after_evolve_loop(...)

end subroutine run1_star
\end{lstlisting}

El corazón del simulador MESA es la sección \textit{step\_loop}. Es en ella donde se sitúa toda la maquinaria mediante la que MESA evalúa y resuelve las ecuaciones de la estructura estelar.\par

Por otro lado, en la estructura de datos \textit{star\_info} MESA almacena toda la información sobre la estrella que está evolucionando. Por convención en el código, el nombre de la variable se utiliza a lo largo del mismo, incluyendo también el que aparece en  \textit{run\_star\_extras.f} para hacer referencia a esta estructura\footnote{En Fortran, el operador de porcentaje (\%) se utiliza para acceder a los componentes de la estructura, así que s\% x hace referencia al campo x contenido en la estructura s.}.\par

La estructura \textit{star\_info} contiene el modelo estelar propiamente dicho, es decir, información sobre las zonas o capas que componen la estrella que se está simulando, su perfil termodinámico, perfil de composición, etc. También contiene todos los valores que hemos asignado a los parámetros que aparecen en el fichero de proyecto \textit{inlist}, así como los valores por defecto para el resto de parámetros físicos que no hemos indicado de forma explícita.\par

Además de todo lo que hemos comentado hasta ahora, MESA también pone a nuestra disposición un conjunto de listas de valores que nos facilitará el poder pasar valores fijados en el fichero \textit{inlist} del proyecto a nuestras rutinas de código. Estos vectores de elementos, con un tamaño máximo de 100 valores y que no forman parte del conjunto estándar de parámetros, son: \textit{x\_ctrl}, \textit{x\_integer\_ctrl}, y \textit{x\_logical\_ctrl} y nos permiten almacenar valores de tipo real, entero y lógico respectivamente (ver Listado \ref{lst:paso_params}).\par

\begin{lstlisting}[language=Fortran, caption={Paso de parámetros no estándar a través del fichero de proyecto inlist}, label={lst:paso_params}]
&controls
  x_ctrl(1) = 3.14
  x_ctrl(2) = 2.78
  x_integer_ctrl(1) = 42
  x_logical_ctrl(1) = .true.
/ ! end of controls inlist
\end{lstlisting}

\subsubsection{Extensión ficheros de salida}
Anteriormente hacíamos referencia a cómo controlar la información estándar que MESA ofrece a través de los dos tipos de fichero de salida: \textit{history} y \textit{profile} (ver apartado PONER REFERENCIA para más detalles). Adicionalmente a esta opción, tenemos la posibilidad de enviar a alguno de estos dos ficheros información que, o bien la calcula ya de por sí MESA, pero no está pensada para ser volcada en estos ficheros, o bien puede representar valores que vienen derivados de nuestras rutinas y que, por lo tanto, no forman parte de la parametrización estándar de MESA.\par

Para cualquiera de los dos casos comentados existen un par de rutinas dentro del fichero \textit{run\_star\_extras.f} que nos permiten, mediante sencillas modificaciones en los métodos \textit{how\_many\_extra\_history\_columns} y \textit{data\_for\_extra\_history\_columns}, añadir cuántos parámetros queramos\footnote{En realidad, existe un máximo de 100 parámetros adicionales.}. Para ello, simplemente hay que indicar la cantidad de nuevos parámetros en los que estamos interesados, el nombre de los mismos y de dónde obtener sus valores.\par

INTRODUCIR UNA CAPTURA DEL FICHERO DE CONFIGURACIÓN Y DEL CÓDIGO ASIGNANDO VALORES

\subsection{Módulos de interés de MESA}
Para empezar, vamos a enumerar los módulos de MESA que hemos encontrado de especial relevancia a estudiar para el objetivo de este trabajo. Principalmente hemos basado nuestra búsqueda en localizar aquéllos que pueden tener relación directa con los procesos de destrucción del Li o con las reacciones nucleares que se producen en el interior estelar. Para cada uno de ellos ofrecemos un pequeño resumen.

\subsubsection{Módulo químico - chem}
El módulo \textit{chem} MESA está compuesto de una colección de datos, funciones y subrutinas encaminadas a la gestión de los elementos químicos y sus isótopos. Contiene información básica sobre cada uno ellos, desde el hidrógeno hasta el uranio. Adicionalmente incluye rutinas para realizar conversiones entre pesos, números atómicos y el nombre de los isótopos. Contiene listados completos sobre las abundancias solares de los diferentes elementos según diferentes estudios (para más información ver \cite{Paxton2011}).\par

\subsubsection{Reacciones termonucleares – rates}
El módulo \textit{rates} contiene los cocientes de reacciones termonucleares de \cite{Caughlan1988}, y \cite{Angulo1999}, siendo esta última opción la utilizada por defecto \cite{Paxton2011}. El conjunto de cocientes de reacciones incluido abarca más de 300 elementos e incluye las reacciones débiles necesarias para la quema de hidrógeno (emisión de positrones y captura de electrones), así como las reacciones de conversión neutrón-protón.

\subsubsection{Reacciones nucleares – net} \label{reac_nuc}
El módulo \textit{net} implementa redes de reacción nuclear. Por defecto, incluye una red básica de 8 isótopos: \isotope[1]{H}, \isotope[3]{He}, \isotope[4]{He}, \isotope[12]{C}, \isotope[14]{N}, \isotope[16]{O}, \isotope[20]{Ne}, \isotope[24]{Mg}, y otras extendidas para cálculos más detallados incluyendo el ciclo CNO, captura de partículas $\alpha$ y reacciones de captura protónica. Además de utilizar las redes existentes, MESA nos permite definir nuestras propias redes de reacción construidas a partir de las ya existentes (opción más sencilla) o completamente desde cero \cite{Paxton2011}.\par

Como indicaremos más detalladamente cuando analicemos la configuración final del fichero de proyecto inlist, las reacciones que más nos interesan son las de las cadenas protón-protón en sus variantes I, II y III (ver sección \ref{sim_mesa} para más detalles).\par

\subsubsection{Teoría de longitud de mezcla – mlt}
El módulo \textit{mlt} implementa la teoría de longitud de mezcla estándar, o por sus siglas en inglés mixing lenght theory (MLT) \cite{Paxton2011}. En la dinámica de fluidos, el modelo o teoría de longitud de mezcla es una parametrización que intenta describir la transferencia de calor y material causado por una inestabilidad convectiva en el seno de un fluido. Este modelo ha sido utilizado en numerosos campos, incluyendo la ciencia atmosférica, oceanografía y estructura estelar, que es el caso que nos atañe. La “longitud de mezcla” se define como la distancia recorrida por una celda o burbuja de material sometida a una inestabilidad térmica convectiva.\par

\subsubsection{Módulo de difusión – diffusion} \label{subsec_diffusion}
El módulo \textit{diffusion} lo utiliza MESA para calcular la difusión de partículas y la sedimentación gravitacional resolviendo las ecuaciones de Burger \cite{Burgers1969} a través del método propuesto por \cite{Thoul1993}. El módulo de difusión trata los elementos presentes en el modelo estelar como pertenecientes a "clases" definidas por el usuario en términos de rangos de masas atómicas. Para cada clase, el usuario especifica un isótopo representativo y todos los miembros de esa clase son tratados idénticamente, con sus velocidades de difusión determinadas por el isótopo representativo. La rutina utiliza la fracción de masa de la clase correspondiente para obtener su coeficiente de difusión \cite{Paxton2015}. Como detallaremos más adelante, esta opción por defecto del módulo no es la más adecuada para obtener las abundancias de Li.\par

El cálculo de la difusión puede restringirse a zonas en las que la temperatura es superior a un valor mínimo o en las que la fracción de masa del elemento indicado es superior a cierto umbral.\par

\subsubsection{Otros parámetros de interés}
Además de los módulos listados anteriormente existen otra serie de parámetros que ofrece MESA que están relacionados también con el planteamiento teórico presentado en este trabajo. En concreto nos estamos refiriendo a los parámetros de masa estelar, contenido de H y He presente en la nube protoestelar, así como la metalicidad de la misma (Z).\par

Finalmente, también tendremos en cuenta la rotación superficial de la estrella, ya que, como se ha venido diciendo, este fenómeno físico es considerado un factor que influye en el agotamiento del Li, tanto potenciándolo, como llegando a inhibirlo.\par

\section{¿Cómo visualizar los resultados de MESA?}
\subsection{Visualización con pgstart}
COMENTAR LA PARTE de pgstart
\subsection{Visualización con terceras herramientas}
En determinadas ocasiones, las opciones de visualización ofrecidas por MESA no es la más adecuada o conveniente para nuestros propósitos. Quizás porque no muestra el parámetro en el que estamos interesados, o si lo hace, deseamos otro tipo de formato de visualización. Aunque en la mayoría de los casos, la necesidad de recurrir a terceras herramientas para la visualización se debe a que necesitamos realizar algún tipo de preprocesamiento sobre los mismos. (ENUMERAR LAS DIFERENTES ETAPAS DE TRATAMIENTO DE LOS DATOS)

En nuestra investigación hemos recurrido a la herramientas \textit{Octave} (ver \ref{sec:tool_octave}) para las etapas de (ENUMERAR LAS ETAPAS) y a \textit{TopCat} (ver \ref{sec:tool_topcat}) para ....

\section{Extendiendo MESA - Parte I} \label{mesa_parte1}
Una vez obtenida una visión relativamente aceptable de las capacidades del simulador, el siguiente paso lógico consistió en ejercitar simulaciones teniendo en mente los siguientes propósitos:

\begin{itemize}
    \item Obtener la soltura suficiente con la herramienta, en lo que se refiere a su funcionamiento y parametrización, con el fin de obtener un punto de partida que nos sirva de referencia a la hora de comprobar los resultados obtenidos al incorporar nuestras rutinas
    \item Extender la funcionalidad del simulador MESA para que éste sea capaz de simular la presencia de campos magnéticos, tanto de intensidad fija como variable, el frenado magnético que estos producen, así como la evolución del parámetro $\amlt$ en la evolución de su estrella anfitriona
\end{itemize}

Con estos dos objetivos principales en mente, la concepción de los diferentes escenarios de simulación se realizó de forma condicionada al planteamiento teórico analizado en la primera parte de este trabajo. En particular, estos escenarios deberían de tener en cuenta los siguientes procesos/características que intervienen, de alguna u otra forma, en la evolución de las abundancias de Li:

\begin{itemize}
    \item Reacciones nucleares en las que interviene el Li
    \item Metalicidad y velocidad de rotación de la estrella
    \item Procesos de difusión
    \item Procesos de sedimentación
    \item Presencia de campo magnético de intensidad fija y variable
    \item Evolución temporal de $\amlt$
\end{itemize}

A la hora de extrapolar el subyacente teórico de este trabajo a los escenarios de simulación, se decidió hacer una división en base al soporte que ofrecía MESA de manera estándar para los conceptos teóricos que hemos acabado de enumerar y para aquéllos que no. De esta forma, nos encontramos que, a excepción del punto 5 y 6, el resto estaban soportados por MESA. A partir de esta división se establecieron dos líneas de trabajo complementarias. La primera encaminada a simular los procesos teóricos 1-4 que influyen sobre la evolución del Li, por un lado, y a utilizar los valores generados por esas simulaciones como marco de referencia por otro. La segunda línea de trabajo se encaminó a extender MESA para que soportara la influencia de campos magnéticos y $\amlt$ variable en la evolución de su estrella anfitriona.\par

\subsection{Simulaciones acordes al marco teórico estándar de MESA}
En esta línea de trabajo nos centramos en hacer uso de las funciones y parametrizaciones estándar que ofrece MESA para dar soporte a los puntos teóricos enumerados anteriormente y que recogemos de nuevo aquí:

\begin{itemize}
    \item Reacciones nucleares en las que interviene el Li
    \item Metalicidad y velocidad de rotación de la estrella
    \item Procesos de difusión
    \item Procesos de sedimentación
\end{itemize}

Como estamos interesados en observar cómo actúan esos y procesos parámetros físicos sobre la abundancia del Li en las fotosferas de estrellas de tipo solar (1.0 $\msun$) a lo largo de su evolución desde la PMS hasta la TAMS, nuestros modelos de simulación comienzan por fijar los siguientes parámetros:

\begin{itemize}
    \item Establecer una metalicidad similar al Sol
    \item Detener la simulación al alcanzar la Termination Age Main Sequence (TAMS)
    \item Obtener las abundancias para los elementos H, He, Be
\end{itemize}

La composición química del Sol, incluyendo su metalicidad y masa inicial del modelo quedó fijada por medio de los siguientes parámetros (Listado \ref{lst:ini_params}) de MESA:

\begin{lstlisting}[language=Fortran, caption={Parametrización de la masa inicial de la estrella}, label={lst:ini_params}]
&star_job
  ! Solar composition, 
  ! check the following file:
  ! $MESA_DIR/star/test_suite/
  ! example_solar_model/inlist_solar_model
  set_uniform_initial_composition = .true.
  initial_h1 = 7.0001067228923031D-01
  initial_h2 = 0
  initial_he3 = 2.7955281905284692D-05
  initial_he4 = 2.7952486377094160D-01

  ! set metal fractions z fractions
  initial_zfracs = 3 !GS98

&controls
  ! starting specifications
  ! or 1.0d0 in Msun units
  initial_mass = 0.8d0 
\end{lstlisting}

Posteriormente necesitamos indicar la condición de parada en la TAMS. Este momento en la evolución de una estrella queda asociado a la etapa en la que la estrella abandona la secuencia principal al haber agotado prácticamente todo el H que se encontraba en su núcleo. MESA nos ofrece la posibilidad establecer una condición de parada cuando las abundancias de ciertos elementos en el núcleo de la estrella alcanzan un cierto valor máximo o mínimo (Figura 8XXXXXXX?). En nuestro caso, utilizamos esta opción para indicar que cuando la abundancia del isótopo \isotope[1]{H} caiga por debajo de 0.000000001, la simulación se detenga (ver Listado \ref{lst:stop_params}.\par

Para poder obtener las abundancias de Li AL final de la TAMS, primeramente necesitamos conocer qué reacciones nucleares son las responsables de su creación y si éstas están o no incluidas por defecto en la configuración de MESA. Como se adelantó en la sección \ref{sec:li_reac_nuc}, estas reacciones son las de cadena protón-protón en sus variantes I, II y III.\par

\begin{lstlisting}[language=Fortran, caption={Parametrización de la condición de parada de la simulación cuando se produce el agotamiento de H en el núcleo de la estrella}, label=, label={lst:stop_params}]
&controls
  ! stop when the center mass fraction of 
  ! h1 drops below this limit
  xa_central_lower_limit_species(1) = 'h1'
  xa_central_lower_limit(1) = 1d-9
\end{lstlisting}

Por tanto, ya tenemos identificas la red de reacciones a activar en la configuración de MESA: reacciones nucleares de tipo protón-protón. Es en este punto donde volvemos a hacer referencia al módulo net (ver \ref{reac_nuc}). Si recordamos, se indicaba que MESA viene pre configurado con una red de reacciones nucleares en las que sólo trata con 8 isótopos. Si queremos incluir más, debemos modificar la red inicial por otra que los incluya.\par

Con los parámetros anteriores incluidos en el proyecto, indicamos a MESA que sustituimos la red de reacciones nucleares por defecto por una nueva denominada \textit{pp\_extras.net}. Las redes son simples ficheros de texto en los que se indica al simulador las reacciones nucleares que debe tener en cuenta y cuál es el aspecto de éstas. En concreto, en la elegida le indicamos que tenga en cuentas las reacciones protón-protón en sus variantes I, II, III.\par

Una vez activada la nueva red de reacciones nucleares en nuestro fichero de proyecto, conseguimos obtener las abundancias para el Li, las cuales son informadas en los ficheros de salida estándar que genera MESA. Adicionalmente estamos interesados en obtener el perfil de evolución de estos elementos a lo largo del tiempo de evolución de la estrella y en función del radio de la estrella (Figura 13XXXXXXXXXX). Para ello recurrimos a utilizar las capacidades gráficas de MESA y mostramos la abundancia de estos elementos. Dado que sus concentraciones son muy pequeñas en comparación del H, debemos de proceder a ajustar el eje de abscisas del gráfico.\par

Dado que nuestro modelo solo abarca las fases evolutivas PMS y MS, decidimos eliminar las reacciones en las que interviene el Fe ya que no vamos a evolucionar nuestra estrella hasta el punto en que comienza a generarse este elemento. Adicionalmente, configuramos MESA para que, a cada paso en la simulación y para cada celda, tenga en cuenta si se producen o dejan de existir isótopos a consecuencia de las reacciones nucleares que están teniendo lugar. Dados los valores de abundancias tan pequeños que se obtienen para el Li y Be, del orden de $10^-9$ para el primero y $10^-31$ para el segundo, optamos por activar esta opción para que el tratamiento de las reacciones nucleares sea lo más riguroso y exacto posible.\par

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{img/tesis/abundances.pdf}
    \caption {Gráfica de abundancias en la que aparecen los elementos H, He, Fe, Li y Be. El eje de abscisas ha sido reescalado}
    \label{fig:abundances}
\end{figure}

Continuando con el marco físico a simular, pasamos a centrarnos en la activación de los mecanismos que, acorde a la teoría, tienen influencia en el agotamiento del Li:
\begin{itemize}
    \item Proceso de difusión
    \item Proceso de rotación
\end{itemize}

Como ya se indicaba en el apartado \ref{subsec_diffusion}:
\begin{center}
    \begin{minipage}{0.9\linewidth}
        \vspace{5pt}%margen superior de minipage
        {\small
        (MESA) calcula la difusión de partículas y la sedimentación gravitacional resolviendo las ecuaciones de Burger a través del método propuesto por Thoul \cite{Thoul1993}. El módulo de difusión trata los elementos presentes en el modelo estelar como pertenecientes a "clases" definidas por el usuario en términos de rangos de masas atómicas.
        }
        \vspace{5pt}%margen inferior de la minipage
    \end{minipage}
\end{center}

De esta información se desprende que el hecho de que al activar la difusión también activamos la sedimentación gravitacional y, de manera indirecta, el módulo \textit{mlt} encargado de calcular los coeficientes de difusión que entran en juego en proceso de mezclado producido a consecuencia de los movimientos convectivos que se produce en el interior de la estrella.\par

Cabe también mencionar el tratamiento por “clases” que realiza MESA. Los isótopos presentes en nuestra red de reacciones nucleares que tienen un peso atómico similar son tratados de forma conjunta por la rutina de difusión en lo que se refiere a su velocidad de difusión. Ésta queda determinada por el isótopo representativo de la “clase”. La agrupación en clases tiene la ventaja de que las simulaciones se ejecutan de una manera más rápida ya que se ahorran cálculos. Por otra parte, tiene el inconveniente de que los resultados pueden llegar a ser menos exactos. En este trabajo hemos optado por desactivar la agrupación en clases a costa de aumentar los tiempos de simulación.\par

En lo que se refiere a la rotación de las capas superficiales de la estrella, MESA permite especificar una velocidad en km/s, que en nuestro caso ha quedado fijada en 20 km/s al ser un valor similar al que tenemos en el Sol.\par

En el Listado \ref{lst:net_params} mostramos la configuración adicional a añadir al proyecto de MESA para poder simular los conceptos teóricos discutidos hasta este momento.
\begin{lstlisting}[language=Fortran, caption={Parametrización de los procesos relacionados con el Li y de las gráficas de salida.}, label={lst:net_params}]
&star_job
  ! Change default net in order to include 
  !reactions for Li and Be
  change_initial_net = .true.
  new_net_name = 'pp_extras.net'
  ! If enable_adaptive_network is true, 
  ! then at each step, 
  ! the system calculates a new set of isos. 
  ! This can cause significant slowdowns
  enable_adaptive_network = .true.

  ! Activate surface rotation and set velocity
  change_v_flag = .true.
  new_v_flag = .true.
  set_initial_surface_rotation_v = .true.
  new_surface_rotation_v = 20 ! km/sec

  ! Rotation off until near ZAMS
  change_rotation_flag = .false.
  new_rotation_flag = .true.

&controls
  ! toggle element diffusion
  do_element_diffusion = .true.
  ! If true, don't lump elements into 
  ! classes for diffusion. 
  ! Instead, each isotope in the network 
  ! is treated as its 
  ! own separate class. This can cause 
  ! significant slowdowns 
  ! for large nets, so it is off by default.
  diffusion_use_full_net = .true.
  diffusion_use_cgs_solver = .false.

&pgstar
  ! show HR diagram
  ! this plots the history of L,Teff 
  ! over many timesteps
  HR_win_flag = .true.

  ! configure file output 
  HR_file_flag = .true.
  HR_file_interval = 1

  ! show the abundance profile
  Abundance_win_flag = .true.
  num_abundance_line_labels = 1

  ! set how many elements must be shown
  Abundance_num_isos_to_show = 5
  Abundance_which_isos_to_show(1) = 'li7'
  Abundance_which_isos_to_show(2) = 'be7'
  Abundance_which_isos_to_show(3) = 'h1'
  Abundance_which_isos_to_show(4) = 'he4'
  Abundance_which_isos_to_show(5) = 'fe56'

  ! set the max and min value for the y-axis
  Abundance_log_mass_frac_min = -30
\end{lstlisting}

Llegado este punto, hemos conseguido obtener una configuración de proyecto base que recoge las características esenciales de la estrella a simular en MESA, pone en marcha los procesos físicos relacionados con la evolución del Li y es capaz de informar, tanto a través de los ficheros de salida estándar como de las representaciones gráficas, sobre las abundancias de Li a lo largo de la evolución de la estrella en sus etapas de PMS y MS. Como decimos, esta configuración de proyecto y los valores arrojados por la simulación que lo utiliza, nos servirán para establecer el marco de referencia contra el que comparar los valores obtenidos en las nuevas simulaciones que extienden la funcionalidad de MESA.\par

--CONTINUAR POR AQUÍ --

\subsection{Simulaciones extendiendo el marco teórico estándar de MESA}
Una vez que hemos llegado a un punto en el que tenemos una configuración de proyecto que nos permite obtener las abundancias de Li de forma detallada, y además hemos conseguido activar los procesos de difusión y rotación superficial en nuestro modelo, damos paso a una segunda línea de trabajo consistente en extender el funcionamiento de MESA para incorporar nuestras propias rutinas de extensión al simulador. Las consideraciones previas a tener en cuenta son las siguientes:

--INCLUIR AQUÍ ENUMERACIÓN DE PASOS A GRANDES RASGOS--
\begin{itemize}
    \item Entender el ciclo de control aplicado a los pasos de simulación
    \item Extender la rutina de inicialización para incorporar nuestros parámetros
    \item Extender 
\end{itemize}

El planteamiento teórico no es especialmente difícil, lo complicado del mismo es la parte que afecta a MESA. En concreto, localizar las secciones concretas de código en las que se calcula el valor de la gravedad local que afecta a cada elemento de masa que tiene en cuenta la simulación. A esto hay que sumarle el hecho de que MESA está construido de una forma modular, así que es de esperar que no tengamos una única sección de código en el programa donde se calcule este valor de gravedad, sino que cada módulo que, de algún u otro modo, contribuya a la difusión química de los elementos tiene que ser revisado.\par

Con diferencia, el localizar dónde invocar a nuestra rutina de código es la parte más complicada de todo el presente trabajo. Tenemos que detectar los módulos implicados en la difusión de los elementos, con el riesgo que implica que nos dejemos alguno por identificar, analizar su código fuente en detalle para llegar a entender su funcionamiento y realizar la invocación en el punto exacto a nuestra rutina, para alterar así el valor de gravedad calculado por MESA para un determinado diferencial de masa. Adicionalmente, se hace totalmente necesario el revisar la bibliografía en la que se basa la implementación de las rutinas presentes en cada módulo para dotar de contexto y significado a las líneas de código. De otro modo, es casi imposible extraer el trasfondo científico que se encuentra en ellas codificado.\par

--REVISAR PARA ELIMINAR REFERENCIAS A GRAVEDAD. HACERLO GENÉRICO--

Inicialmente teníamos dos módulos candidatos por los que empezar a estudiar su implementación, el de difusión y el de mlt. Por tener éste último un menor número de líneas de código (aproximadamente unas 1800) y encontrarse la implementación restringida a un solo módulo, empezamos por él. Tras una revisión a fondo del mismo se llegó a generar una versión modificada del mismo que calculaba un nuevo valor de gravedad local relacionado con el elemento de masa diferencial. Este nuevo valor tenía en cuenta tanto la masa del planeta, como su distancia a la estrella.\par

Seguidamente pasamos a analizar el código fuente del módulo de difusión que, como mencionábamos, se basa en la implementación de las ecuaciones propuestas en el trabajo de Thoul, et al. (1994). Ya desde los primeros momentos nos dimos cuenta de que el esfuerzo de localizar dónde deberíamos de invocar a nuestra rutina se iba a complicar. La implementación de este módulo es bastante más laboriosa, larga (más de 3500 líneas) y se encuentra repartida entre varios módulos que, a diferencia de lo que ocurría con el de mlt, algunos de ellos no ofrecían la posibilidad de ser extendidos de una manera cómoda. Además de esto, la implementación que se sigue en MESA acaba generando una matriz de coeficientes de difusión en los que el efecto de la gravedad está representado de manera implícita, es decir, no se calcula un valor de gravedad local asociada a un elemento diferencial de masa en un paso previo, de forma análoga a como ocurre en el módulo de mlt, para finalmente acabar derivando el valor del coeficiente de difusión. En este caso se resuelve el sistema de ecuaciones y la contribución de la gravedad sobre el elemento de masa queda recogida en las soluciones al mismo. A la propia dificultad de entender la implementación del código, se sumaba el hecho de tener que deshacer el cálculo de los coeficientes para poder incluir el efecto de nuestra rutina, y esto era algo que se antojaba harto complicado al no disponerse del conocimiento necesario. Estábamos en un punto muerto.

--REVISAR PARA ELIMINAR REFERENCIAS A GRAVEDAD--

\subsection{Rutina de inicialización para las simulaciones}
EXPLICAR EL CONJUNTO DE PARÁMETROS QUE SE LEEN DEL INLIST, Y SIGNIFICADO. TAMBIÉN LOS FLAGS DE DEPURACIÓN
\begin{lstlisting}[language=Fortran, caption={Rutina de inicialización de las simulacions.}, label={lst:extras_ctrl}]
      subroutine extras_controls(id, ierr)
         integer, intent(in) :: id
         integer, intent(out) :: ierr
         type (star_info), pointer :: s
         ierr = 0
         call star_ptr(id, s, ierr)
         if (ierr /= 0) return

         original_diffusion_dt_limit = s% diffusion_dt_limit
         s% other_wind => Reimers_then_Blocker
         ! inject our torque routine
         s% other_torque => other_torque_hook

         !debug flags
         debug_use_other_torque = s% x_logical_ctrl(1)
         debug_reset_other_torque = s% x_logical_ctrl(2)
         debug_get_cz_info = s% x_logical_ctrl(3)
         debug_get_core_info = s% x_logical_ctrl(4)
         debug_new_alpha = s% x_logical_ctrl(7)
         debug_mag_field = s% x_logical_ctrl(8)
         debug_j_dot = s% x_logical_ctrl(9)

         !If true, once the radiative core is developed, report always true
         !in is_radiative_core function
         keep_on_rad_core = s% x_logical_ctrl(6)

         !eps thershold
         eps_threshold = s% x_ctrl(2)

         !disk locking
         disk_lt = s% x_ctrl(3)
         disk_omega = s% x_ctrl(4)

         !Variable MLT alpha
         var_mlt_alpha = s% x_logical_ctrl(10)

         ! Once you have set the function pointers you want,
         ! then uncomment this (or set it in your star_job inlist)
         ! to disable the printed warning message,
         ! Uncomment these lines if you wish to use the functions in this file,
         ! otherwise we use a null_ version which does nothing.
         s% extras_startup => extras_startup
         s% extras_start_step => extras_start_step
         s% extras_check_model => extras_check_model         
         s% extras_finish_step => extras_finish_step
         s% extras_after_evolve => extras_after_evolve
         s% how_many_extra_history_columns => how_many_extra_history_columns
         s% data_for_extra_history_columns => data_for_extra_history_columns
         s% how_many_extra_profile_columns => how_many_extra_profile_columns
         s% data_for_extra_profile_columns => data_for_extra_profile_columns

         s% job% warn_run_star_extras =.false.             
      end subroutine extras_controls
\end{lstlisting}

\subsection{Rutina de par de torsión}
EXPLICAR QUE SUSTITUIMOS LA RUTINA POR DEFECTO CON LA NUESTRA
\begin{lstlisting}[language=Fortran, caption={Rutina de par de torsión.}, label={lst:torque_hook}]
      subroutine other_torque_hook(id, ierr)
         use const_def
         integer, intent(in) :: id
         integer, intent(out) :: ierr
         type (star_info), pointer :: s

         ierr = 0
         call star_ptr(id, s, ierr)
         if (ierr /= 0) return

         !If disk locking is actived & the star is younger that disk 
         !locking period &
         !star rotates faster than disk locking rotational velocitiy
         if((disk_lt>0).and.(s%star_age<disk_lt).and.(s%omega(1)>disk_omega)) then
            call other_torque_disk_lock(id, ierr)
         else
            call other_torque_mag_brk(id, ierr)
         endif

      end subroutine other_torque_hook
\end{lstlisting}

\subsection{Rutina de frenado magnético}
EXPLICAR QUE LOS DOS TIPOS DE RUTINAS: INTESIDAD FIJA, INTENSIDAD VARIABLE
\begin{lstlisting}[language=Fortran, caption={Rutina de frenado magnético.}, label={lst:torque_mb_hook}]
      ! This routine implements a magnetic braking effect.
      ! It distributes among the zones which conforms the convective shell the
      ! loss of angular momentum
      subroutine other_torque_mag_brk(id, ierr)
         use const_def
         integer, intent(in) :: id
         integer, intent(out) :: ierr
         type (star_info), pointer :: s
         integer :: k
         real(dp) :: B, j_dot, r_st
         type (star_zone_info), target :: sz_info, core_info
         type (star_zone_info), pointer :: sz_info_ptr, core_info_ptr
         real(dp), dimension(:), pointer :: mag_brk_jdot
         integer :: jdot_routine !how to distribute the j_dot
         !controls if jdot distribution must only affect the convective zone
         logical :: only_cz
         !controls if jdot distribution must wait till a radiative core is develop
         logical :: wait_rad_core          
         integer :: activated !signals when the jdot routine is activated

         !Pointer to structure which conveys information about the convectice
         !and core zones
         sz_info_ptr => sz_info
         core_info_ptr => core_info

         ierr = 0
         call star_ptr(id, s, ierr)
         if (ierr /= 0) return

         !Reset support structures
         allocate(mag_brk_jdot(s% nz))
         s% extra_jdot(:) = 0
         s% extra_omegadot(:) = 0
         activated = 0
         call reset_x_ctrl(s, idx_low_x_ctrl, idx_high_x_ctrl)
         call reset_core_info(core_info_ptr)
         call reset_convective_info(sz_info_ptr)

         !Wait till radiative core is develop?
         wait_rad_core = s% x_logical_ctrl(5)

         !Loss of angular momentum distribution method
         jdot_routine = s% x_integer_ctrl(1)
         only_cz = .true.
         if (jdot_routine /= 0) then
               only_cz = .false.
         end if

         !Get information about the convective zone
         if (only_cz) then
            !Get information about the outter convective zone
            call get_convective_info(s, sz_info_ptr)
         else
            !Get information about the outter convective zone till star surface
            call get_convective_to_surf_info(s, sz_info_ptr)
         end if

         !Get information about the core
         call get_core_info(s, core_info_ptr)

         !Calculate amount of loss of angular moment
         !Magentic field intensity
         B = s% x_ctrl(1)
         if (B < 0) then
            B = calculate_mag_field_intensity_gb(s)
            j_dot = calculate_jdot_rate_gb(s, B)
         else
            j_dot = calculate_jdot_rate_cantiello(s, B)
         end if

         ! The MB routine is activated under the following conditions:
         ! - use_other_torque flag is activated in inlist
         ! AND
         ! - the star is losing mass
         ! AND
         ! - the magnetic field intensitive is bigger than 0.0 (allow to 
         ! execute the routine if use_other_torque=.true.)
         ! AND
         ! (
         !   - a radiative core developed isn't required (configured in inlist)
         !   OR
         !   - a radiative core is required AND this was developed
         !)
         if ((s% use_other_torque) .and. (s% mstar_dot<0.0) .and. (B>0.0) .and. &
            (.not. wait_rad_core .or. (wait_rad_core .and. is_core_rad(s)))) then

            activated = 1

            !Distribute the loss of angular momentum
            call distribute_j_dot(s, j_dot, sz_info_ptr, mag_brk_jdot)

            !It happens that s% extra_jdot is longer than s% nz but mag_brk_jdot
            !is just defined
            !for s% nz elements
            s% extra_jdot(1:s% nz) = mag_brk_jdot
         end if
      end subroutine other_torque_mag_brk
\end{lstlisting}

\subsubsection{Rutina de frenado magnético de intensidad constante}
EXPLICAR QUE SUSTITUIMOS LA RUTINA POR DEFECTO CON LA NUESTRA
\begin{lstlisting}[language=Fortran, caption={Rutina de par de torsión.}, label={lst:jdot_cantiello}]
      real function calculate_jdot_rate_cantiello(s, bf_star) result(new_j_dot)
         use const_def
         type (star_info), pointer, intent(in) :: s
         real(dp), intent(in) :: bf_star

         real(dp) :: r_st, m_st, i_st, alfven_r
         real(dp) :: omega_surf, m_dot, eta_surf, v_inf, v_esc

         !Star data
         r_st = s% r(1)
         m_st = s% m(1)
         omega_surf = s% omega(1)
         !omega_surf = s% omega_avg_surf

         ! escape and infinite velocities
         ! 100000 transform from km/s to cm/s
         v_esc = (618 * ((Rsun/r_st)*(m_st/Msun))**0.5) * 100000
         v_inf = 1.92 * v_esc

         !m_dot = s% star_mdot !This gives the mass loss rate in Mstar/year
         m_dot = s% mstar_dot !This in g/s

         eta_surf = ((r_st * bf_star)**2)/(abs(m_dot) * v_inf)

         !Formula 2.3 Cantiello's MESA assigment
         new_j_dot = two_thirds * m_dot * omega_surf * (r_st**2) * eta_surf
    end function
\end{lstlisting}

\subsection{Rutina de distribución de pérdida de momento angular}
EXPLICAR QUE LOS DOS TIPOS DE RUTINAS: INTESIDAD FIJA, INTENSIDAD VARIABLE
\begin{lstlisting}[language=Fortran, caption={Rutina de distribución de pérdida de momento angular.}, label={lst:dis_jdot}]
      subroutine distribute_j_dot(s, total_j_dot, sz_info, mb_jdot_list)
         type (star_info), pointer, intent(in) :: s
         real(dp), intent(in) :: total_j_dot
         type (star_zone_info), pointer, intent(in) :: sz_info
         real(dp), dimension(:), pointer, intent(out) :: mb_jdot_list
         integer :: k
         real(dp) :: sum_jdot, dm_jdot, dm_bar_jdot, factor

         !By default, no lost of angular moment
         mb_jdot_list(:) = 0.0

         do k = sz_info% top_zone, sz_info% bot_zone, 1
            !Here the jdot distribution strategy is defined
            !Simple rule of three distribution loss of angular momentum based on the 
            !angular momentum of the zone vs total angular momentum of the convective zone

            !IMPORTANT: Don't forget to divide by dm(k) in order to get an "specific" jdot
            mb_jdot_list(k) = ((s% dm(k) * s% r(k)**2 * total_j_dot) / &
                  (sz_info% d_mass * sz_info% d_radius**2)) / s% dm(k)

         end do
      end subroutine distribute_j_dot

\end{lstlisting}

\endinput
%--------------------------------------------------------------------
% FIN DEL CAPÍTULO. 
%--------------------------------------------------------------------

### 2024-01-29 project/overleaf/Tesis III/capitulos/capitulo05.tex
Add Tesis project

\chapter{Resultados de frenado magnético de intensidad fija}\label{ch:quinto-capitulo}
\section{Configuración de los modelos} \label{marco_teorico}
A COMENTAR
Parametrización de los modelos
Rango de valores a asignar a los parámetros libres
Ciclo de control
Para cada paso de simulación
- Obtenemos o calculamos la intensidad del campo magnético
- Calculamos la pérdida de momento angular inducida por el campo magnético
- Distribuimos la pérdida de momento angular entre las capas de la estrella
- Obtenemos o calculamos el valor de $\amlt$

\section{Modelos de evolución estelar}
De acuerdo con la formulación de las secciones anteriores, los dos únicos parámetros libres de nuestra implementación son $B$ y $\Omega$. Las simulaciones numéricas trazaron la historia rotacional y $A(\isotope[7]{Li})$ de una estrella 1 $\msun$ para una variedad de valores iniciales para $B$ y $\Omega$ (ver también Tabla \ref{tab:phy_mesa}). MESA utiliza el enfoque en capas (shellular) \cite{Meynet1997} para dar cuenta de los efectos hidrostáticos de la rotación en modelos estelares 1D.\par

\begin{table}
	\centering
	\caption{Summary of adopted physics in MESA \cite[based on][]{Choi2016}.}
	\label{tab:phy_mesa}
	\begin{tabular}{ll} 
		\hline
		Parameter & Adopted prescriptions and values\\
		\hline
		Solar Abundance & $X_{\odot}=0.7154, Y_{\odot}=0.2703, Z_{\odot}=0.0142$\\
		Equation of State & OPAL+SCVH+MacDonald+HELM+PC\\
		Opacity & OPAL Type I for log T $\geq$ 4 \\ & Ferguson for logT $\leq$ 4\\
		Reaction Rates & JINA REACLIB\\
		Boundary Conditions & ATLAS12; $\tau$=100 tables + photosphere\\
		Difussion & Track \isotope[1]{H}, \isotope[2]{He}, \isotope[7]{Li}, \isotope[7]{Be}\\
		Rotation & Differential rotation at PMS \& MS\\
		Convection & MLT + Ledoux, $\alpha_{MLT}$ = 1.82\\
		Overshoot & time-dependent, diffusive, \\ & $f_{ov,core}=0.0160$,\\ 
		& $f_{ov,sh}=0.0174$\\
		Semiconvection & $\alpha_{sc}=0.1$\\
		Thermohaline & $\alpha_{th}=666$\\
		Rotational Mixing & Include SH, ES, GSF, SSI \& DSI\\
		Magnetic Effects & Magnetic braking based on idealized \\ & monopole field\\
		Magnetic Field & B(G) variable between [3.0 - 5.0]\\
		Mass Loss & activated, $\Dot{M}_{max} = 10^{-3} \: \msun \: yr^{-1}$\\
		Angular Moment Loss & activated, $\Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A}$\\
		\hline
	\end{tabular}
\end{table}

Adoptamos abundancias a escala solar y asumimos una abundancia inicial solar ($Z = Z_{\odot, \mathrm{pr} = 0,0142}$). También adoptamos los siguientes valores nominales para expresar las propiedades estelares en unidades SI, $\rsun = 6,957x10^{10}\, cm$ y $\msun = 1,988x10^{33}\, g$ que son coherentes con las resoluciones de la IAU \cite{Mamajek2015}. Para una descripción detallada de la física adoptada en este trabajo, remitimos al lector a \cite{Choi2016}. Ese trabajo se utilizó como punto de partida para el nuestro en lo que respecta a la parametrización del proyecto MESA, que se calibró para reproducir las abundancias de elementos medidas en la superficie solar.\par

Los modelos incluyeron la rotación durante la PMS ya que existen evidencias que abogan por una fuerte relación establecida entre la destrucción de Li y la rotación en esa fase (\cite{Bouvier2016,Bouvier2018}). Modelamos la rotación como AML se calcula como resultado del par aplicado a la zona de convección por un viento acoplado magnéticamente. Además, no tuvimos en cuenta ni la influencia de los campos magnéticos internos ni su existencia durante la fase T-Tauri.\par 
---CORREGIR ESTO SEGUN EL NUEVO PAPERAdoptamos un enfoque teórico sencillo y pragmático para establecer cuándo la estrella está alcanzando la fase ZAMS. El criterio elegido se basó en la existencia simultánea de una extensa capa convectiva y un núcleo radiativo. A partir de este momento, se activó la rutina MB, actuando como un mecanismo adicional a los existentes en el código evolutivo MESA que participó en el AML de la estrella. ---HASTA AQUÍ---\par

Supusimos que el campo magnético no variaba su intensidad a lo largo de la evolución de la estrella hasta alcanzar el TAMS. Además, también se consideraron otros efectos de rotación durante la evolución de los modelos: el transporte de AM desde el interior radiativo hasta la envoltura convectiva y la redistribución de AM asociada a cambios en la estructura interna durante el proceso de contracción hacia la EM.\par

Nos centramos en el papel indirecto que desempeña el MB en la destrucción del Li. Esto es consecuencia de su influencia en la historia rotacional de las estrellas de tipo solar. Nuestro principal objetivo era comprobar cómo el MB podría contribuir a explicar la evolución del Li en el Sol y en otras estrellas de tipo solar. Se ejercitaron un conjunto de escenarios diferentes con intensidades de campo magnético que oscilaban entre 3,0 y 5,0 G. Calculamos la evolución de modelos estelares 1 $\msun$ con metalicidad inicial solar y $\oomegac$ variable entre $0.0084$ y $0.0336$. Los modelos más avanzados para disposiciones más complejas del campo magnético, así como la evolución de su intensidad a lo largo de la vida de la estrella, se dejaron para trabajos posteriores.\par

Como expresa la Ec.~\ref{eq:j_dot_mesa}, la cantidad de AML depende de $R_A$ (o alternativamente de $\eta_*$), $\Omega$ y $\Dot{M}$. Para un valor grande de $\eta_*$, la estrella experimenta una desaceleración significativa. Controlamos el valor de $\eta_*$ en los modelos variando el campo magnético ($B$) y la velocidad de rotación estelar ($\Omega$). Con respecto a $\Dot{M}$ y como se indica en la tabla \ref{tab:phy_mesa}, para calcular la pérdida de masa se utilizó la fórmula empírica desarrollada por Reimers \cite{Reimers1975} para estrellas en la rama gigante asintótica (AGB). Para una estrella de tipo solar el $\Dot{M}$ durante la EM es relativamente pequeño, alrededor de $10^{-14}\msun \, yr^{-1}$ \cite{Noerdlinger2008}. \par

\subsection{Evolución de Li sin MB}

\begin{figure}
    \centering
	\includegraphics[width=0.7\textwidth]{img/paper1/li_var_vel_0_0g.pdf}
	\caption{La evolución de la abundancia superficial del \isotope[7]{Li} en relación con el \isotope[1]{H}, en función del tiempo para varios modelos de 1 $\msun$. La línea negra continua representa el modelo de referencia según \cite{Choi2016}. El resto de líneas son modelos que incluyen rotación PMS con $\oomegac$ entre 0.0084 y 0.0336, respectivamente. La estrella púrpura y el cuadrado son abundancias superficiales de Li para el Sol actual \cite{Asplund2009} y la media para el cúmulo de las Pléyades \cite{Sestito2005} respectivamente. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:li_var_vel_0g}
\end{figure}

La figura \ref{fig:li_var_vel_0g} muestra la evolución temporal de la abundancia superficial de Li para varios modelos $1\msun$ inicializados con diferentes velocidades de rotación. Las simulaciones tuvieron en cuenta los efectos de rotación y AML causados por los vientos estelares pero no los de MB. La estrella púrpura y el cuadrado son las abundancias superficiales de Li para el Sol actual \cite{Asplund2009} y para las Pléyades, respectivamente \cite{Sestito2005}.\par

Obsérvese cómo la abundancia de Li en la superficie estelar disminuyó con el tiempo para todos los modelos simulados. La línea negra continua representa el modelo de referencia que adopta los parámetros de superación de la envoltura calibrados para el Sol, como se documenta en \cite{Choi2016}. Todos los modelos quemaron demasiado Li antes de la ZAMS y, por tanto, no coincidieron con la abundancia media de Li en superficie de las Pléyades. También fue destacable el hecho de que apenas hubo diferencias entre los distintos modelos en cuanto a la abundancia de Li durante gran parte de la ZAMS. Sólo después de un millón de años se destruyó el Li en un grado acentuado, ya que antes no se alcanzó la temperatura necesaria en BCZ. Después, los diferentes modelos destruyeron el Li de forma muy similar debido a dos razones principales. Por un lado, las zonas convectivas que desarrollaron los modelos tenían tamaños muy similares, por lo que la temperatura en la BCZ era prácticamente la misma. Por otro lado, el diferencial de rotación entre el núcleo y la zona convectiva no comenzó a desarrollarse hasta que alcanzó los $ \approx 10^6$ años, alcanzando su máxima diferencia en la ZAMS alrededor de los $ \approx 10^7$ años. Fue en este momento cuando debieron aumentar los efectos combinados de la turbulencia y la diferencia rotacional, dando lugar a notables diferencias en la evolución de A(Li) (véase la figura \ref{fig:li_var_vel_0g_z1}). Posteriormente, el modelo de referencia (línea negra) no agotó eficientemente el Li en la EM y no consiguió (de nuevo) igualar la abundancia actual de Li en la superficie solar. Los otros modelos que incluían la rotación durante el PMS con valores de $\oomegac$ entre 0.0084 y 0.0336 fueron capaces de quemar Li de forma más realista. Entre ellos, sólo uno (línea verde) se aproximaba a la abundancia actual de Li del Sol, pero su velocidad de rotación era mucho mayor (véase la figura \ref{fig:rot_vel_0g}) que los $2\,\kms$ del Sol \cite{Gill2012}. \par

Durante gran parte del PMS la estrella giró como un cuerpo sólido (véase la Figura \ref{fig:rot_vel_0g}) y esto se debió a que la estrella tenía un interior totalmente convectivo. No fue hasta el final de la trayectoria de Hayashi cuando la estrella comenzó a desarrollar un núcleo radiativo. Fue en esta etapa cuando apareció una diferencia de velocidad angular entre los límites superior e inferior de las zonas radiativa y convectiva respectivamente. El grado de rotación diferencial estaba directamente influido por la $\Omega$ inicial. A medida que los modelos se inicializaban con una velocidad angular mayor, se acentuaba la diferencia de velocidad entre la BCZ y la superficie estelar; cuanto mayor era la velocidad inicial, mayor era el gradiente de velocidad entre los límites inferior y superior de la CZ. Como consecuencia, la fuerza de la turbulencia localizada en la BCZ aumentó, de modo que el Li pudo alcanzar regiones con temperaturas cercanas a $\tli$, donde finalmente se quemó y destruyó (véase la figura \ref{fig:li_var_vel_0g}). Otras investigaciones \cite{Bouvier2018, Baraffe2017} apuntan a una tendencia diametralmente opuesta a la aquí expuesta, es decir, a mayor velocidad de rotación, mayor abundancia de Li en la superficie de la estrella. \par

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/li_var_vel_0_0g_z1.pdf}
	\caption {Similar a la Figura \ref{fig:li_var_vel_0g} pero ampliando la ZAMS. Los modelos con una mayor velocidad de rotación inicial alcanzan ya la ZAMS con una menor cantidad de medida de Li en la superficie estelar.}
	\label{fig:li_var_vel_0g_z1}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.7\textwidth]{img/paper1/rot_vel_var_vel_0_0g.pdf}
	\caption{Evolución de la velocidad angular en la superficie (línea continua) y en el límite inferior (línea discontinua) de la zona convectiva superior, en función del tiempo para varios modelos de 1 $\msun$. Los modelos incluyen rotación PMS con valores $\oomegac$ entre 0,0084 y 0,0336. La estrella púrpura es la velocidad angular superficial para el Sol actual \cite{Gill2012}. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:rot_vel_0g}
\end{figure}

Otros efectos estructurales bien conocidos de la rotación son la disminución de la temperatura efectiva ($\teff$) y, en menor medida, de la luminosidad estelar ($L$). Ambos efectos pueden observarse gráficamente en el diagrama HR de la figura \ref{fig:hr_var_vel_0g}, que muestra una vista ampliada de las trayectorias evolutivas desde la ZAMS hasta la TAMS para varios modelos $\msun$ inicializados con diferentes velocidades de rotación. Si comparamos el modelo no rotatorio (línea sólida negra) con los rotatorios podemos reconocer que al final del PMS, estos últimos alcanzan la ZAMS con un $\teff$ menor que los primeros. Estos resultados coinciden con los de estudios anteriores \cite[véase por ejemplo ][]{Eggenberger2012,Piau2001,Pinsonneault1989}.\par

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/hr_var_vel_0_0g_z1.pdf}
	\caption{Un ejemplo de cuadrícula solar de 1$\msun$ de trayectoria evolutiva estelar desde el PMS hasta el TAMS cubriendo un amplio rango de velocidades angulares. La rotación se activa en los modelos en el PMS y esos modelos llegan antes a la ZAMS y a un $\teff$ menor que el de no rotación (línea negra continua). La luminosidad se expresa en términos de $\lsun$.}
	\label{fig:hr_var_vel_0g}
\end{figure}

\subsection{Evolución del Li con MB de intensidad fija}
La figura \ref{fig:li_var_vel_4_0g} muestra la evolución temporal de la abundancia superficial de Li para varios modelos de 1 $\msun$. Estos modelos se inicializaron con diferentes velocidades rotacionales y tuvieron en cuenta los efectos del MB causado por un campo magnético de intensidad 4G. Si lo comparamos con la Figura \ref{fig:li_var_vel_0g} en la que se despreciaron los efectos del MB, observamos cómo se alteraron los perfiles de abundancia de Li durante el PMS y el MS. Durante el PMS podemos describir el efecto como modesto, algo esperado y en línea con el hecho de que el AML causado por MB (ver Ec.~\ref{eq:j_dot}) depende directamente de la tasa de pérdida de masa. Si tenemos en cuenta que para las estrellas de tipo solar los modelos predicen una tasa de pérdida de masa total modesta, ese valor es incluso mucho menor en esta fase. Por el contrario, durante la EM se observa que la AML es mucho más significativa, provocando que se destruya una menor cantidad de Li.\par

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/li_var_vel_4_0g.pdf}
	\caption{Evolución de la abundancia superficial del \isotope[7]{Li} respecto al \isotope[1]{H}, en función del tiempo para varios modelos de 1 $\msun$. Los modelos incluyen un campo magnético con una intensidad de 4G y rotación PMS con $\oomegac$ entre 0,0084 y 0,0336, respectivamente. La estrella púrpura y el cuadrado son la abundancia superficial de Li para el Sol actual \cite{Asplund2009} y el cúmulo de las Pléyades \cite{Sestito2005} respectivamente. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:li_var_vel_4_0g}
\end{figure}

El efecto de la rutina MB puede apreciarse aún más claramente en las Figuras~\ref{fig:rot_vel_4g}, \ref{fig:rot_vel_4g_z1} \& Apéndices \footnote{Los apéndices comprenden una serie de cuadrículas en función del tiempo y para varios modelos de 1 $\msun$ en los que se muestra, por un lado, la evolución de la abundancia superficial del \isotope[7]{Li} respecto al \isotope[1]{H} tanto para intensidades de campo magnético variables como para velocidades angulares y, por otro, la evolución de la velocidad de rotación superficial.}. En estas figuras representamos los perfiles de rotación para la superficie de las estrellas y para el fondo de la envoltura convectiva. Son modelos de 1 $\msun$ inicializados con diferentes velocidades de rotación y considerando la influencia del MB. De forma similar a los perfiles de evolución de $A(\isotope[7]{Li})$ comentados en el párrafo anterior, el efecto de la rutina se hizo visible una vez alcanzada la ZAMS. Si comparamos la evolución de las curvas aquí presentadas con las de la Figura \ref{fig:rot_vel_0g} vemos como la estrella, en lugar de seguir aumentando $\Omega$, comenzó a ralentizarse tras haber alcanzado su máximo en el paso por la ZAMS. Nótese que en este punto las velocidades angulares en la superficie de la estrella y en la ZAMS alcanzaron su máxima diferencia. Por otro lado, el efecto MB hizo que las velocidades angulares entre ambas zonas de la estrella disminuyeran hasta que, para una edad cercana a la del Sol (ver Figura \ref{fig:rot_vel_4g_z1}), la estrella prácticamente rotaba como un sólido rígido. Estos resultados también eran coherentes con los obtenidos por \cite{Eggenberger2010} en cuanto al efecto del campo magnético, en particular su influencia en la pérdida de momento angular, sobre la velocidad de rotación de la estrella.\par

De forma similar, también observamos que los modelos con menor velocidad angular generalmente acababan mostrando valores más altos para la abundancia de Li en la superficie (ver Figuras~ \ref{fig:li_var_vel_4_0g}, \ref{fig:grid_li_var_vel} y \ref{fig:grid_li_var_g}). En ninguno de esos casos obtuvimos valores de Li en la superficie superiores a los mostrados por el modelo sin rotación.\par

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/rot_vel_var_vel_4_0g.pdf}
	\caption{La evolución de la velocidad de rotación de la superficie, en función del tiempo para varios modelos de 1 $\msun$. Los modelos incluyen un campo magnético con una intensidad de 4\,G, rotación PMS con $\oomegac$ entre 0,0084 y 0,0336, respectivamente y MB. La estrella púrpura es la velocidad angular superficial para el Sol actual \cite{Gill2012}. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:rot_vel_4g}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/rot_vel_var_vel_4_0g_z1.pdf}
	\caption{Similar a la Figura \ref{fig:rot_vel_4g} pero ahora mostrando en detalle la velocidad de rotación de la superficie a medida que la estrella se aproxima a TAMS.}
	\label{fig:rot_vel_4g_z1}
\end{figure}

El MB también dejó su huella en el diagrama HR afectando significativamente al $\teff$ de la estrella. Para visualizar este efecto tomamos como referencia la Figura \ref{hr_vc_0336_var_g_z1} en la que todos los modelos se iniciaron con el mismo valor $\oomegac=0,0336$ pero la intensidad del campo magnético simulado fue diferente. Obsérvese que los modelos con MB produjeron estrellas más calientes debido a su influencia. La menor velocidad con la que giraba la estrella debido al efecto MB provocó el aumento del $\teff$, siendo esta diferencia de prácticamente 95K entre los modelos simulados con 0.0G y 5.0G respectivamente para $log(\llsun)=0$.

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/hr_vc_0336_var_g_z1.pdf}
	\caption{Similar a la Figura \ref{fig:hr_var_vel_0g} pero ahora mostrando en detalle los efectos del frenado magnético en las trayectorias evolutivas para diferentes intensidades de campo magnético y $\oomegac=0.0336$. La presencia de un campo magnético produce estrellas más calientes debido a la influencia del frenado magnético en la velocidad de rotación de la estrella.}
	\label{hr_vc_0336_var_g_z1}
\end{figure}

Como se describe en la sección \ref{sec_mesh}, la rutina MB distribuyó la cantidad total de AML calculada según la Ec.~\ref{eq:k_jdot} entre las distintas capas que componían la CZ. En la Figura \ref{fig:cz_vc_028_var_b} podemos observar la evolución de la CZ más externa normalizada con respecto al radio de la estrella para varios modelos de 1 $\msun$. Todos los modelos se inicializaron con $\oomegac=0.0336$ e intensidades de campo magnético que variaban entre 0.0G y 5.0G. De acuerdo con los modelos establecidos de evolución estelar, en una estrella de tipo solar la CZ cubre prácticamente toda ella durante gran parte del PMS. A medida que se aproxima a la ZAMS, la CZ va disminuyendo como consecuencia de la aparición de un núcleo radiativo y mantiene un radio aproximadamente constante hasta la etapa final de la MS. En este punto aumenta significativamente como respuesta a la expansión generalizada del radio de la estrella. En cuanto al efecto deL MB sobre el tamaño de la CZ, observamos que a medida que aumentaba la intensidad del campo magnético, el tamaño de la CZ disminuía (véase la figura \ref{fig:cz_vc_028_var_b_z1_new}). El núcleo radiativo se desplazó hacia el exterior para incluir una fracción cada vez mayor de la masa estelar, haciendo que la temperatura en la base de la CZ cayera por debajo de $\tli$. Este efecto fue más evidente durante la MS. La disminución del tamaño de la CZ estuvo en consonancia con el hecho de que se destruyera menos Li haciendo que menos material estelar alcanzara zonas con temperaturas superiores a $\tli$.\par

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/cz_vc_028_var_g.pdf}
	\caption{Evolución del tamaño de la zona convectiva en función del tiempo para varios modelos de 1 $\msun$. Todos los modelos se inicializaron con $\oomegac=0.0336$ y las intensidades de campo magnético varían entre 0.0G y 5.0G. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:cz_vc_028_var_b}
\end{figure}

En la Figura \ref{fig:mdot_vc_028_var_b} podemos ver la evolución de $\Dot{M}$ durante PMS y MS para varios modelos de 1 $\msun$. Todos los modelos se inicializaron con $\oomegac=0.0336$ e intensidades de campo magnético que variaban entre 0.0G y 5.0G. En las primeras etapas del PMS fue donde se concentró la mayor pérdida de masa, que fue disminuyendo a medida que se acercaba a la ZAMS. Como la estrella también disminuyó su radio durante la PMS, aumentó $\Omega$ obedeciendo al principio de conservación del AM. Al llegar a la ZAMS, el radio estelar se mantuvo más o menos estable durante gran parte de la MS (excepto en su etapa final, véase la figura \ref{fig:mdot_vc_028_var_b_z1}), pero siguió perdiendo masa mientras aumentaba $\Omega$, aunque de forma menos agresiva si lo comparamos con la PMS. Sin embargo, como consecuencia tanto de la aparición del núcleo radiativo durante el seguimiento de Henyey como de la existencia de una CZ (véase la figura \ref{fig:mb_act_var_vel_vc_028}), se activó la rutina MB, provocando que la velocidad angular de la estrella comenzara a disminuir a lo largo de toda la MS. Cuanto más intenso es el campo magnético, mayor es el efecto de frenado.\par

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/mdot_vc_028_var_g.pdf}
	\caption{Evolución de la pérdida de masa $\Dot{M}$ en función del tiempo para varios modelos de 1 $\msun$. Los modelos incluyen diferentes intensidades de campo magnético entre 0.0G y 5.0G y rotación PMS con $\oomegac=0.028$. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:mdot_vc_028_var_b}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/mdot_vc_028_var_g_z1.pdf}
	\caption{La evolución de la pérdida de masa $\Dot{M}$ a medida que la estrella se acerca a TAMS, en función del tiempo para varios modelos de 1 $\msun$. Los modelos incluyen una intensidad de campo magnético variable entre 0.0G y 5.0G y rotación PMS con $\oomegac=0.028$. Cuanto más intenso sea el campo magnético, menor será $\Dot{M}$.}
	\label{fig:mdot_vc_028_var_b_z1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/mb_act_vc_028_var_g.pdf}
	\caption{La activación de la rutina de frenado magnético en función de la presencia de un núcleo radiativo. Los modelos incluyen un campo magnético con una intensidad comprendida entre 0.0G y 5.0G y una rotación PMS con $\oomegac=0.0228$. Las líneas continuas señalan la activación (on) y desactivación (off) de la rutina de frenado magnético. Las líneas horizontales discontinuas informan sobre la naturaleza del núcleo de la estrella: radiativo (rad) o convectivo (conv). Por decisión de implementación, una vez activada la rutina, permanece activada incluso si la naturaleza del núcleo de la estrella cambia a convectiva. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:mb_act_var_vel_vc_028}
\end{figure}

\subsection{Evolución alternativa del Li con MB}
Hasta este punto, las simulaciones de los diferentes modelos se basaron en la parametrización recogida en la Tabla \ref{tab:phy_mesa}, que a su vez fue adoptada de \cite{Choi2016}. Si recordamos la evolución del Li mostrada en las Figuras \ref{fig:li_var_vel_0g} y \ref{fig:li_var_vel_4_0g}, destacamos que en ambos casos se quemaba demasiado Li antes de llegar a la ZAMS y, por tanto, no coincidía con la abundancia media de Li en superficie de las Pléyades. Por otro lado, utilizando una parametrización ligeramente distinta de la empleada hasta ahora en la que los parámetros de convección y \textit{overshooting} se han reajustado según la Tabla \ref{tab:phy_alt_mesa}, las simulaciones pudieron reproducir con mayor fidelidad tanto las abundancias de Li del cúmulo de las Pléyades como las del Sol (véase la Figura \ref{fig:li_3_0g_0314vc}).\par

\begin{table}
	\centering
	\caption{Alternative MTL and overshooting parameters.}
	\label{tab:phy_alt_mesa}
	\begin{tabular}{ll} 
		\hline
		Parameter & Adopted prescriptions and values\\
		\hline
		Convection & MLT + Ledoux, $\alpha_\mathrm{MLT}$ = 1.70\\
		Overshoot & time-dependent, diffusive, $f_\mathrm{ov,core}=0.016$, \\ & $f_\mathrm{ov,sh}=0.002$\\
		\hline
	\end{tabular}
\end{table}

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/li_3_0g_0314vc.pdf}
	\caption{La evolución de la abundancia superficial del \isotope[7]{Li} respecto al \isotope[1]{H}, en función del tiempo para un modelo de 1 $\msun$. El modelo incluye un campo magnético con una intensidad de 3G y rotación PMS con $\oomegac= 0.0314$, respectivamente. Los parámetros MLT y overshooting se han fijado en $\alpha_\mathrm{MLT}=1.70$, $f_\mathrm{ov,core}=0.016$, $f_\mathrm{ov,sh}=0.002$. La estrella púrpura y el cuadrado son la abundancia superficial de Li para el Sol actual \cite{Asplund2009} y el cúmulo de las Pléyades \cite{Sestito2005} respectivamente. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:li_3_0g_0314vc}
\end{figure}

Del mismo modo, la evolución de la velocidad angular para esta nueva configuración se muestra en la Figura \ref{fig:rot_vel_var_vel_mlt_3_0g}. Aquí todavía no reproducimos correctamente la velocidad angular del Sol. Se hizo evidente que la influencia de los parámetros libres, relativamente arbitrarios, asociados a MLT condicionaba significativamente la evolución de la abundancia de Li. Sin embargo, sin la inclusión de MB no fue posible ajustarlo para las Pléyades y el Sol con la misma trayectoria evolutiva. Este hecho abre la puerta a otras parametrizaciones compatibles con el Sol para reproducir las observaciones en cúmulos estelares jóvenes y para el Sol.\par

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img/paper1/rot_vel_3_0g_0314vc.pdf}
	\caption{La evolución de la velocidad de rotación de la superficie, en función del tiempo para un modelo de 1 $\msun$. El modelo incluye un campo magnético con una intensidad de 3G, rotación PMS con $\oomegac=0.0314$, parámetros en la Tabla \ref{tab:phy_alt_mesa} y MB. La estrella púrpura es la velocidad angular superficial para el Sol actual \cite{Gill2012}. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:rot_vel_var_vel_mlt_3_0g}
\end{figure}

\section{Conclusiones} \label{sec_conclusions}
Hemos demostrado, a través de los diferentes modelos estelares simulados, que los efectos inducidos por la combinación de ambos mecanismos de rotación y frenado magnético ofrecen una forma plausible de reconciliar los datos observacionales con los modelos teóricos. Esto último es de gran importancia tanto para el transporte de AM como para el transporte de elementos químicos. También somos muy conscientes de que aún estamos lejos de comprender los mecanismos físicos exactos que gobiernan estos procesos, por lo que es necesario seguir profundizando en estas áreas de estudio. En particular, el estudio de la evolución de los campos magnéticos durante la PMS y la MS, y su impacto en la AML. \par

Se realizarán futuras y mejoradas implementaciones de nuestra rutina para estudiar los resultados sobre las huellas, la composición superficial del Li, la rotación, etc. Es probable que en un modelo estelar en rotación simulado con MB desde el principio, la rotación diferencial sea muy reducida y, por tanto, las abundancias de Li observadas en estrellas jóvenes podrían explicarse adecuadamente. Proponemos que este resultado podría alcanzarse mediante una ley de interdependencia entre $\Omega$ y $B$. Así, durante el PMS, cuando la estrella rota más rápido, el efecto MB es más eficiente. En cambio, durante el MS, cuando el arranque se ralentiza, el MB será menos intenso. Este enfoque establecería un mecanismo de autorregulación sobre la velocidad angular de la estrella que acabaría influyendo directamente en la evolución del Li. Es igualmente importante comprender mejor el papel general de la pérdida de masa en AML y sobre todo en estrellas más frías y de baja masa. Hoy en día es un reto determinar la velocidad terminal del viento estelar para este tipo de estrellas que juega un papel clave en la AML.\par

Subrayamos también que nuestros modelos no coincidieron al mismo tiempo con la abundancia de Li solar observada y $\Omega$. Por lo tanto, no podemos asegurar que hayamos modelado correctamente la historia rotacional del Sol. En vista de estas deficiencias de nuestros modelos, debemos analizar los resultados obtenidos con cautela y no sacar conclusiones prematuras. También hemos mostrado cómo una parametrización MLT alternativa podría producir resultados acordes con las observaciones. En el transcurso de este trabajo se han apoyado las siguientes conclusiones:
\begin{enumerate}
	\item La inclusión del campo magnético conduce a modelos más fríos y a un menor agotamiento de Li en la EM.
	\item Una combinación de rotación durante la PMS y efecto MB durante la MS produce un comportamiento diferente y potencialmente más prometedor que los producidos por los modelos estándar. Así, nuestro enfoque apunta a reproducir el A(Li) observado y la velocidad de rotación solar al mismo tiempo.
	\item $\teff$ de las trayectorias evolutivas estándar representan límites superiores, ya que estos modelos no tienen en cuenta el efecto de frenado magnético ni la rotación.
	\item La extensión de la zona convectiva disminuye cuando aumenta la intensidad del campo magnético.
	\item El MB durante la PMS y/o el ajuste de los parámetros libres de exceso de MLT y $\amlt$ parece ser también necesario para explicar las abundancias de Li en cúmulos jóvenes.
\end{enumerate}

\endinput
%--------------------------------------------------------------------
% FIN DEL CAPÍTULO. 
%--------------------------------------------------------------------

### 2024-01-29 project/overleaf/Tesis III/capitulos/capitulo06.tex
Add Tesis project

\chapter{Rotación y campos magnéticos de intensidad variable}\label{ch:sexto-capitulo}

\section{Introducción}
--EXPLICAR POR QUÉ OPTAMOS POR LA INTENSIDAD VARIABLE

\section{Frenado magnético según Gallet \& Bouvier (2013)}
Suponer que la intensidad del campo magnético de una estrella permanece constante a lo largo de su evolución es una simplificación necesaria en algunos escenarios, pero no se corresponde con las observaciones (INDICAR LAS REFERENCIAS). \par

De acuerdo a \cite{Gallet2013} la intensidad del campo magnético debería variar a lo largo de la vida de la estrella en lugar de permanecer fija durante toda su evolución. Para ello nos basamos en un formalismo que nos permite calcular una intensidad de campo magnético variable. Este modelo adopta una prescripción de tipo dinamo que permite establecer una relación entre $\Omega$ , $\teff$, y la intensidad del campo magnético ($B$), así como con la tasa de pérdida de masa ($\mwind$) causada por el viento estelar. Comenzaremos enumerando los aspectos más relevantes y las suposiciones realizadas en la modelización de la evolución de la intensidad del campo magnético.\par

Un parámetro relevante para caracterizar la influencia de un campo magnético dado en el viento estelar es $\Omega$. Suponiendo que el campo magnético es generado por una dinamo, su fuerza es proporcional a alguna potencia de $\Omega$,

\begin{ceqn}
	\begin{equation}
		f_*B_* \propto \Omega_*^b \label{eq:mf_strenght}
	\end{equation}
\end{ceqn}

donde $f_*$ es el factor de llenado, es decir, la fracción de la superficie estelar que está magnetizada, $B_*$ es la fuerza del campo magnético, $\Omega_*$ es la velocidad angular de la estrella en la superficie estelar, y $b$ es el exponente de la dinamo \cite{Gallet2013}. El producto de $f_*$ por $B_*$ nos permitirá obtener la intensidad del campo magnético. \par

Según este enfoque, la intensidad del campo magnético variará proporcionalmente a la temperatura efectiva ($\teff$) de la estrella de la siguiente manera: 
\begin{ceqn}
	\begin{equation}
		B_* \approx 1.13 \Beq \label{eq:mf_bstar}
	\end{equation}
\end{ceqn}

donde $B_*$ es proporcional a la intensidad del campo magnético de equipartición, quedando $\Beq$ definido como

\begin{ceqn}
	\begin{equation}
		\Beq = \sqrt{8\pi P_*} = \sqrt{\frac{8\pi\rho_* \boltzmann \teff}{\mu\massH}}\label{eq:mf_beq}    
	\end{equation}
\end{ceqn}

donde $P_*$ es la presión del gas fotosférico, $\rho_*$ es la densidad fotosférica, $\boltzmann$ es la constante de Boltzmann, $\massH$ es la masa de un átomo de hidrógeno, y $\mu$ el peso atómico medio \cite{Cranmer2011}. \par

Como se indica en \cite{Cranmer2011} $\mu$ puede estimarse utilizando las ecuaciones de estado del plasma OPAL \footnote{\url{https://opalopacity.llnl.gov/EOS_2005/}},

\begin{ceqn}
	\begin{equation}
		\mu \approx \frac{7}{4} + \frac{1}{2} \; \textrm{tanh}\Bigg(\frac{3500-\teff}{600}\;\Bigg) \label{eq:mf_atom_weight}
	\end{equation}
\end{ceqn}

Una vez que estemos en condiciones de calcular $\Beq$, podremos calcular $B_*$.\par

Por otro lado, \cite{Cranmer2011} descubrió que la intensidad del campo magnético está influida principalmente por $f_*$ y no por el periodo de rotación de la estrella. Es decir, las variaciones en la velocidad angular de la estrella no alteran significativamente la intensidad de su campo magnético. Además, $f_*$ depende en gran medida del número de Rossby ($\rossby$). Para calcular $\rossby$ de una estrella determinada, es necesario conocer el tiempo de vuelta ($\turnover$),

\begin{ceqn}
	\begin{equation}
		\rossby = \frac{\prot}{\turnover} \label{eq:mf_rossby}
	\end{equation}
	\begin{equation}
		\turnover = 314.24\;exp\left[-\Bigg(\frac{\teff}{1952.5}\Bigg)-\Bigg(\frac{\teff}{6250}\Bigg)^{18} \;\right]+0.002 \label{eq:mf_turnover}
	\end{equation}
\end{ceqn}

\cite{Cranmer2011} Crammer proporciona dos ajustes diferentes para $f_*$ que definen respectivamente los límites superior e inferior para la intensidad media del campo magnético superficial ($f_*B_*$),

\begin{ceqn}
	\begin{align}
		\fmin &= \frac{0.5}{(1+(x/0.16)^{2.6})^{1.3}} \label{eq:fmin}\\
		\fmax &= \frac{1}{1+(x/0.31)^{2.5}} \label{eq:fmax}
	\end{align}
\end{ceqn}

donde x = $\rossby$/$\rossbysun$, y $\rossbysun$ = 1.96. \par

Por último, podemos calcular el factor de llenado de una estrella aplicando el ajuste realizado por \cite{Gallet2013} que reproduce más fielmente el factor de llenado medio del Sol actual ($\fsun$ = 0.001-0.01),

\begin{ceqn}
	\begin{equation}
		\fstar = \frac{0.55}{(1+(x/0.16)^{2.3})^{1.22}} \label{eq:fstart}
	\end{equation}
\end{ceqn}

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{img/paper2/mag_field_limits_var_vel_g3.pdf}
	\caption{La evolución de la intensidad del campo magnético y sus límites superior e inferior, en función del tiempo y $\oomegac$ para varios modelos de 1 $\msun$. El modelo incluye rotación PMS con $\oomegac$ = 0,1425. Las líneas continuas representan la intensidad del campo magnético, mientras que las líneas discontinuas representan la evolución angular de la estrella. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:mag_field_limits_var_vel_g3}
\end{figure}

Con este formalismo podemos calcular el campo magnético medio como una función dependiente del tiempo, de $\rho$, $\teff$, y $\Omega$. Los valores temporales de estas tres variables se obtendrán a partir de las distintas simulaciones realizadas con ayuda de MESA. \par

Una vez que hemos caracterizado cómo evoluciona la intensidad del campo magnético, necesitamos modelizar cómo se produce la pérdida de momento angular.  En el formalismo anterior utilizamos un modelo aplicable a campos magnéticos de intensidad no variable y de baja intensidad ($1,0\,\Gauss \leq B \leq 10,0\,\Gauss$). Como se muestra en la Figura \ref{fig:mag_field_limits_var_vel_g3}, las simulaciones con el presente formalismo arrojan valores de campo magnético que muestran intensidades máximas en torno a $1,0K\,\Gauss$, lo que hace inviable el uso del modelo empleado anteriormente. \par

Comencemos por enumerar los aspectos más relevantes y las suposiciones realizadas en la modelización de la evolución de la rotación, el frenado magnético y el momento angular. Si suponemos un flujo de salida esférico, el par aplicado por un viento estelar $\torquewind$ acoplado magnéticamente viene dictado por,

\begin{ceqn}
	\begin{equation}
		\torquewind \propto \Omega_* \; \mwind \; \ralfven^{2} \label{eq:mb_torque}
	\end{equation}
\end{ceqn}

donde $\mwind$ es la tasa de pérdida de masa, y $\ralfven$ el valor medio del radio de Alfvén. \par

Las estrellas con masas iniciales similares, pero con diferentes ratios de pérdida de masa ($\Dot{M}$), acabarán evolucionando de forma muy diferente. Las partículas ionizadas transportadas por el viento solar no sólo contribuyen a la pérdida de masa, sino también a la pérdida de energía cinética que se deposita en el medio interestelar. Dada una estrella con un viento esféricamente simétrico, $\mwind$ se caracteriza por la siguiente expresión:

\begin{ceqn}
	\begin{equation}
		\mwind = 4\pi r^2\rho\nu \label{eq:mass_loss}
	\end{equation}
\end{ceqn}

donde $r$ es el radio estelar, $\rho$ la densidad de masa, y $\nu$ la velocidad del viento estelar.

Se ha observado que las estrellas de masa intermedia fuertemente magnéticas suelen tener tasas de rotación mucho más lentas que otras estrellas de su población parental \cite{Mathys2006}. En esas estrellas, los campos magnéticos interactúan con la pérdida de masa, donde el radio Alfv'{e}n juega un papel importante. $\ralfven$ se define como el punto en el que la densidad de energía del campo magnético y la densidad de energía cinética están equilibradas. En caso de que $\ralfven$ sea mayor que el radio estelar, entonces el flujo del viento tendrá que seguir el campo magnético. Como consecuencia, el material abandona la superficie estelar con una mayor AM específica, ya que el radio de co-rotación ha aumentado y corresponde aproximadamente a $\ralfven$ que puede expresarse como \cite{Matt2012}

\begin{ceqn}
	\begin{equation}
		\ralfven = K_1\left[\frac{\Bp^{2}\;R_*^{2}}{\mwind\;\sqrt{K_2^2\vesc^2 + \Omega_*^2R_*^{2}}\ }\right]^{m}R_*  \label{eq:mb_ralfven}
	\end{equation}
\end{ceqn}

donde $\Bp$ es la intensidad del campo magnético en la superficie estelar, $\vesc$ es la velocidad de escape, $m = 0,1675$, $K_1 = 1,30$ y $K_2 = 0,0506$ \cite{Gallet2013}.

\begin{ceqn}
	\begin{equation}
		\vesc = \sqrt{\frac{2\,G\,\mstar}{\rstar}} \label{eq:vesc}
	\end{equation}
\end{ceqn}

Por último, tenemos que el AML puede calcularse del siguiente modo:

\begin{ceqn}
	\begin{equation}
		\Dot{J} = \Omega_* \; \mwind \; \ralfven^{2} \label{eq:j_dot}
	\end{equation}
\end{ceqn}

donde $\mwind$ es la tasa de pérdida de masa. Esta expresión se implementa en MESA a partir de los valores expuestos directamente durante las simulaciones.

\section{Efectos de la longitud de mezcla}
La teoría de la longitud de mezcla (MLT) introducida por Böhm-Vitense (BV) ha sido comúnmente adoptada para modelar la convección estelar en códigos de evolución estelar, incluyendo MESA. El parámetro libre más importante en esta teoría es la longitud de mezcla $l$ que se define como:

\begin{ceqn}
	\begin{equation}
		l = \alpha H_p\label{eq:mixlength}
	\end{equation}
\end{ceqn}

donde $H_p$ es la altura de la escala de presión, y $\alpha$ un parámetro libre que se determina de antemano y se mantiene fijo durante las simulaciones de evolución estelar. \par

La influencia de los parámetros libres, relativamente arbitrarios, asociados a MLT condiciona significativamente la evolución de la abundancia de Li. Además, no existen argumentos sólidos que sugieran que el parámetro de longitud de mezcla sea el mismo en todas las estrellas y en todas las fases evolutivas \cite{Pasetto2014}. En \cite{Navarro2020} mostramos cómo una parametrización MLT alternativa podría producir resultados en línea con las observaciones. El ajuste del sobreimpulso (over-shooting) de MLT y del parámetro libre $\amlt$ parece ser necesario para explicar las abundancias de Li en los OC's.\par

De manera similar a la intensidad del campo magnético, un formalismo que nos permita evolucionar a lo largo del tiempo el valor de $\amlt$ en función de determinados parámetros de la estrella, es un planteamiento razonable a la hora de eliminar una restricción relativamente arbitraria impuesta en nuestros modelos.

\subsection{MLT $\alpha$ variable según Sonoi et al. (2019)}
En \cite{Sonoi2018}, los autores introdujeron una forma de calibrar $\amlt$ para estrellas de tipo solar en la que su valor aumenta al aumentar $\gsurf$ o disminuir $\teff$,

\begin{ceqn}
	\begin{equation}
		f(x,y) = a_0 + (a_1 + (a_3 + a_5x +a_6y)x + a_4y)x + a_2y\label{eq:alpha_ml}
	\end{equation}
\end{ceqn}
donde
\begin{ceqn}
	\begin{align}
		x &= \frac{\teff-5777}{1000} \label{eq:eq:alpha_x}\\
		y &= \gsurf-4.44 \label{eq:eq:alpha_y}
	\end{align}
\end{ceqn}

y $a_i$ son los coeficientes resultantes de la función de ajuste \ref{eq:alpha_ml} a los valores $\alpha$ calibrados para el modelo de convección MLT-BV \cite{Sonoi2018}: $a_0=1,790295$, $a_1=-0,14954$, $a_2=0,069574$, $a_3=-0,00829$, $a_4=0,013165$, $a_5=0,080333$, $a_6=-0,03306$. \par
--COMENTAR MÁS EN DETALLE CÓMO POR EJEMPLO SE HACE VARIAR EN MESA. TAMBIÉN ALGO MÁS DEL FORMALISMO--

\section{Extendiendo MESA - Parte II}
\subsection{Rutina de frenado magnético de intensidad variable}
EXPLICAR QUE SUSTITUIMOS LA RUTINA POR DEFECTO CON LA NUESTRA
\begin{lstlisting}[language=Fortran, caption={Rutina de par de torsión.}, label={lst:jdot_gb}]
	real function calculate_jdot_rate_gb(s, bf_star) result(new_j_dot)
	use const_def
	type (star_info), pointer, intent(in) :: s
	real(dp), intent(in) :: bf_star
	real(dp), parameter :: k1 = 1.30
	real(dp), parameter :: k2 = 0.0506
	real(dp), parameter :: m = 0.1675
	real(dp), parameter :: omega_sun = 2.87d-6
	real(dp) :: r_st, m_st, omega_surf, j_dot
	real(dp) :: v_esc, m_dot, alfven_r, alfven_r_num, alfven_r_den

	!Star data
	r_st = s% r(1)
	m_st = s% m(1)
	omega_surf = s% omega(1)
	m_dot = abs(s% mstar_dot) !This in g/s

	!eq 3 [1]
	v_esc = (2 * standard_cgrav * m_st / r_st)**0.5

	alfven_r_num = bf_star**2 * r_st**2

	alfven_r_den = m_dot*((k2**2 * v_esc**2)+(omega_surf**2 * r_st**2))**0.5

	alfven_r = k1 * (alfven_r_num / alfven_r_den)**m * r_st

	j_dot = omega_surf * m_dot * alfven_r**2

	new_j_dot = -j_dot
	end function

\end{lstlisting}

\subsection{Rutina de cálculo de intensidad de campo magnético}
EXPLICAR QUE SUSTITUIMOS LA RUTINA POR DEFECTO CON LA NUESTRA
\begin{lstlisting}[language=Fortran, caption={Rutina de cálculo de intensidad de campo magnético.}, label={lst:jdot_mf_intensity}]
	real function calculate_mag_field_intensity_gb(s) result(new_b)
	use const_def
	type (star_info), pointer, intent(in) :: s

	! rossby_sun = 1.96, Eq 9 [1]
	real(dp) :: r_st, rossby_sun = 1.96, rossby, rossby_norm, m_hydrogen
	real(dp) :: p_rot, p_rot_sun = 25.38 !Carrington solar  rotation
	real(dp) :: tau_c_sun, tau_c, f_star, f_min, f_max, b_star
	real(dp) :: bf_min, bf_max, bf_star, b_equi, mu_avg
	real(dp) :: p_gas_photo, p_photo, p_rad_photo, p_photo2

	!tau_c -> convective turnover time (d)
	!exp ->  natural exponential function
	! Eq 36 [2]
	tau_c = 314.241*exp(-s% Teff/1952.5)*exp(-(s% Teff/6250)**18) + 0.002 
	tau_c_sun = 314.241*exp(-teffsun/1952.5)*exp(-(teffsun/6250)**18) + 0.002

	!Star data
	r_st = s% r(1)
	p_rot = (2*pi*r_st) / (s% v_rot_avg_surf * 86400) ! day

	!Rossby number = p_rot / tau_c
	rossby = p_rot / tau_c
	rossby_sun = p_rot_sun / tau_c_sun
	rossby_norm = rossby / rossby_sun

	!magnetic filling factor fmin & fmax
	!eq 9 [1]
	f_min = 0.5 / (1. + (rossby_norm / 0.16)**2.6)**1.3 
	!eq 10 [1]
	f_max = 1 / (1. + (rossby_norm / 0.31)**2.5)
	!f_star = f_min
	!eq 11 [1]
	f_star = 0.55 / (1. + (rossby_norm / 0.16)**2.3)**1.22

	!Eq 3 [2]
	!mean atomic weight
	mu_avg = 1.75 + 0.5 * tanh((3500. - s% Teff) / 600.)
	!photospheric pressure 
	m_hydrogen = mp !proton mass       
	p_photo = s%rho(s% photosphere_cell_k)*boltzm*s% Teff/(mu_avg*m_hydrogen)
	!Eq 8 [1]
	!Equipartition magnetic filed strength
	b_equi = sqrt(8.* pi * p_photo)

	!Eq 7 [1], magnetic field strength b_star is proportional to the
	!equipartition magnetic field strength b_equi
	b_star = 1.13 * b_equi
	bf_min = f_min * b_star
	bf_max = f_max * b_star
	bf_star = f_star * b_star

	!return the min value, as it does the paper
	new_b = bf_star
	end function

\end{lstlisting}

\subsection{Rutina de $\amlt$ variable}
EXPLICAR QUE LOS DOS TIPOS DE RUTINAS: INTESIDAD FIJA, INTENSIDAD VARIABLE
\begin{lstlisting}[language=Fortran, caption={Rutina de $\amlt$ variable.}, label={lst:var_amlt}]
	!Calibration of the mixing-length parameter alpha for the MLT
	!and FST models by matching with CO5BOLD models
	real function calculate_alpha(s) result(new_alpha)
	type (star_info), pointer, intent(in) :: s

	!Parameters taken from Table 2 - MTL(BV) Eddington

	real(dp) :: a0 = 1.790295, a1 = -0.149542, a2 = 0.069574, a3 = -0.008292
	real(dp) :: a4 = 0.013165, a5 = 0.080333, a6 = -0.033066
	real(dp) :: x, y

	!f(x, y) = a0 + (a1 + (a3 + a5x + a6y)x + a4y)x + a2y
	!where x = (Teff-5777)/1000 and y = log g-4.44
	!Namely, x and y represent deviation from the solar effective 
	!temperature and surface gravity, respectively.
	x = (s% Teff - 5777) / 1000
	y = s% log_surface_gravity - 4.44

	!Equation 4
	new_alpha = a0 + (a1 + (a3 + a5*x + a6*y)*x + a4*y)*x + a2*y
	s% x_ctrl(47) = new_alpha
	end function

\end{lstlisting}

MESA no proporciona un \textit{hook} particular para modificar la rutina por defecto que trae el simulador. Para poder hacer evolucionar el valor de $\amlt$ recurrimos a incorporar nuestra rutina en la que MESA ejecuta al final de cada paso de simulación. De este modo conseguimos que en la siguiente iteración, utilice el valor que hemos calculado.
\begin{lstlisting}[language=Fortran, caption={Rutina de MESA para ejecutar lógica adicional tras la ejecución de un paso de simulación.}, label={lst:extra_start_step}]
	integer function extras_start_step(id, id_extra)
	integer, intent(in) :: id, id_extra
	integer :: ierr
	real(dp) :: x, y, new_alpha
	type (star_info), pointer :: s
	ierr = 0
	call star_ptr(id, s, ierr)
	if (ierr /= 0) return
	extras_start_step = keep_going

	if (var_mlt_alpha .eqv. .true.) then
	s% mixing_length_alpha = calculate_alpha(s)
	end if

	end function extras_start_step
\end{lstlisting}

\endinput
%--------------------------------------------------------------------
% FIN DEL CAPÍTULO. 
%--------------------------------------------------------------------

### 2024-01-29 project/overleaf/Tesis III/capitulos/capitulo07.tex
Add Tesis project

--- 2024-01-29 project/overleaf/Tesis III/capitulos/capitulo05.tex
@@
-\chapter{Resultados de frenado magnético de intensidad fija}\label{ch:quinto-capitulo}
+\chapter{Resultados de frenado magnético de intensidad variable}\label{ch:septimo-capitulo}
 \section{Configuración de los modelos} \label{marco_teorico}
@@
 \section{Modelos de evolución estelar}
-De acuerdo con la formulación de las secciones anteriores, los dos únicos parámetros libres de nuestra implementación son $B$ y $\Omega$.
-Las simulaciones numéricas trazaron la historia rotacional y $A(\isotope[7]{Li})$ de una estrella 1 $\msun$ para una variedad de valores iniciales para $B$ y $\Omega$ (ver también Tabla \ref{tab:phy_mesa}).
-MESA utiliza el enfoque en capas (shellular) \cite{Meynet1997} para dar cuenta de los efectos hidrostáticos de la rotación en modelos estelares 1D.\par
-
-\begin{table}
-	\centering
-	\caption{Summary of adopted physics in MESA \cite[based on][]{Choi2016}.}
-	\label{tab:phy_mesa}
-	\begin{tabular}{ll} 
-		\hline
-		Parameter & Adopted prescriptions and values\\
-		\hline
-		Solar Abundance & $X_{\odot}=0.7154, Y_{\odot}=0.2703, Z_{\odot}=0.0142$\\
-		Equation of State & OPAL+SCVH+MacDonald+HELM+PC\\
-		Opacity & OPAL Type I for log T $\geq$ 4 \\ & Ferguson for logT $\leq$ 4\\
-		Reaction Rates & JINA REACLIB\\
-		Boundary Conditions & ATLAS12; $\tau$=100 tables + photosphere\\
-		Difussion & Track \isotope[1]{H}, \isotope[2]{He}, \isotope[7]{Li}, \isotope[7]{Be}\\
-		Rotation & Differential rotation at PMS \& MS\\
-		Convection & MLT + Ledoux, $\alpha_{MLT}$ = 1.82\\
-		Overshoot & time-dependent, diffusive, \\ & $f_{ov,core}=0.0160$,\\ 
-		& $f_{ov,sh}=0.0174$\\
-		Semiconvection & $\alpha_{sc}=0.1$\\
-		Thermohaline & $\alpha_{th}=666$\\
-		Rotational Mixing & Include SH, ES, GSF, SSI \& DSI\\
-		Magnetic Effects & Magnetic braking based on idealized \\ & monopole field\\
-		Magnetic Field & B(G) variable between [3.0 - 5.0]\\
-		Mass Loss & activated, $\Dot{M}_{max} = 10^{-3} \: \msun \: yr^{-1}$\\
-		Angular Moment Loss & activated, $\Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A}$\\
-		\hline
-	\end{tabular}
-\end{table}
-
-Adoptamos abundancias a escala solar y asumimos una abundancia inicial solar ($Z = Z_{\odot, \mathrm{pr} = 0,0142}$).
-También adoptamos los siguientes valores nominales para expresar las propiedades estelares en unidades SI, $\rsun = 6,957x10^{10}\, cm$ y $\msun = 1,988x10^{33}\, g$ que son coherentes con las resoluciones de la IAU \cite{Mamajek2015}.
-Para una descripción detallada de la física adoptada en este trabajo, remitimos al lector a \cite{Choi2016}.
-Ese trabajo se utilizó como punto de partida para el nuestro en lo que respecta a la parametrización del proyecto MESA, que se calibró para reproducir las abundancias de elementos medidas en la superficie solar.\par
-
-Los modelos incluyeron la rotación durante la PMS ya que existen evidencias que abogan por una fuerte relación establecida entre la destrucción de Li y la rotación en esa fase (\cite{Bouvier2016,Bouvier2018}).
-Modelamos la rotación como AML se calcula como resultado del par aplicado a la zona de convección por un viento acoplado magnéticamente.
-Además, no tuvimos en cuenta ni la influencia de los campos magnéticos internos ni su existencia durante la fase T-Tauri.\par
----CORREGIR ESTO SEGUN EL NUEVO PAPERAdoptamos un enfoque teórico sencillo y pragmático para establecer cuándo la estrella está alcanzando la fase ZAMS.
-El criterio elegido se basó en la existencia simultánea de una extensa capa convectiva y un núcleo radiativo.
-A partir de este momento, se activó la rutina MB, actuando como un mecanismo adicional a los existentes en el código evolutivo MESA que participó en el AML de la estrella.
----HASTA AQUÍ---\par
-
-Supusimos que el campo magnético no variaba su intensidad a lo largo de la evolución de la estrella hasta alcanzar el TAMS.
-Además, también se consideraron otros efectos de rotación durante la evolución de los modelos: el transporte de AM desde el interior radiativo hasta la envoltura convectiva y la redistribución de AM asociada a cambios en la estructura interna durante el proceso de contracción hacia la EM.\par
-
-Nos centramos en el papel indirecto que desempeña el MB en la destrucción del Li.
-Esto es consecuencia de su influencia en la historia rotacional de las estrellas de tipo solar.
-Nuestro principal objetivo era comprobar cómo el MB podría contribuir a explicar la evolución del Li en el Sol y en otras estrellas de tipo solar.
-Se ejercitaron un conjunto de escenarios diferentes con intensidades de campo magnético que oscilaban entre 3,0 y 5,0 G.
-Calculamos la evolución de modelos estelares 1 $\msun$ con metalicidad inicial solar y $\oomegac$ variable entre $0.0084$ y $0.0336$.
-Los modelos más avanzados para disposiciones más complejas del campo magnético, así como la evolución de su intensidad a lo largo de la vida de la estrella, se dejaron para trabajos posteriores.\par
-
-Como expresa la Ec.~\ref{eq:j_dot_mesa}, la cantidad de AML depende de $R_A$ (o alternativamente de $\eta_*$), $\Omega$ y $\Dot{M}$.
-Para un valor grande de $\eta_*$, la estrella experimenta una desaceleración significativa.
-Controlamos el valor de $\eta_*$ en los modelos variando el campo magnético ($B$) y la velocidad de rotación estelar ($\Omega$).
-Con respecto a $\Dot{M}$ y como se indica en la tabla \ref{tab:phy_mesa}, para calcular la pérdida de masa se utilizó la fórmula empírica desarrollada por Reimers \cite{Reimers1975} para estrellas en la rama gigante asintótica (AGB).
-Para una estrella de tipo solar el $\Dot{M}$ durante la EM es relativamente pequeño, alrededor de $10^{-14}\msun \, yr^{-1}$ \cite{Noerdlinger2008}.
-\par
-
-\subsection{Evolución de Li sin MB}
-
-\begin{figure}
-    \centering
-	\includegraphics[width=0.7\textwidth]{img/paper1/li_var_vel_0_0g.pdf}
-	\caption{La evolución de la abundancia superficial del \isotope[7]{Li} en relación con el \isotope[1]{H}, en función del tiempo para varios modelos de 1 $\msun$.
-La línea negra continua representa el modelo de referencia según \cite{Choi2016}.
-El resto de líneas son modelos que incluyen rotación PMS con $\oomegac$ entre 0.0084 y 0.0336, respectivamente.
-La estrella púrpura y el cuadrado son abundancias superficiales de Li para el Sol actual \cite{Asplund2009} y la media para el cúmulo de las Pléyades \cite{Sestito2005} respectivamente.
-La línea vertical discontinua hace referencia a la ZAMS.}
-	\label{fig:li_var_vel_0g}
-\end{figure}
-
-La figura \ref{fig:li_var_vel_0g} muestra la evolución temporal de la abundancia superficial de Li para varios modelos $1\msun$ inicializados con diferentes velocidades de rotación.
-Las simulaciones tuvieron en cuenta los efectos de rotación y AML causados por los vientos estelares pero no los de MB.
-La estrella púrpura y el cuadrado son las abundancias superficiales de Li para el Sol actual \cite{Asplund2009} y para las Pléyades, respectivamente \cite{Sestito2005}.\par
-
-Obsérvese cómo la abundancia de Li en la superficie estelar disminuyó con el tiempo para todos los modelos simulados.
-La línea negra continua representa el modelo de referencia que adopta los parámetros de superación de la envoltura calibrados para el Sol, como se documenta en \cite{Choi2016}.
-Todos los modelos quemaron demasiado Li antes de la ZAMS y, por tanto, no coincidieron con la abundancia media de Li en superficie de las Pléyades.
-También fue destacable el hecho de que apenas hubo diferencias entre los distintos modelos en cuanto a la abundancia de Li durante gran parte de la ZAMS.
-Sólo después de un millón de años se destruyó el Li en un grado acentuado, ya que antes no se alcanzó la temperatura necesaria en BCZ.
-Después, los diferentes modelos destruyeron el Li de forma muy similar debido a dos razones principales.
-Por un lado, las zonas convectivas que desarrollaron los modelos tenían tamaños muy similares, por lo que la temperatura en la BCZ era prácticamente la misma.
-Por otro lado, el diferencial de rotación entre el núcleo y la zona convectiva no comenzó a desarrollarse hasta que alcanzó los $ \approx 10^6$ años, alcanzando su máxima diferencia en la ZAMS alrededor de los $ \approx 10^7$ años.
-Fue en este momento cuando debieron aumentar los efectos combinados de la turbulencia y la diferencia rotacional, dando lugar a notables diferencias en la evolución de A(Li) (véase la figura \ref{fig:li_var_vel_0g_z1}).
-Posteriormente, el modelo de referencia (línea negra) no agotó eficientemente el Li en la EM y no consiguió (de nuevo) igualar la abundancia actual de Li en la superficie solar.
-Los otros modelos que incluían la rotación durante el PMS con valores de $\oomegac$ entre 0.0084 y 0.0336 fueron capaces de quemar Li de forma más realista.
-Entre ellos, sólo uno (línea verde) se aproximaba a la abundancia actual de Li del Sol, pero su velocidad de rotación era mucho mayor (véase la figura \ref{fig:rot_vel_0g}) que los $2\,\kms$ del Sol \cite{Gill2012}.
-\par
-
-Durante gran parte del PMS la estrella giró como un cuerpo sólido (véase la Figura \ref{fig:rot_vel_0g}) y esto se debió a que la estrella tenía un interior totalmente convectivo.
-No fue hasta el final de la trayectoria de Hayashi cuando la estrella comenzó a desarrollar un núcleo radiativo.
-Fue en esta etapa cuando apareció una diferencia de velocidad angular entre los límites superior e inferior de las zonas radiativa y convectiva respectivamente.
-El grado de rotación diferencial estaba directamente influido por la $\Omega$ inicial.
-A medida que los modelos se inicializaban con una velocidad angular mayor, se acentuaba la diferencia de velocidad entre la BCZ y la superficie estelar; cuanto mayor era la velocidad inicial, mayor era el gradiente de velocidad entre los límites inferior y superior de la CZ.
-Como consecuencia, la fuerza de la turbulencia localizada en la BCZ aumentó, de modo que el Li pudo alcanzar regiones con temperaturas cercanas a $\tli$, donde finalmente se quemó y destruyó (véase la figura \ref{fig:li_var_vel_0g}).
-Otras investigaciones \cite{Bouvier2018, Baraffe2017} apuntan a una tendencia diametralmente opuesta a la aquí expuesta, es decir, a mayor velocidad de rotación, mayor abundancia de Li en la superficie de la estrella.
-\par
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/li_var_vel_0_0g_z1.pdf}
-	\caption {Similar a la Figura \ref{fig:li_var_vel_0g} pero ampliando la ZAMS.
-Los modelos con una mayor velocidad de rotación inicial alcanzan ya la ZAMS con una menor cantidad de medida de Li en la superficie estelar.}
-	\label{fig:li_var_vel_0g_z1}
-\end{figure}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\textwidth]{img/paper1/rot_vel_var_vel_0_0g.pdf}
-	\caption{Evolución de la velocidad angular en la superficie (línea continua) y en el límite inferior (línea discontinua) de la zona convectiva superior, en función del tiempo para varios modelos de 1 $\msun$.
-Los modelos incluyen rotación PMS con valores $\oomegac$ entre 0,0084 y 0,0336.
-La estrella púrpura es la velocidad angular superficial para el Sol actual \cite{Gill2012}.
-La línea vertical discontinua hace referencia a la ZAMS.}
-	\label{fig:rot_vel_0g}
-\end{figure}
-
-Otros efectos estructurales bien conocidos de la rotación son la disminución de la temperatura efectiva ($\teff$) y, en menor medida, de la luminosidad estelar ($L$).
-Ambos efectos pueden observarse gráficamente en el diagrama HR de la figura \ref{fig:hr_var_vel_0g}, que muestra una vista ampliada de las trayectorias evolutivas desde la ZAMS hasta la TAMS para varios modelos $\msun$ inicializados con diferentes velocidades de rotación.
-Si comparamos el modelo no rotatorio (línea sólida negra) con los rotatorios podemos reconocer que al final del PMS, estos últimos alcanzan la ZAMS con un $\teff$ menor que los primeros.
-Estos resultados coinciden con los de estudios anteriores \cite[véase por ejemplo ][]{Eggenberger2012,Piau2001,Pinsonneault1989}.\par
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/hr_var_vel_0_0g_z1.pdf}
-	\caption{Un ejemplo de cuadrícula solar de 1$\msun$ de trayectoria evolutiva estelar desde el PMS hasta el TAMS cubriendo un amplio rango de velocidades angulares.
-La rotación se activa en los modelos en el PMS y esos modelos llegan antes a la ZAMS y a un $\teff$ menor que el de no rotación (línea negra continua).
-La luminosidad se expresa en términos de $\lsun$.}
-	\label{fig:hr_var_vel_0g}
-\end{figure}
-
-\subsection{Evolución del Li con MB de intensidad fija}
-La figura \ref{fig:li_var_vel_4_0g} muestra la evolución temporal de la abundancia superficial de Li para varios modelos de 1 $\msun$.
-Estos modelos se inicializaron con diferentes velocidades rotacionales y tuvieron en cuenta los efectos del MB causado por un campo magnético de intensidad 4G.
-Si lo comparamos con la Figura \ref{fig:li_var_vel_0g} en la que se despreciaron los efectos del MB, observamos cómo se alteraron los perfiles de abundancia de Li durante el PMS y el MS.
-Durante el PMS podemos describir el efecto como modesto, algo esperado y en línea con el hecho de que el AML causado por MB (ver Ec.~\ref{eq:j_dot}) depende directamente de la tasa de pérdida de masa.
-Si tenemos en cuenta que para las estrellas de tipo solar los modelos predicen una tasa de pérdida de masa total modesta, ese valor es incluso mucho menor en esta fase.
-Por el contrario, durante la EM se observa que la AML es mucho más significativa, provocando que se destruya una menor cantidad de Li.\par
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/li_var_vel_4_0g.pdf}
-	\caption{Evolución de la abundancia superficial del \isotope[7]{Li} respecto al \isotope[1]{H}, en función del tiempo para varios modelos de 1 $\msun$.
-Los modelos incluyen un campo magnético con una intensidad de 4G y rotación PMS con $\oomegac$ entre 0,0084 y 0,0336, respectivamente.
-La estrella púrpura y el cuadrado son la abundancia superficial de Li para el Sol actual \cite{Asplund2009} y el cúmulo de las Pléyades \cite{Sestito2005} respectivamente.
-La línea vertical discontinua hace referencia a la ZAMS.}
-	\label{fig:li_var_vel_4_0g}
-\end{figure}
-
-El efecto de la rutina MB puede apreciarse aún más claramente en las Figuras~\ref{fig:rot_vel_4g}, \ref{fig:rot_vel_4g_z1} \& Apéndices \footnote{Los apéndices comprenden una serie de cuadrículas en función del tiempo y para varios modelos de 1 $\msun$ en los que se muestra, por un lado, la evolución de la abundancia superficial del \isotope[7]{Li} respecto al \isotope[1]{H} tanto para intensidades de campo magnético variables como para velocidades angulares y, por otro, la evolución de la velocidad de rotación superficial.}.
-En estas figuras representamos los perfiles de rotación para la superficie de las estrellas y para el fondo de la envoltura convectiva.
-Son modelos de 1 $\msun$ inicializados con diferentes velocidades de rotación y considerando la influencia del MB.
-De forma similar a los perfiles de evolución de $A(\isotope[7]{Li})$ comentados en el párrafo anterior, el efecto de la rutina se hizo visible una vez alcanzada la ZAMS.
-Si comparamos la evolución de las curvas aquí presentadas con las de la Figura \ref{fig:rot_vel_0g} vemos como la estrella, en lugar de seguir aumentando $\Omega$, comenzó a ralentizarse tras haber alcanzado su máximo en el paso por la ZAMS.
-Nótese que en este punto las velocidades angulares en la superficie de la estrella y en la ZAMS alcanzaron su máxima diferencia.
-Por otro lado, el efecto MB hizo que las velocidades angulares entre ambas zonas de la estrella disminuyeran hasta que, para una edad cercana a la del Sol (ver Figura \ref{fig:rot_vel_4g_z1}), la estrella prácticamente rotaba como un sólido rígido.
-Estos resultados también eran coherentes con los obtenidos por \cite{Eggenberger2010} en cuanto al efecto del campo magnético, en particular su influencia en la pérdida de momento angular, sobre la velocidad de rotación de la estrella.\par
-
-De forma similar, también observamos que los modelos con menor velocidad angular generalmente acababan mostrando valores más altos para la abundancia de Li en la superficie (ver Figuras~ \ref{fig:li_var_vel_4_0g}, \ref{fig:grid_li_var_vel} y \ref{fig:grid_li_var_g}).
-En ninguno de esos casos obtuvimos valores de Li en la superficie superiores a los mostrados por el modelo sin rotación.\par
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/rot_vel_var_vel_4_0g.pdf}
-	\caption{La evolución de la velocidad de rotación de la superficie, en función del tiempo para varios modelos de 1 $\msun$.
-Los modelos incluyen un campo magnético con una intensidad de 4\,G, rotación PMS con $\oomegac$ entre 0,0084 y 0,0336, respectivamente y MB.
-La estrella púrpura es la velocidad angular superficial para el Sol actual \cite{Gill2012}.
-La línea vertical discontinua hace referencia a la ZAMS.}
-	\label{fig:rot_vel_4g}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/rot_vel_var_vel_4_0g_z1.pdf}
-	\caption{Similar a la Figura \ref{fig:rot_vel_4g} pero ahora mostrando en detalle la velocidad de rotación de la superficie a medida que la estrella se aproxima a TAMS.}
-	\label{fig:rot_vel_4g_z1}
-\end{figure}
-
-El MB también dejó su huella en el diagrama HR afectando significativamente al $\teff$ de la estrella.
-Para visualizar este efecto tomamos como referencia la Figura \ref{hr_vc_0336_var_g_z1} en la que todos los modelos se iniciaron con el mismo valor $\oomegac=0,0336$ pero la intensidad del campo magnético simulado fue diferente.
-Obsérvese que los modelos con MB produjeron estrellas más calientes debido a su influencia.
-La menor velocidad con la que giraba la estrella debido al efecto MB provocó el aumento del $\teff$, siendo esta diferencia de prácticamente 95K entre los modelos simulados con 0.0G y 5.0G respectivamente para $log(\llsun)=0$.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/hr_vc_0336_var_g_z1.pdf}
-	\caption{Similar a la Figura \ref{fig:hr_var_vel_0g} pero ahora mostrando en detalle los efectos del frenado magnético en las trayectorias evolutivas para diferentes intensidades de campo magnético y $\oomegac=0.0336$.
-La presencia de un campo magnético produce estrellas más calientes debido a la influencia del frenado magnético en la velocidad de rotación de la estrella.}
-	\label{hr_vc_0336_var_g_z1}
-\end{figure}
-
-Como se describe en la sección \ref{sec_mesh}, la rutina MB distribuyó la cantidad total de AML calculada según la Ec.~\ref{eq:k_jdot} entre las distintas capas que componían la CZ.
-En la Figura \ref{fig:cz_vc_028_var_b} podemos observar la evolución de la CZ más externa normalizada con respecto al radio de la estrella para varios modelos de 1 $\msun$.
-Todos los modelos se inicializaron con $\oomegac=0.0336$ e intensidades de campo magnético que variaban entre 0.0G y 5.0G.
-De acuerdo con los modelos establecidos de evolución estelar, en una estrella de tipo solar la CZ cubre prácticamente toda ella durante gran parte del PMS.
-A medida que se aproxima a la ZAMS, la CZ va disminuyendo como consecuencia de la aparición de un núcleo radiativo y mantiene un radio aproximadamente constante hasta la etapa final de la MS.
-En este punto aumenta significativamente como respuesta a la expansión generalizada del radio de la estrella.
-En cuanto al efecto deL MB sobre el tamaño de la CZ, observamos que a medida que aumentaba la intensidad del campo magnético, el tamaño de la CZ disminuía (véase la figura \ref{fig:cz_vc_028_var_b_z1_new}).
-El núcleo radiativo se desplazó hacia el exterior para incluir una fracción cada vez mayor de la masa estelar, haciendo que la temperatura en la base de la CZ cayera por debajo de $\tli$.
-Este efecto fue más evidente durante la MS.
-La disminución del tamaño de la CZ estuvo en consonancia con el hecho de que se destruyera menos Li haciendo que menos material estelar alcanzara zonas con temperaturas superiores a $\tli$.\par
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/cz_vc_028_var_g.pdf}
-	\caption{Evolución del tamaño de la zona convectiva en función del tiempo para varios modelos de 1 $\msun$.
-Todos los modelos se inicializaron con $\oomegac=0.0336$ y las intensidades de campo magnético varían entre 0.0G y 5.0G.
-La línea vertical discontinua hace referencia a la ZAMS.}
-	\label{fig:cz_vc_028_var_b}
-\end{figure}
-
-En la Figura \ref{fig:mdot_vc_028_var_b} podemos ver la evolución de $\Dot{M}$ durante PMS y MS para varios modelos de 1 $\msun$.
-Todos los modelos se inicializaron con $\oomegac=0.0336$ e intensidades de campo magnético que variaban entre 0.0G y 5.0G.
-En las primeras etapas del PMS fue donde se concentró la mayor pérdida de masa, que fue disminuyendo a medida que se acercaba a la ZAMS.
-Como la estrella también disminuyó su radio durante la PMS, aumentó $\Omega$ obedeciendo al principio de conservación del AM.
-Al llegar a la ZAMS, el radio estelar se mantuvo más o menos estable durante gran parte de la MS (excepto en su etapa final, véase la figura \ref{fig:mdot_vc_028_var_b_z1}), pero siguió perdiendo masa mientras aumentaba $\Omega$, aunque de forma menos agresiva si lo comparamos con la PMS.
-Sin embargo, como consecuencia tanto de la aparición del núcleo radiativo durante el seguimiento de Henyey como de la existencia de una CZ (véase la figura \ref{fig:mb_act_var_vel_vc_028}), se activó la rutina MB, provocando que la velocidad angular de la estrella comenzara a disminuir a lo largo de toda la MS.
-Cuanto más intenso es el campo magnético, mayor es el efecto de frenado.\par
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/mdot_vc_028_var_g.pdf}
-	\caption{Evolución de la pérdida de masa $\Dot{M}$ en función del tiempo para varios modelos de 1 $\msun$.
-Los modelos incluyen diferentes intensidades de campo magnético entre 0.0G y 5.0G y rotación PMS con $\oomegac=0.028$.
-La línea vertical discontinua hace referencia a la ZAMS.}
-	\label{fig:mdot_vc_028_var_b}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/mdot_vc_028_var_g_z1.pdf}
-	\caption{La evolución de la pérdida de masa $\Dot{M}$ a medida que la estrella se acerca a TAMS, en función del tiempo para varios modelos de 1 $\msun$.
-Los modelos incluyen una intensidad de campo magnético variable entre 0.0G y 5.0G y rotación PMS con $\oomegac=0.028$.
-Cuanto más intenso sea el campo magnético, menor será $\Dot{M}$.}
-	\label{fig:mdot_vc_028_var_b_z1}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/mb_act_vc_028_var_g.pdf}
-	\caption{La activación de la rutina de frenado magnético en función de la presencia de un núcleo radiativo.
-Los modelos incluyen un campo magnético con una intensidad comprendida entre 0.0G y 5.0G y una rotación PMS con $\oomegac=0.0228$.
-Las líneas continuas señalan la activación (on) y desactivación (off) de la rutina de frenado magnético.
-Las líneas horizontales discontinuas informan sobre la naturaleza del núcleo de la estrella: radiativo (rad) o convectivo (conv).
-Por decisión de implementación, una vez activada la rutina, permanece activada incluso si la naturaleza del núcleo de la estrella cambia a convectiva.
-La línea vertical discontinua hace referencia a la ZAMS.}
-	\label{fig:mb_act_var_vel_vc_028}
-\end{figure}
-
-\subsection{Evolución alternativa del Li con MB}
-Hasta este punto, las simulaciones de los diferentes modelos se basaron en la parametrización recogida en la Tabla \ref{tab:phy_mesa}, que a su vez fue adoptada de \cite{Choi2016}.
-Si recordamos la evolución del Li mostrada en las Figuras \ref{fig:li_var_vel_0g} y \ref{fig:li_var_vel_4_0g}, destacamos que en ambos casos se quemaba demasiado Li antes de llegar a la ZAMS y, por tanto, no coincidía con la abundancia media de Li en superficie de las Pléyades.
-Por otro lado, utilizando una parametrización ligeramente distinta de la empleada hasta ahora en la que los parámetros de convección y \textit{overshooting} se han reajustado según la Tabla \ref{tab:phy_alt_mesa}, las simulaciones pudieron reproducir con mayor fidelidad tanto las abundancias de Li del cúmulo de las Pléyades como las del Sol (véase la Figura \ref{fig:li_3_0g_0314vc}).\par
-
-\begin{table}
-	\centering
-	\caption{Alternative MTL and overshooting parameters.}
-	\label{tab:phy_alt_mesa}
-	\begin{tabular}{ll} 
-		\hline
-		Parameter & Adopted prescriptions and values\\
-		\hline
-		Convection & MLT + Ledoux, $\alpha_\mathrm{MLT}$ = 1.70\\
-		Overshoot & time-dependent, diffusive, $f_\mathrm{ov,core}=0.016$, \\ & $f_\mathrm{ov,sh}=0.002$\\
-		\hline
-	\end{tabular}
-\end{table}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/li_3_0g_0314vc.pdf}
-	\caption{La evolución de la abundancia superficial del \isotope[7]{Li} respecto al \isotope[1]{H}, en función del tiempo para un modelo de 1 $\msun$.
-El modelo incluye un campo magnético con una intensidad de 3G y rotación PMS con $\oomegac= 0.0314$, respectivamente.
-Los parámetros MLT y overshooting se han fijado en $\alpha_\mathrm{MLT}=1.70$, $f_\mathrm{ov,core}=0.016$, $f_\mathrm{ov,sh}=0.002$.
-La estrella púrpura y el cuadrado son la abundancia superficial de Li para el Sol actual \cite{Asplund2009} y el cúmulo de las Pléyades \cite{Sestito2005} respectivamente.
-La línea vertical discontinua hace referencia a la ZAMS.}
-	\label{fig:li_3_0g_0314vc}
-\end{figure}
-
-Del mismo modo, la evolución de la velocidad angular para esta nueva configuración se muestra en la Figura \ref{fig:rot_vel_var_vel_mlt_3_0g}.
-Aquí todavía no reproducimos correctamente la velocidad angular del Sol.
-Se hizo evidente que la influencia de los parámetros libres, relativamente arbitrarios, asociados a MLT condicionaba significativamente la evolución de la abundancia de Li.
-Sin embargo, sin la inclusión de MB no fue posible ajustarlo para las Pléyades y el Sol con la misma trayectoria evolutiva.
-Este hecho abre la puerta a otras parametrizaciones compatibles con el Sol para reproducir las observaciones en cúmulos estelares jóvenes y para el Sol.\par
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{img/paper1/rot_vel_3_0g_0314vc.pdf}
-	\caption{La evolución de la velocidad de rotación de la superficie, en función del tiempo para un modelo de 1 $\msun$.
-El modelo incluye un campo magnético con una intensidad de 3G, rotación PMS con $\oomegac=0.0314$, parámetros en la Tabla \ref{tab:phy_alt_mesa} y MB.
-La estrella púrpura es la velocidad angular superficial para el Sol actual \cite{Gill2012}.
-La línea vertical discontinua hace referencia a la ZAMS.}
-	\label{fig:rot_vel_var_vel_mlt_3_0g}
-\end{figure}
-
-\section{Conclusiones} \label{sec_conclusions}
-Hemos demostrado, a través de los diferentes modelos estelares simulados, que los efectos inducidos por la combinación de ambos mecanismos de rotación y frenado magnético ofrecen una forma plausible de reconciliar los datos observacionales con los modelos teóricos.
-Esto último es de gran importancia tanto para el transporte de AM como para el transporte de elementos químicos.
-También somos muy conscientes de que aún estamos lejos de comprender los mecanismos físicos exactos que gobiernan estos procesos, por lo que es necesario seguir profundizando en estas áreas de estudio.
-En particular, el estudio de la evolución de los campos magnéticos durante la PMS y la MS, y su impacto en la AML.
-\par
-
-Se realizarán futuras y mejoradas implementaciones de nuestra rutina para estudiar los resultados sobre las huellas, la composición superficial del Li, la rotación, etc.
-Es probable que en un modelo estelar en rotación simulado con MB desde el principio, la rotación diferencial sea muy reducida y, por tanto, las abundancias de Li observadas en estrellas jóvenes podrían explicarse adecuadamente.
-Proponemos que este resultado podría alcanzarse mediante una ley de interdependencia entre $\Omega$ y $B$.
-Así, durante el PMS, cuando la estrella rota más rápido, el efecto MB es más eficiente.
-En cambio, durante el MS, cuando el arranque se ralentiza, el MB será menos intenso.
-Este enfoque establecería un mecanismo de autorregulación sobre la velocidad angular de la estrella que acabaría influyendo directamente en la evolución del Li.
-Es igualmente importante comprender mejor el papel general de la pérdida de masa en AML y sobre todo en estrellas más frías y de baja masa.
-Hoy en día es un reto determinar la velocidad terminal del viento estelar para este tipo de estrellas que juega un papel clave en la AML.\par
-
-Subrayamos también que nuestros modelos no coincidieron al mismo tiempo con la abundancia de Li solar observada y $\Omega$.
-Por lo tanto, no podemos asegurar que hayamos modelado correctamente la historia rotacional del Sol.
-En vista de estas deficiencias de nuestros modelos, debemos analizar los resultados obtenidos con cautela y no sacar conclusiones prematuras.
-También hemos mostrado cómo una parametrización MLT alternativa podría producir resultados acordes con las observaciones.
-En el transcurso de este trabajo se han apoyado las siguientes conclusiones:
-\begin{enumerate}
-	\item La inclusión del campo magnético conduce a modelos más fríos y a un menor agotamiento de Li en la EM.
-	\item Una combinación de rotación durante la PMS y efecto MB durante la MS produce un comportamiento diferente y potencialmente más prometedor que los producidos por los modelos estándar.
-Así, nuestro enfoque apunta a reproducir el A(Li) observado y la velocidad de rotación solar al mismo tiempo.
-	\item $\teff$ de las trayectorias evolutivas estándar representan límites superiores, ya que estos modelos no tienen en cuenta el efecto de frenado magnético ni la rotación.
-	\item La extensión de la zona convectiva disminuye cuando aumenta la intensidad del campo magnético.
-	\item El MB durante la PMS y/o el ajuste de los parámetros libres de exceso de MLT y $\amlt$ parece ser también necesario para explicar las abundancias de Li en cúmulos jóvenes.
-\end{enumerate}
+\subsection{Evolución del Li con MB de intensidad variable}
+\subsection{Evolución del Li con MB de intensidad y $\amlt$ variables}
+\section{Comparativa MB intensidad fija vs variable}
+\section{Conclusiones finales}
 

### 2024-01-29 project/overleaf/Tesis III/capitulos/capitulo08.tex
Add Tesis project

% !TeX root = ../libro.tex
% !TeX encoding = utf8

\chapter{Ciencia de datos}\label{ch:octavo-capitulo}
\section{Introducción}
La ciencia de datos es un campo interdisciplinario que utiliza algoritmos, procedimientos y procesos para examinar grandes cantidades de datos con el fin de descubrir patrones ocultos, generar conocimientos y dirigir la toma de decisiones. Utiliza técnicas y teorías extraídas de muchos campos dentro del contexto de las matemáticas, estadísticas, programación, análisis, IA y aprendizaje automático para descubrir conocimientos accionables a partir de conjuntos de datos.\par

En el contexto de la astrofísica, la ciencia de datos juega un papel crucial. Los astrofísicos y físicos de partículas utilizan instrumentación de última generación en telescopios y aceleradores para estudiar el Universo tanto en las escalas más grandes como en las más pequeñas. Los enormes conjuntos de datos a escala de petabytes que resultan de las diferentes observaciones y experimentos se explotan en busca de indicios, pistas que pueden arrojar luz sobre la naturaleza de la materia oscura, la energía oscura, la evolución de los agujeros negros y la física más allá del "modelo estándar" de la cosmología y la física de partículas.\par

El desafío es reformular el modelado de datos que los astrofísicos necesitan hacer de tal manera que se puedan aplicar arquitecturas de aprendizaje automático de última generación, sin comprometer el alto nivel de control de errores sistemáticos, ni la comprensión detallada de la incertidumbre estadística, que las preguntas de física que se están abordando con extrema precisión en el siglo XXI exigen. Por lo tanto, las simulaciones nos permiten probar qué teorías son consistentes con el Universo que podemos observar. Esta es una clara demostración del poder y la necesidad de la ciencia de datos para avanzar en nuestra comprensión del universo.\par

\section{Proceso de la ciencia de datos}
El proceso de ciencia de datos es un enfoque sistemático que implica varios pasos para extraer valiosos conocimientos de los datos. Comienza con la definición del problema, que implica entender los objetivos que se persiguen y formular preguntas que pueden ser respondidas con datos. Los siguientes pasos implican la recopilación y limpieza de los datos, seguidos por el análisis exploratorio de los mismos para entender los patrones y tendencias subyacentes. Esto es continuado por la construcción y evaluación del modelo para predecir resultados o descubrir estructuras dentro de los datos. Finalmente, el modelo se despliega, se contrastan los resultados con las observaciones, y los hallazgos se comunican a las partes interesadas.\par

Desde una perspectiva científica, el proceso de ciencia de datos es crucial ya que proporciona una metodología estructurada para manejar y analizar grandes volúmenes de datos. Permite a los científicos probar hipótesis, descubrir patrones, hacer predicciones e impulsar la toma de decisiones. Cada paso en el proceso sirve a un propósito específico y contribuye al objetivo general de extraer información significativa de los datos. Por ejemplo, la limpieza de datos asegura la calidad de los datos, el análisis exploratorio de datos ayuda a entender los datos, y la construcción del modelo permite hacer predicciones e inferencias.\par

Además, la naturaleza iterativa del proceso de ciencia de datos se alinea con el método científico de investigación, que implica formar hipótesis, realizar experimentos y refinar teorías basadas en los resultados. Este proceso iterativo nos permite un aprendizaje y mejora continua. Además, el énfasis en la comunicación en el proceso de ciencia de datos asegura que los hallazgos puedan ser compartidos e implementados de manera efectiva, contribuyendo al avance del conocimiento en varios campos. Por lo tanto, el proceso de ciencia de datos no sólo es un aspecto fundamental de la toma de decisiones basada en datos, sino también un contribuyente significativo al progreso científico.\par

Dentro del proceso de datos podemos enumerar una serie de etapas encaminadas a establecer una sistemática a la hora de su ejecución. El proceso no debe contemplarse como un marco rígido que no admite alteraciones al mismos, sino como que debe ser utilizado como una herramienta que debe ser adaptada a nuestras necesidades. Dependiendo de los objetivos establecidos, de la calidad de los datos que están a nuestra disposición, algunas de las siguientes etapas se harán más necesarias que otras, más intensas en lo que a la asignación de recursos se refiere. Entre los diferentes podemos enumerar los siguientes:

\begin{enumerate}
	\item \textit{Definición del problema} - Este es el primer paso donde definimos el problema que estás tratando de resolver. Implica entender los objetivos de la investigación, para a continuación formular preguntas que pueden ser respondidas con datos e identificar las fuentes de datos necesarias.
	\item \textit{Recopilación de datos} - En este paso, recopilamos los datos necesarios para nuestro análisis. Esto podría implicar la consulta de sitios web, el acceso a bases de datos, generación de simulaciones.
	\item \textit{Limpieza de datos} - Una vez que tenemos los datos recopilados, es hora de limpiarlos. Esto implica manejar valores faltantes, eliminar duplicados, corregir errores y lidiar con valores atípicos.
	\item \textit{Análisis exploratorio de datos} - Este es un paso crucial donde exploramos y visualizamos los datos para entender los patrones subyacentes, las tendencias y los valores atípicos. Es una paso fundamental porque nos ayuda a generar conocimiento y formar hipótesis para un análisis más profundo.
	\item \textit{Ingeniería de características} - Esto implica crear nuevas características a partir de las existentes para representar mejor los patrones subyacentes en los datos. Es un paso esencial para mejorar el rendimiento de los modelos de aprendizaje automático. 
	\item \textit{Construcción del modelo} - Aquí, elegimos un modelo apropiado, lo entrenamos con los datos que previamente hemos obtenido para luego evaluar su rendimiento. Normalmente esto implica dividir los datos en conjuntos de entrenamiento y prueba, seleccionar un algoritmo correcto y fijar sus parámetros.
	\item \textit{Evaluación del modelo} - En este paso, evaluamos el rendimiento del utilizando métricas apropiadas. Es importante usar una métrica que se alinee con los objetivos que hemos prefijados en el paso inicial.
	\item \textit{Despliegue del modelo} - Una vez que estamos satisfecho con el rendimiento del modelo, es hora de desplegarlo. Esto podría implicar integrar el modelo en un sistema de producción, establecer un sistema de monitoreo y desarrollar un plan para el mantenimiento del mismo.
	\item \textit{Comunicación} - Finalmente, comunicamos nuestros hallazgos a las partes interesadas. Esto podría implicar la creación de artículos científicos, informes, o presentaciones que expliquen claramente tu metodología, hallazgos e implicaciones de negocio de nuestro trabajo.
\end{enumerate}

Como hemos adelantado, los pasos anteriores no de dejan de ser un marco de trabajo genérico que debe de ser adaptado a nuestras necesidades específicas. En nuestro caso en particular, no hemos recurrido al uso de herramientas de aprendizaje automático, por lo que la correspondiente etapa no ha formado parte de nuestro proceso de ciencia de datos. Del mismo modo, nuestro modelo no ha sido incorporado, integrado con ningún sistema. De momento este no ha sido el caso pero no sería una insensatez pensar que pudiese pasar a incorporarse como una extensión a la herramienta que hemos utilizado para nuestras simulaciones, si lo autores de la misma lo considerasen oportuno.

\section{Herramientas}
\subsection{Modules for Experiments in Stellar Astrophysics - (MESA)}
Para la realización de este tesis doctoral nos hemos apoyado en la herramienta de evolución estelar \textit{Modules for Experiments in Stellar Astrophysics} (MESA). MESA ha jugado un papel fundamental en nuestra investigación y por ello, debido a su importancia, le hemos dedicado un capítulo en exclusiva en el que entramos a fondo en sus características, estructura, módulos que lo componen, y muy especialmente en cómo extender la física que incorpora sus modelos mediante la programación de nuestras propias rutinas.\par

\subsection{Octave} \label{sec:tool_octave}
GNU Octave es un lenguaje de programación de alto nivel destinado principalmente a cálculos numéricos. Es un conjunto de software libre y de código abierto que proporciona una interfaz de línea de comandos conveniente para resolver problemas lineales y no lineales numéricamente. La sintaxis de Octave es en gran medida compatible con MATLAB, lo que lo convierte en una alternativa popular para los usuarios que necesitan una solución rentable.

Octave proporciona capacidades para la solución numérica de problemas lineales y no lineales, y para realizar otros experimentos numéricos. También proporciona amplias capacidades gráficas para la visualización y manipulación de datos. Octave puede ser extendido por paquetes, permitiendo a los usuarios añadir más funcionalidades según sea necesario. Puede ser ejecutado en modo GUI, como una consola, o invocado como parte de un script de shell. Octave está disponible para varios sistemas operativos, incluyendo GNU/Linux, macOS, BSD y Microsoft Windows.

\subsection{Tool for OPerations on Catalogues And Tables (TOPCAT)} \label{sec:tool_topcat}
TOPCAT, acrónimo de \textit{Tool for OPerations on Catalogues And Tables}, es un visor y editor gráfico altamente versátil e interactivo para datos tabulares. Está diseñado para trabajar con datos astronómicos, pero puede ser utilizado para cualquier tipo de datos en formato tabular. Proporciona un entorno dinámico donde los datos pueden ser cargados, visualizados, editados y analizados. TOPCAT soporta varios formatos de datos y puede gestionar tanto archivos locales como remotos, lo que lo convierte en una herramienta flexible para la manipulación de datos.\par

En cuanto a la extracción de datos, TOPCAT permite a los usuarios ver y editar datos de celdas a través de un navegador, y tiene visores para imágenes y espectros. También se incluye un visor de trazado del cielo para datos astronómicos. Además, proporciona funciones estadísticas para analizar los datos, y capacidades de coincidencia cruzada para comparar diferentes catálogos. La herramienta también soporta \textit{scripting} para tareas automatizadas. Estas características hacen de TOPCAT una herramienta integral para la extracción y análisis de datos en el campo de la astronomía y más allá.\par

Aunque TOPCAT fue diseñado principalmente para trabajar con datos astronómicos, también puede ser utilizado para cualquier tipo de datos tabulares. Esto significa que puede ser útil en una variedad de campos que requieren la manipulación y análisis de datos tabulares.  Por ejemplo, en la investigación científica, los datos a menudo se presentan en forma tabular y requieren análisis detallados. TOPCAT puede ser útil aquí para visualizar y analizar estos datos. Además, en el campo de la informática, los datos tabulares son comunes, por ejemplo, en bases de datos o archivos CSV. Por lo tanto, aunque TOPCAT tiene un enfoque en la astronomía, sus capacidades lo hacen aplicable en una variedad de campos donde se manejan datos tabulares. Sin embargo, los detalles específicos pueden depender del tipo de datos y de las necesidades del usuario.\par

TOPCAT es particularmente útil para el \textit{Virual Observatory} (VO). Ésta es una aplicación de escritorio para el análisis interactivo de datos en formato tabular, especialmente catálogos de fuentes. TOPCAT puede acceder a servicios de datos externos. También puede comunicarse con otras herramientas astronómicas a través del protocolo \textit{Simple Application Messaging Protocol} (SAMP). Esto permite una integración perfecta de TOPCAT en el entorno del VO, permitiendo a los usuarios recuperar y analizar datos de varios servicios.\par

Cuando se trata de extraer datos de diferentes catálogos astronómicos, TOPCAT proporciona una gama de funcionalidades, como las de leer y escribir tablas en varios formatos como FITS, VOTable, CSV. Esto hace posible trabajar con una amplia variedad de catálogos astronómicos. Por ejemplo, puedes usar TOPCAT para acceder a un catálogo de estrellas contenidas en la base de datos Gaia. Además, TOPCAT ofrece características para el cruce de correspondencias, que es crucial al comparar diferentes catálogos. También admite \textit{scripting} para tareas automatizadas, lo cual puede ser particularmente útil al trabajar con catálogos grandes o realizar tareas repetitivas.\par

En resumen, TOPCAT es una herramienta integral que proporciona capacidades robustas para trabajar con el VO y extraer datos de varios catálogos astronómicos. Está diseñado para facilitar la manipulación de tablas, para que los usuarios puedan concentrarse en hacer ciencia.\par

\section{Catálogos de datos}
Una etapa fundamental de toda investigación es la de la confrontación de los datos arrojados con los modelos contra los disponibles en la literatura ya existente,

\subsection{Gaia}
El catálogo Gaia es un mapa tridimensional completo de la Vía Láctea, creado utilizando datos del telescopio espacial Gaia. Éste contiene datos de alrededor de 1.46 mil millones de estrellas, entre los que se encuentra sus posiciones en el cielo, paralaje y movimiento propio. Además, incluye sus magnitudes en diferentes bandas espectrales para una gran cantidad de ellas. El catálogo Gaia es particularmente útil para estudiar estrellas similares al Sol, ya que proporciona un censo único de estrellas dentro de 100 pc de nuestro Sol. Esto lo convierte en un recurso invaluable para los astrónomos y astrofísicos que buscan comprender mejor las propiedades y la evolución de estas estrellas.\par

\subsection{Gaia-ESO Spectroscopic Survey (GES)}
GES es un catálogo público enfocado al análisis espectroscópico de un número importante de estrellas recogidas en el catálogo Gaia. Se llevó a cabo con el instrumento FLAMES en el VLT. Su objetivo es proporcionar una visión homogénea de las distribuciones de cinemática y abundancias elementales en la Vía Láctea. Dentro de la iniciativa GES se han observado más de 100 000 estrellas, proporcionando velocidades radiales y rotacionales proyectadas, parámetros estelares tales como: temperatura efectiva, gravedad superficial y metalicidad, abundancias de varios elementos, entre los que se encuentra el Litio, y parámetros específicos para rastrear la acreción y actividad en estrellas jóvenes. El catálog GES es único en lo que se refiere a la observación de estrellas de todos los tipos espectrales con análisis dedicados y especializados, lo que la hace particularmente útil para estudiar estrellas similares al Sol.\par

\subsection{Combinando Gaia y GES}

La combinación de los datos de los catálogos Gaia y GES puede proporcionar una visión completa de las propiedades y evolución de las estrellas similares al Sol. El catálogo Gaia proporciona datos de posición precisos y paralajes, mientras que la GES ofrece información espectroscópica detallada. Esta combinación permite una comprensión más completa de estas estrellas, incluyendo su cinemática, composiciones químicas y potencial para albergar planetas. Al proporcionar una visión más completa y detallada de estas estrellas, podemos obtener una comprensión más profunda de las propiedades y evolución de las estrellas similares al Sol, lo que representa una fuente de información que se antoja invaluable para la investigación en astrofísica. Estamos ante un conjunto de herramientas poderoso para cualquier astrofísico que estudie estrellas similares al Sol.\par

En el caso de las abundancias de Litio, y como ya hemos apuntado, éste es un elemento clave en el estudio de la astrofísica estelar, ya que su abundancia puede proporcionar información valiosa sobre la edad y la historia de las estrellas. Sin embargo, la determinación precisa de las abundancias de Litio puede ser un desafío debido a la complejidad de las líneas espectrales del Litio. GES nos asiste en nuestra investigación aportando su análisis especializado y dedicado de estrellas de todos los tipos espectrales, convirtiéndose así en una herramienta fundamental a la hora de afrontar este desafío.\par

Por otra parte, tenemos información precisa sobre la velocidad angular de las estrellas. Gaia es capaz de medir la velocidad angular de las estrellas con una precisión sin precedentes, y ésta nos aporta información es crucial para estudiar y entender la rotación de las estrellas, su evolución y estructura interna. Combinado esta información con las abundancia de Litio podemos establecer hipótesis y validar las mismas acerca del impacto de la velocidad angular sobre el Litio. Paralelamente, la velocidad angular juega un papel crucial en el análisis de la presencia e intensidad de un campo magnético en las estrellas. Los campos magnéticos son responsables de la pérdida de momento angular en las estrellas jóvenes y son la principal fuente de energía detrás de una amplia gama de fenómenos dinámicos (llamaradas, emisión de rayos X, manchas estelares) que ocurren en las capas superficiales del Sol y otras estrellas. Por lo tanto, la medición precisa de la velocidad angular de una estrella puede proporcionar información valiosa sobre la presencia e intensidad de su campo magnético.\par

Otra fuente de información valiosa para nuestra investigación procedente de ambos catálogos es la referida a los cúmulos abiertos. Los cúmulos abiertos son grupos de estrellas que se formaron a partir de la misma nube de gas y polvo. Por lo tanto, todas las estrellas de un cúmulo abierto tienen aproximadamente la misma edad y composición química inicial. Esto los convierte en laboratorios naturales para estudiar la evolución de las estrellas a través de sus diferentes etapas. Gaia ha observado una gran cantidad de cúmulos abiertos, proporcionando datos detallados sobre sus miembros. En paralelo, GES también ha observado un gran número de ellos en un rango amplio de edades. La combinación de ambas fuentes permite estudiar la estructura y dinámica de este tipo de cúmulos, y usarlos para restringir y mejorar los modelos de evolución estelar, algo que hemos puesto en práctica en nuestra investigación.\par

Como acabamos de adelantar, la investigación de los campos magnéticos en estrellas similares al Sol. y en particular de aquéllas que se encuentran en cúmulos abiertos, se beneficia enormemente de los datos proporcionados por ambos catálogos. Los campos magnéticos juegan un papel importante en todas las etapas de la evolución estelar. En estrellas similares al Sol, se generan en las capas convectivas exteriores. Estudiar los campos magnéticos a gran escala de estas estrellas puede mejorar nuestra comprensión sobre los mismos y proporcionar restricciones observacionales para los modelos que estudian cómo se generan, la topología que presentan y sus intensidades. Los datos de los catálogos nos asisten a la hora de identificar y estudiar las estrellas con alta actividad magnética, lo que puede proporcionar pistas sobre la dinámica interna de las estrellas y su evolución. Combinando esta información con la velocidad angular, nos puede proporcionar con indicios de cómo se interrelacionan ambos fenómenos. Adicionalmente, cruzando estos datos con la información de los cúmulos abiertos, obtenemos una visión más clara de cómo los campos magnéticos varían entre las estrellas de la misma edad y composición química.\par

La abundancia de Litio puede proporcionar información valiosa sobre la edad y la historia de las estrellas. En particular, puede ayudar a rastrear la mezcla convectiva en el interior de las estrellas, que está influenciada por la rotación estelar. La rotación, a su vez, puede ser modulada por la presencia de campos magnéticos a través del frenado magnético.

Además, en estrellas jóvenes y de baja masa, la presencia de fuertes campos magnéticos puede inhibir la convección, alterando el transporte de energía desde el interior de la estrella. Esto puede afectar la tasa de quema de Litio, lo que resulta en una mayor abundancia de Litio en la superficie de la estrella de lo que se esperaría de lo contrario.

Por lo tanto, al estudiar las abundancias de Litio en las estrellas, los astrónomos pueden obtener pistas sobre la intensidad de los campos magnéticos en el interior de las estrellas. Sin embargo, la interpretación de las abundancias de

Los campos magnéticos en las estrellas pueden tener un impacto significativo en la abundancia de Litio presente en ellas. El Litio se quema fácilmente a una temperatura relativamente baja (2.5 x 10^6 K) en el interior estelar³. Por lo tanto, es un diagnóstico extraordinariamente sensible de la estructura y evolución estelar³.

La abundancia de Litio puede proporcionar información valiosa sobre la edad y la historia de las estrellas. En particular, puede ayudar a rastrear la mezcla convectiva en el interior de las estrellas, que está influenciada por la rotación estelar. La rotación, a su vez, puede ser modulada por la presencia de campos magnéticos a través del frenado magnético.

Además, en estrellas jóvenes y de baja masa, la presencia de fuertes campos magnéticos puede inhibir la convección, alterando el transporte de energía desde el interior de la estrella. Esto puede afectar la tasa de quema de Litio, lo que resulta en una mayor abundancia de Litio en la superficie de la estrella de lo que se esperaría de lo contrario¹.

Por lo tanto, al estudiar las abundancias de Litio en las estrellas, los astrónomos pueden obtener pistas sobre la intensidad de los campos magnéticos en el interior de las estrellas. Sin embargo, la interpretación de las abundancias de Litio puede ser compleja, ya que también pueden ser influenciadas por otros factores, como la metalicidad de la estrella y la eficiencia de la convección¹.

En resumen, aunque el Litio no proporciona una medida directa de los campos magnéticos estelares, su abundancia en las estrellas puede ofrecer valiosas pistas sobre la actividad magnética en el interior estelar. Esto hace que las mediciones de las abundancias de Litio, como las proporcionadas por la Encuesta Espectroscópica Gaia-ESO (GES), sean una herramienta importante para la investigación de los campos magnéticos en las estrellas.

Source: Conversation with Bing, 1/25/2024
(1) [2006.01899] Study of the effects of magnetic braking on the lithium .... https://arxiv.org/abs/2006.01899.
(2) Study of the effects of magnetic braking on the lithium abundances of .... https://academic.oup.com/mnras/article/496/2/1343/5851759.
(3) Enhanced lithium depletion in Sun-like stars with orbiting planets - Nature. https://www.nature.com/articles/nature08483.pdf.
(4) undefined. https://doi.org/10.1093/mnras/staa1569.

Source: Conversation with Bing, 1/25/2024
(1) Effect of magnetic fields on Lithium depletion of solar-type stars. https://www.cambridge.org/core/services/aop-cambridge-core/content/view/054CD22CB5342846A1EC5552F7DED8FE/S1743921313002512a.pdf/effect-of-magnetic-fields-on-lithium-depletion-of-solar-type-stars.pdf.

En resumen, cada uno de estos enfoques ofrece una perspectiva única y valiosa sobre la naturaleza y la evolución de estrellas. La combinación de los catálogos Gaia y GES proporciona una visión completa y detallada de las estrellas similares al Sol. La información sobre las abundancias de Litio, la velocidad angular y los datos de los cúmulos abiertos, en particular, puede arrojar luz sobre la cinemática, la composición química, la edad y la historia de estas estrellas, así como ser una herramienta muy potente para ofrecer una perspectiva única y valiosa sobre la naturaleza y la evolución de los campos magnéticos estelares.  En resumen, una fuente de datos invaluable para cualquier astrofísico que estudie estrellas similares al Sol.\par

Source: Conversation with Bing, 1/25/2024
The Gaia-ESO Survey: Calibrating the lithium–age relation with open .... https://arxiv.org/pdf/2009.00610.pdf.
Interferometric Observations of Magnetic Fields in Forming Stars. https://www.frontiersin.org/articles/10.3389/fspas.2019.00003/full.
ESO Survey: 3D dynamics of young groups and clusters from GES and Gaia .... https://arxiv.org/pdf/2311.08358.pdf.
Stellar magnetic fields and surface structures - Uppsala University. https://www.physics.uu.se/research/astronomy-and-space-physics/research/stars/magnetism/.
The Role of Magnetic Fields in Protostellar Outflows and Star Formation. https://www.frontiersin.org/articles/10.3389/fspas.2019.00054/full.
The Gaia-ESO Survey: Membership probabilities for stars in 32 open .... https://arxiv.org/pdf/2006.09423v1.pdf.
undefined. https://doi.org/10.3389/fspas.2019.00054.
Source: Conversation with Bing, 1/25/2024
[1310.5562] Magnetic fields of Sun-like stars - arXiv.org. https://arxiv.org/abs/1310.5562.
undefined. https://doi.org/10.1017/S1743921314002026.
Membership of stars in open clusters using random forest with gaia data .... https://link.springer.com/article/10.1140/epjs/s11734-021-00205-x.
Evidence of New Magnetic Transitions in Sun-like Stars from Gaia Data. https://www.spacedaily.com/reports/Evidence of New Magnetic Transitions in Sun like Stars from Gaia Data 999.html.
undefined. https://doi.org/10.48550/arXiv.1310.5562.

\endinput
%--------------------------------------------------------------------
% FIN DEL CAPÍTULO. 
%--------------------------------------------------------------------

### 2024-01-29 project/overleaf/Tesis III/preliminares/declaracion-originalidad.tex
Add Tesis project

% !TeX root = ../libro.tex
% !TeX encoding = utf8
%
%*******************************************************
% Declaración de originalidad
%*******************************************************

\thispagestyle{empty}

\hfill\vfill

\textsc{Declaración de originalidad}\\\bigskip

D./Dña. \miNombre \\\medskip

Declaro explícitamente que el trabajo presentado como Tesis Doctoral, correspondiente al curso académico \miCurso, es original, entendido esto en el sentido de que no he utilizado para la elaboración del trabajo fuentes sin citarlas debidamente.
\medskip

En Granada a \today 
\vspace{3cm}
\begin{center} 
Fdo: \miNombre

\end{center}

\vfill

\cleardoublepage
\endinput

### 2024-01-29 project/overleaf/Tesis III/preliminares/introduccion.tex
Add Tesis project

% !TeX root = ../libro.tex
% !TeX encoding = utf8
%
%*******************************************************
% Introducción
%*******************************************************

% \manualmark
% \markboth{\textsc{Introducción}}{\textsc{Introducción}}

\chapter{Introducción}

La investigación de las aparentes anomalías en la abundancia superficial de litio (Li) observadas en el Sol y en cúmulos globulares estelares jóvenes dentro de contextos astrofísicos contemporáneos es muy prometedora para avanzar en nuestra comprensión de los mecanismos que influyen en el agotamiento del Li a lo largo de la evolución estelar. Este estudio profundiza en la intrincada interacción entre la mezcla rotacional y los efectos hidrostáticos rotacionales en estrellas de pre-secuencia principal (PMS) y de secuencia principal (MS) empleando simulaciones de modelos en rotación.\par

La evolución del Li para modelos solares se examina bajo la influencia combinada de la Teoría de la Longitud de Mezcla (MLT) y el frenado magnético (MB), donde la intensidad del campo magnético ($B$) y la parametrización MLT ($\amlt$) juegan un papel clave.

-- ESTO HABRÁ QUE COLOCARLO EN SU SITIO CUANDO HABLEMOS DE LOS MODELOS CON B Y ALFA VARIABLE --

La evolución del Li para modelos solares se examina bajo la influencia combinada de la Teoría de la Longitud de Mezcla (MLT) y el frenado magnético (MB), donde la intensidad del campo magnético ($B$) y la parametrización MLT ($\amlt$) evolucionan dinámicamente con parámetros solares clave, como son la velocidad angular ($\Omega$), la temperatura efectiva ($\teff$) o la densidad ($\rho$). Se propone un enfoque novedoso, que evita valores fijos para estos parámetros y produce resultados coherentes con los datos observacionales.\par

Se generan modelos solares precisos, que reflejan la naturaleza dinámica de $B$ y $\amlt$ a lo largo de la evolución estelar, y se contrastan con datos observacionales de cúmulos abiertos jóvenes obtenidos a través del Gaia-ESO Survey (GES). La inclusión de valores variables de $B$ y $\amlt$ reproduce congruentemente resultados en línea con trabajos anteriores en los que estas aproximaciones se han abordado por separado. Damos un paso más al incorporarlos conjuntamente en nuestros modelos y estudiamos el efecto combinado que producen sobre la historia rotacional y las abundancias superficiales de Li en modelos solares, obteniendo resultados que siguen en línea con esos trabajos, y compatibles con los datos observacionales.\par

Los resultados sugieren un límite inferior robusto de 1.133 dex para las abundancias superficiales de Li en estrellas similares al Sol, alineándose con las observaciones solares y arrojando luz sobre el intrincado equilibrio de los procesos físicos que gobiernan el contenido de Li en las atmósferas estelares. Asimismo, obtenemos valores teóricos de $\amlt$ acordes con el intervalo [1.76, 1.78] obtenido en trabajos anteriores. A la vista de estos prometedores resultados, nuestros modelos ofrecen una alternativa consistente y completa a aquellos con valores fijos, y con un tratamiento aislado de $B$ y $\amlt$. Hemos conseguido obtener simultáneamente resultados, que aunque no coinciden exactamente con las medidas reales del Sol, son sin embargo compatibles con su abundancia superficial de Li, velocidad angular y valores $\amlt$ predichos.\par

\endinput

### 2024-01-29 project/overleaf/Tesis III/preliminares/summary.tex
Add Tesis project

% !TeX root = ../libro.tex
% !TeX encoding = utf8
%
%*******************************************************
% Summary
%*******************************************************

\selectlanguage{english}
\chapter{Summary}

Investigating the apparent anomalies in lithium (Li) surface abundance observed in the Sun and young stellar globular clusters within contemporary astrophysical contexts holds significant promise for advancing our comprehension of the mechanisms influencing Li depletion throughout stellar evolution. This study delves into the intricate interplay between rotational mixing and rotational hydrostatic effects in pre-main sequence (PMS) and main sequence (MS) stars by employing simulated grids of rotating models. The Li evolution of solar models is scrutinized under the combined influence of Mixing Length Theory (MLT) and magnetic braking (MB), where the magnetic field strength ($B$) and MLT parameterization ($\amlt$) dynamically evolve with key solar parameters. A novel approach, avoiding fixed values for these parameters, is proposed, yielding results consistent with observational data.\par

Accurate solar models, reflective of the dynamic nature of $B$ and $\amlt$ throughout stellar evolution, are generated and tested against observational data from young open clusters obtained through the Gaia-ESO Survey (GES). The inclusion of variable $B$ and $\amlt$ values congruently reproduces results in line with previous work in which these approaches have been addressed separately. We go a step further by incorporating them jointly in our models and study the combined effect they produce on the rotational history and surface Li abundances in solar models, obtaining results that are still in line with those works, and compatible with observational data.\par

The findings suggest a robust lower limit of 1.133 dex for surface Li abundances in Sun-like stars, aligning with solar observations and shedding light on the intricate balance of physical processes governing Li content in stellar atmospheres. Likewise, we obtain theoretical values of $\amlt$ in accordance with the [1.76, 1.78] interval  obtained in previous works. In view of these promising results, our models offer a consistent and comprehensive alternative to those with fixed values, and with an isolated treatment of $B$ and $\amlt$. We have managed to simultaneously obtain results, which although are not exactly matching the Sun's actual measurements, are however compatible with its surface Li abundance, angular velocity and predicted $\amlt$ values.\par

% Al finalizar el resumen en inglés, volvemos a seleccionar el idioma español para el documento
\selectlanguage{spanish} 
\endinput

### 2024-01-30 project/overleaf/Tesis III/apendices/apendice02.tex
WIP

% !TeX root = ../libro.tex
% !TeX encoding = utf8
\chapter{Agradecimientos}\label{ap:agradecimientos}
Deseamos agradecer el generoso trabajo y la ayuda ofrecida por las comunidades MESA y TOPCAT. Los autores agradecen el apoyo financiero de los fondos públicos españoles (incluidos los fondos FEDER) para la investigación en el marco de los proyectos ESP2017-87676-C5-2-R y ESP2017-87676-C5-5-R. AGH también agradece el apoyo financiero del Fondo Europeo de Desarrollo Regional/Junta de Andalucía-Consejería de Economía y Conocimiento en el marco del proyecto E-FQM-041-UGR18 de la Universidad de Granada. JCS también agradece el apoyo del proyecto RYC-2012-09913 bajo el programa 'Ram\'on y Cajal' del Ministerio de Ciencia y Educación de España.
\endinput
%------------------------------------------------------------------------------------
% FIN DEL APÉNDICE. 
%------------------------------------------------------------------------------------

### 2024-01-30 project/overleaf/Tesis III/apendices/apendice03.tex
WIP

% !TeX root = ../libro.tex
% !TeX encoding = utf8
\chapter{Software}\label{ap:software}
Las simulaciones utilizadas en este trabajo se han ejecutado con la versión 10398 de MESA \footnote{\url{http://mesa.sourceforge.net/release/2018/03/21/r10398.html}}. Las diferentes figuras se han generado haciendo uso de GNU Octave 5.1.0. \footnote{\url{https://www.gnu.org/software/octave/}}. Los conjuntos de datos de Gaia y GES se han analizado utilizando TOPCAT 4.8-7 \footnote{\url{https://www.g-vo.org/topcat/topcat/}}.

\endinput
%------------------------------------------------------------------------------------
% FIN DEL APÉNDICE. 
%------------------------------------------------------------------------------------

### 2024-01-30 project/overleaf/Tesis III/capitulos/capitulo07.tex
WIP

@@
 \section{Conclusiones finales}
+\subsection{Caballero et al. (2020) Comparison}
+En \cite{Caballero2020} ya abordamos el estudio de la abundancia superficial de litio en el Sol.
+Aquel trabajo previo sentó las bases para el desarrollo del presente.
+En él avanzábamos como futuras líneas de investigación un conocimiento más profundo sobre cómo los campos magnéticos están ligados a la rotación y cómo evolucionan en el tiempo.
+También mostramos cómo ajustando el valor de $\amlt$ podíamos influir en las abundancias de Li.
+Por tanto, una comparación entre los resultados obtenidos anteriormente y los de este estudio pone en perspectiva las diferencias y similitudes entre ambos estudios, así como los avances realizados desde entonces.
+La Tabla \ref{tab:caballero2023_2020} resume las principales diferencias entre ambos trabajos, en línea con lo ya avanzado, es decir, incorporamos formalismos que nos permiten evolucionar temporalmente el $B$ y $\amlt$.\par
+
+\begin{table}
+	\centering
+	\begin{threeparttable}
+		\begin{tabular}{lll} 
+			\hline
+			Parameter & This work & \cite{Caballero2020}\\
+			\hline
+			Convection & MLT variable $\alpha_{\textrm{MLT}}$ & MLT fixed $\alpha_{\textrm{MLT}}=1.82$\\
+			Overshoot & deactivated & $f_{\textrm{ov,core}}=0.0160$,\\ & & $f_{\textrm{ov,sh}}=0.0174$\\
+			Magnetic Braking & \cite{Gallet2013} & \cite{Ud-Doula2008} \\ Formalism & & \\
+			Magnetic Field & B(G) variable & B(G) pre-set [3.0G-5.0G]\\ Strength & & \\
+			Number of OC's & 64 & 1 \\
+			Source of data & GDR3.0 GES DR5.0 & \cite{Sestito2005} \\
+			\hline
+		\end{tabular}
+	\end{threeparttable}
+	\caption{Comparison with \cite{Caballero2020} work.}
+	\label{tab:caballero2023_2020}
+\end{table}
+
+La Figura \ref{fig:li_var_vel_4_0g4} muestra la evolución de la abundancia de Li para $\omegaini$ entre 0.0084 y 0.0336 en presencia de un campo magnético con una intensidad de 4G.
+Hemos aumentado la versión original con el mismo conjunto de datos OCs recogidos aplicados en este nuevo trabajo, de forma que sea posible una comparación consistente entre ambos formalismos.
+Si centramos nuestra atención en la simulación con $\omegaini$ = 0.0336, la simulación reproduce A(Li) = 1.024 dex, valor que está en línea tanto con el del Sol (1.1 $\pm$ 0.1 dex) como con los valores obtenidos en la Figura \ref{fig:li_var_vel_var_g_3} del presente trabajo (1.133 dex) para $\omegaini$ = 0.14 y 0.1425.
+Aunque en este último caso los modelos se inicializaron con valores de $\omegaini$ más altos que en el primero, lo que resulta en $B$ hasta 3 órdenes de magnitud más intensivos ($\approx$ 1KG frente a 4G) la simulación en ambos trabajos acaba produciendo valores muy similares tanto para la velocidad rotacional como para el A(Li).
+Esto se explica en parte porque una mayor intensidad de campo magnético implica un efecto MB más intenso (véanse las figuras \ref{fig:rot_vel_var_vel_var_g3} y \ref{fig:rot_vel_var_g3}).
+\& \ref{fig:rot_vel_var_vel_4_0g4}).
+En el primer caso, el efecto de desaceleración es más agudo que en el segundo, por lo que deja de destruir Li antes.\par
+
+Sin embargo, este efecto no es suficiente para contrarrestar eficazmente la destrucción de Li.
+El otro elemento decisivo que entra en juego es el valor de $\amlt$.
+Como se ejemplifica en \cite{Caballero2020} su valor tiene un impacto inversamente proporcional en la destrucción de Li, donde un valor más bajo resultó en una menor destrucción de Li, particularmente durante el PMS.
+En la figura \ref{fig:alpha_mlt_var_vel_g3} observamos la evolución temporal del valor $\amlt$.
+Hasta la última fase de la PMS, el valor $\amlt$ es sustancialmente inferior a 1,82, el valor preestablecido e invariante utilizado en nuestro trabajo anterior, lo que se traduce en una menor destrucción de Li y permite al modelo alcanzar la ZAMS con una reserva de Li $\approx$ 29\% dex mayor.
+Esta mayor reserva de Li compensa la mayor destrucción inducida por los modelos inicializados con $\omegaini$ mayor velocidad angular.\par
+
+Del mismo modo, procedemos a mostrar la evolución de la velocidad de rotación superficial en la Figura \ref{fig:rot_vel_var_vel_4_0g4} y seleccionamos la simulación con $\omegaini$ = 0.0336.
+De nuevo, los valores similares se explican por la mayor eficiencia del MB en el nuevo modelo.
+Aunque las simulaciones alcanzan la ZAMS con valores de velocidad de rotación de hasta el 200\% para los modelos inicializados con $\omegaini$ más altos, el campo magnético más intenso produce una desaceleración más pronunciada para el mismo intervalo de tiempo.
+Este hecho es especialmente notable en el periodo de tiempo comprendido entre la ZAMS y 1 Ga.
+La velocidad de rotación obtenida en el ecuador de 4.82 km/s, es un valor bastante próximo a 4.72 km/s, es decir, el valor simulado para $\omegaini$ = 0.0336 en la Figura \ref{fig:rot_vel_var_vel_var_g3}.\par
+
+A la vista de los valores obtenidos, podemos concluir que el nuevo modelo introducido en el presente trabajo no sólo es congruente con los resultados de los trabajos en los que se basa, sino también con los de \cite{Caballero2020}.
+Adicionalmente, esta nueva versión expone la ventaja de no tener que recurrir al uso de unos valores prefijados para $B$ y $\amlt$, y sin recurrir a activar el overshooting.
+En definitiva, esto representa una mejora significativa y apoya nuestra línea de investigación orientada a evitar valores preestablecidos haciéndolos evolucionar dinámicamente con parámetros solares clave.\par
+
+\begin{figure}
+    \centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/li_var_vel_4_0g4.pdf}
+	\caption{La evolución de la abundancia superficial del \isotope[7]{Li} en relación con el \isotope[1]{H}, en función del tiempo para varios modelos de 1 $\msun$.
+Los modelos incluyen un campo magnético con una intensidad de 4G.
+La línea negra continua representa el modelo de referencia según \cite{Choi2016}.
+El resto de líneas son modelos que incluyen rotación inicial con $\omegaini$ entre 0.0084 y 0.0336, respectivamente.
+La estrella morada son las abundancias superficiales de Li para el Sol actual \cite{Asplund2009}.
+La línea vertical discontinua hace referencia a la ZAMS.}
+	\label{fig:li_var_vel_4_0g4}
+\end{figure}
+
+\begin{figure}
+    \centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/rot_vel_var_vel_4_0g4.pdf}
+	\caption{La evolución de la velocidad de rotación de la superficie, en función del tiempo para varios modelos de 1 $\msun$.
+Los modelos incluyen un campo magnético con una intensidad de 4G, rotación inicial con $\omegaini$ entre 0.0084 y 0.0336, respectivamente y MB.
+La estrella púrpura es la velocidad angular superficial para el Sol actual \cite{Gill2012}.
+La línea vertical discontinua hace referencia a la ZAMS.}
+	\label{fig:rot_vel_var_vel_4_0g4}
+\end{figure}
+
+\subsection{Gossage et al. (2021) Comparison}
+Debido a la existencia de similitudes entre el trabajo de \cite{Gossage2021} y éste, procedemos a realizar una comparación entre ambos.
+Entre las similitudes encontramos el hecho de que ambos trabajos se centran en el frenado magnético, haciendo uso de MESA para las simulaciones.
+Entre las principales diferencias se encuentra que nuestros modelos incorporan una variable $B$ y $\amlt$.
+Finalmente, ambos comparan sus resultados de simulación (periodos de rotación y abundancias de Li) con arranques localizados en OCs.\par
+
+En \cite{Gossage2021}, los autores también abordan el problema de AML desde un punto de partida similar al nuestro, pero con algunas variaciones notables.
+En la Tabla \ref{tab:gassage_vs_navarro} comparamos las similitudes y diferencias entre ambos estudios.
+En ambos trabajos los modelos consideran la rotación no sólo en el MS sino también en el PMS, con la distinción de que en el trabajo de Gossage tienen en cuenta el bloqueo del disco.
+Este aspecto no es tan relevante en nuestras simulaciones ya que nos hemos centrado en los perfiles de A(Li) y el trabajo de Gossage se centra principalmente en modelar los perfiles de rotación de las estrellas, que se ven afectados de forma significativa por el mecanismo de bloqueo del disco.
+Del mismo modo, también se consideran los efectos del frenado magnético en la pérdida de momento angular, en nuestro caso para estrellas con 1$\msun$ (gemelas-solares) y en el suyo con 0.1$\msun$$le$$\mstar$$\le$1.3$\msun$.
+Ambos modelos utilizan también MLT, pero en nuestro caso el valor $\amlt$ no está prefijado, sino que varía a lo largo de la simulación.\par
+
+Estos dos trabajos se complementan ya que se basan en formalismos diferentes.
+Por un lado, nuestro trabajo se basa en la propuesta de \cite{Gallet2013} cuyos parámetros más relevantes son $\rossby$ y la presencia de un campo magnético de intensidad variable (B), mientras que Gossage presenta dos formalismos: \cite{Matt2015} y \cite{Garraffo2018}.
+El primero se caracteriza principalmente por $\rossby$ y la $\mstar$, mientras que el segundo por $\rossby$ y un parámetro que da cuenta de la complejidad de los campos magnéticos, siendo igual a 1 para un campo dipolar y mayor para multipolos de orden superior.
+En cuanto a las observaciones de OCs, en este trabajo hemos tenido acceso a un mayor número de ellas, y hemos cubierto un rango más amplio de edades [0.001-6.761] Gyr.
+Por último, ambos trabajos simulan sus respectivos modelos en MESA pero utilizando versiones diferentes.\par
+
+\cite{Gossage2021} consiguen reproducir con mayor precisión el perfil de rotación del Sol a su edad actual \cite[ver][panel (a) de las Figuras 2, 4 \& 6]{Gossage2021} aunque como describen los autores del trabajo, calibraron sus modelos para ello, incluyendo un parámetro adicional libre de difusión ($v_0$).
+En nuestro trabajo, nuestro modelo no incluye este tipo de parametrización adicional y sigue siendo capaz de reproducir una velocidad de rotación ecuatorial de 4.72 km/s, un valor no especialmente alejado del valor nominal establecido de 2.0 km/s.
+En lo que respecta a A(Li), su modelo no consigue reproducir el valor para estrellas similares al Sol, incluido el propio Sol \cite[ver][panel (d) en las Figuras 2 \& 4, y (b) en la Figura 6]{Gossage2021}.
+Recalibrando algunos de los parámetros de su modelo y/o incorporando mecanismos adicionales que contribuyan a la destrucción del Li, los autores esperan obtener valores en línea con los datos observacionales.
+En nuestro caso (véase la sección \ref{sec_conclusiones}) pudimos reproducir A(Li) con mayor precisión.
+
+\begin{table}
+	\centering
+	\begin{threeparttable}
+		\begin{tabular}{lll} 
+			\hline
+			Parameter & This work & \cite{Gossage2021}\\
+			\hline
+
+			Rotation & At PMS \& MS & At PMS \& MS\\
+			Solar Mass & $\mstar$=1$\msun$ & 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$ \\
+			Convection & MLT variable $\alpha_{\textrm{MLT}}$ & MLT fixed $\alpha_{\textrm{MLT}}$\\
+			Magnetic Braking & \cite{Gallet2013} & \cite{Matt2015}, \\ Formalism & & \cite{Garraffo2018}\\
+			Magnetic Field & B(G) variable & B(G) pre-set\\ Strength & & \\
+			Disk locking & Not present & Present\\
+			Number of OC's & 64 & 11 \\
+			Source of data & GDR3.0 GES DR5.0 & Several works \\
+			MESA version & r10398 & r11701\\
+			\hline
+		\end{tabular}
+	\end{threeparttable}
+	\caption{Comparison with \cite{Gossage2021} work.}
+	\label{tab:gassage_vs_navarro}    
+\end{table}
+
+\section{Conclusions and future works} \label{sec_conclusions}
+Hemos demostrado, a través de los diferentes modelos estelares simulados, que los efectos inducidos por la combinación de ambos mecanismos de rotación y frenado magnético ofrecen una forma plausible de reconciliar los datos observacionales con los modelos teóricos.
+Proponemos que este resultado podría alcanzarse mediante una ley de interdependencia entre $\Omega$ y $B$.
+A diferencia de nuestro trabajo anterior, estos nuevos modelos eliminan las restricciones de preestablecer valores iniciales y fijos a lo largo de su evolución temporal para la intensidad del campo magnético y el parámetro $\amlt$.
+Hemos contrastado los resultados de los modelos con las medidas obtenidas por las misiones Gaia y GES para estrellas de características similares a nuestro Sol, centrándonos en sus abundancias de Li.\par
+
+Nuestros modelos parecen sugerir un límite inferior a las abundancias superficiales de Li de 1.133 dex para estrellas similares al Sol, valor que se encuentra dentro del margen de error obtenido para el Sol (1.1 $\pm$ 0.1 dex).
+Por otro lado, predicen una velocidad de rotación en el ecuador (4.72 km/s) superior a la del Sol (2,0 km/s), por lo que podrían quemar más Li y reproducir finalmente los reservorios de Li existentes en el Sol.
+Junto con esta mayor velocidad angular, se desarrollaron fuertes campos magnéticos que desencadenaron un intenso frenado magnético que ralentizó su velocidad angular, pero no lo suficiente como para igualar el valor $\Omega$ actual del Sol.
+Además, hay que señalar que nuestros modelos no reproducen la intensidad media del campo magnético del Sol.\par
+
+Aunque hemos mostrado cómo tanto una intensidad de campo magnético en evolución como una parametrización MLT podrían producir resultados acordes con las observaciones, debemos analizar los resultados obtenidos con cautela y no extraer conclusiones prematuras.
+A pesar de estas deficiencias, podemos extraer la siguiente conclusión del presente trabajo:
+
+\begin{enumerate}
+	\item Un campo magnético con intensidad variable es una alternativa válida a los modelos con un valor prefijado, ya que producen resultados en línea con trabajos anteriores para la historia rotacional de estrellas similares al Sol.
+	\item Un formalismo $\amlt$ adaptativo variable podría eliminar la necesidad de prefijar su valor.
+Utilizándolo, pudimos reproducir valores en línea con trabajos anteriores.
+Además, también podría servir para derivar un intervalo de valores potenciales en los casos en que un valor por adelantado $\amlt$ es desconocido o incluso no deseable.
+	\item Nuestros modelos parecen sugerir un límite inferior para A(Li)=1.133 dex a las observaciones proporcionadas por Gaia y GES para estrellas gemelas solares.   
+	\item Aunque los modelos en el presente trabajo se inicializan a valores $\omegaini$ más altos que en nuestro trabajo anterior, 0.1425 y 0.0336 respectivamente, lo que resulta en $B$ hasta 3 órdenes de magnitud más intensos, $\approx$ 1KG frente a 4G, la simulación en ambos trabajos acaba produciendo valores muy similares tanto para la velocidad rotacional como para el A(Li).
+En el primer caso, esto se explica porque una mayor intensidad de campo magnético implica un efecto MB intensivo.
+Para el segundo, los valores $\amlt$ más bajos producidos por el formalismo adaptativo preservan un mayor depósito de Li durante el PMS.
+\end{enumerate}
+
+Nos gustaría concluir con las siguientes reflexiones para los próximos pasos.
+Es imperativo aumentar nuestra comprensión de los interiores estelares y los mecanismos de agotamiento del Li para la muestra de gemelos solares con el fin de realizar comparaciones más precisas.
+Además, ampliar el rango de masas en nuestros modelos nos permitirá confrontar los resultados con un rango más amplio de observaciones.
+Es un hecho que nuestros modelos siguen destruyendo demasiado Li al comparar los valores simulados con los de Gaia y GES.\par
+
+Tomando como punto de partida que las estrellas de rotación más rápida destruyen más Li, son necesarios otros mecanismos, además del MB, para explicar los datos observacionales.
+Entre ellos, el bloqueo al disco protoestelar podría contribuir a contrarrestar la aceleración de la estrella durante el PMS.
+De este modo, actuarían dos mecanismos de frenado a diferentes edades de la estrella.
+El bloqueo del disco actuaría durante la PMS y el MB principalmente durante la MS.
+Con este enfoque se establecería un mecanismo de autorregulación sobre la velocidad angular de la estrella que acabaría influyendo directamente tanto en la evolución del Li como en la intensidad del campo magnético.
+Para esta última se esperan intensidades menores que finalmente podrían converger con la de nuestro Sol.\par
+
+Por otro lado, una menor velocidad de rotación también implica una menor destrucción de Li para nuestros modelos.
+Una forma de compensar este menor Li remanente es recurrir al over-shooting (ver \cite{Caballero2020}) un efecto que no ha sido simulado en nuestros modelos.
+Activando este mecanismo, permitiríamos que el Li fuera alcanzado por burbujas de material lo suficientemente calientes como para destruirlo, contribuyendo así a disminuir su abundancia.
+Lo ideal, como hemos hecho con $\amlt$, sería no introducir el overshooting como un parámetro fijo, sino dependiente de otros parámetros estelares fundamentales.\par
 

### 2024-01-30 project/overleaf/Tesis III/capitulos/capitulo08.tex
WIP

@@
 
+\begin{lstlisting}[language=SQL, caption={Consulta TOPCAT sobre el catálogo GES DR5 para obtener los componentes de los cúmulos abiertos para los cuales se ha determinado una pertenecia mínima del 0.95\%.
+Las columnas obtenidas informan sobre la identificador de la estrella, $\teff$, $\gsurf$, $\feh$, A(Li) y sus errores asociados, y el identificador del cúmulo al que pertenece.}, label={lst:consulta}]
+	SELECT object,teff,e_teff,logg,e_logg,feh,e_feh,li1,e_li1,ges_fld
+	FROM ges_dr5 
+	WHERE ges_fld IN (
+		SELECT DISTINCT ges_fld
+		FROM ges_dr5 
+		WHERE ges_type 
+		LIKE 'GE_CL%' 
+			OR ges_type LIKE 'GE_SD_OC%' 
+			OR ges_type LIKE 'AR_CL%' 
+			OR ges_type LIKE 'AR_SD_OC' )
+	AND mem3d >= 0.95
+	AND li1 IS NOT NULL
+	AND feh IS NOT NULL
+	AND ges_fld IS NOT NULL
+	AND teff IS NOT NULL
+	AND logg IS NOT NULL
+	AND nn_teff IS NOT NULL
+	AND nn_logg IS NOT NULL
+	AND nn_feh IS NOT NULL
+	ORDER BY ges_fld
+\end{lstlisting}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/tesis/tap_query.pdf}
+	\caption{PDF}
+	\label{fig:tap_query}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/tesis/tap_ges_service.pdf}
+	\caption{PDF}
+	\label{fig:tap_ges_service}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/tesis/tap_multi_sources.pdf}
+	\caption{PDF}
+	\label{fig:tap_multi_services}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/tesis/tap_schema.pdf}
+	\caption{PDF}
+	\label{fig:tap_schema}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/tesis/tap_schema_detail.pdf}
+	\caption{PDF}
+	\label{fig:tap_schema_detail}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/tesis/tap_service.pdf}
+	\caption{PDF}
+	\label{fig:tap_service}
+\end{figure}
+
 \section{Catálogos de datos}
@@
 
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/tesis/open_cluster_sample.pdf}
+	\caption{Tabla obtenida de \cite{Bragaglia2022}.
+Las columnas 1 a 7 listan la denominación del cúmulo, las coordenadas de ascensión recta y declinación, la edad estimada del cúmulo, la distancia en parsecs (módulo de distancia convertido), la posición Z en coordenadas cartesianas galácticas, y la distancia desde el centro galáctico suponiendo que el Sol está a 8340pc.
+Las columnas 8 y 9 enumeran el número de espectros y objetivos, mientras que la media de [Fe/H] y la desviación estándar se dan en la columna 10.}
+	\label{fig:open_cluster_sample}
+\end{figure}
+
 La abundancia de Litio puede proporcionar información valiosa sobre la edad y la historia de las estrellas.
@@
 Los campos magnéticos en las estrellas pueden tener un impacto significativo en la abundancia de Litio presente en ellas.
-El Litio se quema fácilmente a una temperatura relativamente baja (2.5 x 10^6 K) en el interior estelar³.
+El Litio se quema fácilmente a una temperatura relativamente baja (2.5 x $10^6$ K) en el interior estelar³.
 Por lo tanto, es un diagnóstico extraordinariamente sensible de la estructura y evolución estelar³.

### 2024-01-31 project/overleaf/Tesis III/capitulos/capitulo08.tex
WIP Tesis

@@
 De momento este no ha sido el caso pero no sería una insensatez pensar que pudiese pasar a incorporarse como una extensión a la herramienta que hemos utilizado para nuestras simulaciones, si lo autores de la misma lo considerasen oportuno.
+
+\section{Catálogos de datos}
+Una etapa fundamental de toda investigación es la de la confrontación de los datos arrojados con los modelos contra los disponibles en la literatura ya existente,
+
+\subsection{Gaia}
+El catálogo Gaia es un mapa tridimensional completo de la Vía Láctea, creado utilizando datos del telescopio espacial Gaia.
+Éste contiene datos de alrededor de 1.46 mil millones de estrellas, entre los que se encuentra sus posiciones en el cielo, paralaje y movimiento propio.
+Además, incluye sus magnitudes en diferentes bandas espectrales para una gran cantidad de ellas.
+El catálogo Gaia es particularmente útil para estudiar estrellas similares al Sol, ya que proporciona un censo único de estrellas dentro de 100 pc de nuestro Sol.
+Esto lo convierte en un recurso invaluable para los astrónomos y astrofísicos que buscan comprender mejor las propiedades y la evolución de estas estrellas.\par
+
+\subsection{Gaia-ESO Spectroscopic Survey (GES)}
+El Gaia-ESO Survey (GES) \cite{Gilmore2012,Randich2013,Randich2022} es un catálogo público enfocado al análisis espectroscópico de un número importante de estrellas recogidas en el catálogo Gaia.
+Se llevó a cabo con el instrumento Espectrógrafo Multielemento de Fibra de Gran Conjunto (FLAMES) y el Espectrógrafo Echelle UV-Visual (UVES) en el VLT.
+Su objetivo es proporcionar una visión homogénea de las distribuciones de cinemática y abundancias elementales en la Vía Láctea, en el campo y en cúmulos abiertos (OC), hasta magnitud 19.\par
+
+Dentro de la iniciativa GES se han observado más de 110 000 estrellas, proporcionando velocidades radiales y rotacionales proyectadas, parámetros estelares tales como: temperatura efectiva, gravedad superficial y metalicidad, abundancias de varios elementos, entre los que se encuentra el Litio, para aproximadamente 1/3 de la muestra, y parámetros específicos para rastrear la acreción y actividad en estrellas jóvenes.
+El catálogo GES es único en lo que se refiere a la observación de estrellas de todos los tipos espectrales con análisis dedicados y especializados, lo que la hace particularmente útil para estudiar estrellas similares al Sol.\par
 
@@
 
-\section{Catálogos de datos}
-Una etapa fundamental de toda investigación es la de la confrontación de los datos arrojados con los modelos contra los disponibles en la literatura ya existente,
-
-\subsection{Gaia}
-El catálogo Gaia es un mapa tridimensional completo de la Vía Láctea, creado utilizando datos del telescopio espacial Gaia.
-Éste contiene datos de alrededor de 1.46 mil millones de estrellas, entre los que se encuentra sus posiciones en el cielo, paralaje y movimiento propio.
-Además, incluye sus magnitudes en diferentes bandas espectrales para una gran cantidad de ellas.
-El catálogo Gaia es particularmente útil para estudiar estrellas similares al Sol, ya que proporciona un censo único de estrellas dentro de 100 pc de nuestro Sol.
-Esto lo convierte en un recurso invaluable para los astrónomos y astrofísicos que buscan comprender mejor las propiedades y la evolución de estas estrellas.\par
-
-\subsection{Gaia-ESO Spectroscopic Survey (GES)}
-GES es un catálogo público enfocado al análisis espectroscópico de un número importante de estrellas recogidas en el catálogo Gaia.
-Se llevó a cabo con el instrumento FLAMES en el VLT.
-Su objetivo es proporcionar una visión homogénea de las distribuciones de cinemática y abundancias elementales en la Vía Láctea.
-Dentro de la iniciativa GES se han observado más de 100 000 estrellas, proporcionando velocidades radiales y rotacionales proyectadas, parámetros estelares tales como: temperatura efectiva, gravedad superficial y metalicidad, abundancias de varios elementos, entre los que se encuentra el Litio, y parámetros específicos para rastrear la acreción y actividad en estrellas jóvenes.
-El catálog GES es único en lo que se refiere a la observación de estrellas de todos los tipos espectrales con análisis dedicados y especializados, lo que la hace particularmente útil para estudiar estrellas similares al Sol.\par
-
+\section{Selección de datos}
 \subsection{Combinando Gaia y GES}
@@
 Adicionalmente, cruzando estos datos con la información de los cúmulos abiertos, obtenemos una visión más clara de cómo los campos magnéticos varían entre las estrellas de la misma edad y composición química.\par
+
+Como indica su nombre, GES pretende complementar los datos sobre paralajes y movimientos propios del satélite Gaia \cite{Mignard2005} con información extremadamente precisa sobre velocidades radiales (RVs), Li y química en general.
+La primera publicación de datos de Gaia (GDR1) \cite{Brown2016} data de 2016 y contiene información sobre los primeros 14 meses de funcionamiento.
+La segunda publicación de datos de Gaia (GDR2), \cite{Brown2018}, publicada en 2018, contenía posiciones, paralajes y movimientos propios de unos 1.300 millones de fuentes, junto con fotometría.
+La tercera publicación de datos de Gaia, la Gaia early data release 3 (GEDR3) \cite{Brown2021} y la Gaia data release 3 (GDR3) \cite{Brown2022}, transmiten medidas actualizadas y más precisas.
+GES las mejora y en su última versión (DR5.0) incluye todos los parámetros astrofísicos derivados por el consorcio Gaia-ESO \cite{Gilmore2022}.
+De interés para nuestra línea de investigación son:
+\begin{itemize}
+	\item Velocidades radiales y proyectadas
+	\item Parámetros estelares (temperatura efectiva, gravedad superficial y metalicidad)
+	\item Abundancias de varios elementos, entre ellos el Li
+	\item Parámetros específicos para rastrear la acreción y la actividad en estrellas jóvenes
+	\item Probabilidad de pertenencia a cúmulos
+\end{itemize}
+
+\subsection{Selección de cúmulos abiertos}
+GES DR5.0 proporciona información sobre 114 324 estrellas correspondientes a OCs, véase \cite{Gilmore2022} para más detalles. Seleccionamos aquellos registros que cumplen alguna de las condiciones:
+
+\begin{itemize}
+	\item GE\_CL: observado por GES, campo del programa OC
+	\item GE\_SD\_OC: observado por GES, campo estándar OC
+	\item AR\_CL: Observación del Archivo ESO, campo de programa OC
+	\item AR\_SD\_OC: Observación de archivo, campo estándar OC
+\end{itemize}
+
+Como resultado de este primer filtrado, el número de registros seleccionados se reduce a 43.299.
+Sobre este subconjunto de datos realizamos otro filtrado, esta vez dirigido a seleccionar aquellos que ofrecen información no nula sobre los siguientes atributos:
+\begin{itemize}
+	\item Metalicidad
+	\item Abundancia de Li
+	\item Identificador de campo GES
+	\item Temperatura efectiva
+	\item Gravedad superficial
+	\item Probabilidad de pertenencia a un cúmulo (>= 0.95\%)
+\end{itemize}
+
+Este segundo filtrado reduce la población de estrellas a considerar a un total de 5.895 registros que se distribuyen entre los 64 OCs listados en la Tabla \ref{tab:oc_full_list}.\par
+
+Como datos de referencia tomamos un subconjunto de las observaciones obtenidas por GES.
+Este subconjunto consiste en aquellas estrellas que pertenecen a OCs, y entre ellas se encuentran potenciales gemelas "ocultas" de nuestro Sol.
+Para identificar estos últimos, se realiza una selección adicional, esta vez utilizando los valores dados por nuestros modelos durante las simulaciones para la metalicidad ($\feh$), la temperatura efectiva ($\teff$) y la gravedad superficial ($\gsurf$).
+Por último, nos quedan las estrellas para las que las observaciones del GES han medido valores congruentes con los resultados de nuestros modelos, a partir de los cuales obtenemos sus abundancias de Li.\par
+
+\subsection{La selección de los gemelos solares más cercanos}
+Una vez preseleccionadas las estrellas pertenecientes a los OC, nuestro interés se centra en identificar las más parecidas a nuestro Sol, es decir, las gemelas solares.
+Esto nos lleva a preguntarnos qué parámetros estelares debemos tener en cuenta para clasificar una estrella como gemela solar.
+Encontrarlos nos ayudaría a averiguar el origen de nuestro Sol y a obtener información sobre las condiciones que prevalecían en los OC en los que nacieron.\par
+
+Ya se han hecho varios intentos de encontrar hermanos y gemelos solares.
+Los primeros se refieren a estrellas que se formaron en el mismo cúmulo que el Sol y, por tanto, deberían tener edades y abundancias químicas muy similares a las del Sol \cite{Adibekyan2018}[ver][y referencias en él].
+Por otro lado, según \cite{Strobel1996} se consideran gemelos solares las estrellas que poseen parámetros físicos fundamentales como $\teff$, $\gsurf$, $\feh$, propiedades fotométricas, composición química, edad, luminosidad, rotación y campos magnéticos son similares, si no idénticos, a los del Sol.
+Localizar estos gemelos solares supone un reto formidable debido a los estrictos criterios y a su dispersión por toda la Vía Láctea.\par
+
+No es raro que, los métodos de búsqueda aplicados se orienten a encontrar las candidatas basándose en sus propiedades cinemáticas, y luego se comparan sus temperaturas, metalicidades y abundancias químicas con las de nuestro Sol.
+En nuestro estudio, ampliamos los criterios de selección añadiendo la gravedad superficial y la edad de la estrella.
+Estos dos nuevos parámetros representan un ajuste fino del proceso de búsqueda de candidatos a gemelos.
+Utilizamos la edad de los OC indicada en la tabla \ref{tab:oc_reduced_list} y los datos obtenidos a partir del conjunto de simulaciones MESA iniciadas con diferentes $\omegaini$, donde $\omegaini = \oomegac$, $\Omega$ es la velocidad angular de la estrella en la superficie estelar, y $\omegac$ es la velocidad superficial en el ecuador de una estrella en rotación donde la fuerza centrífuga equilibra la gravedad newtoniana.
+La edad derivada de estas simulaciones proporciona un parámetro crucial para nuestro análisis.
+Esta información nos permite determinar los demás parámetros de simulación, que posteriormente utilizaremos para compararlos con los valores asociados a las estrellas pertenecientes a los OC preseleccionados.\par
+
+\begin{table}
+	\centering
+	\begin{tabular}{ll} 
+		\hline
+		Parámetro estelar & Intervalo de selección\\
+		\hline
+		OC Age & $\pm0.1\%$ \, for age < 1 Ga \\
+		& $\pm0.15\%$ \, for age $\geq$ 1 Ga \\
+		$\teff$ & $\pm 50.0 \, \Kelvin$\\
+		$\gsurf$ & $\pm 0.05 \, \dex$\\
+		$\feh$ & $\pm 0.05 \, \dex$\\
+		\hline
+	\end{tabular}
+	\caption{Parámetros e intervalos de selección aplicados durante el proceso de triaje sobre los componentes del OC.}
+	\label{tab:sel_params}
+\end{table}
+
+De las aproximadamente 5 900 estrellas identificadas, seleccionamos aquellas cuyos valores de $\feh$, $\teff$ y $\gsurf$ se encuentran en los intervalos correspondientes definidos en la Tabla \ref{tab:sel_params} y según el siguiente proceso de filtrado: a partir de los valores dados por nuestras simulaciones para los parámetros listados en la Tabla \ref{tab:sel_params}, procedemos a seleccionar de los distintos OC aquellas estrellas que tienen valores dentro de los intervalos establecidos.
+En concreto, para la edad estimada de un determinado cúmulo, establecemos un intervalo de $\pm$0,1 para OCs jóvenes (edad < 1 Ga) y $\pm$0.15 para contrapartes más viejas (edad >= 1 Ga) \cite{Cantat-Gaudin2020}[refer to][for comprehensive reference values]{Cantat-Gaudin2020}.
+Este intervalo dependiente de la edad sirve como base para seleccionar los pasos de tiempo de simulación que se alinean con los límites de edad inferior y superior de cada grupo.
+Una vez encontrados estos dos pasos temporales, se reúnen los valores límite asociados para $\teff$, $\gsurf$ y $\feh$.
+Con estos rangos de valores procedemos entonces a seleccionar en cada uno de los OC aquellas estrellas que tienen valores para estos mismos parámetros dentro de ellos.
+De esta forma obtenemos finalmente las estrellas candidatas a ser comparadas con los resultados de la simulación.\par
+
+\begin{table}
+	\centering
+	\begin{tabular}{l l l l l l l} 
+		\hline
+		ObjectId & $\teff$(K) & $\gsurf$(dex) & FeH(dex) & ALi(dex) & eALi(dex) & Age(Ga)\\
+		\hline
+		8510914-1157003 & 5843 & 4.42 & -0.03 & 1.8 & 0.08 & 3.981\\ 
+		8510991-1146169 & 5797 & 4.4 & -0.01 & 1.65 & 0.08 & 3.981\\ 
+		8512177-1144050 & 5818 & 4.46 & 0 & 1.99 & 0.06 & 3.981\\ 
+		8513941-1200571 & 5792 & 4.41 & -0.02 & 1.49 & 0.13 & 3.981\\ 
+		8515559-1148381 & 5794 & 4.43 & -0.04 & 1.56 & 0.1 & 3.981\\ 
+		8520350-1147480 & 5875 & 4.47 & -0.03 & 1.21 & 0.24 & 3.981\\ 
+		\hline
+	\end{tabular}
+	\caption{M67 - Configuración de filtro: Identificador estelar GES, $\teff$(K) [5787.4892, 5887.6823], $\gsurf$(dex) [4.3979, 4.4977], $\feh$(dex) [-0.05, 0.05], Edad(Ga) [3.977, 3.985]}
+	\label{tab:oc_m67}
+\end{table}
+
+\begin{table}
+	\centering
+	\begin{tabular}{l l l l l l l} 
+		\hline
+		ObjectId & $\teff$(K) & $\gsurf$(dex) & FeH(dex) & ALi(dex) & eALi(dex) & Age(Ga)\\
+		\hline
+		7571111-6048156 & 5084 & 4.54 & 0 & 2.29 & 0.08 & 0.24\\ 
+		7595031-6044149 & 5108 & 4.52 & -0.05 & 1.92 & 0.08 & 0.24\\ 
+		7595280-6032498 & 5093 & 4.49 & -0.03 & 2.29 & 0.08 & 0.24\\ 
+		\hline
+	\end{tabular}
+	\caption{NGC2516 - Configuración de filtro: Identificador estelar GES, $\teff$(K) [5010.0324, 5111.0095], $\gsurf$(dex) [4.4405, 4.5405], $\feh$(dex) [-0.05, 0.05], Age(Ga) [0.23964, 0.24036]}
+	\label{tab:oc_ngc2516}
+\end{table}
+
+Las tablas \ref{tab:oc_m67} y \ref{tab:oc_ngc2516} resumen el subconjunto de componentes de los OC M67 y NGC2516 seleccionados durante las simulaciones para modelos inicializados con $\omegaini = 0.14$.
+Cada una de las tablas enumera el identificador GES de la estrella y los rangos de $\teff$, $\gsurf$, $\feh$, A(Li) y edad estimada utilizados en la simulación.
+
+\newpage
+\KOMAoptions{paper=landscape,pagesize}
+\recalctypearea
+
+\begin{longtable}[c]{|l l l l || c c c c c c c c c c|}
+	\hline
+	& & & & & & & & $\omegaini$ & & & & & \\
+	Open Cluster & Age (Gyr) & [Fe/H] & $N_*$ & 0.095 & 0.10 & 0.105 & 0.11 & 0.115 & 0.12 & 0.125 & 0.13 & 0.14 & 0.1425\\
+	\hline
+	Berkeley 21 & 2.138 & -0.21 & 744 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1 & 1 & 1\\
+	Berkeley 39 & 5.623 & -0.14 & 899 & 1 & 1 & 1 & 1 & 1 & 1 & 2 & 2 & 2 & 2\\
+	IC 2602 & 0.036 & -0.06 & 1836 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0\\
+	IC 4665 & 0.033 & 0.01 & 567 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0 & 0 & 0\\
+	Messier 67 & 3.981 & -0.02 & 131 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6\\
+	NGC 2141 & 1.862 & -0.04 & 853 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+	NGC 2355 & 1 & -0.13 & 208 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+	NGC 2420 & 1.698 & -0.15 & 562 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+	NGC 2425 & 2.399 & -0.13 & 528 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+	NGC 2451 & 0.035 & -0.08 & 1656 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 2 & 1\\
+	NGC 2516 & 0.24 & -0.04 & 759 & 0 & 0 & 0 & 1 & 1 & 1 & 3 & 4 & 4 & 3\\
+	NGC 3532 & 0.398 & -0.01 & 1145 & 1 & 1 & 0 & 1 & 1 & 1 & 1 & 1 & 0 & 1\\
+	NGC 6005 & 1.259 & 0.22 & 560 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1\\
+	NGC 6259 & 0.269 & 0.18 & 494 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1\\
+	NGC 6281 & 0.513 & -0.04 & 320 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
+	NGC 6405 & 0.035 & -0.02 & 701 & 1 & 1 & 1 & 0 & 1 & 3 & 2 & 0 & 0 & 0\\
+	NGC 6633 & 0.692 & -0.03 & 1662 & 2 & 2 & 2 & 1 & 1 & 0 & 0 & 0 & 1 & 1\\
+	NGC 6709 & 0.191 & -0.02 & 730 & 2 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
+	Trumpler 20 & 1.862 & 0.13 & 1213 & 3 & 3 & 3 & 2 & 2 & 2 & 2 & 2 & 2 & 2\\
+	\hline
+	\caption{Lista de los OC seleccionados.
+Para cada OC se indica el nombre, la edad estimada, la metalicidad y el número de componentes.
+Además, se muestran los diferentes $\omegaini$ utilizados en las distintas simulaciones, donde $\omegaini = \oomegac$, $\Omega$ es la velocidad angular de la estrella en la superficie estelar, y $\omegac$ es la velocidad superficial en el ecuador de una estrella en rotación donde la fuerza centrífuga equilibra la gravedad newtoniana.
+Para cada entrada del OC, se informa por $\omegaini$ de cuántos componentes del OC cuyos parámetros estelares, tras compararlos con los equivalentes calculados por las simulaciones, se seleccionan.
+Los componentes OC seleccionados se utilizan como referencia en las imágenes que se muestran en las secciones siguientes.}
+	\label{tab:oc_reduced_list}	
+\end{longtable}
+\newpage
+\KOMAoptions{paper=portrait,pagesize}
+\recalctypearea
 

### 2024-01-31 project/octave/plotting_routines.m

@@ -1563,10 +1563,10 @@ function age_vs_teff_plots(gauss_fields, rotational_vels, is_var_vel, ytick, axi
   plot(sun_age, sun_log_g, 's', 'markersize', 10, 'color', [0.5,0.1,0.8], 'markerfacecolor', [0.5,0.1,0.8]);
 
   % Plot reference marks
-  line("xdata",[2.5e7,2.5e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", [0.3,0.3,0.3]);
-  line("xdata",[3.5e7,3.5e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", [0.3,0.3,0.3]);
-  line("xdata",[5.4e7,5.4e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", [0.3,0.3,0.3]);
-  line("xdata",[11.2e7,11.2e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", [0.3,0.3,0.3]);
+  line("xdata",[2.5e7,2.5e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", "cyan");
+  line("xdata",[3.5e7,3.5e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", "cyan");
+  line("xdata",[5.4e7,5.4e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", "magenta");
+  line("xdata",[11.2e7,11.2e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", "magenta");
 
   l = legend(labels, "location", leg_loc);
   set (l, "fontsize", legend_font_size);
@@ -2633,18 +2633,18 @@ function plot_3_5G_var_vel(rot_vels)
   age_vs_li_plots(gauss_fields(idx_3_5G,:), rot_vels, true, 0.5, [1.0e5,1.0e10,0,3.5], 'southwest', 'A(Li7) - 3.5G & var. rotational velocity', 'li_var_vel_3_5g');
[…]

### 2024-02-01 project/overleaf/Tesis III/capitulos/capitulo03.tex
WIP Tesis

@@
 
-\subsubsection{El mecanismo de dinamo}
+\subsection{El mecanismo de dinamo}
 Una dinamo es el proceso por el cual la energía mecánica se convierte
@@
 
-\subsubsection{Campos magnéticos internos}
+\subsection{Campos magnéticos internos}
 Nuestro Sol es una estrella activa en lo que se refiere al campo magnético.
@@
 
-\subsubsection{Campos magnéticos superficiales} \label{surf_mf}
+\subsection{Campos magnéticos superficiales} \label{surf_mf}
 La intensidad del campo magnético medida en la superficie del Sol es de aproximadamente $B\approx1\,\Gauss$ \cite{Weber1967,DAntona2000,Morin2012}, el doble del valor del de la Tierra.
@@
 
-\section{Efectos del frenado magnético}
+\section{Efectos del frenado magnético - Parte I}
 En su enfoque más aceptado, el MB está vinculado al trabajo de Skumanich \cite{Skumanich1972} en el que desarrolla una ley empírica de su evolución en paralelo a la edad de la estrella.
@@
 
-\subsection{Formalismos de frenado magnético} \label{mod_mb}
+\subsection{Formalismo de frenado magnético de intesidad fija} \label{mod_mb}
 Comenzamos enumerando los aspectos más relevantes y las suposiciones realizadas en la modelización de la evolución de la rotación, el frenado magnético y el momento angular en MESA.
@@
 
+\section{Extendiendo MESA - Parte II}
+
+\subsection{Rutina de frenado magnético de intensidad constante}
+EXPLICAR QUE SUSTITUIMOS LA RUTINA POR DEFECTO CON LA NUESTRA
+\begin{lstlisting}[language=Fortran, caption={Rutina de par de torsión.}, label={lst:jdot_cantiello}]
+	real function calculate_jdot_rate_cantiello(s, bf_star) result(new_j_dot)
+	use const_def
+	type (star_info), pointer, intent(in) :: s
+	real(dp), intent(in) :: bf_star
+
+	real(dp) :: r_st, m_st, i_st, alfven_r
+	real(dp) :: omega_surf, m_dot, eta_surf, v_inf, v_esc
+
+	!Star data
+	r_st = s% r(1)
+	m_st = s% m(1)
+	omega_surf = s% omega(1)
+	!omega_surf = s% omega_avg_surf
+
+	! escape and infinite velocities
+	! 100000 transform from km/s to cm/s
+	v_esc = (618 * ((Rsun/r_st)*(m_st/Msun))**0.5) * 100000
+	v_inf = 1.92 * v_esc
+
+	!m_dot = s% star_mdot !This gives the mass loss rate in Mstar/year
+	m_dot = s% mstar_dot !This in g/s
+
+	eta_surf = ((r_st * bf_star)**2)/(abs(m_dot) * v_inf)
+
+	!Formula 2.3 Cantiello's MESA assigment
+	new_j_dot = two_thirds * m_dot * omega_surf * (r_st**2) * eta_surf
+	end function
+\end{lstlisting}
+
+\subsection{Rutina de distribución de pérdida de momento angular}
+EXPLICAR QUE LOS DOS TIPOS DE RUTINAS: INTESIDAD FIJA, INTENSIDAD VARIABLE
+\begin{lstlisting}[language=Fortran, caption={Rutina de distribución de pérdida de momento angular.}, label={lst:dis_jdot}]
+	subroutine distribute_j_dot(s, total_j_dot, sz_info, mb_jdot_list)
+	type (star_info), pointer, intent(in) :: s
+	real(dp), intent(in) :: total_j_dot
+	type (star_zone_info), pointer, intent(in) :: sz_info
+	real(dp), dimension(:), pointer, intent(out) :: mb_jdot_list
+	integer :: k
+	real(dp) :: sum_jdot, dm_jdot, dm_bar_jdot, factor
+
+	!By default, no lost of angular moment
+	mb_jdot_list(:) = 0.0
+
+	do k = sz_info% top_zone, sz_info% bot_zone, 1
+	!Here the jdot distribution strategy is defined
+	!Simple rule of three distribution loss of angular momentum based on the 
+	!angular momentum of the zone vs total angular momentum of the convective zone
+
+	!IMPORTANT: Don't forget to divide by dm(k) in order to get an "specific" jdot
+	mb_jdot_list(k) = ((s% dm(k) * s% r(k)**2 * total_j_dot) / &
+	(sz_info% d_mass * sz_info% d_radius**2)) / s% dm(k)
+
+	end do
+	end subroutine distribute_j_dot
+
+\end{lstlisting}
+
 \endinput

### 2024-02-01 project/overleaf/Tesis III/capitulos/capitulo04.tex
WIP Tesis

@@
 
-\subsubsection{Rutina de frenado magnético de intensidad constante}
-EXPLICAR QUE SUSTITUIMOS LA RUTINA POR DEFECTO CON LA NUESTRA
-\begin{lstlisting}[language=Fortran, caption={Rutina de par de torsión.}, label={lst:jdot_cantiello}]
-      real function calculate_jdot_rate_cantiello(s, bf_star) result(new_j_dot)
-         use const_def
-         type (star_info), pointer, intent(in) :: s
-         real(dp), intent(in) :: bf_star
-
-         real(dp) :: r_st, m_st, i_st, alfven_r
-         real(dp) :: omega_surf, m_dot, eta_surf, v_inf, v_esc
-
-         !Star data
-         r_st = s% r(1)
-         m_st = s% m(1)
-         omega_surf = s% omega(1)
-         !omega_surf = s% omega_avg_surf
-
-         ! escape and infinite velocities
-         ! 100000 transform from km/s to cm/s
-         v_esc = (618 * ((Rsun/r_st)*(m_st/Msun))**0.5) * 100000
-         v_inf = 1.92 * v_esc
-
-         !m_dot = s% star_mdot !This gives the mass loss rate in Mstar/year
-         m_dot = s% mstar_dot !This in g/s
-
-         eta_surf = ((r_st * bf_star)**2)/(abs(m_dot) * v_inf)
-
-         !Formula 2.3 Cantiello's MESA assigment
-         new_j_dot = two_thirds * m_dot * omega_surf * (r_st**2) * eta_surf
-    end function
-\end{lstlisting}
-
-\subsection{Rutina de distribución de pérdida de momento angular}
-EXPLICAR QUE LOS DOS TIPOS DE RUTINAS: INTESIDAD FIJA, INTENSIDAD VARIABLE
-\begin{lstlisting}[language=Fortran, caption={Rutina de distribución de pérdida de momento angular.}, label={lst:dis_jdot}]
-      subroutine distribute_j_dot(s, total_j_dot, sz_info, mb_jdot_list)
-         type (star_info), pointer, intent(in) :: s
-         real(dp), intent(in) :: total_j_dot
-         type (star_zone_info), pointer, intent(in) :: sz_info
-         real(dp), dimension(:), pointer, intent(out) :: mb_jdot_list
-         integer :: k
-         real(dp) :: sum_jdot, dm_jdot, dm_bar_jdot, factor
-
-         !By default, no lost of angular moment
-         mb_jdot_list(:) = 0.0
-
-         do k = sz_info% top_zone, sz_info% bot_zone, 1
-            !Here the jdot distribution strategy is defined
-            !Simple rule of three distribution loss of angular momentum based on the 
-            !angular momentum of the zone vs total angular momentum of the convective zone
-
-            !IMPORTANT: Don't forget to divide by dm(k) in order to get an "specific" jdot
-            mb_jdot_list(k) = ((s% dm(k) * s% r(k)**2 * total_j_dot) / &
-                  (sz_info% d_mass * sz_info% d_radius**2)) / s% dm(k)
-
-         end do
-      end subroutine distribute_j_dot
-
-\end{lstlisting}
-
 \endinput

### 2024-02-01 project/overleaf/Tesis III/capitulos/capitulo06.tex
WIP Tesis

@@
 
-\section{Frenado magnético según Gallet \& Bouvier (2013)}
+\section{Efectos del frenado magnético - Parte II}
+
+\subsection{Formalismo de frenado magnético de intesidad fija}
+
+\subsubsection{Frenado magnético según Gallet \& Bouvier (2013)}
 Suponer que la intensidad del campo magnético de una estrella permanece constante a lo largo de su evolución es una simplificación necesaria en algunos escenarios, pero no se corresponde con las observaciones (INDICAR LAS REFERENCIAS).
@@
 
-Una vez que estemos en condiciones de calcular $\Beq$, podremos calcular $B_*$.\par
+Finalmente, estamos en condiciones de calcular $\Beq$ y $B_*$.\par
 
@@
 
-Por último, podemos calcular el factor de llenado de una estrella aplicando el ajuste realizado por \cite{Gallet2013} que reproduce más fielmente el factor de llenado medio del Sol actual ($\fsun$ = 0.001-0.01),
+Por último, podemos calcular el factor de llenado de una estrella aplicando el ajuste realizado por \cite{Gallet2013} que reproduce más fielmente el factor de llenado medio del Sol actual ($\fsun$ = [0.001-0.01]) ver \cite{Cranmer2011},
 
@@
 	\caption{La evolución de la intensidad del campo magnético y sus límites superior e inferior, en función del tiempo y $\oomegac$ para varios modelos de 1 $\msun$.
-El modelo incluye rotación PMS con $\oomegac$ = 0,1425.
+El modelo incluye rotación PMS con $\oomegac$ = 0.1425.
 Las líneas continuas representan la intensidad del campo magnético, mientras que las líneas discontinuas representan la evolución angular de la estrella.
@@
 \par
+
+Una vez que hemos caracterizado cómo evoluciona la intensidad del campo magnético, necesitamos modelar cómo se produce la pérdida de momento angular.
+A medida que la estrella se contrae durante la fase de pre-secuencia principal (PMS) hacia el comienzo de la secuencia principal (ZAMS), su velocidad angular aumenta, lo que conduce a un aumento gradual del campo magnético medio.
+En nuestro análisis, observamos intensidades máximas de hasta $1.0,\mathrm{kG}$, como se muestra en la figura \ref{fig:mag_field_limits_var_vel_g3}.
+A continuación, entra en una fase de saturación (aunque con una tendencia descendente) hasta $1.0$ Ga, y finalmente disminuye bruscamente como resultado de la menor velocidad angular de la estrella.
+En el formalismo anterior utilizamos un modelo aplicable a campos magnéticos de intensidad no variable y de baja intensidad que oscilan entre unos pocos Gauss ($1.0,\mathrm{G} \leq B \leq 10.0,\mathrm{G}$).
+Las intensidades significativamente mayores requeridas en el presente formalismo hacen inadecuado el modelo empleado anteriormente.\par
 
@@
 
-donde $\Bp$ es la intensidad del campo magnético en la superficie estelar, $\vesc$ es la velocidad de escape, $m = 0,1675$, $K_1 = 1,30$ y $K_2 = 0,0506$ \cite{Gallet2013}.
+donde $\Bp$ es la intensidad del campo magnético en la superficie estelar, $\vesc$ es la velocidad de escape, $m = 0.1675$, $K_1 = 1.30$ y $K_2 = 0.0506$ \cite{Gallet2013}.
 
 \begin{ceqn}
-	\begin{equation}
-		\vesc = \sqrt{\frac{2\,G\,\mstar}{\rstar}} \label{eq:vesc}
-	\end{equation}
+	\begin{align}
+		\Bp &= f_*B_* \label{eq:bp}\\
+		\vesc &= \sqrt{\frac{2\,G\,\mstar}{\rstar}} \label{eq:vesc}
+	\end{align}
 \end{ceqn}
@@
 Además, no existen argumentos sólidos que sugieran que el parámetro de longitud de mezcla sea el mismo en todas las estrellas y en todas las fases evolutivas \cite{Pasetto2014}.
-En \cite{Navarro2020} mostramos cómo una parametrización MLT alternativa podría producir resultados en línea con las observaciones.
-El ajuste del sobreimpulso (over-shooting) de MLT y del parámetro libre $\amlt$ parece ser necesario para explicar las abundancias de Li en los OC's.\par
+En \cite{Navarro2020} mostramos cómo una parametrización alternativa de la longitud de mezcla (ML) podría producir resultados en línea con las observaciones.
+El ajuste del sobreimpulso (over-shooting) de ML y del parámetro libre $\amlt$ parece ser necesario para explicar las abundancias de Li en los OC's.\par
 
@@
 \subsection{MLT $\alpha$ variable según Sonoi et al. (2019)}
-En \cite{Sonoi2018}, los autores introdujeron una forma de calibrar $\amlt$ para estrellas de tipo solar en la que su valor aumenta al aumentar $\gsurf$ o disminuir $\teff$,
+En \cite{Sonoi2018}, los autores introdujeron una forma de calibrar $\amlt$ para estrellas de tipo solar en la que su valor aumenta al aumentar $\gsurf$ o disminuir $\teff$.
+Calibraron valores de $\amlt$ para varios modelos 3D simulados usando el código CO$^5$BOLD \cite{Freytag2011} y los transfirieron a modelos 1D desarrollados por \cite{Ludwig1998}.
+La parametrización que obtuvieron fue:
 
@@
 \end{ceqn}
+
 donde
+
 \begin{ceqn}
@@
 
-y $a_i$ son los coeficientes resultantes de la función de ajuste \ref{eq:alpha_ml} a los valores $\alpha$ calibrados para el modelo de convección MLT-BV \cite{Sonoi2018}: $a_0=1,790295$, $a_1=-0,14954$, $a_2=0,069574$, $a_3=-0,00829$, $a_4=0,013165$, $a_5=0,080333$, $a_6=-0,03306$.
+y $a_i$ son los coeficientes resultantes de la función de ajuste \ref{eq:alpha_ml} a los valores $\alpha$ calibrados para el modelo de convección MLT \cite{Sonoi2018}: $a_0=1.790295$, $a_1=-0.14954$, $a_2=0.069574$, $a_3=-0.00829$, $a_4=0.013165$, $a_5=0.080333$, $a_6=-0.03306$.
 \par
@@
 
-\section{Extendiendo MESA - Parte II}
+\section{Extendiendo MESA - Parte III}
 \subsection{Rutina de frenado magnético de intensidad variable}

### 2024-02-01 project/overleaf/Tesis III/capitulos/capitulo09.tex
WIP Tesis

% !TeX root = ../libro.tex
% !TeX encoding = utf8

\chapter{Conclusiones finales}\label{ch:septimo-capitulo}

\section{Conclusiones finales}
\subsection{Campos magnéticos de intensidad fija vs. variable}
En \cite{Caballero2020} ya abordamos el estudio de la abundancia superficial de litio en el Sol. Aquel trabajo previo sentó las bases para el desarrollo del presente. En él avanzábamos como futuras líneas de investigación un conocimiento más profundo sobre cómo los campos magnéticos están ligados a la rotación y cómo evolucionan en el tiempo. También mostramos cómo ajustando el valor de $\amlt$ podíamos influir en las abundancias de Li. Por tanto, una comparación entre los resultados obtenidos anteriormente y los de este estudio pone en perspectiva las diferencias y similitudes entre ambos estudios, así como los avances realizados desde entonces. La Tabla \ref{tab:caballero2023_2020} resume las principales diferencias entre ambos trabajos, en línea con lo ya avanzado, es decir, incorporamos formalismos que nos permiten evolucionar temporalmente el $B$ y $\amlt$.\par

\begin{table}
	\centering
	\begin{threeparttable}
		\begin{tabular}{lll} 
			\hline
			Parameter & This work & \cite{Caballero2020}\\
			\hline
			Convection & MLT variable $\alpha_{\textrm{MLT}}$ & MLT fixed $\alpha_{\textrm{MLT}}=1.82$\\
			Overshoot & deactivated & $f_{\textrm{ov,core}}=0.0160$,\\ & & $f_{\textrm{ov,sh}}=0.0174$\\
			Magnetic Braking & \cite{Gallet2013} & \cite{Ud-Doula2008} \\ Formalism & & \\
			Magnetic Field & B(G) variable & B(G) pre-set [3.0G-5.0G]\\ Strength & & \\
			Number of OC's & 64 & 1 \\
			Source of data & GDR3.0 GES DR5.0 & \cite{Sestito2005} \\
			\hline
		\end{tabular}
	\end{threeparttable}
	\caption{Comparison with \cite{Caballero2020} work.}
	\label{tab:caballero2023_2020}
\end{table}

La Figura \ref{fig:li_var_vel_4_0g4} muestra la evolución de la abundancia de Li para $\omegaini$ entre 0.0084 y 0.0336 en presencia de un campo magnético con una intensidad de 4G. Hemos aumentado la versión original con el mismo conjunto de datos OCs recogidos aplicados en este nuevo trabajo, de forma que sea posible una comparación consistente entre ambos formalismos. Si centramos nuestra atención en la simulación con $\omegaini$ = 0.0336, la simulación reproduce A(Li) = 1.024 dex, valor que está en línea tanto con el del Sol (1.1 $\pm$ 0.1 dex) como con los valores obtenidos en la Figura \ref{fig:li_var_vel_var_g_3} del presente trabajo (1.133 dex) para $\omegaini$ = 0.14 y 0.1425. Aunque en este último caso los modelos se inicializaron con valores de $\omegaini$ más altos que en el primero, lo que resulta en $B$ hasta 3 órdenes de magnitud más intensivos ($\approx$ 1KG frente a 4G) la simulación en ambos trabajos acaba produciendo valores muy similares tanto para la velocidad rotacional como para el A(Li). Esto se explica en parte porque una mayor intensidad de campo magnético implica un efecto MB más intenso (véanse las figuras \ref{fig:rot_vel_var_vel_var_g3} y \ref{fig:rot_vel_var_g3}). \& \ref{fig:rot_vel_var_vel_4_0g4}). En el primer caso, el efecto de desaceleración es más agudo que en el segundo, por lo que deja de destruir Li antes.\par

Sin embargo, este efecto no es suficiente para contrarrestar eficazmente la destrucción de Li. El otro elemento decisivo que entra en juego es el valor de $\amlt$. Como se ejemplifica en \cite{Caballero2020} su valor tiene un impacto inversamente proporcional en la destrucción de Li, donde un valor más bajo resultó en una menor destrucción de Li, particularmente durante el PMS. En la figura \ref{fig:alpha_mlt_var_vel_g3} observamos la evolución temporal del valor $\amlt$. Hasta la última fase de la PMS, el valor $\amlt$ es sustancialmente inferior a 1,82, el valor preestablecido e invariante utilizado en nuestro trabajo anterior, lo que se traduce en una menor destrucción de Li y permite al modelo alcanzar la ZAMS con una reserva de Li $\approx$ 29\% dex mayor. Esta mayor reserva de Li compensa la mayor destrucción inducida por los modelos inicializados con $\omegaini$ mayor velocidad angular.\par

Del mismo modo, procedemos a mostrar la evolución de la velocidad de rotación superficial en la Figura \ref{fig:rot_vel_var_vel_4_0g4} y seleccionamos la simulación con $\omegaini$ = 0.0336. De nuevo, los valores similares se explican por la mayor eficiencia del MB en el nuevo modelo. Aunque las simulaciones alcanzan la ZAMS con valores de velocidad de rotación de hasta el 200\% para los modelos inicializados con $\omegaini$ más altos, el campo magnético más intenso produce una desaceleración más pronunciada para el mismo intervalo de tiempo. Este hecho es especialmente notable en el periodo de tiempo comprendido entre la ZAMS y 1 Ga. La velocidad de rotación obtenida en el ecuador de 4.82 km/s, es un valor bastante próximo a 4.72 km/s, es decir, el valor simulado para $\omegaini$ = 0.0336 en la Figura \ref{fig:rot_vel_var_vel_var_g3}.\par

A la vista de los valores obtenidos, podemos concluir que el nuevo modelo introducido en el presente trabajo no sólo es congruente con los resultados de los trabajos en los que se basa, sino también con los de \cite{Caballero2020}. Adicionalmente, esta nueva versión expone la ventaja de no tener que recurrir al uso de unos valores prefijados para $B$ y $\amlt$, y sin recurrir a activar el overshooting. En definitiva, esto representa una mejora significativa y apoya nuestra línea de investigación orientada a evitar valores preestablecidos haciéndolos evolucionar dinámicamente con parámetros solares clave.\par

\begin{figure}
	\centering
	\includegraphics[width=0.7\textwidth]{img/paper2/li_var_vel_4_0g4.pdf}
	\caption{La evolución de la abundancia superficial del \isotope[7]{Li} en relación con el \isotope[1]{H}, en función del tiempo para varios modelos de 1 $\msun$. Los modelos incluyen un campo magnético con una intensidad de 4G. La línea negra continua representa el modelo de referencia según \cite{Choi2016}. El resto de líneas son modelos que incluyen rotación inicial con $\omegaini$ entre 0.0084 y 0.0336, respectivamente. La estrella morada son las abundancias superficiales de Li para el Sol actual \cite{Asplund2009}. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:li_var_vel_4_0g4}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.7\textwidth]{img/paper2/rot_vel_var_vel_4_0g4.pdf}
	\caption{La evolución de la velocidad de rotación de la superficie, en función del tiempo para varios modelos de 1 $\msun$. Los modelos incluyen un campo magnético con una intensidad de 4G, rotación inicial con $\omegaini$ entre 0.0084 y 0.0336, respectivamente y MB. La estrella púrpura es la velocidad angular superficial para el Sol actual \cite{Gill2012}. La línea vertical discontinua hace referencia a la ZAMS.}
	\label{fig:rot_vel_var_vel_4_0g4}
\end{figure}

\subsection{Comparación con Gossage et al. (2021)}
Debido a la existencia de similitudes entre el trabajo de \cite{Gossage2021} y éste, procedemos a realizar una comparación entre ambos. Entre las similitudes encontramos el hecho de que ambos trabajos se centran en el frenado magnético, haciendo uso de MESA para las simulaciones. Entre las principales diferencias se encuentra que nuestros modelos incorporan una variable $B$ y $\amlt$. Finalmente, ambos comparan sus resultados de simulación (periodos de rotación y abundancias de Li) con arranques localizados en OCs.\par

En \cite{Gossage2021}, los autores también abordan el problema de AML desde un punto de partida similar al nuestro, pero con algunas variaciones notables. En la Tabla \ref{tab:gassage_vs_navarro} comparamos las similitudes y diferencias entre ambos estudios. En ambos trabajos los modelos consideran la rotación no sólo en el MS sino también en el PMS, con la distinción de que en el trabajo de Gossage tienen en cuenta el bloqueo del disco. Este aspecto no es tan relevante en nuestras simulaciones ya que nos hemos centrado en los perfiles de A(Li) y el trabajo de Gossage se centra principalmente en modelar los perfiles de rotación de las estrellas, que se ven afectados de forma significativa por el mecanismo de bloqueo del disco. Del mismo modo, también se consideran los efectos del frenado magnético en la pérdida de momento angular, en nuestro caso para estrellas con 1$\msun$ (gemelas-solares) y en el suyo con 0.1$\msun$$le$$\mstar$$\le$1.3$\msun$. Ambos modelos utilizan también MLT, pero en nuestro caso el valor $\amlt$ no está prefijado, sino que varía a lo largo de la simulación.\par

Estos dos trabajos se complementan ya que se basan en formalismos diferentes. Por un lado, nuestro trabajo se basa en la propuesta de \cite{Gallet2013} cuyos parámetros más relevantes son $\rossby$ y la presencia de un campo magnético de intensidad variable (B), mientras que Gossage presenta dos formalismos: \cite{Matt2015} y \cite{Garraffo2018}. El primero se caracteriza principalmente por $\rossby$ y la $\mstar$, mientras que el segundo por $\rossby$ y un parámetro que da cuenta de la complejidad de los campos magnéticos, siendo igual a 1 para un campo dipolar y mayor para multipolos de orden superior. En cuanto a las observaciones de OCs, en este trabajo hemos tenido acceso a un mayor número de ellas, y hemos cubierto un rango más amplio de edades [0.001-6.761] Gyr. Por último, ambos trabajos simulan sus respectivos modelos en MESA pero utilizando versiones diferentes.\par

\cite{Gossage2021} consiguen reproducir con mayor precisión el perfil de rotación del Sol a su edad actual \cite[ver][panel (a) de las Figuras 2, 4 \& 6]{Gossage2021} aunque como describen los autores del trabajo, calibraron sus modelos para ello, incluyendo un parámetro adicional libre de difusión ($v_0$). En nuestro trabajo, nuestro modelo no incluye este tipo de parametrización adicional y sigue siendo capaz de reproducir una velocidad de rotación ecuatorial de 4.72 km/s, un valor no especialmente alejado del valor nominal establecido de 2.0 km/s. En lo que respecta a A(Li), su modelo no consigue reproducir el valor para estrellas similares al Sol, incluido el propio Sol \cite[ver][panel (d) en las Figuras 2 \& 4, y (b) en la Figura 6]{Gossage2021}. Recalibrando algunos de los parámetros de su modelo y/o incorporando mecanismos adicionales que contribuyan a la destrucción del Li, los autores esperan obtener valores en línea con los datos observacionales. En nuestro caso (véase la sección \ref{sec_conclusiones}) pudimos reproducir A(Li) con mayor precisión.

\begin{table}
	\centering
	\begin{threeparttable}
		\begin{tabular}{lll} 
			\hline
			Parameter & This work & \cite{Gossage2021}\\
			\hline

			Rotation & At PMS \& MS & At PMS \& MS\\
			Solar Mass & $\mstar$=1$\msun$ & 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$ \\
			Convection & MLT variable $\alpha_{\textrm{MLT}}$ & MLT fixed $\alpha_{\textrm{MLT}}$\\
			Magnetic Braking & \cite{Gallet2013} & \cite{Matt2015}, \\ Formalism & & \cite{Garraffo2018}\\
			Magnetic Field & B(G) variable & B(G) pre-set\\ Strength & & \\
			Disk locking & Not present & Present\\
			Number of OC's & 64 & 11 \\
			Source of data & GDR3.0 GES DR5.0 & Several works \\
			MESA version & r10398 & r11701\\
			\hline
		\end{tabular}
	\end{threeparttable}
	\caption{Comparison with \cite{Gossage2021} work.}
	\label{tab:gassage_vs_navarro}    
\end{table}

\section{Conclusiones y trabajos futuros} \label{sec_conclusions}
Hemos demostrado, a través de los diferentes modelos estelares simulados, que los efectos inducidos por la combinación de ambos mecanismos de rotación y frenado magnético ofrecen una forma plausible de reconciliar los datos observacionales con los modelos teóricos. Proponemos que este resultado podría alcanzarse mediante una ley de interdependencia entre $\Omega$ y $B$. A diferencia de nuestro trabajo anterior, estos nuevos modelos eliminan las restricciones de preestablecer valores iniciales y fijos a lo largo de su evolución temporal para la intensidad del campo magnético y el parámetro $\amlt$. Hemos contrastado los resultados de los modelos con las medidas obtenidas por las misiones Gaia y GES para estrellas de características similares a nuestro Sol, centrándonos en sus abundancias de Li.\par

Nuestros modelos parecen sugerir un límite inferior a las abundancias superficiales de Li de 1.133 dex para estrellas similares al Sol, valor que se encuentra dentro del margen de error obtenido para el Sol (1.1 $\pm$ 0.1 dex). Por otro lado, predicen una velocidad de rotación en el ecuador (4.72 km/s) superior a la del Sol (2,0 km/s), por lo que podrían quemar más Li y reproducir finalmente los reservorios de Li existentes en el Sol. Junto con esta mayor velocidad angular, se desarrollaron fuertes campos magnéticos que desencadenaron un intenso frenado magnético que ralentizó su velocidad angular, pero no lo suficiente como para igualar el valor $\Omega$ actual del Sol. Además, hay que señalar que nuestros modelos no reproducen la intensidad media del campo magnético del Sol.\par

Aunque hemos mostrado cómo tanto una intensidad de campo magnético en evolución como una parametrización MLT podrían producir resultados acordes con las observaciones, debemos analizar los resultados obtenidos con cautela y no extraer conclusiones prematuras. A pesar de estas deficiencias, podemos extraer la siguiente conclusión del presente trabajo:

\begin{enumerate}
	\item Un campo magnético con intensidad variable es una alternativa válida a los modelos con un valor prefijado, ya que producen resultados en línea con trabajos anteriores para la historia rotacional de estrellas similares al Sol.
	\item Un formalismo $\amlt$ adaptativo variable podría eliminar la necesidad de prefijar su valor. Utilizándolo, pudimos reproducir valores en línea con trabajos anteriores. Además, también podría servir para derivar un intervalo de valores potenciales en los casos en que un valor por adelantado $\amlt$ es desconocido o incluso no deseable.
	\item Nuestros modelos parecen sugerir un límite inferior para A(Li)=1.133 dex a las observaciones proporcionadas por Gaia y GES para estrellas gemelas solares.   
	\item Aunque los modelos en el presente trabajo se inicializan a valores $\omegaini$ más altos que en nuestro trabajo anterior, 0.1425 y 0.0336 respectivamente, lo que resulta en $B$ hasta 3 órdenes de magnitud más intensos, $\approx$ 1KG frente a 4G, la simulación en ambos trabajos acaba produciendo valores muy similares tanto para la velocidad rotacional como para el A(Li). En el primer caso, esto se explica porque una mayor intensidad de campo magnético implica un efecto MB intensivo. Para el segundo, los valores $\amlt$ más bajos producidos por el formalismo adaptativo preservan un mayor depósito de Li durante el PMS.
\end{enumerate}

Nos gustaría concluir con las siguientes reflexiones para los próximos pasos. Es imperativo aumentar nuestra comprensión de los interiores estelares y los mecanismos de agotamiento del Li para la muestra de gemelos solares con el fin de realizar comparaciones más precisas. Además, ampliar el rango de masas en nuestros modelos nos permitirá confrontar los resultados con un rango más amplio de observaciones. Es un hecho que nuestros modelos siguen destruyendo demasiado Li al comparar los valores simulados con los de Gaia y GES.\par

Tomando como punto de partida que las estrellas de rotación más rápida destruyen más Li, son necesarios otros mecanismos, además del MB, para explicar los datos observacionales. Entre ellos, el bloqueo al disco protoestelar podría contribuir a contrarrestar la aceleración de la estrella durante el PMS. De este modo, actuarían dos mecanismos de frenado a diferentes edades de la estrella. El bloqueo del disco actuaría durante la PMS y el MB principalmente durante la MS. Con este enfoque se establecería un mecanismo de autorregulación sobre la velocidad angular de la estrella que acabaría influyendo directamente tanto en la evolución del Li como en la intensidad del campo magnético. Para esta última se esperan intensidades menores que finalmente podrían converger con la de nuestro Sol.\par

Por otro lado, una menor velocidad de rotación también implica una menor destrucción de Li para nuestros modelos. Una forma de compensar este menor Li remanente es recurrir al over-shooting (ver \cite{Caballero2020}) un efecto que no ha sido simulado en nuestros modelos. Activando este mecanismo, permitiríamos que el Li fuera alcanzado por burbujas de material lo suficientemente calientes como para destruirlo, contribuyendo así a disminuir su abundancia. Lo ideal, como hemos hecho con $\amlt$, sería no introducir el overshooting como un parámetro fijo, sino dependiente de otros parámetros estelares fundamentales.\par

\endinput
%--------------------------------------------------------------------
% FIN DEL CAPÍTULO. 
%--------------------------------------------------------------------

### 2024-02-02 project/overleaf/Tesis III/capitulos/capitulo05.tex
WIP Tesis

@@
+% !TeX root = ../libro.tex
+% !TeX encoding = utf8
+
 \chapter{Resultados de frenado magnético de intensidad fija}\label{ch:quinto-capitulo}
@@
 \begin{table}
+	\begin{threeparttable}
 		\centering
-	\caption{Summary of adopted physics in MESA \cite[based on][]{Choi2016}.}
-	\label{tab:phy_mesa}
 		\begin{tabular}{ll} 
 			\hline
-		Parameter & Adopted prescriptions and values\\
+			Parámetro & Prescripciones y valores adoptados\\
 			\hline
-		Solar Abundance & $X_{\odot}=0.7154, Y_{\odot}=0.2703, Z_{\odot}=0.0142$\\
-		Equation of State & OPAL+SCVH+MacDonald+HELM+PC\\
-		Opacity & OPAL Type I for log T $\geq$ 4 \\ & Ferguson for logT $\leq$ 4\\
-		Reaction Rates & JINA REACLIB\\
-		Boundary Conditions & ATLAS12; $\tau$=100 tables + photosphere\\
-		Difussion & Track \isotope[1]{H}, \isotope[2]{He}, \isotope[7]{Li}, \isotope[7]{Be}\\
-		Rotation & Differential rotation at PMS \& MS\\
-		Convection & MLT + Ledoux, $\alpha_{MLT}$ = 1.82\\
-		Overshoot & time-dependent, diffusive, \\ & $f_{ov,core}=0.0160$,\\ 
-		& $f_{ov,sh}=0.0174$\\
-		Semiconvection & $\alpha_{sc}=0.1$\\
-		Thermohaline & $\alpha_{th}=666$\\
-		Rotational Mixing & Include SH, ES, GSF, SSI \& DSI\\
-		Magnetic Effects & Magnetic braking based on idealized \\ & monopole field\\
-		Magnetic Field & B(G) variable between [3.0 - 5.0]\\
-		Mass Loss & activated, $\Dot{M}_{max} = 10^{-3} \: \msun \: yr^{-1}$\\
-		Angular Moment Loss & activated, $\Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A}$\\
+			Abundancia Solar & $X_{\odot}=0.7154, Y_{\odot}=0.2703, Z_{\odot}=0.0142$\\
+			Ecuación de estado & OPAL+SCVH+MacDonald+HELM+PC\\
+			Opacidad & OPAL Tipo I para log T $\geq$ 4 \\ & Ferguson para logT $<$ 4\\
+			Tasas de Reacción & JINA REACLIB\\
+			Condiciones de Contorno & ATLAS12; $\tau$=100 tablas + fotoesfera\\
+			Difusión & Rastreo de \isotope[1]{H}, \isotope[2]{He}, \isotope[7]{Li}, \isotope[7]{Be}\\
+			Esquema de Rotación & Rotación diferencial en PMS \& MS\\ & Incluye SH\tnote{1}  , ES\tnote{2}  , GSF\tnote{3}  , SSI\tnote{4}  , DSI\tnote{5}\\
+			Termohalina & $\alpha_{\textrm{th}}=666$\\
+			Convección & $\alpha_{MLT}$ = 1.82 + Ledoux\\
+			Semiconvección & $\alpha_{sc}=0.1$\\
+			Sobreimpulso & $f_{ov,core}=0.0160$, \& $f_{ov,sh}=0.0174$\\
+			Campo Magnético & B(G) prefijado con valores puntuales entre [3.0 - 5.0]\\
+			Pérdida de Masa & $\Dot{M}_{\textrm{max}} = 10^{-3} \: \msun \: yr^{-1}$\\
+			Périda de Momento Angular & $\Dot{J} = \frac{2}{3} \mwind \;\Omega_*\; \ralfven^2$\\
 			\hline
 		\end{tabular}
+		\begin{tablenotes}\footnotesize
+			\item (1) Solberg-Hoiland, (2) Eddington–Sweet
+			\item (3) Goldreich–Schubert–Fricke, (4) Secular Shear Instability
+			\item (5) Dynamical Shear Instability
+		\end{tablenotes}
+	\end{threeparttable}
+	\caption{Resumen de la física adoptada en MESA \cite{Choi2016}[basado en][].}
+	\label{tab:phy_mesa}
 \end{table}
 
-Adoptamos abundancias a escala solar y asumimos una abundancia inicial solar ($Z = Z_{\odot, \mathrm{pr} = 0,0142}$).
-También adoptamos los siguientes valores nominales para expresar las propiedades estelares en unidades SI, $\rsun = 6,957x10^{10}\, cm$ y $\msun = 1,988x10^{33}\, g$ que son coherentes con las resoluciones de la IAU \cite{Mamajek2015}.
+Adoptamos abundancias a escala solar y asumimos una abundancia inicial solar ($Z = Z_{\odot, \mathrm{pr} = 0.0142}$).
+También adoptamos los siguientes valores nominales para expresar las propiedades estelares en unidades SI, $\rsun = 6.957x10^{10}\, cm$ y $\msun = 1.988x10^{33}\, g$ que son coherentes con las resoluciones de la IAU \cite{Mamajek2015}.
 Para una descripción detallada de la física adoptada en este trabajo, remitimos al lector a \cite{Choi2016}.

### 2024-02-02 project/overleaf/Tesis III/capitulos/capitulo07.tex
WIP Tesis

--- 2024-01-29 project/overleaf/Tesis III/capitulos/capitulo07.tex
@@
+% !TeX root = ../libro.tex
+% !TeX encoding = utf8
+
 \chapter{Resultados de frenado magnético de intensidad variable}\label{ch:septimo-capitulo}
-\section{Configuración de los modelos} \label{marco_teorico}
+\section{Configuración de los modelos} \label{marco_teorico_ii}
+Como venimos exponiendo en el capítulo anterior, con la consideración de campos magnéticos de intensidad variable ($B$) dependientes de los parámetros estelares $\Omega$ y $\teff$ llevamos nuestra línea de trabajo un paso más allá.
+La parametrización base utilizada es básicamente la misma que hemos utilizado al abordar los campos magnéticos de intensidad fija.
+La innovación fundamental radica en la introducción de una intensidad de campo magnético variable ($B$) y un parámetro $\amlt$ variable.
+En cuanto al tratamiento de la mezcla convectiva con sobreimpulso (overshooting), es evidente una desviación del modelo anterior; desactivamos deliberadamente esta característica en nuestro modelo actual.
+Esto aísla el impacto de una variable $\amlt$ en A(Li) de los efectos asociados con el sobreimpulso convectivo.
+En particular, la mezcla convectiva con sobreimpulso se aborda habitualmente mediante extensiones ad hoc de la Teoría de la Longitud de Mezcla (MLT), introduciendo así un parámetro libre adicional.
+La Tabla \ref{tab:phy_mesa_ii} enumera las similitudes y diferencias en la configuración de los modelos utilizados en \cite{Caballero2020} y en este trabajo.\par
+
+\begin{table}
+	\begin{threeparttable}
+		\centering
+		\begin{tabular}{ll} 
+			\hline
+			Parámetro & Prescripciones y valores adoptados\\
+			\hline
+			Abundancia Solar & $X_{\odot}=0.7154, Y_{\odot}=0.2703, Z_{\odot}=0.0142$\\
+			Ecuación de estado & OPAL+SCVH+MacDonald+HELM+PC\\
+			Opacidad & OPAL Tipo I para log T $\geq$ 4 \\ & Ferguson para logT $<$ 4\\
+			Tasas de Reacción & JINA REACLIB\\
+			Condiciones de Contorno & ATLAS12; $\tau$=100 tablas + fotoesfera\\
+			Difusión & Rastreo de \isotope[1]{H}, \isotope[2]{He}, \isotope[7]{Li}, \isotope[7]{Be}\\
+			Esquema de Rotación & Rotación diferencial en PMS \& MS\\ & Incluye SH\tnote{1}  , ES\tnote{2}  , GSF\tnote{3}  , SSI\tnote{4}  , DSI\tnote{5}\\
+			Termohalina & $\alpha_{\textrm{th}}=666$\\
+			\textbf{Convección} & $\alpha_{\textrm{MLT}}$ variable dependiente de $\teff$\\ & \& $\gsurf$ + Ledoux\\
+			Semiconvección & $\alpha_{\textrm{sc}}=0.1$\\
+			\textbf{Sobreimpulso} & $f_{\textrm{ov,core}}=0.0$, \& $f_{\textrm{ov,sh}}=0.0$\\
+			\textbf{Campo Magnético} & B(G) variable, dependiente de $\rho$, $\teff$ \&  $\Omega$\\
+			Pérdida de Masa & $\Dot{M}_{\textrm{max}} = 10^{-3} \: \msun \: yr^{-1}$\\
+			Périda de Momento Angular & $\Dot{J} = \Omega_* \; \mwind \; \ralfven^2$\\
+			\hline
+		\end{tabular}
+		\begin{tablenotes}\footnotesize
+			\item (1) Solberg-Hoiland, (2) Eddington–Sweet
+			\item (3) Goldreich–Schubert–Fricke, (4) Secular Shear Instability
+			\item (5) Dynamical Shear Instability
+		\end{tablenotes}
+	\end{threeparttable}
+	\caption{Resumen de la física adoptada en MESA \cite{Choi2016,Caballero2020}[basado en][]. Resaltados en negrita los parámetros con diferente configuración de los trabajos referenciados.}
+	\label{tab:phy_mesa_ii}
+
+\end{table}
+
 A COMENTAR
@@
 \section{Comparativa MB intensidad fija vs variable}
-\section{Conclusiones finales}
+\section{Conclusiones}
 

### 2024-02-06 project/overleaf/Tesis III/capitulos/capitulo05.tex
WIP Tesis

@@
 Calculamos la evolución de modelos estelares 1 $\msun$ con metalicidad inicial solar y $\oomegac$ variable entre $0.0084$ y $0.0336$.
-Los modelos más avanzados para disposiciones más complejas del campo magnético, así como la evolución de su intensidad a lo largo de la vida de la estrella, se dejaron para trabajos posteriores.\par
+Hemos adoptado el siguiente enfoque teórico para establecer cuándo la estrella está alcanzando la ZAMS: cuando tanto la fracción de masa central del \isotope[1]{H} se ha reducido en $\Delta( \isotope[1]{H})= 0.0001$ respecto a su valor inicial, como la luminosidad de la estrella ($L$) está casi totalmente ($L \geq 0.99$) generada por las reacciones nucleares del \isotope[1]{H} ocurren simultáneamente.
+\par
 
@@
 
-\section{Conclusiones} \label{sec_conclusions}
+\section{Conclusiones} \label{sec_conclusions_i}
 Hemos demostrado, a través de los diferentes modelos estelares simulados, que los efectos inducidos por la combinación de ambos mecanismos de rotación y frenado magnético ofrecen una forma plausible de reconciliar los datos observacionales con los modelos teóricos.

### 2024-02-06 project/overleaf/Tesis III/capitulos/capitulo07.tex
WIP Tesis

@@
 \section{Configuración de los modelos} \label{marco_teorico_ii}
+
+A COMENTAR
+Parametrización de los modelos
+Rango de valores a asignar a los parámetros libres
+Ciclo de control
+Para cada paso de simulación
+- Obtenemos o calculamos la intensidad del campo magnético
+- Calculamos la pérdida de momento angular inducida por el campo magnético
+- Distribuimos la pérdida de momento angular entre las capas de la estrella
+- Obtenemos o calculamos el valor de $\amlt$
+
+\section{Modelos de evolución estelar}
 Como venimos exponiendo en el capítulo anterior, con la consideración de campos magnéticos de intensidad variable ($B$) dependientes de los parámetros estelares $\Omega$ y $\teff$ llevamos nuestra línea de trabajo un paso más allá.
@@
 
-A COMENTAR
-Parametrización de los modelos
-Rango de valores a asignar a los parámetros libres
-Ciclo de control
-Para cada paso de simulación
-- Obtenemos o calculamos la intensidad del campo magnético
-- Calculamos la pérdida de momento angular inducida por el campo magnético
-- Distribuimos la pérdida de momento angular entre las capas de la estrella
-- Obtenemos o calculamos el valor de $\amlt$
+Los modelos incluyen rotación desde el PMS.
+Hemos calculado la evolución de modelos estelares de $1\,\msun$ en metalicidad inicial solar con $\omegaini = \oomegac$ que varía entre $0.12$ y $0.1425$.
+A continuación, los resultados se comparan con las edades estelares de estrellas en OC recogidas por la misión Gaia y GES y referenciadas en la Tabla \ref{tab:oc_reduced_list}.\par
 
-\section{Modelos de evolución estelar}
 \subsection{Evolución del Li con MB de intensidad variable}
+Existen evidencias que abogan por una fuerte relación establecida entre la destrucción de Li y la rotación estelar, de forma que cuanto mayor es la velocidad angular, mayor es la destrucción de Li (ver \cite{Bouvier2018, Caballero2020}).
+El efecto del MB en la evolución temporal de A(Li) viene dado por su capacidad para eliminar AM de la estrella, y por tanto influir en su velocidad de rotación, haciendo que la estrella rote más lentamente y destruya menos Li.\par
+
+En la figura \ref{fig:li_var_vel_var_g_3} se muestra la evolución temporal de la abundancia superficial de Li para varios modelos de 1 $\msun$.
+Estos modelos se inicializaron con diferentes velocidades de rotación y tuvieron en cuenta los efectos del MB causado por un campo magnético variable.
+Si lo comparamos con la Figura \ref{fig:li_var_vel_0g} en la que se despreciaron los efectos del MB, observamos cómo se alteraron los perfiles de abundancia de Li a lo largo del PMS y el MS.
+Durante el PMS podemos describir el efecto como modesto, algo esperado y en línea con el hecho de que el AML causado por MB (ver Ec.~\ref{eq:j_dot}) depende directamente de la evolución de $\Omega$ y de la tasa de pérdida de masa.
+Si tenemos en cuenta que para las estrellas de tipo solar los modelos predicen una tasa de pérdida de masa total modesta, ese valor es incluso mucho menor en esta fase.
+Es en la fase de aproximación a la ZAMS cuando la velocidad angular alcanza su máximo.
+Esto, según nuestro modelo, desempeña un papel crucial tanto en la pérdida de masa como en la intensidad del campo magnético.
+Cuanto mayor sea $\Omega$, mayor será $\Dot{M}$ (véase la Figura \ref{fig:mdot_var_vel_g3}), y mayor será la intensidad del campo magnético (véase la Figura \ref{fig:mag_field_var_vel_g3}).
+Estos efectos se combinan y conducen a una acentuación del efecto de frenado magnético.
+Como consecuencia, se produce una ralentización de la destrucción del Li una vez que los modelos entran en el EM.
+Los resultados de nuestras simulaciones para esos modelos inicializados respectivamente a $\omegaini$ = 0.14 y 0.1425 reproducen A(Li) en línea con la de la abundancia del \isotope[7]{Li} de la superficie del Sol (1,1 $\pm$ 0,1 dex).
+Este último da un valor de 1,133 dex, lo que representaría una desviación de aproximadamente el 3\% del valor nominal (véase la Figura \ref{fig:li_var_vel_var_g_3}).\par
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/li_var_vel_var_g_3.pdf}
+	\caption{Se muestra la evolución de la abundancia relativa en superficie del \isotope[7]{Li} en comparación con el \isotope[1]{H} para varios modelos de 1 $\msun$.
+Éstos incluyen un campo magnético de intensidad variable, tasas de rotación iniciales que oscilan entre 0.12 y 0.1425 $\omegaini$, y MB.
+La estrella púrpura representa la abundancia actual de Li en la superficie del Sol \cite{Asplund2009}.
+Los otros puntos de color representan las abundancias superficiales de \isotope[7]{Li} para estrellas con parámetros dentro de los intervalos de selección especificados, correspondientes a la curva de evolución del mismo color.
+La línea vertical discontinua indica la Secuencia Principal de Edad Cero (ZAMS).}
+	\label{fig:li_var_vel_var_g_3}
+\end{figure}
+
+-----------------------------------------------------------------------
+----ESTA FIGURA HABRÁ QUE QUITARLA PORQUE SE INCLUYE YA ANTERIORMENTE--
+-----------------------------------------------------------------------
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/li_var_vel_0_0g_0.pdf}
+	\caption{Se representa la relación entre la abundancia superficial del isotope[7]{Li} y el isotope[1]{H} en función del tiempo para varios modelos de 1 $\msun$ sin campo magnético.
+El modelo de referencia, propuesto por \cite{Choi2016}, está representado por la línea negra continua.
+Los modelos adicionales con tasas de rotación iniciales que van de 0.0084 a 0.0336 se ilustran con las líneas restantes.
+Las abundancias superficiales de Li para el Sol actual, indicadas por una estrella púrpura, se basan en el estudio de \cite{Asplund2009}.
+La línea vertical discontinua corresponde a la Secuencia Principal de Edad Cero (ZAMS).}
+	\label{fig:li_var_vel_0g_ii}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=0.7\textwidth]{img/paper2/mdot_var_vel_g3.pdf}
+	\caption{Evolución de la pérdida de masa $\Dot{M}$ en función del tiempo para varios modelos de 1 $\msun$.
+Los modelos incluyen intensidad de campo magnético variable y rotación inicial con $\omegaini$ entre 0.12 y 0.1425.
+La línea vertical discontinua hace referencia a la ZAMS.}
+	\label{fig:mdot_var_vel_g3}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/mag_field_var_vel_g3.pdf}
+	\caption{La evolución de la intensidad del campo magnético, en función del tiempo y $\omegaini$ para varios modelos de 1 $\msun$.
+Los modelos incluyen una rotación inicial con $\omegaini$ entre 0.12 y 0.1425.
+Las líneas continuas representan la intensidad del campo magnético, mientras que las líneas de puntos representan la evolución angular de la estrella.
+La línea vertical discontinua hace referencia a la ZAMS.}
+	\label{fig:mag_field_var_vel_g3}
+\end{figure}
+
+En la Figura \ref{fig:rot_vel_var_vel_var_g3} se representan los perfiles de rotación de las estrellas para la superficie y para el fondo de la envoltura convectiva.
+Muestra cómo la velocidad angular aumenta de valor a medida que la estrella disminuye de radio a lo largo del PMS.
+Durante esta etapa de su evolución, la estrella tiene una estructura casi convectiva, como puede verse en la figura \ref{fig:cz_var_vel_var_g3}.
+A medida que se acerca a una edad de $\approx 10^6$ años, la zona convectiva de la estrella comienza a reducir su tamaño y el núcleo de la estrella alcanza las condiciones de temperatura, presión y densidad necesarias para desarrollar un núcleo radiativo.
+A su vez, la diferencia en las velocidades angulares de los distintos modelos se hace más evidente, siendo mayor en aquellos que fueron inicializados con una velocidad angular inicial más alta.
+Este hecho implica que la pérdida de masa también es mayor en aquellas estrellas que rotan más rápido.
+Como hemos descrito anteriormente, una mayor pérdida de masa implica también una mayor pérdida de momento angular.
+Los efectos combinados de los aumentos de la temperatura efectiva, la velocidad angular y la pérdida de masa hacen que la intensidad del campo magnético alcance su máximo en la fase de aproximación de la ZAMS (véase la figura \ref{fig:mag_field_var_vel_g3}).
+Observamos que los modelos con menor velocidad angular generalmente terminan exhibiendo valores más altos para la abundancia de Li en la superficie (ver Figuras~ \ref{fig:li_var_vel_var_g_3}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).\par
+
+Para los modelos en los que obtenemos valores de A(Li) en sintonía con los del Sol ($\omegaini$ = 0.14 y 0.1425), tenemos en cambio medidas para el $\Omega$ y $B$ que no están en sintonía con los solares respectivos.
+La velocidad de rotación de la superficie del Sol es de 2km/s con un error mínimo de 3m/s, y la intensidad media del campo magnético es de 1G.
+Para la simulación con $\omegaini$=0.1425, obtenemos una velocidad de rotación en el ecuador de 4.72 km/s.
+Aunque se trata de una medida del mismo orden de magnitud, representa una desviación del 235\%.
+Para la intensidad media del campo magnético tenemos 36.9G, muy lejos del valor de referencia.\par
+
+Como se describe en la sección \ref{mod_mb}, la rutina MB, que se activa cuando la estrella desarrolla un núcleo radiativo y una zona convectiva sobre él (ver Figura \ref{fig:mb_act_var_vel_g3}), distribuyó la cantidad total de LMA calculada según la Ec.~\ref{eq:k_jdot} entre las distintas capas que componían la CZ.
+En la Figura \ref{fig:cz_var_vel_var_g3} podemos observar la evolución de la CZ más externa normalizada con respecto al radio de la estrella para varios modelos de 1 $\msun$.
+De acuerdo con los modelos establecidos de evolución estelar, en una estrella de tipo solar la CZ cubre prácticamente en su totalidad gran parte del PMS.
+A partir de la ZAMS, su tamaño comienza a disminuir gradualmente en una primera fase y luego se hace más pronunciado en torno a los $4.0x10^8$ años de edad.
+Como puede verse en la figura \ref{fig:rot_vel_var_vel_var_g3}, la estrella, tras alcanzar su velocidad angular máxima al pasar por la ZAMS, comienza a desacelerar debido al frenado magnético.\par
+
+Es la presencia del MB la que impide que la estrella siga aumentando su velocidad angular.
+De este modo, la destrucción del Li se ve atenuada por una menor velocidad de rotación.
+Los modelos comienzan a desacelerarse gradualmente una vez superada la ZAMS y durante el periodo inicial de su estancia en la EM.
+La intensidad del campo magnético se mantiene cerca de su máximo, disminuye ligeramente y luego se reduce de forma más acusada.
+Su comportamiento refleja el de la evolución de la velocidad angular, consecuencia de los efectos de la MB.
+El proceso de desaceleración continúa progresivamente hasta que, en torno a la edad actual del Sol, se obtienen valores de velocidad angular muy similares a los del Sol.
+La desaceleración conduce a un menor efecto de las fuerzas centrífugas, lo que permite una contracción del radio de la estrella y, por tanto, un menor tamaño del CZ.
+Además, una menor velocidad angular tiene un efecto sobre la intensidad del campo magnético, como puede verse en la Figura \ref{fig:mag_field_var_vel_g3}.
+El acoplamiento entre la velocidad angular, la intensidad del campo magnético y el radio de la estrella es evidente y coherente.\par
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/rot_vel_var_vel_var_g3.pdf}
+	\caption{La evolución de la velocidad de rotación de la superficie, en función del tiempo para varios modelos de 1 $\msun$.
+Los modelos incluyen un campo magnético de intensidad variable, rotación inicial con $\omegaini$ entre 0.12 y 0.1425, respectivamente y MB.
+La estrella púrpura es la velocidad angular superficial para el Sol actual \cite{Gill2012}.
+La línea vertical discontinua hace referencia a la ZAMS.}
+	\label{fig:rot_vel_var_vel_var_g3}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/mb_act_var_vel_g3.pdf}
+	\caption{La activación de la rutina de frenado magnético en función de la presencia de un núcleo radiativo.
+Las líneas continuas señalan la activación (on) y desactivación (off) de la rutina de frenado magnético.
+Las líneas horizontales discontinuas informan sobre la naturaleza del núcleo de la estrella: radiativo (rad) o convectivo (conv).
+Por decisión de implementación, una vez activada la rutina, permanece activada incluso si la naturaleza del núcleo de la estrella cambia a convectiva.
+La línea vertical discontinua hace referencia a la ZAMS.}
+	\label{fig:mb_act_var_vel_g3}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/cz_var_vel_var_g3.pdf}
+	\caption{SUSTITUIR POR GRÁFICA MOSTRANDO SOLO LA CZ.
+Evolución del tamaño de la zona convectiva en función del tiempo para varios modelos de 1 $\msun$.
+El tamaño se expresa en términos de $\rsun$.
+Los modelos incluyen un campo magnético de intensidad variable, rotación inicial con $\omegaini$ entre 0.12 y 0.1425.
+La línea vertical discontinua hace referencia a la ZAMS.}
+	\label{fig:cz_var_vel_var_g3}
+\end{figure}
+
 \subsection{Evolución del Li con MB de intensidad y $\amlt$ variables}
+Los efectos de la rotación y el frenado magnético también pueden observarse en el diagrama H-R y en la evolución del $\amlt$.
+La inclusión de la rotación implica la aparición de fuerzas centrífugas que afectan tanto a la estructura de la estrella como a la composición química de los distintos estratos que la componen, así como a su temperatura y luminosidad.
+La rotación diferencial entre los límites del núcleo radiativo y las capas convectivas provoca efectos de mezcla en la denominada tacoclina.
+Los efectos de esta mezcla y de sus efectos hidrostáticos se rigen por la MLT, en la que el $\amlt$ desempeña un papel fundamental.
+Como ya se ha dicho, la MLT adolece de la arbitrariedad del valor de $\amlt$.
+La figura \ref{fig:alpha_mlt_var_vel_g3} muestra cómo evoluciona el parámetro $\amlt$ con el tiempo.\par
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/alpha_mlt_var_vel_g3.pdf}
+	\caption{La evolución de $\amlt$, en función del tiempo y $\omegaini$ para varios modelos de 1 $\msun$.
+Los modelos incluyen una rotación inicial con $\omegaini$ entre 0.12 y 0.1425.
+La línea vertical discontinua hace referencia a la ZAMS.
+La estrella púrpura y el rombo son los $\amlt$ dados por \cite{Sonoi2018} y \cite{Samadi2005}, respectivamente.}
+	\label{fig:alpha_mlt_var_vel_g3}
+\end{figure}
+
+Además, estas mismas fuerzas centrífugas, acentuadas en estrellas con mayor velocidad angular, refuerzan también el oscurecimiento gravitatorio \cite[véase por ejemplo ][]{Eggenberger2012,Paxton2019,Gossage2021}.
+La fuerza centrífuga hace que la estrella desplace su masa hacia fuera con respecto al eje de rotación, siendo el efecto más pronunciado en las regiones ecuatoriales que en los polos.
+La consecuencia es que la presión ($\gsurf$) a la que está sometido el gas es menor en las primeras que en las segundas.
+Este efecto hace que las estrellas que rotan a gran velocidad aparezcan menos densas ($\rho$), menos luminosas ($L$) y, por tanto, con una temperatura efectiva ($\teff$) más baja a medida que se acercan a la ZAMS (véase la figura \ref{fig:hr_var_vel_var_g_z13}).
+Si comparamos el modelo no rotatorio (línea sólida negra en la Figura \ref{fig:hr_var_vel_0g}) con los rotatorios, podemos reconocer que al final del PMS, estos últimos alcanzan la ZAMS con un $\teff$ menor que los primeros.
+Esto también repercute en la evolución de $\amlt$ ya que, como se ha comentado anteriormente, en nuestro modelo su valor depende proporcionalmente de $\teff$ y $\gsurf$.
+Ambos parámetros son comparativamente menores en las estrellas que giran más rápido que en las que lo hacen más despacio.
+Como consecuencia $\amlt$ arroja un valor menor en su evolución temporal en aquellas fases en las que la estrella rota más rápido, siendo este efecto más acentuado en la aproximación ZAMS.\par
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/hr_var_vel_var_g_z13.pdf}
+	\caption{Un ejemplo de cuadrícula solar de 1$\msun$ de huellas evolutivas estelares que cubre un rango de velocidades angulares.
+Muestra en detalle los efectos combinados del oscurecimiento gravitatorio y el frenado magnético en las huellas evolutivas.
+Los modelos incluyen un campo magnético de intensidad variable, rotación inicial con $\omegaini$ entre 0.12 y 0.1425.
+La presencia de un campo magnético produce estrellas más calientes debido a la influencia del frenado magnético en la velocidad de rotación de la estrella.}
+	\label{fig:hr_var_vel_var_g_z13}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/hr_var_vel_0_0g_z10.pdf}
+	\caption{Similar a la Figura \ref{fig:hr_var_vel_var_g_z13} pero mostrando ahora las trayectorias evolutivas estelares en ausencia (0G) de un campo magnético.
+La rotación se activa en los modelos en el PMS y esos modelos llegan antes a la ZAMS y a un $\teff$ menor que el que no rota (línea negra sólida).
+La luminosidad se expresa en términos de $\lsun$.
+$\lsun$ = 3.761.
+(Figura tomada de \cite{Caballero2020}.)}
+	\label{fig:hr_var_vel_0g}
+\end{figure}
+
+La figura \ref{fig:teff_logg_var_vel_g3} muestra cómo se comportan $\teff$ y $\gsurf$ a lo largo de la evolución de los modelos.
+Para aquellos con una mayor velocidad de rotación, observamos que tanto su temperatura como su gravedad superficial son menores que en los modelos de rotación más lenta, en consonancia con el oscurecimiento de la gravedad.
+Cabe destacar que en la fase de aproximación ZAMS observamos para el modelo de rotación más rápida (amarillo) que su gravedad superficial es mayor que la del resto de modelos más lentos.
+Esto puede explicarse observando la Figura \ref{fig:lograd_var_vel_g3}, que muestra la evolución del radio estelar para los distintos modelos, en función del tiempo y $\omegaini$.
+Alrededor del intervalo de $2.5x10^{7}$ y $3.5x10^{7}$ Ga (delimitado por las líneas cian) y después de la ZAMS, desde alrededor de $5.4x10^{7}$ a $11.2x10^{7}$ Ga (delimitado por las líneas magenta) el radio estelar del modelo más rápido es menor que el del resto, produciendo un $\gsurf$ mayor, y esto a su vez significa una menor pérdida de masa (véase la Fig.
+\ref{fig:mdot_var_vel_g3}).
+Durante este período el modelo más rápido lo hace a un ritmo menor y expone una gravedad superficial mayor.
+Recordemos que el radio estelar tiene una influencia inversamente cuadrática en el valor de $\gsurf$.
+Esta "anomalía" desaparece en cuanto el radio estelar vuelve a ser mayor para el modelo más rápido.\par
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/teff_logg_var_vel_g3.pdf}
+	\caption{La evolución de $\teff$ y $\gsurf$, en función del tiempo y $\omegaini$ para varios modelos de 1 $\msun$.
+Los modelos incluyen una rotación inicial con $\omegaini$ entre 0.12 y 0.1425.
+Los intervalos de edad [$2.5x10^{7},3.5x10^{7}$] Ga y [$5.4x10^{7},11.2x10^{7}$] Ga, delimitados por las líneas cian y magenta respectivamente, destacan periodos en los que el modelo más rápido con $\omegaini$=0,1425 expone una gravedad superficial superior a la del resto de los más lentos.
+La estrella morada es el $\teff$, y el cuadrado morado es la $\gsurf$ para el Sol actual \cite{Gill2012}.
+La línea vertical discontinua hace referencia a la ZAMS.}
+	\label{fig:teff_logg_var_vel_g3}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\textwidth]{img/paper2/lograd_var_vel_g3.pdf}
+	\caption{La evolución del radio estelar, en función del tiempo y $\omegaini$ para varios modelos de 1 $\msun$ y sus.
+Los modelos incluyen rotación inicial con $\omegaini$ entre 0.12 y 0.1425.
+La estrella púrpura es el radio para el Sol actual \cite{Gill2012}.
+La línea vertical discontinua hace referencia a la ZAMS.}
+	\label{fig:lograd_var_vel_g3}
+\end{figure}
+
 \section{Comparativa MB intensidad fija vs variable}

### 2024-02-07 project/overleaf/Tesis III/capitulos/capitulo09.tex
WIP Tesis

@@
 \section{Conclusiones finales}
-\subsection{Campos magnéticos de intensidad fija vs. variable}
+\section{Comparativa MB intensidad fija vs variable}
 En \cite{Caballero2020} ya abordamos el estudio de la abundancia superficial de litio en el Sol.
@@
 			\hline
-			Parameter & This work & \cite{Caballero2020}\\
+			Parámetro & Intensidad Variable & Intensidad fija\\
 			\hline
-			Convection & MLT variable $\alpha_{\textrm{MLT}}$ & MLT fixed $\alpha_{\textrm{MLT}}=1.82$\\
-			Overshoot & deactivated & $f_{\textrm{ov,core}}=0.0160$,\\ & & $f_{\textrm{ov,sh}}=0.0174$\\
-			Magnetic Braking & \cite{Gallet2013} & \cite{Ud-Doula2008} \\ Formalism & & \\
-			Magnetic Field & B(G) variable & B(G) pre-set [3.0G-5.0G]\\ Strength & & \\
+			Convección & MLT variable $\alpha_{\textrm{MLT}}$ & MLT prefijado $\alpha_{\textrm{MLT}}=1.82$\\
+			Sobreimpulso & desactivado & $f_{\textrm{ov,core}}=0.0160$,\\ & & $f_{\textrm{ov,sh}}=0.0174$\\
+			Formalismo Frenado & \cite{Gallet2013} & \cite{Ud-Doula2008} \\ Magnético & & \\
+			Intensidad Campo & B(G) variable & B(G) prefijado [3.0G-5.0G]\\ Magnético & & \\
 			Number of OC's & 64 & 1 \\
-			Source of data & GDR3.0 GES DR5.0 & \cite{Sestito2005} \\
+			Conjuntos de datos & GDR3.0 GES DR5.0 & \cite{Sestito2005} \\
 			\hline
@@
 	\end{threeparttable}
-	\caption{Comparison with \cite{Caballero2020} work.}
-	\label{tab:caballero2023_2020}
+	\caption{Comparación con el trabajo \cite{Caballero2020}.} \label{tab:caballero2023_2020}
 \end{table}
@@
 La Figura \ref{fig:li_var_vel_4_0g4} muestra la evolución de la abundancia de Li para $\omegaini$ entre 0.0084 y 0.0336 en presencia de un campo magnético con una intensidad de 4G.
-Hemos aumentado la versión original con el mismo conjunto de datos OCs recogidos aplicados en este nuevo trabajo, de forma que sea posible una comparación consistente entre ambos formalismos.
-Si centramos nuestra atención en la simulación con $\omegaini$ = 0.0336, la simulación reproduce A(Li) = 1.024 dex, valor que está en línea tanto con el del Sol (1.1 $\pm$ 0.1 dex) como con los valores obtenidos en la Figura \ref{fig:li_var_vel_var_g_3} del presente trabajo (1.133 dex) para $\omegaini$ = 0.14 y 0.1425.
+Hemos aumentado la versión original con el mismo conjunto de datos OCs recogidos ----REVISAR aplicados en este nuevo trabajo, de forma que sea posible una comparación consistente entre ambos formalismos.---- Si centramos nuestra atención en la simulación con $\omegaini$ = 0.0336, la simulación reproduce A(Li) = 1.024 dex, valor que está en línea tanto con el del Sol (1.1 $\pm$ 0.1 dex) como con los valores obtenidos en la Figura \ref{fig:li_var_vel_var_g_3} del presente trabajo (1.133 dex) para $\omegaini$ = 0.14 y 0.1425.
 Aunque en este último caso los modelos se inicializaron con valores de $\omegaini$ más altos que en el primero, lo que resulta en $B$ hasta 3 órdenes de magnitud más intensivos ($\approx$ 1KG frente a 4G) la simulación en ambos trabajos acaba produciendo valores muy similares tanto para la velocidad rotacional como para el A(Li).
@@
 En la figura \ref{fig:alpha_mlt_var_vel_g3} observamos la evolución temporal del valor $\amlt$.
-Hasta la última fase de la PMS, el valor $\amlt$ es sustancialmente inferior a 1,82, el valor preestablecido e invariante utilizado en nuestro trabajo anterior, lo que se traduce en una menor destrucción de Li y permite al modelo alcanzar la ZAMS con una reserva de Li $\approx$ 29\% dex mayor.
-Esta mayor reserva de Li compensa la mayor destrucción inducida por los modelos inicializados con $\omegaini$ mayor velocidad angular.\par
+Hasta la última fase de la PMS, el valor $\amlt$ es sustancialmente inferior a 1.82, el valor preestablecido e invariante utilizado en nuestro trabajo anterior, lo que se traduce en una menor destrucción de Li y permite al modelo alcanzar la ZAMS con una reserva de Li $\approx$ 29\% dex mayor.
+Esta mayor reserva de Li compensa la mayor destrucción inducida por los modelos inicializados con una $\omegaini$ mayor velocidad angular.\par
 
@@
 Este hecho es especialmente notable en el periodo de tiempo comprendido entre la ZAMS y 1 Ga.
-La velocidad de rotación obtenida en el ecuador de 4.82 km/s, es un valor bastante próximo a 4.72 km/s, es decir, el valor simulado para $\omegaini$ = 0.0336 en la Figura \ref{fig:rot_vel_var_vel_var_g3}.\par
+La velocidad de rotación obtenida en el ecuador de 4.82 km/s es un valor bastante cercano a 4.72 km/s, es decir, el valor simulado para $\omegaini$ = 0.0336 (Fig.~\ref{fig:rot_vel_var_vel_var_g3}).\par
 
 A la vista de los valores obtenidos, podemos concluir que el nuevo modelo introducido en el presente trabajo no sólo es congruente con los resultados de los trabajos en los que se basa, sino también con los de \cite{Caballero2020}.
-Adicionalmente, esta nueva versión expone la ventaja de no tener que recurrir al uso de unos valores prefijados para $B$ y $\amlt$, y sin recurrir a activar el overshooting.
-En definitiva, esto representa una mejora significativa y apoya nuestra línea de investigación orientada a evitar valores preestablecidos haciéndolos evolucionar dinámicamente con parámetros solares clave.\par
+Adicionalmente, esta nueva versión expone la ventaja de no tener que recurrir al uso de valores prefijados para $B$ y $\amlt$, y sin recurrir a activar el sobreimpulso.
+En definitiva, esto supone una mejora significativa y apoya nuestra línea de investigación encaminada a evitar valores prefijados haciéndolos evolucionar dinámicamente con parámetros solares clave.\par
 
@@
 
-\subsection{Comparación con Gossage et al. (2021)}
+\section{Comparativa con Gossage et al. (2021)}
 Debido a la existencia de similitudes entre el trabajo de \cite{Gossage2021} y éste, procedemos a realizar una comparación entre ambos.
@@
 Entre las principales diferencias se encuentra que nuestros modelos incorporan una variable $B$ y $\amlt$.
-Finalmente, ambos comparan sus resultados de simulación (periodos de rotación y abundancias de Li) con arranques localizados en OCs.\par
+Finalmente, ambos comparan los resultados de sus simulaciones (periodos de rotación y abundancias de Li) con arranques localizados en OCs.\par
 
-En \cite{Gossage2021}, los autores también abordan el problema de AML desde un punto de partida similar al nuestro, pero con algunas variaciones notables.
+En \cite{Gossage2021}, los autores también abordan el problema de la AML desde un punto de partida similar al nuestro, pero con algunas variaciones notables.
 En la Tabla \ref{tab:gassage_vs_navarro} comparamos las similitudes y diferencias entre ambos estudios.
 En ambos trabajos los modelos consideran la rotación no sólo en el MS sino también en el PMS, con la distinción de que en el trabajo de Gossage tienen en cuenta el bloqueo del disco.
-Este aspecto no es tan relevante en nuestras simulaciones ya que nos hemos centrado en los perfiles de A(Li) y el trabajo de Gossage se centra principalmente en modelar los perfiles de rotación de las estrellas, que se ven afectados de forma significativa por el mecanismo de bloqueo del disco.
-Del mismo modo, también se consideran los efectos del frenado magnético en la pérdida de momento angular, en nuestro caso para estrellas con 1$\msun$ (gemelas-solares) y en el suyo con 0.1$\msun$$le$$\mstar$$\le$1.3$\msun$.
-Ambos modelos utilizan también MLT, pero en nuestro caso el valor $\amlt$ no está prefijado, sino que varía a lo largo de la simulación.\par
+Este aspecto no es tan relevante en nuestras simulaciones, ya que nos hemos centrado en los perfiles de A(Li), y el trabajo de Gossage se centra principalmente en modelar los perfiles de rotación de las estrellas, que se ven significativamente afectados por el mecanismo de bloqueo del disco.
+Del mismo modo, también se consideran los efectos del frenado magnético en la pérdida de momento angular, en nuestro caso para estrellas con 1$\msun$ (gemelas solares) y en el suyo con 0.1$\msun$$le$$\mstar$$\le$1.3$\msun$.
+Ambos modelos utilizan también MLT pero en nuestro caso el valor $\amlt$ no está prefijado sino que varía a lo largo de la simulación.\par
 
@@
 El primero se caracteriza principalmente por $\rossby$ y la $\mstar$, mientras que el segundo por $\rossby$ y un parámetro que da cuenta de la complejidad de los campos magnéticos, siendo igual a 1 para un campo dipolar y mayor para multipolos de orden superior.
-En cuanto a las observaciones de OCs, en este trabajo hemos tenido acceso a un mayor número de ellas, y hemos cubierto un rango más amplio de edades [0.001-6.761] Gyr.
-Por último, ambos trabajos simulan sus respectivos modelos en MESA pero utilizando versiones diferentes.\par
+En cuanto a las observaciones de OCs, aquí cubrimos un rango más amplio de edades [0.001-6.761] Ga.
+Por último, ambos trabajos simulan sus respectivos modelos en MESA, pero utilizan versiones diferentes.\par
 
-\cite{Gossage2021} consiguen reproducir con mayor precisión el perfil de rotación del Sol a su edad actual \cite[ver][panel (a) de las Figuras 2, 4 \& 6]{Gossage2021} aunque como describen los autores del trabajo, calibraron sus modelos para ello, incluyendo un parámetro adicional libre de difusión ($v_0$).
+\cite{Gossage2021} consiguen reproducir con mayor precisión el perfil de rotación del Sol a su edad actual \cite[ver][panel (a) de las Figuras 2, 4 \& 6]{Gossage2021} aunque, como describen los autores del trabajo, calibraron sus modelos para ello, incluyendo un parámetro adicional libre de difusión ($v_0$).
 En nuestro trabajo, nuestro modelo no incluye este tipo de parametrización adicional y sigue siendo capaz de reproducir una velocidad de rotación ecuatorial de 4.72 km/s, un valor no especialmente alejado del valor nominal establecido de 2.0 km/s.
-En lo que respecta a A(Li), su modelo no consigue reproducir el valor para estrellas similares al Sol, incluido el propio Sol \cite[ver][panel (d) en las Figuras 2 \& 4, y (b) en la Figura 6]{Gossage2021}.
+En cuanto a A(Li), su modelo no reproduce el valor para estrellas similares al Sol, incluido el propio Sol \cite[ver][panel (d) de las Figuras 2 \& 4, y (b) de la Figura 6]{Gossage2021}.
 Recalibrando algunos de los parámetros de su modelo y/o incorporando mecanismos adicionales que contribuyan a la destrucción del Li, los autores esperan obtener valores en línea con los datos observacionales.
-En nuestro caso (véase la sección \ref{sec_conclusiones}) pudimos reproducir A(Li) con mayor precisión.
+En nuestro caso (véase la sección \ref{sec_conclusiones}), pudimos reproducir A(Li) con mayor precisión.
 
@@
 			\hline
-			Parameter & This work & \cite{Gossage2021}\\
+			Parámetro & Este trabajo & \cite{Gossage2021}\\
 			\hline
 
-			Rotation & At PMS \& MS & At PMS \& MS\\
-			Solar Mass & $\mstar$=1$\msun$ & 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$ \\
-			Convection & MLT variable $\alpha_{\textrm{MLT}}$ & MLT fixed $\alpha_{\textrm{MLT}}$\\
-			Magnetic Braking & \cite{Gallet2013} & \cite{Matt2015}, \\ Formalism & & \cite{Garraffo2018}\\
-			Magnetic Field & B(G) variable & B(G) pre-set\\ Strength & & \\
-			Disk locking & Not present & Present\\
-			Number of OC's & 64 & 11 \\
+			Rotación & En PMS \& MS & En PMS \& MS\\
+			Masa solar & $\mstar$=1$\msun$ & 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$ \\
+			Convección & MLT variable $\alpha_{\textrm{MLT}}$ & MLT prefijada $\alpha_{\textrm{MLT}}$\\
+			Formalismo Frenado & \cite{Gallet2013} & \cite{Matt2015}, \\ Magnético & & \cite{Garraffo2018}\\
+			Intensidad Campo & B(G) variable & B(G) prefijada\\ Magnético & & \\
+			Bloqueo de Disco & No activado & Activado\\
+			Número de OC's & 64 & 11 \\
 			Source of data & GDR3.0 GES DR5.0 & Several works \\
-			MESA version & r10398 & r11701\\
+			Versión MESA & r10398 & r11701\\
 			\hline
@@
 	\end{threeparttable}
-	\caption{Comparison with \cite{Gossage2021} work.}
-	\label{tab:gassage_vs_navarro}    
+	\caption{Comparación con \cite{Gossage2021}.}\label{tab:gassage_vs_navarro}    
 \end{table}
 
-\section{Conclusiones y trabajos futuros} \label{sec_conclusions}
+\section{Trabajos futuros} \label{sec_conclusions}
 Hemos demostrado, a través de los diferentes modelos estelares simulados, que los efectos inducidos por la combinación de ambos mecanismos de rotación y frenado magnético ofrecen una forma plausible de reconciliar los datos observacionales con los modelos teóricos.

### 2024-02-28 project/overleaf/paper2/mblithium.tex
Update the CZ figure eliminating the core size track

@@
 % Keep the title short and informative.
-\title[Rotation, magnetic braking \& Li abundances II]{Variable magnetic field and mixing-length convection parameter to reproduce Li abundances and rotation period of the Sun and solar-type stars}
+\title[Rotation, magnetic braking \& Li abundances]{Variable magnetic field and adaptive mixing-length to reproduce Li abundances and rotation periods of solar-type stars in clusters}
 
@@
 R. Caballero Navarro,$^{1}$\thanks{E-mail: rcaballeron@correo.ugr.es}
-A. Garc\'ia Hern\'andez,$^{1,2}$\thanks{E-mail: agh@ugr.es}
-J.~C. Su\'arez$^{1,2}$\thanks{E-mail: jcsuarez@ugr.es}
+A. Garc\'ia Hern\'andez,$^{1}$\thanks{E-mail: agh@ugr.es}
+J.~C. Su\'arez$^{1}$\thanks{E-mail: jcsuarez@ugr.es}
 \\
@@
 $^{1}$Dept. Theoretical Physics and Cosmology, University of Granada (UGR), 18071, Granada, Spain\\
-$^{2}$Instituto de Astrof\'isica de Andaluc\'ia (CSIC), Glorieta de la Astronom\'ia S/N, 18008, Granada, Spain\\
 }
@@
 \begin{abstract}
-Finding the reasons why lithium (Li) surface abundance in the Sun and young stellar globular clusters are seemingly anomalous in present-day scenarios, will deliver a better understanding about the mechanisms which influence the Li depletion during the stellar evolution, in particular during the pre-main sequence (PMS) and main sequence (MS).\par
-
-The present work documents an approach in which neither the magnetic field strength nor the MLT parametrisation remains fixed throughout the evolution of the star.
-The influence of free parameters associated with $\amlt$ and $B$ significantly condition the the Li abundance.
-We document an alternative to prefixing values to these parameters which ends up providing results that are compatible with observational data for Li abundances of solar siblings hosted in young clusters.\par
-
-We explore, by simulating several grids of PMS and MS rotating models, how the effects of rotational mixing and of the rotational hydrostatic effects governs the Li abundances.
-The combined effects of the MLT and MB constraint the Li evolution of the solar models.
-We calculate accurate solar models for which the $\amlt$ and $B$ values are variable over time, i.e.
-not remaining fixed through the simulations, and make them dependent on important solar parameters.
-The data obtained from simulations are confronted with observational data measured on young open clusters collected by the Gaia-ESO Survey (GES).
-The results show how faster rotating models destroy more Li in the late PMS and during the MS, and how the effects of magnetic braking mitigate this destruction.
-The interconnection between the star's internal convective motions, effective temperature, surface gravity and rotational velocity, as well as its chemical composition, determine the strength of the magnetic field and impact the effectiveness of the convective motions involved in the destruction of Li.
+Investigating the apparent anomalies in lithium (Li) surface abundance observed in the Sun and young stellar globular clusters within contemporary astrophysical contexts holds significant promise for advancing our understanding of the mechanisms influencing Li depletion throughout stellar evolution.
+This study delves into the intricate interplay between rotational mixing and rotational hydrostatic effects in pre-main-sequence (PMS) and main-sequence (MS) stars by employing simulated grids of rotating models.
+The Li evolution of solar models is scrutinized under the combined influence of Mixing Length Theory (MLT) and magnetic braking (MB), where the magnetic field strength ($B$) and MLT parameterization ($\amlt$) dynamically evolve with key solar parameters.
+A novel approach, avoiding fixed values for these parameters, is proposed, yielding results consistent with observational data.\par
+
+We generate and test precise solar models that take into account the dynamic nature of $B$ and $\amlt$ during stellar evolution.
+To validate these models, we compare them with observational data from young open clusters obtained through the Gaia-ESO Survey (GES).
+Our results are consistent with previous studies that have separately addressed these approaches.
+Additionally, we integrate both approaches in our models and examine the combined impact they have on the rotational history and surface Li abundances in solar models.
+The outcomes of our study are in agreement with previous research and are compatible with observational data.\par
+
+The results indicate that there is a strong minimum limit of 1.133 dex for the surface Li abundances in stars similar to the Sun.
+This aligns with observations of the Sun and provides insight into the complex balance of physical processes that control the amount of Li in stellar atmospheres.
+Similarly, our calculations yield theoretical values of $\amlt$ that fall within the range of [1.76, 1.78] obtained in previous studies.
+Based on these promising findings, our models offer a consistent and comprehensive alternative to those that use fixed values and only consider the variables $B$ and $\amlt$ separately.
+We have successfully obtained results that, although not an exact match to the Sun's actual measurements, are still consistent with its surface Li abundance, angular velocity, and predicted $\amlt$ values.
 
@@
 \section{Introduction} \label{sec_intro}
-The angular momentum loss (AML) is considered to be governed by magnetically-coupled stellar winds which exert a braking force on the star that eventually causes a loss of angular momentum \citep{UdDoula2002}.
-
-Over the last few years, decades, data from different observational campaigns have allowed studies of the rotational evolution of low and medium-mass stars \citep{Hartman2010,Gallet2013,Bouvier2016}. 
-Among them we find studies \citep{Ud-Doula2008,Cranmer2011,Gallet2015,Amard2016} that focus on modelling the effects of magnetic braking (MB) and how its coupling with stellar winds impact the rotational evolution of the star.
-These models rely on a number of parameters which are fixed, among them the magnetic field strength B (we will come back to this point later).
-This approach is not sufficient to explain the observational data, and interaction with other processes in the angular velocity evolution of low-mass stars is necessary: the distribution of angular momentum loss in the interior of the star along the pre-main sequence (PMS) and main sequence (MS) \citep{Charbonnel2005, Eggenberger2008, Eggenberger2009, Navarro2020}, and the star - protostellar disc interaction during the PMS \citep{Bouvier2008,Gallet2013,Eggenberger2012,Zanni2012,Bouvier2016}.\par
-
-Discrepancies in Li abundances have been detected in young cluster for which there is as yet no coherent explanation \citep[see][and references therein]{Navarro2020}.
-Li is destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches $\tli \approx 2.5 x 10^6\; K$.
-The solar-type stars are known to have a convection zone (CZ) that covers much of the stellar radius during the PMS.
-In this way, the convective movements that occur in its interior manage to drag the Li to areas in which the BCZ limit exceeds the $\tli$ and Li is burned \citep{Iben1965}.
-This process slows down as the star approaches the zero-age MS (ZAMS) when the convection zone retreats and moves away from areas where the internal temperature is high enough to destroy the Li.\par
-
-The impact of rotation both on PMS and Li depletion for solar-type stars has already extensively debated in the past \citep{Pinsonneault1997,Jeffries2004,Somers2014} and revised more recently on the basis of the availability of more accurate measures \citep{Gallet2013,Bouvier2016, Bouvier2018, Franciosini2022}.
-Of particular interest for this paper are the Li observational data measured on young open clusters.
-The Gaia-ESO Survey (GES) \citep{Gilmore2012,Randich2013,Randich2022} has produced very high quality spectroscopy records of around 110 000 Milky Way stars, in the field and in open clusters, down to magnitude 19, It has delivered good-quality radial velocities and stellar parameters for a large fraction of its unique target stars.
-Elemental abundances were derived for up to 31 elements for targets observed with the Fibre Large Array Multi Element Spectrograph (FLAMES) and the UV-Visual Echelle Spectrograph (UVES).
-Lithium abundances are delivered for about 1/3 of the sample.\par
-
-As indicated by its name, GES intends to complement the data about parallaxes and proper motions from the Gaia satellite \citep{Mignard2005} with extremely precise information on radial velocities (RVs), Li, and chemistry in general.
-The first Gaia data release (GDR1) \citep{Brown2016} dated on 2016 and contained information on the first 14 months of operation.
-Gaia data release 2 (GDR2) \citep{Brown2018}, released on 2018, contained positions, parallaxes, and proper motions for about 1.3 billion sources, together with photometry.
+Throughout the star's evolution, directly observable parameters such as its rotational velocity, effective temperature, surface gravity or abundance of chemical elements undergo variations.
+These parameters exert an influence, either directly or indirectly, on the angular momentum loss (AML) caused by the star's magnetic field.
+On the contrary, we encounter stellar convection motions, which, in most stellar evolution models, are described by the mixing-length theory (MLT).
+MLT makes use of the mixing length scale which is proportional to the height of the local pressure scale, and to the mixing-length parameter ($\amlt$).
+This parameter, $\amlt$, is considered a fixed free parameter and must be determined by comparing the stellar models with some reference, in most cases our Sun.
+Therefore, assuming that the magnetic field strength ($B$) and $\amlt$ remain constant throughout the stellar evolution is a necessary simplification in some scenarios but might not correspond with reality.\par
+
+There are different mechanisms that affect the AML in solar-type stars e.g.
+tidal forces in binary systems or stellar winds.
+In close binary systems, angular momentum and kinetic energy can be exchanged between the two stars and their orbits through tidal forces \citep{Li2018}.
+In the latter case, the magnetically-coupled stellar winds efficiently remove angular momentum,  leading to the spin-down of stars with age.
+This is due to the braking force exerted by stellar winds, which results in the eventual loss of angular momentum \citep{UdDoula2002}.\par
+
+In recent years, the rotational evolution of low- and intermediate-mass stars has been studied thanks to several observational campaigns \citep{Hartman2010, Gallet2013, Bouvier2016}.
+Some specific investigations, such as those by \citet{Ud-Doula2008, Cranmer2011, Gallet2015, Amard2016}, focus on modeling the effects of magnetic braking (MB) and its coupling with stellar winds on the rotational evolution of stars.
+These models rely on a number of parameters which are fixed, among them the magnetic field strength $B$.
+This approach is not sufficient to explain the observational data, and interaction with other processes in the angular velocity evolution of low-mass stars is necessary: e.g., the distribution of AML in the interior of the star along the pre-main sequence (PMS) and main sequence (MS) \citep{Charbonnel2005, Eggenberger2008, Eggenberger2009, Caballero2020}, and the star - protostellar disc interaction during the PMS \citep{Bouvier2008,Gallet2013,Eggenberger2012,Zanni2012,Bouvier2016}.\par
+
+The impact of rotation on both PMS and Li depletion in solar-type stars has been extensively debated \citep{Pinsonneault1997,Jeffries2004,Somers2014}, with recent revisions based on more accurate measurements \citep{Gallet2013, Bouvier2016, Bouvier2018, Franciosini2022}.
+Li is destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches $\tli \approx 2.5 x 10^6\; \mathrm{K}$.
+Solar-type stars are known to have a convection zone (CZ) that covers much of the stellar radius during the PMS.
+In this way, the convective movements that occur in its interior manage to drag Li to areas in which the BCZ limit exceeds $\tli$ and Li is burned \citep{Iben1965}.
+This process slows down as the star approaches the zero-age MS (ZAMS) when the convection zone retreats and moves away from areas where the internal temperature is high enough to destroy the Li.
+Discrepancies in Li abundances have been detected in young cluster for which there is as yet no coherent explanation \citep[see][and references therein]{Caballero2020}.\par
+
+The Gaia-ESO Survey (GES) \citep{Gilmore2012,Randich2013,Randich2022} has produced very high quality spectroscopy records of around 110 000 Milky Way stars, in the field and in open clusters (OC), down to magnitude 19.
+It has delivered good-quality radial velocities and stellar parameters for a large fraction of its unique target stars.
+Elemental abundances for up to 31 elements were derived for targets observed with the Fiber Large Array Multi-Element Spectrograph (FLAMES) and the UV-Visual Echelle Spectrograph (UVES).
+Li abundances are delivered for about 1/3 of the sample.\par
+
+The primary goal is to explore how MB contributes to the evolution of Li in the Sun and other solar-type stars.
+We aim to develop evolutionary stellar models in which both $B$ and $\amlt$ are free parameters that vary over time based on specific stellar parameters.
+We provide semi-empirical approximations for the calculation of AML resulting from braking caused by variable parameters $B$ and $\amlt$.
+Simulation data are compared with various stellar parameters obtained from Gaia and GES, allowing for the analysis, validation, and refinement of our models against a large candidate set.
+This has been implemented as an extension to Modules for Experiments in Stellar Astrophysics \citep[MESA; ][]{Paxton2011, Paxton2013,Paxton2015, Paxton2018, Paxton2019}.\par
+
+\section{Data} \label{sec_data}
+As indicated by its name, GES intends to complement the data on parallaxes and proper motions from the Gaia satellite \citep{Mignard2005} with extremely precise information on radial velocities (RVs), Li, and chemistry in general.
+The first Gaia data release (GDR1) \citep{Brown2016} dated 2016 and contained information on the first 14 months of operation.
+Gaia data release 2 (GDR2) \citep{Brown2018}, released in 2018, contained positions, parallaxes, and proper motions for about 1.3 billion sources, together with photometry.
 The third data release of Gaia, the Gaia early data release 3 (GEDR3) \citep{Brown2021} and the Gaia data release 3 (GDR3) \citep{Brown2022}, convey updated and more precise measures.
-GES enhances them and in its latest release (DR5.0) includes all the astrophysical parameters derived by the Gaia-ESO consortium for 114 324 stars \citep{Gilmore2022}.
-The interest for the present work are:
+GES enhances them and in its latest release (DR5.0) includes all astrophysical parameters derived by the Gaia-ESO consortium \citep{Gilmore2022}.
+Of interest for the present work are:
 \begin{itemize}
-    \item The radial and projected RVs.
-    \item Stellar parameters (effective temperature, surface gravity and metallicity).
-    \item Abundances of several elements, among them Li.
-    \item Specific parameters for tracing accretion and activity in young stars.
-    \item Cluster probability membership.  
+    \item The radial and projected RVs
+    \item Stellar parameters (effective temperature, surface gravity and metallicity)
+    \item Abundances of several elements, among them Li
+    \item Specific parameters for tracing accretion and activity in young stars
+    \item Cluster probability membership
 \end{itemize}
 
-\section{Data} \label{sec_method}
 The data sets provided by the Gaia mission in GEDR3 and GDR3, as well as those published by GES in DR5.0 form the main references used in this work.
-In general, we focus on the observations made on OCs, and particularly on those analysed in the \cite{Bragaglia2022} and \cite{Randich2022} papers.
+In general, we focus on the observations made on OCs, and particularly on those analysed in the \citet{Bragaglia2022} and \citet{Randich2022} papers.
 On the basis of the results presented in these papers, a set of 64 OCs (see Table \ref{tab:oc_full_list}) are selected as the reference dataset for our work.
@@
 	\centering
-	\begin{tabular}{|l l l l l || c c c c c | c c c c c|} 
-		\hline
-             & & & & & & & $\oomegac$ & & & & & $\oomegac$ & & \\
-		Open Cluster & GES\_FLD & Age (Gyr) & [Fe/H] & $N_*$ & 0.095 & 0.10 & 0.105 & 0.11 & 0.115 & 0.12 & 0.125 & 0.13 & 0.14 & 0.1425\\
-		\hline
-            Berkeley 21 & Br21 & 2.138 & -0.21 & 744 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1 & 1 & 1\\
-            Berkeley 39 & Br39 & 5.623 & -0.14 & 899 & 1 & 1 & 1 & 1 & 1 & 1 & 2 & 2 & 2 & 2\\
-            IC 2602 & IC2602 & 0.036 & -0.06 & 1836 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0\\
-            IC 4665 & IC4665 & 0.033 & 0.01 & 567 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0 & 0 & 0\\
-            Messier 67 & M67 & 3.981 & -0.02 & 131 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6\\
-            NGC 2141 & NGC2141 & 1.862 & -0.04 & 853 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
-            NGC 2355 & NGC2355 & 1 & -0.13 & 208 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
-            NGC 2420 & NGC2420 & 1.698 & -0.15 & 562 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
-            NGC 2425 & NGC2425 & 2.399 & -0.13 & 528 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
-            NGC 2451 & NGC2451 & 0.035 & -0.08 & 1656 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 2 & 1\\
-            NGC 2516 & NGC2516 & 0.24 & -0.04 & 759 & 0 & 0 & 0 & 1 & 1 & 1 & 3 & 4 & 4 & 3\\
-            NGC 3532 & NGC3532 & 0.398 & -0.01 & 1145 & 1 & 1 & 0 & 1 & 1 & 1 & 1 & 1 & 0 & 1\\
-            NGC 6005 & NGC6005 & 1.259 & 0.22 & 560 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1\\
-            NGC 6259 & NGC6259 & 0.269 & 0.18 & 494 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1\\
-            NGC 6281 & NGC6281 & 0.513 & -0.04 & 320 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
-            NGC 6405 & NGC6405 & 0.035 & -0.02 & 701 & 1 & 1 & 1 & 0 & 1 & 3 & 2 & 0 & 0 & 0\\
-            NGC 6633 & NGC6633 & 0.692 & -0.03 & 1662 & 2 & 2 & 2 & 1 & 1 & 0 & 0 & 0 & 1 & 1\\
-            NGC 6709 & NGC6709 & 0.191 & -0.02 & 730 & 2 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
-            Trumpler 20 & Trumpler20 & 1.862 & 0.13 & 1213 & 3 & 3 & 3 & 2 & 2 & 2 & 2 & 2 & 2 & 2\\
+	\begin{tabular}{|l l l l || c c c c c c c c c c|} 
+		\hline
+             & & & & & & & & $\omegaini$ & & & & & \\
+		Open Cluster & Age (Gyr) & [Fe/H] & $N_*$ & 0.095 & 0.10 & 0.105 & 0.11 & 0.115 & 0.12 & 0.125 & 0.13 & 0.14 & 0.1425\\
+		\hline
+            Berkeley 21 & 2.138 & -0.21 & 744 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1 & 1 & 1\\
+            Berkeley 39 & 5.623 & -0.14 & 899 & 1 & 1 & 1 & 1 & 1 & 1 & 2 & 2 & 2 & 2\\
+            IC 2602 & 0.036 & -0.06 & 1836 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0\\
+            IC 4665 & 0.033 & 0.01 & 567 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0 & 0 & 0\\
+            Messier 67 & 3.981 & -0.02 & 131 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6\\
+            NGC 2141 & 1.862 & -0.04 & 853 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            NGC 2355 & 1 & -0.13 & 208 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            NGC 2420 & 1.698 & -0.15 & 562 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            NGC 2425 & 2.399 & -0.13 & 528 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+            NGC 2451 & 0.035 & -0.08 & 1656 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 2 & 1\\
+            NGC 2516 & 0.24 & -0.04 & 759 & 0 & 0 & 0 & 1 & 1 & 1 & 3 & 4 & 4 & 3\\
+            NGC 3532 & 0.398 & -0.01 & 1145 & 1 & 1 & 0 & 1 & 1 & 1 & 1 & 1 & 0 & 1\\
+            NGC 6005 & 1.259 & 0.22 & 560 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1\\
+            NGC 6259 & 0.269 & 0.18 & 494 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1\\
+            NGC 6281 & 0.513 & -0.04 & 320 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
+            NGC 6405 & 0.035 & -0.02 & 701 & 1 & 1 & 1 & 0 & 1 & 3 & 2 & 0 & 0 & 0\\
+            NGC 6633 & 0.692 & -0.03 & 1662 & 2 & 2 & 2 & 1 & 1 & 0 & 0 & 0 & 1 & 1\\
+            NGC 6709 & 0.191 & -0.02 & 730 & 2 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
+            Trumpler 20 & 1.862 & 0.13 & 1213 & 3 & 3 & 3 & 2 & 2 & 2 & 2 & 2 & 2 & 2\\
             \hline
@@
  	\caption{List of selected OCs.
-For each OC, name, GES denomination, estimated age, metallicity and number of components are listed.
-The number of OC components selected by the different simulations are shown separately.
-This information is split into two groups, each corresponding to the set of simulations used in the figures shown in the following sections.}
+For each OC, name, estimated age, metallicity and number of components are listed.
+Additionally, the different $\omegaini$ used in the different simulations are shown, where $\omegaini = \oomegac$, $\Omega$ is the star angular velocity at stellar surface, and $\omegac$ is the surface velocity at the equator of a rotating star where the centrifugal force balances the Newtonian gravity.
+For each OC entry, it is informed per $\omegaini$ how many OC components whose stellar parameters, after comparing them with the equivalent ones calculated by the simulations, are selected.
+The sorted-out OC members are used as a reference in the pictures shown in the following sections.}
   	\label{tab:oc_reduced_list}
@@
 
-\begin{table*}
-	\centering
-	\begin{tabular}{l l l l l l l} 
-		\hline
-            ObjectId & $\teff$(K) & $\gsurf$(dex) & FeH(dex) & ALi(dex) & eALi(dex) & Age(Gyr)\\
-		\hline
-            8510914\_1157003 & 5843 & 4.42 & -0.03 & 1.8 & 0.08 & 3.981\\ 
-            8510991\_1146169 & 5797 & 4.4 & -0.01 & 1.65 & 0.08 & 3.981\\ 
-            8512177\_1144050 & 5818 & 4.46 & 0 & 1.99 & 0.06 & 3.981\\ 
-            8513941\_1200571 & 5792 & 4.41 & -0.02 & 1.49 & 0.13 & 3.981\\ 
-            8515559\_1148381 & 5794 & 4.43 & -0.04 & 1.56 & 0.1 & 3.981\\ 
-            8520350\_1147480 & 5875 & 4.47 & -0.03 & 1.21 & 0.24 & 3.981\\ 
-            \hline
-	\end{tabular}
- 	\caption{M67 - Filter setup: $\teff$(K) [5787.4892, 5887.6823], $\gsurf$(dex) [4.3979, 4.4977], $\feh$(dex) [-0.05, 0.05], Age(Gyr) [3.977, 3.985]}
-	\label{tab:oc_m67}
-\end{table*}
-
-\begin{table*}
-	\centering
-	\begin{tabular}{l l l l l l l} 
-		\hline
-            ObjectId & $\teff$(K) & $\gsurf$(dex) & FeH(dex) & ALi(dex) & eALi(dex) & Age(Gyr)\\
-		\hline
-            7571111-6048156 & 5084 & 4.54 & 0 & 2.29 & 0.08 & 0.24\\ 
-            7595031-6044149 & 5108 & 4.52 & -0.05 & 1.92 & 0.08 & 0.24\\ 
-            7595280-6032498 & 5093 & 4.49 & -0.03 & 2.29 & 0.08 & 0.24\\ 
-            \hline
-	\end{tabular}
- 	\caption{NGC2516 - Filter setup: $\teff$(K) [5010.0324, 5111.0095], $\gsurf$(dex) [4.4405, 4.5405], $\feh$(dex) [-0.05, 0.05], Age(Gyr) [0.23964, 0.24036]}
-	\label{tab:oc_ngc2516}
-
-\end{table*}
-
 \subsection{The selection of clusters}
-GES DR5.0 provides information on 114 324 stars corresponding to OCs, see \cite{Gilmore2022} for details. Of interest for the present work are those records that meet the condition:
+GES DR5.0 provides information on 114 324 stars corresponding to OCs, see \citet{Gilmore2022} for details. We select those records which satisfy any of the conditions:
 \begin{itemize}
@@
 
-This second filtering reduces the population of stars to be considered to a total of 5 895 records which are distributed among the 58 OCs listed in Table \ref{tab:oc_full_list}.\par
-
-\section{Method} \label{sec_method}
-The AML is considered to be governed by magnetically-coupled stellar winds which exert a braking force on the star that eventually causes a loss of angular momentum.
-In addition, throughout the star's evolution, directly observable parameters such as its rotational velocity, effective temperature, surface gravity or abundance of chemical elements undergo variations.
-These parameters have an influence, either directly or indirectly, on the angular momentum loss caused by the star's magnetic field.
-On the other hand, we have stellar convection which in most stellar evolution models is prescribed by MLT.
-That makes use of the mixing-length scale to express which in turn is proportional to the height of the local pressure scale, and to $\amlt$.
-$\amlt$ is considered a free-parameter and must be determined by comparing the stellar models with some calibrator, typically our Sun, remaining fixed.
-Therefore, assuming that the magnetic field strength ($B$) and $\amlt$ remain constant throughout the stellar evolution is a necessary simplification in some scenarios but might not correspond with reality.\par
-
-The method of work in the present work has as its initial premise the development of evolutionary stellar models in which both $B$, as well as $\amlt$ are not fixed as free parameters but their values vary over time as a function of certain stellar parameters.
+This second filtering reduces the population of stars to be considered to a total of 5 895 records which are distributed among the 64 OCs listed in Table \ref{tab:oc_full_list}.\par
+
 As benchmark data we take a subset of the observations obtained by GES.
-This subset consists of those stars that belong to OCs, and among them are "hidden" potential siblings of our Sun.
+This subset consists of those stars that belong to OCs, and among them are "hidden" potential twins of our Sun.
 To identify the latter, an additional screening is performed, this time using the values given by our models during the simulations for metallicity ($\feh$), effective temperature ($\teff$), and surface gravity ($\gsurf$).
-Finally we are left with those stars for which the GES observations have measured values congruent with our model results and from which we obtain their Li abundances.\par
-
-In the remainder of this paper we present in detail how we select those GES stars most similar to our Sun, and describe semi-empirical approximations for the calculation of the AML as a result of the braking caused by a magnetic field of varying strength, and to make the $\amlt$ value variable.
-This has been implemented as an extension to MESA.\par
-
-\subsection{The selection of closest solar siblings}
-Once we have pre-selected the stars belonging to the OCs listed in the previous chapter, our interest focus on identifying those that are most similar to our Sun i.e.
-solar siblings.
-Finding such solar twins is a difficult task since they are disseminated across our Galaxy.
+Finally we are left with those stars for which the GES observations have measured values congruent with our model results, from which we obtain their Li abundances.\par
+
+\subsection{The selection of closest solar twins}
+Once we have pre-selected the stars belonging to the OCs, our interest is focused on identifying those that are most similar to our Sun, i.e.
+solar twins.
+This brings the question, which stellar parameters should be considered to classify a star as solar twin?
 Finding them would help us to figure out the origin of our Sun and and gain information about what conditions prevailed in the OCs in which they were born.\par
 
-Several attempts have been made to find solar siblings in the past \citep[see][and references therein]{Adibekyan2018}.
-In most of them, the search method is oriented to search the candidates based on its kinematic properties and then their metallicities and chemical abundances are compared with those of our Sun.\par
-
-In our case we extend the selection criteria by adding additional constraints: the surface gravity and age of the star.
-These two new parameters represent a fine-tuning of the search process for sibling candidates to our Sun.
-We use the age of the OCs reported in Table \ref{tab:oc_reduced_list} and the data obtained from the set of simulations performed using Modules for Experiments in Stellar Astrophysics stellar evolution code \citep[MESA; ][]{Paxton2011, Paxton2013,Paxton2015, Paxton2018, Paxton2019}.
-Our models extend the investigations documented in \cite{Navarro2020} on the influence of rotation and magnetic braking on the star Li depletion during the pre-main sequence (PMS) and the main sequence (MS).
-The data obtained from the simulations are confronted by comparing different stellar parameters obtained by Gaia and GES.
-This allows us to analyse, validate and refine our models against a very large set of candidates.
-The end result of this process is to obtain a subset of candidate sibling stars of our Sun that help us to support (or refute) our line of investigation of how the Li abundance behaves throughout the evolution of solar-type stars.\par
-
-From the approximately 5 900 identified stars, we proceed to select those whose $\feh$, $\teff$, and $\gsurf$ values are within the following limits:
-\begin{itemize}
-    \item $0.05 \, \dex <= \feh <= 0.05 \, \dex$
-    \item $50.0 \, \Kelvin <= \teff <= 50.0 \, \Kelvin$
-    \item $0.05 \, \dex <= \gsurf <= 0.05 \, \dex$
-\end{itemize}
-
-The method of selection is as follows: Based on the values given by our simulations for the for the aforementioned, we proceed to select in the different OCs those stars that have values within the established ranges.
-That is, for the estimated age of a particular cluster, we calculate an interval of $\pm$0.1 for young OCs (age < 1Gyr) and $\pm$0.15 for old OCs (age >= 1Gyr) \citep[see][and reference values therein]{Cantat-Gaudin2020}.
-This age interval is used to select in our simulations the time step that is closest to the lower and upper age limits.
+Several attempts have already been made to find solar siblings and solar twins.
+The former refers to stars that formed in the same cluster as the Sun and therefore they should have ages and chemical abundances very similar to those of the Sun \citep[see][and references therein]{Adibekyan2018}.
+However, according to \citet{Strobel1996} solar twins are considered stars possessing fundamental physical parameters such as $\teff$, $\gsurf$, $\feh$, photometric properties, chemical composition, age, luminosity, rotation, and magnetic fields that are similar, if not identical, to those of the Sun.
+Locating these solar twins presents a formidable challenge due to strict criteria and their dispersion throughout the Milky Way.\par
+
+It is not unusual that, the applied search methods are oriented to find the candidates based on its kinematic properties, and then their temperatures, metallicities and chemical abundances are compared with those of our Sun.
+In our study, we extend the selection criteria by adding the surface gravity and the age of the star.
+These two new parameters represent a fine-tuning of the search process for twin candidates.
+We used the age of the OCs reported in Table \ref{tab:oc_reduced_list} and the data obtained from the set of MESA simulations initiated with different $\omegaini$, where $\omegaini = \oomegac$, $\Omega$ is the angular velocity of the star at the stellar surface, and $\omegac$ is the surface velocity at the equator of a rotating star where the centrifugal force balances Newtonian gravity.
+The age derived from these simulations provides a crucial parameter for our analysis.
+This information enables us to determine the remaining simulation parameters, which we subsequently use for comparison with the values associated with stars belonging to the short-listed OCs.\par
+
+\begin{table}
+	\centering
+	\begin{tabular}{ll} 
+		\hline
+		Stellar Parameter & Selection Interval\\
+		\hline
+        OC Age & $\pm0.1\%$ \, for age < 1 Ga \\
+        & $\pm0.15\%$ \, for age $\geq$ 1 Ga \\
+        $\teff$ & $\pm 50.0 \, \Kelvin$\\
+        $\gsurf$ & $\pm 0.05 \, \dex$\\
+        $\feh$ & $\pm 0.05 \, \dex$\\
+		\hline
+	\end{tabular}
+    \caption{Selection parameters and intervals applied during the triage process over the OC's components.}
+    \label{tab:sel_params}
+\end{table}
+
+From approximately 900 identified stars, we select those whose $\feh$, $\teff$, and $\gsurf$ are in the correspondent intervals defined in Table \ref{tab:sel_params} and according to the following classification process: based on the values given by our simulations for the parameters listed in Table \ref{tab:sel_params}, we proceed to select from the different OCs those stars that have values within the established ranges.
+Specifically, for the estimated age of a given cluster, we establish an interval of $\pm$0.1 for young OCs (age < 1 Ga) and $\pm$0.15 for older counterparts (age >= 1 Ga) \citep[refer to][for comprehensive reference values]{Cantat-Gaudin2020}.
+This age-dependent interval serves as the basis for selecting simulation time steps that align with the lower and upper age limits of each cluster.
 Once these two time steps are found, the associated limit values for $\teff$, $\gsurf$ and $\feh$ are gathered.
-With these value ranges we proceed then to select in each of the OCs those stars that have values for these same parameters within them.
-In this way, we finally obtain the candidate stars for relevant for being compared with the simulation results.\par
-
-Tables \ref{tab:oc_m67} and \ref{tab:oc_ngc2516} summarize the subset of components of OC M67 and NGC2516 selected during the simulations for models initialised with $\oomegac$ = 0.14.
-Each of the tables lists the GES identifier of the star and the $\teff$, $\gsurf$, $\feh$, A(Li) and estimated age ranges used in the simulation.
-
+With these value ranges, we proceed then to select in each of the OCs those stars that have values for these same parameters within them.
+In this way, we finally obtain the candidate stars to be compared with the simulation results.\par
+
+\begin{table*}
+	\centering
+	\begin{tabular}{l l l l l l l} 
+		\hline
+            ObjectId & $\teff$(K) & $\gsurf$(dex) & FeH(dex) & ALi(dex) & eALi(dex) & Age(Gyr)\\
+		\hline
+            8510914-1157003 & 5843 & 4.42 & -0.03 & 1.8 & 0.08 & 3.981\\ 
+            8510991-1146169 & 5797 & 4.4 & -0.01 & 1.65 & 0.08 & 3.981\\ 
+            8512177-1144050 & 5818 & 4.46 & 0 & 1.99 & 0.06 & 3.981\\ 
+            8513941-1200571 & 5792 & 4.41 & -0.02 & 1.49 & 0.13 & 3.981\\ 
+            8515559-1148381 & 5794 & 4.43 & -0.04 & 1.56 & 0.1 & 3.981\\ 
+            8520350-1147480 & 5875 & 4.47 & -0.03 & 1.21 & 0.24 & 3.981\\ 
+            \hline
+	\end{tabular}
+ 	\caption{M67 - Filter setup: GES star identifier, $\teff$(K) [5787.4892, 5887.6823], $\gsurf$(dex) [4.3979, 4.4977], $\feh$(dex) [-0.05, 0.05], Age(Ga) [3.977, 3.985]}
+	\label{tab:oc_m67}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{l l l l l l l} 
+		\hline
+            ObjectId & $\teff$(K) & $\gsurf$(dex) & FeH(dex) & ALi(dex) & eALi(dex) & Age(Gyr)\\
+		\hline
+            7571111-6048156 & 5084 & 4.54 & 0 & 2.29 & 0.08 & 0.24\\ 
+            7595031-6044149 & 5108 & 4.52 & -0.05 & 1.92 & 0.08 & 0.24\\ 
+            7595280-6032498 & 5093 & 4.49 & -0.03 & 2.29 & 0.08 & 0.24\\ 
+            \hline
+	\end{tabular}
+ 	\caption{NGC2516 - Filter setup: GES star identifier, $\teff$(K) [5010.0324, 5111.0095], $\gsurf$(dex) [4.4405, 4.5405], $\feh$(dex) [-0.05, 0.05], Age(Ga) [0.23964, 0.24036]}\label{tab:oc_ngc2516}
+
+\end{table*}
+
+The Tables \ref{tab:oc_m67} and \ref{tab:oc_ngc2516} summarize the subset of components of the OC's M67 and NGC2516 selected during the simulations for models initialised with $\omegaini = 0.14$.
+Each of the tables lists the GES identifier of the star and $\teff$, $\gsurf$, $\feh$, A(Li) and estimated age ranges used in the simulation.
+
+\section{The modelling} \label{modelling}
 \subsection{Modelling magnetic intensity} \label{mod_mbi}
-In our models we make the case that the magnetic field strength should vary over the lifetime of the star rather than remain fixed throughout its evolution.
-For this we rely on a formalism \citep{Gallet2013} that allows us to calculate a variable magnetic field strength.
-This model adopts a dynamo-type prescription that allows a relationship to be established between the $\Omega$ , $\teff$, and the strength of the magnetic field ($B$), as well as with the mass-loss ($\mwind$) rate caused by the stellar wind.
+Our modeling approach is based on variations of the magnetic field strength throughout a star's evolution, following the formalism introduced by \citet{Gallet2013}.
+The adopted model employs a dynamo-type prescription, establishing relationships among the star's angular velocity ($\Omega$), effective temperature ($\teff$), magnetic field strength ($B$), and the mass-loss rate ($\mwind$) induced by the stellar wind.
 We start by enumerating the most relevant aspects and assumptions made in modelling the evolution of magnetic field intensity.
-\\par
+\par
 
@@
 where $f_*$ is the filling factor, i.e.
-the fraction of the star surface which is magnetized, $B_*$ is the strength of the magnetic filed, $\Omega_*$ is the star angular velocity at stellar surface, and $b$ is the dynamo exponent \citep{Gallet2013}.
-The multiplication of $f_*$ by $B_*$ will allow us to obtain the magnetic field strength.
+the fraction of the star surface which is magnetized, $B_*$ is the strength of the magnetic field, $\Omega_*$ is the star angular velocity at stellar surface, and $b$ is the dynamo exponent \citep{Gallet2013}.
+The multiplication of $f_*$ by $B_*$ will allow us to obtain the mean surface magnetic field.
 \par
 
-According to this approach, the magnetic field strength will vary proportionally to the effective temperature ($\teff$) of the star as follows: 
+According to this approach, the magnetic field strength will vary proportionally to $\teff$ as follows: 
 \begin{ceqn}
@@
 \end{ceqn}
-$B_*$ is proportional to the equipartition magnetic field strength, being $\Beq$ is defined as 
+where $B_*$ is proportional to the equipartition magnetic field strength, and being $\Beq$ defined as 
 \begin{ceqn}
@@
 
-As indicated in \cite{Cranmer2011} $\mu$ can be estimated using the OPAL plasma equations of state\footnote{\url{https://opalopacity.llnl.gov/EOS_2005/}},
+As indicated in \citet{Cranmer2011}, $\mu$ can be estimated using the OPAL plasma equations of state\footnote{\url{https://opalopacity.llnl.gov/EOS_2005/}},
 \begin{ceqn}
@@
 \end{ceqn}
-Once we are in a position to calculate $\Beq$, we can calculate $B_*$.\par
-
-On the other hand, \cite{Cranmer2011} found that the magnetic field strength is mainly influenced by $f_*$ rather than by the rotation period of the star.
-That is, variations in the star's angular velocity do not significantly alter its magnetic field strength.
+Finally, we calculate both $\Beq$ and $B_*$.\par
+
+On the other hand, \citet{Cranmer2011} found that the surface mean magnetic field strength is mainly influenced by $f_*$ rather than by the rotation period of the star.
+That is, variations in the star's angular velocity do not significantly alter its surface mean magnetic field strength.
 In addition, $f_*$ has a strong dependence on the Rossby number ($\rossby$).
@@
 \end{ceqn}
-\cite{Cranmer2011} Crammer provides two different settings for $f_*$ which define respectively the upper and lower limits for the mean surface magnetic field strength ($f_*B_*$),
+\citet{Cranmer2011} provides two different settings for $f_*$ which define respectively the upper and lower limits for the mean surface magnetic field strength ($f_*B_*$),
 
@@
 
-Finally, we are able to calculate the filling factor for a star by applying the adjustment made by \cite{Gallet2013} which more closely reproduces the average filling factor of the present Sun ($\fsun$ = 0.001-0.01),
+Finally, we are able to calculate the filling factor for a star by applying the adjustment made by \citet{Gallet2013} which more closely reproduces the average filling factor of the present Sun $\fsun$ ([0.001-0.01], see \citep{Cranmer2011}),
 \begin{ceqn}
@@
 	\includegraphics[clip,width=\columnwidth]{figures/paper2/mag_field_limits_var_vel_g3.pdf}
-    \caption{The evolution of magnetic field intensity and its upper and lower limits, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
-The model include PMS rotation with $\oomegac$ = 0.1425.
+    \caption{The evolution of magnetic field intensity and its upper and lower limits, as a function of time and $\omegaini$ for a 1 $\msun$ model.
+The model include rotation with $\omegaini$ = 0.1425.
 The solid lines represent the magnetic field strength, while the dotted lines represent the angular evolution of the star.
@@
 With this formalism we are able to calculate the mean magnetic field as a time-dependent function of $\rho$, $\teff$, and $\Omega$.
-The time values for these three variables will be obtained from the different simulations carried out with the help of MESA.
+The time values for these three variables will be obtained from the different simulations carried out with MESA.
 
@@
 Once we have characterised how the magnetic field strength evolves, we need to model how the loss of angular momentum occurs.
-In \cite{Navarro2020} we use a model applicable to magnetic fields of non-variable intensity and of few Gauss ($1.0\,\Gauss \leq B \leq 10.0\,\Gauss$).
-As shown in Figure \ref{fig:mag_field_limits_var_vel_g3}, the simulations yield magnetic field values that show maximum intensities of around $1.0K\,\Gauss$, making the use of the model employed previously non viable.
-\par
-
-Let's start by enumerating the most relevant aspects and assumptions made in modelling the evolution of rotation, magnetic braking and angular momentum.
+As the star undergoes contraction during the pre-main-sequence (PMS) phase towards the Zero-Age Main Sequence (ZAMS), its angular velocity increases, leading to a gradual augmentation in the mean magnetic field.
+In our analysis, we observed intensities that reached $1.0,\mathrm{kG}$, as depicted in Figure \ref{fig:mag_field_limits_var_vel_g3}.
+It then enters a saturation stage (although with a downward trend) up to $1.0$ Ga, and finally declines sharply as a result of the star's slower angular velocity.
+In contrast, our previous work in \citet{Caballero2020} employed a model suited for magnetic fields of non-variable intensity and ranging from a few Gauss ($1.0,\mathrm{G} \leq B \leq 10.0,\mathrm{G}$).
+The significantly higher intensities required in the present formalism render the previously employed model inadequate.\par
+
+Let us start by enumerating the most relevant aspects and assumptions made in modeling the evolution of rotation, magnetic braking and angular momentum.
 If we assume a spherical outflow, the torque applied by a magnetically-coupled stellar winds $\torquewind$ is dictated by,
@@
 where $\mwind$ is the mass loss rate, and $\ralfven$ the averaged value of the Alfvén radius.\par
-Stars with similar initial masses but different mass loss ($\Dot{M}$) ratios will end up evolving very differently.
+Stars with similar initial masses but different mass loss ($\Dot{M}$) rates will end up evolving very differently.
 The ionized particles carried by the solar wind not only contribute to the mass loss but also to the loss of kinetic energy that is deposited in the interstellar medium.
@@
 \begin{ceqn}
-\begin{equation}
-\vesc = \sqrt{\frac{2\,G\,\mstar}{\rstar}} \label{eq:vesc}
-\end{equation}
+\begin{align}
+\Bp &= f_*B_* \label{eq:bp}\\
+\vesc &= \sqrt{\frac{2\,G\,\mstar}{\rstar}} \label{eq:vesc}
+\end{align}
 \end{ceqn}
@@
 \end{ceqn}
-where $\mwind$ is the mass loss rate. This expression is implemented in MESA based on values directly exposed during the simulations.
-
-\subsection{Modelling MLT $\alpha$} \label{mod_mltalpha}
-The mixing-length theory (MLT) introduced by Böhm-Vitense (BV) has been commonly adopted to model stellar convection in stellar evolutionary codes, including MESA.
-The most important free parameter in this theory is the mixing length $l$ which is defined as
-\begin{ceqn}
-\begin{equation}
- l = \alpha H_p\label{eq:mixlength}
-\end{equation}
-\end{ceqn}
-where $H_p$ is the pressure scale height, and $\alpha$ a free parameter which is determined in advance and keep fixed during the stellar evolution simulations.\par
-
- It became evident that the influence of the free, relatively arbitrary, parameters associated with MLT significantly conditioned the evolution of the Li abundance.
-Moreover, there is no strong arguments which suggest that the mixing-length parameter is the same in all stars and at all evolutionary phases \citep{Pasetto2014}.
-In \cite{Navarro2020} we showed how an alternative MLT parameterization could produce results in line with the observations.
-Adjustment of MLT over-shooting and $\amlt$ free parameter seems to be required for explaining Li abundances in OC's.\par
-
-In \cite{Sonoi2018}, the authors introduced a way for calibrating $\amlt$ for solar-like stars in which its value increases with increasing $\gsurf$ or decreasing $\teff$,
+where $\mwind$ is the mass loss rate. This expression is implemented in MESA on the basis of values directly exposed during the simulations.
+
+\subsection{Modelling adaptive MLT $\alpha$} \label{mod_mltalpha}
+The mixing-length theory (MLT) introduced by \citet{BohmVit58} has been commonly adopted to model stellar convection in stellar evolutionary codes, including MESA.
+The most prominent parameter in this theory is the mixing length $l$ which is defined as
+\begin{ceqn}
+\begin{equation}
+ l = \amlt H_p\label{eq:mixlength}
+\end{equation}
+\end{ceqn}
+where $H_p$ is the pressure scale height, and $\amlt$ a free parameter which is determined in advance and normally keep fixed during the stellar evolution simulations.\par
+
+ It became evident that the influence of the free parameter associated with the MLT significantly conditioned the evolution of the Li abundance.
+Moreover, there are no strong arguments that suggest that the mixing-length parameter is the same in all stars and at all evolutionary phases \citep{Pasetto2014}.
+In \citet{Caballero2020} the authors showed how an alternative mixing-length (ML) parameterization could produce results in line with the observations.
+Adjustment of ML over-shooting and $\amlt$ free parameter seems to be required for explaining Li abundances in OC's.\par
+
+In \citet{Sonoi2018}, the authors introduced a way for calibrating $\amlt$ for solar-like stars in which its value increases with increasing $\gsurf$ or decreasing $\teff$.
+They calibrated the values of $\amlt$ for several simulated 3D models using the CO$^5$BOLD code \citep{Freytag2011} and transferred them to the 1D models developed by \citet{Ludwig1998}.
+The parameterization they obatained was:
 
@@
 \end{ceqn}
-and $a_i$ are the resulting coefficients of the fitting function \ref{eq:alpha_ml} to the calibrated $\alpha$ values for the MLT-BV convection model \citep{Sonoi2018}: $a_0=1.790295$, $a_1=-0.14954$, $a_2=0.069574$, $a_3=-0.00829$, $a_4=0.013165$, $a_5=0.080333$, $a_6=-0.03306$.\par
-
-\section{Results} \label{sec_3}
-
-\subsection{Grid Models} \label{sec_grid}
-In the current work we have replaced two pre-set parameters (alpha and intensity) for the different simulations of our model.
-Not only have we eliminated the need to assign a value to them, but we allow their values to vary as the simulations evolve over time.
-The single free parameter of our implementation is $\Omega$.
-The numerical simulations traced the rotational history and $A(\isotope[7]{Li})$ of a $1\, \msun$ star for a variety of initial values for $\Omega$ (see also Table \ref{tab:phy_mesa}).\par
+and $a_i$ are the resulting coefficients of the fitting function to the calibrated $\amlt$ values for the MLT convection model \citep{Sonoi2018}: $a_0=1.790295$, $a_1=-0.14954$, $a_2=0.069574$, $a_3=-0.00829$, $a_4=0.013165$, $a_5=0.080333$, $a_6=-0.03306$.\par
+
+\subsection{Grid of Models} \label{sec_grid}
+Following \citet{Caballero2020} we restrict our simulations to solar-type stars with a mass of $1\,\msun$.
+The evolutionary models of stellar interiors incorporate rotation effects, account for the magnetic braking on stellar evolution, and track the rotational history and the surface abundance of $A(\isotope[7]{Li})$ for various initial angular velocities ($\omegaini$).
+The main differences lie in the parameterization of new model: we have put the focus on reducing the number of free parameters involved in the simulations.
+A pivotal innovation lies in the introduction of a variable magnetic field strength ($B$) and a variable $\amlt$ parameter.
+As for the treatment of convective overshoot mixing, a departure from the previous model is evident; we deliberately deactivate this feature in our current model.
+This isolates the impact of a variable $\amlt$ on A(Li) from the effects associated with convective overshooting.
+Notably, convective overshoot mixing is commonly addressed through ad hoc extensions of the Mixing Length Theory (MLT), introducing an additional free parameter.
+The Table \ref{tab:phy_mesa} enumerates the similarities and differences in the configuration of the models used in \citet{Caballero2020} and in this paper.\par
+
+For a comprehensive overview of the similarities and differences between the configurations of the models used in \citet{Caballero2020} and the present study, refer to Table \ref{tab:phy_mesa}.
 
 \begin{table}
+\begin{threeparttable}
 	\centering
-	\caption{Summary of adopted physics in MESA \citep[based on][]{Choi2016,Navarro2020}.}
-	\label{tab:phy_mesa}
 	\begin{tabular}{ll} 
@@
 		Equation of State & OPAL+SCVH+MacDonald+HELM+PC\\
-		Opacity & OPAL Type I for log T $\geq$ 4 \\ & Ferguson for logT $\leq$ 4\\
+		Opacity & OPAL Type I for log T $\geq$ 4 \\ & Ferguson for logT $<$ 4\\
 		Reaction Rates & JINA REACLIB\\
@@
 		Diffusion & Track \isotope[1]{H}, \isotope[2]{He}, \isotope[7]{Li}, \isotope[7]{Be}\\
-		Rotation & Differential rotation at PMS \& MS\\
-		Convection & MLT + Ledoux, variable $\alpha_{{\rm MLT}}$\\
-		Overshoot deactivated & $f_{{\rm ov,core}}=0.0$, \& $f_{{\rm ov,sh}}=0.0$\\
+		Rotation Schema & Differential rotation at PMS \& MS\\ & Include SH\tnote{1}  , ES\tnote{2}  , GSF\tnote{3}  , SSI\tnote{4}  , DSI\tnote{5}\\
+		Thermohaline & $\alpha_{{\rm th}}=666$\\
+		\textbf{Convection} & $\alpha_{{\rm MLT}}$ variable depending on $\teff$\\ & \& $\gsurf$ + Ledoux\\
             Semiconvection & $\alpha_{{\rm sc}}=0.1$\\
-		Thermohaline & $\alpha_{{\rm th}}=666$\\
-		Rotational Mixing & Include SH, ES, GSF, SSI \& DSI\\
-            Mixing Length Theory & $\alpha_{{\rm MLT}}$ variable depending on $\teff$\\ & \& $\gsurf$\\
-		Magnetic Effects & Magnetic braking based on idealized \\ & dipole field\\
-		Magnetic Field & B(G) variable, depending on $\rho$, $\teff$ \&  $\Omega$\\\
-		Mass Loss & activated, $\Dot{M}_{{\rm max}} = 10^{-3} \: \msun \: yr^{-1}$\\
-		Angular Moment Loss & activated, $\Dot{J} = \Omega_* \; \mwind \; \ralfven^2$\\
+		\textbf{Overshoot} & $f_{{\rm ov,core}}=0.0$, \& $f_{{\rm ov,sh}}=0.0$\\
+		\textbf{Magnetic Field} & B(G) variable, depending on $\rho$, $\teff$ \&  $\Omega$\\
+		Mass Loss & $\Dot{M}_{{\rm max}} = 10^{-3} \: \msun \: yr^{-1}$\\
+		Angular Moment Loss & $\Dot{J} = \Omega_* \; \mwind \; \ralfven^2$\\
 		\hline
 	\end{tabular}
+    \begin{tablenotes}\footnotesize
+        \item (1) Solberg-Hoiland, (2) Eddington–Sweet
+        \item (3) Goldreich–Schubert–Fricke, (4) Secular Shear Instability
+        \item (5) Dynamical Shear Instability
+    \end{tablenotes}
+    \end{threeparttable}
+    \caption{Summary of adopted physics in MESA \citep[based on][]{Choi2016,Caballero2020}. Highlighted in bold are the parameters with different configuration from referenced works.}
+	\label{tab:phy_mesa}
+
 \end{table}
 
-Most of the configuration used in the simulations remain the same as in our previous work, with the exception of the overshooting (deactivated), the $\amlt$ pre-set value, and the formalism used for calculating $\Dot{J}$ (see \cite{Navarro2020} for further details).\par
-
-The models continue to including rotation during the PMS.
-We adopted a simple and pragmatic theoretical approach to establish when the star is reaching the ZAMS phase: the simultaneous existence of an extensive convective layer and a radiative core.
-From this moment on, the MB routine was activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participated in the star AML.
-This time, the restriction of a magnetic field with a constant intensity throughout the evolution of the star until it reached the TAMS is removed in favor of variable one.
-Something similar happens with $\amlt$ which indeed is  a key factor to determine the near-surface structure of stellar models.
-
-In this new paper we continue to focusing on the indirect role played by the MB on the Li destruction.
-The main objective continues to be to check how MB could contribute to explain the evolution of Li in the Sun and other solar-type stars.
-We computed the evolution of $1\,\msun$ stellar models at solar initial metallicity with $\oomegac$ varying between $0.12$ and $0.1425$.
-The results are then compared with the data collected by the Gaia mission and GES for the stars belonging to the OC's referenced in Table \ref{tab:oc_list}.\par
+The models include rotation since the PMS.
+We computed the evolution of $1\,\msun$ stellar models at solar initial metallicity with $\omegaini = \oomegac$ varying between $0.12$ and $0.1425$.
+We have adopted the following theoretical approach to establish when the star is reaching the ZAMS: as both the central \isotope[1]{H} mass fraction has been reduced by $\Delta( \isotope[1]{H})= 0.0001$ from its initial value, and the star's luminosity ($L$) is almost fully ($L \geq 0.99$) generated by the nuclear reactions of \isotope[1]{H} occur simultaneously.
+The results are then compared with the stellar ages from stars in OCs collected by the Gaia mission and GES and referenced in Table \ref{tab:oc_reduced_list}.\par
 
@@
 For large $\ralfven$ values, the star undergoes a significant deceleration.
-The value of $\ralfven$ depends on the following varying parameters: magnetic field ($B$) intensity, $\teff$ and$\Omega$.
+The value of $\ralfven$ depends on the following varying parameters: magnetic field ($B$) intensity, $\teff$ and $\Omega$.
 With regards to $\Dot{M}$ and as reported in Table \ref{tab:phy_mesa}, the empirical formula developed by Reimers \citep{Reimers1975} for stars in the asymptotic giant branch (AGB) was used for calculating the mass-loss.
@@
 After that, an AM value is assigned to each cell $k$ ($J_k$).
-At this point, our MB turns on, modifying $J_k$.
+At this point, our MB routine will turn on if the star has both an extensive convective layer, and a radiative core.
+From this moment on, the MB routine was activated, acting as an additional mechanism to those existing in the MESA evolutionary code that participated in the star AML and modifying $J_k$.
 This was done by providing an additional contribution ($\Dot{J}_{k}$).
@@
 
-\subsection{Li evolution with MB}
-Figure \ref{fig:li_var_vel_var_g_3} shows the temporal evolution of surface Li abundance for several 1 $\msun$ models.
-Those models were initialized with different rotational velocities and took into consideration the effects of MB caused by a variable magnetic field which its intensity evolution is shown in Figure \ref{fig:mag_field_var_vel_g3}.
-If we compare it with Figure \ref{fig:li_var_vel_0g} in which the effects of MB were neglected, we notice how the profiles of Li abundance were altered during PMS and MS.
-During the PMS we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see Eq.~\ref{eq:j_dot}) depends directly on the mass loss rate (see Figure \ref{fig:mdot_var_vel_g3}) and the $\Omega$ (see Figure \ref{fig:rot_vel_var_vel_var_g3}) evolution.
+\section{Results} \label{sec_3}
+There are evidences that advocate for a strong established relationship between Li destruction and stellar rotation, so that the higher the angular velocity, the more Li destruction occurs (see \citet{Bouvier2018, Caballero2020}).
+The effect of the MB on the time evolution of A(Li) is given by its ability to remove AM from the star, and thus influence its rotational velocity, causing the star to rotate more slowly and to destroy less Li.\par
+
+In Figure \ref{fig:li_var_vel_var_g_3} it is shown the temporal evolution of surface Li abundance for several 1 $\msun$ models.
+Those models were initialized with different rotational velocities and took into consideration the effects of MB caused by a variable magnetic field.
+If we compare it with Figure \ref{fig:li_var_vel_0g} in which the effects of MB were neglected, we notice how the profiles of Li abundance were altered across PMS and MS.
+During the PMS we can describe the effect as modest, somewhat expected and in line with the fact that the AML caused by MB (see Eq.~\ref{eq:j_dot}) depends directly on the the $\Omega$ evolution and mass loss rate.
 If we take into account that for solar-type stars the models predict a modest total mass loss rate, that value is even much lower in this phase.
-It is in the approach phase to the TAMS that the angular velocity reaches its maximum.
+It is in the approach phase to the ZAMS that the angular velocity reaches its maximum.
 This, according to our model, plays a crucial role in both the mass loss and the magnetic field strength.
-The higher the $\Omega$, the higher the $\Dot{M}$, and therefore the higher the magnetic field strength.
+The higher the $\Omega$, the higher the $\Dot{M}$  (see Figure \ref{fig:mdot_var_vel_g3}), and the higher the magnetic field strength (see Figure \ref{fig:mag_field_var_vel_g3}).
 These effects combine and lead to an accentuation of the magnetic braking effect.
 As a consequence, there is a slowing down of Li destruction once the models enter the MS.
-The results of our simulations for those models respectively initialized to $\oomegac$ = 0.14 and 0.1425 reproduce A(Li) in line with that of the Sun surface \isotope[7]{Li} abundance  (1.1 $\pm$ 0.1 dex).
-The latter gives a value of 1.133 dex, which would represent a deviation of about 3\% from the nominal value.\par
-
-\begin{figure}
-	\includegraphics[trim = 25mm 10mm 15mm 10mm, clip, width=\columnwidth]{figures/paper2/li_var_vel_0_0g_0.pdf}
-    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
-The solid black line represents the reference model according to \citet{Choi2016}.
-The rest of lines are models which include PMS rotation with $\oomegac$ between 0.0084 and 0.0336, respectively.
-The purple star and square are surface Li abundances for the present-day Sun \citep{Asplund2009} and the average for the Pleiades cluster \citep{Sestito2005} respectively.
-The dashed vertical line makes reference to the ZAMS.
-(Figure taken from \citet{Navarro2020}.)}
+The results of our simulations for those models respectively initialized to $\omegaini$ = 0.14 and 0.1425 reproduce A(Li) in line with that of the Sun surface \isotope[7]{Li} abundance  (1.1 $\pm$ 0.1 dex).
+The latter gives a value of 1.133 dex, which would represent a deviation of about 3\% from the nominal value (see Figure \ref{fig:li_var_vel_var_g_3}).\par
+
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/li_var_vel_var_g_3.pdf}
+    \caption{Evolution of the relative surface abundance of \isotope[7]{Li} compared to \isotope[1]{H} is shown for various 1 $\msun$ models.
+These include a magnetic field with varying intensity, initial rotation rates ranging from 0.12 to 0.1425 $\omegaini$, and MB.
+The purple star represents the present-day Sun's surface Li abundance \citep{Asplund2009}.
+The other colored dots represent surface \isotope[7]{Li} abundances for stars with parameters within the specified selection intervals, corresponding to the evolution curve of the same color.
+The dashed vertical line indicates the Zero Age Main Sequence (ZAMS).}
+    \label{fig:li_var_vel_var_g_3}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[clip, width=\columnwidth]{figures/paper2/li_var_vel_0_0g_0.pdf}
+    \caption{Ratio of surface \isotope[7]{Li} abundance to \isotope[1]{H} as a function of time is depicted for various 1 $\msun$ models without magnetic field.
+The reference model, as proposed by \citet{Choi2016}, is represented by the solid black line.
+Additional models with initial rotation rates ranging from 0.0084 to 0.0336 are illustrated by the remaining lines.
+The surface Li abundances for the present-day Sun, indicated by a purple star, are based on the study by \citet{Asplund2009}.
+The dashed vertical line indicates the Zero Age Main Sequence (ZAMS).}
     \label{fig:li_var_vel_0g}
@@
 \begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/mdot_var_vel_g3.pdf}
+    \caption{The evolution of mass loss $\Dot{M}$ as a function of time for several 1 $\msun$ models.
+The models include variable magnetic field intensity and initial rotation with $\omegaini$ between 0.12 and 0.1425.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:mdot_var_vel_g3}
+\end{figure}
+
+\begin{figure}
 	\includegraphics[clip,width=\columnwidth]{figures/paper2/mag_field_var_vel_g3.pdf}
-    \caption{The evolution of magnetic field intensity, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
-The models include PMS rotation with $\oomegac$ between 0.12 and 0.1425.
+    \caption{The evolution of magnetic field intensity, as a function of time and $\omegaini$ for several 1 $\msun$ models.
+The models include initial rotation with $\omegaini$ between 0.12 and 0.1425.
 The solid lines represent the magnetic field strength, while the dotted lines represent the angular evolution of the star.
@@
 
-In Figure \ref{fig:rot_vel_var_vel_var_g3} we depict the rotation profiles for the surface of the stars and for the bottom of the convective envelop.
+In Figure \ref{fig:rot_vel_var_vel_var_g3} we depict the rotation profiles of stars for the surface and for the bottom of the convective envelop.
 It shows how the angular velocity increases in value as the star decreases in radius along the PMS.
 During this stage of its evolution, the star has a nearly convective structure, as can be seen in Figure \ref{fig:cz_var_vel_var_g3}.
-As it approaches the ZAMS at $\approx 10^6$ years, the convective zone of the star begins to shrink in size and the core of the star reaches the temperature, pressure and density conditions necessary to develop a radiative core.
+As it approaches an age of $\approx 10^6$ years, the convective zone of the star begins to shrink in size and the core of the star reaches the temperature, pressure and density conditions necessary to develop a radiative core.
 In turn, the difference in the angular velocities of the different models becomes more evident, being higher in those that were initialised with a higher initial angular velocity.
@@
 The combined effects of the increases in effective temperature, angular velocity and mass loss cause the magnetic field strength to reach its maximum in the approach phase of the ZAMS (see Figure \ref{fig:mag_field_var_vel_g3}).
-We observed that the models with lower angular velocity generally ended up exhibiting higher values for the Li abundance on the surface (see Figures~ \ref{fig:li_var_vel_var_g_3}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).
-For these same simulated models initialized to $\oomegac$ = 0.14 and 0.1425, the values obtained for the $\Omega$ and the magnetic field strength are far from the measurements obtained for the Sun.
-$\osun$ is 2km/s with a minimum error of 3m/s, and the mean magnetic field strength is 1G.
-For the simulation with $\oomegac$=0.1425, we obtain a rotational velocity of 4.72km/s, which represents a 235\% deviation, and for the average magnetic field strength we have 36.9G, far away from the reference value.\par
-
-\begin{figure}
-	\includegraphics[clip,width=\columnwidth]{figures/paper2/li_var_vel_var_g_3.pdf}
-    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
-The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425, respectively and MB.
-The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
-The rest of the dots in different colours represent surface \isotope[7]{Li} abundances of stars with similar parameters to the evolution curve represented with the same colour.The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:li_var_vel_var_g_3}
-\end{figure}
-
-As described in section \ref{mod_mb}, the MB routine distributed the total amount of AML calculated according to Eq.~\ref{eq:k_jdot} among the different layers that composed the CZ.
+We observed that the models with lower angular velocity generally ended up exhibiting higher values for the Li abundance on the surface (see Figures~ \ref{fig:li_var_vel_var_g_3}, \ref{fig:grid_li_var_vel} \& \ref{fig:grid_li_var_g}).\par
+
+For the models in which we obtain values of A(Li) in line with those of the Sun ($\omegaini$ = 0.14 and 0.1425), we have on the other hand measurements for the $\Omega$ and $B$ which are not are not in tune with the respective solar ones.
+The Sun surface rotational velocity is 2km/s with a minimum error of 3m/s, and the mean magnetic field strength is 1G.
+For the simulation with $\omegaini$=0.1425, we obtain a rotational velocity at equator of 4.72 km/s.
+Although it is a measure on the same order of magnitude, it represents a 235\% deviation.
+For the average magnetic field strength we have 36.9G, far away from the reference value.\par
+
+As described in section \ref{mod_mb}, the MB routine, which is activated when the star develops a radiative core and a convective zone above it (see Figure \ref{fig:mb_act_var_vel_g3}), distributed the total amount of AML calculated according to Eq.~\ref{eq:k_jdot} among the different layers that composed the CZ.
 In Figure \ref{fig:cz_var_vel_var_g3} we can observe the evolution of the most external CZ normalized with respect to the radius of the star for several 1 $\msun$ models.
-The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425.
 In accordance with the established models of stellar evolution, in a solar-type star the CZ covers practically all of it for a large part of the PMS.
 From the ZAMS on-wards, its size starts to decrease gradually in a first phase and then becomes more pronounced at around $4.0x10^8$ years of age.
-This change in the size of the CZ is due to magnetic braking.
-As can be seen in Figure \ref{fig:rot_vel_var_vel_var_g3}, the star, after reaching its maximum angular velocity when passing the ZAMS, starts to decelerate due to magnetic braking.
-In our models the magnetic braking routine is activated when the star develops a radiative core and the existence of a convective zone above it (see Figure \ref{fig:mb_act_var_vel_g3}).
+As can be seen in Figure \ref{fig:rot_vel_var_vel_var_g3}, the star, after reaching its maximum angular velocity when passing the ZAMS, starts to decelerate due to magnetic braking.\par
+
 It is the presence of the MB that prevents the star from further increasing its angular velocity.
@@
 \begin{figure}
-	\includegraphics[clip,width=\columnwidth]{figures/paper2/cz_var_vel_var_g3.pdf}
-    \caption{The evolution of radiative and convective zones size as a function of time for several 1 $\msun$ models.
-The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425.
-The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:cz_var_vel_var_g3}
-\end{figure}
-
-\begin{figure}
 	\includegraphics[clip,width=\columnwidth]{figures/paper2/rot_vel_var_vel_var_g3.pdf}
     \caption{The evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
-The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425, respectively and MB.
+The models include a magnetic field with variable intensity, initial rotation with $\omegaini$ between 0.12 and 0.1425, respectively and MB.
 The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
@@
     \caption{The activation of the magnetic braking routine as a function of the presence of a radiative core.
-The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425.
 The solid lines signal the magnetic braking routine activation (on) and deactivation (off).
@@
 
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/cz_var_vel_var_g3.pdf}
+    \caption{The evolution of the convective zone size as a function of time for several 1 $\msun$ models.
+The size is expressed in terms of $\rsun$.
+The models include a magnetic field with variable intensity, initial rotation with $\omegaini$ between 0.12 and 0.1425.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:cz_var_vel_var_g3}
+\end{figure}
+
 The effects of rotation and magnetic braking can also be seen in the H-R diagram and in the evolution of the $\amlt$.
@@
 The differential rotation between the boundaries of the radiative core and the convective layers leads to mixing effects in the so-called tachocline.
-The effects of this mixing and of its hydrostatic effects are governed by the MLT in which the $\amlt$ plays a major role in it.
-As aforementioned, MLT suffers from the arbitrariness of the value of $\amlt$, and it was necessary to establish its value based on physical constraints occurring in the interior of the star.
+The outcomes of this mixing and its hydrostatic impacts are controlled by the MLT, where $\amlt$ assumes a crucial role.
+As previously stated, the arbitrariness of the $\amlt$ value introduces a level of uncertainty into the MLT.
 Figure \ref{fig:alpha_mlt_var_vel_g3} shows how the $\amlt$ parameter evolves over time.\par
 
+%The effects of this mixing and of its hydrostatic effects are governed by the MLT in which the $\amlt$ plays a major role in it.
+As aforementioned, MLT suffers from the arbitrariness of the value of $\amlt$.
+
 \begin{figure}
 	\includegraphics[clip,width=\columnwidth]{figures/paper2/alpha_mlt_var_vel_g3.pdf}
-    \caption{The evolution of $\amlt$, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
-The models include PMS rotation with $\oomegac$ between 0.12 and 0.1425.
-The dashed vertical line makes reference to the ZAMS.}
+    \caption{The evolution of $\amlt$, as a function of time and $\omegaini$ for several 1 $\msun$ models.
+The models include initial rotation with $\omegaini$ between 0.12 and 0.1425.
+The dashed vertical line makes reference to the ZAMS.
+The purple star and diamond are the $\amlt$ given by \citet{Sonoi2018} and \citet{Samadi2005}, respectively.}
     \label{fig:alpha_mlt_var_vel_g3}
@@
 Additionally, these same centrifugal forces, accentuated in stars with a higher angular velocity, reinforce the gravity darkening \citep[see e.g.
-][]{Gossage2021,Paxton2019,Eggenberger2012} as well.
-The centrifugal force causes the star to shift its mass outwards with respect to the rotation axis, the effect being more pronounced in the equatorial regions than at the poles.
+][]{Eggenberger2012,Paxton2019,Gossage2021} as well.
+The centrifugal force causes the star to shift its mass outward with respect to the rotation axis, the effect being more pronounced in the equatorial regions than at the poles.
 The consequence is that the pressure ($\gsurf$) to which the gas is subjected is lower in the former than in the latter.
-This effect causes those stars that rotate fast enough to appear less dense ($\rho$), less luminous ($L$) and therefore with a lower effective temperature ($\teff$) as they approach the ZAMS (see Figure \ref{fig:hr_var_vel_var_g_z13}).
-If we compare the non-rotating model (black solid line in Figure \ref{fig:hr_var_vel_0g}) with the rotating ones we can recognize that at the end of the PMS, the latter reach the ZAMS with a lower $\teff$ than the former.
+This effect causes stars that rotate fast enough to appear less dense ($\rho$), less luminous ($L$) and therefore with a lower effective temperature ($\teff$) as they approach the ZAMS (see Figure \ref{fig:hr_var_vel_var_g_z13}).
+If we compare the non-rotating model (black solid line in Figure \ref{fig:hr_var_vel_0g}) with the rotating ones, we can recognize that at the end of the PMS, the latter reach the ZAMS with a lower $\teff$ than the former.
 This also has an impact on the evolution of $\amlt$ because, as discussed above, in our model its value depends proportionally on $\teff$ and $\gsurf$.
@@
 	\includegraphics[clip,width=\columnwidth]{figures/paper2/hr_var_vel_var_g_z13.pdf}
-    \caption{Similar to Figure \ref{fig:hr_var_vel_0g} but now showing in detail the combined effects of the gravity darkening and magnetic braking on the evolutionary tracks for several 1 $\msun$ models.
-The models include a magnetic field with variable intensity, PMS rotation with $\oomegac$ between 0.12 and 0.1425.
-The presence of a magnetic field produces hotter stars due to the influence of the magnetic braking on the rotational velocity of the star.
-$\lsun$ = 3.761.}
+    \caption{An example solar 1$\msun$ grid of stellar evolutionary tracks covering a range of angular velocities.
+It shows in detail the combined effects of the gravity darkening and magnetic braking on the evolutionary tracks.
+The models include a magnetic field with variable intensity, initial rotation with $\omegaini$ between 0.12 and 0.1425.
+The presence of a magnetic field produces hotter stars due to the influence of the magnetic braking on the rotational velocity of the star.}
     \label{fig:hr_var_vel_var_g_z13}
@@
 	\includegraphics[trim = 10mm 10mm 15mm 10mm, clip,width=\columnwidth]{figures/paper2/hr_var_vel_0_0g_z10.pdf}
-    \caption{An example solar 1$\msun$ grid of stellar evolutionary track from PMS till TAMS covering a wide range of angular velocities.
+    \caption{Similar to Figure \ref{fig:hr_var_vel_var_g_z13} but now showing stellar evolutionary tracks in the absence (0G) of a magnetic field.
 The rotation is activated in the models in the PMS and those models reach before the ZAMS and at a lower $\teff$ than the non-rotating one (solid black line).
 The luminosity is expressed in terms of $\lsun$.
-(Figure taken from \citet{Navarro2020}.)}
+$\lsun$ = 3.761.
+(Figure taken from \citet{Caballero2020}.)}
     \label{fig:hr_var_vel_0g}
@@
 
-Figures \ref{fig:teff_logg_var_vel_g3} and \ref{fig:teff_logg_var_vel_g_z13} show how $\teff$ and $\gsurf$ behave over the evolution of the models.
-For those with a higher rotation speed we observe that both their temperature and surface gravity are lower than in the slower rotating models, in line with the gravity darkening.
-It is worth noting that in the ZAMS approach phase we observe for the faster rotating model (yellow) that its surface gravity is higher than that of the rest of the slower models.
-This can be explained by looking at Figure \ref{fig:mdot_var_vel_g3}, which shows the mass loss of the star.
-Around the interval $2.5x10^{7}$ and $3.5x10^{7}$ years the mass loss of the fastest model is smaller than that of the rest.
-Since the surface gravity is directly proportional to the amount of mass of the star, if the mass of the star decreases, the surface gravity decreases as well.
-During this period the faster model does so at a lower rate and so we have a higher surface gravity.
-This "anomaly" disappears as soon as the mass loss becomes more pronounced and becomes noticeably larger between $3.5x10^{7}$ and $4.5x10^{7}$ years, shortly before reaching ZAMS.
-In this phase the surface gravity of the fastest rotator is lower than that of the other models.
-We have a similar situation past ZAMS, from around $4.5x10^{7}$ till $11.0x10^{7}$ years.
-In this case the explanation for the higher surface gravity of the model that rotates faster than the others is due to the size of the star, how its radius behaves.
-Figures \ref{fig:lograd_var_vel_g3} and \ref{fig:lograd_var_vel_g_z13} show the time evolution of the radius for the different models.
-In them we can see how for the fastest rotator, in the above-mentioned time interval, its size is smaller.
-As it decreases more than the others, and although it also continues to lose more mass, its $\gsurf$ is greater.
-Let us remember that the radius has an inversely quadratic influence on the value of the $\gsurf$.\par
+Figure \ref{fig:teff_logg_var_vel_g3} shows how $\teff$ and $\gsurf$ behave over the evolution of the models.
+For those with a higher rotation speed, we observe that both their temperature and surface gravity are lower than in the slower rotating models, in line with the gravity darkening.
+It is worth noting that in the ZAMS approach phase we observe for the fastest rotating model (yellow) that its surface gravity is higher than that of the rest of the slower models.
+This can be explained by looking at Figure \ref{fig:lograd_var_vel_g3}, which shows the evolution of the stellar radius for the different models, as a function of time and $\omegaini$.
+Around the interval $2.5x10^{7}$ and $3.5x10^{7}$ Ga (delimited by the cyan lines) and after the ZAMS, from around $5.4x10^{7}$ to $11.2x10^{7}$ Ga (delimited by the magenta lines) the stellar radius of the fastest model is smaller than that of the rest, producing a higher $\gsurf$, and this in turn means less mass loss (see Fig.
+\ref{fig:mdot_var_vel_g3}).
+During this period the fastest model does so at a lower rate and expose a higher surface gravity.
+Let us recall that the stellar radius has an inversely quadratic influence on the value of $\gsurf$.
+This "anomaly" disappears as soon as the stellar radius becomes again bigger for the fastest model.\par
 
@@
 	\includegraphics[clip,width=\columnwidth]{figures/paper2/teff_logg_var_vel_g3.pdf}
-    \caption{The evolution of $\teff$ and $\gsurf$, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
-The models include PMS rotation with $\oomegac$ between 0.12 and 0.1425.
+    \caption{The evolution of $\teff$ and $\gsurf$, as a function of time and $\omegaini$ for several 1 $\msun$ models.
+The models include initial rotation with $\omegaini$ between 0.12 and 0.1425.
+The age intervals [$2.5x10^{7},3.5x10^{7}$] Ga and [$5.4x10^{7},11.2x10^{7}$] Ga, delimited by the cyan and magenta lines respectively, highlight periods in which the fastest model with $\omegaini$=0.1425 exposes a surface gravity higher than that of the rest of the slower ones.
 The purple star is the $\teff$, and the purple square is the $\gsurf$ for the present-day Sun \citep{Gill2012}.
@@
 \begin{figure}
-	\includegraphics[clip,width=\columnwidth]{figures/paper2/teff_logg_var_vel_g_z13.pdf}
-    \caption{Similar to \ref{fig:teff_logg_var_vel_g3} but zooming in on the ZAMS.
-The models with a higher initial rotational velocity reach already the ZAMS with a lower $\teff$ and higher $\gsurf$.}
-    \label{fig:teff_logg_var_vel_g_z13}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[clip,width=\columnwidth]{figures/paper2/mdot_var_vel_g3.pdf}
-    \caption{The evolution of mass loss $\Dot{M}$ as a function of time for several 1 $\msun$ models.
-The models include variable magnetic field intensity and PMS rotation with $\oomegac$ between 0.12 and 0.1425.
-The dashed vertical line makes reference to the ZAMS.}
-    \label{fig:mdot_var_vel_g3}
-\end{figure}
-
-\begin{figure}
 	\includegraphics[clip,width=\columnwidth]{figures/paper2/lograd_var_vel_g3.pdf}
-    \caption{The evolution of the stellar radius, as a function of time and $\oomegac$ for several 1 $\msun$ models and their.
-The models include PMS rotation with $\oomegac$ between 0.12 and 0.1425.
+    \caption{The evolution of the stellar radius, as a function of time and $\omegaini$ for several 1 $\msun$ models and their.
+The models include initial rotation with $\omegaini$ between 0.12 and 0.1425.
 The purple star is the radius for the present-day Sun \citep{Gill2012}.
@@
 
-\begin{figure}
-	\includegraphics[clip,width=\columnwidth]{figures/paper2/lograd_var_vel_g_z13.pdf}
-    \caption{Similar to Figure \ref{fig:lograd_var_vel_g3} but zooming in on the ZAMS. The fastest model exhibits the smallest radius from around $4.5x10^{7}$ till $11.0x10^{7}$ years.}
-    \label{fig:lograd_var_vel_g_z13}
-\end{figure}
-
-\subsection{Gossage et al. (2021) Comparison}
-In \cite{Gossage2021}, the authors also approach the problem of AML from a similar starting point to ours, but with some notable differences.
-In the Table \ref{tab:gassage_vs_navarro} we compare the most notable similarities and differences between the two studies.
-In both works the models consider rotation not only in the MS but also in the PMS, with the difference that in Gossage's work they take into account disk-locking.
-This aspect is not so important in our simulations since in our case we have focused on the A(Li) profiles.
-In contrast, Gossage's work focuses mainly on modelling rotation profiles of stars, which are significantly affected by disk-locking.
-Similarly, the effects of magnetic braking on angular momentum loss are also considered.
-In our case for stars with 1$\msun$ (solar-twins) and in their with 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$.
-Both models also use MLT but in our case the MLT alpha value is not pre-fixed but varies throughout the simulation.
-These two works complement each other as they are based on different formalisms.
-On the one hand, our work is based on the proposal of \cite{Gallet2013} whose most relevant parameters are Ro and the presence of a magnetic field of varying strength (B), and on the other hand Gossage presents two formalisms: \cite{Matt2015} and \cite{Garraffo2018}.
-The former is mainly characterised by $\rossby$ and the $\mstar$, while the latter by $\rossby$ and the complexity of the magnetic field ($\magfc$).
-As far as OCs observations are concerned, in this work we have had access to a larger number of them, and covered a wider range of ages [0.001-6.761] Gyr.
-Finally, both papers simulate their respective models in MESA.\par
+\subsection{Caballero et al. (2020) Comparison}
+In \citet{Caballero2020} we have already addressed the study of the lithium surface abundance in the Sun.
+That previous work laid the foundations for the development of the present.
+In it, we advanced as future lines of research a deeper understanding about how magnetic fields are linked to rotation and how they evolve in time.
+We also showed how by adjusting the value of $\amlt$ we were able to influence the in the Li abundances.
+Therefore, a comparison between the results obtained previously and those of this study puts into perspective the differences and similarities between the two studies, as well as the advances made since then.
+The Table \ref{tab:caballero2023_2020} summarises the main differences between the two works, in line with what is already advanced i.e.
+we incorporate formalisms that allow us to temporally evolve the $B$ and $\amlt$.\par
 
@@
 	\centering
-	\caption{Comparison with \cite{Gossage2021} work.}
-	\label{tab:gassage_vs_navarro}
     \begin{threeparttable}
@@
 		\hline
-		Parameter & This work & \cite{Gossage2021}\\
+		Parameter & This work & \citet{Caballero2020}\\
+		\hline
+		Convection & MLT variable $\alpha_{{\rm MLT}}$ & MLT fixed $\alpha_{{\rm MLT}}=1.82$\\
+            Overshoot & deactivated & $f_{{\rm ov,core}}=0.0160$,\\ & & $f_{{\rm ov,sh}}=0.0174$\\
+		Magnetic Braking & \citet{Gallet2013} & \citet{Ud-Doula2008} \\ Formalism & & \\
+		Magnetic Field & B(G) variable & B(G) pre-set [3.0G-5.0G]\\ Strength & & \\
+            Number of OC's & 64 & 1 \\
+            Source of data & GDR3.0 GES DR5.0 & \citet{Sestito2005} \\
+		\hline
+	\end{tabular}
+    \end{threeparttable}
+    \caption{Comparison with \citet{Caballero2020} work.} \label{tab:caballero2023_2020}
+\end{table}
+
+The Figure \ref{fig:li_var_vel_4_0g4} shows the evolution of Li abundance for $\omegaini$ between 0.0084 and 0.0336 in the presence of a magnetic field with an intensity of 4G.
+We have augmented the original version with the same set of OCs data gathered applied in this new work, so that a consistent comparison between both formalism is possible.
+If we focus our attention on the simulation with $\omegaini$ = 0.0336, the simulation reproduces  A(Li) = 1.024 dex, a value which is in line with both that of the Sun (1.1 $\pm$ 0.1 dex) and the values obtained in Figure \ref{fig:li_var_vel_var_g_3} for the present work (1.133 dex) for $\omegaini$ = 0.14 and 0.1425.
+Although in the latter case the models were initialized at higher $\omegaini$ values than in the former, resulting in $B$ up to 3 orders of magnitude more intensive ($\approx$ 1KG versus 4G) the simulation in both works end up in producing very similar values for both the rotational velocity and the A(Li).
+This is partially explained because a higher magnetic field strength implies an intensive MB effect (refer to Figures \ref{fig:rot_vel_var_vel_var_g3} \& \ref{fig:rot_vel_var_vel_4_0g4}).
+In the former case, the deceleration effect is sharper than in the latter, so it ceases to destroy Li before.\par
+
+However, this effect is not sufficient to effectively counteract the destruction of Li.
+The other decisive element that comes into play is the value of $\amlt$.
+As exemplified in \citet{Caballero2020} its value has an inversely proportional impact on Li destruction, where a lower value resulted in less Li destruction, particularly during PMS.
+In the figure \ref{fig:alpha_mlt_var_vel_g3} we observed the temporal $\amlt$ value evolution.
+Until the latest phase of the PMS, the value $\amlt$ is substantially less than 1.82, the pre-set and invariant value used in our previous work, resulting in less Li destruction and allowing the model to reach the ZAMS with a $\approx$ 29\% dex higher Li reservoir.
+This major Li reserve compensates for the greater destruction induced by the models initialized with a $\omegaini$ higher angular velocity.\par
+
+Similarly, we proceed to show the evolution of the surface rotational velocity in Figure \ref{fig:rot_vel_var_vel_4_0g4} and select the simulation with $\omegaini$ = 0.0336.
+Again, the similar values are explained by the higher efficiency of the MB in the new model.
+Although the simulations reach the ZAMS with rotational velocity values up to 200\% for the models initialised with higher $\omegaini$, the stronger magnetic field produces a more pronounced deceleration for the same time interval.
+This fact is especially noticeable in the time period between the ZAMS and 1 Ga.
+The rotational velocity obtained at the equator of 4.82 km/s is a value quite close to 4.72 km/s, that is, the value simulated for $\omegaini$ = 0.0336 (Fig.~\ref{fig:rot_vel_var_vel_var_g3}).\par
+
+In view of the values obtained, we can draw the conclusion that the new model introduced in the current work is not only congruent with the results of the works on which it is based, but also with those of \citet{Caballero2020}.
+Additionally, this new version exposes the advantage of not having to resort to the use of preset values for $B$ and $\amlt$, and without recurring to activate the overshooting.
+All in all, this represents a significant improvement and supports our line of research aimed to avoid preset values by making them to evolve dynamically with key solar parameters.\par
+
+\begin{figure}
+	\includegraphics[clip, width=\columnwidth]{figures/paper2/li_var_vel_4_0g4.pdf}
+    \caption{The evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
+The models include a magnetic field with an intensity of 4G.
+The solid black line represents the reference model according to \citet{Choi2016}.
+The rest of lines are models which include initial rotation with $\omegaini$ between 0.0084 and 0.0336, respectively.
+The purple star is the surface Li abundances for the present-day Sun \citep{Asplund2009}.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:li_var_vel_4_0g4}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[clip,width=\columnwidth]{figures/paper2/rot_vel_var_vel_4_0g4.pdf}
+    \caption{The evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
+The models include a magnetic field with an intensity of 4G, initial rotation with $\omegaini$ between 0.0084 and 0.0336, respectively and MB.
+The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
+The dashed vertical line makes reference to the ZAMS.}
+    \label{fig:rot_vel_var_vel_4_0g4}
+\end{figure}
+
+\subsection{Gossage et al. (2021) Comparison}
+Due to the existence of similarities between \citet{Gossage2021} work and this one, we proceed to make a comparison between the two of them.
+Among the similarities we find the fact that both papers focus on magnetic braking, making use of MESA for the simulations.
+Among the main differences is that our models incorporate a variable $B$ and $\amlt$.
+Finally, both compare their simulation results (rotation periods and Li abundances) with starts located in OCs.\par
+
+In \citet{Gossage2021}, the authors also approach the problem of AML from a similar starting point to ours, but with some notable variations.
+In the Table \ref{tab:gassage_vs_navarro} we compare the similarities and differences between the two studies.
+In both works the models consider rotation not only in the MS but also in the PMS, with the distinction that in Gossage's work they take into account disk-locking.
+This aspect is not so relevant in our simulations since we have focused on the A(Li) profiles, and Gossage's work focuses mainly on modeling rotation profiles of stars, which are significantly impacted by the disk-locking mechanism.
+Similarly, the effects of magnetic braking on angular momentum loss are also considered, in our case for stars with 1$\msun$ (solar twins) and in their with 0.1$\msun$$\le$$\mstar$$\le$1.3$\msun$.
+Both models also use MLT but in our case the value $\amlt$ is not prefixed but varies throughout the simulation.\par
+
+These two works complement each other as they are based on different formalisms.
+On the one hand, our work is based on the proposal of \citet{Gallet2013} whose most relevant parameters are $\rossby$ and the presence of a magnetic field of varying strength (B), whereas Gossage presents two formalisms: \citet{Matt2015} and \citet{Garraffo2018}.
+The former is mainly characterised by $\rossby$ and the $\mstar$, while the latter by $\rossby$ and a parameter that accounts for the complexity of the magnetic fields, being equal to 1 for a dipole field and larger for higher-order multipoles.
+As far as OCs observations are concerned, here we covered a wider range of ages [0.001-6.761] Ga.
+Finally, both papers simulate their respective models in MESA but use different versions.\par
+
+\citet{Gossage2021} manage to reproduce more accurately the Sun's rotation profile at its current age \citep[see][panel (a) in Figures 2, 4 \& 6]{Gossage2021} although, as the authors of the paper describe, they calibrated their models to do so, by including an additional diffusion-free parameter ($v_0$).
+In our work, our model does not include this type of additional parameterization and is still able to reproduce an equatorial rotation speed of 4.72 km/s, a value not particularly far from the nominal set value of 2.0 km/s.
+As far as A(Li) is concerned, their model does not reproduce the value for Sun-like stars, including the Sun itself \citep[see][panel (d) in Figures 2 \& 4, and (b) in Figure 6]{Gossage2021}.
+By recalibrating some of the parameters of their model and/or incorporating additional mechanisms that contribute to Li destruction, the authors hope to obtain values in line with observational data.
+In our case (see Section \ref{sec_conclusions}), we were able to reproduce A(Li) with greater precision.
+
+\begin{table}
+	\centering
+    \begin{threeparttable}
+	\begin{tabular}{lll} 
+		\hline
+		Parameter & This work & \citet{Gossage2021}\\
 		\hline
@@
 		Convection & MLT variable $\alpha_{{\rm MLT}}$ & MLT fixed $\alpha_{{\rm MLT}}$\\
-		Magnetic Braking & \cite{Gallet2013} & \cite{Matt2015}, \\ Formalism & & \cite{Garraffo2018}\\
-            Magnetic Braking & $\rossbystar$ \& $\bstar$ & $\rossbystar$ \& $\mstar$, \\ Parameters & & $\rossbystar$ \& $\magfc$\tnote{(1)}\\
-		Magnetic Field & B(G) variable & Not used\\ Strength & & \\
+		Magnetic Braking & \citet{Gallet2013} & \citet{Matt2015}, \\ Formalism & & \citet{Garraffo2018}\\
+		Magnetic Field & B(G) variable & B(G) pre-set\\ Strength & & \\
             Disk locking & Not present & Present\\
@@
     \end{threeparttable}
-        \begin{tablenotes}\footnotesize
-        \item (1) Magnetic field complexity
-        \end{tablenotes}
+	\caption{Comparison with \citet{Gossage2021} work.}\label{tab:gassage_vs_navarro}    
 \end{table}
@@
 \section{Conclusions and future works} \label{sec_conclusions}
-We have shown, through the different simulated stellar models, that the effects induced by the combination of both rotation and magnetic braking mechanisms offer a plausible way to reconcile the observational data with the theoretical models.
+Through various simulated stellar models, we've demonstrated that the interplay of rotation and magnetic braking can align observational data with theoretical models, potentially through an $\Omega$-$B$ interdependence law.
+These models, unlike our previous ones, don't preset initial values for magnetic field strength and the $\amlt$ parameter.
+We've validated these models with Gaia and GES data for Sun-like stars, especially their Li abundances.\par
+
+Our models suggest a lower Li abundance limit of 1.133 dex for Sun-like stars, within the Sun's error margin (1.1 $\pm$ 0.1 dex).
+They predict a higher equatorial rotational velocity (4.72 km/s) than the Sun (2.0 km/s), enabling more Li burning.
+Despite developing strong magnetic fields and intensive magnetic braking, our models don't match the Sun's omega value and average magnetic field strength.\par
+
+While our evolving magnetic field strength and MLT parametrization align with observations, we must interpret these results cautiously. Key findings include:
+\begin{enumerate}
+    \item Varying magnetic field intensity is a valid alternative to preset models, aligning with previous work on Sun-like stars' rotational history.
+    \item A variable adaptive $\amlt$ formalism eliminates the need for preset values, reproducing values consistent with previous work and providing potential value ranges when upfront $\amlt$ values are unknown or undesirable.
+    \item Our models suggest a lower limit for A(Li)=1.133 dex to the Gaia and GES observations for solar twin stars.
+    \item Despite initializing models at higher $\omegaini$ values than in our previous work, resulting in $B$ up to 3 orders of magnitude more intensive, $\approx$ 1KG versus 4G, both works' simulations produce similar values for rotational velocity and A(Li).
+\end{enumerate}
+
+In conclusion, enhancing our understanding of stellar interiors and Li depletion mechanisms for solar twins is crucial for more accurate comparisons.
+Broadening our models' mass range will enable us to compare results with a wider observational range.
+Our models currently over-deplete Li compared to Gaia and GES data.\par
+
+Considering that faster rotating stars deplete more Li, additional mechanisms, like disk locking to the protostellar disk, are needed to explain observations.
+This introduces two braking mechanisms acting at different star ages, establishing a self-regulating mechanism influencing Li evolution and magnetic field strength, which could align with the Sun's lower intensities.\par
+
+Lower rotational velocity means less Li destruction in our models.
+To compensate for this, we could use overshooting, an effect not simulated in our models, which would allow Li to be reached by hot material bubbles, decreasing its abundance.
+Ideally, like $\amlt$, overshooting should depend on other fundamental stellar parameters.\par
+
+%We have shown, through the different simulated stellar models, that the effects induced by the combination of both rotation and magnetic braking mechanisms offer a plausible way to reconcile the observational data with the theoretical models.
 We propose that this result could be achieved by a law of interdependence between $\Omega$ and $B$.
 In contrast to our previous work, these new models remove the restrictions of pre-setting initial and fixed values along their time evolution for the magnetic field strength and the $\amlt$ parameter.
-We cross-checked the model results with measurements obtained from the Gaia and GES mission for stars with similar characteristics to our Sun, focusing on their Li abundances.
-
-Our models seem to suggest a lower limit to the surface Li abundances of 1.133 dex for Sun-like stars, value which is within the margin of error obtained for the Sun (1.1 $\pm$ 0.1 dex).
-On the other hand, they predict a higher angular velocity (4.72 km/s) than the Sun (2.0 km/s) so that they could burn more Li and finally reproduced the Sun Li existing reservoirs.
+We have cross-checked the model results with measurements obtained from the Gaia and GES mission for stars with similar characteristics to our Sun, focusing on their Li abundances.\par
+
+%Our models seem to suggest a lower limit to the surface Li abundances of 1.133 dex for Sun-like stars, value which is within the margin of error obtained for the Sun (1.1 $\pm$ 0.1 dex).
+On the other hand, they predict a higher rotational velocity at equator (4.72 km/s) than the Sun (2.0 km/s) so that they could burn more Li and finally reproduced the Sun Li existing reservoirs.
 Coupled with this higher angular velocity, strong magnetic fields developed, triggering an intensive magnetic braking that slowed its angular velocity but not enough to match the present $\Omega$ value of the Sun.
@@
 
-As far as Li abundance and $\Omega$ are concerned, our models, while not exactly reproducing the values of the Sun, show promising results.
-Although we have shown how both an evolving magnetic field strength and, a MLT parametrization could produce results in line with the observations, we must analyse them results obtained with caution and not draw any premature conclusions.
-Nonetheless, we can draw the following finding from the present work:
-
-\begin{enumerate}
-    \item A magnetic field with varying intensity is a valid alternative to explain the rotational history of Sun-like stars.
-    \item A variable $\amlt$ eliminates the need to pre-fix its value in advance for the entire evolution of the star.
-    \item Our models seem to suggest a lower limit to the observations provided by Gaia and GES for solar sibling stars.
-    \item The strength of the magnetic field regulates the size of the star's convection zone.
-The magnetic field has a direct implication on the rotational history of the star, and the speed with which the star rotates impacts the size of the convective zone.
-    \item Inclusion of the magnetic field leads to cooler models and lower Li depletion in the MS.
-    \item Disk-locking during PMS seem to be also required for explaining Li abundances in young clusters.
-\end{enumerate}
-
-We would like to conclude with the following thoughts for the next steps.
-It is a fact that our models still destroy too much Li when comparing the produced Li abundances obtained with those from Gaia and GES.
-Taking as starting point that faster rotating stars destroy more Li, other mechanisms, in addition to magnetic braking, are necessary to explain the observational data.
-Among them, disk locking to the protostellar disk could contribute to counteract the acceleration of the star during PMS.
-Thus, we would have two braking mechanisms acting at different ages of the star.
-The disk locking would act during the PMS and the magnetic braking mainly during the MS.
-This approach would establish a self-regulating mechanism over the angular velocity of the star that would end up directly influencing the both Li evolution and the magnetic field strength.
-For the latter, lower intensities are expected which could finally converge with that of our Sun.
-On the other hand, a lower rotational speed also means less Li destruction for our models.
-One way to compensate for this lower Li remaining is to resort to over-shooting (see \cite{Navarro2020}) an effect that has not been simulated in our models.
+%Although we have shown how both an evolving magnetic field strength and, a MLT parametrization could produce results in line with the observations, we must analyse the results obtained with caution and not draw any premature conclusions.
+In spite of these shortcomings, we can draw the following finding from the present work:
+
+%\begin{enumerate}
+%    \item A magnetic field with varying intensity is a valid alternative to the models with a pre-fixed value as they produce results in line with previous work for the rotational history of Sun-like stars.
+%    \item A variable adaptive $\amlt$ formalism could eliminate the need to prefix its value.
+By using it, we were able to reproduce values in line with previous works.
+Additionally, it could also serve to derive an interval of potential values in cases when an upfront $\amlt$ value is unknown or even not desirable.
+%    \item Our models seem to suggest a lower limit for A(Li)=1.133 dex to the observations provided by Gaia and GES for solar twin stars.   
+%    \item Although the models in the present work are initialized at higher $\omegaini$ values than in our previous work, 0.1425 and 0.0336 respectively, resulting in $B$ up to 3 orders of magnitude more intensive, $\approx$ 1KG versus 4G, the simulation in both works end up producing very similar values for both the rotational velocity and the A(Li).
+For the former, this is explained because a higher magnetic field strength implies an intensive MB effect.
+For the latter, the lower $\amlt$ values produced by the adaptive formalism preserves a higher Li reservoir during the PMS.
+%\end{enumerate}
+
+%We would like to conclude with the following thoughts for the next steps.
+It is imperative to increase our understanding of stellar interiors and the mechanisms of Li depletion for solar twins sample in order to perform more accurate comparisons.
+Also, expanding the mass range in our models will allow us to confront the results with a wider range of observations.
+It is a fact that our models still destroy too much Li when comparing the simulated values with those from Gaia and GES.\par
+
+%Taking as starting point that faster rotating stars destroy more Li, other mechanisms, in addition to MB, are necessary to explain the observational data.
+Among them, disk locking to the protostellar disk could contribute to counteracting the acceleration of the star during PMS.
+In this way, two braking mechanisms act at different ages of the star.
+Disk locking will act during PMS and MB mainly during MS.
+With such approach it would be established a self-regulating mechanism over the angular velocity of the star that would end up directly influencing both the Li evolution and the magnetic field strength.
+For the latter, lower intensities are expected which could finally converge with that of our Sun.\par
+
+%On the other hand, a lower rotational speed also means less Li destruction for our models.
+One way to compensate for this remaining lower Li is to resort to overshooting \citep[see ][]{Caballero2020} an effect that has not been simulated in our models.
 By activating this mechanism, we would allow the Li to be reached by bubbles of material hot enough to destroy it, thus contributing to decrease its abundance.
-Ideally, as we have done with $\amlt$, overshooting should not be introduced as a free parameter but dependent on other fundamental stellar parameters.\par
+Ideally, as we have done with $\amlt$, overshooting should not be introduced as a fixed parameter but dependent on other fundamental stellar parameters.\par
 
@@
 We wish to acknowledge the generous work and help offered by the MESA and TOPCAT communities.
-Authors acknowledges funding support from Spanish public funds (including FEDER fonds) for research under projects ESP2017-87676-C5-2-R and ESP2017-87676-C5-5-R.
+The authors acknowledge funding support from Spanish public funds (including FEDER fonds) for research under projects ESP2017-87676-C5-2-R and ESP2017-87676-C5-5-R.
+AGH also acknowledges the funding support of the 'European Regional Development Fund / Andaluca-Consejera de Economa y Conocimiento' under project E-FQM-041-UGR18 by the Universidad de Granada.
 JCS also acknowledges support from project RYC-2012-09913 under the 'Ram\'on y Cajal' program of the Spanish Ministry of Science and Education.
@@
 The simulations used in this paper been executed with the MESA release 10398 \footnote{\url{http://mesa.sourceforge.net/release/2018/03/21/r10398.html}}.
-The different figures have been generated making use of GNU Octave 5.1.0.
+The different figures have been generated using GNU Octave 5.1.0.
 \footnote{\url{https://www.gnu.org/software/octave/}}.
-The Gaia and GES data sets have been analyzed using TOPCAT 4.8-7 \footnote{\url{https://www.g-vo.org/topcat/topcat/}}.
+Gaia and GES data sets have been analyzed using TOPCAT 4.8-7 \footnote{\url{https://www.g-vo.org/topcat/topcat/}}.
 
@@
 		\hline
-             & & & & & & & $\oomegac$ & & & & & $\oomegac$ & & \\
+             & & & & & & & $\omegaini$ & & & & & $\omegaini$ & & \\
 		Open Cluster & GES\_FLD & Age (Gyr) & [Fe/H] & $N_*$ & 0.095 & 0.10 & 0.105 & 0.11 & 0.115 & 0.12 & 0.125 & 0.13 & 0.14 & 0.1425\\
@@
 For each OC, name, GES denomination, estimated age, metallicity and number of components are listed.
-The number of OC components selected by the different simulations are shown separately.
-This information is split into two groups, each corresponding to the set of simulations used in the figures shown in the following sections.
-Finally, those OCs that have been selected in any of the simulations are highlighted in grey.}
+Two sets of simulations for different $\omegaini$ are shown, where $\omegaini = \oomegac$, $\Omega$ is the star angular velocity at stellar surface, and $\omegac$ is the surface velocity at the equator of a rotating star where the centrifugal force balances the Newtonian gravity.
+Those OCs with member that have been selected in any of the simulations are highlighted in grey.}
   	\label{tab:oc_full_list}
@@
 \caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
-Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\oomegac$ varies between 0.0084 and 0.0336, respectively.
-The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
+The top two figures show a set of models with variable magnetic field with intensity, and $\omegaini$ varying between 0.095 and 0.1425.
+In the bottom one the magnetic field intensity was set to $0.0\,\Gauss$, and $\omegaini$ varying between 0.0 and 0.0336.
+The purple star is the surface Li abundance for the present-day Sun \citep{Asplund2009}.
 The dashed vertical line makes reference to the ZAMS.}
@@
     \end{subfigure}
-\caption{Grid showing the evolution of surface \isotope[7]{Li} abundance relative to \isotope[1]{H}, as a function of time for several 1 $\msun$ models.
-Each figure shows a set of models in which $\oomegac$ has been fixed and the magnetic field with intensity varies between $0.0\,\Gauss$ and $5.0\,\Gauss$, respectively.
-The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009} and the Pleiades cluster \citep{Sestito2005} respectively.
+\caption{Grid showing of the evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
+The top two figures show a set of models with variable magnetic field with intensity, and $\omegaini$ varying between 0.095 and 0.1425.
+In the bottom one the magnetic field intensity was set to $0.0\,\Gauss$, and $\omegaini$ varying between 0.0 and 0.0336.
+The purple star and square are surface Li abundance for the present-day Sun \citep{Asplund2009}.
 The dashed vertical line makes reference to the ZAMS.}
@@
     \end{subfigure}
-\caption{Grid showing of the evolution of surface rotational velocity, as a function of time for several 1 $\msun$ models.
-Each figure shows a set of models in which the magnetic field with intensity has been fixed and $\oomegac$ varies between 0.0084 and 0.0336.
-The purple star is the surface angular velocity for the present-day Sun \citep{Gill2012}.
+\caption{Grid showing of the evolution of $\amlt$, as a function of time for several 1 $\msun$ models.
+The two figures show a set of models with variable magnetic field with intensity, and $\omegaini$ varying between 0.095 and 0.1425.
+The purple star and diamond are the $\amlt$ given by \citet{Sonoi2018} and \citet{Samadi2005}, respectively.
 The dashed vertical line makes reference to the ZAMS.}

### 2024-02-28 project/octave/plotting_routines.m

@@ -1038,7 +1038,7 @@ function plot_size_cz2(A, color, width, ytick, axis_limits, offset, linestyle)
   %Plot values
   %plot(A(:,1), A(:,3) .- A(:,4), color, 'linewidth', width);
   plot(A(:,1), A(:,3) .- A(:,4) .+ offset , color, 'linewidth', width, 'linestyle', linestyle);
-  plot(A(:,1), A(:,5) .- A(:,6) .+ offset , color, 'linewidth', width, 'linestyle', '--');
+  %plot(A(:,1), A(:,5) .- A(:,6) .+ offset , color, 'linewidth', width, 'linestyle', '--');
   %plot(A(:,1), 10 .** A(:,2), color, 'linewidth', width, 'linestyle', '--');
 
   %Axis scales
@@ -1475,11 +1475,11 @@ function age_vs_cz_size_plots(gauss_fields, rotational_vels, is_var_vel, ytick, 
       %Generate serie labels
       if (is_var_vel)
         labels = {labels{:}, ['CZ_{rad}-', strtrim(rotational_vels(j,:))]};
-        labels = {labels{:}, ['Core_{rad}-', strtrim(rotational_vels(j,:))]};
+        %labels = {labels{:}, ['Core_{rad}-', strtrim(rotational_vels(j,:))]};
         %labels = {labels{:}, ['ZAMS-', strtrim(rotational_vels(j,:))]};
       else
         labels = {labels{:}, ['CZ_{rad}-', strtrim(gauss_fields(i,:))]};
-        labels = {labels{:}, ['Core_{rad}-', strtrim(gauss_fields(i,:))]};
+        %labels = {labels{:}, ['Core_{rad}-', strtrim(gauss_fields(i,:))]};
         %labels = {labels{:}, ['ZAMS-', strtrim(gauss_fields(i,:))]};
       endif
     end
@@ -1496,7 +1496,7 @@ function age_vs_cz_size_plots(gauss_fields, rotational_vels, is_var_vel, ytick, 
   set (l, "fontsize", legend_font_size);
   %legend boxoff
   xlabel('star age (Ga)', 'fontsize', axis_font_size);
-  ylabel('Size conv. zone (radius CZ & Core/radius star)', 'fontsize', axis_font_size);
[…]

### 2024-03-04 project/octave/plotting_routines.m
Update paper1 figures for using Gaia data

@@ -141,7 +141,7 @@ global teff_top_limit = 5800;
 
 
 
-global data_parent_folder = '/home/rcaballeron/MESA/workspace/sun-jupiter-system/Docs/runs/run_paper3d';
+global data_parent_folder = '/home/rcaballeron/MESA/workspace/sun-jupiter-system/Docs/runs/run_thesis';
 global tables_parent_folder = '/home/rcaballeron/MESA/workspace/sun-jupiter-system/project/tables';
 global filename = '1M_photosphere_history.data';
 
@@ -1298,6 +1298,7 @@ function hr_plots(gauss_fields, rotational_vels, is_var_vel, xy_ticks, axis_limi
     for j=1:rows(rotational_vels)
       sub_folder = strcat(gauss_fields(i,:), '_', rotational_vels(j,:));
       full_path = strcat(data_parent_folder, '/', sub_folder, '/', filename);
+      full_path
 
       fmt = get_parsing_fmt([star_age_col, log_Teff_col, log_L_col]);
 
@@ -2582,7 +2583,7 @@ function plot_0G_var_vel_z1(rot_vels,idx)
   age_vs_li_plots(gauss_fields(idx_0_0G,:), rot_vels, true, 0.1, [1.0e7, 1.0e8, 2.1, 2.4], 'northeast', 'A(Li7) - 0G & var. rotational velocity', 'li_var_vel_0_0g_z1_', num2str(idx));
 end
 
-function plot_2G_var_vel(rot_vels)
+function plot_2G_var_vel(rot_vels, idx)
   global gauss_fields;
   global idx_2_0G;
   gauss_fields = ['2g'];
@@ -2590,47 +2591,47 @@ function plot_2G_var_vel(rot_vels)
   %rotational_vels = ['035crit';'034crit';'033crit';'032crit';'031crit'];
   rotational_vels = ['026crit';'027crit';'028crit';'029crit';'030crit'];
 
-  age_vs_li_plots(gauss_fields(idx_2_0G,:), rot_vels, true, 0.5, [1.0e5,1.0e10,0,3.5], 'southwest', 'A(Li7) - 2.0G & var. rotational velocity', 'li_var_vel_2_0g');
[…]

### 2024-03-06 project/overleaf/Tesis III/capitulos/capitulo01.tex
Fix bibliography packages. WIP chapter reviews

@@
 
-\chapter{¿Por qué el Litio?}\label{ch:primer-capitulo}
+\chapter{¿Por qué el litio?}\label{ch:primer-capitulo}
 
 \section{Introducción}
-El estudio de las abundancias de Litio en las estrellas, particularmente aquellas similares al Sol, es crucial por varias razones.
-En primer lugar, el Litio es uno de los pocos elementos producidos en la nucleosíntesis del Big Bang.
+El estudio de las abundancias de litio en las estrellas, particularmente aquellas similares al Sol, es crucial por varias razones.
+En primer lugar, el litio es uno de los pocos elementos producidos en la nucleosíntesis del Big Bang.
 Su abundancia proporciona una prueba crítica para la teoría del Big Bang y nos permite sondear las condiciones del universo primitivo.
-Además, el Litio es sensible a las temperaturas estelares, y su tasa de destrucción aumenta rápidamente a temperaturas superiores a 2.5 millones de Kelvin, que son típicas para los interiores estelares.\par
+Además, el litio es sensible a las temperaturas estelares, y su tasa de destrucción aumenta rápidamente a temperaturas superiores a 2.5 millones de Kelvin, que son típicas para los interiores estelares.\par
 
-En segundo lugar, la abundancia de Litio puede arrojar luz sobre la estructura interna y los procesos de mezcla dentro de las estrellas.
-En los modelos estelares estándar, se espera que el Litio se agote en el sobre convectivo exterior de una estrella debido a su transporte a capas más profundas y calientes donde se destruye.
-Sin embargo, las observaciones a menudo muestran más agotamiento de Litio del que predicen estos modelos, lo que sugiere procesos de mezcla adicionales.
-Estudiar las abundancias de Litio puede ayudar a refinar nuestra comprensión de estos procesos y mejorar los modelos estelares.\par
+En segundo lugar, la abundancia de litio puede arrojar luz sobre la estructura interna y los procesos de mezcla dentro de las estrellas.
+En los modelos estelares estándar, se espera que el litio se agote en la zona convectiva exterior de una estrella debido a su transporte a capas más profundas y calientes donde se destruye.
+Sin embargo, las observaciones a menudo muestran más agotamiento de litio del que predicen estos modelos, lo que sugiere procesos de mezcla adicionales.
+Estudiar las abundancias de litio puede ayudar a refinar nuestra comprensión de estos procesos y mejorar los modelos estelares.\par
 
-El estudio de las abundancias de Litio es particularmente relevante al examinar cúmulos abiertos.
+El estudio de las abundancias de litio también es particularmente relevante al examinar cúmulos abiertos.
 Los cúmulos abiertos son grupos de estrellas que se han formado a partir de la misma nube molecular gigante, lo que significa que comparten una edad y composición química inicial comunes.
 Esto los convierte en excelentes laboratorios para estudiar la evolución estelar y la nucleosíntesis.
-El contenido de Litio en estas estrellas puede proporcionar valiosos conocimientos sobre los procesos de mezcla interna y la edad del cúmulo.
-A medida que las estrellas de un cúmulo envejecen, su abundancia de Litio superficial disminuye debido a la mezcla y la quema.
-Al comparar las abundancias de Litio observadas en las estrellas de un cúmulo abierto con modelos teóricos, podemos estimar la edad del cúmulo y obtener conocimientos sobre la eficiencia de los procesos de mezcla.\par
+El contenido de litio en estas estrellas puede proporcionar valiosos conocimientos sobre los procesos de mezcla interna y la edad del cúmulo.
+A medida que las estrellas de un cúmulo envejecen, su abundancia de litio superficial disminuye debido a la mezcla y la quema.
+Al comparar las abundancias de litio observadas en las estrellas de un cúmulo abierto con modelos teóricos, podemos estimar la edad del cúmulo y obtener conocimientos sobre la eficiencia de los procesos de mezcla.\par
 
-Además, dado que todas las estrellas en un cúmulo inicialmente tienen la misma abundancia de Litio, cualquier diferencia observada debe ser debido a procesos que ocurren dentro de las estrellas.
-Esto nos permite investigar cómo factores como la masa, la temperatura, la rotación y la presencia de campos magnéticos afectan a la destrucción de Litio, contribuyendo a nuestra comprensión de los interiores estelares.
-Por lo tanto, el estudio de las abundancias de Litio en cúmulos abiertos juega un papel crucial en el avance de nuestro conocimiento de las estrellas y su evolución.\par
+Además, dado que todas las estrellas en un cúmulo inicialmente tienen la misma abundancia de litio, cualquier diferencia observada debe ser debido a procesos que ocurren dentro de las estrellas.
+Esto nos permite investigar cómo factores como la masa, la temperatura, la rotación y la presencia de campos magnéticos afectan a la destrucción de litio, contribuyendo a nuestra comprensión de los interiores estelares.
+Por lo tanto, el estudio de las abundancias de litio en cúmulos abiertos juega un papel crucial en el avance de nuestro conocimiento de las estrellas y su evolución.\par
+
+En conclusión, el estudio de las abundancias de litio en estrellas es fundamental para nuestra comprensión del universo.
+Desde su origen en la nucleosíntesis del Big Bang hasta su sensibilidad a las temperaturas estelares y su papel en la estructura interna de las estrellas, el litio nos brinda valiosos conocimientos sobre la evolución cósmica.
+Además, al examinar cúmulos abiertos, podemos desentrañar los procesos de mezcla y estimar la edad de estos grupos estelares.
+En última instancia, el litio nos conecta con los misterios del cosmos y nos permite explorar las profundidades del espacio y el tiempo.\par
 

### 2024-03-06 project/overleaf/Tesis III/capitulos/capitulo02.tex
Fix bibliography packages. WIP chapter reviews

@@
 \section{Introducción}
-La abundancia de Li observada en la fotosfera de las estrellas es un indicador de su composición interior y de los procesos de mezclado que en su él tienen lugar.
+La abundancia de litio observada en la fotosfera de las estrellas es un indicador de su composición interior y de los procesos de mezclado que en él tienen lugar.
 Adicionalmente, estas abundancias (además de otras métricas) se utilizan para comprobar la validez de los modelos estelares.
-Para que esto sea posible hay que tomar como premisa inicial que la abundancia de Li generada en la nucleosíntesis asociada al Big Bang es conocida y este elemento solo se destruye a través de reacciones nucleares.
-A pesar de décadas de esfuerzos teóricos, se sigue sin encontrar una explicación coherente con los modelos para las discrepancias encontradas en las comparaciones sobre la abundancia de Li para estrellas pertenecientes a cúmulos de diferentes edades y que se encuentran en el mismo estado evolutivo, o bien en la presecuencia principal (Pre-Main Sequence, PMS), o bien en la secuencia principal (Main Sequence, MS).
-Los modelos no son capaces de explicar las abundancias detectadas en las etapas tardías de la MS \cite{Tschape2001}.\par
+Para que esto sea posible hay que tomar como premisa inicial que la abundancia de Litio generada en la nucleosíntesis asociada al Big Bang es conocida, y que este elemento solo se destruye a través de reacciones nucleares.\par
 
-Es sabido que parte de la pérdida de Li se produce durante la PMS y que además ésta se acentúa según decrece la masa de la estrella y, a igualdad de masa, según aumenta la metalicidad de la misma.
-Para reproducir la dependencia entre la edad y masa de las estrellas con la merma de las concentraciones de Li, estos modelos necesitan de una combinación de procesos de mezclado cada vez más complejos, como overshooting, mezclado debido a procesos rotacionales o difusión microscópica, o la presencia de campos magnéticos.A pesar de décadas de esfuerzos teóricos, aún no se ha podido encontrar una explicación coherente derivada de modelos para las discrepancias de abundancia de Li detectadas en estrellas pertenecientes a cúmulos de diferentes edades y que se encuentran en el mismo estado evolutivo, ya sea en la PMS o en la MS.
-Además, estos modelos teóricos no son capaces de explicar las abundancias de elementos químicos detectadas en las etapas tardías de la MS \cite{Tschape2001}.\par
+En el ámbito de la astrofísica, persiste un enigma que desafía décadas de investigación teórica: las discrepancias entre los modelos de evolución estelar y las mediciones de abundancia de litio en estrellas.
+Estas discrepancias se manifiestan en estrellas que pertenecen a cúmulos de diferentes edades y que se encuentran en el mismo estado evolutivo, ya sea en la presecuencia principal (Pre-Main Sequence, PMS) o en la secuencia principal (Main Sequence, MS).
+A pesar de los esfuerzos, los modelos actuales no logran explicar las abundancias observadas en las etapas tardías de la MS \citep{Tschape2001}.\par
 
-En la presente tesis estamos interesados en investigar cómo la presencia de campos magnéticos pueden influir en las abundancias de Li detectadas en las atmósferas de las estrellas de tipo solar.
-El Li es un elemento que se destruye fácilmente en las capas interiores de este tipo de estrellas.
+Es sabido que parte de la pérdida de litio se produce durante la PMS y que además ésta se acentúa según decrece la masa de la estrella y, a igualdad de masa, según aumenta la metalicidad de la misma.
+Para reproducir la dependencia entre la edad y masa de las estrellas con la merma de las concentraciones de litio, estos modelos necesitan de una combinación de procesos de mezclado cada vez más complejos, como rebasamiento (overshooting), mezclado debido a procesos rotacionales o difusión microscópica, o la presencia de campos magnéticos.\par
+
+En la presente tesis estamos interesados en investigar cómo la presencia de campos magnéticos pueden influir en las abundancias de litio detectadas en las atmósferas de las estrellas de tipo solar.
+El litio es un elemento que se destruye fácilmente en las capas interiores de este tipo de estrellas.
 Este proceso ocurre principalmente durante la PMS en mayor medida, aunque puede darse también durante la MS.
 Adicionalmente, cabe la posibilidad de que se dé en las capas exteriores si existe un proceso eficiente de mezclado en la estrella, en el que puede influir la presencia de campos magnéticos.
-Por tanto, el estudio de la abundancia de Li en la superficie de la estrella puede ser clave para entender el proceso de evolución del momento angular de la misma.\par
+Por tanto, el estudio de la abundancia de litio en la superficie de la estrella puede ser clave para entender el proceso de evolución del momento angular de la misma.\par
+
+---ESTO PARECE ESTAR FUERA DE SITIO --
 
@@
 
-\section{El Li primordial}
-Los modelos teóricos de estrellas no informan sobre la cantidad inicial de Li que tiene una estrella, sólo describen cómo se agota.
-Por tanto, para hacer una estimación precisa de la abundancia inicial de Li, es un requisito poder comparar previamente las observaciones y los modelos.
-Nuestro Sol representa una excepción única, ya que nos permite conocer la abundancia actual de este elemento en su fotosfera, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex \footnote{Contracción procedente del inglés para decimal exponent.}$ \cite{Jeffries2004}, donde $A(\isotope[7]{Li})$ se define según la Ec.~\ref{eq:A_Li}\par
+-- HASTA AQUÍ --
+
+\section{El litio primordial}
+Los modelos teóricos existentes de evolución estelar no informan sobre la cantidad inicial de litio que tiene una estrella, sólo describen cómo se agota.
+Por tanto, para hacer una estimación precisa de la abundancia inicial de litio, es un requisito poder comparar previamente las observaciones y los modelos.
+Nuestro Sol representa una excepción única, ya que nos permite conocer la abundancia actual de este elemento en su fotosfera, $A(\isotope[7]{Li}) = 1.1 \pm 0.1 \, dex \footnote{Contracción procedente del inglés para decimal exponent.}$ \citep{Jeffries2004}, donde $A(\isotope[7]{Li})$ se define según la Ecuación \ref{eq:A_Li}
 
@@
 \end{ceqn}
-donde $N$ es el número de átomos del elemento en cuestión por unidad de volumen y 12 representa la abundancia del H, el cual se toma como punto de referencia para evitar que aparezcan números negativos.\par
 
-Por otro lado, la abundancia inicial de $A(\isotope[7]{Li}) = 3.34 \, dex$ se obtiene a partir de mediciones de meteoritos \cite{Randich2006}.
-Según la teoría establecida para las estrellas recién nacidas, tenemos que la abundancia inicial de Li puede estimarse con bastante precisión a partir de mediciones fotosféricas en estrellas de tipo T-Tauri, o de las estrellas F más calientes que forman parte de cúmulos algo más antiguos.
-Para estas últimas, la teoría actual de la evolución estelar sugiere que el Li aún no debe estar agotado.
-Los resultados obtenidos a partir de las medidas en ambos tipos de estrellas nos permiten fijar la abundancia inicial de Li en el intervalo $3.0 \, dex < A(\isotope[7]{Li}) < 3.4 \, dex$ \cite{Randich2006}.\par
+donde $N$ es el número de átomos del elemento en cuestión por unidad de volumen y 12 representa la abundancia de $H$, el cual se toma como punto de referencia para evitar que aparezcan números negativos.\par
 
-\section{Evolución del Li en la presecuencia principal (PMS)}
+Por otro lado, la abundancia inicial de $A(\isotope[7]{Li}) = 3.34$ dex se obtiene a partir de mediciones de meteoritos \citep{Randich2006}.
+Según la teoría establecida para las estrellas recién nacidas, tenemos que la abundancia inicial de litio puede estimarse con bastante precisión a partir de mediciones fotosféricas en estrellas de tipo T-Tauri, o de las estrellas F más calientes que forman parte de cúmulos algo más antiguos.
+Para estas últimas, la teoría actual de la evolución estelar sugiere que el litio aún no debe estar agotado.
+Los resultados obtenidos a partir de las medidas en ambos tipos de estrellas nos permiten fijar la abundancia inicial de litio en el intervalo $3.0 < A(\isotope[7]{Li}) < 3.4$ dex \citep{Randich2006}.\par
+
+\section{Evolución del litio en la presecuencia principal (PMS)}
 La PMS es la continuación directa de la fase protoestelar.
-Inicialmente, a medida que el objeto estelar joven evoluciona a lo largo del tiempo, su luminosidad es fundamentalmente consecuencia directa de un proceso de acreción suficientemente fuerte que es capaz de mantener la fusión activa de deuterio \cite{Stahler1983}.
+Inicialmente, a medida que el objeto estelar joven evoluciona a lo largo del tiempo, su luminosidad es fundamentalmente consecuencia directa de un proceso de acreción suficientemente fuerte que es capaz de mantener la fusión activa de deuterio \citep{Stahler1983}.
 Una vez que cesa la acumulación de material, la protoestrella, rodeada sólo por un disco residual de material, desciende a lo largo de su traza de Hayashi de una forma casi vertical.
-Esta fase de la evolución estelar se localiza en el diagrama Hertzspring-Russell \footnote{Diagrama en el que se enfrenta la luminosidad de la estrella contra su temperatura} y en la misma la estrella es totalmente convectiva.
-A medida que la estrella desciende por la traza de Hayashi se van contrayendo y su radio disminuye, a la par que lo hace su luminosidad.
-Su temperatura superficial ($\teff$)tiende a mantenerse  en torno $3000..5000 K$ aunque la interior va en aumento a consecuencia de la contracción del gas que conforma la estrella y el aumento de presión.
+Esta fase de la evolución estelar se localiza en el diagrama Hertzspring-Russell (HR) \footnote{Diagrama en el que se enfrenta la luminosidad de la estrella contra su temperatura} y en la misma la estrella es totalmente convectiva.
+A medida que la estrella desciende por la traza de Hayashi se va contrayendo y su radio disminuye, a la par que lo hace su luminosidad.
+Su temperatura superficial ($\teff$) tiende a mantenerse  en torno $3000..5000$ K aunque ésta va en aumento a consecuencia de la contracción del gas que conforma la estrella y el aumento de presión.
 Llegado un punto con suficiente temperatura en el interior de la estrella, ésta desarrolla una zona convectiva y la estrella entra en la traza de Heney, donde se producen reacciones nucleares, marcando así la entrada en la MS.
@@
 
-Durante la evolución estelar en esta fase, el Li se quema a temperaturas relativamente bajas (2.5 - 3.0 $10^{6} K$) y, en las estrellas de baja masa (< 1.2 $\msun$), el mezclado que se produce en la zona convectiva de la estrella puede hacer que el material ya agotado de Li alcance en relativamente poco tiempo la fotosfera de la estrella.
-Por esta razón, las mediciones de abundancia de Li fotosférica nos brindan uno de los pocos métodos de sondeo de interiores estelares y, a su vez, representan un importante banco de pruebas contra el que enfrentar los resultados que se obtienen a partir de los modelos evolutivos para PMS.
-Entender cómo se produce el agotamiento del Li durante la PMS también nos ofrece la posibilidad de poder estimar la edad de las estrellas más jóvenes y, por supuesto, es un punto de partida para poder cuantificar cualquier agotamiento subsiguiente de Li que se produzca a lo largo de la MS.\par
+Durante la evolución estelar en esta fase, el Li se quema a temperaturas relativamente bajas ($2.5..3.0 \, 10^{6}$ K) y, en las estrellas de baja masa (< 1.2 $\msun$), el mezclado que se produce en la zona convectiva de la estrella puede hacer que el material ya agotado de litio alcance, en relativamente poco tiempo, la fotosfera de la estrella.
+Por esta razón, las mediciones de abundancia de litio fotosféricas nos brindan uno de los pocos métodos de sondeo de interiores estelares y, a su vez, representan un importante banco de pruebas contra el que enfrentar los resultados que se obtienen a partir de los modelos evolutivos para estrellas que se encuentran en la PMS.
+Entender cómo se produce el agotamiento del litio durante esta fase evolutiva también nos ofrece la posibilidad de poder estimar la edad de las estrellas más jóvenes y, por supuesto, es un punto de partida para poder cuantificar cualquier agotamiento subsiguiente de litio que se produzca a lo largo de la MS.\par
 
@@
 
-La Figura \ref{fig:isocrhones} muestra la fase de PMS para modelos de distintas masas iniciales.
-Asimismo, se indican la fase de quema de Li y el momento de aparición del núcleo radiativo.
-Como se puede ver, las estrellas con M < 0.35 $\msun$ que se encuentran en la PMS tienen una estructura relativamente sencilla: son totalmente convectivas hasta que consiguen alcanzar la Termination Age Main Sequence (TAMS).
+La Figura \ref{fig:isocrhones} muestra la fase PMS para modelos de distintas masas iniciales.
+Asimismo, se indican la fase de quema de litio y el momento de aparición del núcleo radiativo.
+Como se puede observar, las estrellas con una masa $\mstar$ < 0.35 $\msun$ que se encuentran en la PMS tienen una estructura relativamente sencilla: son totalmente convectivas hasta que consiguen alcanzar la edad de finalización de secuencia principal (Termination Age Main Sequence, TAMS).
 A medida que la estrella va reduciendo su volumen según desciende a lo largo de la traza de Hayashi, su núcleo se va calentando, pero el gradiente de temperatura se mantiene muy cerca del gradiente adiabático, el cual delimita la frontera entre un núcleo radiativo o convectivo.
-Esta situación se mantiene para la práctica totalidad de la estrella, con la excepción de las capas superficiales.
-El Li comienza a arder en reacciones de captura protónica cuando la temperatura central, Tc=2.5 $10^{6} K$ y, debido a que la tasa de reacción es tan sensible a la temperatura para densidades típicas de PMS \cite{Randich2006} y a que el mezclado convectivo se produce de forma tan rápida, todo el Li se quema en una pequeña fracción de tiempo.
-Por otra parte, se ha contrastado que la edad a la que se produce el agotamiento de Li es inversamente proporcional a la masa de la estrella, es decir, aumenta según disminuye la masa hasta que se alcanza el límite de M < 0.06 $\msun$, ya que en este tipo de estrellas nunca llegará a alcanzarse una temperatura lo suficientemente alta como para poder quemar Li.\par
+Esta situación se mantiene para la práctica totalidad de la estrella, con la excepción de las capas superficiales.\par
 
-En las estrellas M > 0.35 $\msun$, el proceso de agotamiento del Li es bastante más complejo.
-Tienen densidades centrales más bajas y a medida que la Tc aumenta durante la contracción del PMS, la opacidad cae lo suficiente como para que el gradiente de temperatura se vuelva subadiabático, es decir, inhibe el proceso de mezclado y se forma un núcleo radiativo que empuja hacia afuera para incluir una fracción rápidamente creciente de la masa estelar.
-Para M <  $\msun$ solo existe una pequeña ventana de oportunidad en la que poder quemar un poco de Li antes de que se desarrolle el núcleo radiativo (aproximadamente a 2 Myr para 1 $\msun$).
-Como se puede observar en la Figura \ref{fig:isocrhones}, para M < 0.6 $\msun$ el núcleo radiativo se desarrolla antes de que la quema de Li esté completa.
-A falta de mezcla por convección, el material agotado de Li no puede llegar a la fotosfera y una vez que la temperatura en la base de la zona de convección cae por debajo del umbral de combustión de Li, el agotamiento de Li fotosférico se detiene.
-Para una estrella de 1 $\msun$, el agotamiento de Li fotosférico comienza aproximadamente a los 2 Myr y debería de terminar al alcanzar los 15 Myr.\par
+El litio comienza a arder en reacciones de captura protónica cuando la temperatura central, $T_c=2.5 \, 10^{6}$ K y, debido a que la tasa de reacción es tan sensible a la temperatura para densidades típicas de PMS \citep{Randich2006} y a que el mezclado convectivo se produce de forma tan rápida, todo el litio se quema en una pequeña fracción de tiempo.
+Por otra parte, se ha contrastado que la edad a la que se produce el agotamiento de litio es inversamente proporcional a la masa de la estrella, es decir, aumenta según disminuye la masa hasta que se alcanza el límite de $\mstar$ < 0.06 $\msun$, ya que en este tipo de estrellas nunca llegará a alcanzarse una temperatura lo suficientemente alta como para poder quemar litio.\par
 
-Los estudios sobre el Li en los cúmulos muy jóvenes indican que las estrellas de tipo solar sufren muy poco (si es que lo hay) agotamiento de Li durante las fases de la PMS  \cite{Jeffries2004}.
-Al mismo tiempo, se ha dedicado una gran cantidad de esfuerzo observacional al muestreo de abundancia de Li y también del Be \cite{Mena2011, DelgadoMena2014} con el objetivo de poner restricciones empíricas más severas a los mecanismos propuestos.
+En las estrellas $\mstar$ > 0.35 $\msun$, el proceso de agotamiento del Li es bastante más complejo.
+Tienen densidades centrales más bajas y a medida que la $T_c$ aumenta durante la contracción del PMS, la opacidad cae lo suficiente como para que el gradiente de temperatura se vuelva subadiabático, es decir, inhibe el proceso de mezclado y se forma un núcleo radiativo que empuja hacia afuera para incluir una fracción rápidamente creciente de la masa estelar.
+Para $\mstar$ <  $\msun$ solo existe una pequeña ventana de oportunidad en la que poder quemar un poco de litio antes de que se desarrolle el núcleo radiativo (aproximadamente a 2 Ma para 1 $\msun$).
+Como se puede observar en la Figura \ref{fig:isocrhones}, para $\mstar$ < 0.6 $\msun$ el núcleo radiativo se desarrolla antes de que la quema de litio esté completa.
+A falta de mezcla por convección, el material agotado de litio no puede llegar a la fotosfera y una vez que la temperatura en la base de la zona de convección cae por debajo del umbral de combustión de litio, el agotamiento de litio fotosférico se detiene.
+Para una estrella de 1 $\msun$, el agotamiento de litio fotosférico comienza aproximadamente a los 2 Ma y debería de terminar al alcanzar los 15 Ma.\par
 
-\section{Evolución del Li en la secuencia principal (MS)}
-Como ya mencionó en el trabajo \cite{Zappala1972} hace algo más de cuatro décadas, la medición de Li en cúmulos más antiguos que los Hyades es una herramienta clave para investigar el agotamiento que sufre el Li en la MS, además de las escalas de tiempo en las que ocurre este proceso.
-Ahora, más de 40 años después, se han llevado nuevas mediciones de Li para varios cúmulos antiguos, incluyendo NGC 752 y M 67 \cite{Sestito2006} que han permitido caracterizar de manera más completa la dispersión de Li en estrellas de tipo solar y con metalicidad y edad similar al Sol \cite{Sestito2005}.\par
+Los estudios sobre el litio en los cúmulos muy jóvenes indican que las estrellas de tipo solar sufren muy poco (si es que lo hay) agotamiento de litio durante las fases de la PMS  \citep{Jeffries2004}.
+Al mismo tiempo, se ha dedicado una gran cantidad de esfuerzo observacional al muestreo de abundancia de litio y también del berilio \citep{Mena2011, DelgadoMena2014} con el objetivo de poner restricciones empíricas más severas a los mecanismos propuestos.\par
 
-Basándonos en los resultados de los modelos de evolución estándar que solo tienen en cuenta los procesos de convección como los mecanismos a través de los cuales se produce el mezclado, solo debería agotarse el Li en la PMS, ya que es en esta fase cuando las estrellas exhiben una envolvente convectiva profunda con una base lo suficientemente caliente como para quemar el Li.
+\section{Evolución del litio en la secuencia principal (MS)}
+Como ya mencionaron \cite{Zappala1972} en su trabajo hace más de cinco décadas, la medición de litio en cúmulos más antiguos que los Hyades es una herramienta clave para investigar el agotamiento que sufre el litio en la MS, además de las escalas de tiempo en las que ocurre este proceso.
+Tres décadas después, se realizaron nuevas mediciones de litio para varios cúmulos antiguos, incluyendo NGC 752 y M 67 \citep{Sestito2006} que permitieron caracterizar de manera más completa la dispersión de litio en estrellas de tipo solar y con metalicidad y edad similar al Sol \citep{Sestito2005}.
+Recientemente, gracias a los datos aportados por el satélite Gaia \citep{Brown2016, Brown2018, Brown2021, Brown2022} y a las abundacias de litio obtenidas a partir de ellos por el estudio Gaia ESO (Gaia ESO Survey, GES) \citep{Gilmore2012,Randich2013,Randich2022} tenemos a nuestra disposición un mayor volumen de datos que nos permiten refinar aun más nuestros modelos.\par
+
+Basándonos en los resultados de los modelos de evolución estándar que solo tienen en cuenta los procesos de convección como los mecanismos a través de los cuales se produce el mezclado, solo debería agotarse el litio en la PMS, ya que es en esta fase cuando las estrellas exhiben una envolvente convectiva profunda con una base lo suficientemente caliente como para quemar el litio.
 Por otra parte, estos mismos modelos predicen que este proceso se detiene en la MS para estrellas de tipo solar o aquéllas más calientes.
-Sin embargo, los estudios llevados a cabo con cúmulos abiertos han revelado A(Li) 10 veces inferior respecto a los valores esperados en estrellas en su MS \cite{Sestito2005}, y de hecho, nuestro propio Sol presenta un agotamiento una A(Li) 100 veces inferior si comparamos con la abundancia de Li detectada en los meteoritos \cite{Lodders2003}.\par
+Sin embargo, los estudios llevados a cabo con cúmulos abiertos han revelado abundancias de litio A(Li) 10 veces inferior respecto a los valores esperados en estrellas en su MS \citep{Sestito2005}, y de hecho, nuestro propio Sol presenta un agotamiento una A(Li) 100 veces inferior si la comparamos con la detectada en los meteoritos \citep{Lodders2003}.\par
 
@@
 	\includegraphics[width=0.5\textwidth]{img/tesis/li_abundances_vs_age.pdf}
-	\caption{Abundancia media de Li en función de la edad.
+	\caption{Abundancia media de litio en función de la edad.
 Las masas y edades han sido estimadas a partir de las temperaturas efectivas hacienda uso de isócronas.
-Las abundancias de Li se denotan mediante la notación log n(Li) = N (Li)/N (H) + 12.
+Las abundancias de litio se denotan mediante la notación log n(Li) = N (Li)/N (H) + 12.
 Los diferentes símbolos indican estrellas con diferentes masas, a saber: 1$\pm$0.02 $\msun$ (círculos), 1.05$\pm$0.02 $\msun$ (cuadrados) y 1.1 $\pm$0.02 $\msun$ (triángulos).
 El Sol ($\odot$) también se muestra.
-La línea horizontal indica la abundancia primigenia de Li, figura de \cite{Randich2006}.}
+La línea horizontal indica la abundancia primigenia de litio, figura de \cite{Randich2006}.}
 	\label{fig:li_abundances_vs_age}
@@
 
-La Figura \ref{fig:li_abundances_vs_age} muestra que la evolución del Li desde aproximadamente 100 Myr hasta 6-8 Gyr es similar para los 3 rangos de masas.
+La Figura \ref{fig:li_abundances_vs_age} muestra que la evolución del litio desde aproximadamente 100 Ma hasta 6-8 Ga es similar para los 3 rangos de masas.
 Se observa que los valores log n(Li) declinan a una ratio casi constante a medida que la estrella envejece.
-Esto ocurre así hasta aproximadamente los 2 Gyr, momento a partir del cual el agotamiento del Li parece virtualmente detenerse y las abundancias tienden a converger hacia los valores del Spite plateau, la línea de base en la A(Li) encontrada para estrellas antiguas (o de población II) y que por tanto se formaron a partir de material no alterado por otros procesos, que orbitan el halo galáctico.
+Esto ocurre así hasta aproximadamente los 2 Ga, momento a partir del cual el agotamiento del litio parece virtualmente detenerse y las abundancias tienden a converger hacia los valores del Spite plateau, la línea de base en la A(Li) encontrada para estrellas antiguas (o de población II) y que por tanto se formaron a partir de material no alterado por otros procesos, que orbitan el halo galáctico.
 \par
 
-\section{Mecanismos de destrucción de Li}
-En este apartado procedemos a enumerar las diferentes opciones que actualmente defienden las principales líneas de investigación para intentar establecer una relación causa-efecto para la discrepancia de la abundancia de Li detectada en las fotosferas de estrellas de tipo solar.\par
+\section{Mecanismos de destrucción de litio}
+En este apartado procedemos a enumerar las diferentes opciones que actualmente defienden las principales líneas de investigación para intentar establecer una relación causa-efecto para la discrepancia de la abundancia de litio detectada en las fotosferas de estrellas de tipo solar.\par
 
-\subsection{Li y metalicidad}
-Durante más de tres décadas, las estrellas con baja metalicidad ([Fe/H] < -1.0 dex), también denominadas de Población II, se han caracterizado por exhibir una relación bastante estable entre la abundancia de Li y su metalicidad \cite{Guiglion2016}.
-Este nivel quedó inicialmente vinculado a la abundancia primordial asociada al Li, aunque esta situación ha cambiado a la luz de los resultados obtenidos por la misión Wilkinson Microwave Anisotropy Probe (WMAP), que han propiciado que la abundancia de este elemento predicha por el modelo Estándar de Nucleosíntesis para el Big Bang (SBBN) haya pasado a estimarse en 2.6 dex \cite{Spergel2003}.
-Este dato, sin embargo, entra en contradicción con la abundancia de Li medida en enanas viejas de baja metalicidad  (A(Li) $\simeq$ 2.2 dex) situadas en el halo de la galaxia conocido como el Spite plateau \cite{Spite1982}.
-Además, para estrellas de baja metalicidad (\ref{fig:li_abundances_sbbn}) se ha encontrado que la abundancia de Li no parece guardar relación con la metalicidad de la estrella \cite{Fu2015}.\par
+\subsection{Litio y metalicidad}
+Durante las últimas décadas, las estrellas con baja metalicidad ([Fe/H] < -1.0 dex), también denominadas de Población II, se han caracterizado por exhibir una relación bastante estable entre la abundancia de litio y su metalicidad \citep{Guiglion2016}.
+Este nivel quedó inicialmente vinculado a la abundancia primordial asociada al litio, aunque esta situación cambió a la luz de los resultados obtenidos por la misión de la sonda de anisotropía de microondas Wilkinson (Wilkinson Microwave Anisotropy Probe, WMAP), que propiciaron que la abundancia de este elemento predicha por el modelo estándar de nucleosíntesis para el Big Bang (Standard Big Bang Nucleosynthesis, SBBN) pasara a estimarse en 2.6 dex \citep{Spergel2003}.
+Este dato, sin embargo, entra en contradicción con la abundancia de litio medida en enanas viejas de baja metalicidad  (A(Li) $\simeq$ 2.2 dex) situadas en el halo de la galaxia conocido como el Spite plateau \citep{Spite1982}.
+Además, para estrellas de baja metalicidad (\ref{fig:li_abundances_sbbn}) se ha encontrado que la abundancia de litio no parece guardar relación con la metalicidad de la estrella \citep{Fu2015}.\par
 
@@
 	\includegraphics[width=0.5\textwidth]{img/tesis/li_abundances_sbbn.pdf}
-	\caption{La abundancia de Li derivada para estrellas de baja metalicidad situadas en la secuencia principal es aproximadamente tres veces menor que el valor de Li primordial predicho por la SBBN, figura de \cite{Fu2015}.}
+	\caption{La abundancia de litio derivada para estrellas de baja metalicidad situadas en la secuencia principal es aproximadamente tres veces menor que el valor de litio primordial predicho por la SBBN, figura de \cite{Fu2015}.}
 	\label{fig:li_abundances_sbbn}
@@
 
-Se han realizado investigaciones para intentar vislumbrar si existe una relación entre las concentraciones de Li y la metalicidad de la nube de la que surgen las estrellas \cite{Pinsonneault1997, BarradoyNavascues2001}).
-Parece que esto no es así y el agotamiento del Li observado no sugiere que dependa de manera concluyente del contenido en elementos metálicos presente en esa nube primordial \cite{Guiglion2016}.
-Se ha contrastado que tanto para los cúmulos jóvenes (alta metalicidad) como para los antiguos (baja metalicidad), cuyos niveles de metalicidad varían aproximadamente $\pm$0.2 dex con respecto a la metalicidad de nuestro Sol, parecen compartir una distribución de Li muy similar \cite{Sestito2005}.
-Adicionalmente, se ha medido que la abundancia de Li medida en el Sol es mucho más baja que la que presentada por otras estrellas similares situadas en su vecindad \cite{Reddy2003}.
+Se han realizado investigaciones para intentar vislumbrar si existe una relación entre las concentraciones de litio y la metalicidad de la nube de la que surgen las estrellas \citep{Pinsonneault1997, BarradoyNavascues2001}).
+Parece que esto no es así y el agotamiento del litio observado no sugiere que dependa de manera concluyente del contenido en elementos metálicos presente en esa nube primordial \citep{Guiglion2016}.
+Adicionalmente, se ha contrastado que tanto para los cúmulos jóvenes (alta metalicidad) como para los antiguos (baja metalicidad), cuyos niveles de metalicidad varían aproximadamente $\pm$0.2 dex con respecto a la metalicidad de nuestro Sol, parecen compartir una distribución de litio muy similar \citep{Sestito2005}.
+Asímismo, se ha medido que la abundancia de litio medida en el Sol es mucho más baja que la que presentada por otras estrellas similares situadas en su vecindad \citep{Reddy2003}.
 Esta variación es más acusada que la detectada para otros tipos de elementos presentes en él.\par
 
-Por otro lado, otros autores \cite{Sestito2005} defienden que la amplia gama de abundancias de Li observadas en estrellas similares al Sol y cercanas a éste es más fácilmente explicable si se establece una dependencia entre la abundancia de Li, la edad y masa de la estrella.\par
+Por otro lado, otros autores \citep{Sestito2005} defienden que la amplia gama de abundancias de litio observadas en estrellas similares al Sol y cercanas a éste es más fácilmente explicable si se establece una dependencia entre la abundancia de Li, la edad y masa de la estrella.\par
+
+A la luz de los resultados y conclusiones expuestas en los diferentes trabajos, se hace evidente que no existe un consenso unánime sobre la posible relación entre la metalicidad de una estrella y su abundancia de litio.\par
+
+--HASTA AQUÍ REVISADO--
 

### 2024-03-16 project/overleaf/Tesis III/capitulos/capitulo02.tex
WIP

@@
 Por tanto, el estudio de la abundancia de litio en la superficie de la estrella puede ser clave para entender el proceso de evolución del momento angular de la misma.\par
-
----ESTO PARECE ESTAR FUERA DE SITIO --
-
-Los denominados modelos estándar, modelos que incluyen la convección sólo como proceso de mezcla y no consideran otras opciones de transporte como la difusión o la pérdida de momento angular (Angular Moment Loss, AML), han sido los principales implicados en la elaboración de estas predicciones \cite{Sestito2005}.
-El Li se destruye a una temperatura $\tli \approx 2.5 x 10^6\; K$ cuando un átomo de Li colisiona con un protón produciendo dos átomos de He, algo que tiene lugar durante las reacciones protón-protón tipo II (P-P II), y por tanto se destruye directamente en las envolturas estelares cuando la temperatura en la base de la zona de convección (Base Convective Zone, BCZ) alcanza ese valor.
-El Sol en particular, y las estrellas de tipo solar en general, se caracterizan por tener una CZ que cubre gran parte del radio estelar durante la PMS lo que hace que su límite inferior supere $\tli$ \cite{Iben1965}.
-Este agotamiento se detiene durante la aproximación al inicio de la MS (Zero-Age Main Sequence, ZAMS) cuando la zona de convección retrocede y la temperatura en la BCZ es más fría que $\tli$.
-En los modelos estelares estándar sólo la masa y la composición química inicial determinan a qué distancia de la superficie estelar se alcanza la temperatura $\tli$, por lo que se espera que estrellas dentro de cúmulos con masa similar alcancen la ZAMS con iguales abundancias superficiales de Li.
-Además, también deberían mostrar una evolución del Li muy similar hasta su aproximación a la edad terminal de la MS (Terminal-Age Main Sequence, TAMS).
-Durante este mismo periodo los mecanismos de convección desencadenan un proceso de mezcla que homogeneiza la composición química de la envoltura convectiva, desde su límite inferior hasta la superficie estelar.
-Sin embargo, se han observado diferentes abundancias de Li en distintas poblaciones estelares \cite[][y referencias en las mismas]{Somers2014}.\par
-
--- HASTA AQUÍ --
 
@@
 
---HASTA AQUÍ REVISADO--
+\subsection{Litio y convección}
+Una gran cantidad de trabajo teórico y observacional se ha dedicado a la comprensión de litio y su evolución.
+Aún así, las evidencias acumuladas no terminan de concordar con las predicciones de los modelos estándar \citep{Sestito2005}; entendiendo por estándar a aquellos modelos que incluyen sólo los mecanismos de convección como proceso de mezcla y no tienen en cuenta fenómenos que podrían influir también como el transporte, como la difusión, las ondas de gravedad (nos referimos a ondas en el fluido de la estrella inducidas por la gravedad), la pérdida de momento angular o la presencia de planetas.\par
 
-\subsection{Li y convección}
-Una gran cantidad de trabajo teórico y observacional se ha dedicado a la comprensión de Li y su evolución, las evidencias acumuladas no concuerdan con las predicciones de los modelos estándar \cite{Sestito2005}; entendiendo por estándar a aquellos modelos que incluyen sólo los mecanismos de convección como proceso de mezcla y no tienen en cuenta fenómenos que podrían influir también como el transporte, como la difusión, las ondas de gravedad (nos referimos a ondas en el fluido de la estrella inducidas por la gravedad), la pérdida de momento angular o la presencia de planetas.\par
+En las estrellas de tipo solar, la quema de litio se produce a una temperatura aproximada de $2.5 10^{6}$ K mediante las reacciones de captura protónica que acontecen en interiores estelares.
+Por lo tanto, si se mantiene de manera continuada en el tiempo el mecanismo que transporta el litio entre la zona de convección exterior químicamente mixta y las regiones más profundas que poseen temperaturas lo suficientemente altas, éste acabará siendo destruido y su abundancia fotosférica disminuirá, llegando en algunos casos prácticamente a desparecer de la superficie estelar \citep{DelgadoMena2014}.\par
 
-En las estrellas de tipo solar, la quema de Li se produce a una temperatura aproximada de 2.5 $10^{6} K$ mediante las reacciones de captura protónica que acontecen en interiores estelares.
-Por lo tanto, si se mantiene de manera continuada en el tiempo el mecanismo que transporta el Li entre la zona de convección exterior químicamente mixta y las regiones más profundas que poseen temperaturas lo suficientemente altas, éste acabará siendo destruido y su abundancia fotosférica disminuirá, llegando en algunos casos prácticamente a desparecer de la superficie estelar \cite{DelgadoMena2014}.\par
+A su vez, la abundancia de litio medida en la fotosfera nos sirve también para realizar la calibración de modelos de evolución estelar.
+Sin embargo, es preciso mencionar que para que este planteamiento sea coherente, hay que suponer que la abundancia inicial de litio es conocida y que éste es única y exclusivamente destruido mediante reacciones nucleares.\par
 
-A su vez, la abundancia de Li medida en la fotosfera nos sirve también para realizar la calibración de modelos de evolución estelar.
-Sin embargo, es preciso mencionar que para que este planteamiento sea coherente, hay que suponer que la abundancia inicial de Li es conocida y que éste es única y exclusivamente destruido mediante reacciones nucleares.\par
-
-\subsection{Li y difusión}
+\subsection{Litio y difusión}
 Un factor determinante para conocer la fase evolutiva en la que se encuentra una estrella es su composición química interna.
@@
 
-En el trabajo de \cite{Richard2004}, mostraron que la menor abundancia de Li observada en las estrellas de tipo Población II es el resultado del agotamiento de Li producido por el efecto de difusión atómica que entra en competencia con los procesos de mezclado en las zonas radiativas de estas estrellas.\par
+En el trabajo de \cite{Richard2004} mostraron que la menor abundancia de litio observada en las estrellas de tipo Población II es el resultado del agotamiento del mismo producido por el efecto de difusión atómica que entra en competencia con los procesos de mezclado en las zonas radiativas de estas estrellas.La difusión microscópica acorta la vida útil de la secuencia principal y conduce a un agotamiento de los elementos superficiales.
+El modelo solar requiere de su inclusión o de lo contrario ni siquiera la edad del Sol se podría producir correctamente \citep{Thoul1993}.\par
 
-La difusión microscópica acorta la vida útil de la secuencia principal y conduce a un agotamiento de los elementos superficiales.
-El modelo solar requiere la inclusión de difusión microscópica, de lo contrario ni siquiera la edad del Sol se podría producir correctamente \cite{Thoul1993}.\par
-
-\subsection{Li y exoplanetas}
-El descubrimiento de un gran número de exoplanetas durante las dos últimas décadas \cite{Mayor1995, Bonfils2018} ha significado un importante empuje para la comunidad científica en el campo de la astrofísica.
+\subsection{Litio y exoplanetas} \label{sec:litio_exoplanetas}
+El descubrimiento de un gran número de exoplanetas durante las dos últimas décadas \citep{Mayor1995, Bonfils2018} ha significado un importante empuje para la comunidad científica en el campo de la astrofísica.
 Aparte de las consecuencias directas derivadas de poder estudiar los nuevos planetas descubiertos más allá de nuestro sistema solar, lo que ya es de por sí extremadamente interesante, no lo es menos la observación de sus estrellas anfitrionas, ya que éstas pueden llegar a aportar información muy valiosa acerca de las características globales, composición y formación de los sistemas planetarios extrasolares que albergan.\par
 
-La hipótesis de que la variabilidad de concentración de Li en las estrellas puede ser debida a una posible correlación entre la presencia de planetas es una línea de trabajo de diversos estudios \cite{Israelian2009, DelgadoMena2014, Figueira2014}.
-Estas investigaciones apuntan a que esta presencia podría ser el desencadenante de la discrepancia en la abundancia de Li medidas en estrellas de tipo solar.
-Para nuestro Sol, ésta es unas 140 veces inferior a la que le correspondería a su modelo estelar \cite{Israelian2009}, aunque la temperatura en la BCZ no es lo suficientemente alta como para quemarlo.\par
+La hipótesis de que la variabilidad de concentración de litio en las estrellas puede ser debida a una posible correlación entre la presencia de planetas es una línea de trabajo de diversos estudios \citep{Israelian2009, DelgadoMena2014, Figueira2014}.
+Estas investigaciones apuntan a que esta presencia podría ser el desencadenante de la discrepancia en la abundancia de litio medidas en estrellas de tipo solar.
+Para nuestro Sol, ésta es unas 140 veces inferior a la que le correspondería a su modelo estelar \cite{Israelian2009}, aunque la temperatura en en el límite inferior de la zona convectiva (Bottom Convective Zone, BCZ) no es lo suficientemente alta como para quemarlo.\par
 
@@
 Esta nueva situación ha derivado en la realización de nuevos trabajos, o en la revisión de otros anteriores, en los que se han podido realizar nuevos estudios comparativos más amplios y sin sesgo en lo que se refiere a poder comparar estrellas con y sin sistemas planetarios.
-Estos nuevos trabajos han arrojado evidencias, tanto a favor como en contra, sobre la existencia de una relación directa entre la menor abundancia de Li en las estrellas con planetas con respecto a las que no tienen.
-Entre los primeros, tenemos el trabajo de \cite{Israelian2009}, en el que sus autores afirman que tras el análisis comparativo de un total de 133 estrellas en el rango de temperaturas que va de los $5600..5900 K$, de las que 30 de ellas albergan planetas y 103 no, encontraron que la gran mayoría de aquéllas que poseían planetas presentaron una disminución acusada en la abundancia de Li respecto de las que no tenían planetas (\ref{fig:li_abundances_planets}).
-También indican que por encima y por debajo de ese rango de temperaturas no parece existir relación alguna con la presencia o no de planetas, extremo que justifican al alegar que para estrellas con una temperatura superficial por encima de los $5850 K$, las capas convectivas de las estrellas más masivas que el Sol no son lo suficientemente profundas y además se encuentran demasiado alejadas como para alcanzar las regiones interiores de las estrella donde se alcanza una temperatura suficiente para quemar el Li.\par
+Estos nuevos trabajos han arrojado evidencias, tanto a favor \citep{Bouvier2008, Israelian2009} como en contra \citep{Baumann2010}, sobre la existencia de una relación directa entre la menor abundancia de litio en las estrellas con planetas con respecto a las que no tienen.\par
 
-Incluso su trabajo va más allá al afirmar que la menor abundancia de Li en estrellas con una temperatura efectiva restringida al rango $\tsun \pm 80$ K ($\tsun = 5777$ K) es una constante en las estrellas con planetas que se encuentran en su MS y que este resultado es independiente de la temperatura efectiva o la masa de las mismas.
-Por otro lado, el Li apenas se ve afectado en las estrellas menos evolucionadas, más calientes ($\teff > 6000$ K) y masivas, ya que éstas presentan una capa convectiva delgada.
-Por tanto, en este tipo de estrellas no existe una relación directa entre la temperatura de las mismas y la concentración de Li.\par
+Respecto a los primeros, la línea argumental postula que la presencia de planetas alrededor de una estrella podría afectar la evolución de la abundancia de litio fotosférico al afectar historia rotacional de su estrella anfitriona.
+Mediante el desarrollo de modelos de evolución de tipo solar incluyendo rotación, tanto para rotadores lentos (unas 5 veces mayor la velocidad de rotación del Sol) como rápidos (unas 70 veces mayor), que se hacen evolucionar desde la PMS hasta la edad actual de nuestro Sol, han conseguido encontrar pruebas de que las estrellas con una rotación menor desarrollan un alto grado de rotación diferencial entre el núcleo radiativo y la envolvente convectiva, mientras que los rotadores rápidos evolucionan con poco desacoplamiento entre núcleo y envoltura.
+A raíz de esta fuerte rotación diferencial en la base de la envoltura convectiva, en los rotadores lentos el proceso de destrucción del litio se vuelve más eficiente, concluyendo que las estrellas que albergan exoplanetas y presentan un agotamiento del litio debieron de ser rotadores lentos durante ZAMS y que esto se debió a una interacción duradera entre la estrella y el disco plotoplanetario durante la PMS.
+Esto se ve reforzado por análisis comparativos realizados entre estrellas que algerban planetas y otras que no.
+En la Figura \ref{fig:li_abundances_planets} podemos observar como la gran mayoría de aquéllas que poseían planetas presentaron una disminución acusada en la abundancia de litio respecto de las que no tenían planetas.\par
+
+Entre los que apuntan a que no existe relación directa alguna entre la menor abundancia de Li detectada en la atmósfera de las estrellas que albergan planetas y las que no los tienen, se argumentación radica en el hecho de que las evidencias encontradas por otros investigadores en la dirección de la existencia de una relación entre las abundancias de litio detectadas y las estrellas que albergan planetas, se deben única y exclusivamente a consideraciones sesgadas y errores sistemáticos producidos durante el análisis de los datos recopilados.
+Por el contrario, estos mismos autores sostienen que sí existen evidencias claras entre la abundancia de litio de una estrella con su edad y la metalicidad (Z) de la misma.
+Por otra parte, la ingestión de un planeta “fallido” podría ser la causa por la que se produce un aumento de la abundancia de este elemento en la superficie de las estrellas.\par
 
@@
 	\includegraphics[width=0.5\textwidth]{img/tesis/li_abundances_planets.pdf}
-	\caption{Abundancia de Li vs temperatura efectiva en estrellas de tipo solar con y sin planetas.
+	\caption{Abundancia de litio vs temperatura efectiva en estrellas de tipo solar con y sin planetas.
 Las estrellas con planetas vienen representadas por los círculos rojos y las sin planetas por los círculos vacíos.
@@
 
-Recientemente, se ha postulado otra alternativa por la que la presencia de planetas alrededor de una estrella podría afectar la evolución de la abundancia de Li fotosférico \cite{Bouvier2008}.
-El autor mantiene que, en el contexto de las estrellas de tipo solar en las que se han detectado exoplanetas respecto de las que no los albergan, la menor concentración de Li en las primeras puede ser el resultado de su historia rotacional.
-Afirma que, mediante el desarrollo de modelos de evolución de tipo solar incluyendo rotación, tanto para rotadores lentos (unas 5 veces mayor la velocidad de rotación del Sol) como rápidos (unas 70 veces mayor), que hacen evolucionar desde la PMS hasta la edad actual de nuestro Sol, han conseguido encontrar pruebas de que las estrellas con una rotación menor desarrollan un alto grado de rotación diferencial entre el núcleo radiativo y la envolvente convectiva, mientras que los rotadores rápidos evolucionan con poco desacoplamiento entre núcleo y envoltura.
-A raíz de esta fuerte rotación diferencial en la base de la envoltura convectiva, en los rotadores lentos el proceso de destrucción del Li se vuelve más eficiente, concluyendo que las estrellas que albergan exoplanetas y presentan un agotamiento del Li debieron de ser rotadores lentos durante ZAMS y que esto se debió a una interacción duradera entre la estrella y el disco plotoplanetario durante la PMS.
-De este planteamiento también se desprende que la vida de los discos debe ser más longeva (>5 Myr) de lo esperado y, posiblemente, también sea una condición necesaria para que se produzca la formación de planetas masivos y su posterior migración hacia zonas más exteriores del sistema planetario.\par
+\subsection{Litio y rotación}
+El impacto de la rotación tanto en la PMS como en el agotamiento del Li para las estrellas de tipo solar ha sido ampliamente debatido en el pasado \citep{Pinsonneault1997,Jeffries2004,Somers2014} y revisado más recientemente sobre la base de la disponibilidad de medidas más precisas \citep{Bouvier2016}.
+Los resultados sugieren una conexión entre la velocidad de rotación de las estrellas y la abundancia de litio detectada en sus atmósferas estelares, de modo que las estrellas que giran más rápido destruyen menos litio que las que lo hacen de manera más lenta.
+De alguna manera los resultados publicados por \cite{Bouvier2016} son inesperados ya que apuntan en una dirección diametralmente opuesta a los de trabajos anteriores, en los que se predecía que las estrellas que rotan más rápidamente deberían de destruir una mayor cantidad de litio.\par
 
-Asimismo, no faltan tampoco los referentes bibliográficos que defienden que no existe relación directa alguna entre la menor abundancia de Li detectada en la atmósfera de las estrellas que albergan planetas y las que no los tienen.
-En el trabajo de \cite{Baumann2010} se procede a analizar 117 estrellas con propiedades semejantes a nuestro Sol.
-Sus autores sostienen que las evidencias encontradas por otros investigadores en la dirección de la existencia de una relación entre las abundancias de Li detectadas y las estrellas que albergan planetas, se deben única y exclusivamente a consideraciones sesgadas y errores sistemáticos producidos durante el análisis de los datos recopilados.
-Por el contrario, estos mismos autores sostienen que sí existen evidencias claras entre la abundancia de Li de una estrella con su edad y la metalicidad de la misma.
-Por otra parte, la ingestión de un planeta “fallido” podría ser la causa por la que se produce un aumento de la abundancia de este elemento en la superficie de las estrellas.\par
+Para explicar esta tendencia, es necesario recurrir a mecanismos adicionales que vinculen el agotamiento del litio y la rotación.
+Esta aparente contradicción de los resultados de diferentes trabajos no tiene por qué se tal.
+De hecho se plantean diferentes ideas que pueden ayudar a entender las discrepancias, entre las que se encuentra el acoplamiento de las estrellas con sus disco proto-estelar (ver \ref{sec:litio_exoplanetas}), que influiría en los mecanismos de mezclado ocasionados por la rotación \citep{Bouvier2008, Eggenberger2012}, la influencia de campos magnéticos \citep{Eggenberger2009} que tienen la propiedad de transmitir el momento angular (AM) de forma mucho más eficaz que inducir la mezcla \citep{Denissenkov2007}.
+Como consecuencia de este incremento en la eficiencia del transporte del AM, se reduce la cantidad de rotación diferencial entre las zonas radiativa y convectiva de la estrella (se fomenta una rotación de cuerpo sólido), así como las inestabilidades rotacionales inducidas.
+La acreción de material \citep{Baraffe2010} también se postula como una posibilidad para explicar la menor abundancia de litio para estrellas en el mismo rango de masas y edad.
+Episodios periódicos de acreción de material por parte de la estrella conducen a desarrollar temperaturas centrales significativamente más altas.
+A consecuencia de ello, una mayor cantidad de litio puede ser destruido.\par
 
-\subsection{Li y rotación}
-El impacto de la rotación tanto en la PMS como en el agotamiento del Li para las estrellas de tipo solar ha sido ampliamente debatido en el pasado \cite{Pinsonneault1997,Jeffries2004,Somers2014} y revisado más recientemente sobre la base de la disponibilidad de medidas más precisas \cite{Bouvier2016}.
-Los resultados sugieren una conexión entre la velocidad de rotación de las estrellas y la A(Li) detectada en sus atmósferas estelares, de modo que las estrellas que giran más rápido destruyen menos Li que las que lo hacen de manera más lenta.
-De alguna manera los resultados publicados por \cite{Bouvier2016} son inesperados ya que apuntan en una dirección diametralmente opuesta a los de trabajos anteriores, en los que se predecía que las estrellas que rotan más rápidamente deberían de destruir una mayor cantidad de Li.\par
-
-Para explicar esta tendencia, es necesario recurrir a mecanismos adicionales que vinculen el agotamiento del Li y la rotación.
-Esta aparente contradicción de los resultados de diferentes trabajos no tiene por qué se tal.
-De hecho se plantean diferentes ideas que pueden ayudar a entender las discrepancias, entre las que se encuentra el acoplamiento de las estrellas con sus disco proto-estelar, que influiría en los mecanismos de mezclado ocasionados por la rotación \cite{Bouvier2008, Eggenberger2012}, la influencia de campos magnéticos \cite{Eggenberger2009} que tienen la propiedad de transmitir el momento angular (AM) de forma mucho más eficaz que inducir la mezcla \cite{Denissenkov2007}.
-Como consecuencia de este incremento en la eficiencia del transporte del AM, se reduce la cantidad de rotación diferencial entre las zonas radiativa y convectiva de la estrella (se fomenta una rotación de cuerpo sólido), así como las inestabilidades rotacionales inducidas.
-La acreción de material \cite{Baraffe2010} también se postula como una posibilidad para explicar la menor A(Li) para estrellas en el mismo rango de masas y edad.
-Episodios periódicos de acreción de material por parte de la estrella conducen a desarrollar temperaturas centrales significativamente más altas.
-A consecuencia de ello, una mayor cantidad de Li puede ser destruido.\par
-
-\subsection{A qué se debe esta merma de Li}
-Como hemos enumerado en los puntos anteriores, existe una propuesta amplia de potenciales mecanismos que podrían, por una parte, gobernar el ciclo de vida del Li, y por otra, ayudar a entender a qué se deben las discrepancias entre las observaciones y los resultados arrojados por los modelos.
-Sirva como ejemplo la línea de trabajo de \cite{Ramirez2012} en el que sus autores miden las abundancias de Li de una gran muestra de estrellas para encontrar correlaciones entre éstas y otras propiedades.
-Tales correlaciones deberían poder ayudar a determinar exactamente cómo se destruye el Li en las estrellas y cómo la abundancia galáctica de este elemento ha cambiado durante la vida de la Vía Láctea.\par
+\subsection{A qué se debe esta merma de litio}
+Como hemos enumerado en las secciones anteriores, existe una propuesta amplia de potenciales mecanismos que podrían, por una parte, gobernar el ciclo de vida del litio, y por otra, ayudar a entender a qué se deben las discrepancias entre las observaciones y los resultados arrojados por los modelos.
+Sirva como ejemplo la línea de trabajo de \cite{Ramirez2012} en el que sus autores miden las abundancias de litio de una gran muestra de estrellas para encontrar correlaciones entre éstas y otras propiedades.
+Tales correlaciones deberían poder ayudar a determinar exactamente cómo se destruye el litio en las estrellas y cómo la abundancia galáctica de este elemento ha cambiado durante la vida de la Vía Láctea.\par
 
 De manera genérica, estos mecanismos pueden agruparse en dos ideas principales.
-Por un lado, los que abogan por la propuesta de que la abundancia de este elemento en la superficie de las estrellas podría ser menor de la que los modelos predicen debido a que el Li podría haberse transformado en elementos más pesados mediante reacciones nucleares.
-El mayor inconveniente de este planteamiento es que la temperatura requerida para fusionar el Li y crear otros elementos es normalmente más elevada que la que puede ser alcanzada por las capas externas convectivas presentes en estrellas similares a nuestro Sol.
-Por esa razón, se ha propuesto una segunda explicación alternativa que plantea que el Li podría haberse mezclado en las capas más profundas de la estrella y de este modo, haber desaparecido de las zonas exteriores \cite{Pinsonneault1997}.\par
+Por un lado, los que abogan por la propuesta de que la abundancia de este elemento en la superficie de las estrellas podría ser menor de la que los modelos predicen debido a que el litio podría haberse transformado en elementos más pesados mediante reacciones nucleares.
+El mayor inconveniente de este planteamiento es que la temperatura requerida para fusionar el litio y crear otros elementos es normalmente más elevada que la que puede ser alcanzada por las capas externas convectivas presentes en estrellas similares a nuestro Sol.
+Por esa razón, se ha propuesto una segunda explicación alternativa que plantea que el litio podría haberse mezclado en las capas más profundas de la estrella y de este modo, haber desaparecido de las zonas exteriores \citep{Pinsonneault1997}.\par
 
 \section{El litio y las reacciones nucleares} \label{sec:li_reac_nuc}
-Hasta aquí hemos presentado los posibles mecanismos que influyen de manera más o menos directa en la destrucción del Li, pero finalmente éste se destruye porque se produce una "quema" del mismo cuando las condiciones en el interior estelar son las adecuadas para que se produzcan una serie de reacciones nucleares.
-Como consecuencia, para poder obtener las abundancias de Li aL final de la TAMS, primeramente necesitamos conocer qué reacciones nucleares son las responsables de su creación.
-De acuerdo a Cox y Giuli \cite{Cox1968}, en las condiciones de temperatura y densidad que se dan en el interior de una estrella de tipo solar (1.0 $\msun$), el tiempo de vida de los elementos más ligeros: $\isotope[2]{H}$, $\isotope[3]{H}$, $\isotope[6]{Li}$, $\isotope[9]{B}$, $\isotope[10]{B}$ y $\isotope[11]{B}$ y los resultantes de combinarse con el H a través de las reacciones de cadena protón-protón en sus variantes I, II y III, tienen un tiempo de vida que va desde los pocos segundos hasta los aproximadamente $10^4$ años.
+Hasta aquí hemos presentado los posibles mecanismos que influyen de manera más o menos directa en la destrucción del litoi, pero finalmente éste se destruye porque se produce una "quema" del mismo cuando las condiciones en el interior estelar son las adecuadas para que se produzcan una serie de reacciones nucleares.
+Como consecuencia, para poder obtener las abundancias de litio al final de la TAMS, primeramente necesitamos conocer qué reacciones nucleares son las responsables de su creación.
+De acuerdo a Cox y Giuli \citep{Cox1968}, en las condiciones de temperatura y densidad que se dan en el interior de una estrella de tipo solar (1.0 $\msun$), el tiempo de vida de los elementos más ligeros: $\isotope[2]{H}$, $\isotope[3]{H}$, $\isotope[6]{Li}$, $\isotope[9]{B}$, $\isotope[10]{B}$ y $\isotope[11]{B}$ y los resultantes de combinarse con el H a través de las reacciones de cadena protón-protón en sus variantes I, II y III, tienen un tiempo de vida que va desde los pocos segundos hasta los aproximadamente $10^4$ años.
 Por lo tanto, todos estos elementos deberían de consumirse en las etapas iniciales de vida de la estrella, tan pronto su interior ronde los $10^7$ K.\par
@@
 	\centering
-	\includegraphics[width=0.4\textwidth]{img/tesis/pp-I.png}
+	\includegraphics[width=0.35\textwidth]{img/tesis/pp-I.png}
 	\caption {Cadena de reacciones protón-protón I (pp-I)}
@@
 	\centering
-	\includegraphics[width=0.4\textwidth]{img/tesis/pp-II.png}
+	\includegraphics[width=0.35\textwidth]{img/tesis/pp-II.png}
 	\caption {Cadena de reacciones protón-protón II (pp-II)}
@@
 	\centering
-	\includegraphics[width=0.4\textwidth]{img/tesis/pp-III.png}
+	\includegraphics[width=0.35\textwidth]{img/tesis/pp-III.png}
 	\caption {Cadena de reacciones protón-protón III (pp-III)}
@@
 Los tres tipos de variantes ocurren de manera simultánea en el interior de las estrellas, pero unas son más frecuentes que otras.
-De acuerdo a las investigaciones teóricas, parcialmente soportadas por las medidas de neutrinos procedentes del Sol, se calcula que el 86\% de las reacciones se corresponden con las de tipo I, un 14\% con la de tipo II, y solo un 0.02\% con las de tipo III \cite{Scholz2018}.
+De acuerdo a las investigaciones teóricas, parcialmente soportadas por las medidas de neutrinos procedentes del Sol, se calcula que el 86\% de las reacciones se corresponden con las de tipo I, un 14\% con la de tipo II, y solo un 0.02\% con las de tipo III \cite{Scholz2018}.\par
 
 De manera similar, la vida media de los isótopos involucrados en el ciclo CNO (Figura \ref{fig:ciclo-cno}) al recombinarse con el H dominante en esas primeras etapas debería de rondar los $10^6$-$10^7$ años.
-Sin embargo, este ciclo de fusión a través del cual las estrellas también generan He a partir del H es más representativo de estrellas con una masa superior a 1.3 $\msun$.
+Sin embargo, este ciclo de fusión a través del cual las estrellas también generan He a partir del H es más representativo de estrellas con una masa superior a 1.3 $\msun$.\par
 
@@
 	\centering
-	\includegraphics[width=0.5\textwidth]{img/tesis/ciclo CNO.png}
+	\includegraphics[width=0.4\textwidth]{img/tesis/ciclo CNO.png}
 	\caption {Cadena de reacciones del ciclo CNO}

### 2024-03-16 project/overleaf/Tesis III/capitulos/capitulo04.tex
WIP

@@
 % !TeX encoding = utf8
+
 \chapter{Modules for Experiments in Stellar Astrophysics - MESA}\label{ch:cuarto-capitulo}
 \section{Introducción}
-Para la realización de este tesis doctoral nos hemos apoyado en la herramienta de evolución estelar \textit{Modules for Experiments in Stellar Astrophysics} (MESA).
+Como hemos anticipado, la herramienta de evolución estelar MESA ha sido utilizada extensivamente en nuestra línea de investigación.
 Este simulador incorpora módulos que se actualizan de manera periódica en base a los avances derivados de los trabajos más novedosos sobre ecuaciones de estado, opacidad, velocidades de reacción nuclear, datos de difusión de elementos y condiciones de límite atmosférico.
-Es una herramienta muy útil para poder contrastar mediciones obtenidas con los resultados arrojados por un determinado modelo, o para hacer predicciones a partir de modelos y contrastar los resultados que obtenemos de estos con las mediciones disponibles en otros trabajos de investigación.\par
+Es una herramienta muy versátil a la hora de poder contrastar las valores obtenidos con los resultados esperados por un determinado modelo, o para realizar predicciones a partir de nuevos modelos y contrastar sus pronósticos con los datos disponibles de otros trabajos de investigación.\par
 
-En el caso particular de esta tesis doctoral estamos interesados en estudiar los mecanismos que intervienen en la destrucción de Li en estrellas similares a nuestro Sol, y particularmente en ella.
-Como hemos comentado a día de hoy no existe una opinión unánime en la comunidad científica, más allá de que los modelos de evolución estelar no producen resultados coherentes con las observaciones, de qué procesos gobiernan el proceso de agotamiento del Li estelar y de cuándo se ponen en marcha o se detienen.
-Existen diversos planteamientos de procesos que intentan dar una respuesta a este enigma.
-En este trabajo nos vamos a centrar en cómo influyen en modelos de estrellas de tipo solar, con 1.0 $\msun$ y con una metalicidad (Z) similar al Sol, los efectos de la rotación y de la pérdida de momento angular causada por la presencia de campos magnéticos que inducen un efecto de frenado magnético.
+En el caso particular de nuestra línea de investigación estamos interesados en estudiar los mecanismos que intervienen en la destrucción del litio en estrellas similares a nuestro Sol, y particularmente en ella.
+Resaltar nuevamente el hecho de que a día de hoy no existe una opinión unánime en la comunidad científica, más allá de que los modelos de evolución estelar no producen resultados coherentes con las observaciones, sobre cuáles son los mecanismos que gobiernan el proceso de agotamiento del litio estelar, y de cuándo se ponen en marcha o se detienen.
+Existen diversos planteamientos de procesos que intentan dar una respuesta a este enigma.\par
+
+En nuestra investigación nos centramos en cómo influyen en modelos de estrellas de tipo solar, aquéllos con 1.0 $\msun$ y con una metalicidad (Z) similar al Sol, los efectos de la rotación y de la pérdida de momento angular causada por la presencia de campos magnéticos que inducen un efecto de frenado magnético.
 Estos campos magnéticos tendrán, en una primera aproximación, una intensidad fija a lo largo de toda la evolución temporal del modelo.
-Posteriormente se extenderá el modelo para simular campos magnéticos de intensidad variable.
-Adicionalmente, los modelos simulados se basan en la teoría de la longitud de mezcla para modelar la convección estelar.
+Posteriormente se extenderá el modelo para simular campos magnéticos de intensidad variable.\par
+
+Adicionalmente, los modelos simulados se basan en la teoría de la longitud de mezcla (Mixing Length Theory, MLT) para modelar la convección estelar.
 Este formalismo depende del parámetro libre longitud de mezcla ($\amlt$), parámetro que mayoritariamente tiene un valor preestablecido y fijo durante toda la simulación.
@@
 
-En estas circunstancias, nos encontramos en que el primer problema a solventar es el de cómo incorporar estas extensiones al simulador MESA, al mismo tiempo que se respeta su estructura modular y, sobre todo, sin provocar efectos indeseables en el resto de parámetros de la simulación.
-Para dar solución a este punto tenemos que basarnos en los planteamientos teóricos documentados en el Capítulo \ref{ch:tercer-capitulo} e implementarlos, en forma de rutinas en código Fortran, que puedan ser utilizadas en conjunto con el resto de módulos de MESA.\par
+Una vez hemos definido el los objetivos principales a alcanzar, el siguiente paso lógico es definir una estrategia para poder alcanzarlos, y esta pasa por hacer uso de MESA.
+Al estudiar en detalle la funcionalidad que MESA\footnote{\url{http://mesa.sourceforge.net/release/2018/03/21/r10398.html}} oferta por defecto, se constata que el simulador no ofrece capacidades ni para la incorporación de campos magnéticos, ni para la variabilidad temporal del valor $\amlt$.
+Esto implica que, como parte de nuestra línea de investigación, debemos añadir como un nuevo objetivo la ampliación de las capacidades técnicas del simulador MESA.
+Adicionalmente, éste debe ser el primero a alcanzar antes de poder realizar las simulaciones adecuadas para nuestra línea de investigación.\par
 
+El primer desafío a solventar es el de entender cómo incorporar las extensiones necesarias al simulador MESA.
+Al mismo tiempo, hay que respetar su estructura modular de manera que no provoquemos efectos colaterales indeseables en otros módulos y/o el resto de parámetros utilizados por el simulador.
+Para dar solución a este punto tenemos que basarnos en los planteamientos teóricos documentados en el Capítulo \ref{ch:tercer-capitulo} e implementarlos, en forma de rutinas en código Fortran, para que puedan ser utilizadas en conjunto con el resto de módulos de MESA.
 Por otra parte, se hace también necesario el comprender los diferentes parámetros y procesos que MESA utiliza a la hora de calcular las abundancias de los diferentes elementos que podemos encontrar en una estrella a lo largo de su evolución, tanto en su interior como en su atmósfera.
-Por tanto, aquí tenemos otra área de estudio que pasa por conocer las diferentes capacidades ya disponibles en el simulador que tengan impacto sobre el agotamiento del Li, entender cuándo pasan a ser relevantes en el proceso evolutivo de la estrella, cuáles son los parámetros que gobiernan su funcionamiento, las relaciones que guardan entre ellos y, por último, qué valores son los más recomendables para dar soporte al escenario de simulación que acabamos de plantear.\par
+Por tanto, aquí tenemos otra área de estudio que pasa por conocer las diferentes capacidades ya disponibles en el simulador que tengan impacto sobre el agotamiento del litio, entender cuándo pasan a ser relevantes en el proceso evolutivo de la estrella, cuáles son los parámetros que gobiernan su funcionamiento, las relaciones que guardan entre ellos y, por último, qué valores son los más recomendables para dar soporte al escenario de simulación que acabamos de plantear.\par
 
-Una vez tengamos la parametrización adecuada para nuestro planteamiento teórico, pasaremos a simular de manera sistemática, haciendo uso de nuestras nuevas rutinas, un conjunto de modelos con diferentes valores para los parámetros relevantes de nuestros modelos: velocidad angular, intensidad del campo magnético y valor de $\amlt$.
-Los resultados de abundancia de Li que obtengamos en cada uno de los escenarios planteados los enfrentaremos con los que arroja MESA sin hacer uso de nuestra rutina.
-Esta comparación de resultados nos deberá de dar un primer indicio de si nuestro planteamiento realmente tiene algún efecto notable sobre la abundancia Li detectado en la superficie estelar.
-Finalmente, en base a los resultados obtenidos, analizaremos las consecuencias que nuestras rutinas han tenido sobre ellos y presentaremos una serie de conclusiones de por qué obtenemos estos datos y cómo se interpretan los mismos en base a los estudios teóricos y observaciones obtenidas por otras líneas de investigación.\par
+Una vez tengamos la parametrización adecuada para nuestro planteamiento teórico, pasaremos a simular de manera sistemática, haciendo uso de nuestras nuevas rutinas y de un conjunto de modelos con diferentes valores para sus parámetros relevantes: velocidad angular, intensidad del campo magnético y valor de $\amlt$.
+Los resultados de abundancia de litio que obtengamos en cada uno de los escenarios planteados los enfrentaremos con los que arroja MESA sin hacer uso de nuestra rutina.
+Esta comparación de resultados nos deberá de dar un primer indicio de si nuestro planteamiento realmente tiene algún efecto notable sobre la abundancia litio calculadas por el simulador con y sin nuestras rutinas.
+Finalmente, en base a los valores obtenidos, analizaremos las consecuencias que nuestras rutinas han tenido sobre ellos y presentaremos una serie de conclusiones de por qué obtenemos estos datos y cómo se interpretan los mismos en base a los estudios teóricos y observaciones obtenidas por otras líneas de investigación.\par
 
@@
 Como indica su contribuidor y desarrollador principal \author{QUEPONERAQUÍ??} \cite[xxxxxx]{Pasetto2014} (Paxton, et al., 2011, 2013, 2015) MESA es un conjunto de librerías de código abierto, robustas y eficientes escritas en el lenguaje de programación Fortran 95 que es susceptible de ser utilizado en una amplia gama de aplicaciones en astrofísica estelar computacional.
-Está categorizado dentro de los denominados códigos de evolución estelar unidimensional (1-D) y combina un gran número de módulos numéricos y físicos que le permiten simular una amplia gama de escenarios de evolución estelar, que van desde los que incluyen estrellas de muy baja masa hasta lo que caracterizan estrellas masivas, además de tener en cuenta fases avanzadas de evolución, como el flash del helio, pulsos térmicos o la rama asintótica gigante (AGB).
+Está categorizado dentro de los denominados códigos de evolución estelar unidimensional (1-D) y combina un gran número de módulos numéricos y físicos que le permiten simular un gran número de escenarios de evolución estelar, que van desde los que incluyen estrellas de muy baja masa hasta los que caracterizan estrellas masivas, además de tener en cuenta fases avanzadas de evolución, como el flash del helio, pulsos térmicos o la rama asintótica gigante (XXXXXXXX XXXXXXXX AGB).
 Utiliza un modelo de malla o zonas adaptables, que representa las diferentes capas y celdas en las que se estructura el modelo estelar, y emplea sofisticados controles de evolución temporal.\par
@@
 
-Por un lado, MESA aborda la física estelar, su estructura y su evolución con métodos numéricos modernos y sofisticados.
-Por otro, utiliza una física actualizada que le confiere una amplia gama de aplicaciones en diferentes escenarios.
-Los métodos numéricos y computacionales empleados por MESA le permiten evolucionar de manera consistentemente a través de fases exigentes planteadas por los modelos estelares.\par
-
-MESA se basa en el principio de código abierto: cualquiera puede descargar el código fuente en el que se basa, compilarlo y ejecutarlo para sus propias necesidades de investigación, además de poder ampliar su funcionalidad, como se mostrará a continuación en este trabajo.
+MESA se basa en el principio de código abierto: cualquiera puede descargar el código fuente en el que se basa, compilarlo y ejecutarlo para sus propias necesidades de investigación, además de poder ampliar su funcionalidad, como se mostrará a lo largo del presente documento.
 El propósito de ser distribuido como código abierto es el de involucrar y llegar al mayor número de miembros de la comunidad astrofísica para que hagan uso de la herramienta, y al mismo tiempo alentarla para que hagan contribuciones al proyecto, ya sea en forma de pruebas, detección y corrección de errores o añadiendo nuevas capacidades al simulador.\par
@@
 \section{¿Cómo trabajar con MESA?}
-Cuando se descarga e instala MESA (en el presente trabajo no entramos al detalle de cómo hacer esto) veremos que bajo el directorio de instalación existe una serie amplia de subdirectorios adicionales.
+Cuando se descarga e instala MESA (en el presente documento no entramos al detalle de cómo hacer esto) veremos que bajo el directorio de instalación existe una serie amplia de subdirectorios adicionales.
 La mayoría de ellos contienen el código para los diferentes módulos que componen MESA y que proporciona alguna funcionalidad específica.
 De todos ellos, el módulo más importante es \textit{star}, ya que es éste el módulo que interacciona y controla la ejecución de los demás.
-Adicionalmente es también el módulo encargado gestionar el estado interno de la estrella y de calcular el paso temporal a utilizar en cada iteración durante el proceso de simulación.\par
+Adicionalmente, es también el módulo encargado de gestionar el estado interno de la estrella y de calcular el paso temporal a utilizar en cada iteración durante el proceso de simulación.\par
 
@@
 
-La diferencia entre las secciones \textit{star\_job} y controls puede llegar a ser muy sutil.
-Aun así, intentaremos dar unas reglas generales para tratar de averiguar en qué sección esperaríamos encontrar determinadas opciones de configuración.
+Las diferencias entre las secciones \textit{star\_job} y \textit{controls} pueden llegar a ser muy sutiles.
+Aun así, intentaremos dar unas reglas generales para tratar de averiguar en qué sección esperaríamos encontrar determinadas opciones de configuración.\par
 
@@
 
-Por otra parte, la sección controls es la adecuada para opciones de configuración encaminadas a responder a las siguientes cuestiones:
+Por otra parte, la sección \textit{controls} es la adecuada para opciones de configuración encaminadas a responder a las siguientes cuestiones:
 \begin{itemize}
@@
 Para conocer en más detalle las opciones, lo mejor es remitirse a la documentación estándar de MESA disponible en su sitio oficial\footnote{http://mesa.sourceforge.net}.
-Otra opción disponible, es la de consultar los ficheros de código donde se encuentran los valores por defecto asignados a cada una de las opciones del simulador.
+Otra opción a nuestro alcance es la de consultar los ficheros de código donde se encuentran los valores por defecto asignados a cada una de las opciones del simulador, aunque para esto es recomendable poseer conocimientos en lenguajes de programación.
 Para cada una de las secciones comentadas anteriormente, MESA dispone de un fichero específico (\$MESA\_DIR hace referencia al directorio donde se encuentra instalado el software):
@@
 
-Las opciones tienden a estar agrupadas según la funcionalidad del simulador que regulan.
-Es recomendable que cuando se esté buscando una determinada opción, antes de entrar al detalle de cada una de las entradas en los ficheros, se intente identificar mediante las preguntas listadas anteriormente en qué fichero podríamos encontrarla y, una vez accedamos al contenido del mismo, revisemos las diferentes cabeceras disponibles a modo de comentarios que en ellos se encuentra.
+Las opciones tienden a estar agrupadas según la funcionalidad del módulo que regulan.
+Es recomendable que cuando se esté buscando una determinada opción, antes de entrar al detalle de cada una de las entradas en los ficheros, se intente identificar mediante las preguntas listadas anteriormente en qué fichero podríamos encontrarla y, una vez accedamos al contenido del mismo, revisemos las diferentes cabeceras disponibles a modo de comentarios que en ellos se encuentran.\par
 
-\subsubsection{Controlando la salida}
+\subsubsection{Controlando la salida} \label{sec:ctrl_salida}
 MESA genera por defecto una cantidad ingente de información.
 Existen dos tipos de ficheros de salida con información de la simulación: uno con información histórica sobre cantidades globales (masa, luminosidad) almacenadas a diferentes intervalos de tiempo y otro tipo con perfiles del interior estelar que almacena información con cantidades que varían de forma espacial (densidad, presión) en un determinado instante de tiempo.
-El contenido de estos ficheros no se encuentra directamente accesible en el fichero inlist, sino en los siguientes ficheros específicos para cada uno de ellos:
+El contenido de estos ficheros no se encuentra directamente accesible en el fichero \textit{inlist}, sino en los siguientes ficheros específicos para cada uno de ellos:
 
@@
 
-Adicionalmente, si se diese la situación (como ha sido en este trabajo) de que se quieren mostrar valores de propiedades diferentes a los que aparecen por defecto en alguno de estos dos ficheros, cabe la posibilidad de añadirlos haciendo uso, como veremos más adelante, del fichero \textit{run\_star\_extras.f}.
+Adicionalmente, si se diese la situación (como ha sido en nuestro caso) de que se quieran mostrar valores de propiedades diferentes a los que aparecen por defecto en alguno de estos dos ficheros, cabe la posibilidad de añadirlos haciendo uso, como veremos más adelante, del fichero \textit{run\_star\_extras.f}.\par
 
@@
 A veces se hace necesario extender la funcionalidad de MESA porque la que viene por defecto en el simulador no cumple con las necesidades del problema que queremos abordar.
-Para estas situaciones, MESA ofrece la opción de extender su funcionalidad de un modo relativamente sencillo y elegante y sin necesidad de tener que modificar el código fuente del simulador en sí.
-Con esto se garantiza que el simulador no queda alterado en su funcionamiento estándar a raíz de una alteración del código base y permite que otros usuarios de MESA no tengan que alterar sus instalaciones para poder ejecutar simulaciones que hacen uso de funcionalidades no estándar.
+Para estas situaciones, MESA ofrece la opción de ampliar su funcionalidad de un modo relativamente sencillo y elegante, y sin necesidad de tener que modificar el código fuente del simulador en sí.
+Con esto se garantiza que el mismo no queda alterado en su funcionamiento estándar a raíz de una alteración del código base, a la vez que permite que otros usuarios de MESA no tengan que alterar sus instalaciones para poder ejecutar simulaciones que hacen uso de funcionalidades no estándar.
 \par
@@
 MESA proporciona el archivo \textit{run\_star\_extras.f} que ofrece una variedad de puntos de extensión (hooks) destinados a implementar acciones que no pueden hacerse únicamente mediante el uso del fichero de proyecto \textit{inlists}.
-Entre estas acciones adicionales tenemos la posibilidad de cambiar parámetros en cada paso de la simulación o la de reemplazar muchas de las rutinas de física predeterminadas que trae el simulador.\par
+Entre estas acciones adicionales tenemos la posibilidad de cambiar parámetros en cada paso de la simulación, o la de reemplazar muchas de las rutinas de física predeterminadas que trae el simulador.\par
 
@@
 Las rutinas definidas en ese fichero que se incorpora son las que queremos personalizar, pero sin afectar a las versiones originales.
-Para ello, procederemos a reemplazar esta declaración de inclusión por el contenido del archivo referenciado, así conseguimos tener nuestra propia copia local y personal.\par
+Para ello, procederemos a reemplazar esta declaración de inclusión por el contenido del archivo referenciado, consiguiendo de este modo tener nuestra propia copia local y personal.\par
 
@@
 Para ello se necesitan seguir los siguientes pasos:
+
 \begin{enumerate}
@@
 Las diferentes rutinas presentes en \textit{run\_star\_extras.f} se invocan en diferentes puntos durante la ejecución de MESA.
-En el Listado \ref{lst:bucle_ctrl}, y a modo de pseudo\-código en Fortran, podemos encontrar un aspecto del bucle principal de control sobre el que itera MESA.\par
-\begin{lstlisting}[language=Fortran, caption={Bucle de control principal sobre el que itera MESA en cada paso temporal}, label={lst:bucle_ctrl}]
+En el Listado \ref{lst:bucle_ctrl}, y a modo de pseudo\-código en Fortran, podemos encontrar el aspecto del bucle principal de control sobre el que itera MESA.\par
+
+\begin{lstlisting}[language=Fortran, float, caption={Bucle de control principal sobre el que itera MESA en cada paso temporal}, label={lst:bucle_ctrl}]
 subroutine run1_star(...)
@@
 
-Por otro lado, en la estructura de datos \textit{star\_info} MESA almacena toda la información sobre la estrella que está evolucionando.
+Por otro lado, en la estructura de datos \textit{star\_info}, MESA almacena toda la información sobre la estrella que está evolucionando.
 Por convención en el código, el nombre de la variable se utiliza a lo largo del mismo, incluyendo también el que aparece en \textit{run\_star\_extras.f} para hacer referencia a esta estructura\footnote{En Fortran, el operador de porcentaje (\%) se utiliza para acceder a los componentes de la estructura, así que s\% x hace referencia al campo x contenido en la estructura s.}.\par
@@
 
-Además de todo lo que hemos comentado hasta ahora, MESA también pone a nuestra disposición un conjunto de listas de valores que nos facilitará el poder pasar valores fijados en el fichero \textit{inlist} del proyecto a nuestras rutinas de código.
+Además de todo lo que hemos comentado hasta ahora, MESA también pone a nuestra disposición un conjunto de listas de valores que nos facilitará el poder pasar parámetros fijados en el fichero \textit{inlist} del proyecto a nuestras rutinas de código.
 Estos vectores de elementos, con un tamaño máximo de 100 valores y que no forman parte del conjunto estándar de parámetros, son: \textit{x\_ctrl}, \textit{x\_integer\_ctrl}, y \textit{x\_logical\_ctrl} y nos permiten almacenar valores de tipo real, entero y lógico respectivamente (ver Listado \ref{lst:paso_params}).\par
 
-\begin{lstlisting}[language=Fortran, caption={Paso de parámetros no estándar a través del fichero de proyecto inlist}, label={lst:paso_params}]
+\begin{lstlisting}[language=Fortran, float, caption={Paso de parámetros no estándar a través del fichero de proyecto inlist}, label={lst:paso_params}]
 &controls
@@
 \subsubsection{Extensión ficheros de salida}
-Anteriormente hacíamos referencia a cómo controlar la información estándar que MESA ofrece a través de los dos tipos de fichero de salida: \textit{history} y \textit{profile} (ver apartado PONER REFERENCIA para más detalles).
-Adicionalmente a esta opción, tenemos la posibilidad de enviar a alguno de estos dos ficheros información que, o bien la calcula ya de por sí MESA, pero no está pensada para ser volcada en estos ficheros, o bien puede representar valores que vienen derivados de nuestras rutinas y que, por lo tanto, no forman parte de la parametrización estándar de MESA.\par
+Anteriormente hacíamos referencia a cómo controlar la información estándar que MESA ofrece a través de los dos tipos de fichero de salida: \textit{history} y \textit{profile} (ver \ref{sec:ctrl_salida}).
+Adicionalmente a esta opción, tenemos la posibilidad de enviar a alguno de estos dos ficheros información que, o bien la calcula ya de por sí MESA pero no está pensada para ser volcada en estos ficheros, o bien puede representar valores que vienen derivados de nuestras rutinas y que, por lo tanto, no forman parte de la parametrización estándar de MESA.\par
 
-Para cualquiera de los dos casos comentados existen un par de rutinas dentro del fichero \textit{run\_star\_extras.f} que nos permiten, mediante sencillas modificaciones en los métodos \textit{how\_many\_extra\_history\_columns} y \textit{data\_for\_extra\_history\_columns}, añadir cuántos parámetros queramos\footnote{En realidad, existe un máximo de 100 parámetros adicionales.}.
-Para ello, simplemente hay que indicar la cantidad de nuevos parámetros en los que estamos interesados, el nombre de los mismos y de dónde obtener sus valores.\par
+Para cualquiera de los dos casos comentados existen un par de rutinas dentro del fichero \textit{run\_star\_extras.f} que nos permiten, mediante sencillas modificaciones en los métodos \textit{how\_many\_extra\_history\_columns} y \textit{data\_for\_extra\_history\_columns}, añadir cuántos parámetros queramos\footnote{En realidad, existe un máximo de 100 parámetros adicionales.} (ver Listado \ref{lst:extra_column}).
+Para ello, simplemente hay que indicar la cantidad de nuevos parámetros en los que estamos interesados, el nombre de los mismos (ver Listado \ref{lst:extra_name_columns}) y de dónde obtener sus valores.\par
 
-INTRODUCIR UNA CAPTURA DEL FICHERO DE CONFIGURACIÓN Y DEL CÓDIGO ASIGNANDO VALORES
+\begin{lstlisting}[language=Fortran, float, caption={Configuración del número de parámetros adicionales a incluir en el fichero de salida.}, label={lst:extra_columns}]
+  integer function how_many_extra_history_columns(id, id_extra)
+    integer, intent(in) :: id, id_extra
+    integer :: ierr
+    type (star_info), pointer :: s
+    ierr = 0
+    call star_ptr(id, s, ierr)
+    if (ierr /= 0) return
+
+    how_many_extra_history_columns = 43
+  end function how_many_extra_history_columns
+\end{lstlisting}
+
+\begin{lstlisting}[language=Fortran, float, caption={Definición de los nombres de los campos adicionales a aparecer en el fichero de salida.}, label={lst:extra_name_columns}]
+  subroutine data_for_extra_history_columns(id, id_extra, n, names, vals, ierr)
+    integer, intent(in) :: id, id_extra, n
+    character (len=maxlen_history_column_name) :: names(n)
+    real(dp) :: vals(n)
+    integer, intent(out) :: ierr
+    integer :: i
+    type (star_info), pointer :: s
+
+    ierr = 0
+    i = 0
+    call star_ptr(id, s, ierr)
+    if (ierr /= 0) return
+    !if (s% use_other_torque .eqv. .true.) then
+      !routine activated
+      names(1) = 'activated'
+      vals(1) = s% x_ctrl(25)
+
+      !magnetic field torque
+      names(2) = 'B'
+      vals(2) = s% x_ctrl(61)
+      ...
+
+      !alfven_r
+      names(43) = 'alfven_r'
+      vals(43) = s% x_ctrl(62)
+    !end if
+
+  end subroutine data_for_extra_history_columns
+
+\end{lstlisting}
+
+\begin{lstlisting}[language=Fortran, float, caption={Ejemplo de asignación de valores calculados para la zona convectiva a los parámetros extras definidos.}, label={lst:extra_name_columns}]
+  subroutine other_torque_mag_brk(id, ierr)	
+    ...
+
+    !Routine activated
+    s% x_ctrl(25) = activated
+
+    !Convective zone info
+    s% x_ctrl(26) = sz_info_ptr% top_radius / s% r(1)
+    s% x_ctrl(27) = sz_info_ptr% bot_radius / s% r(1)
+    s% x_ctrl(28) = sz_info_ptr% top_mass / s% m(1)
+    s% x_ctrl(29) = sz_info_ptr% bot_mass/ s% m(1)
+    s% x_ctrl(30) = sz_info_ptr% top_zone
+    s% x_ctrl(31) = sz_info_ptr% bot_zone
+    s% x_ctrl(32) = sz_info_ptr% top_vrot
+    s% x_ctrl(33) = sz_info_ptr% bot_vrot
+    s% x_ctrl(34) = sz_info_ptr% top_omega
+    s% x_ctrl(35) = sz_info_ptr% bot_omega
+    ...
+
+  end subroutine other_torque_mag_brk
+\end{lstlisting}
 
 \subsection{Módulos de interés de MESA}
-Para empezar, vamos a enumerar los módulos de MESA que hemos encontrado de especial relevancia a estudiar para el objetivo de este trabajo.
-Principalmente hemos basado nuestra búsqueda en localizar aquéllos que pueden tener relación directa con los procesos de destrucción del Li o con las reacciones nucleares que se producen en el interior estelar.
+Para empezar, vamos a enumerar los módulos de MESA que hemos encontrado de especial relevancia a estudiar para los objetivos de nuestra línea de investigación.
+Principalmente hemos basado nuestra búsqueda en localizar aquéllos que pueden tener relación directa con los procesos de destrucción del litio o con las reacciones nucleares que se producen en el interior estelar.
 Para cada uno de ellos ofrecemos un pequeño resumen.
@@
 
-Como indicaremos más detalladamente cuando analicemos la configuración final del fichero de proyecto inlist, las reacciones que más nos interesan son las de las cadenas protón-protón en sus variantes I, II y III (ver sección \ref{sim_mesa} para más detalles).\par
+Como indicaremos más detalladamente cuando analicemos la configuración final del fichero de proyecto \textit{inlist}, las reacciones que más nos interesan son las de las cadenas protón-protón en sus variantes I, II y III (ver sección \ref{sim_mesa} para más detalles).\par
 
 \subsubsection{Teoría de longitud de mezcla – mlt}
-El módulo \textit{mlt} implementa la teoría de longitud de mezcla estándar, o por sus siglas en inglés mixing lenght theory (MLT) \cite{Paxton2011}.
-En la dinámica de fluidos, el modelo o teoría de longitud de mezcla es una parametrización que intenta describir la transferencia de calor y material causado por una inestabilidad convectiva en el seno de un fluido.
+El módulo \textit{mlt} implementa la teoría de longitud de mezcla estándar (MLT) \citep{Paxton2011}.
+En la dinámica de fluidos, MLT es una parametrización que intenta describir la transferencia de calor y material causado por una inestabilidad convectiva en el seno de un fluido.
 Este modelo ha sido utilizado en numerosos campos, incluyendo la ciencia atmosférica, oceanografía y estructura estelar, que es el caso que nos atañe.
-La “longitud de mezcla” se define como la distancia recorrida por una celda o burbuja de material sometida a una inestabilidad térmica convectiva.\par
+La longitud de mezcla se define como la distancia recorrida por una celda o burbuja de material sometida a una inestabilidad térmica convectiva.\par
 
 \subsubsection{Módulo de difusión – diffusion} \label{subsec_diffusion}
-El módulo \textit{diffusion} lo utiliza MESA para calcular la difusión de partículas y la sedimentación gravitacional resolviendo las ecuaciones de Burger \cite{Burgers1969} a través del método propuesto por \cite{Thoul1993}.
-El módulo de difusión trata los elementos presentes en el modelo estelar como pertenecientes a "clases" definidas por el usuario en términos de rangos de masas atómicas.
+El módulo \textit{diffusion} lo utiliza MESA para calcular la difusión de partículas y la sedimentación gravitacional resolviendo las ecuaciones de Burger \citep{Burgers1969} a través del método propuesto por \cite{Thoul1993}.
+El módulo de difusión trata los elementos presentes en el modelo estelar como pertenecientes a clases definidas por el usuario en términos de rangos de masas atómicas.
 Para cada clase, el usuario especifica un isótopo representativo y todos los miembros de esa clase son tratados idénticamente, con sus velocidades de difusión determinadas por el isótopo representativo.
-La rutina utiliza la fracción de masa de la clase correspondiente para obtener su coeficiente de difusión \cite{Paxton2015}.
-Como detallaremos más adelante, esta opción por defecto del módulo no es la más adecuada para obtener las abundancias de Li.\par
+La rutina utiliza la fracción de masa de la clase correspondiente para obtener su coeficiente de difusión \citep{Paxton2015}.
+Como detallaremos más adelante, esta opción por defecto del módulo no es la más adecuada para obtener las abundancias de litio.\par
 
@@
 \subsubsection{Otros parámetros de interés}
-Además de los módulos listados anteriormente existen otra serie de parámetros que ofrece MESA que están relacionados también con el planteamiento teórico presentado en este trabajo.
-En concreto nos estamos refiriendo a los parámetros de masa estelar, contenido de H y He presente en la nube protoestelar, así como la metalicidad de la misma (Z).\par
+Además de los módulos listados anteriormente existen otra serie de parámetros que ofrece MESA que están relacionados también con el planteamiento teórico de nuestra investigación.
+En concreto nos estamos refiriendo a los parámetros de masa estelar, contenido de \isotope[1]{H}, \isotope[3]{He} y \isotope[4]{He} presente en la nube protoestelar, así como la metalicidad de la misma (Z).\par
 
-Finalmente, también tendremos en cuenta la rotación superficial de la estrella, ya que, como se ha venido diciendo, este fenómeno físico es considerado un factor que influye en el agotamiento del Li, tanto potenciándolo, como llegando a inhibirlo.\par
+Finalmente, también tendremos en cuenta la rotación superficial de la estrella, ya que, como se ha venido comentando, este fenómeno físico es considerado un factor que influye en el agotamiento del litio, tanto potenciándolo, como llegando a inhibirlo.\par
 
@@
 COMENTAR LA PARTE de pgstart
+
 \subsection{Visualización con terceras herramientas}
@@
 Quizás porque no muestra el parámetro en el que estamos interesados, o si lo hace, deseamos otro tipo de formato de visualización.
-Aunque en la mayoría de los casos, la necesidad de recurrir a terceras herramientas para la visualización se debe a que necesitamos realizar algún tipo de preprocesamiento sobre los mismos.
-(ENUMERAR LAS DIFERENTES ETAPAS DE TRATAMIENTO DE LOS DATOS)
+Aunque en la mayoría de los casos, la necesidad de recurrir a terceras herramientas para la visualización se debe a que necesitamos realizar algún tipo de preprocesamiento (ver \ref{sec:gemelos_solares} para más detalles) sobre los mismos antes de proceder a su representación e interpretación final.\par
 
-En nuestra investigación hemos recurrido a la herramientas \textit{Octave} (ver \ref{sec:tool_octave}) para las etapas de (ENUMERAR LAS ETAPAS) y a \textit{TopCat} (ver \ref{sec:tool_topcat}) para ....
+En nuestra investigación hemos recurrido a la herramientas Octave (ver \ref{sec:tool_octave}) para la extracción, transformación, filtrado, enriquecimiento y visualización de los datos.
+En Octave establecemos el un flujo de trabajo sobre los ficheros de salida de MESA, cuyas etapas:
+\begin{enumerate}
+	\item Extracción de datos - En esta fase, extraemos los datos de los ficheros de salida de MESA. El objetivo es obtener los datos brutos arrojados por el simulador para su posterior procesamiento.
+	\item Transformación de datos - Los datos extraídos se someten a una serie de transformaciones para limpiarlos, enriquecerlos y estructurarlos de manera coherente.
+Las tareas comunes de transformación incluyen:
+	\begin{enumerate}
+		\item Limpieza de datos - Eliminación de valores nulos, corrección de errores y estandarización
+		\item Conversión de formatos - Cambio de tipos de datos, ajuste de formatos de fecha, etc
+		\item Agregación - Resumen de datos mediante cálculos (por ejemplo, sumas, promedios)
+		\item Derivación - Creación de nuevas columnas a partir de las existentes (por ejemplo, calcular velocidades radiales a partir de las angulares)
+	\end{enumerate}
+	\item Carga -  En esta fase, los datos transformados pasan a ser visualizados en diferentes tipos de gráficas y quedan almacenados para su posterior interpretación. 
+\end{enumerate}
 
 \section{Extendiendo MESA - Parte I} \label{mesa_parte1}
-Una vez obtenida una visión relativamente aceptable de las capacidades del simulador, el siguiente paso lógico consistió en ejercitar simulaciones teniendo en mente los siguientes propósitos:
+Una vez obtenida una visión relativamente aceptable de las capacidades del simulador, la etapa a considerar es la de obtener un obtener un marco de referencia para nuestras simulaciones.
+Éste nos debe de asistir en la consecución de los siguientes objetivos:
 
@@
 
-Con estos dos objetivos principales en mente, la concepción de los diferentes escenarios de simulación se realizó de forma condicionada al planteamiento teórico analizado en la primera parte de este trabajo.
-En particular, estos escenarios deberían de tener en cuenta los siguientes procesos/características que intervienen, de alguna u otra forma, en la evolución de las abundancias de Li:
+Con estos dos objetivos principales en mente, la concepción de los diferentes escenarios de simulación se realizó de forma condicionada al planteamiento teórico que introduciremos en el capítulo \ref{parte1}.
+En particular, estos escenarios soportan los siguientes procesos/características que intervienen, de alguna u otra forma, en la evolución de las abundancias de litio:
 
 \begin{itemize}
-    \item Reacciones nucleares en las que interviene el Li
+    \item Reacciones nucleares en las que interviene el litio
     \item Metalicidad y velocidad de rotación de la estrella
@@
     \item Procesos de sedimentación
-    \item Presencia de campo magnético de intensidad fija y variable
-    \item Evolución temporal de $\amlt$
+    \item Presencia de campo magnético de intensidad fija
 \end{itemize}
 
-A la hora de extrapolar el subyacente teórico de este trabajo a los escenarios de simulación, se decidió hacer una división en base al soporte que ofrecía MESA de manera estándar para los conceptos teóricos que hemos acabado de enumerar y para aquéllos que no.
-De esta forma, nos encontramos que, a excepción del punto 5 y 6, el resto estaban soportados por MESA.
+A la hora de extrapolar el subyacente teórico de nuestra línea de investigación a los escenarios de simulación, se opta por hacer una división en base al soporte que ofrece MESA de manera estándar para los conceptos teóricos que hemos acabado de enumerar y para aquéllos que no.
+De esta forma, nos encontramos que a excepción del punto 5, el resto se encuentran soportados por MESA.
 A partir de esta división se establecieron dos líneas de trabajo complementarias.
-La primera encaminada a simular los procesos teóricos 1-4 que influyen sobre la evolución del Li, por un lado, y a utilizar los valores generados por esas simulaciones como marco de referencia por otro.
-La segunda línea de trabajo se encaminó a extender MESA para que soportara la influencia de campos magnéticos y $\amlt$ variable en la evolución de su estrella anfitriona.\par
+La primera encaminada a simular los procesos teóricos 1-4 ya soportados por MESA, y una segunda, más avanzada, encaminada a extender el simulador con nuestra propia rutina de frenado magnético.\par
 
 \subsection{Simulaciones acordes al marco teórico estándar de MESA}
-En esta línea de trabajo nos centramos en hacer uso de las funciones y parametrizaciones estándar que ofrece MESA para dar soporte a los puntos teóricos enumerados anteriormente y que recogemos de nuevo aquí:
+En esta línea de trabajo nos centramos en hacer uso de las funciones y parametrizaciones estándar que ofrece MESA para dar soporte a los puntos teóricos 1-4 enumerados anteriormente y que recogemos de nuevo aquí:
 
@@
 
-Como estamos interesados en observar cómo actúan esos y procesos parámetros físicos sobre la abundancia del Li en las fotosferas de estrellas de tipo solar (1.0 $\msun$) a lo largo de su evolución desde la PMS hasta la TAMS, nuestros modelos de simulación comienzan por fijar los siguientes parámetros:
+Como estamos interesados en observar cómo actúan esos procesos y parámetros físicos sobre la abundancia de litio en las fotosferas de estrellas de tipo solar (1.0 $\msun$) a lo largo de su evolución desde la PMS hasta la TAMS, nuestros modelos de simulación comienzan por fijar los siguientes parámetros:
 
@@
     \item Establecer una metalicidad similar al Sol
-    \item Detener la simulación al alcanzar la Termination Age Main Sequence (TAMS)
+    \item Detener la simulación al alcanzar la TAMS
     \item Obtener las abundancias para los elementos H, He, Be
@@
 
-La composición química del Sol, incluyendo su metalicidad y masa inicial del modelo quedó fijada por medio de los siguientes parámetros (Listado \ref{lst:ini_params}) de MESA:
+La composición química del Sol, incluyendo su metalicidad y masa inicial del modelo queda fijada por medio de los siguientes parámetros (Listado \ref{lst:ini_params}) de MESA:
 
-\begin{lstlisting}[language=Fortran, caption={Parametrización de la masa inicial de la estrella}, label={lst:ini_params}]
+\begin{lstlisting}[language=Fortran, float, caption={Parametrización de la masa inicial de la estrella}, label={lst:ini_params}]
 &star_job
@@
 Posteriormente necesitamos indicar la condición de parada en la TAMS.
-Este momento en la evolución de una estrella queda asociado a la etapa en la que la estrella abandona la secuencia principal al haber agotado prácticamente todo el H que se encontraba en su núcleo.
-MESA nos ofrece la posibilidad establecer una condición de parada cuando las abundancias de ciertos elementos en el núcleo de la estrella alcanzan un cierto valor máximo o mínimo (Figura 8XXXXXXX?).
-En nuestro caso, utilizamos esta opción para indicar que cuando la abundancia del isótopo \isotope[1]{H} caiga por debajo de 0.000000001, la simulación se detenga (ver Listado \ref{lst:stop_params}.\par
+Este momento en la evolución de una estrella queda asociado a la etapa en la que la estrella abandona la MS al haber agotado prácticamente todo el H que se encontraba en su núcleo.
+MESA nos ofrece la posibilidad establecer una condición de parada cuando las abundancias de ciertos elementos en el núcleo de la estrella alcanzan un cierto valor máximo o mínimo.
+En nuestro caso, utilizamos esta opción para indicar que cuando la abundancia del isótopo \isotope[1]{H} caiga por debajo de 0.000000001, la simulación se detenga (ver Listado \ref{lst:stop_params}).\par
 
-Para poder obtener las abundancias de Li AL final de la TAMS, primeramente necesitamos conocer qué reacciones nucleares son las responsables de su creación y si éstas están o no incluidas por defecto en la configuración de MESA.
+Para poder obtener las abundancias de litio al final de la TAMS, primeramente necesitamos conocer qué reacciones nucleares son las responsables de su creación y si éstas están o no incluidas por defecto en la configuración de MESA.
 Como se adelantó en la sección \ref{sec:li_reac_nuc}, estas reacciones son las de cadena protón-protón en sus variantes I, II y III.\par
 
-\begin{lstlisting}[language=Fortran, caption={Parametrización de la condición de parada de la simulación cuando se produce el agotamiento de H en el núcleo de la estrella}, label=, label={lst:stop_params}]
+\begin{lstlisting}[language=Fortran, float, caption={Parametrización de la condición de parada de la simulación cuando se produce el agotamiento de H en el núcleo de la estrella}, label={lst:stop_params}]
 &controls
@@
 
-Por tanto, ya tenemos identificas la red de reacciones a activar en la configuración de MESA: reacciones nucleares de tipo protón-protón.
-Es en este punto donde volvemos a hacer referencia al módulo net (ver \ref{reac_nuc}).
+Por tanto, ya tenemos identificas la red de reacciones a activar en la configuración de MESA.
+Es en este punto donde volvemos a hacer referencia al módulo \textit{net} (ver \ref{reac_nuc}).
 Si recordamos, se indicaba que MESA viene pre configurado con una red de reacciones nucleares en las que sólo trata con 8 isótopos.
@@
 
-Con los parámetros anteriores incluidos en el proyecto, indicamos a MESA que sustituimos la red de reacciones nucleares por defecto por una nueva denominada \textit{pp\_extras.net}.
+Con los parámetros anteriores incluidos en el proyecto, indicamos a MESA que sustituimos la red de reacciones nucleares por defecto por una nueva denominada \textit{mesa\_49.net} (Listado \ref{lst:mesa49}).
 Las redes son simples ficheros de texto en los que se indica al simulador las reacciones nucleares que debe tener en cuenta y cuál es el aspecto de éstas.
-En concreto, en la elegida le indicamos que tenga en cuentas las reacciones protón-protón en sus variantes I, II, III.\par
+En concreto, en la elegida se soportan, entre otras, las reacciones protón-protón en sus variantes I, II, III.\par
 
-Una vez activada la nueva red de reacciones nucleares en nuestro fichero de proyecto, conseguimos obtener las abundancias para el Li, las cuales son informadas en los ficheros de salida estándar que genera MESA.
-Adicionalmente estamos interesados en obtener el perfil de evolución de estos elementos a lo largo del tiempo de evolución de la estrella y en función del radio de la estrella (Figura 13XXXXXXXXXX).
-Para ello recurrimos a utilizar las capacidades gráficas de MESA y mostramos la abundancia de estos elementos.
-Dado que sus concentraciones son muy pequeñas en comparación del H, debemos de proceder a ajustar el eje de abscisas del gráfico.\par
+\begin{lstlisting}[language=Fortran, float, caption={Red de reacciones mesa\_49.net. Esta red incluye un tratamiento detallado de la para los elementos ligeros H, He, Li y Be}, label={lst:mesa49}]
+  add_isos_and_reactions(
+    neut
+    h  1  2  ! hydrogen
+    he 3  4  ! helium
+    li7      ! lithium
+    be7
+    be 9 10  ! berylium
+    b8       ! boron
+    c 12 13  ! carbon
+    n 13 15  ! nitrogen
+    o 14 18  ! oxygen
+    f 17 19  ! fluorine
+    ne 18 22 ! neon
+    na 21 24 ! sodium
+    mg 23 26 ! magnesium
+    al 25 27 ! aluminum
+    si 27 30 ! silicon
+    p 30 31  ! phosphorus
+    s 31 34  ! sulfur
+  )
 
-Dado que nuestro modelo solo abarca las fases evolutivas PMS y MS, decidimos eliminar las reacciones en las que interviene el Fe ya que no vamos a evolucionar nuestra estrella hasta el punto en que comienza a generarse este elemento.
-Adicionalmente, configuramos MESA para que, a cada paso en la simulación y para cada celda, tenga en cuenta si se producen o dejan de existir isótopos a consecuencia de las reacciones nucleares que están teniendo lugar.
-Dados los valores de abundancias tan pequeños que se obtienen para el Li y Be, del orden de $10^-9$ para el primero y $10^-31$ para el segundo, optamos por activar esta opción para que el tratamiento de las reacciones nucleares sea lo más riguroso y exacto posible.\par
+  add_isos(
+    ca40 ! calcium
+    ti48 ! titanium
+    fe56 ! iron
+ )
+\end{lstlisting}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{img/tesis/abundances.pdf}
-    \caption {Gráfica de abundancias en la que aparecen los elementos H, He, Fe, Li y Be. El eje de abscisas ha sido reescalado}
-    \label{fig:abundances}
-\end{figure}
-
-Continuando con el marco físico a simular, pasamos a centrarnos en la activación de los mecanismos que, acorde a la teoría, tienen influencia en el agotamiento del Li:
+Continuando con el marco físico a simular, pasamos a centrarnos en la activación de los mecanismos que, acorde a la teoría, tienen influencia en el agotamiento del litio:
 \begin{itemize}
@@
         {\small
-        (MESA) calcula la difusión de partículas y la sedimentación gravitacional resolviendo las ecuaciones de Burger a través del método propuesto por Thoul \cite{Thoul1993}.
+        (MESA) calcula la difusión de partículas y la sedimentación gravitacional resolviendo las ecuaciones de Burger a través del método propuesto por Thoul \citep{Thoul1993}.
 El módulo de difusión trata los elementos presentes en el modelo estelar como pertenecientes a "clases" definidas por el usuario en términos de rangos de masas atómicas.
@@
 
-De esta información se desprende que el hecho de que al activar la difusión también activamos la sedimentación gravitacional y, de manera indirecta, el módulo \textit{mlt} encargado de calcular los coeficientes de difusión que entran en juego en proceso de mezclado producido a consecuencia de los movimientos convectivos que se produce en el interior de la estrella.\par
+Activando la difusión en MESA también activamos la sedimentación gravitacional.
+De manera indirecta, queda también involucrado el módulo \textit{mlt}.
+Recordamos que este módulo es el encargado de calcular los coeficientes de difusión que entran en juego en el proceso de mezclado producido a consecuencia de los movimientos convectivos que se produce en el interior de la estrella.\par
 
-Cabe también mencionar el tratamiento por “clases” que realiza MESA.
+Cabe también mencionar el tratamiento por clases que realiza MESA.
 Los isótopos presentes en nuestra red de reacciones nucleares que tienen un peso atómico similar son tratados de forma conjunta por la rutina de difusión en lo que se refiere a su velocidad de difusión.
-Ésta queda determinada por el isótopo representativo de la “clase”.
+Ésta queda determinada por el isótopo representativo de la clase.
 La agrupación en clases tiene la ventaja de que las simulaciones se ejecutan de una manera más rápida ya que se ahorran cálculos.
@@
 
+Una vez activada la nueva red de reacciones nucleares en nuestro fichero de proyecto MESA, conseguimos obtener las abundancias para el litio, las cuales son informadas en los ficheros de salida estándar que genera MESA.
+Adicionalmente estamos interesados en obtener el perfil de evolución de estos elementos a lo largo del tiempo de evolución de la estrella y en función del radio de la estrella (Figura \ref{fig:abundances}).
+Para ello recurrimos a utilizar las capacidades gráficas de MESA y mostramos la abundancia de estos elementos.
+Dado que sus concentraciones son muy pequeñas en comparación al H, debemos de proceder a ajustar el eje de abscisas del gráfico.\par
+
+Adicionalmente, configuramos MESA para que, a cada paso en la simulación y para cada celda, tenga en cuenta si se producen o dejan de existir isótopos a consecuencia de las reacciones nucleares que están teniendo lugar.
+Dados los valores de abundancias tan pequeños que se obtienen para el litio, del orden de -20 respecto al logaritmo de la masa total de la estrella, optamos por activar el tratamiento de las reacciones nucleares sea lo más riguroso y exacto posible.\par
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.5\textwidth]{img/tesis/abundances.pdf}
+	\caption {Gráfica de abundancias en la que aparecen los elementos H, He, Fe, Li y Be. El eje de abscisas ha sido reescalado para posibilitar la representación de la abundancia de litio y berilio.}
+	\label{fig:abundances}
+\end{figure}
+
 En lo que se refiere a la rotación de las capas superficiales de la estrella, MESA permite especificar una velocidad en km/s, que en nuestro caso ha quedado fijada en 20 km/s al ser un valor similar al que tenemos en el Sol.\par
 
-En el Listado \ref{lst:net_params} mostramos la configuración adicional a añadir al proyecto de MESA para poder simular los conceptos teóricos discutidos hasta este momento.
-\begin{lstlisting}[language=Fortran, caption={Parametrización de los procesos relacionados con el Li y de las gráficas de salida.}, label={lst:net_params}]
+En el Listado \ref{lst:li_evol} mostramos la configuración adicional a añadir al proyecto de MESA para poder simular los conceptos teóricos discutidos hasta este momento.
+\begin{lstlisting}[language=Fortran, float, caption={Parametrización de los procesos relacionados con el tratamiento del litio y su evolución temporal.}, label={lst:li_evol}]
 &star_job
   ! Change default net in order to include 
-  !reactions for Li and Be
+  !reactions for Li
   change_initial_net = .true.
-  new_net_name = 'pp_extras.net'
+  new_net_name = 'mesa_49.net'
   ! If enable_adaptive_network is true, 
@@
 
-Llegado este punto, hemos conseguido obtener una configuración de proyecto base que recoge las características esenciales de la estrella a simular en MESA, pone en marcha los procesos físicos relacionados con la evolución del Li y es capaz de informar, tanto a través de los ficheros de salida estándar como de las representaciones gráficas, sobre las abundancias de Li a lo largo de la evolución de la estrella en sus etapas de PMS y MS.
-Como decimos, esta configuración de proyecto y los valores arrojados por la simulación que lo utiliza, nos servirán para establecer el marco de referencia contra el que comparar los valores obtenidos en las nuevas simulaciones que extienden la funcionalidad de MESA.\par
+Llegado este punto, hemos conseguido obtener una configuración de proyecto base que recoge las características esenciales de la estrella a simular en MESA: pone en marcha los procesos físicos relacionados con la evolución del litio y es capaz de informar, tanto a través de los ficheros de salida estándar como de las representaciones gráficas, sobre las abundancias de este elemento a lo largo de la evolución de la estrella en sus etapas de PMS y MS.
+Esta configuración de proyecto y los valores arrojados por la simulación que lo utiliza, nos servirán para establecer el marco de referencia contra el que comparar los valores obtenidos en las nuevas simulaciones que extienden la funcionalidad de MESA.\par
 
---CONTINUAR POR AQUÍ --
+--CONTINUAR POR AQUI--
 

### 2024-03-16 project/overleaf/Tesis III/capitulos/capitulo06.tex
WIP Tesis
WIP

@@
 --EXPLICAR POR QUÉ OPTAMOS POR LA INTENSIDAD VARIABLE
-
+La segunda línea de trabajo se encaminó a extender MESA para que soportara la influencia de campos magnéticos y $\amlt$ variable en la evolución de su estrella anfitriona.
+Adicionalmente a lo de la parte 1
+\begin{itemize}
+	\item Presencia de campo magnético de intensidad variable
+	\item Evolución temporal de $\amlt$
+\end{itemize}
 \section{Efectos del frenado magnético - Parte II}
 
-\subsection{Formalismo de frenado magnético de intesidad fija}
+\subsection{Formalismo de frenado magnético de intesidad variable}
 

### 2024-03-16 project/overleaf/Tesis III/capitulos/capitulo08.tex
WIP

--- 2024-01-30 project/overleaf/Tesis III/capitulos/capitulo08.tex
@@
 La ciencia de datos es un campo interdisciplinario que utiliza algoritmos, procedimientos y procesos para examinar grandes cantidades de datos con el fin de descubrir patrones ocultos, generar conocimientos y dirigir la toma de decisiones.
-Utiliza técnicas y teorías extraídas de muchos campos dentro del contexto de las matemáticas, estadísticas, programación, análisis, IA y aprendizaje automático para descubrir conocimientos accionables a partir de conjuntos de datos.\par
+Utiliza técnicas y teorías extraídas de muchos campos dentro del contexto de las matemáticas, estadísticas, programación, análisis y aprendizaje automático para descubrir conocimientos accionables a partir de conjuntos de datos.\par
 
@@
 Del mismo modo, nuestro modelo no ha sido incorporado, integrado con ningún sistema.
-De momento este no ha sido el caso pero no sería una insensatez pensar que pudiese pasar a incorporarse como una extensión a la herramienta que hemos utilizado para nuestras simulaciones, si lo autores de la misma lo considerasen oportuno.
+De momento este no ha sido el caso pero no sería descabellado pensar que pudiese pasar a incorporarse como una extensión a la herramienta que hemos utilizado para nuestras simulaciones, si lo autores de la misma lo considerasen oportuno.
+
+\section{Catálogos de datos}
+En la vastedad del cosmos, los catálogos de datos se alzan como pilares fundamentales para los investigadores en astrofísica.
+Estos catálogos, que contienen descripciones detalladas y metadatos, permiten comprender y acceder a la riqueza de información cósmica.
+Proporcionan contexto esencial para comprender los datos recopilados por telescopios y observatorios, permitiendo a los científicos explorar las coordenadas celestiales, las magnitudes estelares y las características espectrales.\par
+
+Sus propiedades fundamentales son múltiples: en primer lugar, los catálogos permiten una búsqueda eficiente.
+En un campo donde la cantidad de datos es abrumadora, esta eficiencia es vital.
+Además, fomentan la colaboración científica al proporcionar un acceso centralizado.
+Los metadatos, que describen y resumen la entradas de los catálogos, son la clave para evaluar la calidad y actualidad de los datos.\par
+
+Los catálogos proporcionan una base de referencia para comparar las propiedades simuladas con las observaciones reales.
+Además, nos permiten calibrar los parámetros de los modelos y refinarlos mediante la comparación detallada entre las características de las estrellas simuladas y las observadas.\par
+
+\subsection{Gaia}
+El catálogo Gaia \citep{Mignard2005} es un mapa tridimensional completo de la Vía Láctea, creado utilizando datos del telescopio espacial Gaia.
+Éste contiene datos de alrededor de 1.46 millardos de estrellas, entre los que se encuentra sus posiciones en el cielo, paralaje y movimiento propio.
+Además, incluye sus magnitudes en diferentes bandas espectrales para una gran cantidad de ellas.
+El catálogo Gaia es particularmente útil para estudiar estrellas similares al Sol, ya que proporciona un censo único de estrellas dentro de 100 pc de nuestro Sol.
+Esto lo convierte en un recurso invaluable para los astrónomos y astrofísicos que buscan comprender mejor las propiedades y la evolución de estas estrellas.\par
+
+\subsection{Gaia-ESO Spectroscopic Survey - (GES)}
+El estudio Gaia-ESO (Gaia-ESO Survey, GES) \citep{Gilmore2012,Randich2013,Randich2022} es un catálogo público enfocado al análisis espectroscópico de un número importante de estrellas recogidas en el catálogo Gaia.
+Se llevó a cabo con el instrumento espectrógrafo multielemento de fibra de gran conjunto (Fibre Large Array Multi Element Spectrograph, FLAMES) y el espectrógrafo Echelle ultravioleta y visual (Ultraviolet and Visual Echelle Spectrograph, UVES) en el VLT.
+Su objetivo es proporcionar una visión homogénea de las distribuciones de cinemática y abundancias elementales en la Vía Láctea, en el campo y en cúmulos abiertos (Open Cluster, OC), hasta magnitud 19.\par
+
+Dentro de la iniciativa GES se han observado más de 110 000 estrellas, proporcionando velocidades radiales y rotacionales proyectadas, parámetros estelares tales como: temperatura efectiva, gravedad superficial y metalicidad, abundancias de varios elementos, entre los que se encuentra el litio, para aproximadamente 1/3 de la muestra, y parámetros específicos para rastrear la acreción y actividad en estrellas jóvenes.
+El catálogo GES es único en lo que se refiere a la observación de estrellas de todos los tipos espectrales con análisis dedicados y especializados, lo que la hace particularmente útil para estudiar estrellas similares al Sol.\par
 
 \section{Herramientas}
-\subsection{Modules for Experiments in Stellar Astrophysics - (MESA)}
-Para la realización de este tesis doctoral nos hemos apoyado en la herramienta de evolución estelar \textit{Modules for Experiments in Stellar Astrophysics} (MESA).
+\subsection{Modules for Experiments in Stellar Astrophysics - MESA}
+Para la realización de este tesis doctoral nos hemos apoyado en la herramienta de evolución estelar MESA (Modules for Experiments in Stellar Astrophysics).
 MESA ha jugado un papel fundamental en nuestra investigación y por ello, debido a su importancia, le hemos dedicado un capítulo en exclusiva en el que entramos a fondo en sus características, estructura, módulos que lo componen, y muy especialmente en cómo extender la física que incorpora sus modelos mediante la programación de nuestras propias rutinas.\par
@@
 Es un conjunto de software libre y de código abierto que proporciona una interfaz de línea de comandos conveniente para resolver problemas lineales y no lineales numéricamente.
-La sintaxis de Octave es en gran medida compatible con MATLAB, lo que lo convierte en una alternativa popular para los usuarios que necesitan una solución rentable.
+La sintaxis de Octave es en gran medida compatible con MATLAB, lo que lo convierte en una alternativa popular para los usuarios que necesitan una solución rentable.\par
 
-Octave proporciona capacidades para la solución numérica de problemas lineales y no lineales, y para realizar otros experimentos numéricos.
-También proporciona amplias capacidades gráficas para la visualización y manipulación de datos.
+Esta herramienta también proporciona amplias capacidades gráficas para la visualización y manipulación de datos.
 Octave puede ser extendido por paquetes, permitiendo a los usuarios añadir más funcionalidades según sea necesario.
-Puede ser ejecutado en modo GUI, como una consola, o invocado como parte de un script de shell.
+Puede ser ejecutado en modo visual, como una consola, o invocado como parte de un script de entorno.
 Octave está disponible para varios sistemas operativos, incluyendo GNU/Linux, macOS, BSD y Microsoft Windows.
 
-\subsection{Tool for OPerations on Catalogues And Tables (TOPCAT)} \label{sec:tool_topcat}
+\subsection{Tool for OPerations on Catalogues And Tables - TOPCAT} \label{sec:tool_topcat}
 TOPCAT, acrónimo de \textit{Tool for OPerations on Catalogues And Tables}, es un visor y editor gráfico altamente versátil e interactivo para datos tabulares.
@@
 Además, proporciona funciones estadísticas para analizar los datos, y capacidades de coincidencia cruzada para comparar diferentes catálogos.
-La herramienta también soporta \textit{scripting} para tareas automatizadas.
+La herramienta también soporta scripting para tareas automatizadas.
 Estas características hacen de TOPCAT una herramienta integral para la extracción y análisis de datos en el campo de la astronomía y más allá.\par
 
-Aunque TOPCAT fue diseñado principalmente para trabajar con datos astronómicos, también puede ser utilizado para cualquier tipo de datos tabulares.
-Esto significa que puede ser útil en una variedad de campos que requieren la manipulación y análisis de datos tabulares.
-Por ejemplo, en la investigación científica, los datos a menudo se presentan en forma tabular y requieren análisis detallados.
-TOPCAT puede ser útil aquí para visualizar y analizar estos datos.
-Además, en el campo de la informática, los datos tabulares son comunes, por ejemplo, en bases de datos o archivos CSV.
-Por lo tanto, aunque TOPCAT tiene un enfoque en la astronomía, sus capacidades lo hacen aplicable en una variedad de campos donde se manejan datos tabulares.
-Sin embargo, los detalles específicos pueden depender del tipo de datos y de las necesidades del usuario.\par
-
-TOPCAT es particularmente útil para el \textit{Virual Observatory} (VO).
+TOPCAT es particularmente útil para el observatorio virtual (Virtual Observatory, VO).
 Ésta es una aplicación de escritorio para el análisis interactivo de datos en formato tabular, especialmente catálogos de fuentes.
 TOPCAT puede acceder a servicios de datos externos.
-También puede comunicarse con otras herramientas astronómicas a través del protocolo \textit{Simple Application Messaging Protocol} (SAMP).
+También puede comunicarse con otras herramientas astronómicas a través del protocolo simple de mensajería de aplicaciones (Simple Application Messaging Protocol, SAMP).
 Esto permite una integración perfecta de TOPCAT en el entorno del VO, permitiendo a los usuarios recuperar y analizar datos de varios servicios.\par
@@
 Esto hace posible trabajar con una amplia variedad de catálogos astronómicos.
-Por ejemplo, puedes usar TOPCAT para acceder a un catálogo de estrellas contenidas en la base de datos Gaia.
+Por ejemplo, se puede usar TOPCAT para acceder a un catálogo de estrellas contenidas en la base de datos Gaia.
 Además, TOPCAT ofrece características para el cruce de correspondencias, que es crucial al comparar diferentes catálogos.
-También admite \textit{scripting} para tareas automatizadas, lo cual puede ser particularmente útil al trabajar con catálogos grandes o realizar tareas repetitivas.\par
+En el Listado \ref{lst:consulta} podemos observar la consulta que hemos utilizado para seleccionar del catálogo GES las estrellas pertenecientes a OC's con un nivel de confianza >= 95\%.
+Para cada una de ellas obtenemos su temperatura efectiva, la gravedad superficial, su metalicidad, la abundancia de Li, y los valores de error asociados a cada uno de esos parámetros.\par
 
-En resumen, TOPCAT es una herramienta integral que proporciona capacidades robustas para trabajar con el VO y extraer datos de varios catálogos astronómicos.
-Está diseñado para facilitar la manipulación de tablas, para que los usuarios puedan concentrarse en hacer ciencia.\par
-
-\begin{lstlisting}[language=SQL, caption={Consulta TOPCAT sobre el catálogo GES DR5 para obtener los componentes de los cúmulos abiertos para los cuales se ha determinado una pertenecia mínima del 0.95\%.
+\begin{lstlisting}[language=SQL, float, caption={Consulta TOPCAT sobre el catálogo GES DR5 para obtener los componentes de los cúmulos abiertos para los cuales se ha determinado una pertenecia mínima del 0.95\%.
 Las columnas obtenidas informan sobre la identificador de la estrella, $\teff$, $\gsurf$, $\feh$, A(Li) y sus errores asociados, y el identificador del cúmulo al que pertenece.}, label={lst:consulta}]
@@
 
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\textwidth]{img/tesis/tap_query.pdf}
-	\caption{PDF}
-	\label{fig:tap_query}
-\end{figure}
+Antes de poder realizar la consulta hay que localizar el catálogo sobre el que ejecutar la búsqueda.
+Como acabamos de comentar, TOPCAT es una herramienta polivalente y que sobresale a la hora de acceder a diferentes fuentes datos.
+Mediante el protocolo de acceso a tablas (Table Access Protocol, TAP) TOPCAT ofrece la posibilidad de realizar búsquedas entre los catálogos que soportan TAP mediante el uso de palabras claves o propiedades de los catálogos.
+En la Figura \ref{fig:tap_ges_service} vemos el resultado ofrecido tras realizar una búsqueda en los catálogos disponibles que están relacionados con GES.\par
 
@@
 	\includegraphics[width=0.7\textwidth]{img/tesis/tap_ges_service.pdf}
-	\caption{PDF}
+	\caption{Búsqueda de catálogos con información de GES.
+Mediante el uso de TAP, TOPCAT es capaz de ofrecer una lista de catálogos que poseen información referente a GES.
+En nuestro caso, hemos seleccionado el catálogo publicado por la ESO.}
 	\label{fig:tap_ges_service}
@@
 
+Un ejemplo similar, pero esta vez orientado a localizar catálogos con información de cúmulos abiertos es el que nos ofrece la Figura \ref{fig:tap_query_oc}.
+Una vez seleccionado el catálogo, TOPCAT nos ofrece la posibilidad de obtener detalles adicionales sobre los esquemas y tablas que los conforman.
+Esta información es muy útil a la hora de realizar nuestras consultas, ya que de ella podemos extraer no solo las columnas incluidas en la diferentes tablas, sino también el tipo de información que contiene (necesario cuando el nombre de la columna no es suficientemente explicativo), tipo de datos (numérico, texto, fecha...), columnas claves, referencia entre tablas (claves foráneas), y más información relevante.
+Este conjunto de información adicional es lo que se conoce como meta-información o meta-datos asociados a la tabla (ver Figura \ref{fig:tap_table_metadata}).\par
+
 \begin{figure}
 	\centering
-	\includegraphics[width=0.7\textwidth]{img/tesis/tap_multi_sources.pdf}
-	\caption{PDF}
-	\label{fig:tap_multi_services}
+	\includegraphics[width=0.7\textwidth]{img/tesis/tap_service.pdf}
+	\caption{Búsqueda de catálogos con información relacionada con cúmulos abiertos.}
+	\label{fig:tap_query_oc}
 \end{figure}
@@
 	\includegraphics[width=0.7\textwidth]{img/tesis/tap_schema.pdf}
-	\caption{PDF}
-	\label{fig:tap_schema}
+	\caption{Vista con meta-información asociada a una tabla.}
+	\label{fig:tap_table_metadata}
 \end{figure}
+
+TOPCAT también es capaz de mostrar las diferentes tablas contenidas dentro de un esquema.
+Un esquema define cómo se organizan los datos dentro, las tablas que los contienen.
+Esto incluye restricciones lógicas, como nombres de tablas, campos, tipos de datos y las relaciones existente entre ellas (ver Figura \ref{fig:tap_schema_detail}).\par
 
@@
 	\includegraphics[width=0.7\textwidth]{img/tesis/tap_schema_detail.pdf}
-	\caption{PDF}
+	\caption{Vista con lista de tablas contenida en un esquema.}
 	\label{fig:tap_schema_detail}
@@
 
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\textwidth]{img/tesis/tap_service.pdf}
-	\caption{PDF}
-	\label{fig:tap_service}
-\end{figure}
-
-\section{Catálogos de datos}
-Una etapa fundamental de toda investigación es la de la confrontación de los datos arrojados con los modelos contra los disponibles en la literatura ya existente,
-
-\subsection{Gaia}
-El catálogo Gaia es un mapa tridimensional completo de la Vía Láctea, creado utilizando datos del telescopio espacial Gaia.
-Éste contiene datos de alrededor de 1.46 mil millones de estrellas, entre los que se encuentra sus posiciones en el cielo, paralaje y movimiento propio.
-Además, incluye sus magnitudes en diferentes bandas espectrales para una gran cantidad de ellas.
-El catálogo Gaia es particularmente útil para estudiar estrellas similares al Sol, ya que proporciona un censo único de estrellas dentro de 100 pc de nuestro Sol.
-Esto lo convierte en un recurso invaluable para los astrónomos y astrofísicos que buscan comprender mejor las propiedades y la evolución de estas estrellas.\par
-
-\subsection{Gaia-ESO Spectroscopic Survey (GES)}
-GES es un catálogo público enfocado al análisis espectroscópico de un número importante de estrellas recogidas en el catálogo Gaia.
-Se llevó a cabo con el instrumento FLAMES en el VLT.
-Su objetivo es proporcionar una visión homogénea de las distribuciones de cinemática y abundancias elementales en la Vía Láctea.
-Dentro de la iniciativa GES se han observado más de 100 000 estrellas, proporcionando velocidades radiales y rotacionales proyectadas, parámetros estelares tales como: temperatura efectiva, gravedad superficial y metalicidad, abundancias de varios elementos, entre los que se encuentra el Litio, y parámetros específicos para rastrear la acreción y actividad en estrellas jóvenes.
-El catálog GES es único en lo que se refiere a la observación de estrellas de todos los tipos espectrales con análisis dedicados y especializados, lo que la hace particularmente útil para estudiar estrellas similares al Sol.\par
-
-\subsection{Combinando Gaia y GES}
-
+\section{Selección de datos}
+\subsection{El binomio Gaia-GES}
 La combinación de los datos de los catálogos Gaia y GES puede proporcionar una visión completa de las propiedades y evolución de las estrellas similares al Sol.
@@
 Esta combinación permite una comprensión más completa de estas estrellas, incluyendo su cinemática, composiciones químicas y potencial para albergar planetas.
-Al proporcionar una visión más completa y detallada de estas estrellas, podemos obtener una comprensión más profunda de las propiedades y evolución de las estrellas similares al Sol, lo que representa una fuente de información que se antoja invaluable para la investigación en astrofísica.
+Al proporcionar una perspectiva más amplia y detallada de estas estrellas, podemos obtener una comprensión más profunda de las propiedades y evolución de las estrellas similares al Sol, lo que representa una fuente de información que se antoja invaluable para la investigación en astrofísica.
 Estamos ante un conjunto de herramientas poderoso para cualquier astrofísico que estudie estrellas similares al Sol.\par
 
-En el caso de las abundancias de Litio, y como ya hemos apuntado, éste es un elemento clave en el estudio de la astrofísica estelar, ya que su abundancia puede proporcionar información valiosa sobre la edad y la historia de las estrellas.
-Sin embargo, la determinación precisa de las abundancias de Litio puede ser un desafío debido a la complejidad de las líneas espectrales del Litio.
+En el caso de las abundancias de litio, y como ya hemos apuntado, éste es un elemento clave en el estudio de la astrofísica estelar, ya que su abundancia puede proporcionar información valiosa sobre la edad y la historia de las estrellas.
+Sin embargo, la determinación precisa de las abundancias de litio puede ser un desafío debido a la complejidad de las líneas espectrales de este elemento.
 GES nos asiste en nuestra investigación aportando su análisis especializado y dedicado de estrellas de todos los tipos espectrales, convirtiéndose así en una herramienta fundamental a la hora de afrontar este desafío.\par
@@
 Por otra parte, tenemos información precisa sobre la velocidad angular de las estrellas.
-Gaia es capaz de medir la velocidad angular de las estrellas con una precisión sin precedentes, y ésta nos aporta información es crucial para estudiar y entender la rotación de las estrellas, su evolución y estructura interna.
-Combinado esta información con las abundancia de Litio podemos establecer hipótesis y validar las mismas acerca del impacto de la velocidad angular sobre el Litio.
+Gaia es capaz de medir la velocidad angular de las mismas con una precisión sin precedentes, y ésta nos aporta información crucial para estudiar y entender su rotación, evolución y estructura interna.
+Combinado esta información con las abundancias de litio podemos establecer hipótesis y validar las mismas acerca del impacto de la velocidad angular sobre este elemento.
 Paralelamente, la velocidad angular juega un papel crucial en el análisis de la presencia e intensidad de un campo magnético en las estrellas.
@@
 
-Como acabamos de adelantar, la investigación de los campos magnéticos en estrellas similares al Sol.
-y en particular de aquéllas que se encuentran en cúmulos abiertos, se beneficia enormemente de los datos proporcionados por ambos catálogos.
+Como acabamos de adelantar, la investigación de los campos magnéticos en estrellas similares al Sol, y en particular de aquéllas que se encuentran en cúmulos abiertos, se beneficia enormemente de los datos proporcionados por ambos catálogos.
 Los campos magnéticos juegan un papel importante en todas las etapas de la evolución estelar.
-En estrellas similares al Sol, se generan en las capas convectivas exteriores.
+En estrellas similares al Sol se generan en las capas convectivas exteriores.
 Estudiar los campos magnéticos a gran escala de estas estrellas puede mejorar nuestra comprensión sobre los mismos y proporcionar restricciones observacionales para los modelos que estudian cómo se generan, la topología que presentan y sus intensidades.
-Los datos de los catálogos nos asisten a la hora de identificar y estudiar las estrellas con alta actividad magnética, lo que puede proporcionar pistas sobre la dinámica interna de las estrellas y su evolución.
-Combinando esta información con la velocidad angular, nos puede proporcionar con indicios de cómo se interrelacionan ambos fenómenos.
+Los datos de los catálogos nos asisten a la hora de identificar y estudiar las estrellas con alta actividad magnética, lo que puede proporcionar pistas sobre la dinámica interna de las mismas y su evolución.
+Combinando esta información con la velocidad angular, nos puede proporcionar indicios de cómo se interrelacionan ambos fenómenos.
 Adicionalmente, cruzando estos datos con la información de los cúmulos abiertos, obtenemos una visión más clara de cómo los campos magnéticos varían entre las estrellas de la misma edad y composición química.\par
+
+\subsection{Selección de parámetros}
+Como indica su nombre, GES pretende complementar los datos sobre paralajes y movimientos propios del satélite Gaia con información extremadamente precisa sobre velocidades radiales (Radial Velocity, RV), litio y química en general.
+La primera publicación de datos de Gaia (GDR1) \citep{Brown2016} data de 2016 y contiene información sobre los primeros 14 meses funcionamiento de la misión.
+La segunda publicación de datos de Gaia (GDR2) \citep{Brown2018} publicada en 2018, contiene posiciones, paralajes y movimientos propios de unos 1 300 millones de fuentes, junto con información fotométrica.
+La tercera publicación de datos de Gaia, la Gaia Early Data Release 3 (GEDR3) \citep{Brown2021} y la Gaia Data Release 3 (GDR3) \citep{Brown2022}, aportan medidas actualizadas y más precisas.
+GES las mejora y en su última versión (DR5.0) incluye todos los parámetros astrofísicos derivados por el consorcio Gaia-ESO \citep{Gilmore2022}.
+De interés para nuestra línea de investigación son:
+\begin{itemize}
+	\item Velocidades radiales y proyectadas
+	\item Parámetros estelares (temperatura efectiva, gravedad superficial y metalicidad)
+	\item Abundancias de varios elementos, entre ellos el litio
+	\item Parámetros específicos para rastrear la acreción y la actividad en estrellas jóvenes
+	\item Probabilidad de pertenencia a cúmulos
+\end{itemize}
+
+\subsection{Selección de cúmulos abiertos}
+GES DR5.0 proporciona información sobre 114 324 estrellas véase \citep[véase][para más detalles]{Gilmore2022}.
+Para nuestra investigación estamos interesados en considerar aquéllas que pertenecen a OCs, para ello seleccionamos aquellos registros GES que cumplen alguna de las siguientes condiciones:
+
+\begin{itemize}
+	\item GE\_CL: observado por GES, campo del programa OC
+	\item GE\_SD\_OC: observado por GES, campo estándar OC
+	\item AR\_CL: Observación del Archivo ESO, campo de programa OC
+	\item AR\_SD\_OC: Observación de archivo, campo estándar OC
+\end{itemize}
+
+Como resultado de este primer filtrado, el número de registros seleccionados se reduce a 43 299.
+Sobre este subconjunto de datos realizamos otro filtrado, esta vez dirigido a seleccionar aquellos que ofrecen información no nula sobre los siguientes atributos:
+
+\begin{itemize}
+	\item Metalicidad
+	\item Abundancia de litio
+	\item Identificador de campo GES
+	\item Temperatura efectiva
+	\item Gravedad superficial
+	\item Probabilidad de pertenencia a un cúmulo (>= 0.95\%)
+\end{itemize}
+
+Este segundo filtrado reduce la población de estrellas a considerar a un total de 5 895 registros que se distribuyen entre los 64 OCs listados en la Tabla \ref{tab:oc_full_list}.\par
+
+Como datos de referencia tomamos un subconjunto de las observaciones obtenidas por GES.
+Este subconjunto consiste en aquellas estrellas que pertenecen a OCs, y entre ellas se encuentran potenciales gemelas "ocultas" de nuestro Sol.
+Para identificar estas últimas, procedemos a realizar una selección adicional, esta vez utilizando los valores dados por nuestros modelos durante las simulaciones para la metalicidad ($\feh$), la temperatura efectiva ($\teff$) y la gravedad superficial ($\gsurf$).
+Por último, nos quedan las estrellas para las que las observaciones de GES han medido valores congruentes con los resultados de nuestros modelos, a partir de los cuales obtenemos finalmente sus abundancias de litio.\par
+
+\subsection{Selección de gemelos solares} \label{sec:gemelos_solares}
+Una vez preseleccionadas las estrellas pertenecientes a los OCs, el interés se centra en identificar las más parecidas a nuestro Sol, es decir, las gemelas solares.
+Esto nos lleva a preguntarnos qué parámetros estelares debemos tener en cuenta para clasificar una estrella como gemela solar.
+Encontrarlos nos ayudaría a averiguar el origen de nuestro Sol y a obtener información sobre las condiciones que prevalecían en los OC en los que nacieron.\par
+
+Ya se han realizado varios intentos de encontrar hermanos y gemelos solares.
+Los primeros se refieren a estrellas que se formaron en el mismo cúmulo que el Sol y, por tanto, deberían tener edades y abundancias químicas muy similares a las de éste \citep[ver][y referencias en él]{Adibekyan2018}.
+Por otro lado, según \cite{Strobel1996}, se consideran gemelos solares a aquellas estrellas que poseen parámetros físicos fundamentales como $\teff$, $\gsurf$, $\feh$, propiedades fotométricas, composición química, edad, luminosidad, rotación y campos magnéticos similares, si no idénticos, a los del Sol.
+Localizar estos gemelos solares supone un reto formidable debido a los estrictos criterios y a su dispersión por toda la Vía Láctea.\par
+
+No es de extrañar que los métodos de búsqueda aplicados se orienten a encontrar las candidatas basándose en sus propiedades cinemáticas.
+Luego se comparan sus temperaturas, metalicidades y abundancias químicas con las de nuestro Sol.
+En nuestro estudio, ampliamos los criterios de selección añadiendo la gravedad superficial y la edad de la estrella.
+Estos dos nuevos parámetros representan un ajuste fino del proceso de búsqueda de candidatos a gemelos.
+Los catálogos Gaia y GES no proporcionan la edad estimada de los OCs, así que necesitamos recurrir a los resultados obtenidos en el trabajo de \cite{Bragaglia2022} (ver Figura \ref{fig:open_cluster_sample}).
+Utilizamos la edad de los OCs indicada en la tabla \ref{tab:oc_reduced_list} y los datos obtenidos a partir del conjunto de simulaciones MESA iniciadas con diferentes $\omegaini$, donde $\omegaini = \oomegac$, $\Omega$ es la velocidad angular de la estrella en la superficie estelar, y $\omegac$ es la velocidad superficial en el ecuador de una estrella en rotación donde la fuerza centrífuga equilibra la gravedad newtoniana.
+La edad derivada de estas simulaciones proporciona un parámetro clave para nuestro análisis.
+Esta información nos permite determinar los demás parámetros de simulación, que posteriormente utilizaremos para compararlos con los valores asociados a las estrellas pertenecientes a los OCs preseleccionados.\par
+
+\begin{table}
+	\centering
+	\begin{tabular}{ll} 
+		\hline
+		Parámetro estelar & Intervalo de selección\\
+		\hline
+		Edad OC & $\pm0.1\%$ \, para edad < 1 Ga \\
+		& $\pm0.15\%$ \, para edad $\geq$ 1 Ga \\
+		$\teff$ & $\pm 50.0 \, \Kelvin$ \\
+		$\gsurf$ & $\pm 0.05 \, \dex$\\
+		$\feh$ & $\pm 0.05 \, \dex$\\
+		\hline
+	\end{tabular}
+	\caption{Parámetros e intervalos de selección aplicados durante el proceso de triaje sobre los componentes del OC.}
+	\label{tab:sel_params}
+\end{table}
+
+De las aproximadamente 5 900 estrellas identificadas, seleccionamos aquellas cuyos valores de $\feh$, $\teff$ y $\gsurf$ se encuentran en los intervalos correspondientes definidos en la Tabla \ref{tab:sel_params} y según el siguiente proceso de filtrado: a partir de los valores dados por nuestras simulaciones para los parámetros listados en la Tabla \ref{tab:sel_params}, procedemos a seleccionar de los distintos OCs aquellas estrellas que tienen valores dentro de los intervalos establecidos.
+En concreto, para la edad estimada de un determinado cúmulo, establecemos un intervalo de $\pm$0.1 para OCs jóvenes (edad < 1 Ga) y $\pm$0.15 para contrapartes más viejas (edad >= 1 Ga) \citep{Cantat-Gaudin2020}[véase][para más detalle sobre los valores referidos].
+Este intervalo dependiente de la edad sirve como base para seleccionar los pasos de tiempo de simulación que se alinean con los límites de edad inferior y superior de cada grupo.
+Una vez encontrados estos dos pasos temporales, se reúnen los valores límite asociados para $\teff$, $\gsurf$ y $\feh$.
+Con estos rangos de valores procedemos entonces a seleccionar en cada uno de los OCs aquellas estrellas que tienen valores para estos mismos parámetros dentro de ellos.
+De esta forma obtenemos finalmente las estrellas candidatas a ser comparadas con los resultados de la simulación.\par
+
+El resultado del proceso de filtrado puede observarse en las siguientes tablas.
+En la Tabla \ref{tab:oc_reduced_list} se agrupan los OCs que han aportado alguno de sus miembros en las simulaciones que hemos realizado.
+Para un determinado OC se informa del número de estrellas que se han seleccionado en las diferentes simulaciones.
+Por otra parte, las tablas \ref{tab:oc_m67} y \ref{tab:oc_ngc2516} resumen el subconjunto de componentes de los OCs M67 y NGC2516 seleccionados durante las simulaciones para modelos inicializados con $\omegaini = 0.14$.
+Cada una de las tablas enumera el identificador GES de la estrella y los rangos de $\teff$, $\gsurf$, $\feh$, A(Li), el error estimado de la A(Li) y edad estimada utilizados en la simulación.
+Estos valores representan los rangos de valores en los que se ha basado el filtrado.
+
+\newpage
+\KOMAoptions{paper=landscape,pagesize}
+\recalctypearea
+
+\begin{longtable}[c]{|l l l l || c c c c c c c c c c|}
+	\hline
+	& & & & & & & & $\omegaini$ & & & & & \\
+	Cúmulo Abierto & Edad(Ga) & [Fe/H] & $N_*$ & 0.095 & 0.10 & 0.105 & 0.11 & 0.115 & 0.12 & 0.125 & 0.13 & 0.14 & 0.1425\\
+	\hline
+	Berkeley 21 & 2.138 & -0.21 & 744 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1 & 1 & 1\\
+	Berkeley 39 & 5.623 & -0.14 & 899 & 1 & 1 & 1 & 1 & 1 & 1 & 2 & 2 & 2 & 2\\
+	IC 2602 & 0.036 & -0.06 & 1836 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0\\
+	IC 4665 & 0.033 & 0.01 & 567 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 0 & 0 & 0\\
+	Messier 67 & 3.981 & -0.02 & 131 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6\\
+	NGC 2141 & 1.862 & -0.04 & 853 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+	NGC 2355 & 1 & -0.13 & 208 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+	NGC 2420 & 1.698 & -0.15 & 562 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+	NGC 2425 & 2.399 & -0.13 & 528 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1\\
+	NGC 2451 & 0.035 & -0.08 & 1656 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 0 & 2 & 1\\
+	NGC 2516 & 0.24 & -0.04 & 759 & 0 & 0 & 0 & 1 & 1 & 1 & 3 & 4 & 4 & 3\\
+	NGC 3532 & 0.398 & -0.01 & 1145 & 1 & 1 & 0 & 1 & 1 & 1 & 1 & 1 & 0 & 1\\
+	NGC 6005 & 1.259 & 0.22 & 560 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 & 1\\
+	NGC 6259 & 0.269 & 0.18 & 494 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1\\
+	NGC 6281 & 0.513 & -0.04 & 320 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
+	NGC 6405 & 0.035 & -0.02 & 701 & 1 & 1 & 1 & 0 & 1 & 3 & 2 & 0 & 0 & 0\\
+	NGC 6633 & 0.692 & -0.03 & 1662 & 2 & 2 & 2 & 1 & 1 & 0 & 0 & 0 & 1 & 1\\
+	NGC 6709 & 0.191 & -0.02 & 730 & 2 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
+	Trumpler 20 & 1.862 & 0.13 & 1213 & 3 & 3 & 3 & 2 & 2 & 2 & 2 & 2 & 2 & 2\\
+	\hline
+	\caption{Lista de los OCs seleccionados.
+Para cada OC se indica el nombre, la edad estimada, la metalicidad y el número de componentes.
+Además, se muestran los diferentes $\omegaini$ utilizados en las distintas simulaciones, donde $\omegaini = \oomegac$, $\Omega$ es la velocidad angular de la estrella en la superficie estelar, y $\omegac$ es la velocidad superficial en el ecuador de una estrella en rotación donde la fuerza centrífuga equilibra la gravedad newtoniana.
+Para cada entrada del OC se informa por $\omegaini$ de cuántos componentes del OC cuyos parámetros estelares, tras compararlos con los equivalentes calculados por las simulaciones, se seleccionan.
+Los componentes OC seleccionados se utilizan como referencia en las figuras que se muestran en las secciones siguientes.}
+	\label{tab:oc_reduced_list}	
+\end{longtable}
+\newpage
+\KOMAoptions{paper=portrait,pagesize}
+\recalctypearea
+
+\begin{table}
+	\centering
+	\begin{tabular}{l l l l l l l} 
+		\hline
+		Id Objeto & $\teff$(K) & $\gsurf$(dex) & FeH(dex) & ALi(dex) & eALi(dex) & Edad(Ga)\\
+		\hline
+		8510914-1157003 & 5843 & 4.42 & -0.03 & 1.8 & 0.08 & 3.981\\ 
+		8510991-1146169 & 5797 & 4.4 & -0.01 & 1.65 & 0.08 & 3.981\\ 
+		8512177-1144050 & 5818 & 4.46 & 0 & 1.99 & 0.06 & 3.981\\ 
+		8513941-1200571 & 5792 & 4.41 & -0.02 & 1.49 & 0.13 & 3.981\\ 
+		8515559-1148381 & 5794 & 4.43 & -0.04 & 1.56 & 0.1 & 3.981\\ 
+		8520350-1147480 & 5875 & 4.47 & -0.03 & 1.21 & 0.24 & 3.981\\ 
+		\hline
+	\end{tabular}
+	\caption{M67 - Configuración de filtro: Identificador estelar GES, $\teff$(K) [5787.4892, 5887.6823], $\gsurf$(dex) [4.3979, 4.4977], $\feh$(dex) [-0.05, 0.05], Edad(Ga) [3.977, 3.985]}
+	\label{tab:oc_m67}
+\end{table}
+
+\begin{table}
+	\centering
+	\begin{tabular}{l l l l l l l} 
+		\hline
+		Id Objeto & $\teff$(K) & $\gsurf$(dex) & FeH(dex) & ALi(dex) & eALi(dex) & Edad(Ga)\\
+		\hline
+		7571111-6048156 & 5084 & 4.54 & 0 & 2.29 & 0.08 & 0.24\\ 
+		7595031-6044149 & 5108 & 4.52 & -0.05 & 1.92 & 0.08 & 0.24\\ 
+		7595280-6032498 & 5093 & 4.49 & -0.03 & 2.29 & 0.08 & 0.24\\ 
+		\hline
+	\end{tabular}
+	\caption{NGC2516 - Configuración de filtro: Identificador estelar GES, $\teff$(K) [5010.0324, 5111.0095], $\gsurf$(dex) [4.4405, 4.5405], $\feh$(dex) [-0.05, 0.05], Edad(Ga) [0.23964, 0.24036]}
+	\label{tab:oc_ngc2516}
+\end{table}
 
@@
 	\caption{Tabla obtenida de \cite{Bragaglia2022}.
-Las columnas 1 a 7 listan la denominación del cúmulo, las coordenadas de ascensión recta y declinación, la edad estimada del cúmulo, la distancia en parsecs (módulo de distancia convertido), la posición Z en coordenadas cartesianas galácticas, y la distancia desde el centro galáctico suponiendo que el Sol está a 8340pc.
+Las columnas 1 a 7 listan la denominación del cúmulo, las coordenadas de ascensión recta y declinación, la edad estimada del cúmulo, la distancia (módulo de distancia convertido) en parsecs (pc) , la posición Z en coordenadas cartesianas galácticas, y la distancia desde el centro galáctico suponiendo que el Sol está a 8340 pc.
 Las columnas 8 y 9 enumeran el número de espectros y objetivos, mientras que la media de [Fe/H] y la desviación estándar se dan en la columna 10.}
@@
 \end{figure}
-
-La abundancia de Litio puede proporcionar información valiosa sobre la edad y la historia de las estrellas.
-En particular, puede ayudar a rastrear la mezcla convectiva en el interior de las estrellas, que está influenciada por la rotación estelar.
-La rotación, a su vez, puede ser modulada por la presencia de campos magnéticos a través del frenado magnético.
-
-Además, en estrellas jóvenes y de baja masa, la presencia de fuertes campos magnéticos puede inhibir la convección, alterando el transporte de energía desde el interior de la estrella.
-Esto puede afectar la tasa de quema de Litio, lo que resulta en una mayor abundancia de Litio en la superficie de la estrella de lo que se esperaría de lo contrario.
-
-Por lo tanto, al estudiar las abundancias de Litio en las estrellas, los astrónomos pueden obtener pistas sobre la intensidad de los campos magnéticos en el interior de las estrellas.
-Sin embargo, la interpretación de las abundancias de
-
-Los campos magnéticos en las estrellas pueden tener un impacto significativo en la abundancia de Litio presente en ellas.
-El Litio se quema fácilmente a una temperatura relativamente baja (2.5 x $10^6$ K) en el interior estelar³.
-Por lo tanto, es un diagnóstico extraordinariamente sensible de la estructura y evolución estelar³.
-
-La abundancia de Litio puede proporcionar información valiosa sobre la edad y la historia de las estrellas.
-En particular, puede ayudar a rastrear la mezcla convectiva en el interior de las estrellas, que está influenciada por la rotación estelar.
-La rotación, a su vez, puede ser modulada por la presencia de campos magnéticos a través del frenado magnético.
-
-Además, en estrellas jóvenes y de baja masa, la presencia de fuertes campos magnéticos puede inhibir la convección, alterando el transporte de energía desde el interior de la estrella.
-Esto puede afectar la tasa de quema de Litio, lo que resulta en una mayor abundancia de Litio en la superficie de la estrella de lo que se esperaría de lo contrario¹.
-
-Por lo tanto, al estudiar las abundancias de Litio en las estrellas, los astrónomos pueden obtener pistas sobre la intensidad de los campos magnéticos en el interior de las estrellas.
-Sin embargo, la interpretación de las abundancias de Litio puede ser compleja, ya que también pueden ser influenciadas por otros factores, como la metalicidad de la estrella y la eficiencia de la convección¹.
-
-En resumen, aunque el Litio no proporciona una medida directa de los campos magnéticos estelares, su abundancia en las estrellas puede ofrecer valiosas pistas sobre la actividad magnética en el interior estelar.
-Esto hace que las mediciones de las abundancias de Litio, como las proporcionadas por la Encuesta Espectroscópica Gaia-ESO (GES), sean una herramienta importante para la investigación de los campos magnéticos en las estrellas.
-
-Source: Conversation with Bing, 1/25/2024
-(1) [2006.01899] Study of the effects of magnetic braking on the lithium .... https://arxiv.org/abs/2006.01899.
-(2) Study of the effects of magnetic braking on the lithium abundances of .... https://academic.oup.com/mnras/article/496/2/1343/5851759.
-(3) Enhanced lithium depletion in Sun-like stars with orbiting planets - Nature. https://www.nature.com/articles/nature08483.pdf.
-(4) undefined. https://doi.org/10.1093/mnras/staa1569.
-
-Source: Conversation with Bing, 1/25/2024
-(1) Effect of magnetic fields on Lithium depletion of solar-type stars.
-https://www.cambridge.org/core/services/aop-cambridge-core/content/view/054CD22CB5342846A1EC5552F7DED8FE/S1743921313002512a.pdf/effect-of-magnetic-fields-on-lithium-depletion-of-solar-type-stars.pdf.
-
-En resumen, cada uno de estos enfoques ofrece una perspectiva única y valiosa sobre la naturaleza y la evolución de estrellas.
-La combinación de los catálogos Gaia y GES proporciona una visión completa y detallada de las estrellas similares al Sol.
-La información sobre las abundancias de Litio, la velocidad angular y los datos de los cúmulos abiertos, en particular, puede arrojar luz sobre la cinemática, la composición química, la edad y la historia de estas estrellas, así como ser una herramienta muy potente para ofrecer una perspectiva única y valiosa sobre la naturaleza y la evolución de los campos magnéticos estelares.
-En resumen, una fuente de datos invaluable para cualquier astrofísico que estudie estrellas similares al Sol.\par
-
-Source: Conversation with Bing, 1/25/2024
-The Gaia-ESO Survey: Calibrating the lithium–age relation with open .... https://arxiv.org/pdf/2009.00610.pdf.
-Interferometric Observations of Magnetic Fields in Forming Stars. https://www.frontiersin.org/articles/10.3389/fspas.2019.00003/full.
-ESO Survey: 3D dynamics of young groups and clusters from GES and Gaia .... https://arxiv.org/pdf/2311.08358.pdf.
-Stellar magnetic fields and surface structures - Uppsala University. https://www.physics.uu.se/research/astronomy-and-space-physics/research/stars/magnetism/.
-The Role of Magnetic Fields in Protostellar Outflows and Star Formation. https://www.frontiersin.org/articles/10.3389/fspas.2019.00054/full.
-The Gaia-ESO Survey: Membership probabilities for stars in 32 open .... https://arxiv.org/pdf/2006.09423v1.pdf.
-undefined. https://doi.org/10.3389/fspas.2019.00054.
-Source: Conversation with Bing, 1/25/2024
-[1310.5562] Magnetic fields of Sun-like stars - arXiv.org. https://arxiv.org/abs/1310.5562.
-undefined. https://doi.org/10.1017/S1743921314002026.
-Membership of stars in open clusters using random forest with gaia data .... https://link.springer.com/article/10.1140/epjs/s11734-021-00205-x.
-Evidence of New Magnetic Transitions in Sun-like Stars from Gaia Data. https://www.spacedaily.com/reports/Evidence of New Magnetic Transitions in Sun like Stars from Gaia Data 999.html.
-undefined. https://doi.org/10.48550/arXiv.1310.5562.
 

### 2024-03-17 project/overleaf/Tesis III/capitulos/capitulo03.tex
WIP

@@
 \end{ceqn}
-donde $\nu_\infty$ es la velocidad terminal del viento definida como la velocidad que alcanza el viento, la materia que fluye a gran distancia de la estrella central, donde ya no es acelerada por la fuerza impulsora del viento pero su desaceleración debida a la interacción con el medio interestelar (ISM) es despreciable \cite{Niedzielski2002}.\par
+donde $\nu_\infty$ es la velocidad terminal del viento definida como la velocidad que alcanza el viento, la materia que fluye a gran distancia de la estrella central, donde ya no es acelerada por la fuerza impulsora del viento pero su desaceleración debida a la interacción con el medio interestelar (Inter-Stellar Medium, ISM) es despreciable \cite{Niedzielski2002}.\par
 

### 2024-03-17 project/overleaf/Tesis III/capitulos/capitulo04.tex
WIP

@@
 
-      ! redo, retry, or backup must be done 
-      ! inside the step_loop
-      if (result == redo) then
-        result = star_prepare_to_redo(...)
-      end if
-      if (result == retry) then
-        result = star_prepare_to_retry(...)
-      end if
-      if (result == backup) then
-        result = star_do1_backup(...)
-        just_did_backup = .true.
-      else
-        just_did_backup = .false.
-      end if
       if (result == terminate) then
@@
     end do step_loop
-
-    ! once we get here, the only options
-    ! are keep_going or terminate.
 
@@
    call after_evolve_loop(...)
-
 end subroutine run1_star
@@
 
-\begin{lstlisting}[language=Fortran, float, caption={Parametrización de la masa inicial de la estrella}, label={lst:ini_params}]
+\begin{lstlisting}[language=Fortran, float, caption={Parametrización de la metalicidad y masa inicial de la estrella}, label={lst:ini_params}]
   &star_job
     ! Solar composition, 
-  ! check the following file:
-  ! $MESA_DIR/star/test_suite/
-  ! example_solar_model/inlist_solar_model
     set_uniform_initial_composition = .true.
-  initial_h1 = 7.0001067228923031D-01
-  initial_h2 = 0
-  initial_he3 = 2.7955281905284692D-05
-  initial_he4 = 2.7952486377094160D-01
+    initial_h1 = 0.715399992
+    initial_h2 = 1.430799984e-05
+    initial_he3 = 4.48623528494e-05
+    initial_he4 = 0.270255137647
 
     ! set metal fractions z fractions
-  initial_zfracs = 3 !GS98
+    initial_zfracs = 6 ! AGSS09_zfracs
 
@@
     ! or 1.0d0 in Msun units
-  initial_mass = 0.8d0 
+    initial_mass = 1.0d0 
 \end{lstlisting}
@@
 
---CONTINUAR POR AQUI--
-
 \subsection{Simulaciones extendiendo el marco teórico estándar de MESA}
-Una vez que hemos llegado a un punto en el que tenemos una configuración de proyecto que nos permite obtener las abundancias de Li de forma detallada, y además hemos conseguido activar los procesos de difusión y rotación superficial en nuestro modelo, damos paso a una segunda línea de trabajo consistente en extender el funcionamiento de MESA para incorporar nuestras propias rutinas de extensión al simulador.
+Una vez que hemos llegado a un punto en el que tenemos una configuración de proyecto que nos permite obtener las abundancias de litio de forma detallada, y además hemos conseguido activar los procesos de difusión y rotación superficial en nuestro modelo, damos paso a una segunda línea de trabajo consistente en extender el funcionamiento de MESA para incorporar nuestras propias rutinas de al simulador.
 Las consideraciones previas a tener en cuenta son las siguientes:
 
---INCLUIR AQUÍ ENUMERACIÓN DE PASOS A GRANDES RASGOS--
 \begin{itemize}
-    \item Entender el ciclo de control aplicado a los pasos de simulación
-    \item Extender la rutina de inicialización para incorporar nuestros parámetros
-    \item Extender 
+	\item Definir el formalismo matemático que soporta el proceso a incorporar a MESA
+	\item Identificar los parámetros fijos y variables del formalismo
+	\item Mapear los parámetros variables a sus equivalentes en MESA
+	\item Identificar el \textit{hook} de MESA adecuado para incluir la rutina
+	\item Implementar la rutina y asegurar su compatibilidad con MESA
+	\item Identificar los parámetros de configuración
+	\item Mapear los parámetros de configuración en el fichero \textit{inlist}
+	\item Realizar las simulaciones
 \end{itemize}
 
-El planteamiento teórico no es especialmente difícil, lo complicado del mismo es la parte que afecta a MESA.
-En concreto, localizar las secciones concretas de código en las que se calcula el valor de la gravedad local que afecta a cada elemento de masa que tiene en cuenta la simulación.
-A esto hay que sumarle el hecho de que MESA está construido de una forma modular, así que es de esperar que no tengamos una única sección de código en el programa donde se calcule este valor de gravedad, sino que cada módulo que, de algún u otro modo, contribuya a la difusión química de los elementos tiene que ser revisado.\par
+Los puntos 1 a 4 y 7 del listado anterior serán extensivamente documentados en las partes \ref{parte1} y \ref{parte2}.
+En lo que atañe a los puntos 5 y 6, dado que es algo común a ambas partes, lo comentaremos en este apartado.\par
 
-Con diferencia, el localizar dónde invocar a nuestra rutina de código es la parte más complicada de todo el presente trabajo.
-Tenemos que detectar los módulos implicados en la difusión de los elementos, con el riesgo que implica que nos dejemos alguno por identificar, analizar su código fuente en detalle para llegar a entender su funcionamiento y realizar la invocación en el punto exacto a nuestra rutina, para alterar así el valor de gravedad calculado por MESA para un determinado diferencial de masa.
-Adicionalmente, se hace totalmente necesario el revisar la bibliografía en la que se basa la implementación de las rutinas presentes en cada módulo para dotar de contexto y significado a las líneas de código.
-De otro modo, es casi imposible extraer el trasfondo científico que se encuentra en ellas codificado.\par
+Los parámetros de configuración son valores o ajustes que se utilizan para personalizar el comportamiento del simulador MESA, y por extensión, de nuestras rutinas.
+MESA ofrece una gran cantidad de de estos parámetros.
+La manera de especificarlos es mediante su inclusión en el fichero \textit{inlist}.
+Este mecanismo se hace también extensivo a las rutinas que se añaden al simulador.
+En el Listado \ref{lst:config_inlist} mostramos parte del contenido del fichero \textit{inlist}.
+Lo primero es indicar al simulador que debe utilizar nuestra rutina de par de torsión en lugar de la que trae por defecto MESA.
+Es a través de este \textit{hook} la forma de implementar nuestras rutinas de frenado magnético.
+A continuación listamos los diferentes parámetros de configuración para las diferentes rutinas.
+Estos se indican a través de las variables \textit{x\_ctrl}, \textit{x\_integer\_ctrl} y \textit{x\_logical\_ctrl} según estos parámetros sean valores reales, enteros o lógicos respectivamente.\par
 
---REVISAR PARA ELIMINAR REFERENCIAS A GRAVEDAD. HACERLO GENÉRICO--
+\begin{lstlisting}[language=Fortran, float, caption={Parámetros de configuración para las rutinas incorporadas al simulador MESA.}, label={lst:config_inlist}]
+  ...
+  !HOOKs
+  ! make use of 'hooks'
+  use_other_torque = .true.
 
-Inicialmente teníamos dos módulos candidatos por los que empezar a estudiar su implementación, el de difusión y el de mlt.
-Por tener éste último un menor número de líneas de código (aproximadamente unas 1800) y encontrarse la implementación restringida a un solo módulo, empezamos por él.
-Tras una revisión a fondo del mismo se llegó a generar una versión modificada del mismo que calculaba un nuevo valor de gravedad local relacionado con el elemento de masa diferencial.
-Este nuevo valor tenía en cuenta tanto la masa del planeta, como su distancia a la estrella.\par
+  ! **********************************
+  ! Magnetic braking input parameters
+  ! **********************************
+  ! B - magnetic field torque (G).
+  ! If set to -1, a variable magnetic field strengh is calculated
+  !x_ctrl(1) = 4.0d0  
+  x_ctrl(1) = -1
 
-Seguidamente pasamos a analizar el código fuente del módulo de difusión que, como mencionábamos, se basa en la implementación de las ecuaciones propuestas en el trabajo de Thoul, et al.
-(1994).
-Ya desde los primeros momentos nos dimos cuenta de que el esfuerzo de localizar dónde deberíamos de invocar a nuestra rutina se iba a complicar.
-La implementación de este módulo es bastante más laboriosa, larga (más de 3500 líneas) y se encuentra repartida entre varios módulos que, a diferencia de lo que ocurría con el de mlt, algunos de ellos no ofrecían la posibilidad de ser extendidos de una manera cómoda.
-Además de esto, la implementación que se sigue en MESA acaba generando una matriz de coeficientes de difusión en los que el efecto de la gravedad está representado de manera implícita, es decir, no se calcula un valor de gravedad local asociada a un elemento diferencial de masa en un paso previo, de forma análoga a como ocurre en el módulo de mlt, para finalmente acabar derivando el valor del coeficiente de difusión.
-En este caso se resuelve el sistema de ecuaciones y la contribución de la gravedad sobre el elemento de masa queda recogida en las soluciones al mismo.
-A la propia dificultad de entender la implementación del código, se sumaba el hecho de tener que deshacer el cálculo de los coeficientes para poder incluir el efecto de nuestra rutina, y esto era algo que se antojaba harto complicado al no disponerse del conocimiento necesario.
-Estábamos en un punto muerto.
+  ! eps thhreshold use for detecting nuclear reactions in the core (erg/g/s)
+  x_ctrl(2) = 0.1d-1
 
---REVISAR PARA ELIMINAR REFERENCIAS A GRAVEDAD--
+  ! How to redistribute the loss of angular momentum
+  ! 0 = Only among the zones of outmost convective zone
+  ! 1 = till the outmost zone, r(1)
+  x_integer_ctrl(1) = 0
+
+  ! Controls if the mb routine is executed only when a radiative core is develop
+  ! .false. = don't wait, 
+  ! .true. = wait
+  x_logical_ctrl(5) = .true.
+  !keep_on_rad_core
+  x_logical_ctrl(6) = .true.
+
+  ! **********************************
+  ! Variable MLT alpha
+  ! **********************************
+  ! (de)activate variable MLT alpha routine
+  x_logical_ctrl(10) = .true.
+
+  ! **********************************
+  ! Disk locking input parameters
+  ! **********************************
+  ! disk lifetime in years. Set to -1=deactivated
+  !x_ctrl(3) = 3.0d6 
+  x_ctrl(3) = -1
+  ! disk omega in rad/s
+  x_ctrl(4) = 9.090256e-6
+
+  ! **********************************
+  ! Debug flags
+  ! **********************************
+  ! debug flag for use_other_torque routine
+  x_logical_ctrl(1) = .false. 
+  ! debug flag for reset_other_torque, if true, jdot is discarded
+  x_logical_ctrl(2) = .false.
+  ! debug flag for get_convective_info
+  x_logical_ctrl(3) = .false. 
+  ! debug flag for get_core_info
+  x_logical_ctrl(4) = .false.  
+  ! debug flag for new_alpha
+  x_logical_ctrl(7) = .false.  
+  ! debug flag for calculate_mag_field_intensity
+  x_logical_ctrl(8) = .false.
+  ! debug flag for calculate_jdot_rate  
+  x_logical_ctrl(9) = .false.  
+  ...
+
+\end{lstlisting}
 
 \subsection{Rutina de inicialización para las simulaciones}
-EXPLICAR EL CONJUNTO DE PARÁMETROS QUE SE LEEN DEL INLIST, Y SIGNIFICADO. TAMBIÉN LOS FLAGS DE DEPURACIÓN
-\begin{lstlisting}[language=Fortran, caption={Rutina de inicialización de las simulacions.}, label={lst:extras_ctrl}]
+Una vez el fichero \textit{inlist} contiene los parámetros de configuración para nuestra(s) rutina(s) hay que proceder a leerlos para ponerlos a disposición del código de las mismas.
+Como se indica en el listado de control \ref{lst:bucle_ctrl}, antes de proceder a ejecutar los pasos de la simulación, MESA procede a invocar la rutina \textit{extras\_ctrl}.
+En el Listado \ref{lst:extras_ctrl} se muestra la forma de proceder para obtener los valores indicados en el fichero de configuración.
+Estos valores se asignan a variables que quedan a disposición al código de las diferentes rutinas.
+
+\begin{lstlisting}[language=Fortran, float, caption={Rutina de inicialización de las simulacions.}, label={lst:extras_ctrl}]
       subroutine extras_controls(id, ierr)
@@
 \subsection{Rutina de par de torsión}
-EXPLICAR QUE SUSTITUIMOS LA RUTINA POR DEFECTO CON LA NUESTRA
-\begin{lstlisting}[language=Fortran, caption={Rutina de par de torsión.}, label={lst:torque_hook}]
+El par de torsión, también conocido como torque o momento de torsión, es una magnitud física que describe la capacidad de una fuerza para provocar un giro o rotación en un objeto.
+El par de torsión y el momento angular están estrechamente relacionados en el contexto del movimiento de rotación.
+El momento angular es una cantidad física que describe la tendencia de un objeto giratorio a seguir girando.
+Cuando se aplica un par de torsión a un objeto giratorio, si se hace en la misma dirección en la que el objeto ya gira, aumenta su velocidad angular.
+Por el contrario, si se aplica en la dirección opuesta al giro del objeto, disminuye su velocidad angular.
+En nuestro caso, el frenado magnético producirá una disminución, una toque en sentido contrario al giro de la estrella, provocando en ésta una pérdida de momento angular.\par
+
+En el Listado \ref{lst:torque_hook} se muestra como reemplazar la rutina por defecto de MESA para el cálculo del par de torsión para la nuestra, en la que implementaremos el frenado magnético.
+
+\begin{lstlisting}[language=Fortran, float, caption={Rutina de par de torsión.}, label={lst:torque_hook}]
       subroutine other_torque_hook(id, ierr)
@@
 \subsection{Rutina de frenado magnético}
-EXPLICAR QUE LOS DOS TIPOS DE RUTINAS: INTESIDAD FIJA, INTENSIDAD VARIABLE
-\begin{lstlisting}[language=Fortran, caption={Rutina de frenado magnético.}, label={lst:torque_mb_hook}]
-      ! This routine implements a magnetic braking effect.
-      ! It distributes among the zones which conforms the convective shell the
-      ! loss of angular momentum
+Si entrar en detalles por el momento (los mismos se abordarán en los siguientes capítulos), la pérdida de momento angular causada por el frenado magnético depende directamente de la velocidad de rotación de la estrella, de la cantidad de masa perdida por la misma debido a los vientos estelares y de la intensidad del campo magnético.
+Está pérdida, una vez computada, es distribuida entre las diferentes capas que componen la zona convectiva superficial de la estrella.\par
+
+Por otro lado, nuestro Sol es una estrella activa en lo que se refiere al campo magnético.
+Según los modelos teóricos éste es generado por una dinamo en la tacoclina.
+Ésta se define como la región de transición que se encuentra entre la zona interior radiactiva del Sol y la zona de convección que la rodea.
+En ella se produce un cambio significativo en el comportamiento de la rotación solar.
+En la zona interior radiactiva, la rotación es similar a la de un sólido rígido.
+Esto podría deberse a la influencia de un campo magnético fósil.
+En la zona exterior convectiva, la rotación es diferencial, con los polos rotando más lentamente que las regiones ecuatoriales.
+Es en esta región que presenta un cambio brusco de las velocidades de rotación donde, según los estudios, se genera el campo magnético del Sol.
+Éste alcanza la superficie con la ayuda de procesos convectivos que tienen lugar en la zona convectiva y da lugar a regiones magnéticamente activas en la superficie solar.\par
+
+Como hemos acabo de enumerar, existen una serie de condiciones para que se produzca el efecto de frenado magnético:
+
+\begin{enumerate}
+	\item La estrella rota
+	\item La estrella pierde masa debido a los vientos estelares
+	\item La estrella tiene un nucleo radiativo
+	\item La estrella tiene una zona convectiva que alcanza su superficie
+\end{enumerate}
+
+El Listado \ref{lst:torque_mb_hook} muestra como nuestra rutina de par de momento tiene en cuenta las diferentes condiciones listadas para influenciar o no en el momento angular de la estrella.
+En el caso de que lo haga, provocará una pérdida de momento en función las variables mencionadas, pérdida que será distribuida entre las diferentes capas de la capa convectiva exterior.
+
+\begin{lstlisting}[language=Fortran, float, caption={Rutina de frenado magnético.}, label={lst:torque_mb_hook}]
   subroutine other_torque_mag_brk(id, ierr) 
-         use const_def
-         integer, intent(in) :: id
-         integer, intent(out) :: ierr
-         type (star_info), pointer :: s
-         integer :: k
-         real(dp) :: B, j_dot, r_st
-         type (star_zone_info), target :: sz_info, core_info
-         type (star_zone_info), pointer :: sz_info_ptr, core_info_ptr
-         real(dp), dimension(:), pointer :: mag_brk_jdot
-         integer :: jdot_routine !how to distribute the j_dot
      !controls if jdot distribution must only affect the convective zone
@@
      logical :: wait_rad_core
-         integer :: activated !signals when the jdot routine is activated
+     !signals when the jdot routine is activated          
+     integer :: activated 
+     ...
 
-         !Pointer to structure which conveys information about the convectice
-         !and core zones
-         sz_info_ptr => sz_info
-         core_info_ptr => core_info
-
-         ierr = 0
-         call star_ptr(id, s, ierr)
-         if (ierr /= 0) return
-
-         !Reset support structures
-         allocate(mag_brk_jdot(s% nz))
-         s% extra_jdot(:) = 0
-         s% extra_omegadot(:) = 0
      activated = 0
-         call reset_x_ctrl(s, idx_low_x_ctrl, idx_high_x_ctrl)
-         call reset_core_info(core_info_ptr)
-         call reset_convective_info(sz_info_ptr)
+     ...
 
@@
      only_cz = .true.
-         if (jdot_routine /= 0) then
-               only_cz = .false.
-         end if
+     ...
 
@@
      !Calculate amount of loss of angular moment
-         !Magentic field intensity
+     !Magnetic field intensity
      B = s% x_ctrl(1)
@@
          (.not. wait_rad_core .or. (wait_rad_core .and. is_core_rad(s)))) then
-
        activated = 1
-
        !Distribute the loss of angular momentum
        call distribute_j_dot(s, j_dot, sz_info_ptr, mag_brk_jdot)
-
        !It happens that s% extra_jdot is longer than s% nz but mag_brk_jdot
-            !is just defined
-            !for s% nz elements
+       !is just defined for s% nz elements
        s% extra_jdot(1:s% nz) = mag_brk_jdot

### 2024-03-20 project/overleaf/Tesis III/capitulos/capitulo03.tex
WIP

--- 2024-02-01 project/overleaf/Tesis III/capitulos/capitulo03.tex
@@
 
-El impacto de la rotación tanto en la PMS como en el agotamiento del Li para las estrellas de tipo solar ha sido ampliamente debatido en el pasado \cite{Pinsonneault1997,Jeffries2004,Somers2014} y revisado más recientemente sobre la base de la disponibilidad de medidas más precisas \cite{Bouvier2016}.
+El impacto de la rotación tanto en la PMS como en el agotamiento del litio para las estrellas de tipo solar ha sido ampliamente debatido en el pasado \citep{Pinsonneault1997,Jeffries2004,Somers2014} y revisado más recientemente sobre la base de la disponibilidad de medidas más precisas \citep{Bouvier2016}.
 Sin embargo, estos estudios previos se centraron principalmente en los efectos hidrostáticos y no consideraron la influencia de un acoplamiento entre el campo magnético estelar y su posible efecto de desacelaración.
-Se piensa que la AML tiene una influencia directa en los procesos de mezcla y que el transporte del momento podría producirse por una serie de mecanismos tales como: pérdida de masa, campos magnéticos y ondas gravitatorias (modos g).
-Respecto a estos dos últimos, las ondas gravitatorias \cite{Charbonnel2005, Pincon2016} y los campos magnéticos \cite{Eggenberger2009} tienen la propiedad de transmitir el momento angular (AM) de forma mucho más eficaz que inducir la mezcla \cite{Denissenkov2007}.
+Se considera que la pérdida de momento angular (Angular Moment Loss, AML) tiene una influencia directa en los procesos de mezcla que ocurren en el interior estelar y que el transporte de ese momento podría producirse por una serie de mecanismos tales como: pérdida de masa, campos magnéticos y ondas gravitatorias (modos g).
+Respecto a estos dos últimos mecanismos, las ondas gravitatorias \citep{Charbonnel2005, Pincon2016} y los campos magnéticos \citep{Eggenberger2009} tienen la propiedad de transmitir el momento angular (Angular Moment, AM) de forma mucho más eficaz que inducir el de mezcla \citep{Denissenkov2007}.
 Como consecuencia de este incremento en la eficiencia del transporte del AM, se reduce la cantidad de rotación diferencial entre las zonas radiativa y convectiva de la estrella (se fomenta una rotación de cuerpo sólido), así como las inestabilidades rotacionales inducidas.
-Además, también se originarían campos magnéticos en los límites entre estas zonas, en las llamadas tacoclinas \cite{Aschwanden2014, Guerrero2016}, que interaccionarían con las partículas del viento solar que se verían obligadas a rotar en sincronía con el campo magnético imperante y por tanto, frenarían la rotación estrella.
+Además, también se originarían campos magnéticos en los límites entre estas zonas, en las llamadas tacoclinas \citep{Aschwanden2014, Guerrero2016}, que interaccionarían con las partículas del viento solar que se verían obligadas a rotar en sincronía con el campo magnético imperante y por tanto, frenarían la rotación estrella.
 Esto es lo que se conoce como el efecto de frenado magnético (Magnetic Braking, MB).
@@
     \centering
-    \includegraphics[width=0.5\textwidth]{img/tesis/gravitational_waves.jpg}
+    \includegraphics[width=0.4\textwidth]{img/tesis/gravitational_waves.jpg}
     \caption{Vista esquemática de la estrella.
@@
 
-A día de hoy, los modelos estándar no son capaces de replicar de forma satisfactoria los valores observados de abundancia de Li en las superficies estelares con las predicciones que estos modelos arrojan.
-Esta falta de concordancia nos hace pensar que ciertos mecanismos físicos que influyen en la destrucción del Li están siendo modelizados de forma inadecuada o simplemente no se tienen en cuenta, como por ejemplo el frenado magnético.
+A día de hoy, los modelos estándar no son capaces de reconciliar de forma satisfactoria los valores observados de abundancia de litio en las superficies estelares con las predicciones que estos modelos arrojan.
+Esta falta de concordancia nos hace pensar que ciertos mecanismos físicos que influyen en la destrucción del litio están siendo modelizados de forma inadecuada o simplemente no se están teniendo en cuenta, como por ejemplo el frenado magnético.
 A la vista de estos hallazgos, parece evidente que aún quedan cuestiones abiertas por resolver sobre el tema de los procesos que participan, directa o indirectamente, en los mecanismos de mezcla que no están incorporados en los modelos estelares estándar.
@@
 
-La abundancia de Li observada en la fotosfera de las estrellas es un indicador de su composición interior y de los procesos de mezclado que en su interior tienen lugar.
-Adicionalmente, estas abundancias (además de otras métricas) se utilizan para comprobar la validez de los modelos estelares.
-Para que esto sea posible hay que tomar como premisa inicial que la abundancia de Li generada en la nucleosíntesis asociada al Big Bang es conocida y este elemento solo se destruye a través de reacciones nucleares.
-A pesar de décadas de esfuerzos teóricos, se sigue sin encontrar una explicación coherente con los modelos para las discrepancias encontradas en las comparaciones sobre la abundancia de Li para estrellas pertenecientes a cúmulos de diferentes edades y que se encuentran en el mismo estado evolutivo, o bien en la PMS, o bien en MS.
-Los modelos no son capaces de explicar las abundancias detectadas en las etapas tardías de la MS \cite{Tschape2001}.\par
-
-Es sabido que parte de la pérdida de Li se produce durante la PMS y que además ésta se acentúa según decrece la masa de la estrella y, a igualdad de masa, según aumenta la metalicidad de la misma.
-Para reproducir la dependencia entre la edad y masa de las estrellas con la merma de las concentraciones de Li, estos modelos necesitan de una combinación de procesos de mezclado cada vez más complejos, como overshooting, mezclado debido a procesos rotacionales o difusión microscópica.
-Para las estrellas de tipo solar, nos encontramos con dos factores que tienen que ser explicados desde el punto de vista de la abundancia de Li:
+Es conocido que parte de la pérdida de litio se produce durante la PMS y que además ésta se acentúa según decrece la masa de la estrella y, a igualdad de masa, según aumenta la metalicidad de la misma.
+Para reproducir la dependencia entre la edad y masa de las estrellas con la merma de las concentraciones de litio, estos modelos necesitan de una combinación de procesos de mezclado cada vez más complejos, como overshooting, mezclado debido a procesos rotacionales o difusión microscópica.
+Para las estrellas de tipo solar, nos encontramos con dos factores que tienen que ser explicados desde el punto de vista de la abundancia de litio:
 
 \begin{enumerate}
-    \item La deficiencia generalizada de Li en el Sol y en estrellas de tipo tardío presentes en cúmulos a diferentes edades.
-    \item La gran dispersión en abundancia de Li que existe en estrellas con una determinada temperatura efectiva detectada en los cúmulos abiertos.
+    \item La deficiencia generalizada de litio en el Sol y en estrellas de tipo tardío presentes en cúmulos a diferentes edades.
+    \item La gran dispersión en abundancia de litio que existe en estrellas con una determinada temperatura efectiva detectada en los cúmulos abiertos.
 \end{enumerate}
 
-Existen datos observacionales que sugieren que las estrellas que giran más rápido preservan el Li mejor que las que lo hacen más lentamente.
-Este hecho se puede deber a que por encima de cierto umbral de velocidad de rotación, ésta impide de manera drástica la penetración vertical de penachos generados por movimientos convectivos que, cuando están presentes, contribuyen a hacer más eficiente el proceso de destrucción de Li \cite{Baraffe2017}.\par
+Existen datos observacionales que sugieren que las estrellas que giran más rápido preservan el litio mejor que las que lo hacen más lentamente.
+Este hecho se puede deber a que por encima de cierto umbral de velocidad de rotación, ésta impide de manera drástica la penetración vertical de penachos generados por movimientos convectivos (ver Figura \ref{fig:grav_waves} BUSCAR SI TENEMOS UNA GRÁFICA MÁS APROPIADA EN BARAFE2017)que, cuando están presentes, contribuyen a hacer más eficiente el proceso de destrucción de litio \citep{Baraffe2017}.\par
 
-Adicionalmente y según el modelo estándar de nucleosíntesis para el Big Bang, la abundancia de Li original puede estimarse en 2.6 dex \cite{Spergel2003} pero la abundancia de litio detectadas en meteoritos se conoce que está en el orden A(Li) = [3.26-3.34] dex \cite{Randich2006, Grevesse2007}, lo que apunta hacia un enriquecimiento desde que se produjo el Big Bang.
-Por otro lado, la abundancia medida en las estrellas de tipo solar es del orden de ALi = 1.05 dex \cite{Grevesse2007}, lo que implica la existencia de procesos de destrucción del Li en su interior.
-Esta discrepancia entre las abundancias medidas en meteoritos y en estrellas de tipo solar es lo que ha pasado a denominarse el “problema del litio”.\par
+Adicionalmente y según el modelo SBBN, la abundancia de litio original estimada es de 2.6 dex \citep{Spergel2003} pero la abundancia de litio detectadas en meteoritos se conoce que está en el orden A(Li) = [3.26-3.34] dex \citep{Randich2006, Grevesse2007}, lo que apunta hacia un enriquecimiento desde que se produjo el Big Bang.
+Por otro lado, la abundancia medida en las estrellas de tipo solar es del orden de A(Li) = 1.05 dex \citep{Grevesse2007}, lo que implica la existencia de procesos de destrucción del litio en su interior.
+Esta discrepancia entre las abundancias medidas en meteoritos y en estrellas de tipo solar es lo que ha pasado a denominarse el "problema del litio".\par
 
-Continuando con la importancia del Li y teniendo en cuenta que tanto éste, como los demás elementos ligeros berilio y boro, se queman a temperaturas relativamente bajas en interiores estelares, tenemos que estos elementos sólo sobreviven en las capas más externas de una estrella, donde la temperatura es menor.
+Continuando con la importancia del litio y teniendo en cuenta que tanto éste, como los demás elementos ligeros se queman a temperaturas relativamente bajas en interiores estelares, tenemos que estos elementos sólo sobreviven en las capas más externas de una estrella, donde la temperatura es menor que la necesaria para su destrucción.
 Adicionalmente, su presencia es un trazador muy potente relacionado con los mecanismos de mezcla en funcionamiento en estructuras estelares, ya que nos puede aportar información acerca de qué tal eficientes son a la hora de transportar estos elementos a zonas más interiores donde serían destruidos.\par
@@
 Nuestro Sol, como estrella que es y como ocurre con el resto de los cuerpos celestes, gira en torno a su eje.
-El tiempo que tarda en completar una revolución completa sobre este es de aproximadamente 26 días.
+El tiempo que tarda en completar una revolución completa sobre éste es de aproximadamente 26 días.
 Decimos aproximadamente porque la duración del día solar es diferente dependiendo de la latitud que se tome como referencia.
-Si comparamos las velocidades de rotación solar y la terrestre con respecto al diámetro de los correspondiente cuerpos celestes, se obtiene que un punto situado sobre la superficie solar girar aproximadamente cuatro veces más rápido de lo que lo haría sobre la Tierra.
-
+Si comparamos las velocidades de rotación solar y la terrestre con respecto al diámetro de los correspondiente cuerpos celestes, se obtiene que un punto situado sobre la superficie solar girar sobre cuatro veces más rápido de lo que lo haría sobre la Tierra.
 Esto ocurre porque, a diferencia de la Tierra, el Sol no es un cuerpo rígido.
 Por lo tanto, al estudiar la rotación del Sol, no se puede considerar como una estructura compacta.
-El Sol es en realidad una enorme esfera de plasma, más parecida a una enorme bola de gas que a una estructura sólida rígida.\par
+El Sol es en realidad una enorme esfera de plasma, más parecida a una gigantesca bola de gas que a una estructura sólida rígida.\par
 
 Si tomamos como referencia medidas de la rotación solar a lo largo de un período largo de tiempo podemos observar que, dependiendo de la latitud que tomemos como referencia, la velocidad de rotación se va incrementando conforme la distancia al ecuador solar se reduce.
-Es decir, el Sol gira más rápidamente en el ecuador que en sus polos \ref{fig:rot_solar_vs_latitud}.
-Este comportamiento que se conoce como rotación diferencial fue descrito por Richard Carrington (1826-1875) tras analizar el movimiento de las manchas solares \cite{Carrington1863}.
-Carrington dedujo la siguiente ley empírica mediante donde $\Omega_*$ es la cantidad angular de la rotación diaria y $\phi$ es la latitud heliográfica, que corresponde a la latitud en la Tierra  \cite{Banisch2009}.\par
+Es decir, el Sol gira más rápidamente en el ecuador que en sus polos (ver Figura \ref{fig:rot_solar_vs_latitud}).
+Este comportamiento que se conoce como rotación diferencial fue descrito por Richard Carrington (1826-1875) tras analizar el movimiento de las manchas solares \citep{Carrington1863}.
+Carrington dedujo la siguiente ley empírica mediante donde $\Omega_*$ es la cantidad angular de la rotación diaria y $\phi$ es la latitud heliográfica, que corresponde a la latitud en la Tierra  \citep{Banisch2009}.\par
 
@@
     \caption{Rotación diferencial presente en el Sol.
-Tomando como referencia un meridiano solar y situando sobre este diferentes puntos de referencias a diferentes latitudes, se observa que la velocidad con la que estos giran es mayor a medida la distancia al ecuador solar disminuye.
+Tomando como referencia un meridiano solar y situando sobre este diferentes puntos de referencias a diferentes latitudes, se observa que la velocidad con la que estos giran es mayor a medida que la distancia al ecuador solar disminuye.
 Crédito McGraw-Hill.}
@@
 
-En la región ecuatorial la rotación dura solo 25 días, mientras que en los polos el Sol gira más lentamente \ref{fig:ang_solar_vs_latitud}.
+En la región ecuatorial la rotación dura solo 25 días, mientras que en los polos el Sol gira más lentamente (ver Figura \ref{ver Figura fig:ang_solar_vs_latitud}).
 Aquí el día solar tiene una duración aproximada de 30 días.
-A pesar de esta rotación diferencial entre regiones situadas en latitudes diferentes, no se produce un aplanamiento reseñable del disco solar ya que la velocidad de rotación es pequeña, de media 2km/s \cite{Gill2012}.\par
+A pesar de esta rotación diferencial entre regiones situadas en latitudes diferentes, no se produce un aplanamiento reseñable del disco solar ya que la velocidad de rotación es pequeña, de media 2km/s \citep{Gill2012}.\par
 
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{img/tesis/sun_3.png}
+    \includegraphics[width=0.7\textwidth]{img/tesis/sun_3.png}
     \caption{Valor predicho teóricamente para la velocidad de rotación del Sol en función de la latitud.
@@
 \section{Efectos de los campos magnéticos}
-Aún se desconoce el origen de los campos magnéticos en las estrellas.
-Los campos magnéticos organizados detectados en algunas estrellas O \cite{Wade2010} podrían ser campos fósiles \cite[ver][para más detalles ]{Dudorov2014}, o campos producidos mediante un mecanismo de dinamo \cite{Cantiello2009}.
-Además, la presencia de campos magnéticos con intensidades del orden de kG \cite{Hussain2014} ha sido una condición sine qua non para tratar ciertos datos observados en estrellas de tipo T Tauri en sistemas PMS jóvenes en acreción \cite{Johns-Krull2007}.\par
-
 La masa de las estrellas juega un papel determinante en la existencia o no de una envoltura convectiva externa.
@@
 En cambio, las estrellas menos masivas, como el Sol, tienen envolturas convectivas y un núcleo radiativo.
-En ambos tipos de estrellas se pueden generar fuertes campos magnéticos en la tacoclina pero con notables diferencias \cite[v.g.][para más detalles]{Chabrier2006,Charbonneau2010}.\par
+En ambos tipos de estrellas se pueden generar fuertes campos magnéticos en la tacoclina pero con notables diferencias \citep[v.g.][para más detalles]{Chabrier2006,Charbonneau2010}.\par
 
-En el Sol, la materia que lo compone se encuentra en su mayoría presente en forma de plasma.
-Es este estado, la envoltura electrónica se encuentra desacoplada del núcleo atómico y bajo ciertas condiciones físicas pueden moverse libremente, por lo que pueden poner en movimiento una carga electrónica.
-Bajo estas condiciones, la materia solar puede alcanzar la conductividad del cobre \cite{Banisch2009}.\par
+Aun así, a día de hoy aún no existen un consenso acerca de cuál es el origen de los campos magnéticos en las estrellas.
+Los campos magnéticos organizados detectados en algunas estrellas O \citep{Wade2010} podrían ser campos fósiles \citep[ver][para más detalles ]{Dudorov2014}, pero también podrían deberse a campos producidos mediante un mecanismo de dinamo \citep{Cantiello2009}.
+Además, la presencia de campos magnéticos con intensidades del orden de kG \citep{Hussain2014} ha sido una condición sine qua non para tratar ciertos datos observados en estrellas de tipo T Tauri en sistemas PMS jóvenes en acreción \citep{Johns-Krull2007}.\par
 
@@
 Para que una estrella tenga un campo magnético generado por dinamo, debe contener una región fluida conductora de electricidad que experimente movimientos para generar inducción.
-Sin embargo, existen otras condiciones necesarias para el vigor y la morfología de los movimientos que se derivan del hecho de que el campo magnético no debe decaer debido a la disipación óhmica.
+En el Sol, la materia que lo compone se encuentra en su mayoría presente en forma de plasma.
+Es este estado, la envoltura electrónica se encuentra desacoplada del núcleo atómico y bajo ciertas condiciones físicas pueden moverse libremente, por lo que pueden poner en movimiento una carga electrónica.
+Bajo estas condiciones, la materia solar puede alcanzar la conductividad del cobre \citep{Banisch2009}.
 La necesidad básica para mantener un mecanismo de dinamo es una región fluida eléctricamente conductora en el interior de la estrella.
+Aún así, deben de existir otras condiciones necesarias para el vigor y la morfología no decaiga debido a la disipación óhmica.
+\par
 
-El Sol se rige por un fuerte campo magnético que se genera con una intensidad de campo magnético de B z 10 5 G \cite{Aschwanden2014} en la tacoclina, la delgada capa de cizalla intercalada entre la zona radiativa y la convectiva.
-Los tubos de flujo magnético flotante se elevan a través de la zona de convección (debido a la inestabilidad convectiva que obedece a la ley de Schwarz).
-que obedece al criterio de Schwarzschild) y emergen en la superficie solar en las regiones activas, donde forman manchas solares con intensidades de campo magnético de B z 10 3 G y bucles coronales con intensidades de campo de B z 10 2 G en los puntos de apoyo fotosféricos, y de B z 10 G en las mayores alturas coronales.
-Se cree que la rotación diferencial en la superficie solar enrolla el campo magnético superficial, que se fragmenta bajo la tensión magnética, circula meridionalmente hacia los polos y se reorienta desde el estado de tensión toroidal (con líneas de campo orientadas en dirección este-oeste) en el máximo solar a un campo dipolar poloidal (que conecta el polo norte con el polo sur) en el mínimo solar.
-mínimo solar.
+El Sol se rige por un fuerte campo magnético que se genera con una intensidad de campo magnético de B z 10 5 G --BUSCAR QUÉ ES B z 10-- \citep{Aschwanden2014} en la tacoclina, la delgada capa de cizalla intercalada entre la zona radiativa y la convectiva.
+Los tubos de flujo magnético flotante se elevan a través de la zona de convección (debido a la inestabilidad convectiva que obedece a la ley de de Schwarzschild) y emergen en la superficie solar en las regiones activas, donde forman manchas solares con intensidades de campo magnético de B z 10 3 G --AQUÍ TAMBIÉN-- y bucles coronales con intensidades de campo de B z 10 2 G --Y AQUÍ-- en los puntos de apoyo fotosféricos, y de B z 10 G -- Y AQUÍ-- en las mayores alturas coronales.
+Se cree que la rotación diferencial en la superficie solar enrolla el campo magnético superficial, que se fragmenta bajo la tensión magnética, circula meridionalmente hacia los polos y se reorienta desde el estado de tensión toroidal (con líneas de campo orientadas en dirección este-oeste) en el máximo solar a un campo dipolar poloidal (que conecta el polo norte con el polo sur) en el mínimo solar.\par
 
 \subsection{Campos magnéticos internos}
-Nuestro Sol es una estrella activa en lo que se refiere al campo magnético.
-Según los modelos teóricos este campo magnético es generado por una dinamo en la tacoclina y con una intensidad en torno a $B\approx10^{5}\,\Gauss$ \cite{Maeder2003a,Dudorov2014}.
+--MIRAR QUE NO ENTRAMOS EN CONTRADICCIÓN CY REPETICIÓN CON LOS VALORES ANTERIORES--
+Como acabamos de comentar, el Sol es una estrella activa en lo que se refiere al campo magnético.
+Según los modelos teóricos este campo magnético es generado por una dinamo en la tacoclina y con una intensidad en torno a $B\approx10^{5}\,\Gauss$ \citep{Maeder2003a,Dudorov2014}.
 En el momento de su generación, las líneas de ese campo magnético tendrán una orientación poloidal y se verán afectadas por la rotación diferencial y obligadas a adoptar una tipología toroidal (el efecto $\omega$).
@@
 Esto sería posible gracias a los movimientos convectivos de material que se producen en el CZ de la estrella y suponer que este material se comporta como un conductor perfecto.
-Esta segunda condición permite suponer que las líneas de campo magnético están "congeladas" en el fluido conductor y por tanto deben moverse solidariamente con él (efecto $\alpha$) \cite{Stanley2014}.\par
+Esta segunda condición permite suponer que las líneas de campo magnético están "congeladas" en el fluido conductor y por tanto deben moverse solidariamente con él (efecto $\alpha$) \citep{Stanley2014}.\par
 
@@
 \subsection{Campos magnéticos superficiales} \label{surf_mf}
-La intensidad del campo magnético medida en la superficie del Sol es de aproximadamente $B\approx1\,\Gauss$ \cite{Weber1967,DAntona2000,Morin2012}, el doble del valor del de la Tierra.
+La intensidad del campo magnético medida en la superficie del Sol es de aproximadamente $B\approx1\,\Gauss$ \citep{Weber1967,DAntona2000,Morin2012}, el doble del valor del de la Tierra.
 Además su topología es variada y compleja.
 El modelo de dinamo establece que el campo magnético producido en el interior de la estrella alcanza la superficie con la ayuda de procesos convectivos que tienen lugar en la CZ y da lugar a regiones magnéticamente activas en la superficie solar.
-Además, las estrellas en rotación que exponen una CZ exterior significativa pueden producir campos magnéticos superficiales a través de un mecanismo de dinamo \cite[v.g.][]{Brandenburg2004,Charbonneau2010,Brun2017}.
-Sea cual sea el origen de los campos magnéticos superficiales, se predice un acoplamiento con el viento estelar, provocando así una pérdida de masa y, si ésta es lo suficientemente fuerte, produzcan MB \cite[v.g.][]{UdDoula2002,Ud-Doula2007,Ud-Doula2008,Meynet2010}.\par
+Además, las estrellas en rotación que exponen una CZ exterior significativa pueden producir campos magnéticos superficiales a través de un mecanismo de dinamo \citep[ver][para más detalles]{Brandenburg2004,Charbonneau2010,Brun2017}.
+Sea cual sea el origen de los campos magnéticos superficiales, se predice un acoplamiento con el viento estelar, provocando así una pérdida de masa y, si ésta es lo suficientemente fuerte, produzcan MB \citep[ver][para más detalles]{UdDoula2002,Ud-Doula2007,Ud-Doula2008,Meynet2010}.\par
 
 \section{Efectos del frenado magnético - Parte I}
-En su enfoque más aceptado, el MB está vinculado al trabajo de Skumanich \cite{Skumanich1972} en el que desarrolla una ley empírica de su evolución en paralelo a la edad de la estrella.
+En su enfoque más aceptado, el MB está vinculado al trabajo de Skumanich \citep{Skumanich1972} en el que desarrolla una ley empírica de su evolución en paralelo a la edad de la estrella.
 Para calibrar esta ley, Skumanich se basó en datos obtenidos de estrellas de tipo G encontradas en la MS.
@@
 El AML causado por MB depende directamente de la cantidad de masa perdida por la estrella debido a los vientos estelares.
-El coeficiente de pérdida de masa estimado para una estrella de tipo solar, aproximadamente $10^{-14}\msun \; yr^{-1}$ \cite{Noerdlinger2008}, y el AML resultante derivado de ese valor pueden considerarse relativamente modesto como para influir decisivamente en la evolución de la estrella.
+El coeficiente de pérdida de masa estimado para una estrella de tipo solar, aproximadamente $10^{-14}\msun \; yr^{-1}$ \citep{Noerdlinger2008}, y el AML resultante derivado de ese valor pueden considerarse relativamente modesto como para influir decisivamente en la evolución de la estrella.
 Además, el efecto combinado de los movimientos convectivos y la rotación diferencial conducen a la generación de campos magnéticos.
-Esos campos magnéticos acaban alcanzando la superficie de la estrella, de modo que el AML puede incrementarse en varios órdenes de magnitud \cite{Langer2012}.
-Ese incremento es consecuencia del campo magnético que obliga a las partículas ionizadas del viento estelar a girar con la misma velocidad angular, vientos que se extienden una distancia varias veces superior al radio de la estrella \cite[ver][para más detalles]{UdDoula2002,Ud-Doula2007,Ud-Doula2008}.
+Esos campos magnéticos acaban alcanzando la superficie de la estrella, de modo que el AML puede incrementarse en varios órdenes de magnitud \citep{Langer2012}.
+Ese incremento es consecuencia del campo magnético que obliga a las partículas ionizadas del viento estelar a girar con la misma velocidad angular, vientos que se extienden una distancia varias veces superior al radio de la estrella \citep[ver][para más detalles]{UdDoula2002,Ud-Doula2007,Ud-Doula2008}.\par
 
@@
 Esto es muy importante, por ejemplo, en las estrellas masivas, que se sabe que rotan rápidamente y tienen fuertes vientos estelares.
-También se ha observado que las estrellas de masa intermedia fuertemente magnéticas suelen tener tasas de rotación mucho más lentas que otras estrellas de su población parental \cite{Mathys2006}.
-En esas estrellas, la presencia de dicho campo magnético interactuará con la pérdida de masa y el radio Alfv'{e}n ($R_{A}$) juega un papel importante en esta interacción.\par
-
-En el resto de este trabajo, describimos dos enfoques semi-empíricos para el cálculo del AML como resultado del par aplicado por un viento estelar acoplado magnéticamente.
-Esto se ha implementado como una extensión del código de evolución estelar Modules for Experiments in Stellar Astrophysics \cite[MESA; ][]{Paxton2011, Paxton2013,Paxton2015, Paxton2018, Paxton2019}.
+También se ha observado que las estrellas de masa intermedia fuertemente magnéticas suelen tener tasas de rotación mucho más lentas que otras estrellas de su población parental \citep{Mathys2006}.
+En esas estrellas, la presencia de dicho campo magnético interactuará con la pérdida de masa y el radio Alfven ($R_{A}$) juega un papel importante en esta interacción.
+En las siguientes secciones, describimos dos enfoques semi-empíricos para el cálculo del AML como resultado del par aplicado por un viento estelar acoplado magnéticamente.\par
 
 \subsubsection{Frenado magnético de intesidad fija según Ud-Doula \& Owocki (2002)}
+--EMPEZAR ASÍ---
+Este modelo adopta una prescripción de ...
 Un parámetro relevante para caracterizar la influencia de un campo magnético dado sobre el viento estelar es el denominado parámetro magnético de confinamiento del viento ($\eta_*$).
 Representa una relación energética (Ec.~\ref{eq:wind_conf}) y define un parámetro característico de la eficacia relativa de los campos magnéticos para circunscribir y/o canalizar el flujo de salida del viento \cite{UdDoula2002}.\par
-
-\begin{ceqn}
-\begin{equation}
-    \eta_* = \frac{B^{2}/8 \pi}{3} \label{eq:wind_conf2}
-\end{equation}
-\end{ceqn}
 
@@
 
-Haciendo uso de $\eta_*$, el $R_{A}$ se define como el punto en el que se equilibran la densidad de energía del campo magnético, marcada $B$, y la densidad de energía cinética, definida por $\rho$ y $(\nu)$.
+Haciendo uso de $\eta_*$, el $R_{A}$ se define como el punto en el que se equilibran la densidad de energía del campo magnético, representada por $B$, y la densidad de energía cinética, definida por $\rho$ y $(\nu)$.
 Es decir, donde el coeficiente entre ambos es igual a 1.
@@
 
-\begin{ceqn}
 \begin{equation}
-    \Dot{M} = 4\pi r^2\rho\nu \label{eq:mass_loss}
+    \Dot{M} = 4\pi r^2\rho\nu \label{eq:mass_loss1}
 \end{equation}
-\end{ceqn}
 
-Usando (\ref{eq:wind_conf}) and (\ref{eq:mass_loss}), $\eta_*$ puede ser aproximado por: 
-\begin{ceqn}
+Usando (\ref{eq:wind_conf}) y (\ref{eq:mass_loss1}), $\eta_*$ puede ser aproximado por: 
 \begin{equation}
@@
 \end{equation}
-\end{ceqn}
 
@@
 
-\begin{ceqn}
-\begin{equation}
-    \eta_* = \frac{B^{2}r^{2}}{\Dot{M}\nu} \label{eq:wind_conf2}
-\end{equation}
-\end{ceqn}
-
-\begin{equation}\label{eq:identidad-pitagorica2}
-  \cos^2 x + \sin^2 x = 1.
-\end{equation}
-
-A su vez, de las características observables del viento solar, nos importan especialmente los parámetros relacionados con la cantidad de masa que es capaz de arrancar de las capas externas de la estrella en un intervalo de tiempo dado, y la velocidad que alcanza el propio viento medida a gran distancia de la estrella, la velocidad terminal ($\nu_\infty$).
-En general, un flujo de salida canalizado magnéticamente tendrá una geometría de corriente compleja pero, por conveniencia, la Ec.~\ref{eq:mass_loss} simplemente caracteriza la fuerza del viento en términos de una tasa de pérdida de masa distribuida simétricamente sobre una esfera.
+A su vez, de las características observables del viento solar, nos importan especialmente los parámetros relacionados con la cantidad de masa que es capaz de arrancar de las capas externas de la estrella en un intervalo de tiempo dado, y la velocidad que alcanza el propio viento medida a gran distancia de la estrella, la denominada velocidad terminal ($\nu_\infty$).
+En general, un flujo de salida canalizado magnéticamente tendrá una geometría de corriente compleja pero, por conveniencia, la Ec.~\ref{eq:mass_loss1} simplemente caracteriza la fuerza del viento en términos de una tasa de pérdida de masa distribuida simétricamente sobre una esfera.
 $\nu$ puede caracterizarse por la variación radial de la velocidad del flujo saliente en términos de la ley de velocidad (Ec.~\ref{eq:vel_law}), donde $r$ representa la distancia desde el centro de la estrella hasta el punto donde se va a medir la velocidad, y $R_*$ es el radio de la estrella:\par
 
-\begin{ceqn}
 \begin{equation}
@@
 \end{equation}
-\end{ceqn}
-donde $\nu_\infty$ es la velocidad terminal del viento definida como la velocidad que alcanza el viento, la materia que fluye a gran distancia de la estrella central, donde ya no es acelerada por la fuerza impulsora del viento pero su desaceleración debida a la interacción con el medio interestelar (ISM) es despreciable \cite{Niedzielski2002}.\par
 
-Los vientos impulsados por líneas de las estrellas OB masivas tienen velocidades terminales que son directamente proporcionales a la velocidad de escape fotosférica \cite{Lamers2000} según:\par
+donde $\nu_\infty$ es la velocidad terminal del viento definida como la velocidad que alcanza el viento, la materia que fluye a gran distancia de la estrella central, donde ya no es acelerada por la fuerza impulsora del viento, y su vez la desaceleración debida a la interacción con el medio interestelar (Inter-Stellar Medium, ISM) es despreciable \citep{Niedzielski2002}.\par
+
+Para el caso de estrellas OB masivas, se estima que los vientos impulsados por sus líneas de campo magnéticas tienen velocidades terminales que son directamente proporcionales a la velocidad de escape fotosférica \citep{Lamers2000} según:\par
 
@@
 Haciendo uso de (\ref{eq:vesc}), $\nu_\infty$ puede reformularse como:
-\begin{ceqn}
 \begin{equation}
@@
 \end{equation}
-\end{ceqn}
 donde ${\rsun}$ es el radio, y ${\msun}$ la masa del Sol.
 
-Dado que la versión de MESA utilizada en este trabajo no incluye el tratamiento de los campos magnéticos superficiales (sí tiene en cuenta el transporte por campos magnéticos internos de momento angular y elementos químicos) y, por tanto, tampoco los efectos de frenado magnético, es necesario extender el simulador para incluir este fenómeno.
-En la presente tesis hemos seguido el trabajo teórico sobre el AM en el Sol realizado por \cite{Weber1967}, haciendo uso de la Ec.~ \ref{eq:j_dot} para el cálculo del AML.\par
+Se ha observado que las estrellas de masa intermedia fuertemente magnéticas suelen tener tasas de rotación mucho más lentas que otras estrellas de su población parental \cite{Mathys2006}.
+En esas estrellas, los campos magnéticos interactúan con la pérdida de masa, donde el radio Alfven ($R_{A}$) juega un papel importante.
+Haciendo uso de $\eta_*$ (Ec.
+\ref{eq:wind_conf2}), $R_{A}$ se define como el punto en el que la densidad de energía del campo magnético y la densidad de energía cinética se equilibran.
+En caso de que $R_{A}$ sea mayor que el radio estelar, entonces el flujo del viento tendrá que seguir el campo magnético.
+Como consecuencia, el material abandona la superficie estelar con un AM específico mayor, ya que el radio de co-rotación ha aumentado y corresponde aproximadamente a $R_{A}$.\par
 
-Se ha observado que las estrellas de masa intermedia fuertemente magnéticas suelen tener tasas de rotación mucho más lentas que otras estrellas de su población parental \cite{Mathys2006}.
-En esas estrellas, los campos magnéticos interactúan con la pérdida de masa, donde el radio Alfv'{e}n ($R_{A}$) juega un papel importante.
-Haciendo uso de $\eta_*$ (Ec.
-\ref{eq:wind_conf}), $R_{A}$ se define como el punto en el que la densidad de energía del campo magnético y la densidad de energía cinética se equilibran.
-En caso de que $R_{A}$ sea mayor que el radio estelar, entonces el flujo del viento tendrá que seguir el campo magnético.
-Como consecuencia, el material abandona la superficie estelar con un AM específico mayor, ya que el radio de co-rotación ha aumentado y corresponde aproximadamente a $R_{A}$.
-Siguiendo a \cite{Weber1967} el AML puede calcularse mediante:
+Dado que la versión de MESA utilizada en este trabajo no incluye el tratamiento de los campos magnéticos superficiales (sí tiene en cuenta el transporte por campos magnéticos internos de momento angular y elementos químicos) y por tanto, tampoco los efectos de frenado magnético, es necesario extender el simulador para incluir este fenómeno.
+En la presente tesis hemos seguido el trabajo teórico sobre el AM en el Sol realizado por \cite{Weber1967}, haciendo uso de la Ec.~ \ref{eq:j_dot1} para el cálculo del AML.
+Siguiendo a \cite{Weber1967} la AML puede calcularse mediante:
 
-\begin{ceqn}
 \begin{equation}
- \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A} \label{eq:j_dot}
+ \Dot{J} = \frac{2}{3} \Dot{M}\Omega R^{2}_{A} \label{eq:j_dot1}
 \end{equation}
-\end{ceqn}
 
-donde $\Dot{M}$ es la tasa de pérdida de masa, $\Omega$ la velocidad angular en la superficie de la estrella y $R_A$ el radio Alfv\'{e}n. \par
+donde $\Dot{M}$ es la tasa de pérdida de masa, $\Omega$ la velocidad angular en la superficie de la estrella y $R_A$ el radio Alfven.
+La expresión anterior se puede reescribir según \cite{Ud-Doula2008} de modo que dependa de $\eta_*$ en lugar de $R_A$, quedando así una expresión más conveniente para ser implementada en MESA porque se basa en valores directamente expuestos durante las simulaciones.
 
-La expresión anterior se puede reescribir según \cite{Ud-Doula2008} de modo que dependa de $\eta_*$ en lugar de $R_A$ como sigue:
-
-\begin{ceqn}
 \begin{equation}
@@
 
-\end{ceqn}
-que es una expresión más conveniente para ser implementada en MESA porque se basa en valores directamente expuestos durante las simulaciones.
-
 \section{Efectos de la distrubución de AML}
 \subsection{Formalismo para la distribución de AML} \label{mod_aml}
-Como expresa la Eq.~\ref{eq:mb_torque}, la cantidad de AML depende de $\ralfven$, $\Omega$ y $\Dot{M}$.
+Como expresa la Eq.~\ref{eq:j_dot1}, la cantidad de AML depende de $\ralfven$, $\Omega$ y $\Dot{M}$.
 Para valores grandes de $\ralfven$, la estrella sufre una desaceleración significativa.
 El valor de $\ralfven$ depende de los siguientes parámetros variables: intensidad del campo magnético ($B$), $\teff$ y $\Omega$.
-Con respecto a $\Dot{M}$ y como se indica en la tabla \ref{tab:phy_mesa}, para calcular la pérdida de masa se utilizó la fórmula empírica desarrollada por Reimers \cite{Reimers1975} para estrellas en la rama gigante asintótica (AGB).
-Para una estrella de tipo solar el $\Dot{M}$ durante la EM es relativamente pequeño, alrededor de $10^{-14}\msun \, yr^{-1}$ \cite{Noerdlinger2008}.
+(----TENEMOS FÓRMULA?----) Con respecto a $\Dot{M}$ y como se indica en la tabla \ref{tab:phy_mesa}, para calcular la pérdida de masa se utiliza la fórmula empírica desarrollada por Reimers \citep{Reimers1975} para estrellas en su fase AGB.
+Para una estrella de tipo solar la $\Dot{M}$ durante la EM (-----MS????
+MIRAR EN PAPER-----) es relativamente pequeña, alrededor de $10^{-14}\msun \, yr^{-1}$ \citep{Noerdlinger2008}.
 \par
 
-MESA asigna un valor $\Omega$ para cada celda $k$ ($\Omega_k$) que se ajusta para que el momento angular resultante se conserve tras calcular la nueva masa de la celda $k$ ($m_k$) y su distancia al centro de la estrella ($r_k$).
-A continuación, se asigna un valor de AM a cada celda $k$ ($J_k$).
-En este punto, nuestro MB se activa, modificando $J_k$.
-Para ello, se proporciona una contribución adicional ($\Dot{J}_{k}$).
-Esta contribución es el resultado del par externo ejercido por el campo magnético una vez que se ha distribuido entre las diferentes capas que componen la CZ según dicta la Eq.~\ref{eq:k_jdot}:\par
+\section{Extendiendo MESA - Parte II}
 
-\begin{ceqn}
+\subsection{Rutina de frenado magnético de intensidad constante} \label{sec:mb_cte_b}
+En lo que MESA atañe, el simulador asigna un valor $\Omega$ para cada celda $k$ ($\Omega_k$) que se ajusta para que el AM resultante se conserve tras calcular la nueva masa de la celda $k$ ($m_k$) y su distancia al centro de la estrella ($r_k$).
+A continuación, le asigna un valor de AM a cada celda $k$ ($J_k$).
+En este punto, nuestra rutina de MB se activa, modificando $J_k$.
+Para ello, se proporciona una contribución adicional ($\Dot{J}_{k}$), también de signo negativo porque estamos ante una pérdida.
+Ésta es el resultado del par externo ejercido por el campo magnético una vez que se ha distribuido entre las diferentes capas que componen la CZ según dicta la Ec.~\ref{eq:k_jdot}:\par
+
 \begin{align}
@@
 \end{align}
-\end{ceqn}
 
-La distribución se realiza en dos iteraciones.
-En la primera iteración, a cada celda $k$ se le asigna la cantidad máxima de $\Dot{J}_{k}$ que puede acomodar en el paso de tiempo establecido por la simulación.
-Las cantidades de $\Dot{J}_{k}$ no acomodadas por las celdas se acumulan como momento angular residual ($\jres$).
-En una segunda pasada, el valor $\jres$ residual acumulado recogido en la iteración anterior se redistribuye entre aquellas celdas que pueden acomodar más par.
-De esta forma conseguimos conservar el momento angular.\par
-
-\section{Extendiendo MESA - Parte II}
-
-\subsection{Rutina de frenado magnético de intensidad constante}
 EXPLICAR QUE SUSTITUIMOS LA RUTINA POR DEFECTO CON LA NUESTRA
-\begin{lstlisting}[language=Fortran, caption={Rutina de par de torsión.}, label={lst:jdot_cantiello}]
+\begin{lstlisting}[language=Fortran, caption={Rutina de par de torsión para un campo magnético de intensidad fija.}, label={lst:jdot_cantiello}]
   real function calculate_jdot_rate_cantiello(s, bf_star) result(new_j_dot)
@@
 	omega_surf = s% omega(1)
-	!omega_surf = s% omega_avg_surf
 
@@
 
-	!m_dot = s% star_mdot !This gives the mass loss rate in Mstar/year
 	m_dot = s% mstar_dot !This in g/s
@@
 
-\subsection{Rutina de distribución de pérdida de momento angular}
-EXPLICAR QUE LOS DOS TIPOS DE RUTINAS: INTESIDAD FIJA, INTENSIDAD VARIABLE
+\subsection{Rutina de distribución de pérdida de momento angular} \label{sec:dist_aml1}
+La distribución de AML se realiza en dos iteraciones.
+En la primera, a cada celda $k$ se le asigna la cantidad máxima de $\Dot{J}_{k}$ que puede acomodar en el paso de tiempo establecido por la simulación.
+Las cantidades de $\Dot{J}_{k}$ no acomodadas por las celdas se acumulan como momento angular residual ($\jres$).
+En una segunda pasada, el valor $\jres$ residual acumulado recogido en la iteración anterior se redistribuye entre aquellas celdas que pueden acomodar más par.
+De esta forma conseguimos conservar el momento angular.\par
+
+--ESTE LISTADO NO ES EL BUENO YA QUE NO INCLUYE LA SEGUNDA PASADA--
+
 \begin{lstlisting}[language=Fortran, caption={Rutina de distribución de pérdida de momento angular.}, label={lst:dis_jdot}]

### 2024-03-20 project/overleaf/Tesis III/capitulos/capitulo04.tex
WIP

@@
 Los resultados de abundancia de litio que obtengamos en cada uno de los escenarios planteados los enfrentaremos con los que arroja MESA sin hacer uso de nuestra rutina.
-Esta comparación de resultados nos deberá de dar un primer indicio de si nuestro planteamiento realmente tiene algún efecto notable sobre la abundancia litio calculadas por el simulador con y sin nuestras rutinas.
+Esta comparación de resultados nos deberá de dar un primer indicio de si nuestro planteamiento realmente tiene algún efecto notable sobre la abundancia litio calculada por el simulador con y sin nuestras rutinas.
 Finalmente, en base a los valores obtenidos, analizaremos las consecuencias que nuestras rutinas han tenido sobre ellos y presentaremos una serie de conclusiones de por qué obtenemos estos datos y cómo se interpretan los mismos en base a los estudios teóricos y observaciones obtenidas por otras líneas de investigación.\par
@@
 \subsection{El proyecto MESA}
-Como indica su contribuidor y desarrollador principal \author{QUEPONERAQUÍ??} \cite[xxxxxx]{Pasetto2014} (Paxton, et al., 2011, 2013, 2015) MESA es un conjunto de librerías de código abierto, robustas y eficientes escritas en el lenguaje de programación Fortran 95 que es susceptible de ser utilizado en una amplia gama de aplicaciones en astrofísica estelar computacional.
-Está categorizado dentro de los denominados códigos de evolución estelar unidimensional (1-D) y combina un gran número de módulos numéricos y físicos que le permiten simular un gran número de escenarios de evolución estelar, que van desde los que incluyen estrellas de muy baja masa hasta los que caracterizan estrellas masivas, además de tener en cuenta fases avanzadas de evolución, como el flash del helio, pulsos térmicos o la rama asintótica gigante (XXXXXXXX XXXXXXXX AGB).
+Como indica su contribuidor y desarrollador principal Bill Paxton \citep{Paxton2011, Paxton2013, Paxton2015, Paxton2018} MESA es un conjunto de librerías de código abierto, robustas y eficientes escritas en el lenguaje de programación Fortran 95 que es susceptible de ser utilizado en una amplia gama de aplicaciones en astrofísica estelar computacional.
+Está categorizado dentro de los denominados códigos de evolución estelar unidimensional (1-D) y combina un gran número de módulos numéricos y físicos que le permiten simular una amplia variedad de escenarios de evolución estelar, que van desde los que incluyen estrellas de muy baja masa hasta los que caracterizan estrellas masivas, además de tener en cuenta fases avanzadas de evolución, como el flash del helio, pulsos térmicos o la rama asintótica gigante (Asymptotic Giant Branch, AGB).
 Utiliza un modelo de malla o zonas adaptables, que representa las diferentes capas y celdas en las que se estructura el modelo estelar, y emplea sofisticados controles de evolución temporal.\par
@@
 Por otra parte, MESA se apoya en ficheros de proyecto en los que se fijan los diferentes parámetros iniciales, bajo qué condiciones tiene que evolucionar la estrella, opciones de paradas y representación gráfica que deben utilizarse durante la simulación.
-MESA ofrece en su distribución un buen número de proyectos pre configurados para diferentes escenarios de evolución estelar.
-A continuación, pasamos a comentar en más detalle el contenido del fichero de configuración de proyecto que en MESA se denomina \textit{inlist}.\par
+MESA ofrece en su distribución un buen número de proyectos pre configurados para diferentes escenarios de evolución estelar.\par
 
-\subsubsection{Fichero de configuración}
+\subsection{Fichero de configuración}
 MESA utiliza el fichero \textit{inlist} como el punto de configuración y entrada para las simulaciones.
@@
 \begin{itemize}
-    \item \textit{star\_job} - aquí encontramos las opciones que controlan cómo evoluciona la estrella
-    \item \textit{controls} - opciones para el módulo star de MESA
-    \item \textit{pgstar} - opciones para la salida gráfica por pantalla
+    \item \textit{star\_job} - Sección con opciones que controlan cómo evoluciona la estrella
+    \item \textit{controls} - Sección con opciones para el módulo \textit{star} de MESA
+    \item \textit{pgstar} - Sección con opciones para la salida gráfica por pantalla
 \end{itemize}
@@
 Para cada una de las secciones comentadas anteriormente, MESA dispone de un fichero específico (\$MESA\_DIR hace referencia al directorio donde se encuentra instalado el software):
+
 \begin{itemize}
@@
 
-\subsubsection{Controlando la salida} \label{sec:ctrl_salida}
+\subsection{Controlando la salida} \label{sec:ctrl_salida}
 MESA genera por defecto una cantidad ingente de información.
@@
 
-\subsection{¿Cómo extender MESA?}
+\section{¿Cómo extender MESA?}
 A veces se hace necesario extender la funcionalidad de MESA porque la que viene por defecto en el simulador no cumple con las necesidades del problema que queremos abordar.
@@
 
-MESA proporciona el archivo \textit{run\_star\_extras.f} que ofrece una variedad de puntos de extensión (hooks) destinados a implementar acciones que no pueden hacerse únicamente mediante el uso del fichero de proyecto \textit{inlists}.
+MESA proporciona el archivo \textit{run\_star\_extras.f} que ofrece una variedad de puntos de extensión (\textit{hooks}) destinados a implementar acciones que no pueden hacerse únicamente mediante el uso del fichero de proyecto \textit{inlists}.
 Entre estas acciones adicionales tenemos la posibilidad de cambiar parámetros en cada paso de la simulación, o la de reemplazar muchas de las rutinas de física predeterminadas que trae el simulador.\par
 
-\subsubsection{Activación de capacidades extras}
+\subsection{Activación de capacidades extras}
 El primer paso para hacer uso de las capacidades extendidas de MESA es el de proceder a activarlas.
@@
 Veremos que el contenido del mismo es bastante simple, se limita a incluir otro fichero que es el que contiene el conjunto de rutinas predeterminadas que se entrega con MESA.
-Las rutinas definidas en ese fichero que se incorpora son las que queremos personalizar, pero sin afectar a las versiones originales.
+Las rutinas definidas en ese fichero son las candidatas a ser personalizadas, pero lo haremos sin afectar a las versiones originales.
 Para ello, procederemos a reemplazar esta declaración de inclusión por el contenido del archivo referenciado, consiguiendo de este modo tener nuestra propia copia local y personal.\par
 
-\subsubsection{Extensión mediante hooks}
+\subsection{Extensión mediante hooks}
 MESA proporciona una forma de sustituir la mayoría de las rutinas físicas que ofrece por defecto sin la necesidad de, como apuntábamos anteriormente, tener que modificar el código original del programa mediante el uso de los denominados \textit{hooks} o puntos de extensión.
@@
 \begin{enumerate}
-    \item Activar el hook deseado en el fichero \textit{run\_star\_extra.f} situado en el directorio \textit{src} del proyecto en el que estamos trabajando
+    \item Activar el \textit{hook} deseado en el fichero \textit{run\_star\_extra.f} situado en el directorio \textit{src} del proyecto en el que estamos trabajando
     \item Indicar nuestra versión del módulo de física a sustituir/ampliar
@@
 Las diferentes rutinas presentes en \textit{run\_star\_extras.f} se invocan en diferentes puntos durante la ejecución de MESA.
-En el Listado \ref{lst:bucle_ctrl}, y a modo de pseudo\-código en Fortran, podemos encontrar el aspecto del bucle principal de control sobre el que itera MESA.\par
+En el Listado \ref{lst:bucle_ctrl}, y a modo de pseudo\-código en Fortran, podemos encontrar el aspecto del bucle principal de control sobre el que itera MESA.
+El corazón del simulador MESA es la sección \textit{step\_loop}.
+Es en ella donde se sitúa toda la maquinaria mediante la que MESA evalúa y resuelve las ecuaciones de la estructura estelar.\par
 
@@
 
-El corazón del simulador MESA es la sección \textit{step\_loop}. Es en ella donde se sitúa toda la maquinaria mediante la que MESA evalúa y resuelve las ecuaciones de la estructura estelar.\par
-
 Por otro lado, en la estructura de datos \textit{star\_info}, MESA almacena toda la información sobre la estrella que está evolucionando.
@@
 
-\subsubsection{Extensión ficheros de salida}
-Anteriormente hacíamos referencia a cómo controlar la información estándar que MESA ofrece a través de los dos tipos de fichero de salida: \textit{history} y \textit{profile} (ver \ref{sec:ctrl_salida}).
+\subsection{Extensión ficheros de salida}
+Anteriormente hacíamos referencia a cómo controlar la información estándar que MESA ofrece a través de los dos tipos de fichero de salida: \textit{history} y \textit{profile} (ver Sección \ref{sec:ctrl_salida}).
 Adicionalmente a esta opción, tenemos la posibilidad de enviar a alguno de estos dos ficheros información que, o bien la calcula ya de por sí MESA pero no está pensada para ser volcada en estos ficheros, o bien puede representar valores que vienen derivados de nuestras rutinas y que, por lo tanto, no forman parte de la parametrización estándar de MESA.\par
 
-Para cualquiera de los dos casos comentados existen un par de rutinas dentro del fichero \textit{run\_star\_extras.f} que nos permiten, mediante sencillas modificaciones en los métodos \textit{how\_many\_extra\_history\_columns} y \textit{data\_for\_extra\_history\_columns}, añadir cuántos parámetros queramos\footnote{En realidad, existe un máximo de 100 parámetros adicionales.} (ver Listado \ref{lst:extra_column}).
+Para cualquiera de los dos casos comentados existen un par de rutinas dentro del fichero \textit{run\_star\_extras.f} que nos permiten, mediante sencillas modificaciones en los métodos \textit{how\_many\_extra\_history\_columns} y \textit{data\_for\_extra\_history\_columns}, añadir cuántos parámetros queramos\footnote{En realidad, existe un máximo de 100 parámetros adicionales.} (ver Listado \ref{lst:extra_columns}).
 Para ello, simplemente hay que indicar la cantidad de nuevos parámetros en los que estamos interesados, el nombre de los mismos (ver Listado \ref{lst:extra_name_columns}) y de dónde obtener sus valores.\par
@@
 
-\subsection{Módulos de interés de MESA}
+\section{Módulos de interés de MESA}
 Para empezar, vamos a enumerar los módulos de MESA que hemos encontrado de especial relevancia a estudiar para los objetivos de nuestra línea de investigación.
@@
 
-\subsubsection{Módulo químico - chem}
+\subsection{Módulo químico - chem}
 El módulo \textit{chem} MESA está compuesto de una colección de datos, funciones y subrutinas encaminadas a la gestión de los elementos químicos y sus isótopos.
@@
 
-\subsubsection{Reacciones termonucleares – rates}
-El módulo \textit{rates} contiene los cocientes de reacciones termonucleares de \cite{Caughlan1988}, y \cite{Angulo1999}, siendo esta última opción la utilizada por defecto \cite{Paxton2011}.
+\subsection{Reacciones termonucleares – rates}
+El módulo \textit{rates} contiene los cocientes de reacciones termonucleares de \cite{Caughlan1988}, y \cite{Angulo1999}, siendo esta última opción la utilizada por defecto \citep{Paxton2011}.
 El conjunto de cocientes de reacciones incluido abarca más de 300 elementos e incluye las reacciones débiles necesarias para la quema de hidrógeno (emisión de positrones y captura de electrones), así como las reacciones de conversión neutrón-protón.
 
-\subsubsection{Reacciones nucleares – net} \label{reac_nuc}
+\subsection{Reacciones nucleares – net} \label{reac_nuc}
 El módulo \textit{net} implementa redes de reacción nuclear.
 Por defecto, incluye una red básica de 8 isótopos: \isotope[1]{H}, \isotope[3]{He}, \isotope[4]{He}, \isotope[12]{C}, \isotope[14]{N}, \isotope[16]{O}, \isotope[20]{Ne}, \isotope[24]{Mg}, y otras extendidas para cálculos más detallados incluyendo el ciclo CNO, captura de partículas $\alpha$ y reacciones de captura protónica.
-Además de utilizar las redes existentes, MESA nos permite definir nuestras propias redes de reacción construidas a partir de las ya existentes (opción más sencilla) o completamente desde cero \cite{Paxton2011}.\par
+Además de utilizar las redes existentes, MESA nos permite definir nuestras propias redes de reacción construidas a partir de las ya existentes (opción más sencilla) o completamente desde cero \citep{Paxton2011}.\par
 
-Como indicaremos más detalladamente cuando analicemos la configuración final del fichero de proyecto \textit{inlist}, las reacciones que más nos interesan son las de las cadenas protón-protón en sus variantes I, II y III (ver sección \ref{sim_mesa} para más detalles).\par
+Como indicaremos más detalladamente cuando analicemos la configuración final del fichero de proyecto \textit{inlist}, las reacciones que más nos interesan son las de las cadenas protón-protón en sus variantes I, II y III (ver Sección \ref{sec:li_reac_nuc} para más detalles).\par
 
-\subsubsection{Teoría de longitud de mezcla – mlt}
+\subsection{Teoría de longitud de mezcla – mlt}
 El módulo \textit{mlt} implementa la teoría de longitud de mezcla estándar (MLT) \citep{Paxton2011}.
@@
 
-\subsubsection{Módulo de difusión – diffusion} \label{subsec_diffusion}
+\subsection{Módulo de difusión – diffusion} \label{subsec_diffusion}
 El módulo \textit{diffusion} lo utiliza MESA para calcular la difusión de partículas y la sedimentación gravitacional resolviendo las ecuaciones de Burger \citep{Burgers1969} a través del método propuesto por \cite{Thoul1993}.
@@
 
-\subsubsection{Otros parámetros de interés}
+\subsection{Otros parámetros de interés}
 Además de los módulos listados anteriormente existen otra serie de parámetros que ofrece MESA que están relacionados también con el planteamiento teórico de nuestra investigación.
-En concreto nos estamos refiriendo a los parámetros de masa estelar, contenido de \isotope[1]{H}, \isotope[3]{He} y \isotope[4]{He} presente en la nube protoestelar, así como la metalicidad de la misma (Z).\par
+En concreto nos estamos refiriendo a los parámetros de masa estelar, contenido de \isotope[1]{H}, \isotope[3]{He} y \isotope[4]{He} presente en la nube protoestelar, así como la metalicidad (Z) de la misma.\par
 
@@
 \section{¿Cómo visualizar los resultados de MESA?}
-\subsection{Visualización con pgstart}
-COMENTAR LA PARTE de pgstart
+\subsection{Visualización con pgstar}
+Observar y analizar el resultado de las simulaciones atendiendo únicamente a los valores numéricos recogidos en los ficheros de salida no es la mejor opción a la hora de hacernos una idea de cómo están yendo nuestras simulaciones.
+Al menos, este es el caso en las etapas iniciales donde nuestra atención está más en saber si las simulaciones se están comportando de la manera esperada, o si por al contrario muestra una evolución no esperada.
+En este contexto, \textit{pgstar} es una herramienta extremadamente útil.
+
+La herramienta \textit{pgstar} se encuentra integrada dentro de MESA y permite, a la vez que nuestra simulación se está ejecutando, la representación visual de un buen número de gráficas\footnote{Visitar https://docs.mesastar.org/en/latest/using\_mesa/using\_pgstar.html para obtener información complementaria.} generadas a partir de los datos de perfiles y valores temporales que se van produciendo.
+Entre éstas se encuentran diagramas HR, de abundacia de elementos químicos, de gravedad superficial vs.
+temperatura efectiva, por citar algunas de las posibilidades.
+La herramienta ofrece una gran flexibilidad a la hora de especificar tamaños de gráfica, número de ellas, fuentes, colores, leyendas y un sin fin de opciones adicionales.
+En el Listado \ref{lst:pg_star_config} se muestra un ejemplo de configuración de la herramienta para producir la gráfica de abundancia mostrada en la Figura \ref{fig:abundances}.
+
+\begin{lstlisting}[language=Fortran, float, caption={Ejemplo de configuración de la herramienta pgstart para la generación de una gráfica de abundacias de elementos químicos.}, label={lst:pg_star_config}]
+  ...	
+  Abundance_win_flag = .true.
+  Abundance_show_photosphere_location = .true.
+  num_abundance_line_labels = 1
+
+  Abundance_num_isos_to_show = 5
+  Abundance_which_isos_to_show(1) = 'h1'
+  Abundance_which_isos_to_show(2) = 'he4'
+  Abundance_which_isos_to_show(3) = 'li7'
+  Abundance_which_isos_to_show(4) = 'be7'
+  Abundance_which_isos_to_show(5) = 'fe56'
+
+  Abundance_log_mass_frac_min = -30
+  Abundance_log_mass_frac_max = 10
+  Abundance_file_interval = 1
+  ...
+
+\end{lstlisting}
 
 \subsection{Visualización con terceras herramientas}
-En determinadas ocasiones, las opciones de visualización ofrecidas por MESA no es la más adecuada o conveniente para nuestros propósitos.
+En determinadas ocasiones, las opciones de visualización ofrecidas por MESA no son las más adecuadas o convenientes para nuestros propósitos.
 Quizás porque no muestra el parámetro en el que estamos interesados, o si lo hace, deseamos otro tipo de formato de visualización.
-Aunque en la mayoría de los casos, la necesidad de recurrir a terceras herramientas para la visualización se debe a que necesitamos realizar algún tipo de preprocesamiento (ver \ref{sec:gemelos_solares} para más detalles) sobre los mismos antes de proceder a su representación e interpretación final.\par
+Aunque en la mayoría de los casos, la necesidad de recurrir a terceras herramientas para la visualización se debe a que necesitamos realizar algún tipo de preprocesamiento (ver Sección \ref{sec:gemelos_solares} para más detalles) sobre los mismos antes de proceder a su representación e interpretación final.\par
 
-En nuestra investigación hemos recurrido a la herramientas Octave (ver \ref{sec:tool_octave}) para la extracción, transformación, filtrado, enriquecimiento y visualización de los datos.
-En Octave establecemos el un flujo de trabajo sobre los ficheros de salida de MESA, cuyas etapas:
+En nuestra investigación hemos recurrido a la herramientas Octave (ver Sección \ref{sec:tool_octave}) para la extracción, transformación, filtrado, enriquecimiento y visualización de los datos.
+En Octave establecemos un flujo de trabajo sobre los ficheros de salida de MESA, cuyas etapas más relevantes son:
 \begin{enumerate}
@@
 \section{Extendiendo MESA - Parte I} \label{mesa_parte1}
-Una vez obtenida una visión relativamente aceptable de las capacidades del simulador, la etapa a considerar es la de obtener un obtener un marco de referencia para nuestras simulaciones.
+Una vez obtenida una visión relativamente aceptable de las capacidades del simulador, la etapa a considerar es la de obtener un marco de referencia para nuestras simulaciones.
 Éste nos debe de asistir en la consecución de los siguientes objetivos:
@@
 
-\begin{itemize}
+\begin{enumerate}
     \item Reacciones nucleares en las que interviene el litio
@@
     \item Presencia de campo magnético de intensidad fija
-\end{itemize}
+\end{enumerate}
 
@@
 
-\begin{itemize}
+\begin{enumerate}
     \item Reacciones nucleares en las que interviene el Li
@@
     \item Procesos de sedimentación
-\end{itemize}
+\end{enumerate}
 
@@
 
-\begin{itemize}
+\begin{enumerate}
     \item Establecer una metalicidad similar al Sol
     \item Detener la simulación al alcanzar la TAMS
-    \item Obtener las abundancias para los elementos H, He, Be
-\end{itemize}
+    \item Obtener las abundancias para los elementos H, He, Li
+\end{enumerate}
 
-La composición química del Sol, incluyendo su metalicidad y masa inicial del modelo queda fijada por medio de los siguientes parámetros (Listado \ref{lst:ini_params}) de MESA:
+La composición química del Sol, incluyendo su metalicidad y masa inicial del modelo queda fijada por medio de los siguientes parámetros (ver Listado \ref{lst:ini_params}) de MESA:
 
@@
 Por tanto, ya tenemos identificas la red de reacciones a activar en la configuración de MESA.
-Es en este punto donde volvemos a hacer referencia al módulo \textit{net} (ver \ref{reac_nuc}).
+Es en este punto donde volvemos a hacer referencia al módulo \textit{net} (ver Sección  \ref{reac_nuc}).
 Si recordamos, se indicaba que MESA viene pre configurado con una red de reacciones nucleares en las que sólo trata con 8 isótopos.
@@
 
-Con los parámetros anteriores incluidos en el proyecto, indicamos a MESA que sustituimos la red de reacciones nucleares por defecto por una nueva denominada \textit{mesa\_49.net} (Listado \ref{lst:mesa49}).
+Con los parámetros anteriores incluidos en el proyecto, indicamos a MESA que sustituimos la red de reacciones nucleares por defecto por una nueva denominada \textit{mesa\_49.net} (ver Listado \ref{lst:mesa49}).
 Las redes son simples ficheros de texto en los que se indica al simulador las reacciones nucleares que debe tener en cuenta y cuál es el aspecto de éstas.
@@
         (MESA) calcula la difusión de partículas y la sedimentación gravitacional resolviendo las ecuaciones de Burger a través del método propuesto por Thoul \citep{Thoul1993}.
-El módulo de difusión trata los elementos presentes en el modelo estelar como pertenecientes a "clases" definidas por el usuario en términos de rangos de masas atómicas.
+El módulo de difusión trata los elementos presentes en el modelo estelar como pertenecientes a clases definidas por el usuario en términos de rangos de masas atómicas.
         }
@@
 Una vez activada la nueva red de reacciones nucleares en nuestro fichero de proyecto MESA, conseguimos obtener las abundancias para el litio, las cuales son informadas en los ficheros de salida estándar que genera MESA.
-Adicionalmente estamos interesados en obtener el perfil de evolución de estos elementos a lo largo del tiempo de evolución de la estrella y en función del radio de la estrella (Figura \ref{fig:abundances}).
+Adicionalmente estamos interesados en obtener el perfil de evolución de estos elementos a lo largo del tiempo de evolución de la estrella y en función del radio de la estrella (ver Figura \ref{fig:abundances}).
 Para ello recurrimos a utilizar las capacidades gráficas de MESA y mostramos la abundancia de estos elementos.
@@
 
-En lo que se refiere a la rotación de las capas superficiales de la estrella, MESA permite especificar una velocidad en km/s, que en nuestro caso ha quedado fijada en 20 km/s al ser un valor similar al que tenemos en el Sol.\par
+En lo que se refiere a la rotación de las capas superficiales de la estrella, MESA permite especificar una velocidad en km/s, que en nuestro caso ha quedado fijada en 20 km/s al ser un valor similar al que tenemos en el Sol.
+En el Listado \ref{lst:li_evol} mostramos la configuración adicional a añadir al proyecto de MESA para poder simular los conceptos teóricos discutidos hasta este momento.
 
-En el Listado \ref{lst:li_evol} mostramos la configuración adicional a añadir al proyecto de MESA para poder simular los conceptos teóricos discutidos hasta este momento.
 \begin{lstlisting}[language=Fortran, float, caption={Parametrización de los procesos relacionados con el tratamiento del litio y su evolución temporal.}, label={lst:li_evol}]
@@
 
-Llegado este punto, hemos conseguido obtener una configuración de proyecto base que recoge las características esenciales de la estrella a simular en MESA: pone en marcha los procesos físicos relacionados con la evolución del litio y es capaz de informar, tanto a través de los ficheros de salida estándar como de las representaciones gráficas, sobre las abundancias de este elemento a lo largo de la evolución de la estrella en sus etapas de PMS y MS.
+Hasta aquí, hemos conseguido obtener una configuración de proyecto base que recoge las características esenciales de la estrella a simular en MESA: pone en marcha los procesos físicos relacionados con la evolución del litio y es capaz de informar, tanto a través de los ficheros de salida estándar como de las representaciones gráficas, sobre las abundancias de este elemento a lo largo de la evolución de la estrella en sus etapas de PMS y MS.
 Esta configuración de proyecto y los valores arrojados por la simulación que lo utiliza, nos servirán para establecer el marco de referencia contra el que comparar los valores obtenidos en las nuevas simulaciones que extienden la funcionalidad de MESA.\par
@@
 
-\begin{itemize}
+\begin{enumerate}
 	\item Definir el formalismo matemático que soporta el proceso a incorporar a MESA
@@
 	\item Realizar las simulaciones
-\end{itemize}
+\end{enumerate}
 
-Los puntos 1 a 4 y 7 del listado anterior serán extensivamente documentados en las partes \ref{parte1} y \ref{parte2}.
+Los puntos 1 a 4, 7 y 8 del listado anterior serán extensivamente documentados en las partes \ref{parte1} y \ref{parte2}.
 En lo que atañe a los puntos 5 y 6, dado que es algo común a ambas partes, lo comentaremos en este apartado.\par
@@
 Lo primero es indicar al simulador que debe utilizar nuestra rutina de par de torsión en lugar de la que trae por defecto MESA.
-Es a través de este \textit{hook} la forma de implementar nuestras rutinas de frenado magnético.
+Es a través de este \textit{hook} como procedemos a implementar nuestra(s) rutina(s) de frenado magnético.
 A continuación listamos los diferentes parámetros de configuración para las diferentes rutinas.
@@
 Por el contrario, si se aplica en la dirección opuesta al giro del objeto, disminuye su velocidad angular.
-En nuestro caso, el frenado magnético producirá una disminución, una toque en sentido contrario al giro de la estrella, provocando en ésta una pérdida de momento angular.\par
+En nuestro caso, el frenado magnético producirá una disminución, una torque en sentido contrario al giro de la estrella, provocando en ésta una pérdida de momento angular.\par
 

### 2024-03-20 project/overleaf/Tesis III/capitulos/capitulo05.tex
WIP

@@
 
-\subsection{Evolución alternativa del Li con MB}
+\subsection{Evolución alternativa del Li con MB} \label{sec:alt_mlt}
 Hasta este punto, las simulaciones de los diferentes modelos se basaron en la parametrización recogida en la Tabla \ref{tab:phy_mesa}, que a su vez fue adoptada de \cite{Choi2016}.

### 2024-04-09 project/octave/plotting_routines.m
Add reference marks to CZ plot

@@ -1636,6 +1636,13 @@ function age_vs_radius_plots(gauss_fields, rotational_vels, is_var_vel, ytick, a
   % Plot sun reference
   plot(sun_age, 0, '*', 'markersize', 15, 'color', [0.5,0.1,0.8]);
 
+  % Plot reference marks
+  line("xdata",[2.5e7,2.5e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", "cyan");
+  line("xdata",[3.5e7,3.5e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", "cyan");
+  line("xdata",[5.4e7,5.4e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", "magenta");
+  line("xdata",[11.2e7,11.2e7], "ydata",[axis_limits(3),axis_limits(4)], "linewidth", 2, "linestyle", ":", "color", "magenta");
+
+
   l = legend(labels, "location", leg_loc);
   set (l, "fontsize", legend_font_size);
   %legend boxoff
@@ -3629,6 +3636,10 @@ function main()
   rot_vels10 = rotational_vels([idx_125crit;idx_13crit;idx_14crit;idx_1475crit;idx_155crit],:);
   rot_vels11 = rotational_vels([idx_12crit;idx_125crit;idx_13crit;idx_14crit;idx_1425crit;],:);
 
+  %paper2
+  p2_rot_vels7 = rotational_vels([idx_12crit;idx_125crit;idx_13crit;idx_14crit;idx_1425crit;],:);
+
+
 
 
 
@@ -3818,8 +3829,10 @@ function main()
 
   %plot_teff_vs_mag_field_XG(rotational_vels([idx_1475crit],:),1);
 
-  paper1();
+  %paper1();
   %paper2();
+  %plot_radius_vs_mag_field_XG_z1(p2_rot_vels7,3);
+  plot_age_vs_teff_XG_z1(rot_vels7,3);
 end
 
 

### 2024-04-09 project/overleaf/Tesis III/capitulos/capitulo02.tex

@@
 Este dato, sin embargo, entra en contradicción con la abundancia de litio medida en enanas viejas de baja metalicidad  (A(Li) $\simeq$ 2.2 dex) situadas en el halo de la galaxia conocido como el Spite plateau \citep{Spite1982}.
-Además, para estrellas de baja metalicidad (\ref{fig:li_abundances_sbbn}) se ha encontrado que la abundancia de litio no parece guardar relación con la metalicidad de la estrella \citep{Fu2015}.\par
+Además, para estrellas de baja metalicidad (ver Figura \ref{fig:li_abundances_sbbn}) se ha encontrado que la abundancia de litio no parece guardar relación con la metalicidad de la estrella \citep{Fu2015}.\par
 

### 2024-04-09 project/overleaf/Tesis III/capitulos/capitulo06.tex
WIP

--- 2024-02-01 project/overleaf/Tesis III/capitulos/capitulo06.tex
@@
+% !TeX root = ../libro.tex
+% !TeX encoding = utf8
+
 \chapter{Rotación y campos magnéticos de intensidad variable}\label{ch:sexto-capitulo}
@@
 \section{Introducción}
---EXPLICAR POR QUÉ OPTAMOS POR LA INTENSIDAD VARIABLE
-
-\section{Efectos del frenado magnético - Parte II}
-
-\subsection{Formalismo de frenado magnético de intesidad fija}
-
-\subsubsection{Frenado magnético según Gallet \& Bouvier (2013)}
+--ELABORAR LA INTRODUCCIÓN. REVISAR PAPER2--
 Suponer que la intensidad del campo magnético de una estrella permanece constante a lo largo de su evolución es una simplificación necesaria en algunos escenarios, pero no se corresponde con las observaciones (INDICAR LAS REFERENCIAS).
@@
 
+--EXPLICAR POR QUÉ OPTAMOS POR LA INTENSIDAD VARIABLE
+La segunda línea de trabajo se encaminó a extender MESA para que soportara la influencia de campos magnéticos y $\amlt$ variable en la evolución de su estrella anfitriona.
+Adicionalmente a lo de la parte 1
+\begin{itemize}
+	\item Presencia de campo magnético de intensidad variable
+	\item Evolución temporal de $\amlt$
+\end{itemize}
+\section{Efectos del frenado magnético - Parte II}
+
+\subsection{Formalismo de frenado magnético de intesidad variable} 
 De acuerdo a \cite{Gallet2013} la intensidad del campo magnético debería variar a lo largo de la vida de la estrella en lugar de permanecer fija durante toda su evolución.
 Para ello nos basamos en un formalismo que nos permite calcular una intensidad de campo magnético variable.
+
+\subsubsection{Frenado magnético según Gallet \& Bouvier (2013)}
 Este modelo adopta una prescripción de tipo dinamo que permite establecer una relación entre $\Omega$ , $\teff$, y la intensidad del campo magnético ($B$), así como con la tasa de pérdida de masa ($\mwind$) causada por el viento estelar.
@@
 
-\begin{ceqn}
 \begin{equation}
@@
 \end{equation}
-\end{ceqn}
 
-donde $f_*$ es el factor de llenado, es decir, la fracción de la superficie estelar que está magnetizada, $B_*$ es la fuerza del campo magnético, $\Omega_*$ es la velocidad angular de la estrella en la superficie estelar, y $b$ es el exponente de la dinamo \cite{Gallet2013}.
+donde $f_*$ es el factor de llenado, es decir, la fracción de la superficie estelar que está magnetizada, $B_*$ es la fuerza del campo magnético, $\Omega_*$ es la velocidad angular de la estrella en la superficie estelar, y $b$ es el exponente de la dinamo \citep{Gallet2013}.
 El producto de $f_*$ por $B_*$ nos permitirá obtener la intensidad del campo magnético.
@@
 \end{equation}
-\end{ceqn}
 
-donde $P_*$ es la presión del gas fotosférico, $\rho_*$ es la densidad fotosférica, $\boltzmann$ es la constante de Boltzmann, $\massH$ es la masa de un átomo de hidrógeno, y $\mu$ el peso atómico medio \cite{Cranmer2011}.
+donde $P_*$ es la presión del gas fotosférico, $\rho_*$ es la densidad fotosférica, $\boltzmann$ es la constante de Boltzmann, $\massH$ es la masa de un átomo de hidrógeno, y $\mu$ el peso atómico medio \citep{Cranmer2011}.
 \par
@@
 \end{equation}
-\end{ceqn}
 
-\cite{Cranmer2011} Crammer proporciona dos ajustes diferentes para $f_*$ que definen respectivamente los límites superior e inferior para la intensidad media del campo magnético superficial ($f_*B_*$),
+\cite{Cranmer2011} proporciona dos ajustes diferentes para $f_*$ que definen respectivamente los límites superior e inferior para la intensidad media del campo magnético superficial ($f_*B_*$),
 
-\begin{ceqn}
 \begin{align}
@@
 \end{align}
-\end{ceqn}
 
@@
 
-Por último, podemos calcular el factor de llenado de una estrella aplicando el ajuste realizado por \cite{Gallet2013} que reproduce más fielmente el factor de llenado medio del Sol actual ($\fsun$ = [0.001-0.01]) ver \cite{Cranmer2011},
-
-\begin{ceqn}
-	\begin{equation}
-		\fstar = \frac{0.55}{(1+(x/0.16)^{2.3})^{1.22}} \label{eq:fstart}
-	\end{equation}
-\end{ceqn}
+Por último, podemos calcular el factor de llenado de una estrella aplicando el ajuste realizado por \cite{Gallet2013} que reproduce más fielmente el factor de llenado medio del Sol actual ($\fsun$ = [0.001-0.01], ver \citep{Cranmer2011}),
 
@@
 
-Una vez que hemos caracterizado cómo evoluciona la intensidad del campo magnético, necesitamos modelizar cómo se produce la pérdida de momento angular.
-En el formalismo anterior utilizamos un modelo aplicable a campos magnéticos de intensidad no variable y de baja intensidad ($1,0\,\Gauss \leq B \leq 10,0\,\Gauss$).
-Como se muestra en la Figura \ref{fig:mag_field_limits_var_vel_g3}, las simulaciones con el presente formalismo arrojan valores de campo magnético que muestran intensidades máximas en torno a $1,0K\,\Gauss$, lo que hace inviable el uso del modelo empleado anteriormente.
-\par
-
+--QUIZAS SACARLO A UNA SECCIÓN PROPIA DE AML --
 Una vez que hemos caracterizado cómo evoluciona la intensidad del campo magnético, necesitamos modelar cómo se produce la pérdida de momento angular.
-A medida que la estrella se contrae durante la fase de pre-secuencia principal (PMS) hacia el comienzo de la secuencia principal (ZAMS), su velocidad angular aumenta, lo que conduce a un aumento gradual del campo magnético medio.
+A medida que la estrella se contrae durante la fase de PMS hacia el comienzo de la ZAMS, su velocidad angular aumenta, lo que conduce a un aumento gradual del campo magnético medio.
 En nuestro análisis, observamos intensidades máximas de hasta $1.0,\mathrm{kG}$, como se muestra en la figura \ref{fig:mag_field_limits_var_vel_g3}.
@@
 
+--BLOQUE REPETIDO. HACER REFERENCIA A LA SECCIÓN ANTERIOR, INDICANDO QUE A DIFERENCIA DE ELLA, AQUÍ UTILIZAMOS R-ALVEN--
 Se ha observado que las estrellas de masa intermedia fuertemente magnéticas suelen tener tasas de rotación mucho más lentas que otras estrellas de su población parental \cite{Mathys2006}.
@@
 \end{equation}
-\end{ceqn}
 
-donde $\Bp$ es la intensidad del campo magnético en la superficie estelar, $\vesc$ es la velocidad de escape, $m = 0.1675$, $K_1 = 1.30$ y $K_2 = 0.0506$ \cite{Gallet2013}.
+donde $\Bp$ es la intensidad del campo magnético en la superficie estelar, $\vesc$ es la velocidad de escape, $m = 0.1675$, $K_1 = 1.30$ y $K_2 = 0.0506$ \citep{Gallet2013}.
 
-\begin{ceqn}
 \begin{align}
@@
 
+--HASTA AQUÍ--
+
 \section{Efectos de la longitud de mezcla}
-La teoría de la longitud de mezcla (MLT) introducida por Böhm-Vitense (BV) ha sido comúnmente adoptada para modelar la convección estelar en códigos de evolución estelar, incluyendo MESA.
+La teoría de la longitud de mezcla (Mixing Length Theory, MLT) introducida por \citet{BohmVit58} ha sido comúnmente adoptada para modelar la convección estelar en códigos de evolución estelar, incluyendo MESA.
 El parámetro libre más importante en esta teoría es la longitud de mezcla $l$ que se define como:
 
-\begin{ceqn}
 \begin{equation}
@@
 Además, no existen argumentos sólidos que sugieran que el parámetro de longitud de mezcla sea el mismo en todas las estrellas y en todas las fases evolutivas \cite{Pasetto2014}.
-En \cite{Navarro2020} mostramos cómo una parametrización alternativa de la longitud de mezcla (ML) podría producir resultados en línea con las observaciones.
-El ajuste del sobreimpulso (over-shooting) de ML y del parámetro libre $\amlt$ parece ser necesario para explicar las abundancias de Li en los OC's.\par
+En la Sección \ref{sec:alt_mlt} mostramos cómo una parametrización alternativa de la longitud de mezcla (ML) podría producir resultados en línea con las observaciones.
+El ajuste del overshooting de ML y del parámetro libre $\amlt$ parece ser necesario para explicar las abundancias de litio en los OC's.\par
 
@@
 \subsection{MLT $\alpha$ variable según Sonoi et al. (2019)}
-En \cite{Sonoi2018}, los autores introdujeron una forma de calibrar $\amlt$ para estrellas de tipo solar en la que su valor aumenta al aumentar $\gsurf$ o disminuir $\teff$.
-Calibraron valores de $\amlt$ para varios modelos 3D simulados usando el código CO$^5$BOLD \cite{Freytag2011} y los transfirieron a modelos 1D desarrollados por \cite{Ludwig1998}.
+En \cite{Sonoi2018}, los autores introdujeron una forma de calibrar $\amlt$ para estrellas de tipo solar en la que su valor directamente proporcional a la $\gsurf$, e inversamente proporcional a la $\teff$.
+--CREAN UN SIMPLE EXCEL MOSTRANDO ESTO-- Calibraron diferentes valores de $\amlt$ para varios modelos 3D simulados usando el código CO$^5$BOLD \citep{Freytag2011} y los transfirieron a modelos 1D desarrollados por \cite{Ludwig1998}.
 La parametrización que obtuvieron fue:
 
-\begin{ceqn}
 \begin{equation}
@@
 
-\begin{ceqn}
 \begin{align}
@@
 \end{align}
-\end{ceqn}
 
-y $a_i$ son los coeficientes resultantes de la función de ajuste \ref{eq:alpha_ml} a los valores $\alpha$ calibrados para el modelo de convección MLT \cite{Sonoi2018}: $a_0=1.790295$, $a_1=-0.14954$, $a_2=0.069574$, $a_3=-0.00829$, $a_4=0.013165$, $a_5=0.080333$, $a_6=-0.03306$.
+y $a_i$ son los coeficientes resultantes de la función de ajuste \ref{eq:alpha_ml} a los valores $\alpha$ calibrados para el modelo de convección MLT \citep{Sonoi2018}: $a_0=1.790295$, $a_1=-0.14954$, $a_2=0.069574$, $a_3=-0.00829$, $a_4=0.013165$, $a_5=0.080333$, $a_6=-0.03306$.
 \par
---COMENTAR MÁS EN DETALLE CÓMO POR EJEMPLO SE HACE VARIAR EN MESA. TAMBIÉN ALGO MÁS DEL FORMALISMO--
+--COMENTAR MÁS EN DETALLE CÓMO, POR EJEMPLO, SE HACE VARIAR EN MESA. TAMBIÉN ALGO MÁS DEL FORMALISMO--
 
 \section{Extendiendo MESA - Parte III}
-\subsection{Rutina de frenado magnético de intensidad variable}
-EXPLICAR QUE SUSTITUIMOS LA RUTINA POR DEFECTO CON LA NUESTRA
-\begin{lstlisting}[language=Fortran, caption={Rutina de par de torsión.}, label={lst:jdot_gb}]
+\subsection{Rutina de frenado magnético de intensidad variable} \label{sec:mb_var_b}
+En la Sección \ref{sec:mb_cte_b} comentamos los detalles de como MESA conserva el AM entre las diferentes celdas $k$ y les asigna un valor de AM $J_k$.
+Este mecanismo sigue siendo el mismo para esta nueva rutina.
+En este nuevo escenario la diferencia radica en la nueva rutina de MB (ver Listado \ref{lst:jdot_gb}), concretamente en la implementación de la misma según el nuevo formalismo que hemos introducido para campos magnéticos de intensidad variable, y especialmente en la rutina que calcula la intensidad del campo magnético, devolviendo (potencialmente) un valor diferente a cada paso de la simulación (ver Listado \ref{lst:jdot_mf_intensity}).\par
+
+\begin{lstlisting}[language=Fortran, float, caption={Rutina de par de torsión para un campo magnético de intensidad variable.}, label={lst:jdot_gb}]
   real function calculate_jdot_rate_gb(s, bf_star) result(new_j_dot)
@@
 
-	!eq 3 [1]
 	v_esc = (2 * standard_cgrav * m_st / r_st)**0.5
@@
 	alfven_r_num = bf_star**2 * r_st**2
-
 	alfven_r_den = m_dot*((k2**2 * v_esc**2)+(omega_surf**2 * r_st**2))**0.5
-
 	alfven_r = k1 * (alfven_r_num / alfven_r_den)**m * r_st
@@
   end function	
-
 \end{lstlisting}
@@
 \subsection{Rutina de cálculo de intensidad de campo magnético}
-EXPLICAR QUE SUSTITUIMOS LA RUTINA POR DEFECTO CON LA NUESTRA
-\begin{lstlisting}[language=Fortran, caption={Rutina de cálculo de intensidad de campo magnético.}, label={lst:jdot_mf_intensity}]
+La incorporación del formalismo para calcular una intensidad de campo magnético variable representa una mejora sustancial al modelo presentado en el capítulo \ref{ch:tercer-capitulo}.
+Esta intensidad es funcionalmente dependiente del tiempo, de $\rho$, $\teff$, y $\Omega$, valores que podemos obtener directamente de MESA.\par
+
+\begin{lstlisting}[language=Fortran, float, caption={Rutina de cálculo de intensidad de campo magnético.}, label={lst:jdot_mf_intensity}]
   real function calculate_mag_field_intensity_gb(s) result(new_b)
@@
 	!exp ->  natural exponential function
-	! Eq 36 [2]
 	tau_c = 314.241*exp(-s% Teff/1952.5)*exp(-(s% Teff/6250)**18) + 0.002 
@@
 	!magnetic filling factor fmin & fmax
-	!eq 9 [1]
 	f_min = 0.5 / (1. + (rossby_norm / 0.16)**2.6)**1.3 
-	!eq 10 [1]
 	f_max = 1 / (1. + (rossby_norm / 0.31)**2.5)
-	!f_star = f_min
-	!eq 11 [1]
 	f_star = 0.55 / (1. + (rossby_norm / 0.16)**2.3)**1.22
@@
 	p_photo = s%rho(s% photosphere_cell_k)*boltzm*s% Teff/(mu_avg*m_hydrogen)
-	!Eq 8 [1]
 	!Equipartition magnetic filed strength
@@
 \subsection{Rutina de $\amlt$ variable}
-EXPLICAR QUE LOS DOS TIPOS DE RUTINAS: INTESIDAD FIJA, INTENSIDAD VARIABLE
-\begin{lstlisting}[language=Fortran, caption={Rutina de $\amlt$ variable.}, label={lst:var_amlt}]
-	!Calibration of the mixing-length parameter alpha for the MLT
-	!and FST models by matching with CO5BOLD models
+Adicionalmente, y en línea con el objetivo de reducir el número de parámetros fijos, hemos procedido a calcular, para cada paso temporal de la simulación de MESA, un valor de $\amlt$, eliminando así la necesidad de prefijar su valor para toda la simulación.\par
+
+\begin{lstlisting}[language=Fortran, float, caption={Rutina de $\amlt$ variable.}, label={lst:var_amlt}]
   real function calculate_alpha(s) result(new_alpha)
@@
 
-	!Equation 4
 	new_alpha = a0 + (a1 + (a3 + a5*x + a6*y)*x + a4*y)*x + a2*y
@@
 De este modo conseguimos que en la siguiente iteración, utilice el valor que hemos calculado.
-\begin{lstlisting}[language=Fortran, caption={Rutina de MESA para ejecutar lógica adicional tras la ejecución de un paso de simulación.}, label={lst:extra_start_step}]
+\begin{lstlisting}[language=Fortran, float, caption={Rutina de MESA para ejecutar lógica adicional tras la ejecución de un paso de simulación.}, label={lst:extra_start_step}]
   integer function extras_start_step(id, id_extra)

### 2024-04-09 project/overleaf/Tesis III/capitulos/to be reviewed.tex

aaaa \gls{lcm} \acrshort{gcd} \Gls{latex}

---ESTO PARECE ESTAR FUERA DE SITIO --

Los denominados modelos estándar, modelos que incluyen la convección sólo como proceso de mezcla y no consideran otras opciones de transporte como la difusión o la pérdida de momento angular (Angular Moment Loss, AML), han sido los principales implicados en la elaboración de estas predicciones \cite{Sestito2005}. El Li se destruye a una temperatura $\tli \approx 2.5 x 10^6\; K$ cuando un átomo de Li colisiona con un protón produciendo dos átomos de He, algo que tiene lugar durante las reacciones protón-protón tipo II (P-P II), y por tanto se destruye directamente en las envolturas estelares cuando la temperatura en la base de la zona de convección (Base Convective Zone, BCZ) alcanza ese valor. El Sol en particular, y las estrellas de tipo solar en general, se caracterizan por tener una CZ que cubre gran parte del radio estelar durante la PMS lo que hace que su límite inferior supere $\tli$ \cite{Iben1965}. Este agotamiento se detiene durante la aproximación al inicio de la MS (Zero-Age Main Sequence, ZAMS) cuando la zona de convección retrocede y la temperatura en la BCZ es más fría que $\tli$. En los modelos estelares estándar sólo la masa y la composición química inicial determinan a qué distancia de la superficie estelar se alcanza la temperatura $\tli$, por lo que se espera que estrellas dentro de cúmulos con masa similar alcancen la ZAMS con iguales abundancias superficiales de Li. Además, también deberían mostrar una evolución del Li muy similar hasta su aproximación a la edad terminal de la MS (Terminal-Age Main Sequence, TAMS). Durante este mismo periodo los mecanismos de convección desencadenan un proceso de mezcla que homogeneiza la composición química de la envoltura convectiva, desde su límite inferior hasta la superficie estelar. Sin embargo, se han observado diferentes abundancias de Li en distintas poblaciones estelares \cite[][y referencias en las mismas]{Somers2014}.\par

-- HASTA AQUÍ --
